---
title: "Growth Curve Models"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(ggplot2)
set.seed(2024)
```

# Growth Curve Models

## Trajectories as the Target

Standard repeated measures models ask: "Is the mean outcome different across time points?" Growth curve models ask something richer: "How does each unit's trajectory unfold over time, and what explains variation in those trajectories?"

The shift in framing is fundamental. We are not trying to estimate one average change — we are trying to recover the distribution of individual change processes that generated the data. Each person has their own DGP; growth curve models estimate the parameters of that distribution.

## The Unconditional Growth Model

Start with no predictors. Just time and people.

```{r unconditional-growth}
n_people <- 60
n_times  <- 6  # weeks

# Each person has a random baseline (intercept) and random rate of change (slope)
person_intercepts <- rnorm(n_people, 50, 8)
person_slopes     <- rnorm(n_people, -2, 1.5)  # average decline of 2/week

df_growth <- expand_grid(person = 1:n_people, week = 0:(n_times-1)) |>
  mutate(
    stress = person_intercepts[person] +
             person_slopes[person] * week +
             rnorm(n(), 0, 4),
    person = factor(person)
  )

# Plot individual trajectories
sample_people <- sample(levels(df_growth$person), 12)

ggplot(df_growth |> filter(person %in% sample_people),
       aes(x = week, y = stress, group = person, color = person)) +
  geom_line(alpha = 0.7) + geom_point(size = 1.5, alpha = 0.6) +
  geom_smooth(data = df_growth, aes(group = NULL, color = NULL),
              method = "lm", color = "black", linewidth = 1.5, se = TRUE) +
  scale_color_viridis_d() +
  labs(title = "Growth curve: individual stress trajectories over 6 weeks",
       subtitle = "Colored lines = 12 individuals; black line = average trajectory",
       x = "Week", y = "Stress (0-100)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

```{r fit-growth}
# Unconditional growth model
m_growth <- lmer(stress ~ week + (week | person), data = df_growth)
summary(m_growth)
```

```{r growth-variance}
broom.mixed::tidy(m_growth, effects = "ran_pars") |>
  knitr::kable(digits = 3,
    caption = "Variance components: how much do people differ in their starting point and rate of change?")
```

**Reading the output:**
- Fixed effect of `week` (≈ −2): the average rate of stress decline per week
- Random intercept variance: how much people differ in their starting stress level
- Random slope variance: how much people differ in their rate of change
- Intercept-slope correlation: do people who start high also decline faster (or slower)?

## Adding Predictors of Growth

The power of growth curve models comes from asking: **what predicts the trajectory?** We can add level-2 predictors to explain why some people's intercepts and slopes differ from others.

```{r growth-predictors}
# Add a predictor: treatment condition
df_growth <- df_growth |>
  mutate(treatment = rep(rbinom(n_people, 1, 0.5), each = n_times))

# Treatment might affect both initial level and rate of change
m_growth_pred <- lmer(stress ~ week * treatment + (week | person),
                      data = df_growth)

broom.mixed::tidy(m_growth_pred, effects = "fixed") |>
  knitr::kable(digits = 3,
    caption = "Growth model with treatment: interaction = different slopes by condition")
```

The `week:treatment` interaction coefficient tells you whether the treatment group's stress declines faster or slower than the control group.

## Nonlinear Growth

Some processes do not change linearly. Symptoms might improve rapidly at first, then plateau. Performance curves often show diminishing returns. The growth curve framework accommodates nonlinearity.

```{r nonlinear-growth}
# Quadratic growth
df_growth_q <- expand_grid(person = 1:n_people, week = 0:8) |>
  mutate(
    person_int    = rep(rnorm(n_people, 50, 8), each = 9),
    person_lin    = rep(rnorm(n_people, -5, 2), each = 9),
    person_quad   = rep(rnorm(n_people, 0.5, 0.3), each = 9),
    stress        = person_int + person_lin * week + person_quad * week^2 + rnorm(n(), 0, 3),
    person        = factor(person)
  )

# Fit with quadratic time
m_quad <- lmer(stress ~ week + I(week^2) + (week | person), data = df_growth_q)

# Plot average predicted trajectory
df_growth_q <- df_growth_q |>
  mutate(fitted_stress = fitted(m_quad))

df_growth_q |>
  group_by(week) |>
  summarise(mean_obs = mean(stress), mean_fit = mean(fitted_stress)) |>
  pivot_longer(-week) |>
  ggplot(aes(x = week, y = value, color = name, linetype = name)) +
  geom_line(linewidth = 1.2) + geom_point(size = 2.5) +
  scale_color_manual(values = c("#aec6e8", "#1e3a5f"),
                     labels = c("Fitted (quadratic)", "Observed")) +
  scale_linetype_manual(values = c("solid", "dashed"),
                        labels = c("Fitted (quadratic)", "Observed")) +
  labs(title = "Quadratic growth: curve captures the deceleration",
       x = "Week", y = "Mean Stress", color = NULL, linetype = NULL) +
  theme_minimal(base_size = 13)
```

::: {.callout-warning icon=false}
## Assumption: The Growth Function is Correctly Specified

**What it means:** Linear growth assumes constant rate of change. Quadratic growth assumes a parabolic trajectory. If the true trajectory is neither — e.g., a logarithmic curve or piecewise linear — your model will be misspecified.

**How to check:** Plot the observed means by time point. Does the pattern look linear? Is there a clear deceleration or acceleration that requires a curve? Use AIC to compare linear vs. quadratic models.
:::

::: {.callout-warning icon=false}
## Assumption: Random Slope for Time is Correctly Specified

**What it means:** Allowing only a random intercept (but not slope) for time assumes everyone changes at exactly the same rate. This is almost always wrong for developmental or therapeutic outcomes.

**How to check:** Compare the random-intercept-only model to the random-slope model using a likelihood ratio test. Plot individual trajectories — if they clearly fan out or diverge, a random slope is needed.
:::
