---
title: "Growth Curve Models"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(ggplot2)
set.seed(2024)
```

# Growth Curve Models

::: {.callout-note icon=false}
## The Growth Curve Question

> **What is the scientific question?** Growth curve models are for questions about *trajectories* — how fast something changes, whether change rates differ across people, and whether a predictor shifts the rate of change. If your question is only about mean levels at specific waves, a simpler repeated-measures model suffices.
>
> **What does this model assume about time?** That each person has their own trajectory described by an intercept (their level at time = 0) and a slope (their rate of change). The model estimates the distribution of those intercepts and slopes across people.
>
> **What goes wrong if you ignore individual trajectory variation?** A random-intercept-only model forces everyone to change at the same rate. If people genuinely differ in how fast they change — some improving rapidly, others plateauing — the model misspecifies the DGP, produces poor predictions for individuals, and may bias the average slope estimate.
:::

## Before Fitting: Three Decisions to Make First

### Decision 1 — Is a growth curve model the right tool?

Growth curve models answer: *"How does the outcome trajectory unfold, and what explains variation in trajectories?"* They are not the right tool for every longitudinal question.

| Your question | Right tool |
|:--------------|:-----------|
| Is the group mean higher at wave 3 than wave 1? | RM-ANOVA or mixed model with time as a factor |
| Does a treatment shift the level at one follow-up wave? | Mixed model with treatment × wave interaction |
| How fast does the outcome change, and do people differ in their rate? | **Growth curve model** |
| Does a treatment affect the *rate* of change, not just the level? | **Growth curve model with treatment × slope** |

### Decision 2 — What does time = 0 mean?

The intercept in a growth model is the predicted outcome when time = 0. Time coding determines what the intercept represents — choose it for interpretability.

| Time coding | Intercept means | Best for |
|:------------|:----------------|:---------|
| 0 = first wave | Baseline level | Intervention studies where pre-treatment level matters |
| 0 = midpoint | Mid-study level | Reduces intercept-slope correlation; improves estimation |
| 0 = last wave | Final-wave level | When the endpoint is the primary reference |
| 0 = age (individually varying) | Level at a specific age | Developmental and aging studies |

### Decision 3 — How many waves, and are they equally spaced?

| Waves | What you can reliably estimate |
|:------|:-------------------------------|
| 2 | Intercept + one slope. Growth curve = change score. No information about curvature. |
| 3 | Linear growth. Minimal test for non-linearity. |
| 4+ | Quadratic growth, piecewise change, or splines. Recommended for detecting trajectory shape. |
| Unequal spacing | Use exact elapsed time as the time variable — not wave number. |

With only 2 waves, a growth curve model adds nothing beyond a difference score. Fit a simpler model.

---

## Trajectories as the Target

Standard repeated measures models ask: "Is the mean outcome different across time points?" Growth curve models ask something richer: "How does each unit's trajectory unfold over time, and what explains variation in those trajectories?"

The shift in framing is fundamental. We are not trying to estimate one average change — we are trying to recover the distribution of individual change processes that generated the data. Each person has their own DGP; growth curve models estimate the parameters of that distribution.

## Pre-Fit Checks: Always Plot Before Fitting

```{r pre-fit-checks, fig.width=11, fig.height=4}
library(patchwork)

# Spaghetti plot: individual trajectories
sample_ids <- sample(levels(df_growth$person), 15)

p_spaghetti <- df_growth |>
  filter(person %in% sample_ids) |>
  ggplot(aes(x = week, y = stress, group = person)) +
  geom_line(alpha = 0.45, color = "#1e3a5f") +
  geom_point(size = 1.5, alpha = 0.5, color = "#1e3a5f") +
  stat_summary(aes(group = NULL), fun = mean, geom = "line",
               color = "#e63946", linewidth = 1.5) +
  labs(title = "Spaghetti plot: individual trajectories\n(red = cohort average)",
       subtitle = "Fan-out = slopes differ across people → random slope needed",
       x = "Week", y = "Stress score") +
  theme_minimal(base_size = 11)

# Mean-by-wave with error bars
p_means <- df_growth |>
  group_by(week) |>
  summarise(m = mean(stress), se = sd(stress) / sqrt(n()), .groups = "drop") |>
  ggplot(aes(x = week, y = m, ymin = m - 1.96*se, ymax = m + 1.96*se)) +
  geom_line(linewidth = 1.1, color = "#1e3a5f") +
  geom_errorbar(width = 0.15, color = "#1e3a5f") +
  geom_point(size = 3, color = "#1e3a5f") +
  labs(title = "Mean ± 95% CI by wave",
       subtitle = "Flat line ≈ no growth; curve ≈ nonlinear trajectory",
       x = "Week", y = "Mean stress") +
  theme_minimal(base_size = 11)

# Person-level change: connect each person's first and last observation
p_change <- df_growth |>
  group_by(person) |>
  summarise(start = first(stress), end = last(stress), .groups = "drop") |>
  mutate(direction = if_else(end < start, "Declined", "Increased or flat")) |>
  ggplot(aes(x = start, y = end - start, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual(values = c("Declined" = "#2d6a4f", "Increased or flat" = "#cb181d")) +
  labs(title = "Individual change: first → last wave",
       subtitle = "Spread = heterogeneous slopes; correlation with start = need to model",
       x = "Starting stress", y = "Change (last − first)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_spaghetti | p_means | p_change
```

**What to look for:**

- **Spaghetti plot:** Do trajectories fan out over time? Wide fan-out signals that individuals differ in their rate of change — a random slope is needed.
- **Mean-by-wave:** Does the average look linear? If the means accelerate, decelerate, or level off, a linear model will be misspecified.
- **Individual change plot:** Is the amount of change correlated with the starting level? If people who start high show larger declines, the intercept-slope correlation will be negative.

## The Unconditional Growth Model

Start with no predictors. Just time and people.

```{r unconditional-growth}
n_people <- 60
n_times  <- 6  # weeks

# Each person has a random baseline (intercept) and random rate of change (slope)
person_intercepts <- rnorm(n_people, 50, 8)
person_slopes     <- rnorm(n_people, -2, 1.5)  # average decline of 2/week

df_growth <- expand_grid(person = 1:n_people, week = 0:(n_times-1)) |>
  mutate(
    stress = person_intercepts[person] +
             person_slopes[person] * week +
             rnorm(n(), 0, 4),
    person = factor(person)
  )

# Plot individual trajectories
sample_people <- sample(levels(df_growth$person), 12)

ggplot(df_growth |> filter(person %in% sample_people),
       aes(x = week, y = stress, group = person, color = person)) +
  geom_line(alpha = 0.7) + geom_point(size = 1.5, alpha = 0.6) +
  geom_smooth(data = df_growth, aes(group = NULL, color = NULL),
              method = "lm", color = "black", linewidth = 1.5, se = TRUE) +
  scale_color_viridis_d() +
  labs(title = "Growth curve: individual stress trajectories over 6 weeks",
       subtitle = "Colored lines = 12 individuals; black line = average trajectory",
       x = "Week", y = "Stress (0-100)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

```{r fit-growth}
# Unconditional growth model
m_growth <- lmer(stress ~ week + (week | person), data = df_growth)
summary(m_growth)
```

```{r growth-variance}
broom.mixed::tidy(m_growth, effects = "ran_pars") |>
  knitr::kable(digits = 3,
    caption = "Variance components: how much do people differ in their starting point and rate of change?")
```

**Reading the output:**
- Fixed effect of `week` (≈ −2): the average rate of stress decline per week
- Random intercept variance: how much people differ in their starting stress level
- Random slope variance: how much people differ in their rate of change
- Intercept-slope correlation: do people who start high also decline faster (or slower)?

## Adding Predictors of Growth

The power of growth curve models comes from asking: **what predicts the trajectory?** We can add level-2 predictors to explain why some people's intercepts and slopes differ from others.

```{r growth-predictors}
# Add a predictor: treatment condition
df_growth <- df_growth |>
  mutate(treatment = rep(rbinom(n_people, 1, 0.5), each = n_times))

# Treatment might affect both initial level and rate of change
m_growth_pred <- lmer(stress ~ week * treatment + (week | person),
                      data = df_growth)

broom.mixed::tidy(m_growth_pred, effects = "fixed") |>
  knitr::kable(digits = 3,
    caption = "Growth model with treatment: interaction = different slopes by condition")
```

The `week:treatment` interaction coefficient tells you whether the treatment group's stress declines faster or slower than the control group.

## Choosing the Functional Form for Time

The most consequential decision in a growth curve model is how time enters — linear, quadratic, piecewise, or something else. **Always choose functional form from the data, not from a significance test.**

```{r functional-form-comparison}
# Fit linear growth to the quadratic dataset for comparison
m_linear_q <- lmer(stress ~ week + (week | person), data = df_growth_q,
                   control = lmerControl(optimizer = "bobyqa"))
m_quad     <- lmer(stress ~ week + I(week^2) + (week | person), data = df_growth_q,
                   control = lmerControl(optimizer = "bobyqa"))

anova(m_linear_q, m_quad) |>
  knitr::kable(digits = 3,
    caption = "AIC/BIC and LRT: does adding quadratic time improve fit?")
```

```{r functional-form-plot, fig.width=8, fig.height=3.5}
df_growth_q |>
  mutate(fit_lin  = predict(m_linear_q),
         fit_quad = predict(m_quad)) |>
  group_by(week) |>
  summarise(Observed = mean(stress),
            `Linear`  = mean(fit_lin),
            `Quadratic` = mean(fit_quad), .groups = "drop") |>
  pivot_longer(-week, names_to = "series", values_to = "stress") |>
  mutate(series = factor(series, levels = c("Observed", "Linear", "Quadratic"))) |>
  ggplot(aes(x = week, y = stress, color = series, linetype = series)) +
  geom_line(linewidth = 1.2) + geom_point(size = 2.5) +
  scale_color_manual(values = c(Observed = "#555", Linear = "#cb181d",
                                 Quadratic = "#2d6a4f")) +
  scale_linetype_manual(values = c(Observed = "solid", Linear = "dashed",
                                    Quadratic = "solid")) +
  labs(title = "Functional form: observed vs. linear vs. quadratic fit",
       subtitle = "The pre-fit mean plot already showed curvature — AIC confirms quadratic is needed",
       x = "Week", y = "Mean stress", color = NULL, linetype = NULL) +
  theme_minimal(base_size = 12)
```

**Decision rule:** use the more complex form only when (1) the pre-fit mean-by-wave plot shows curvature AND (2) AIC/BIC clearly favor the more complex model. A quadratic term that just reaches significance with no visible curvature in the pre-fit plot is likely a chance artefact.

## Nonlinear Growth

Some processes do not change linearly. Symptoms might improve rapidly at first, then plateau. Performance curves often show diminishing returns. The growth curve framework accommodates nonlinearity.

```{r nonlinear-growth}
# Quadratic growth
df_growth_q <- expand_grid(person = 1:n_people, week = 0:8) |>
  mutate(
    person_int    = rep(rnorm(n_people, 50, 8), each = 9),
    person_lin    = rep(rnorm(n_people, -5, 2), each = 9),
    person_quad   = rep(rnorm(n_people, 0.5, 0.3), each = 9),
    stress        = person_int + person_lin * week + person_quad * week^2 + rnorm(n(), 0, 3),
    person        = factor(person)
  )

# Fit with quadratic time
m_quad <- lmer(stress ~ week + I(week^2) + (week | person), data = df_growth_q)

# Plot average predicted trajectory
df_growth_q <- df_growth_q |>
  mutate(fitted_stress = fitted(m_quad))

df_growth_q |>
  group_by(week) |>
  summarise(mean_obs = mean(stress), mean_fit = mean(fitted_stress)) |>
  pivot_longer(-week) |>
  ggplot(aes(x = week, y = value, color = name, linetype = name)) +
  geom_line(linewidth = 1.2) + geom_point(size = 2.5) +
  scale_color_manual(values = c("#aec6e8", "#1e3a5f"),
                     labels = c("Fitted (quadratic)", "Observed")) +
  scale_linetype_manual(values = c("solid", "dashed"),
                        labels = c("Fitted (quadratic)", "Observed")) +
  labs(title = "Quadratic growth: curve captures the deceleration",
       x = "Week", y = "Mean Stress", color = NULL, linetype = NULL) +
  theme_minimal(base_size = 13)
```

::: {.callout-warning icon=false}
## Assumption: The Growth Function is Correctly Specified

**What it means:** Linear growth assumes constant rate of change. Quadratic growth assumes a parabolic trajectory. If the true trajectory is neither — e.g., a logarithmic curve or piecewise linear — your model will be misspecified.

**How to check:** Plot the observed means by time point. Does the pattern look linear? Is there a clear deceleration or acceleration that requires a curve? Use AIC to compare linear vs. quadratic models.
:::

## Diagnosing Convergence Problems

Growth curve models with random slopes frequently produce convergence warnings or singular fits. These are not cosmetic — they indicate the model is over-complex for the data.

```{r convergence-check}
# Check for singularity — near-zero variances or boundary correlations
isSingular(m_growth)
isSingular(m_growth_pred)

# Inspect the random effects structure
as.data.frame(VarCorr(m_growth)) |>
  select(grp, var1, var2, sdcor) |>
  knitr::kable(digits = 4,
    caption = "Random effects: check for near-zero SDs or correlations at ±1")
```

| Warning sign | What it means | What to do next |
|:-------------|:--------------|:----------------|
| `isSingular()` is TRUE | Model too complex for the data | Remove random slope or fix correlation to 0 |
| Random slope variance ≈ 0 | Individuals do not differ in their rate of change | Remove `(week \| person)`, keep `(1 \| person)` |
| Intercept-slope correlation = ±1 | Correlation not identified | Use `(week \|\| person)` to fix correlation at 0 |
| "Model failed to converge" | Optimizer stopped early | Add `control = lmerControl(optimizer = "bobyqa")` |
| `allFit()` gives different estimates across optimizers | Unstable likelihood surface | Report estimates from the optimizer that converges; simplify model |

## Post-Fit Diagnostics

After fitting, check whether residuals show remaining time structure. Remaining pattern means the model has not fully captured the trajectory.

```{r post-fit-diagnostics, fig.width=9, fig.height=3.5}
df_growth <- df_growth |>
  mutate(resid_g = residuals(m_growth))

p_resid_wave <- df_growth |>
  group_by(week) |>
  summarise(mr = mean(resid_g), .groups = "drop") |>
  ggplot(aes(x = week, y = mr)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_line(color = "#e63946", linewidth = 1.1) +
  geom_point(color = "#e63946", size = 2.5) +
  labs(title = "Mean residuals by wave",
       subtitle = "Remaining pattern = wrong functional form",
       x = "Week", y = "Mean residual") +
  theme_minimal(base_size = 11)

p_resid_person <- df_growth |>
  group_by(person) |>
  summarise(mr = mean(resid_g), ms = mean(stress), .groups = "drop") |>
  ggplot(aes(x = ms, y = mr)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(alpha = 0.4, color = "#1e3a5f", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#1e3a5f", linewidth = 0.8) +
  labs(title = "Person-mean residuals vs. mean outcome",
       subtitle = "Slope ≠ 0 = random effects not capturing person differences",
       x = "Person mean stress", y = "Person mean residual") +
  theme_minimal(base_size = 11)

p_resid_wave | p_resid_person
```

## Assumption Table

```{r assumption-table, echo=FALSE}
knitr::kable(
  tibble(
    Assumption = c(
      "Correct functional form for time",
      "Individual variation in rate of change (random slope)",
      "Normal distribution of random effects",
      "Independence of within-person residuals",
      "Random effects independent of predictors"
    ),
    `Plain-language meaning` = c(
      "The model's time equation matches the actual trajectory shape",
      "People differ in how fast they change, not just in their starting level",
      "Person-level intercepts and slopes are drawn from a bivariate normal distribution",
      "After removing the individual trajectory, residuals are not autocorrelated",
      "Who starts high or changes fast is unrelated to treatment or other covariates"
    ),
    `What to inspect` = c(
      "Mean-by-wave plot; residuals by wave after fitting",
      "Spaghetti plot (do trajectories fan out?); LRT comparing RE structures",
      "Q-Q plot of estimated random effects (BLUPs)",
      "ACF of within-person residuals; residuals by wave",
      "Correlate BLUPs with treatment or baseline predictors"
    ),
    `Failure looks like` = c(
      "Systematic curve in mean-by-wave residuals; AIC favors more complex time model",
      "Singular fit; near-zero slope variance; visible divergence of individual lines",
      "Heavy tails or bimodality in the BLUP distribution",
      "Significant ACF at lag 1 in within-person residuals",
      "Estimated BLUPs correlate with treatment group or baseline covariate"
    ),
    `What to do next` = c(
      "Add quadratic term; piecewise model at a theoretically motivated wave; splines",
      "Add `(week | person)`; if singular, consider whether slope variation is theoretically expected",
      "With n > 30 clusters, MLM is robust to mild non-normality; use robust SEs for severe violations",
      "Add AR(1) residual structure via `nlme::lme(correlation = corAR1())`",
      "If randomized: not a concern. If observational: center predictors; consider person FE"
    )
  ),
  caption = "Growth curve model assumption checklist"
)
```

::: {.callout-warning icon=false}
## Assumption: Random Slope for Time is Correctly Specified

**What it means:** Allowing only a random intercept (but not slope) for time assumes everyone changes at exactly the same rate. This is almost always wrong for developmental or therapeutic outcomes.

**How to check:** Compare the random-intercept-only model to the random-slope model using a likelihood ratio test. Plot individual trajectories — if they clearly fan out or diverge, a random slope is needed.

**If this fails:** Add `(week | person)` to the model formula. If this causes a singular fit, the data do not support individual slope variation — simplify back to `(1 | person)` and acknowledge that the model assumes homogeneous rates of change.
:::

---

## Reviewer Challenge

::: {.callout-tip}
## What a Reviewer Would Ask

**Challenge 1:** "You fit a linear growth model — but why should stress decline at a constant rate? Isn't it more plausible that improvement is rapid at first and then decelerates?"
The correct response: show the mean-by-wave plot from the pre-fit check. If that plot is approximately linear and AIC does not favor a quadratic model, argue for parsimony. If it shows clear curvature, upgrade to quadratic and show both the comparison and the improvement in residual pattern.

**Challenge 2:** "You have a random slope for time — does the model converge properly? What is the estimated slope variance and intercept-slope correlation?"
Always report: `isSingular()` result, estimated random slope variance, intercept-slope correlation. If singular, explain what simplification was applied and why.

**Challenge 3:** "Your growth model uses wave number as time, but participants were measured on different calendar dates. Doesn't that introduce error into the time variable?"
If measurement timing varies across participants, use the exact elapsed time in days or weeks, not a wave label. This is especially important when intervals are unequal or participant-specific.
:::

## Minimum Reporting Checklist

- [ ] State what time = 0 represents and the units of the slope
- [ ] Show the spaghetti plot and mean-by-wave plot as pre-fit evidence for the chosen functional form
- [ ] Report whether linear, quadratic, or another functional form was selected and why (AIC + pre-fit plot)
- [ ] Report the random effects structure: intercept only vs. intercept + slope
- [ ] Report whether `isSingular()` returns FALSE
- [ ] Report the random slope variance and intercept-slope correlation
- [ ] Show post-fit residuals by wave — confirms the functional form adequately captured the trajectory
- [ ] If a treatment × slope interaction is reported, interpret it as a difference in rates of change, not just a significance test
