---
title: "Survival Analysis for Longitudinal Events"
---

```{r setup, include=FALSE}
library(tidyverse)
library(survival)
library(survminer)
library(broom)
library(patchwork)
set.seed(2024)
```

# Survival Analysis for Longitudinal Events

::: {.callout-note icon=false}
## The Event Timing Question

> **What is the outcome here?** Not a score that changes over time — an event. A customer churns. A borrower defaults. A household makes its first purchase after product acquisition. The scientific question is not "how did the score change?" but "how long until the event happened, and what predicts the rate at which it happens?"
>
> **What makes this different from regression?** Two things: censoring (not everyone experiences the event during the observation window) and time-varying hazard (the rate of the event changes over time). Standard regression cannot handle either correctly.
>
> **What does survival analysis do?** It models the hazard — the instantaneous rate of the event at each moment, conditional on having survived to that point — while correctly accounting for censored observations.
:::

## When the Outcome Is Time-to-Event

Many JDM outcomes of interest are not continuous scores but **events with timing**:

- A customer **churns** from a financial app — leaves the platform
- A borrower **defaults** on their first late payment
- A household **opts out** of a default enrollment they were automatically placed in
- A consumer makes their **first purchase** after receiving a marketing offer
- A subscriber **cancels** automatic fee renewal

For each of these, the scientific question concerns *when* the event happens and what speeds or slows the rate. Standard regression applied to the event indicator (event = 1/0) ignores timing entirely. Standard regression applied to the elapsed time until the event ignores **censoring** — participants who have not yet experienced the event by the end of the study. Discarding them as "incomplete data" would introduce systematic bias (the longer-surviving, more loyal customers would vanish from the analysis).

Survival analysis handles both: it models the hazard function over time while correctly incorporating information from participants who have not yet churned, defaulted, or opted out.

### What Survival Data Look Like: Censoring in Practice

The diagram below shows why standard regression fails for time-to-event outcomes. Each row is a customer. Some churn at a known time (red ✕). Others are still active when the study ends, or leave for reasons unrelated to churn — these are **censored** (grey dot). Dropping censored customers would systematically remove the most loyal users and overstate churn rates. Survival analysis retains them, using their partial observation: the fact that they survived *at least* as long as they were observed.

```{r censoring-diagram, echo=FALSE, fig.width=9, fig.height=3.8, message=FALSE, warning=FALSE}
df_cens <- tibble(
  id    = 1:10,
  label = paste0("Customer ", LETTERS[1:10]),
  end   = c(22, 180, 65, 95, 110, 42, 180, 88, 140, 55),
  type  = c("Churned",   "Active at end", "Churned",  "Withdrew early",
            "Churned",   "Churned",       "Active at end", "Churned",
            "Withdrew early", "Churned")
)

df_cens <- df_cens |>
  mutate(
    label = fct_reorder(label, id, .desc = TRUE),
    color_grp = case_when(
      type == "Churned"          ~ "Churned (event observed)",
      type == "Active at end"    ~ "Censored: still active day 180",
      type == "Withdrew early"   ~ "Censored: withdrew early"
    )
  )

ggplot(df_cens, aes(y = label)) +
  geom_segment(aes(x = 0, xend = end, yend = label, color = color_grp),
               linewidth = 4.5, lineend = "round", alpha = 0.72) +
  geom_point(data = df_cens |> filter(type == "Churned"),
             aes(x = end), shape = 4, size = 4.5, stroke = 2.2,
             color = "#c0392b") +
  geom_point(data = df_cens |> filter(type != "Churned"),
             aes(x = end), shape = 16, size = 4, color = "#555555") +
  geom_vline(xintercept = 180, linetype = "dashed", color = "#888888",
             linewidth = 0.8) +
  annotate("text", x = 182, y = 0.55,
           label = "Study end\n(day 180)", hjust = 0, size = 2.9,
           color = "#888888") +
  scale_x_continuous(limits = c(0, 215),
                     breaks = seq(0, 180, 30)) +
  scale_color_manual(
    values = c("Churned (event observed)"       = "#e63946",
               "Censored: still active day 180" = "#1e3a5f",
               "Censored: withdrew early"       = "#aaaaaa"),
    name = NULL
  ) +
  labs(
    title    = "Censoring: what we observe and what we don't",
    subtitle = "\u2715 = churn event observed     \u25cf = censored (event not yet happened or lost to follow-up)",
    x        = "Days since account opening",
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

Customers B and G survive the full 180 days — they contribute positive information about surviving through day 180 without implying they never churn. Customers D and I withdrew early for unrelated reasons — they contribute positive information through their last observed day. Excluding either group would remove the most persistent users and produce a biased estimate of the survival function.

---

## The Hazard and Survival Functions

**The survival function** $S(t)$ gives the probability of surviving (not experiencing the event) past time $t$:
$$S(t) = P(T > t)$$

**The hazard function** $h(t)$ gives the instantaneous rate of the event at time $t$, given survival to that point:
$$h(t) = \lim_{\Delta t \to 0} \frac{P(t \leq T < t + \Delta t \mid T \geq t)}{\Delta t}$$

They are linked by:
$$S(t) = \exp\left(-\int_0^t h(u)\, du\right)$$

**JDM intuitions:**

- **Churn hazard**: typically highest in the first weeks after account opening (buyer's remorse / friction with the app), then declining as engaged users settle in — a decreasing hazard.
- **Default hazard**: may increase over time as debt accumulates and financial stress compounds — an increasing hazard.
- **First purchase hazard**: may peak a few days after an offer arrives (immediate response window), then decay as the offer recedes from attention — a non-monotone hazard.

---

## Step 1: The Kaplan-Meier Estimator

The Kaplan-Meier (KM) estimator is the nonparametric estimate of the survival function. It makes no assumptions about the shape of the hazard and handles censoring correctly.

### Simulating Time-to-Churn Data

We simulate 120 customers: 60 assigned to a standard onboarding flow, 60 to an enhanced onboarding with a personalized savings goal feature. We track days until app churn over a 180-day observation window. Customers who have not churned by day 180 are censored.

```{r simulate-survival}
n_customers <- 120
condition   <- rep(c(0, 1), each = 60)   # 0 = standard, 1 = enhanced onboarding

# True churn times: exponential with condition-dependent rate
rate_standard <- 0.012  # average churn ~83 days
rate_enhanced <- 0.006  # enhanced roughly halves churn rate

churn_time <- ifelse(
  condition == 0,
  rexp(n_customers, rate = rate_standard),
  rexp(n_customers, rate = rate_enhanced)
)

# Censoring at day 180
follow_up  <- 180
event_time <- pmin(churn_time, follow_up)
event      <- as.integer(churn_time <= follow_up)   # 1 = churned, 0 = censored

df_surv <- tibble(
  customer  = 1:n_customers,
  days      = event_time,
  churned   = event,
  condition = factor(condition, labels = c("Standard onboarding", "Enhanced onboarding"))
)

df_surv |>
  group_by(condition) |>
  summarise(
    n_churned    = sum(churned),
    n_retained   = sum(1 - churned),
    median_days  = median(days[churned == 1]),
    .groups      = "drop"
  ) |>
  knitr::kable(caption = "Churn summary by onboarding condition")
```

### Kaplan-Meier Survival Curve

```{r km-fit, fig.width=8, fig.height=4.5}
km_fit <- survfit(Surv(days, churned) ~ condition, data = df_surv)

plot_data <- broom::tidy(km_fit)

ggplot(plot_data, aes(x = time, y = estimate, color = strata, fill = strata)) +
  geom_step(linewidth = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  scale_color_manual(
    values = c("#1e3a5f", "#2d6a4f"),
    labels = c("Standard onboarding", "Enhanced onboarding"),
    name   = NULL
  ) +
  scale_fill_manual(
    values = c("#1e3a5f", "#2d6a4f"),
    labels = c("Standard onboarding", "Enhanced onboarding"),
    name   = NULL
  ) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  labs(
    title    = "Kaplan-Meier survival curves: time to app churn",
    subtitle = "Shaded = 95% CI. Steps at each churn event. Customers not yet churned are censored at day 180.",
    x        = "Days since account opening",
    y        = "P(still active on platform)"
  ) +
  theme_minimal(base_size = 13)
```

The KM curve starts at $S(0) = 1$ (everyone active) and steps down at each churn event. The **median survival time** is the day at which the curve crosses 0.50 — half the customers in that condition have churned by that point. Censored customers (still active at day 180) contribute their partial information: they reduce the "at risk" count at later times without contributing a step down.

---

## Step 2: The Log-Rank Test

```{r log-rank}
logrank_test <- survdiff(Surv(days, churned) ~ condition, data = df_surv)
logrank_test
```

The log-rank test compares survival curves between conditions without assuming a parametric shape. It is most powerful when the hazard ratio between conditions is approximately constant over time — the proportional hazards assumption.

---

## Step 3: The Cox Proportional Hazards Model

The Cox model is the standard regression tool for survival outcomes:

$$h(t \mid X_j) = \underbrace{h_0(t)}_{\substack{\text{baseline hazard} \\ \text{(unspecified)}}} \cdot \exp(\beta_1 X_{1j} + \beta_2 X_{2j} + \cdots)$$

Each $\exp(\beta_k)$ is a **hazard ratio**: the multiplicative factor by which the instantaneous churn rate changes for a one-unit increase in $X_k$.

- HR = 0.5: the treatment halves the churn rate at every time point
- HR = 1.0: no effect on the churn rate
- HR = 2.0: the predictor doubles the churn rate

```{r cox-fit}
cox_model <- coxph(Surv(days, churned) ~ condition, data = df_surv)

broom::tidy(cox_model, exponentiate = TRUE, conf.int = TRUE) |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  knitr::kable(
    digits = 3,
    caption = "Cox model: hazard ratio for enhanced vs. standard onboarding. HR < 1 means lower churn rate."
  )
```

### The Proportional Hazards Assumption

The Cox model assumes the hazard ratio between conditions is **constant over time**. If the enhanced onboarding's protective effect fades after month 3 (customers disengage with their savings goal), the curves would converge in later periods — violating proportional hazards.

```{r ph-test, fig.width=7, fig.height=4}
cox_test <- cox.zph(cox_model)
cox_test

ggplot(
  data.frame(time = cox_test$time, residual = cox_test$y[, 1]),
  aes(x = time, y = residual)
) +
  geom_point(alpha = 0.4, color = "#1e3a5f") +
  geom_smooth(method = "loess", color = "#e63946", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title    = "Schoenfeld residuals: testing proportional hazards",
    subtitle = "Flat loess line supports PH. Trend suggests time-varying hazard ratio.",
    x = "Days since account opening",
    y = "Schoenfeld residual"
  ) +
  theme_minimal(base_size = 13)
```

---

## Step 4: Frailty Models — Handling Clustered Churn Data

When customers are nested in contexts — bank branches, geographic regions, or platform versions — standard Cox models assume independence that is violated. Customers from the same branch may churn for similar branch-level reasons (local economic shock, branch closure, manager change). The survival analogue of a random intercept is a **frailty** — a latent multiplicative factor that shifts the churn hazard up or down for everyone in a cluster.

$$h_{ij}(t) = h_0(t) \cdot \exp(\beta X_{ij}) \cdot Z_j, \quad Z_j \sim \text{Gamma}(1, \theta)$$

$Z_j$ is the frailty for branch $j$. If $Z_j > 1$, all customers in that branch have higher churn risk — perhaps because local economic conditions make financial management more stressful for everyone there.

```{r simulate-clustered-survival}
n_branches   <- 15
n_per_branch <- 8
n_total      <- n_branches * n_per_branch

branch       <- rep(1:n_branches, each = n_per_branch)
condition_b  <- rep(rep(c(0, 1), each = n_per_branch / 2), n_branches)

# Branch-level frailty (some branches have structurally higher churn)
branch_frailty <- rgamma(n_branches, shape = 2, rate = 2)  # variance = 0.5

base_rate   <- 0.010
churn_time_b <- rexp(
  n_total,
  rate = base_rate * branch_frailty[branch] * exp(0.5 * (1 - condition_b))
)

df_frailty <- tibble(
  customer  = 1:n_total,
  branch    = factor(branch),
  days      = pmin(churn_time_b, follow_up),
  churned   = as.integer(churn_time_b <= follow_up),
  condition = factor(condition_b, labels = c("Standard", "Enhanced"))
)
```

```{r frailty-fit}
cox_frailty <- coxph(
  Surv(days, churned) ~ condition + frailty(branch, distribution = "gamma"),
  data = df_frailty
)

summary(cox_frailty)
```

The frailty variance estimate tells you how much branches differ in their baseline churn hazard. Large frailty variance means the branch you are a customer of matters a great deal for your churn risk — independent of the onboarding condition. This is the survival analogue of a large ICC.

---

## Assumptions

::: {.callout-warning icon=false}
## Assumption: Non-Informative Censoring

**Plain-language meaning:** Censoring must be **independent of the event process**. Customers who are censored (still active at the end of the observation window) must not systematically differ in their imminent churn risk from those who have not yet been censored.

**What the model is assuming away:** Any relationship between the *reason* for censoring and the *rate* of the event. If customers who are about to churn disengage from the app before formally churning — becoming inactive but not yet counted as churned — they are effectively informatively censored.

**JDM example:** If financially distressed customers frequently delete the app before their account officially lapses (a soft churn that precedes the hard churn event), the observed churn times underrepresent the actual churning rate. Customers with high churn risk are leaving the observation window early for risk-related reasons.

**Experimental analog:** In a lab study of time-to-abandonment of a financial task, participants who give up before completing it but whose abandonment does not trigger the "event" (task abandonment) are informatively censored.

**What a skeptical reviewer might say:** "Why would customers who are about to churn be equally likely to remain in your observation window as customers who are loyal? If pre-churn disengagement precedes formal churn, your censoring is informative."

**Observable warning sign:** Censored customers show systematically lower engagement metrics or different baseline characteristics than event-experiencing customers — suggesting informative censoring.

**What to check:** Compare baseline characteristics (account age, engagement score, demographics) between censored and churned customers. If they differ substantially, consider sensitivity analyses or inverse-probability-of-censoring weighting (IPCW).

**What claim survives if the assumption is doubtful:** The KM estimates and Cox HRs are biased (usually toward underestimating the true event rate). If censoring depends on latent event risk, both effect magnitude and direction can be distorted.
:::

::: {.callout-warning icon=false}
## Assumption: Proportional Hazards (Cox Model Only)

**Plain-language meaning:** The hazard ratio between conditions is constant over time. The enhanced onboarding's protective effect at day 30 is the same multiplier as at day 120.

**What the model is assuming away:** Time-varying treatment effects — for example, an onboarding boost that fades after initial engagement, or a default enrollment whose effect builds as habits solidify.

**JDM example:** A personalized savings goal feature may prevent early churn (when customers are still learning the app) but have weaker effects later once the novelty fades. The hazard ratio would drift toward 1 over time, violating PH.

**What a skeptical reviewer might say:** "Why should the effect of enhanced onboarding on churn be a fixed multiplier at all time points? Surely the benefit is concentrated in the first weeks when onboarding is salient."

**What to check:** `cox.zph()` test and Schoenfeld residual plot. If violated, add a time-by-treatment interaction (`tt()` in `survival`) or stratify the baseline hazard by condition.
:::

::: {.callout-warning icon=false}
## Assumption: Correct Frailty Distribution

**Plain-language meaning:** The frailty model assumes branch-level random effects follow a Gamma distribution (the standard choice). If the actual distribution of branch-level churn propensity is very different — heavily skewed or multimodal — the Gamma frailty may not fit well.

**What to check:** Examine residuals from the frailty model. Compare fitted and observed survival curves within branches. For robustness, consider log-normal frailty as an alternative (available in `coxme` or `frailtyEM`).
:::

---

## A Map of Survival Methods

| Goal | Method | Package |
|------|---------|---------|
| Describe survival curves nonparametrically | Kaplan-Meier | `survival::survfit()` |
| Compare two conditions | Log-rank test | `survival::survdiff()` |
| Regression with covariates | Cox proportional hazards | `survival::coxph()` |
| Test proportional hazards | Schoenfeld test | `survival::cox.zph()` |
| Handle nested/clustered data | Frailty model | `survival::coxph(frailty(...))` |
| Time-varying treatment effects | Time-interaction Cox | `survival::tt()` |
| Visualize survival curves | KM plots, forest plots | `survminer` |

---

## Reviewer Challenge

::: {.callout-tip}
## What a Reviewer Would Ask

**Challenge 1:** "You defined churn as account closure — but isn't disengagement (stopping app use) the more behaviorally relevant event, and it would be much harder to define?"
This is a latent event timing problem. Formally measured churn is a discrete, observable event; meaningful disengagement happens earlier and more continuously. Sensitivity analyses using alternative event definitions (e.g., 30-day inactivity vs. formal account closure) help assess whether conclusions depend on the event definition.

**Challenge 2:** "Your observation window is 180 days, but many customers are still active at the end. Doesn't that mean you can't actually estimate median survival for the enhanced condition?"
Correct — if the KM curve does not cross 0.50 within the observation window, the median is undefined from the data. Report the 25th percentile survival time (the day by which 25% of customers have churned) instead, or report survival probabilities at specific time points (e.g., "60% of enhanced customers were still active at day 120").

**Challenge 3:** "The frailty variance estimate is large — does that mean your treatment effect is confounded with branch-level variation?"
Not confounded, but attenuated. The frailty absorbs between-branch variance that would otherwise inflate residual variance. The treatment effect (HR) is estimated *conditional on* branch frailty — it represents the within-branch contrast, which is the more precise and causally interpretable estimate. A large frailty variance is an argument for the frailty model, not against it.
:::
