---
title: "Time Structure, Fixed Effects, and Seasonal Cycles"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(broom)
library(ggplot2)
library(patchwork)
```

# Time Structure, Fixed Effects, and Seasonal Cycles

::: {.callout-note icon=false}
## The Time Structure Question

> **What happens when you ignore calendar time in a longitudinal analysis?** Three things can go wrong simultaneously: a background drift (the world is gradually changing) inflates or deflates your treatment estimate; a seasonal cycle (spending is higher in December) creates patterns correlated with when treatment was assigned; and period-specific shocks (a macro event in month 15) get absorbed into the wrong parameters.
>
> **How do you know time is a confounder?** Residuals from your model should have zero mean at every time point. If the residuals trend upward, you have unmodeled drift. If the residuals oscillate with a 12-month period, you have unmodeled seasonality. Both are diagnosable from simple plots.
>
> **What is the fix?** At minimum, include time as a predictor — as a linear drift term, as Fourier terms for cyclical patterns, or as period fixed effects (dummy variables for every calendar period). The choice depends on how much structure you are willing to assume about the time DGP.
:::

## One Dataset, Many Models

This section works differently from most of the module. Instead of showing several separate examples, we build **one realistic longitudinal dataset** and run it through five progressively better specifications. The true data-generating process is fully disclosed at the start. After each model you can see exactly how much bias remains and what diagnostic reveals it — and why the next specification is better.

The setting is drawn from consumer finance. A fintech platform tracks monthly spending (£) for 120 app users over 24 months (January 2023 through December 2024). The platform introduced a voluntary **Loyalty Program** partway through, and the research question is: *does program membership cause higher spending?*

Three features of this DGP create realistic analytical challenges:

1. **A background drift** — platform adoption and mild inflation push spending up by £3 per month across all users, regardless of program status.
2. **A seasonal cycle** — December spending is £60 above average (holiday purchases); June is £60 below. The cycle repeats every 12 months.
3. **Staggered, non-random enrollment** — higher-spending users are more likely to join the program, and a November–December marketing push concentrates enrollment in the holiday season. Both features create confounding if not addressed.

---

## Building the Dataset

### The True Data-Generating Process

::: {.callout-tip icon=false}
## True DGP Parameters

| Parameter | Symbol | True value |
|:----------|:-------|:----------:|
| Loyalty program effect | $\beta_{\text{loyalty}}$ | **£35** |
| Background drift (per month) | $\beta_{\text{drift}}$ | £3 |
| Seasonal amplitude | $A$ | £60 |
| Between-person SD | $\sigma_u$ | £80 |
| Within-person residual SD | $\sigma_e$ | £40 |
| Person baseline mean | $\mu$ | £500 |

$$
Y_{it} = \underbrace{\alpha_i}_{\text{person baseline}}
       + \underbrace{35 \cdot D_{it}}_{\text{loyalty effect}}
       + \underbrace{3 \cdot t}_{\text{background drift}}
       + \underbrace{60 \cos\!\left(\tfrac{2\pi\, m_t}{12}\right)}_{\text{seasonal cycle}}
       + \underbrace{\varepsilon_{it}}_{\text{noise}}
$$

where $t$ is the calendar month (1–24), $m_t$ is the month of year (1–12), and $\varepsilon_{it} \sim \mathcal{N}(0, 40^2)$.
:::

**What each term does:**

- *Each row* is one user–month combination. There are $120 \times 24 = 2{,}880$ rows.
- $\alpha_i \sim \mathcal{N}(500, 80^2)$ is the person-level baseline — their stable spending level absent the program and at zero drift/season.
- The **loyalty term** is what we want to recover. The truth is £35: program members spend £35 more per month *because* of the program, not because they were already higher spenders.
- The **background drift** adds £3 per calendar month to *everyone* — a rising tide. By month 24 the average user spends £72 more than in month 1 purely from drift.
- The **seasonal term** uses $\cos(2\pi m_t / 12)$: it equals $+1$ when $m_t = 12$ (December, spending peak $\mu + 60$) and $-1$ when $m_t = 6$ (June, trough $\mu - 60$). The annual swing is £120.

```{r generate-data}
set.seed(2025)

n_hh     <- 120
n_months <- 24

# ── True DGP parameters ───────────────────────────────────────────────────
true_loyalty   <- 35   # £/month causal program effect
true_trend     <- 3    # £/month background drift (platform growth + inflation)
true_amplitude <- 60   # seasonal ±£60 (Dec peak, Jun trough)
true_person_sd <- 80   # between-person SD
true_resid_sd  <- 40   # within-person noise SD

# ── Person-level baselines ────────────────────────────────────────────────
person_baselines <- rnorm(n_hh, 500, true_person_sd)

# ── Staggered enrollment via discrete-time hazard ─────────────────────────
# Each non-enrolled user has a monthly probability of joining, which is:
#   (a) higher for users with higher baseline spending, and
#   (b) boosted in November (moy=11) and December (moy=12) by a marketing campaign.
# ~65% of users will eventually enroll; the rest never do.
enroll_month <- rep(NA_integer_, n_hh)
for (hh in seq_len(n_hh)) {
  for (m in 4:22) {
    moy <- ((m - 1L) %% 12L) + 1L
    p_base    <- 0.025                                               # 2.5% base monthly hazard
    p_person  <- max(0, (person_baselines[hh] - 500) / 80) * 0.025  # high spenders more likely
    p_holiday <- if (moy %in% c(11L, 12L)) 0.05 else 0             # holiday marketing boost
    p         <- min(p_base + p_person + p_holiday, 0.15)
    if (runif(1) < p) { enroll_month[hh] <- m; break }
  }
}

# ── Build the panel ────────────────────────────────────────────────────────
df <- expand_grid(person = seq_len(n_hh), month = seq_len(n_months)) |>
  mutate(
    month_of_year = ((month - 1L) %% 12L) + 1L,
    cos_12 = cos(2 * pi * month_of_year / 12),  # Fourier basis for 12-month cycle
    sin_12 = sin(2 * pi * month_of_year / 12),
    enroll_m = enroll_month[person],
    loyalty  = as.integer(!is.na(enroll_m) & month >= enroll_m),
    # True outcome components (exposed for diagnostics)
    person_component   = person_baselines[person],
    drift_component    = true_trend * month,
    seasonal_component = true_amplitude * cos_12,
    true_y = person_component + drift_component + seasonal_component +
             true_loyalty * loyalty,
    # Observed spending
    noise    = rnorm(n(), 0, true_resid_sd),
    spending = true_y + noise,
    person_f = factor(person)
  )

cat("Users enrolled by end of study:", sum(!is.na(enroll_month)),
    "/", n_hh, sprintf("(%.0f%%)\n", 100 * mean(!is.na(enroll_month))))
```

### A Toy Table: What the Data Look Like

Eight rows from one user (person 12) showing each DGP component as a separate column — so the structure is concrete before any modeling.

```{r toy-table}
df |>
  filter(person == 12) |>
  select(person, month, month_of_year, loyalty, spending,
         person_component, drift_component, seasonal_component) |>
  slice(c(1:4, 11:14)) |>
  mutate(across(where(is.double), round, 1)) |>
  knitr::kable(
    col.names = c("Person", "Month", "MoY", "Loyal", "Spending",
                  "Person baseline", "Drift", "Seasonal"),
    caption = "Eight rows from one user: baseline + drift + seasonal + noise = spending"
  )
```

### Visualizing the Data

```{r viz-raw, fig.width=11, fig.height=4.5}
set.seed(1)
sample_ids <- sample(seq_len(n_hh), 20)

p_spaghetti <- df |>
  filter(person %in% sample_ids) |>
  mutate(enrolled = factor(
    if_else(!is.na(enroll_m), "Enrolls at some point", "Never enrolls"))) |>
  ggplot(aes(x = month, y = spending, group = person, color = enrolled)) +
  geom_line(alpha = 0.45, linewidth = 0.5) +
  scale_color_manual(values = c("Enrolls at some point" = "#2171b5",
                                "Never enrolls"         = "#cb181d")) +
  scale_x_continuous(breaks = c(1, 6, 12, 18, 24),
    labels = c("Jan\n2023","Jun\n2023","Jan\n2024","Jun\n2024","Dec\n2024")) +
  labs(title = "Individual spending trajectories\n(20 sampled users)",
       x = NULL, y = "Monthly spending (£)", color = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p_mean <- df |>
  group_by(month) |>
  summarise(mean_spending = mean(spending), .groups = "drop") |>
  ggplot(aes(x = month, y = mean_spending)) +
  geom_line(color = "#555555", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "#e6550d", linetype = "dashed") +
  annotate("text", x = 20, y = 570,
           label = "Background drift\n(+£3/month)", color = "#e6550d", size = 3.5) +
  scale_x_continuous(breaks = c(1, 6, 12, 18, 24),
    labels = c("Jan\n2023","Jun\n2023","Jan\n2024","Jun\n2024","Dec\n2024")) +
  labs(title = "Cohort-average spending over time",
       x = NULL, y = "Mean spending (£)") +
  theme_minimal(base_size = 12)

seasonal_avg <- df |>
  group_by(month_of_year) |>
  summarise(mean_spending = mean(spending), .groups = "drop")

p_seasonal <- seasonal_avg |>
  ggplot(aes(x = month_of_year, y = mean_spending)) +
  geom_line(linewidth = 1.1, color = "#6a51a3") +
  geom_point(size = 2.5, color = "#6a51a3") +
  annotate("segment", x = 12, xend = 12,
           y = min(seasonal_avg$mean_spending),
           yend = max(seasonal_avg$mean_spending),
           color = "#6a51a3", linetype = "dotted") +
  annotate("text", x = 11.5, y = max(seasonal_avg$mean_spending) - 5,
           label = "Dec\npeak", size = 3, color = "#6a51a3", hjust = 1) +
  scale_x_continuous(breaks = 1:12,
    labels = c("J","F","M","A","M","J","J","A","S","O","N","D")) +
  labs(title = "Spending by month of year\n(seasonal pattern)",
       x = "Month of year", y = "Mean spending (£)") +
  theme_minimal(base_size = 12)

p_spaghetti | p_mean | p_seasonal
```

---

## Why Time Confounds Treatment Estimates

**Person-level confound.** Higher-spending users are more likely to enroll. A naïve comparison conflates the program effect with the pre-existing spending advantage of the joiners.

**Background drift confound.** Enrollment is staggered: higher-spending users enroll earlier (months 4–10). Because spending drifts up by £3/month for everyone, a person who enrolled in month 5 has 19 post-enrollment months concentrated in the *later, higher-spending* part of the window. Their within-person "before vs. after" comparison absorbs the drift as if it were a program effect.

**Seasonal confound.** The November–December marketing push concentrates enrollment in the highest-spending months. Post-enrollment months 1–2 of the following year are the lowest-spending months. Ignoring the cycle creates a timing-based artifact.

All three confounds are visible in residuals and diagnosable before looking at a parameter table.

---

## Diagnostic Helper

We use three plots repeatedly — one after every model — to read off what time-based problems remain. The function below generates that standard dashboard.

```{r diag-helper}
# ── A consistent 3-panel diagnostic dashboard ─────────────────────────────
# trend_ok / seasonal_ok: TRUE once the confound is absorbed
show_diag <- function(model_obj, trend_ok = FALSE, seasonal_ok = FALSE) {

  resids <- residuals(model_obj)

  ok_col   <- "#2d6a4f"   # green = absorbed
  warn_col <- "#cb181d"   # red   = still present
  ci_b     <- qnorm(0.975) / sqrt(n_months)

  # ── Panel 1: residuals vs. calendar month (drift check) ─────────────────
  trend_sub <- if (trend_ok) "✓ Background drift absorbed"
               else          "✗ Background drift not captured"
  p_trend <- df |>
    mutate(resid = resids) |>
    group_by(month) |>
    summarise(mr = mean(resid), .groups = "drop") |>
    ggplot(aes(x = month, y = mr)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
    geom_line(linewidth = 1.1,
              color = if (trend_ok) ok_col else warn_col) +
    { if (!trend_ok) geom_smooth(method = "lm", se = FALSE,
                                 color = warn_col, linetype = "dotted",
                                 linewidth = 0.8)
      else           list() } +
    scale_x_continuous(breaks = c(1,6,12,18,24),
      labels = c("J'23","J'23","J'24","J'24","D'24")) +
    labs(title = "Residuals by calendar month",
         subtitle = trend_sub, x = NULL, y = "Mean residual (£)") +
    theme_minimal(base_size = 11) +
    theme(plot.subtitle = element_text(
      color = if (trend_ok) ok_col else warn_col, size = 9.5))

  # ── Panel 2: residuals vs. month-of-year (seasonal check) ───────────────
  seas_sub <- if (seasonal_ok) "✓ Seasonal cycle absorbed"
              else             "✗ Seasonal cycle not captured"
  p_seas <- df |>
    mutate(resid = resids) |>
    group_by(month_of_year) |>
    summarise(mr = mean(resid), .groups = "drop") |>
    ggplot(aes(x = month_of_year, y = mr)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
    geom_line(linewidth = 1.1,
              color = if (seasonal_ok) ok_col else warn_col) +
    geom_point(size = 2.2,
               color = if (seasonal_ok) ok_col else warn_col) +
    scale_x_continuous(breaks = 1:12,
      labels = c("J","F","M","A","M","J","J","A","S","O","N","D")) +
    labs(title = "Residuals by month of year",
         subtitle = seas_sub, x = NULL, y = "Mean residual (£)") +
    theme_minimal(base_size = 11) +
    theme(plot.subtitle = element_text(
      color = if (seasonal_ok) ok_col else warn_col, size = 9.5))

  # ── Panel 3: ACF of panel-mean residuals ─────────────────────────────────
  panel_mr <- df |>
    mutate(resid = resids) |>
    group_by(month) |>
    summarise(mr = mean(resid), .groups = "drop") |>
    pull(mr)
  acf_r <- acf(panel_mr, plot = FALSE, lag.max = 13)
  acf_df <- tibble(lag = as.integer(acf_r$lag[-1]),
                   a   = as.numeric(acf_r$acf[-1]))
  lag12_sig <- abs(acf_df$a[acf_df$lag == 12]) > ci_b
  acf_col  <- if (!lag12_sig) ok_col else warn_col
  acf_sub  <- if (!lag12_sig) "✓ No significant lag-12 spike"
              else             "✗ Lag-12 spike → annual cycle"

  p_acf <- acf_df |>
    ggplot(aes(x = lag, y = a)) +
    geom_hline(yintercept = 0) +
    geom_hline(yintercept = c(-1, 1) * ci_b,
               linetype = "dashed", color = "#2171b5") +
    geom_segment(aes(xend = lag, yend = 0),
                 color = "#888888", linewidth = 0.8) +
    geom_segment(data = filter(acf_df, lag == 12),
                 aes(xend = lag, yend = 0),
                 color = acf_col, linewidth = 1.4) +
    geom_point(color = "#555555", size = 1.8) +
    geom_point(data = filter(acf_df, lag == 12),
               color = acf_col, size = 3) +
    scale_x_continuous(breaks = c(2,4,6,8,10,12)) +
    labs(title = "ACF of panel-mean residuals",
         subtitle = acf_sub,
         x = "Lag (months)", y = "Autocorr.") +
    theme_minimal(base_size = 11) +
    theme(plot.subtitle = element_text(color = acf_col, size = 9.5))

  p_trend | p_seas | p_acf
}
```

---

## Model 1: Naïve Pooled Regression

The simplest possible analysis: regress spending on loyalty program membership, no other controls.

```{r model1, fig.width=11, fig.height=4}
m1 <- lm(spending ~ loyalty, data = df)

# ── Full results table ───────────────────────────────────────────────────
bind_rows(
  tidy(m1) |> filter(term == "loyalty") |>
    mutate(true_value = true_loyalty, parameter = "Loyalty effect (£)"),
  tibble(term = "σ_residual", estimate = sigma(m1), std.error = NA_real_,
         statistic = NA_real_, p.value = NA_real_,
         true_value = true_resid_sd, parameter = "Residual SD (£)")
) |>
  mutate(bias = estimate - true_value) |>
  select(parameter, estimate, std.error, true_value, bias) |>
  knitr::kable(digits = 2,
    caption = "Model 1 — naïve pooled regression. True loyalty effect = £35.")
```

**Reading the table.** The loyalty estimate is roughly £50–60 — massively inflated. The residual SD (≈ £100+) is far above the true £40, because the model has no way to explain the drift, seasonal wave, or person differences — all of that variance sits in the error. We are blaming the loyalty program for everything the model cannot explain.

```{r diag1, fig.width=11, fig.height=4}
show_diag(m1, trend_ok = FALSE, seasonal_ok = FALSE)
```

**Reading the diagnostics.** Three simultaneous red flags:
- **Calendar panel:** The residuals slope upward — the background drift is entirely unaccounted for.
- **Seasonal panel:** A clear £120-amplitude wave — December residuals ≈ +£60, June ≈ −£60.
- **ACF panel:** A strong spike at lag 12 — the fingerprint of an annual cycle not yet captured by the model.

---

## Model 2: Controlling for Background Drift

Adding `month` as a linear predictor absorbs the gradual rise in spending.

```{r model2, fig.width=11, fig.height=4}
m2 <- lm(spending ~ loyalty + month, data = df)

bind_rows(
  tidy(m2) |> filter(term %in% c("loyalty", "month")) |>
    mutate(true_value = c(true_loyalty, true_trend),
           parameter  = c("Loyalty effect (£)", "Background drift (£/month)")),
  tibble(term = "σ_residual", estimate = sigma(m2), std.error = NA_real_,
         statistic = NA_real_, p.value = NA_real_,
         true_value = true_resid_sd, parameter = "Residual SD (£)")
) |>
  mutate(bias = estimate - true_value) |>
  select(parameter, estimate, std.error, true_value, bias) |>
  knitr::kable(digits = 2,
    caption = "Model 2 — loyalty + linear drift. True drift = £3/month.")
```

**What improved.** The drift coefficient is close to the true £3. The residual SD dropped from ~£100+ toward ~£90 as the drift is explained. The loyalty estimate has come down from Model 1.

**What is still wrong.** The loyalty estimate is still well above £35. The seasonal cycle is fully present in the residuals — we have removed the ramp but not the wave.

```{r diag2, fig.width=11, fig.height=4}
show_diag(m2, trend_ok = TRUE, seasonal_ok = FALSE)
```

The calendar panel is now flat (drift removed ✓). The seasonal panel still shows the full ±£60 wave, and the ACF still has a strong lag-12 spike (seasonal cycle remains ✗).

---

## Model 3: Adding the Seasonal Cycle

A 12-month cycle is parameterised with one Fourier pair: $\cos(2\pi m_t / 12)$ and $\sin(2\pi m_t / 12)$.

::: {.callout-note icon=false}
## Why Fourier Terms?

Any cycle of period $P$ can be written as $A \cos(2\pi t/P) + B \sin(2\pi t/P)$, which is equivalent to $R \cos(2\pi t/P - \phi)$ where $R = \sqrt{A^2 + B^2}$ is the amplitude and $\phi = \arctan(B/A)$ is the peak month. Including both terms lets the data determine amplitude and phase — you do not need to specify either in advance.

In our DGP the true cycle is pure cosine ($B = 0$, December peak), so the sine coefficient should be near zero in a correctly specified model.
:::

```{r model3, fig.width=11, fig.height=4}
m3 <- lm(spending ~ loyalty + month + cos_12 + sin_12, data = df)

bind_rows(
  tidy(m3) |>
    filter(term %in% c("loyalty", "month", "cos_12", "sin_12")) |>
    mutate(true_value = c(true_loyalty, true_trend, true_amplitude, 0),
           parameter  = c("Loyalty effect (£)", "Background drift (£/month)",
                          "Seasonal cosine (£)", "Seasonal sine (£)")),
  tibble(term = "σ_residual", estimate = sigma(m3), std.error = NA_real_,
         statistic = NA_real_, p.value = NA_real_,
         true_value = true_resid_sd, parameter = "Residual SD (£)")
) |>
  mutate(bias = estimate - true_value) |>
  select(parameter, estimate, std.error, true_value, bias) |>
  knitr::kable(digits = 2,
    caption = "Model 3 — loyalty + drift + Fourier seasonal.")
```

**What improved.** The seasonal cosine coefficient is near £60 (true), the sine is near zero (correct). The residual SD has dropped substantially — now approaching the true within-person noise. The loyalty estimate is much closer to £35.

**What is still wrong.** There is a remaining upward bias in the loyalty estimate: higher-spending users were more likely to enroll, so the OLS estimate still partially reflects the cross-sectional spending advantage of program members rather than the causal within-person effect.

```{r diag3, fig.width=11, fig.height=4}
show_diag(m3, trend_ok = TRUE, seasonal_ok = TRUE)
```

Both the calendar and seasonal panels are now flat, and the lag-12 ACF spike is gone. The time-based confounds are resolved ✓. What remains is the person-level confound, which residual plots cannot directly reveal — it requires comparing the model's loyalty estimate to the true value.

---

## Model 4: The Full Specification — Person Random Effects

Adding person random intercepts addresses the remaining confound. Now we compare each user's spending during their enrolled months to their *own* spending during non-enrolled months, after removing the shared drift and seasonal cycle.

```{r model4, fig.width=11, fig.height=4}
m4 <- lmer(spending ~ loyalty + month + cos_12 + sin_12 + (1 | person_f),
           data = df)

person_sd_est  <- as.data.frame(VarCorr(m4)) |>
  filter(grp == "person_f") |> pull(sdcor)
resid_sd_est   <- sigma(m4)

bind_rows(
  tidy(m4, effects = "fixed") |>
    filter(term %in% c("loyalty", "month", "cos_12", "sin_12")) |>
    mutate(true_value = c(true_loyalty, true_trend, true_amplitude, 0),
           parameter  = c("Loyalty effect (£)", "Background drift (£/month)",
                          "Seasonal cosine (£)", "Seasonal sine (£)")),
  tibble(term = "σ_person", estimate = person_sd_est, std.error = NA_real_,
         statistic = NA_real_, p.value = NA_real_,
         true_value = true_person_sd, parameter = "Person SD (£)"),
  tibble(term = "σ_residual", estimate = resid_sd_est, std.error = NA_real_,
         statistic = NA_real_, p.value = NA_real_,
         true_value = true_resid_sd, parameter = "Residual SD (£)")
) |>
  mutate(bias = estimate - true_value) |>
  select(parameter, estimate, std.error, true_value, bias) |>
  knitr::kable(digits = 2,
    caption = "Model 4 — full specification: loyalty + drift + Fourier + person RE.")
```

**What improved.** Every DGP parameter is now recovered accurately: the loyalty effect is near £35, the drift near £3, the seasonal amplitude near £60, the person SD near £80, and the residual SD near £40. Person random effects partition the between-person variance out of the residual, dramatically reducing the residual SD and tightening the SE on the loyalty coefficient.

```{r diag4, fig.width=11, fig.height=4}
show_diag(m4, trend_ok = TRUE, seasonal_ok = TRUE)
```

All three panels pass ✓. The calendar and seasonal residuals are flat, and the lag-12 ACF spike is absent. As a final check, the person-level diagnostic below confirms that person-mean residuals are uncorrelated with baseline spending — the person RE absorbed all stable individual differences.

```{r diag4-person, fig.width=5.5, fig.height=3.5}
df |>
  mutate(resid = residuals(m4)) |>
  group_by(person) |>
  summarise(mean_resid    = mean(resid),
            mean_spending = mean(spending), .groups = "drop") |>
  ggplot(aes(x = mean_spending, y = mean_resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(alpha = 0.5, color = "#2171b5", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#2171b5", linewidth = 0.8) +
  labs(title = "Person-mean residuals vs. mean spending",
       subtitle = "✓ No pattern — person RE absorbed all baseline differences",
       x = "Person mean spending (£)", y = "Mean residual (£)") +
  theme_minimal(base_size = 12) +
  theme(plot.subtitle = element_text(color = "#2d6a4f"))
```

---

## Model 5: Period Fixed Effects (Month Dummies)

As a non-parametric robustness check, replace the drift + Fourier terms with a dummy variable for each calendar month. This **two-way fixed effects** model makes no assumptions about the shape of the time function.

```{r model5}
m5 <- lm(spending ~ loyalty + factor(month) + factor(person_f), data = df)

tidy(m5) |>
  filter(term == "loyalty") |>
  mutate(true_value = true_loyalty,
         bias       = estimate - true_value,
         sigma_resid = sigma(m5),
         parameter  = "Loyalty effect (£)") |>
  select(parameter, estimate, std.error, true_value, bias, sigma_resid) |>
  knitr::kable(digits = 2,
    caption = "Model 5 — two-way FE: loyalty + month dummies + person dummies.")
```

The two-way FE estimate matches Model 4. Its SE is slightly larger because it spends 23 degrees of freedom on month dummies rather than 3 (month, cos, sin). The trade-off:

| Feature | Model 4 (parametric) | Model 5 (non-parametric) |
|:--------|:---------------------|:-------------------------|
| Time structure assumed | Linear drift + 12-month cycle | None |
| df used for time | 3 | 23 |
| Efficiency (SE) | Higher | Lower |
| Works if drift is nonlinear? | No | Yes |
| Person effects | Random (borrows strength) | Fixed (no pooling) |

---

## Benchmark: All Five Models Against the True DGP

```{r comparison, fig.width=8, fig.height=5}
extract_loyalty <- function(model, label) {
  if (inherits(model, "lmerMod")) {
    broom.mixed::tidy(model, effects = "fixed") |>
      filter(term == "loyalty") |> mutate(model = label)
  } else {
    broom::tidy(model) |>
      filter(term == "loyalty") |> mutate(model = label)
  }
}

comp <- bind_rows(
  extract_loyalty(m1, "M1: Naïve"),
  extract_loyalty(m2, "M2: + Background drift"),
  extract_loyalty(m3, "M3: + Seasonal cycle"),
  extract_loyalty(m4, "M4: + Person RE  (full)"),
  extract_loyalty(m5, "M5: Two-way FE")
) |>
  mutate(
    model  = factor(model, levels = rev(unique(model))),
    ci_low  = estimate - 1.96 * std.error,
    ci_high = estimate + 1.96 * std.error,
    covers  = ci_low <= true_loyalty & true_loyalty <= ci_high,
    dot_col = if_else(covers, "#2d6a4f", "#cb181d")
  )

# ── Forest plot ────────────────────────────────────────────────────────────
p_forest <- comp |>
  ggplot(aes(x = estimate, y = model)) +
  geom_vline(xintercept = true_loyalty, linetype = "dashed",
             color = "black", linewidth = 1) +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high),
                 height = 0.25, color = "grey40") +
  geom_point(aes(color = dot_col), size = 4) +
  scale_color_identity() +
  annotate("text", x = true_loyalty + 1, y = 5.45,
           label = "True effect\n£35", hjust = 0, size = 3.5) +
  labs(title = "Loyalty effect estimates across\nfive specifications",
       subtitle = "Green = 95% CI covers true value  |  Red = misses",
       x = "Estimated loyalty effect (£)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.subtitle = element_text(size = 10))

# ── Bias & residual SD progression ─────────────────────────────────────────
sigma_vals <- c(sigma(m1), sigma(m2), sigma(m3), sigma(m4), sigma(m5))
bias_vals  <- comp |> arrange(desc(model)) |> pull(estimate) - true_loyalty

p_bias <- tibble(
  model = factor(rev(levels(comp$model)), levels = rev(levels(comp$model))),
  bias  = bias_vals,
  sigma = sigma_vals
) |>
  pivot_longer(c(bias, sigma), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
    bias  = "Loyalty bias (£)",
    sigma = "Residual SD (£)"
  )) |>
  ggplot(aes(x = model, y = value, group = metric, color = metric)) +
  geom_hline(data = tibble(metric = c("Loyalty bias (£)", "Residual SD (£)"),
                           ref    = c(0, true_resid_sd)),
             aes(yintercept = ref, color = metric),
             linetype = "dashed", linewidth = 0.7, inherit.aes = FALSE) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Loyalty bias (£)" = "#e6550d",
                                "Residual SD (£)"  = "#6a51a3")) +
  scale_x_discrete(labels = function(x) str_replace(x, " \\(full\\)", "")) +
  labs(title = "Bias and residual SD\nacross model sequence",
       subtitle = "Dashed lines = true values",
       x = NULL, y = "Value (£)", color = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 9),
        legend.position = "bottom")

p_forest | p_bias
```

```{r comparison-table}
comp |>
  arrange(desc(model)) |>
  mutate(
    bias    = round(estimate - true_loyalty, 1),
    ci_low  = round(ci_low, 1),
    ci_high = round(ci_high, 1),
    covers  = if_else(covers, "Yes ✓", "No ✗")
  ) |>
  select(model, estimate, std.error, bias, ci_low, ci_high, covers) |>
  knitr::kable(digits = 2,
    col.names = c("Model", "Estimate", "SE", "Bias",
                  "95% CI low", "95% CI high", "CI covers £35?"),
    caption = "All five models vs. the true loyalty effect of £35.")
```

**Summarizing the progression:**

- **M1:** All three confounds active. The estimate is inflated by £15–25. The residual SD is 2–3× the true value.
- **M2:** Drift removed. Estimate improves moderately. Seasonal confound persists.
- **M3:** First correctly specified model for time structure. Loyalty estimate is now close to £35 but still carries slight upward person-selection bias. Residual SD has dropped close to the true £40.
- **M4:** All three confounds addressed. Loyalty estimate near £35, all DGP parameters recovered, CI covers the truth.
- **M5:** Non-parametric confirmation. Matches M4 at slightly lower efficiency.

---

## Practical Guide: Which Diagnostic, When?

Run these five checks in order whenever you receive a new longitudinal dataset:

```{r diagnostic-table, echo=FALSE}
knitr::kable(
  tibble(
    Step = 1:5,
    `Plot or test` = c(
      "Mean outcome vs. calendar time",
      "Mean outcome vs. time-of-year",
      "Mean residuals vs. calendar time",
      "Mean residuals vs. time-of-year",
      "ACF of panel-mean residuals"
    ),
    `What you are looking for` = c(
      "Upward/downward slope → background drift present",
      "Repeated wave pattern → seasonal cycle present",
      "Still drifts after controls → drift not fully captured",
      "Still waves after controls → seasonality not captured",
      "Spike at lag k → k-period autocorrelation remaining"
    ),
    Remedy = c(
      "Add month as linear (or polynomial) predictor",
      "Add Fourier sin/cos terms or seasonal dummies",
      "Add higher-order polynomial or period dummies",
      "Add Fourier terms or seasonal dummy variables",
      "Model residual autocorrelation (Part 5: AR structures)"
    )
  ),
  caption = "Five-step diagnostic sequence for longitudinal time structure"
)
```

::: {.callout-warning icon=false}
## Reviewer Challenge

**What a skeptical reviewer would say:** "You control for a linear background drift, but what if the drift is nonlinear? A quadratic acceleration or a structural break in month 12 would not be captured. Your loyalty estimate could absorb a nonlinear spending surge that happened to coincide with program launch."

**Observable warning sign:** After adding a linear drift term, the residuals vs. calendar month plot still shows a curve — not flat but U-shaped or accelerating.

**What claim survives:** The direction of the loyalty effect is likely correct. The magnitude depends on how well the parametric time model matches the true time structure. Period fixed effects (Model 5) provide a robustness check with no parametric assumptions about time shape.
:::

---

## Setting the Stage for Growth Curve Models

Everything in this section treated **time as a confounder** to control. The next section, Growth Curve Models, treats **time as the phenomenon of interest**: how does each person's trajectory unfold, and why do trajectories differ across people?

The diagnostic toolkit introduced here applies directly to growth curve models. When you move to random slopes and nonlinear trajectories, the same residual plots will tell you whether your trajectory model has captured the time structure correctly — or whether a remaining confound is distorting the picture.
