---
title: "Covariance Structures"
---

```{r setup, include=FALSE}
library(tidyverse)
library(nlme)
library(lme4)
library(ggplot2)
set.seed(77)
```

# Covariance Structures

## The Covariance Matrix is a Scientific Statement

When you measure the same person at multiple time points, those measurements co-vary. How they co-vary is not a statistical nuisance — it is a substantive statement about the DGP. The choice of covariance structure encodes an assumption about how the underlying process operates over time.

The covariance structure specifies two things: **variances** (how variable are observations at each time point?) and **covariances** (how correlated are observations at different time points?).

## The Main Structures

### 1. Compound Symmetry (CS)

All variances are equal. All covariances are equal regardless of lag.

$$\Sigma_{CS} = \begin{pmatrix} \sigma^2 & \sigma^2\rho & \sigma^2\rho \\ \sigma^2\rho & \sigma^2 & \sigma^2\rho \\ \sigma^2\rho & \sigma^2\rho & \sigma^2 \end{pmatrix}$$

**When it makes sense:** When there is no reason to think observations closer in time are more similar than distant observations. Very rare in practice.

**When it fails:** Almost always. Time 1 and Time 2 are almost always more correlated than Time 1 and Time 6.

### 2. Autoregressive Order 1 — AR(1)

Correlations decay exponentially with time lag. Observations one step apart correlate $\rho$; two steps apart, $\rho^2$; $k$ steps apart, $\rho^k$.

$$\Sigma_{AR1} = \sigma^2\begin{pmatrix} 1 & \rho & \rho^2 \\ \rho & 1 & \rho \\ \rho^2 & \rho & 1 \end{pmatrix}$$

**When it makes sense:** For many psychological, physiological, and economic time series where the past predicts the present and the influence fades over time. Very commonly appropriate.

### 3. Unstructured (UN)

No constraints. Every variance and covariance is estimated freely.

**When it makes sense:** When you have no prior beliefs about the pattern and enough data to estimate all parameters.

**When it fails:** With many time points or small samples, the unstructured model is overparameterized and produces unstable estimates.

### 4. Toeplitz

Like AR(1) but relaxes the constraint that correlations must follow an exponential decay — each lag has its own correlation, but the correlation only depends on the lag (not the absolute time point). A flexible middle ground.

## Visualizing the Structures

```{r vis-structures}
make_cov_plot <- function(Sigma, title) {
  n <- nrow(Sigma)
  # Convert to correlation
  D <- diag(1 / sqrt(diag(Sigma)))
  Corr <- D %*% Sigma %*% D

  data.frame(Corr) |>
    setNames(paste0("T", 1:n)) |>
    mutate(row = paste0("T", 1:n)) |>
    pivot_longer(-row, names_to = "col", values_to = "r") |>
    mutate(row = factor(row, levels = rev(paste0("T", 1:n))),
           col = factor(col, levels = paste0("T", 1:n))) |>
    ggplot(aes(x = col, y = row, fill = r)) +
    geom_tile(color = "white", linewidth = 1) +
    geom_text(aes(label = round(r, 2)), size = 3.5) +
    scale_fill_gradient2(low = "#f7fbff", mid = "#4292c6", high = "#08306b",
                         midpoint = 0.5, limits = c(-0.1, 1)) +
    labs(title = title, x = NULL, y = NULL) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none")
}

n <- 5
rho <- 0.6

CS   <- matrix(rho, n, n); diag(CS) <- 1; CS <- CS * 4
AR1  <- outer(0:(n-1), 0:(n-1), function(i, j) rho^abs(i-j)) * 4
Toep <- outer(0:(n-1), 0:(n-1), function(i, j) {
  c(1, 0.7, 0.5, 0.3, 0.2)[abs(i-j)+1]
}) * 4
UN <- matrix(c(4.0, 2.4, 1.8, 1.4, 1.0,
               2.4, 4.2, 2.2, 1.6, 1.1,
               1.8, 2.2, 4.5, 2.4, 1.5,
               1.4, 1.6, 2.4, 4.8, 2.2,
               1.0, 1.1, 1.5, 2.2, 5.0), n, n)

library(patchwork)
(make_cov_plot(CS,   "Compound Symmetry") +
 make_cov_plot(AR1,  "AR(1)")) /
(make_cov_plot(Toep, "Toeplitz") +
 make_cov_plot(UN,   "Unstructured"))
```

## Specifying Covariance Structures in R

`lme4` does not support custom covariance structures beyond the random effects structure. For explicit covariance modeling, use `nlme`:

```{r covariance-nlme}
# Simulate data with AR(1) structure
n_people <- 50
n_times  <- 5
rho_true <- 0.7

Y_ar1 <- MASS::mvrnorm(n_people, rep(50, n_times),
                        outer(0:(n_times-1), 0:(n_times-1),
                              function(i,j) rho_true^abs(i-j)) * 9)

df_long <- as.data.frame(Y_ar1) |>
  setNames(paste0("t", 1:n_times)) |>
  mutate(person = 1:n_people) |>
  pivot_longer(-person, names_to = "time", values_to = "Y") |>
  mutate(time_num = as.integer(str_extract(time, "[0-9]")),
         person   = factor(person))

# Fit four competing models
m_cs  <- gls(Y ~ factor(time_num), correlation = corCompSymm(form = ~1 | person), data = df_long)
m_ar1 <- gls(Y ~ factor(time_num), correlation = corAR1(form = ~time_num | person), data = df_long)
m_un  <- gls(Y ~ factor(time_num), correlation = corSymm(form = ~time_num | person),
             weights = varIdent(form = ~1 | factor(time_num)), data = df_long)

# Compare AIC
tibble(
  Model = c("Compound Symmetry", "AR(1)", "Unstructured"),
  AIC   = c(AIC(m_cs), AIC(m_ar1), AIC(m_un)),
  BIC   = c(BIC(m_cs), BIC(m_ar1), BIC(m_un))
) |>
  arrange(AIC) |>
  knitr::kable(digits = 1,
    caption = "Model comparison: AR(1) fits better than CS when data has AR structure")
```

## Assumptions and How to Check Each Structure

::: {.callout-warning icon=false}
## AR(1) Assumption: Stationarity

**What it means:** The mean and variance of observations are constant over time, and the correlation between observations depends only on the time lag, not on the absolute time point.

**How to check:** Plot the mean and variance of your outcome at each time point. If either is trending systematically, stationarity is violated. A detrended model (with time as a fixed effect) may restore stationarity of the residuals.
:::

::: {.callout-warning icon=false}
## Unstructured Assumption: Sufficient Sample Size

**What it means:** The unstructured model estimates $T(T+1)/2$ covariance parameters (for $T$ time points). With 5 time points, that is 15 parameters just for the covariance structure.

**How to check:** Compare the number of observations to the number of covariance parameters. As a rough guide, you want at least 10× as many observations as parameters.
:::

## Model Selection Strategy

Use AIC/BIC to compare competing covariance structures:

1. Start with the AR(1) model as a reasonable default
2. Compare to compound symmetry (more constrained) and Toeplitz (more flexible)
3. Try unstructured if sample size allows
4. Choose the model with the lowest AIC, subject to convergence and interpretability
