---
title: "Cross-Lagged Panel Models"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lavaan)
library(ggplot2)
library(patchwork)
set.seed(2024)
```

# Cross-Lagged Panel Models

::: {.callout-note icon=false}
## The Cross-Lagged Question

> **What is the scientific question?** CLPM and RI-CLPM are for questions about *temporal ordering* — does A at time t predict B at time t+1, after controlling for B's own stability? They are not the right tool for questions about mean levels, trajectories, or causal effects from intervention.
>
> **What does this model assume about time?** That the relationship between the two variables is stationary — the same cross-lagged structure holds across all waves. And that autoregressive effects accurately capture the persistence of each variable.
>
> **The fundamental problem with classic CLPM:** Cross-lagged paths in the classic model mix within-person dynamics with between-person stable differences. People who are chronically financially stressed also tend to borrow more riskily — not because stress causes borrowing, but because both reflect the same underlying stable trait. The classic CLPM conflates these two stories.
:::

## Should You Use a CLPM at All?

Before fitting, assess whether CLPM is appropriate for your design and question.

::: {.callout-warning icon=false}
## Pre-Fit Decision: Is CLPM the Right Tool?

CLPM is likely **not appropriate** when:

| Situation | Problem | Better alternative |
|:----------|:--------|:-------------------|
| Only 2 waves | Cannot distinguish cross-lagged from concurrent correlation | Collect a third wave; use change score analysis |
| Waves are widely or unequally spaced | Lag coefficients are hard to interpret; stationarity assumption more questionable | Use continuous-time models; report the lag interval |
| Strong stable between-person differences | Cross-lagged paths will reflect trait differences, not within-person dynamics | Use RI-CLPM to separate stable from dynamic components |
| Poor reliability of either variable | Attenuation bias in cross-lagged paths; ME inflates residuals | Model measurement error explicitly or use RI-CLPM |
| Variables have strong trends | Non-stationarity violates the core assumption | Detrend variables first; use growth model approach |

CLPM is most trustworthy when you have **3+ equally spaced waves**, the variables are **relatively stable within-person** (high autoregressive effects), and **between-person trait differences are small** relative to within-person fluctuation.
:::

## Two Variables, Evolving Together

Suppose you want to understand the relationship between financial stress and risky borrowing behavior over time. Does financial stress push people toward riskier borrowing? Or does risky borrowing create financial strain? Or do they mutually reinforce each other?

To answer this, you need data on both variables at multiple time points, and you need a model that can disentangle the *temporal ordering* of their relationship. The **cross-lagged panel model (CLPM)** is the standard tool for this question.

## The Classic CLPM

In a two-wave, two-variable CLPM with variables S (financial stress) and B (risky borrowing):

$$S_{t+1} = a_{SS} \cdot S_t + a_{SB} \cdot B_t + \varepsilon_{S,t+1}$$
$$B_{t+1} = a_{BB} \cdot B_t + a_{BS} \cdot S_t + \varepsilon_{B,t+1}$$

Four parameters:
- $a_{SS}$, $a_{BB}$: **autoregressive effects** — stability of each variable over time
- $a_{SB}$: **cross-lagged effect** of B at time $t$ on S at time $t+1$
- $a_{BS}$: **cross-lagged effect** of S at time $t$ on B at time $t+1$

```{r clpm-setup}
# Simulate 3-wave panel data
n <- 200
n_waves <- 3

# True DGP: financial stress -> risky borrowing (not the reverse)
S <- matrix(NA, n, n_waves)
B <- matrix(NA, n, n_waves)

S[, 1] <- rnorm(n, 50, 8)   # baseline financial stress
B[, 1] <- rnorm(n, 45, 7)   # baseline borrowing propensity

for (t in 2:n_waves) {
  S[, t] <- 0.6 * S[, t-1] + 0.1 * B[, t-1] + rnorm(n, 0, 5)  # small B->S effect
  B[, t] <- 0.6 * B[, t-1] + 0.3 * S[, t-1] + rnorm(n, 0, 5)  # strong S->B effect
}

df_clpm <- as.data.frame(cbind(S, B)) |>
  setNames(c(paste0("S", 1:3), paste0("B", 1:3)))
```

```{r fit-clpm}
# Classic CLPM in lavaan
clpm_model <- '
  # Autoregressive effects
  S2 ~ a_SS * S1
  S3 ~ a_SS * S2
  B2 ~ a_BB * B1
  B3 ~ a_BB * B2

  # Cross-lagged effects
  S2 ~ a_SB * B1
  S3 ~ a_SB * B2
  B2 ~ a_BS * S1
  B3 ~ a_BS * S2

  # Residual covariances (within same wave)
  S2 ~~ B2
  S3 ~~ B3

  # Variance of wave 1
  S1 ~~ S1
  B1 ~~ B1
  S1 ~~ B1
'

m_clpm <- sem(clpm_model, data = df_clpm)
parameterEstimates(m_clpm) |>
  filter(op == "~") |>
  select(lhs, rhs, label, est, se, pvalue) |>
  knitr::kable(digits = 3,
    caption = "CLPM: cross-lagged path from S→B is strong; B→S is weak")
```

Good news: the CLPM detects the true asymmetric relationship (financial stress predicts future risky borrowing, not the other way around).

## The Problem with Classic CLPM

The classic CLPM conflates **within-person dynamics** with **between-person differences**. Consider: people who are chronically financially stressed tend to also engage in chronically riskier borrowing (stable between-person correlation). The cross-lagged paths in the classic CLPM pick up both this stable between-person covariation *and* the genuine within-person dynamic — they cannot be separated.

This means even if there is **no within-person causal relationship** between financial stress and risky borrowing (i.e., changes in financial stress within a person do not cause changes in borrowing behavior), the classic CLPM can show significant "cross-lagged" effects simply because chronically stressed people also tend to borrow more riskily.

### The Same Observed Pattern, Three Different Stories

The same observed correlations between financial stress and risky borrowing across waves are consistent with three completely different causal stories. The classic CLPM cannot distinguish them without extending the model.

```{r clpm-stories, echo=FALSE, fig.width=11, fig.height=3.6, message=FALSE, warning=FALSE}
make_clpm_panel <- function(arrows_df, title, note) {
  nodes <- tibble(
    label = c("S₁", "B₁", "S₂", "B₂", "S₃", "B₃"),
    x     = c(1, 1, 3, 3, 5, 5),
    y     = c(2, 1, 2, 1, 2, 1)
  )
  p <- ggplot() +
    geom_point(data = nodes, aes(x = x, y = y),
               shape = 21, size = 14, fill = "#deebf7",
               color = "#1e3a5f", stroke = 1.3) +
    geom_text(data = nodes, aes(x = x, y = y, label = label),
              size = 3.0, fontface = "bold", color = "#1e3a5f")
  for (i in seq_len(nrow(arrows_df))) {
    p <- p + annotate("segment",
                      x     = arrows_df$x1[i], xend = arrows_df$x2[i],
                      y     = arrows_df$y1[i], yend = arrows_df$y2[i],
                      arrow = arrow(length = unit(0.18, "cm"), type = "closed"),
                      color = arrows_df$col[i], linewidth = 1.0,
                      linetype = arrows_df$lty[i])
  }
  p + annotate("text", x = 3, y = 0.45, label = note,
               hjust = 0.5, size = 2.6, color = "gray30", lineheight = 1.3) +
    labs(title = title) +
    xlim(0.2, 5.8) + ylim(0.25, 2.7) +
    theme_void(base_size = 10) +
    theme(plot.title = element_text(face = "bold", size = 9.5, hjust = 0.5,
                                    margin = margin(b = 3)))
}

# AR arrows common to all panels (autoregressive stability)
ar <- tibble(x1 = c(1.5,1.5,3.5,3.5), x2 = c(2.5,2.5,4.5,4.5),
             y1 = c(2,1,2,1),          y2 = c(2,1,2,1),
             col = rep("#888888", 4),  lty = rep("solid", 4))

# Story A: S→B only
a_arrows <- bind_rows(ar,
  tibble(x1=c(1.4,3.4), x2=c(2.6,4.6), y1=c(1.85,1.85), y2=c(1.15,1.15),
         col=rep("#e63946",2), lty=rep("solid",2)))
p_a <- make_clpm_panel(a_arrows,
  "Story A: stress \u2192 risky borrowing",
  "Within-person: stress changes cause\nborrowing changes one wave later")

# Story B: B→S only
b_arrows <- bind_rows(ar,
  tibble(x1=c(1.4,3.4), x2=c(2.6,4.6), y1=c(1.15,1.15), y2=c(1.85,1.85),
         col=rep("#2d6a4f",2), lty=rep("solid",2)))
p_b <- make_clpm_panel(b_arrows,
  "Story B: risky borrowing \u2192 stress",
  "Within-person: borrowing changes cause\nstress changes one wave later")

# Story C: stable trait causes both (RI), no within-person dynamic
ri_nodes <- tibble(
  label = c("RI_S", "RI_B"),
  x = c(2, 2), y = c(2.5, 0.6)
)
p_c <- ggplot() +
  geom_point(data = tibble(label=c("S₁","B₁","S₂","B₂","S₃","B₃"),
                           x=c(1,1,3,3,5,5), y=c(2,1,2,1,2,1)),
             aes(x=x,y=y), shape=21, size=14, fill="#deebf7",
             color="#1e3a5f", stroke=1.3) +
  geom_text(data = tibble(label=c("S₁","B₁","S₂","B₂","S₃","B₃"),
                           x=c(1,1,3,3,5,5), y=c(2,1,2,1,2,1)),
            aes(x=x,y=y,label=label), size=3.0, fontface="bold", color="#1e3a5f") +
  geom_point(data = ri_nodes, aes(x=x,y=y),
             shape=21, size=14, fill="#f3e5f5", color="#9b59b6", stroke=1.3) +
  geom_text(data = ri_nodes, aes(x=x, y=y, label=label),
            size=2.5, fontface="bold", color="#9b59b6") +
  annotate("segment", x=1.95, xend=1.05, y=2.43, yend=2.07, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("segment", x=1.95, xend=3.05, y=2.43, yend=2.07, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("segment", x=1.95, xend=4.95, y=2.43, yend=2.07, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("segment", x=1.95, xend=1.05, y=0.63, yend=0.97, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("segment", x=1.95, xend=3.05, y=0.63, yend=0.97, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("segment", x=1.95, xend=4.95, y=0.63, yend=0.97, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("text", x=3, y=0.45,
           label="Between-person: chronic stress-proneness\nand chronic risk-taking co-occur (no within-person dynamic)",
           hjust=0.5, size=2.6, color="gray30", lineheight=1.3) +
  labs(title = "Story C: stable trait drives both variables") +
  xlim(0.2, 5.8) + ylim(0.25, 2.75) +
  theme_void(base_size=10) +
  theme(plot.title=element_text(face="bold", size=9.5, hjust=0.5,
                                margin=margin(b=3)))

p_a + p_b + p_c
```

All three stories produce **the same bivariate correlation pattern** between $S_t$ and $B_{t+1}$, and between $B_t$ and $S_{t+1}$. The classic CLPM cannot distinguish them because it does not separate stable between-person covariation (Story C) from genuine within-person dynamics (Stories A and B). The RI-CLPM separates these sources explicitly, making it the correct model for the within-person question.

## The Random Intercept CLPM (RI-CLPM)

Hamaker, Kuiper, & Grasman (2015) proposed the **Random Intercept CLPM**, which solves this by separating two sources of variation:

1. **Between-person:** Stable trait levels (random intercepts) that explain why some people are chronically higher on S or B
2. **Within-person:** Fluctuations around each person's stable level, and the cross-lagged dynamics *among those fluctuations*

```{r ri-clpm}
ri_clpm_model <- '
  # Random intercepts (stable between-person levels)
  RI_S =~ 1*S1 + 1*S2 + 1*S3
  RI_B =~ 1*B1 + 1*B2 + 1*B3

  # Within-person components (deviations from stable level)
  wS1 =~ 1*S1; wS2 =~ 1*S2; wS3 =~ 1*S3
  wB1 =~ 1*B1; wB2 =~ 1*B2; wB3 =~ 1*B3

  # Constrain observed variances to 0 (all captured by RI + within)
  S1 ~~ 0*S1; S2 ~~ 0*S2; S3 ~~ 0*S3
  B1 ~~ 0*B1; B2 ~~ 0*B2; B3 ~~ 0*B3

  # Within-person cross-lagged dynamics
  wS2 ~ a_SS*wS1 + a_SB*wB1
  wS3 ~ a_SS*wS2 + a_SB*wB2
  wB2 ~ a_BB*wB1 + a_BS*wS1
  wB3 ~ a_BB*wB2 + a_BS*wS2

  # Random intercept variances and covariance
  RI_S ~~ RI_S; RI_B ~~ RI_B; RI_S ~~ RI_B

  # Within-person residual variances
  wS1 ~~ wS1; wB1 ~~ wB1
  wS2 ~~ wS2; wB2 ~~ wB2
  wS3 ~~ wS3; wB3 ~~ wB3
  wS2 ~~ wB2; wS3 ~~ wB3

  # Mean structure
  RI_S ~ 1; RI_B ~ 1
  wS1 ~ 0; wB1 ~ 0
'

m_ri_clpm <- sem(ri_clpm_model, data = df_clpm, estimator = "ML")

parameterEstimates(m_ri_clpm) |>
  filter(op == "~", lhs %in% c("wS2","wS3","wB2","wB3")) |>
  select(lhs, rhs, label, est, se, pvalue) |>
  knitr::kable(digits = 3,
    caption = "RI-CLPM: within-person cross-lagged effects (controls for stable between-person differences)")
```

## Comparing CLPM vs. RI-CLPM

| Feature | Classic CLPM | RI-CLPM |
|---------|-------------|---------|
| Controls for stable between-person differences? | No | Yes |
| Cross-lagged = within-person dynamics? | No (confounded) | Yes |
| More parameters? | No | Yes (random intercepts) |
| Better for causal within-person questions? | No | Yes |

::: {.callout-warning icon=false}
## Assumption: Stationarity of Cross-Lagged Effects

**What it means:** The CLPM (and RI-CLPM) typically constrains cross-lagged effects to be equal across waves. This assumes the dynamics do not change over the study period.

**How to check:** Fit an unconstrained model (different path coefficients at each wave) and compare fit to the constrained model.
:::

::: {.callout-warning icon=false}
## Assumption: No Omitted Dynamics

**What it means:** Both models assume all relevant variables are included. An omitted third variable that drives both S and B could produce spurious cross-lagged effects.

**How to check:** Think carefully about what else changes over time that could explain both variables. Include plausible third variables as additional nodes.

**If this fails:** Cross-lagged coefficients are observational associations, not causal estimates. Limit the interpretation to "S at time t predicts B at time t+1 above and beyond B's own stability, while controlling for the variables in the model" — not "S causes B."
:::

## Assumption Table

```{r clpm-assumption-table, echo=FALSE}
knitr::kable(
  tibble(
    Assumption = c(
      "Stationarity of cross-lagged effects",
      "No omitted time-varying confounders",
      "Separation of within- and between-person variation (RI-CLPM)",
      "Sufficient waves to identify dynamics"
    ),
    `Plain-language meaning` = c(
      "The S→B and B→S path coefficients are the same at every wave",
      "No third variable drives both S and B over time, creating spurious cross-lagged paths",
      "The random intercept fully captures each person's stable trait level in both variables",
      "At least 3 waves are needed; 2 waves cannot distinguish cross-lagged from concurrent correlation"
    ),
    `What to inspect` = c(
      "Fit a model with wave-varying paths; compare to constrained model with LRT or modification indices",
      "Review your theory: what else changes over time that predicts both variables?",
      "Inspect the variance of the random intercept factors; check if CLPM and RI-CLPM give similar cross-lags",
      "Report the number of waves and the interval between them"
    ),
    `Failure looks like` = c(
      "Large modification indices for freeing specific cross-lagged paths at particular waves",
      "Cross-lagged coefficients change substantially when a third variable is added to the model",
      "CLPM and RI-CLPM cross-lags differ substantially — stable traits were confounding the classic model",
      "Only 2 waves: no information about temporal dynamics beyond the concurrent correlation"
    ),
    `What to do next` = c(
      "Free the path at the problematic wave; interpret as time-varying effect; report the heterogeneity",
      "Add the confound to the model; if not measurable, narrow the causal claim",
      "Use RI-CLPM for within-person claims; report both models and explain why they differ",
      "Collect more waves; analyze as a change-score model instead"
    )
  ),
  caption = "CLPM / RI-CLPM assumption table"
)
```

---

## Reviewer Challenge

::: {.callout-tip}
## What a Reviewer Would Ask

**Challenge 1:** "Your classic CLPM shows stress predicts borrowing — but isn't it possible that chronically stressed people also tend to borrow more? How do you know this is a dynamic within-person effect rather than a stable between-person correlation?"
The correct response: show the RI-CLPM results alongside the classic CLPM. If the RI-CLPM cross-lagged path is smaller or non-significant while the classic CLPM path is large, stable trait differences were driving the apparent dynamic effect. Report both models and interpret accordingly.

**Challenge 2:** "You constrain the cross-lagged paths to be equal across waves — but what if the stress-borrowing relationship changes over the study (e.g., as the economic environment changes)?"
Fit the unconstrained model (free paths at each wave) and compare fit with a likelihood ratio test. If the unconstrained model fits significantly better, report the wave-varying paths and discuss what might explain the change.

**Challenge 3:** "You have only 3 waves — is that enough to reliably estimate cross-lagged dynamics?"
Three waves is the minimum. With 3 waves, you estimate one set of cross-lagged paths (wave 1 → wave 2 and wave 2 → wave 3, constrained equal). This is adequate for a preliminary test but limited in power and unable to detect non-stationarity. More waves provide better estimates. Report this limitation explicitly.
:::

## Minimum Reporting Checklist

- [ ] State the number of waves and the interval between them
- [ ] Report whether cross-lagged paths were constrained equal across waves and the LRT result for that constraint
- [ ] Report both classic CLPM and RI-CLPM results and explain the difference
- [ ] Report the random intercept variance in the RI-CLPM — large variance means stable traits matter
- [ ] Interpret cross-lagged coefficients as "A at time t predicts B at time t+1, controlling for B's own stability" — not as causal effects
- [ ] If significant cross-lagged path: show it holds in RI-CLPM before making within-person claims
