---
title: "Applied Identification Strategies for Panel Data"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(plm)
library(AER)
library(broom)
library(broom.mixed)
library(ggplot2)
library(patchwork)
library(sandwich)
library(lmtest)
set.seed(42)
```

## Section 1: Introduction — Module 3 Meets Panel Data

Module 3 covered six sources of endogeneity and six identification strategies. The discussion there was largely cross-sectional: a researcher observes units at one point in time and tries to isolate causal effects despite confounding. This chapter applies each of those strategies to **longitudinal panel data**, where units are observed repeatedly across time. The time dimension creates both new problems (time-varying confounders, dynamic selection) and new opportunities (the ability to difference out stable confounders using fixed effects, the ability to exploit policy rollout dates as natural experiments).

The table below maps Module 3 threats to their panel-data solutions.

| Module 3 Threat | Panel Solution | Key R Function |
|---|---|---|
| Time-invariant OVB (Part 1) | Fixed effects (Module 1) | `lmer(..., REML=F)` or `plm(model="within")` |
| Time-varying endogeneity (Part 2 IV) | Panel IV | `ivreg(y ~ x + factor(unit) | z + factor(unit))` |
| Policy adoption at known date (Part 5 DiD) | TWFE + event study | `plm(model="within", effect="twoways")` |
| Temporal threshold (Part 4 RDD) | Interrupted Time Series | `lmer(y ~ t + post + t_post + (1|unit))` |
| Non-random treatment adoption (Part 6 PS) | Time-varying PS matching | `matchit()` at each adoption wave |
| Non-random attrition (Part 3 Selection) | Available-case lmer + MI | Module 4 Part 9 |

: Module 3 identification strategies mapped to their panel-data analogs {tbl-colwidths="[30,35,35]"}

The sections that follow walk through hands-on implementations of the most commonly needed strategies: two-way fixed effects DiD with an event study, interrupted time series (the temporal RDD), and panel IV.

---

## Section 2: Panel DiD with Two-Way Fixed Effects

### Setup and simulation

The **two-way fixed effects (TWFE)** estimator is the workhorse of panel difference-in-differences. It absorbs unit fixed effects (removing stable confounders) and time fixed effects (removing common shocks), then estimates the treatment effect from within-unit, within-period variation.

We simulate 60 firms observed over 8 quarters. Thirty firms adopt a management initiative at quarter 5. The true average treatment effect on the treated (ATT) is 10 productivity points.

```{r twfe-simulate}
n_firms    <- 60
n_quarters <- 8

firm_eff    <- rnorm(n_firms, 0, 8)
quarter_eff <- rnorm(n_quarters, 0, 3)

df_twfe <- expand_grid(firm = 1:n_firms, quarter = 1:n_quarters) |>
  mutate(
    treated      = as.integer(firm <= 30),
    post         = as.integer(quarter >= 5),
    D            = treated * post,
    f_eff        = firm_eff[firm],
    q_eff        = quarter_eff[quarter],
    productivity = 50 + 10 * D + f_eff + q_eff + rnorm(n(), 0, 4),
    firm         = factor(firm),
    quarter      = factor(quarter)
  )
```

### Three equivalent estimators

```{r twfe-models}
# 1. Manual 2x2 DiD
did_manual <- with(
  df_twfe |>
    group_by(treated, post) |>
    summarise(y = mean(productivity), .groups = "drop"),
  (y[treated == 1 & post == 1] - y[treated == 1 & post == 0]) -
    (y[treated == 0 & post == 1] - y[treated == 0 & post == 0])
)
cat("Manual DiD estimate:", round(did_manual, 2), "\n")

# 2. Regression DiD with firm-clustered SEs
m_reg_did  <- lm(productivity ~ treated * post, data = df_twfe)
se_cluster <- coeftest(m_reg_did, vcov = vcovCL(m_reg_did, cluster = ~firm))

# 3. TWFE via plm (two-way within)
pdata_twfe <- pdata.frame(
  df_twfe |> mutate(firm_id = as.integer(firm), qtr = as.integer(quarter)),
  index = c("firm_id", "qtr")
)
m_twfe <- plm(productivity ~ D, data = pdata_twfe,
              model = "within", effect = "twoways")
```

```{r twfe-compare, echo=FALSE}
naive_se <- sqrt(diag(vcov(m_reg_did)))["treated:post"]
clust_se <- se_cluster["treated:post", "Std. Error"]

tibble(
  Method           = c("Manual DiD", "Regression DiD (naive SE)",
                       "Regression DiD (clustered SE)", "TWFE via plm"),
  Estimate         = c(did_manual,
                       coef(m_reg_did)["treated:post"],
                       se_cluster["treated:post", "Estimate"],
                       coef(m_twfe)["D"]),
  `Std. Error`     = c(NA, naive_se, clust_se, NA)
) |>
  knitr::kable(digits = 3,
               caption = "All three approaches target the true ATT ≈ 10. Clustered SEs account for within-firm correlation of productivity shocks over time; they may be larger or smaller than naive SEs depending on the dependence structure.")
```

::: {.callout-warning icon=false}
## Why Cluster Standard Errors in Panel DiD?

Observations within the same firm are not independent — productivity shocks persist over time. Treating them as independent (naive OLS SEs) understates uncertainty and inflates Type I error rates. The rule of thumb: cluster at the level of treatment assignment. Here, treatment varies at the firm level, so we cluster on `firm`. With fewer than ~30 clusters, consider wild cluster bootstrap (`fwildclusterboot` package).
:::

### Event study: testing parallel trends

The parallel trends assumption — that treated and control firms would have followed the same trajectory absent treatment — is not directly testable, but we can look for violations by checking whether pre-treatment differences in trends are already present.

```{r event-study-plot, fig.width=8, fig.height=4.5}
m_event <- lm(productivity ~ treated * factor(quarter), data = df_twfe)

event_df <- broom::tidy(m_event, conf.int = TRUE) |>
  filter(str_detect(term, "treated:")) |>
  mutate(
    quarter = as.integer(str_extract(term, "[0-9]+")),
    pre     = quarter < 5
  )

ggplot(event_df, aes(x = quarter, y = estimate, color = pre)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 4.5, linetype = "dotted", color = "gray30") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("FALSE" = "#1e3a5f", "TRUE" = "#aec6e8"),
    labels = c("FALSE" = "Post-treatment", "TRUE" = "Pre-treatment (should \u2248 0)")
  ) +
  annotate("text", x = 4.7, y = max(event_df$conf.high) * 0.9,
           label = "Treatment\nadoption", hjust = 0, size = 3.5, color = "gray30") +
  labs(
    title   = "Event study: pre-period near zero supports parallel trends",
    subtitle = "Post-period coefficients trace the materializing treatment effect",
    x       = "Quarter",
    y       = "Estimated treatment effect vs. Q1",
    color   = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

Pre-period interaction coefficients (light blue) are scattered around zero — consistent with parallel pre-trends. Post-period coefficients (dark blue) step up after quarter 5 and hover near the true effect of 10. If the pre-period coefficients showed a systematic upward or downward slope, that would be evidence of pre-existing differential trends and would undermine the DiD identification assumption.

::: {.callout-tip}
## Reading an Event Study Plot

- **Pre-period coefficients near zero**: parallel trends plausible. Fail this check and DiD is suspect.
- **Sharp jump at treatment date**: consistent with an immediate treatment effect.
- **Gradual post-period increase**: treatment effect builds over time (dynamic effects).
- **Anticipation effects** (pre-period coefficients trending upward before treatment): firms may be responding in advance; adjust the "post" indicator.
:::

---

## Section 3: Interrupted Time Series (Temporal RDD)

### Concept

**Interrupted Time Series (ITS)** applies the RDD logic to time rather than a continuous running variable. When a policy takes effect at a known calendar date $c$, we look for a discontinuity in the time series at that date — a sudden level jump and/or a change in slope. The key identification assumption mirrors RDD's continuity assumption: absent the policy, the time trend would have continued smoothly through $c$.

The ITS model decomposes the response into:

$$Y_{it} = \beta_0 + \beta_1 t + \underbrace{\beta_2 D_t}_{\text{level jump}} + \underbrace{\beta_3 (t - c)\, D_t}_{\text{slope change}} + u_{0i} + \varepsilon_{it}$$

where $D_t = \mathbf{1}[t \geq c]$, $t - c$ is time elapsed since the intervention (zero before), $u_{0i}$ is a unit random intercept (absorbing stable unit differences), and $\varepsilon_{it}$ is idiosyncratic error.

- $\beta_1$: pre-intervention monthly trend
- $\beta_2$: immediate level change at month $c$ (the "jump")
- $\beta_3$: additional slope change after intervention (change in rate of change)

### Simulation

We simulate 40 retail stores observed for 30 months. A new pricing policy takes effect at month 16. The true DGP has a $+5$ level jump and a $+0.3$/month additional slope change.

```{r its-simulate}
n_stores <- 40
n_months <- 30
store_eff <- rnorm(n_stores, 0, 6)

df_its <- expand_grid(store = 1:n_stores, month = 1:n_months) |>
  mutate(
    post   = as.integer(month >= 16),
    t_post = pmax(month - 15, 0),       # months elapsed since policy; 0 before
    s_eff  = store_eff[store],
    sales  = 60 + 0.8 * month + 5 * post + 0.3 * t_post + s_eff + rnorm(n(), 0, 5),
    store  = factor(store)
  )
```

```{r its-model}
m_its <- lmer(sales ~ month + post + t_post + (1 | store), data = df_its)

broom.mixed::tidy(m_its, effects = "fixed") |>
  knitr::kable(digits = 3,
               caption = "ITS model: pre-trend (month), level jump (post), slope change (t_post). True values: 0.8, 5, 0.3.")
```

### Visualization

```{r its-plot, fig.width=8.5, fig.height=4.5}
store_avg <- df_its |>
  group_by(month) |>
  summarise(mean_sales = mean(sales), .groups = "drop")

# Predicted values from lmer for an average store
pred_df <- tibble(
  month   = 1:n_months,
  post    = as.integer(month >= 16),
  t_post  = pmax(month - 15, 0),
  store   = factor(1)          # prediction for store 1 (fixed effects only for plot)
)
pred_df$pred <- predict(m_its, newdata = pred_df, re.form = NA)

ggplot(store_avg, aes(x = month, y = mean_sales)) +
  geom_point(color = "gray60", size = 2) +
  geom_line(data = pred_df, aes(y = pred), color = "#1e3a5f", linewidth = 1.2) +
  geom_vline(xintercept = 15.5, linetype = "dashed", color = "#c0392b") +
  annotate("segment",
           x = 15.5, xend = 15.5, y = 67, yend = 72,
           arrow = arrow(length = unit(0.25, "cm")), color = "#c0392b") +
  annotate("text", x = 16, y = 73,
           label = expression(beta[2]*": level jump"), hjust = 0,
           color = "#c0392b", size = 3.5) +
  annotate("text", x = 22, y = 83,
           label = expression(beta[3]*": slope change"),
           color = "#2980b9", size = 3.5) +
  labs(
    title    = "Interrupted Time Series: pricing policy at month 16",
    subtitle = "Level jump (\u03b22) and slope change (\u03b23) both visible at the cutoff",
    x        = "Month",
    y        = "Store-average sales"
  ) +
  theme_minimal(base_size = 13)
```

### Pre-trend test

The ITS equivalent of the RDD density (McCrary) test is to check whether the pre-intervention trend is genuinely linear, or whether a non-linear pre-trend exists that could masquerade as a level jump. A simple diagnostic:

```{r its-pretrend}
df_pre <- df_its |> filter(month < 16)
m_pretrend <- lmer(sales ~ month + I(month^2) + (1 | store), data = df_pre)
broom.mixed::tidy(m_pretrend, effects = "fixed") |>
  knitr::kable(digits = 4,
               caption = "Pre-period model with quadratic term. If I(month^2) is far from zero, the pre-trend is curved and the level-jump interpretation is suspect.")
```

A quadratic coefficient near zero in the pre-period supports the linearity assumption. A large quadratic coefficient means the trajectory was already bending at the cutoff — the ITS level jump would be confounded.

---

## Section 4: Panel IV

### When fixed effects are not enough

Unit fixed effects remove **time-invariant** confounders. But if a confounder varies both across units and across time — and is correlated with the predictor — FE alone is insufficient. Panel IV combines the within-unit differencing of FE with an instrument that shifts the endogenous predictor but has no direct path to the outcome.

### Simulation: R&D investment and patents

We simulate 200 firms over 6 years. R&D investment is endogenous (high-quality firms invest more). Industry-level tax incentives serve as the instrument (relevant to R&D, excluded from the patent equation except through R&D).

```{r panel-iv-simulate}
n_f  <- 200
n_y  <- 6
firm_q       <- rnorm(n_f, 0, 5)                  # unobserved firm quality
industry     <- rep(1:10, each = n_f / 10)
tax_incentive <- rnorm(10, 0, 2)[industry]         # industry-level component

df_piv <- expand_grid(firm = 1:n_f, year = 1:n_y) |>
  mutate(
    fq      = firm_q[firm],
    # Time-varying quality shock: firm FE removes fq mean, but NOT this shock
    fq_tv   = fq + rnorm(n(), 0, 3),
    tax_inc = tax_incentive[firm] + rnorm(n(), 0, 0.5),  # firm-year realization
    # R&D: endogenous — high-quality firm-years invest more (time-varying confound)
    rd      = 20 + 2 * fq_tv + 3 * tax_inc + rnorm(n(), 0, 5),
    # Patents: causal effect of R&D = 0.5; also depends on time-varying quality
    patents = 5 + 0.5 * rd + 1.5 * fq_tv + rnorm(n(), 0, 3),
    firm    = factor(firm),
    year    = factor(year)
  )
```

```{r panel-iv-models}
# OLS + FE: still biased by time-varying fq-rd correlation
m_ols_fe <- lm(patents ~ rd + factor(firm) + factor(year), data = df_piv)

# IV + FE: use tax_inc as instrument, include firm + year FE
m_iv_fe  <- ivreg(
  patents ~ rd + factor(firm) + factor(year) |
    tax_inc + factor(firm) + factor(year),
  data = df_piv
)

# First-stage F-statistic
fs <- lm(rd ~ tax_inc + factor(firm) + factor(year), data = df_piv)
fs_F <- summary(fs)$fstatistic[1]
cat("First-stage F on instrument:", round(fs_F, 1), "\n")
```

```{r panel-iv-compare, echo=FALSE}
tibble(
  Model    = c("OLS + unit/year FE", "IV + unit/year FE"),
  Estimate = c(coef(m_ols_fe)["rd"], coef(m_iv_fe)["rd"]),
  `Std. Error` = c(
    sqrt(diag(vcov(m_ols_fe)))["rd"],
    sqrt(diag(vcov(m_iv_fe)))["rd"]
  ),
  `True value` = 0.5
) |>
  knitr::kable(digits = 4,
               caption = "OLS + FE is biased by time-varying firm quality (fq_tv), which firm fixed effects cannot remove. IV is consistent under the exclusion restriction; single-sample estimates may deviate from the true 0.5 due to finite-sample noise.")
```

::: {.callout-warning icon=false}
## Weak Instruments in Panel Data

Panel fixed effects remove between-unit variation and retain only **within-unit** variation. This can substantially weaken an instrument that looked strong in cross-sectional data: the industry tax incentive varies mostly across industries, and much of that cross-industry variation is soaked up by the firm fixed effects. Always check the first-stage $F$-statistic using the **demeaned** (within-transformed) data, not the raw data. The conventional threshold is $F > 10$; for inference-robust methods (Anderson-Rubin test), the threshold matters less.

```r
# Safe check: first stage on demeaned data via plm
pdata_piv   <- pdata.frame(df_piv |> mutate(firm_id=as.integer(firm), yr=as.integer(year)),
                             index=c("firm_id","yr"))
fs_within   <- plm(rd ~ tax_inc, data=pdata_piv, model="within", effect="twoways")
summary(fs_within)
```
:::

---

## Section 5: Time-Varying Propensity Score Matching

When units adopt a treatment at **different waves** — a staggered rollout rather than a single cutoff — the two-by-two DiD logic breaks down because the "post" period differs across treated units. Two-way FE with staggered adoption can produce negatively-weighted treatment effect estimates when treatment effects are heterogeneous across cohorts (the "negative weighting" problem documented by Callaway & Sant'Anna, 2021).

One practical alternative is **time-varying propensity score matching**, which handles staggered adoption explicitly:

**Step 1**: For each adoption wave $w$, identify:
- Firms that just adopted in wave $w$ (the "treated" at wave $w$).
- Firms that have not yet adopted by wave $w$ (the "not-yet-treated" controls).

**Step 2**: Compute the propensity score at wave $w$ among eligible units:

```r
glm(adopted_now ~ lag_outcome + size + industry + region,
    family = binomial,
    data   = eligible_at_wave_w)
```

**Step 3**: Match treated firms to the most similar not-yet-treated firms **at that wave** using nearest-neighbor matching within the wave-specific pool.

**Step 4**: Estimate the ATT on the matched sample, using the matched controls' post-adoption outcomes as the counterfactual.

This approach produces a cohort-specific ATT for each adoption wave, which can be aggregated into an overall ATT. It is conceptually equivalent to the Callaway & Sant'Anna (2021) estimator (available in the `did` package) and is doubly robust when combined with outcome regression.

::: {.callout-tip}
## PS-Weighted DiD: Doubly Robust Panel Estimation

Combining time-varying PS weighting with DiD regression produces an estimator that is **doubly robust**: consistent if either (a) the propensity score model is correctly specified, or (b) parallel trends holds conditional on covariates. Neither assumption alone is sufficient for ordinary DiD; neither alone is sufficient for ordinary PSM. The combination of both provides extra insurance. See `drdid` package for implementation.
:::

---

## Section 6: Decision Tree — Which Strategy for Which Panel Problem?

::: {.callout-tip}
## Which Identification Strategy for Which Panel Problem?

| Problem | Method | Key assumption | Module reference |
|---------|---------|----------------|-----------------|
| Stable unit-level confounders | Unit fixed effects | Time-invariant endogeneity | M1 Parts 1–2 |
| Time-varying endogeneity | Panel IV | Instrument relevance + exclusion restriction | M3 Part 2 |
| Parallel policy rollout | TWFE DiD | Parallel trends | M3 Part 5 |
| Staggered adoption | Callaway-Sant'Anna / `drdid` | Parallel trends by cohort | M3 Part 5 |
| Temporal policy cutoff | ITS | Continuity of time trend at cutoff | M3 Part 4 |
| Non-random treatment adoption | Time-varying PS matching | Selection on observables | M3 Part 6 |
| Non-random attrition | Available-case lmer + MI | MAR | M4 Part 9 |

The first question is always: **what creates the identifying variation?** Time, a threshold, an instrument, or a policy rollout? The answer determines the method.
:::
