---
title: "Cross-Lagged Panel Models"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lavaan)
library(ggplot2)
library(patchwork)
set.seed(2024)
```

# Cross-Lagged Panel Models

## Two Variables, Evolving Together

Suppose you want to understand the relationship between financial stress and risky borrowing behavior over time. Does financial stress push people toward riskier borrowing? Or does risky borrowing create financial strain? Or do they mutually reinforce each other?

To answer this, you need data on both variables at multiple time points, and you need a model that can disentangle the *temporal ordering* of their relationship. The **cross-lagged panel model (CLPM)** is the standard tool for this question.

## The Classic CLPM

In a two-wave, two-variable CLPM with variables S (financial stress) and B (risky borrowing):

$$S_{t+1} = a_{SS} \cdot S_t + a_{SB} \cdot B_t + \varepsilon_{S,t+1}$$
$$B_{t+1} = a_{BB} \cdot B_t + a_{BS} \cdot S_t + \varepsilon_{B,t+1}$$

Four parameters:
- $a_{SS}$, $a_{BB}$: **autoregressive effects** — stability of each variable over time
- $a_{SB}$: **cross-lagged effect** of B at time $t$ on S at time $t+1$
- $a_{BS}$: **cross-lagged effect** of S at time $t$ on B at time $t+1$

```{r clpm-setup}
# Simulate 3-wave panel data
n <- 200
n_waves <- 3

# True DGP: financial stress -> risky borrowing (not the reverse)
S <- matrix(NA, n, n_waves)
B <- matrix(NA, n, n_waves)

S[, 1] <- rnorm(n, 50, 8)   # baseline financial stress
B[, 1] <- rnorm(n, 45, 7)   # baseline borrowing propensity

for (t in 2:n_waves) {
  S[, t] <- 0.6 * S[, t-1] + 0.1 * B[, t-1] + rnorm(n, 0, 5)  # small B->S effect
  B[, t] <- 0.6 * B[, t-1] + 0.3 * S[, t-1] + rnorm(n, 0, 5)  # strong S->B effect
}

df_clpm <- as.data.frame(cbind(S, B)) |>
  setNames(c(paste0("S", 1:3), paste0("B", 1:3)))
```

```{r fit-clpm}
# Classic CLPM in lavaan
clpm_model <- '
  # Autoregressive effects
  S2 ~ a_SS * S1
  S3 ~ a_SS * S2
  B2 ~ a_BB * B1
  B3 ~ a_BB * B2

  # Cross-lagged effects
  S2 ~ a_SB * B1
  S3 ~ a_SB * B2
  B2 ~ a_BS * S1
  B3 ~ a_BS * S2

  # Residual covariances (within same wave)
  S2 ~~ B2
  S3 ~~ B3

  # Variance of wave 1
  S1 ~~ S1
  B1 ~~ B1
  S1 ~~ B1
'

m_clpm <- sem(clpm_model, data = df_clpm)
parameterEstimates(m_clpm) |>
  filter(op == "~") |>
  select(lhs, rhs, label, est, se, pvalue) |>
  knitr::kable(digits = 3,
    caption = "CLPM: cross-lagged path from S→B is strong; B→S is weak")
```

Good news: the CLPM detects the true asymmetric relationship (financial stress predicts future risky borrowing, not the other way around).

## The Problem with Classic CLPM

The classic CLPM conflates **within-person dynamics** with **between-person differences**. Consider: people who are chronically financially stressed tend to also engage in chronically riskier borrowing (stable between-person correlation). The cross-lagged paths in the classic CLPM pick up both this stable between-person covariation *and* the genuine within-person dynamic — they cannot be separated.

This means even if there is **no within-person causal relationship** between financial stress and risky borrowing (i.e., changes in financial stress within a person do not cause changes in borrowing behavior), the classic CLPM can show significant "cross-lagged" effects simply because chronically stressed people also tend to borrow more riskily.

### The Same Observed Pattern, Three Different Stories

The same observed correlations between financial stress and risky borrowing across waves are consistent with three completely different causal stories. The classic CLPM cannot distinguish them without extending the model.

```{r clpm-stories, echo=FALSE, fig.width=11, fig.height=3.6, message=FALSE, warning=FALSE}
make_clpm_panel <- function(arrows_df, title, note) {
  nodes <- tibble(
    label = c("S₁", "B₁", "S₂", "B₂", "S₃", "B₃"),
    x     = c(1, 1, 3, 3, 5, 5),
    y     = c(2, 1, 2, 1, 2, 1)
  )
  p <- ggplot() +
    geom_point(data = nodes, aes(x = x, y = y),
               shape = 21, size = 14, fill = "#deebf7",
               color = "#1e3a5f", stroke = 1.3) +
    geom_text(data = nodes, aes(x = x, y = y, label = label),
              size = 3.0, fontface = "bold", color = "#1e3a5f")
  for (i in seq_len(nrow(arrows_df))) {
    p <- p + annotate("segment",
                      x     = arrows_df$x1[i], xend = arrows_df$x2[i],
                      y     = arrows_df$y1[i], yend = arrows_df$y2[i],
                      arrow = arrow(length = unit(0.18, "cm"), type = "closed"),
                      color = arrows_df$col[i], linewidth = 1.0,
                      linetype = arrows_df$lty[i])
  }
  p + annotate("text", x = 3, y = 0.45, label = note,
               hjust = 0.5, size = 2.6, color = "gray30", lineheight = 1.3) +
    labs(title = title) +
    xlim(0.2, 5.8) + ylim(0.25, 2.7) +
    theme_void(base_size = 10) +
    theme(plot.title = element_text(face = "bold", size = 9.5, hjust = 0.5,
                                    margin = margin(b = 3)))
}

# AR arrows common to all panels (autoregressive stability)
ar <- tibble(x1 = c(1.5,1.5,3.5,3.5), x2 = c(2.5,2.5,4.5,4.5),
             y1 = c(2,1,2,1),          y2 = c(2,1,2,1),
             col = rep("#888888", 4),  lty = rep("solid", 4))

# Story A: S→B only
a_arrows <- bind_rows(ar,
  tibble(x1=c(1.4,3.4), x2=c(2.6,4.6), y1=c(1.85,1.85), y2=c(1.15,1.15),
         col=rep("#e63946",2), lty=rep("solid",2)))
p_a <- make_clpm_panel(a_arrows,
  "Story A: stress \u2192 risky borrowing",
  "Within-person: stress changes cause\nborrowing changes one wave later")

# Story B: B→S only
b_arrows <- bind_rows(ar,
  tibble(x1=c(1.4,3.4), x2=c(2.6,4.6), y1=c(1.15,1.15), y2=c(1.85,1.85),
         col=rep("#2d6a4f",2), lty=rep("solid",2)))
p_b <- make_clpm_panel(b_arrows,
  "Story B: risky borrowing \u2192 stress",
  "Within-person: borrowing changes cause\nstress changes one wave later")

# Story C: stable trait causes both (RI), no within-person dynamic
ri_nodes <- tibble(
  label = c("RI_S", "RI_B"),
  x = c(2, 2), y = c(2.5, 0.6)
)
p_c <- ggplot() +
  geom_point(data = tibble(label=c("S₁","B₁","S₂","B₂","S₃","B₃"),
                           x=c(1,1,3,3,5,5), y=c(2,1,2,1,2,1)),
             aes(x=x,y=y), shape=21, size=14, fill="#deebf7",
             color="#1e3a5f", stroke=1.3) +
  geom_text(data = tibble(label=c("S₁","B₁","S₂","B₂","S₃","B₃"),
                           x=c(1,1,3,3,5,5), y=c(2,1,2,1,2,1)),
            aes(x=x,y=y,label=label), size=3.0, fontface="bold", color="#1e3a5f") +
  geom_point(data = ri_nodes, aes(x=x,y=y),
             shape=21, size=14, fill="#f3e5f5", color="#9b59b6", stroke=1.3) +
  geom_text(data = ri_nodes, aes(x=x, y=y, label=label),
            size=2.5, fontface="bold", color="#9b59b6") +
  annotate("segment", x=1.95, xend=1.05, y=2.43, yend=2.07, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("segment", x=1.95, xend=3.05, y=2.43, yend=2.07, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("segment", x=1.95, xend=4.95, y=2.43, yend=2.07, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("segment", x=1.95, xend=1.05, y=0.63, yend=0.97, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("segment", x=1.95, xend=3.05, y=0.63, yend=0.97, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("segment", x=1.95, xend=4.95, y=0.63, yend=0.97, color="#9b59b6",
           linewidth=1.0, linetype="dashed",
           arrow=arrow(length=unit(0.17,"cm"),type="closed")) +
  annotate("text", x=3, y=0.45,
           label="Between-person: chronic stress-proneness\nand chronic risk-taking co-occur (no within-person dynamic)",
           hjust=0.5, size=2.6, color="gray30", lineheight=1.3) +
  labs(title = "Story C: stable trait drives both variables") +
  xlim(0.2, 5.8) + ylim(0.25, 2.75) +
  theme_void(base_size=10) +
  theme(plot.title=element_text(face="bold", size=9.5, hjust=0.5,
                                margin=margin(b=3)))

p_a + p_b + p_c
```

All three stories produce **the same bivariate correlation pattern** between $S_t$ and $B_{t+1}$, and between $B_t$ and $S_{t+1}$. The classic CLPM cannot distinguish them because it does not separate stable between-person covariation (Story C) from genuine within-person dynamics (Stories A and B). The RI-CLPM separates these sources explicitly, making it the correct model for the within-person question.

## The Random Intercept CLPM (RI-CLPM)

Hamaker, Kuiper, & Grasman (2015) proposed the **Random Intercept CLPM**, which solves this by separating two sources of variation:

1. **Between-person:** Stable trait levels (random intercepts) that explain why some people are chronically higher on S or B
2. **Within-person:** Fluctuations around each person's stable level, and the cross-lagged dynamics *among those fluctuations*

```{r ri-clpm}
ri_clpm_model <- '
  # Random intercepts (stable between-person levels)
  RI_S =~ 1*S1 + 1*S2 + 1*S3
  RI_B =~ 1*B1 + 1*B2 + 1*B3

  # Within-person components (deviations from stable level)
  wS1 =~ 1*S1; wS2 =~ 1*S2; wS3 =~ 1*S3
  wB1 =~ 1*B1; wB2 =~ 1*B2; wB3 =~ 1*B3

  # Constrain observed variances to 0 (all captured by RI + within)
  S1 ~~ 0*S1; S2 ~~ 0*S2; S3 ~~ 0*S3
  B1 ~~ 0*B1; B2 ~~ 0*B2; B3 ~~ 0*B3

  # Within-person cross-lagged dynamics
  wS2 ~ a_SS*wS1 + a_SB*wB1
  wS3 ~ a_SS*wS2 + a_SB*wB2
  wB2 ~ a_BB*wB1 + a_BS*wS1
  wB3 ~ a_BB*wB2 + a_BS*wS2

  # Random intercept variances and covariance
  RI_S ~~ RI_S; RI_B ~~ RI_B; RI_S ~~ RI_B

  # Within-person residual variances
  wS1 ~~ wS1; wB1 ~~ wB1
  wS2 ~~ wS2; wB2 ~~ wB2
  wS3 ~~ wS3; wB3 ~~ wB3
  wS2 ~~ wB2; wS3 ~~ wB3

  # Mean structure
  RI_S ~ 1; RI_B ~ 1
  wS1 ~ 0; wB1 ~ 0
'

m_ri_clpm <- sem(ri_clpm_model, data = df_clpm, estimator = "ML")

parameterEstimates(m_ri_clpm) |>
  filter(op == "~", lhs %in% c("wS2","wS3","wB2","wB3")) |>
  select(lhs, rhs, label, est, se, pvalue) |>
  knitr::kable(digits = 3,
    caption = "RI-CLPM: within-person cross-lagged effects (controls for stable between-person differences)")
```

## Comparing CLPM vs. RI-CLPM

| Feature | Classic CLPM | RI-CLPM |
|---------|-------------|---------|
| Controls for stable between-person differences? | No | Yes |
| Cross-lagged = within-person dynamics? | No (confounded) | Yes |
| More parameters? | No | Yes (random intercepts) |
| Better for causal within-person questions? | No | Yes |

::: {.callout-warning icon=false}
## Assumption: Stationarity of Cross-Lagged Effects

**What it means:** The CLPM (and RI-CLPM) typically constrains cross-lagged effects to be equal across waves. This assumes the dynamics do not change over the study period.

**How to check:** Fit an unconstrained model (different path coefficients at each wave) and compare fit to the constrained model.
:::

::: {.callout-warning icon=false}
## Assumption: No Omitted Dynamics

**What it means:** Both models assume all relevant variables are included. An omitted third variable that drives both S and B could produce spurious cross-lagged effects.

**How to check:** Think carefully about what else changes over time that could explain both variables. Include plausible third variables as additional nodes.
:::
