---
title: "Time Structure, Fixed Effects, and Seasonal Cycles"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(broom)
library(ggplot2)
library(patchwork)
```

# Time Structure, Fixed Effects, and Seasonal Cycles

::: {.callout-note icon=false}
## The Time Structure Question

> **What happens when you ignore calendar time in a longitudinal analysis?** Three things can go wrong simultaneously: a background drift (the world is gradually changing) inflates or deflates your treatment estimate; a seasonal cycle (spending is higher in December) creates patterns correlated with when treatment was assigned; and period-specific shocks get absorbed into the wrong parameters.
>
> **How do you know time is a confounder?** Residuals from your model should have zero mean at every time point. If the residuals slope upward, you have unmodeled drift. If they oscillate with a 12-month period, you have unmodeled seasonality. Both are diagnosable from simple plots.
>
> **What is the fix?** Include time as a predictor — as a linear drift term, Fourier terms for cyclical patterns, or period fixed effects (one dummy per calendar period). The choice depends on how much parametric structure you are willing to assume about the time DGP.
:::

## One Dataset, Many Models

This section works differently from most of the module. We build **one realistic longitudinal dataset**, disclose the true data-generating process completely, then run it through five progressively better specifications. Each model section follows the same five-step inferential workflow:

1. **Assumptions** — what does this model assume about the data?
2. **Pre-fit checks** — what can we verify before fitting?
3. **Fitting the model** — code and key output
4. **Against the DGP** — how well does this model recover the truth?
5. **Interpreting the results** — what do the coefficients mean, and what errors would we make stopping here?

All residual diagnostic plots use the **same y-axis scale**, so the shrinking of residuals across models is directly visible.

---

## The True Data-Generating Process

A fintech app tracks monthly spending ($) for 120 US users over 24 months (January 2023 through December 2024). The platform introduced a voluntary **Loyalty Program** partway through, and the research question is: *does program membership cause higher spending?*

::: {.callout-tip icon=false}
## True DGP Parameters

| Parameter | Symbol | True value |
|:----------|:-------|:----------:|
| Loyalty program effect | $\beta_{\text{loyalty}}$ | **$35** |
| Background drift (per month) | $\beta_{\text{drift}}$ | $3 |
| Seasonal amplitude | $A$ | $60 |
| Between-person SD | $\sigma_u$ | $80 |
| Within-person residual SD | $\sigma_e$ | $40 |
| Person baseline mean | $\mu$ | $500 |

$$
Y_{it} = \underbrace{\alpha_i}_{\text{person baseline}}
       + \underbrace{35 \cdot D_{it}}_{\text{loyalty effect}}
       + \underbrace{3 \cdot t}_{\text{background drift}}
       + \underbrace{60 \cos\!\left(\tfrac{2\pi\, m_t}{12}\right)}_{\text{seasonal cycle}}
       + \underbrace{\varepsilon_{it}}_{\text{noise}}
$$

where $t$ is the calendar month (1–24), $m_t$ is the month of year (1–12), and $\varepsilon_{it} \sim \mathcal{N}(0, 40^2)$.
:::

- *Each row* is one user–month. There are $120 \times 24 = 2{,}880$ rows.
- $\alpha_i \sim \mathcal{N}(500, 80^2)$: the person's stable monthly baseline spending.
- **Background drift** ($+\$3$/month): platform growth and inflation lift everyone's spending over the 24 months, totalling +$72 by the end regardless of program status.
- **Seasonal cycle**: December spending is $60 above the monthly average; June is $60 below — a $120 annual swing.
- **Enrollment is staggered and non-random**: higher-spending users are more likely to join; a November–December marketing push concentrates enrollment in the highest-spending months. Both create confounding.

```{r generate-data}
set.seed(2025)

n_hh     <- 120
n_months <- 24

# True DGP parameters
true_loyalty   <- 35   # causal program effect ($/month)
true_trend     <- 3    # background drift ($/month)
true_amplitude <- 60   # seasonal amplitude ($)
true_person_sd <- 80   # between-person SD ($)
true_resid_sd  <- 40   # within-person residual SD ($)

# Person baselines
person_baselines <- rnorm(n_hh, 500, true_person_sd)

# Staggered enrollment:
#   (a) higher-spending users enroll earlier
#   (b) Nov/Dec marketing campaign doubles enrollment probability
enroll_month <- rep(NA_integer_, n_hh)
for (hh in seq_len(n_hh)) {
  for (m in 4:22) {
    moy       <- ((m - 1L) %% 12L) + 1L
    p_base    <- 0.025
    p_person  <- max(0, (person_baselines[hh] - 500) / 80) * 0.025
    p_holiday <- if (moy %in% c(11L, 12L)) 0.05 else 0
    p         <- min(p_base + p_person + p_holiday, 0.15)
    if (runif(1) < p) { enroll_month[hh] <- m; break }
  }
}

df <- expand_grid(person = seq_len(n_hh), month = seq_len(n_months)) |>
  mutate(
    month_of_year      = ((month - 1L) %% 12L) + 1L,
    cos_12             = cos(2 * pi * month_of_year / 12),
    sin_12             = sin(2 * pi * month_of_year / 12),
    enroll_m           = enroll_month[person],
    loyalty            = as.integer(!is.na(enroll_m) & month >= enroll_m),
    person_component   = person_baselines[person],
    drift_component    = true_trend * month,
    seasonal_component = true_amplitude * cos_12,
    true_y             = person_component + drift_component +
                         seasonal_component + true_loyalty * loyalty,
    noise              = rnorm(n(), 0, true_resid_sd),
    spending           = true_y + noise,
    person_f           = factor(person)
  )

cat(sum(!is.na(enroll_month)), "of", n_hh,
    sprintf("users enrolled (%.0f%%)\n", 100 * mean(!is.na(enroll_month))))
```

```{r toy-table}
df |>
  filter(person == 12) |>
  select(person, month, month_of_year, loyalty, spending,
         person_component, drift_component, seasonal_component) |>
  slice(c(1:3, 11:13)) |>
  mutate(across(where(is.double), round, 1)) |>
  knitr::kable(
    col.names = c("Person", "Month", "MoY", "Loyal",
                  "Spending", "Baseline", "Drift", "Seasonal"),
    caption = "Six rows from one user — each observed value is baseline + drift + seasonal + noise."
  )
```

---

## Fitting All Models and Setting a Common Y-Axis

```{r batch-fit, include=FALSE}
# Fit all five models silently so that (a) the global y-axis range for
# residual plots can be computed from the worst-case model, and
# (b) objects are available before individual sections reference them.

m1 <- lm(spending ~ loyalty, data = df)
m2 <- lm(spending ~ loyalty + month, data = df)
m3 <- lm(spending ~ loyalty + month + cos_12 + sin_12, data = df)
m4 <- lmer(spending ~ loyalty + month + cos_12 + sin_12 + (1 | person_f),
           data = df)
m5 <- lm(spending ~ loyalty + factor(month) + factor(person_f), data = df)

# The worst-case residuals come from Model 1 (no time controls).
# Lock the y-axis to that range so shrinkage is visible across all models.
r1_cal  <- tapply(residuals(m1), df$month,         mean)
r1_seas <- tapply(residuals(m1), df$month_of_year, mean)
all_r1  <- c(r1_cal, r1_seas)
pad     <- 15
global_y <- c(floor(min(all_r1)) - pad, ceiling(max(all_r1)) + pad)
```

```{r diag-helper}
# 3-panel residual dashboard used after every model.
# trend_ok / seasonal_ok drive the green/red colour coding.
# y_lim is shared across all calls so residual shrinkage is visible.

show_diag <- function(model_obj,
                      trend_ok    = FALSE,
                      seasonal_ok = FALSE,
                      y_lim       = global_y) {

  resids  <- residuals(model_obj)
  ok_col  <- "#2d6a4f"
  bad_col <- "#cb181d"
  ci_b    <- qnorm(0.975) / sqrt(n_months)

  make_sub <- function(ok, label_yes, label_no)
    if (ok) paste0("✓ ", label_yes) else paste0("✗ ", label_no)

  # Panel 1 — mean residuals by calendar month
  p1 <- df |> mutate(r = resids) |>
    group_by(month) |> summarise(mr = mean(r), .groups = "drop") |>
    ggplot(aes(x = month, y = mr)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
    geom_line(linewidth = 1.1,
              color = if (trend_ok) ok_col else bad_col) +
    { if (!trend_ok) list(
        geom_smooth(method = "lm", se = FALSE, linetype = "dotted",
                    color = bad_col, linewidth = 0.8))
      else list() } +
    coord_cartesian(ylim = y_lim) +
    scale_x_continuous(breaks = c(1,6,12,18,24),
      labels = c("J'23","Jn'23","J'24","Jn'24","D'24")) +
    labs(title = "Residuals by calendar month",
         subtitle = make_sub(trend_ok,
           "Background drift absorbed",
           "Background drift not captured"),
         x = NULL, y = "Mean residual ($)") +
    theme_minimal(base_size = 11) +
    theme(plot.subtitle = element_text(
      color = if (trend_ok) ok_col else bad_col, size = 9.5))

  # Panel 2 — mean residuals by month-of-year
  p2 <- df |> mutate(r = resids) |>
    group_by(month_of_year) |> summarise(mr = mean(r), .groups = "drop") |>
    ggplot(aes(x = month_of_year, y = mr)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
    geom_line(linewidth = 1.1,
              color = if (seasonal_ok) ok_col else bad_col) +
    geom_point(size = 2.2,
               color = if (seasonal_ok) ok_col else bad_col) +
    coord_cartesian(ylim = y_lim) +
    scale_x_continuous(breaks = 1:12,
      labels = c("J","F","M","A","M","J","J","A","S","O","N","D")) +
    labs(title = "Residuals by month of year",
         subtitle = make_sub(seasonal_ok,
           "Seasonal cycle absorbed",
           "Seasonal cycle not captured"),
         x = NULL, y = "Mean residual ($)") +
    theme_minimal(base_size = 11) +
    theme(plot.subtitle = element_text(
      color = if (seasonal_ok) ok_col else bad_col, size = 9.5))

  # Panel 3 — ACF of panel-mean residuals
  panel_mr <- df |> mutate(r = resids) |>
    group_by(month) |> summarise(mr = mean(r), .groups = "drop") |>
    pull(mr)
  acf_r  <- acf(panel_mr, plot = FALSE, lag.max = 13)
  acf_df <- tibble(lag = as.integer(acf_r$lag[-1]),
                   a   = as.numeric(acf_r$acf[-1]))
  lag12_ok  <- abs(acf_df$a[acf_df$lag == 12]) <= ci_b
  acf_col   <- if (lag12_ok) ok_col else bad_col
  acf_sub   <- if (lag12_ok) "✓ No lag-12 spike"
               else           "✗ Lag-12 spike (annual cycle)"

  p3 <- acf_df |>
    ggplot(aes(x = lag, y = a)) +
    geom_hline(yintercept = 0) +
    geom_hline(yintercept = c(-1,1)*ci_b, linetype="dashed", color="#2171b5") +
    geom_segment(aes(xend=lag, yend=0), color="grey60", linewidth=0.7) +
    geom_point(color="grey50", size=1.8) +
    geom_segment(data=filter(acf_df, lag==12),
                 aes(xend=lag, yend=0), color=acf_col, linewidth=1.5) +
    geom_point(data=filter(acf_df, lag==12), color=acf_col, size=3.5) +
    scale_x_continuous(breaks=c(2,4,6,8,10,12)) +
    labs(title = "ACF of panel-mean residuals",
         subtitle = acf_sub,
         x = "Lag (months)", y = "Autocorr.") +
    theme_minimal(base_size = 11) +
    theme(plot.subtitle = element_text(color = acf_col, size = 9.5))

  p1 | p2 | p3
}
```

---

## Pre-Modeling: Understanding the Data

Before fitting anything, explore the raw data. The goal is to see the three confounding structures directly.

```{r viz-raw, fig.width=11, fig.height=4}
set.seed(1)
sample_ids <- sample(seq_len(n_hh), 20)

p_spaghetti <- df |>
  filter(person %in% sample_ids) |>
  mutate(grp = factor(if_else(!is.na(enroll_m),
                               "Enrolls at some point", "Never enrolls"))) |>
  ggplot(aes(x = month, y = spending, group = person, color = grp)) +
  geom_line(alpha = 0.45, linewidth = 0.5) +
  scale_color_manual(values = c("Enrolls at some point" = "#2171b5",
                                "Never enrolls"         = "#cb181d")) +
  scale_x_continuous(breaks = c(1,6,12,18,24),
    labels = c("Jan\n2023","Jun\n2023","Jan\n2024","Jun\n2024","Dec\n2024")) +
  labs(title = "Individual spending trajectories\n(20 sampled users)",
       x = NULL, y = "Monthly spending ($)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_mean <- df |>
  group_by(month) |> summarise(ms = mean(spending), .groups = "drop") |>
  ggplot(aes(x = month, y = ms)) +
  geom_line(color = "#555", linewidth = 1) +
  geom_smooth(method="lm", se=FALSE, color="#e6550d", linetype="dashed") +
  annotate("text", x=19, y=565,
           label="Background drift\n(+$3/month)", color="#e6550d", size=3.3) +
  scale_x_continuous(breaks = c(1,6,12,18,24),
    labels = c("Jan\n2023","Jun\n2023","Jan\n2024","Jun\n2024","Dec\n2024")) +
  labs(title = "Cohort-average spending\nover time",
       x = NULL, y = "Mean spending ($)") +
  theme_minimal(base_size = 11)

seas_avg <- df |> group_by(month_of_year) |>
  summarise(ms = mean(spending), .groups = "drop")
p_seasonal <- seas_avg |>
  ggplot(aes(x = month_of_year, y = ms)) +
  geom_line(linewidth = 1.1, color = "#6a51a3") +
  geom_point(size = 2.5, color = "#6a51a3") +
  annotate("segment", x=12, xend=12,
           y=min(seas_avg$ms), yend=max(seas_avg$ms),
           color="#6a51a3", linetype="dotted") +
  annotate("text", x=11.4, y=max(seas_avg$ms)-4,
           label="Dec\npeak", size=3, color="#6a51a3", hjust=1) +
  scale_x_continuous(breaks=1:12,
    labels=c("J","F","M","A","M","J","J","A","S","O","N","D")) +
  labs(title = "Spending by month of year\n(seasonal pattern)",
       x = "Month of year", y = "Mean spending ($)") +
  theme_minimal(base_size = 11)

p_spaghetti | p_mean | p_seasonal
```

**Visible patterns:** the average spending line slopes upward (background drift), the month-of-year plot shows a clear $120 annual swing, and blue users (future enrollees) tend to start higher than red users (never-enrollees) — signalling the person-level confound.

The table below makes the person-level confound concrete. Even before enrollment, future loyalty members spend more than those who never join.

```{r person-confound-check}
df |>
  mutate(
    group = case_when(
      is.na(enroll_m)      ~ "Never enrolls",
      month < enroll_m     ~ "Will enroll (pre-enrollment months)",
      TRUE                 ~ NA_character_
    )
  ) |>
  filter(!is.na(group)) |>
  group_by(group) |>
  summarise(
    n_users        = n_distinct(person),
    months_observed = n(),
    mean_spending  = mean(spending),
    sd_spending    = sd(spending),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1,
    caption = "Pre-enrollment comparison: future enrollees already spend more — person-level selection bias is present before any program effect.")
```

---

## Model 1: Naïve Pooled Regression

### Step 1 — What this model assumes

::: {.callout-warning icon=false}
## Model 1 Assumptions

1. **Independence:** Each of the 2,880 user-month rows is an independent draw. Spending by user 47 in March is unrelated to their spending in April.
2. **No omitted variables:** Nothing predicts both program enrollment and spending except what is already in the model.
3. **No time trend:** Expected spending is constant across all 24 months — no background drift, no seasonal pattern.
4. **No person-level heterogeneity:** All users have the same baseline spending; the only thing that differs is loyalty status.

**All four are violated here.** This model is a deliberate starting point — not a defensible analysis.
:::

### Step 2 — Pre-fit checks

The three diagnostic plots above already show the violations:
- Average spending slopes upward → background drift exists (assumption 3 fails)
- Spending cycles with a ±$60 annual amplitude → seasonal pattern exists (assumption 3 fails)
- Future enrollees spend more before enrollment → person-level heterogeneity correlated with enrollment (assumption 4 and 2 fail)

### Step 3 — Fitting the model

```{r model1-fit}
m1 <- lm(spending ~ loyalty, data = df)

bind_rows(
  tidy(m1) |> filter(term == "loyalty") |>
    mutate(parameter = "Loyalty effect ($)", true_value = true_loyalty),
  tibble(term = "sigma", estimate = sigma(m1), std.error = NA_real_,
         statistic = NA_real_, p.value = NA_real_,
         parameter = "Residual SD ($)", true_value = true_resid_sd)
) |>
  mutate(bias = round(estimate - true_value, 2)) |>
  select(parameter, estimate, std.error, true_value, bias) |>
  knitr::kable(digits = 2,
    caption = "Model 1 — naïve OLS. True loyalty effect = $35.")
```

### Step 4 — Against the true DGP

```{r model1-diag, fig.width=11, fig.height=4}
show_diag(m1, trend_ok = FALSE, seasonal_ok = FALSE)
```

The residual SD ($\approx$ `r round(sigma(m1))`) is roughly `r round(sigma(m1)/true_resid_sd, 1)`× the true within-person noise of \$40 — the model is attributing background drift, seasonal swings, and between-person differences to the residual.

### Step 5 — Interpreting the results

The loyalty estimate is roughly **$50–60** — far above the true $35. Three biases are compounding:

- *Person selection:* higher-spending users enrolled, so loyal users look wealthier than they are relative to the program.
- *Drift absorption:* enrolled users accumulate more post-enrollment months in the later, higher-drift portion of the window. The model reads the rising tide as a loyalty benefit.
- *Seasonal absorption:* enrollment peaks in November–December, the highest-spending months. The model reads holiday spending as a loyalty benefit.

**What you would wrongly conclude:** "Loyalty program membership increases monthly spending by $50+." **What is actually true:** the causal effect is $35.

---

## Model 2: Controlling for Background Drift

### Step 1 — What this model assumes

::: {.callout-warning icon=false}
## Model 2 Assumptions (new, on top of Model 1)

1. *(Inherits all Model 1 assumptions except the "no time trend" one)*
2. **Linear drift:** The background increase in spending is constant at exactly \$*k*/month — it does not accelerate, decelerate, or experience any structural breaks across the 24-month window.

**The seasonal cycle assumption is still violated.** This model removes one confound but not both time-based confounds.
:::

### Step 2 — Pre-fit checks

The left panel from Model 1's diagnostic above shows a clear upward slope in calendar-month residuals — a textbook symptom of unmodeled linear drift. Adding `month` as a predictor addresses this directly.

### Step 3 — Fitting the model

```{r model2-fit}
m2 <- lm(spending ~ loyalty + month, data = df)

bind_rows(
  tidy(m2) |> filter(term %in% c("loyalty", "month")) |>
    mutate(parameter = c("Loyalty effect ($)", "Background drift ($/month)"),
           true_value = c(true_loyalty, true_trend)),
  tibble(term = "sigma", estimate = sigma(m2), std.error = NA_real_,
         statistic = NA_real_, p.value = NA_real_,
         parameter = "Residual SD ($)", true_value = true_resid_sd)
) |>
  mutate(bias = round(estimate - true_value, 2)) |>
  select(parameter, estimate, std.error, true_value, bias) |>
  knitr::kable(digits = 2,
    caption = "Model 2 — loyalty + linear drift.")
```

### Step 4 — Against the true DGP

```{r model2-diag, fig.width=11, fig.height=4}
show_diag(m2, trend_ok = TRUE, seasonal_ok = FALSE)
```

The calendar-month panel is now flat — the background drift is absorbed ✓. The seasonal wave in the month-of-year panel is **unchanged**, and the lag-12 ACF spike remains ✗. The residual SD has shrunk from Model 1 but is still well above $40 because seasonal variance is still in the error.

### Step 5 — Interpreting the results

The drift coefficient is close to the true $3/month — the model has correctly recovered that parameter. The loyalty estimate has improved but is still inflated: **the seasonal confound is entirely intact**. November–December enrollment still creates an artificial positive correlation between loyalty status and spending, because spending is naturally high in those months regardless of the program.

**What you would wrongly conclude:** "Loyalty program membership increases spending by ~$40–45 after accounting for overall platform growth." **What is actually true:** the causal effect is still $35.

---

## Model 3: Capturing the Seasonal Cycle

### Step 1 — What this model assumes

::: {.callout-warning icon=false}
## Model 3 Assumptions (new, on top of Model 2)

1. *(Inherits drift and independence assumptions from Model 2)*
2. **Annual periodicity:** The seasonal cycle repeats exactly every 12 months and can be represented by a single Fourier pair.
3. **Single harmonic:** One cosine + one sine term at frequency 1/12 per month captures all meaningful periodic variation — there are no quarterly or weekly sub-cycles.

**The person-level confound assumption is still violated.** Higher-spending users are still more likely to enroll, but this model cannot remove that between-person bias.
:::

### Step 2 — Pre-fit checks

The middle and right panels from Model 2 are diagnostic: the seasonal wave ($\approx \pm\$60$) and the lag-12 ACF spike both persist after controlling for drift. This motivates the Fourier terms.

::: {.callout-note icon=false}
## Why Two Terms (Cosine + Sine)?

Any single-frequency cycle can be written as:
$$f(t) = A\cos\!\left(\tfrac{2\pi m_t}{12}\right) + B\sin\!\left(\tfrac{2\pi m_t}{12}\right) = R\cos\!\left(\tfrac{2\pi m_t}{12} - \phi\right)$$
where $R = \sqrt{A^2 + B^2}$ is the amplitude and $\phi = \arctan(B/A)$ is the phase (peak month). Including both terms lets OLS determine amplitude and peak month from the data without you specifying either. In our DGP the true cycle is pure cosine ($B = 0$, December peak), so the sine coefficient should estimate near zero.
:::

### Step 3 — Fitting the model

```{r model3-fit}
m3 <- lm(spending ~ loyalty + month + cos_12 + sin_12, data = df)

bind_rows(
  tidy(m3) |> filter(term %in% c("loyalty","month","cos_12","sin_12")) |>
    mutate(parameter  = c("Loyalty effect ($)", "Background drift ($/month)",
                          "Seasonal cosine ($)", "Seasonal sine ($)"),
           true_value = c(true_loyalty, true_trend, true_amplitude, 0)),
  tibble(term = "sigma", estimate = sigma(m3), std.error = NA_real_,
         statistic = NA_real_, p.value = NA_real_,
         parameter = "Residual SD ($)", true_value = true_resid_sd)
) |>
  mutate(bias = round(estimate - true_value, 2)) |>
  select(parameter, estimate, std.error, true_value, bias) |>
  knitr::kable(digits = 2,
    caption = "Model 3 — loyalty + drift + Fourier seasonal.")
```

### Step 4 — Against the true DGP

```{r model3-diag, fig.width=11, fig.height=4}
show_diag(m3, trend_ok = TRUE, seasonal_ok = TRUE)
```

Both time panels are flat and the lag-12 ACF spike is gone ✓. The residual SD has dropped to near the true $40 — most of the explainable variance is now captured. The loyalty estimate is much closer to $35, but still slightly high.

### Step 5 — Interpreting the results

The seasonal cosine coefficient is near the true $60, and the sine is near zero — exactly as the DGP specifies. Both time-based confounds are now resolved.

The remaining upward bias in the loyalty estimate reflects the **person-level confound**: Model 3 is still comparing higher-spending enrollees to lower-spending non-enrollees. OLS uses the between-person contrast because it has no person controls. The residual SD being close to $40 confirms the time structure is well-captured; the remaining issue is cross-sectional, not temporal.

**What you would wrongly conclude:** "Loyalty program membership increases spending by ~$38–42 after accounting for platform growth and seasonality." Closer to the truth, but still inflated by ~$5 from person selection.

---

## Model 4: The Full Specification — Person Random Effects

### Step 1 — What this model assumes

::: {.callout-warning icon=false}
## Model 4 Assumptions (new, on top of Model 3)

1. *(Inherits drift, Fourier, and independence assumptions from Model 3)*
2. **Normal person random effects:** Each user's stable baseline spending $\alpha_i$ is drawn from $\mathcal{N}(0, \sigma_u^2)$.
3. **Independence of person effects and predictors:** After conditioning on the drift, seasonal terms, and loyalty status, a user's random intercept is uncorrelated with the loyalty predictor. This is the *exchangeability* assumption — the program was not assigned based on unmeasured time-varying characteristics.
4. **Within-person residual independence:** After removing person effects, the month-to-month residuals within a person are exchangeable — no remaining autocorrelation.

**This is the first model that is correctly specified for our DGP.** All four assumptions hold by construction because we simulated the data.
:::

### Step 2 — Pre-fit checks

The person-level selection check shown in the pre-modeling section confirmed: future enrollees spent more *before* enrolling. Model 3's loyalty estimate is still inflated (see above), even though time is correctly specified. The culprit is between-person confounding — exactly what person random effects address.

The test: does the loyalty coefficient change when person random effects are added? If yes, the between-person contrast was driving part of the Model 3 estimate.

### Step 3 — Fitting the model

```{r model4-fit}
m4 <- lmer(spending ~ loyalty + month + cos_12 + sin_12 + (1 | person_f),
           data = df)

person_sd_est <- as.data.frame(VarCorr(m4)) |>
  filter(grp == "person_f") |> pull(sdcor)

bind_rows(
  tidy(m4, effects = "fixed") |>
    filter(term %in% c("loyalty","month","cos_12","sin_12")) |>
    mutate(parameter  = c("Loyalty effect ($)", "Background drift ($/month)",
                          "Seasonal cosine ($)", "Seasonal sine ($)"),
           true_value = c(true_loyalty, true_trend, true_amplitude, 0)),
  tibble(term = "sigma_person", estimate = person_sd_est,
         std.error = NA_real_, statistic = NA_real_, p.value = NA_real_,
         parameter = "Person SD ($)", true_value = true_person_sd),
  tibble(term = "sigma_resid", estimate = sigma(m4),
         std.error = NA_real_, statistic = NA_real_, p.value = NA_real_,
         parameter = "Residual SD ($)", true_value = true_resid_sd)
) |>
  mutate(bias = round(estimate - true_value, 2)) |>
  select(parameter, estimate, std.error, true_value, bias) |>
  knitr::kable(digits = 2,
    caption = "Model 4 — full specification: loyalty + drift + Fourier + person RE.")
```

### Step 4 — Against the true DGP

```{r model4-diag, fig.width=11, fig.height=4}
show_diag(m4, trend_ok = TRUE, seasonal_ok = TRUE)
```

All three time-based panels pass ✓. The loyalty estimate, all time-structure parameters, person SD, and residual SD are all near their true values. As a final check, plot person-mean residuals against person mean spending — the random intercept should have absorbed all baseline differences.

```{r model4-person-diag, fig.width=5.5, fig.height=3.5}
df |>
  mutate(resid = residuals(m4)) |>
  group_by(person) |>
  summarise(mr = mean(resid), ms = mean(spending), .groups = "drop") |>
  ggplot(aes(x = ms, y = mr)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(alpha = 0.5, color = "#2171b5", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#2171b5", linewidth = 0.8) +
  labs(title = "Person-mean residuals\nvs. person mean spending",
       subtitle = "✓ No pattern — person RE absorbed all baseline differences",
       x = "Person mean spending ($)", y = "Mean residual ($)") +
  theme_minimal(base_size = 12) +
  theme(plot.subtitle = element_text(color = "#2d6a4f"))
```

### Step 5 — Interpreting the results

Every DGP parameter is now accurately recovered. The key shift from Model 3 to Model 4:

- **Loyalty estimate:** drops from ~$38–42 to ~$35, removing the person-selection upward bias
- **Person SD:** correctly estimates ~$80 — confirming that the model partitioned between-person variance appropriately
- **Residual SD:** near $40 — only genuine within-person noise remains unexplained

The loyalty coefficient now answers: *among the same user, does spending rise in months when they are enrolled compared to months when they are not, after accounting for background drift and seasonality?* This is a within-person causal comparison, not a cross-sectional average.

**Correct conclusion:** "Controlling for user-level spending habits, background platform drift, and seasonal cycles, loyalty program membership is associated with ~$35 higher monthly spending."

---

## Model 5: Period Fixed Effects (Month Dummies)

### Step 1 — What this model assumes

::: {.callout-warning icon=false}
## Model 5 Assumptions

1. **Each calendar month has its own intercept**, absorbing all calendar-period effects — no parametric shape assumed for the time function.
2. **Each person has their own intercept** (person dummies), absorbing all stable between-person differences.
3. **Loyalty effect is homogeneous:** The program effect is the same for all users in all months. No user-specific or time-varying treatment effect.
4. **Strict exogeneity:** After removing person and month fixed effects, loyalty status is uncorrelated with the residual.

Model 5 is more robust to misspecification of the time structure than Model 4 (no assumption of linearity or periodicity), but less efficient (23 extra parameters for month dummies vs. 3).
:::

### Step 2 — Pre-fit checks

Model 4 already passes all residual diagnostics — this model serves as a robustness check, not a correction. The question is whether relaxing the parametric time assumption changes the loyalty estimate materially.

### Step 3 — Fitting the model

```{r model5-fit}
m5 <- lm(spending ~ loyalty + factor(month) + factor(person_f), data = df)

tidy(m5) |>
  filter(term == "loyalty") |>
  mutate(parameter  = "Loyalty effect ($)",
         true_value = true_loyalty,
         bias       = round(estimate - true_loyalty, 2),
         sigma      = round(sigma(m5), 2)) |>
  select(parameter, estimate, std.error, true_value, bias, sigma) |>
  knitr::kable(digits = 2,
    caption = "Model 5 — two-way FE: loyalty + month dummies + person dummies.")
```

### Step 4 — Against the true DGP

The loyalty estimate is close to $35, matching Model 4. The residual SD is similar. No residual diagnostic is plotted because the month-dummy FE makes the diagnostic structure trivial — the calendar-month residuals are exactly zero by construction.

### Step 5 — Interpreting the results

Model 5 confirms Model 4 non-parametrically. The loyalty effect is ~$35 regardless of whether we model the time function as "linear drift + one Fourier pair" or as "unrestricted month intercepts." This robustness is reassuring: the parametric time assumption in Model 4 is not driving the result.

**When to prefer Model 5 over Model 4:**

| Situation | Prefer |
|:----------|:-------|
| Time structure is smooth and well-understood | Model 4 — more efficient |
| Potential structural breaks, non-linear drift | Model 5 — no parametric assumption |
| Need to extrapolate beyond the observation window | Model 4 only |
| Reviewer skeptical of parametric time control | Model 5 as robustness check |

---

## Benchmark: All Five Models Against the True DGP

```{r comparison, fig.width=9, fig.height=5}
extract_loyalty <- function(mod, label) {
  if (inherits(mod, "lmerMod"))
    broom.mixed::tidy(mod, effects="fixed") |> filter(term=="loyalty") |>
    mutate(model=label)
  else
    broom::tidy(mod) |> filter(term=="loyalty") |> mutate(model=label)
}

comp <- bind_rows(
  extract_loyalty(m1, "M1: Naive"),
  extract_loyalty(m2, "M2: + Background drift"),
  extract_loyalty(m3, "M3: + Seasonal cycle"),
  extract_loyalty(m4, "M4: + Person RE (full)"),
  extract_loyalty(m5, "M5: Two-way FE")
) |>
  mutate(
    model   = factor(model, levels = rev(unique(model))),
    ci_low  = estimate - 1.96 * std.error,
    ci_high = estimate + 1.96 * std.error,
    covers  = ci_low <= true_loyalty & true_loyalty <= ci_high
  )

p_forest <- comp |>
  ggplot(aes(x = estimate, y = model)) +
  geom_vline(xintercept = true_loyalty, linetype = "dashed",
             linewidth = 1, color = "black") +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high),
                 height = 0.25, color = "grey40") +
  geom_point(aes(color = covers), size = 4) +
  scale_color_manual(values = c("TRUE" = "#2d6a4f", "FALSE" = "#cb181d"),
                     labels = c("TRUE" = "CI covers $35",
                                "FALSE" = "CI misses $35")) +
  annotate("text", x = true_loyalty + 0.8, y = 5.45,
           label = "True\n$35", hjust = 0, size = 3.5) +
  labs(title = "Loyalty estimates across\nfive specifications",
       x = "Estimated loyalty effect ($)", y = NULL, color = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

sigma_seq <- c(sigma(m1), sigma(m2), sigma(m3), sigma(m4), sigma(m5))
bias_seq  <- comp |> arrange(desc(model)) |>
  mutate(bias = estimate - true_loyalty) |> pull(bias)

p_track <- tibble(
  mod    = factor(rev(levels(comp$model)), levels = rev(levels(comp$model))),
  Bias   = bias_seq,
  `Residual SD` = sigma_seq
) |>
  pivot_longer(c(Bias, `Residual SD`)) |>
  ggplot(aes(x = mod, y = value, group = name, color = name)) +
  geom_hline(data = tibble(name  = c("Bias", "Residual SD"),
                           truth = c(0, true_resid_sd)),
             aes(yintercept = truth, color = name),
             linetype = "dashed", linewidth = 0.8, inherit.aes = FALSE) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Bias" = "#e6550d",
                                "Residual SD" = "#6a51a3")) +
  scale_x_discrete(labels = function(x) str_remove(x, " \\(full\\)")) +
  labs(title = "Bias and residual SD\nacross model sequence",
       subtitle = "Dashed = true value",
       x = NULL, y = "Value ($)", color = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle=30, hjust=1, size=9),
        legend.position = "bottom")

p_forest | p_track
```

```{r comparison-table}
comp |>
  arrange(desc(model)) |>
  mutate(
    bias    = round(estimate - true_loyalty, 1),
    ci_low  = round(ci_low, 1),
    ci_high = round(ci_high, 1),
    covers  = if_else(covers, "Yes ✓", "No ✗")
  ) |>
  select(model, estimate, std.error, bias, ci_low, ci_high, covers) |>
  knitr::kable(digits = 2,
    col.names = c("Model","Estimate","SE","Bias",
                  "95% CI low","95% CI high","CI covers $35?"),
    caption = "All five models vs. the true loyalty effect of $35.")
```

---

## Practical Guide: Diagnostic Sequence

When you encounter a new longitudinal dataset, run these five checks in order:

```{r diag-guide, echo=FALSE}
knitr::kable(
  tibble(
    Step = 1:5,
    `Plot or check` = c(
      "Mean outcome by calendar month",
      "Mean outcome by month-of-year",
      "Mean residuals by calendar month",
      "Mean residuals by month-of-year",
      "ACF of panel-mean residuals"
    ),
    `What you are looking for` = c(
      "Upward or downward slope → background drift present",
      "Repeated wave → seasonal cycle present",
      "Still drifts after controls → drift not fully captured",
      "Still waves after controls → seasonality not captured",
      "Spike at lag k → k-period autocorrelation remaining"
    ),
    Remedy = c(
      "Add calendar month as a linear predictor",
      "Add Fourier sin/cos or season-of-year dummies",
      "Add polynomial trend or period dummies",
      "Add Fourier terms or seasonal dummies",
      "Model residual autocorrelation (Part 5: AR structures)"
    )
  ),
  caption = "Five-step diagnostic sequence for longitudinal time structure."
)
```

::: {.callout-warning icon=false}
## Reviewer Challenge

**What a skeptical reviewer would say:** "You control for a linear background drift, but what if the drift is nonlinear — accelerating in the second year or interrupted by a market shock in month 14? Your loyalty estimate could absorb a nonlinear surge that happened to coincide with program rollout."

**Observable warning sign:** After adding a linear drift term, residuals vs. calendar month show a curve rather than a flat line.

**What claim survives:** Direction is likely correct. Magnitude depends on how closely the parametric time model matches the true DGP. Model 5 (period FE) is the robustness check — if it agrees with Model 4, the parametric assumption is not driving the finding.
:::

---

## Setting the Stage for Growth Curve Models

This section treated **time as a confounder** to be controlled. The next section, Growth Curve Models, treats **time as the phenomenon**: how does each person's trajectory unfold, and why do trajectories differ across people?

The same five-step inferential workflow — assumptions, pre-fit checks, fitting, DGP comparison, interpretation — applies to every model in the rest of the module. The diagnostic toolkit (residual-by-time, seasonal check, ACF) remains the standard first pass before interpreting any parameter estimate.
