---
title: "Latent Growth Curve Models (SEM)"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lavaan)
library(lme4)
library(ggplot2)
set.seed(111)
```

# Latent Growth Curve Models

## Two Languages for the Same Model

The growth curve model from Part 3 and the **Latent Growth Curve Model (LGC)** from structural equation modeling are mathematically equivalent. They answer the same questions using different parameterizations and different software.

Understanding both languages is valuable because:
1. Many psychology researchers encounter LGC models in SEM training (Mplus, lavaan) before they encounter MLM-based growth models
2. The SEM framing makes certain things explicit that are implicit in MLM — particularly the role of measurement error and the ability to model multiple growth processes simultaneously
3. Reviewing both frames deepens understanding of what the model is actually doing

## The SEM Framing

In the latent growth curve model, the trajectory is represented by **two latent variables**:
- **Intercept factor (η_i):** The latent initial level for each person
- **Slope factor (η_s):** The latent rate of change for each person

Observed scores at each time point are modeled as:
$$Y_{ti} = 1 \cdot \eta_i + \lambda_t \cdot \eta_s + \varepsilon_{ti}$$

where $\lambda_t$ are **factor loadings** that encode the time metric (e.g., $\lambda_1 = 0, \lambda_2 = 1, \lambda_3 = 2$ for linear growth), and $\varepsilon_{ti}$ are measurement residuals.

The intercept factor has loadings of 1.0 at all time points. The slope factor has loadings equal to the time coding.

## Fitting an LGC Model in lavaan

### JDM Context: Savings Growth After Loyalty Program Enrollment

A financial services firm tracks 150 customers' monthly net savings for 5 months after they enroll in a loyalty-and-coaching program. The key substantive questions:

- **On average**, how much do customers save at enrollment, and how fast does savings grow?
- **How much do they vary?** Do some customers start high and stay high? Do some start low but grow quickly?
- **Do initial level and growth rate covary?** Customers who start with high savings might grow more slowly (ceiling effects) or more quickly (financially engaged customers who compound good habits).

These are precisely the questions the latent growth curve model answers. The intercept factor captures each customer's baseline savings level; the slope factor captures their personal savings growth trajectory.

```{r lgc-simulation}
# 150 loyalty program customers tracked for 5 months
# Outcome: monthly net savings (€)
n <- 150
intercepts <- rnorm(n, 500, 80)   # wide between-customer variation in baseline savings
slopes     <- rnorm(n,  15, 12)   # some customers grow savings fast; others plateau

Y <- sapply(0:4, function(t) intercepts + slopes * t + rnorm(n, 0, 35))
colnames(Y) <- paste0("savings_m", 1:5)   # monthly savings at months 1–5
df_lgc <- as.data.frame(Y)
```

### Visualizing What the Latent Factors Mean

Before fitting the model, it helps to see what the intercept and slope factors represent in individual trajectories.

```{r lgc-visual, fig.width=7, fig.height=4.5}
# Draw 12 representative individual trajectories
df_vis <- tibble(
  id        = 1:12,
  intercept = c(intercepts[1:12]),
  slope     = c(slopes[1:12])
) |>
  expand_grid(time = 0:4) |>
  mutate(savings = intercept + slope * time + rnorm(n(), 0, 25),
         id = factor(id))

grand_intercept <- mean(intercepts)
grand_slope     <- mean(slopes)

ggplot(df_vis, aes(x = time, y = savings)) +
  geom_line(aes(group = id), alpha = 0.35, color = "#4393c3") +
  geom_point(aes(group = id), alpha = 0.35, color = "#4393c3", size = 1.5) +
  # Mean growth line
  geom_abline(intercept = grand_intercept, slope = grand_slope,
              color = "#d73027", linewidth = 1.5) +
  # Label: intercept at time = 0
  annotate("point", x = 0, y = grand_intercept, color = "#d73027", size = 4) +
  annotate("text",  x = 0.15, y = grand_intercept - 30,
           label = paste0("Mean intercept: \u20ac", round(grand_intercept), "/month at enrollment"),
           hjust = 0, size = 3.5, color = "#d73027") +
  # Label: slope (one-month increment on mean line)
  annotate("segment", x = 3, xend = 4,
           y = grand_intercept + 3 * grand_slope,
           yend = grand_intercept + 4 * grand_slope,
           color = "#d73027", linewidth = 1.2,
           arrow = arrow(length = unit(0.25, "cm"), ends = "last")) +
  annotate("text",  x = 4.05, y = grand_intercept + 3.5 * grand_slope,
           label = paste0("Mean slope:\n+\u20ac", round(grand_slope), "/month"),
           hjust = 0, size = 3.5, color = "#d73027") +
  scale_x_continuous(breaks = 0:4,
                     labels = paste0("Month ", 1:5)) +
  labs(
    title    = "Latent growth curve model: loyalty program savings trajectories",
    subtitle = "Blue = individual customer trajectories; Red = average growth line\nIntercept = baseline savings; Slope = monthly growth rate",
    x = NULL, y = "Monthly net savings (\u20ac)"
  ) +
  coord_cartesian(xlim = c(0, 4.8)) +
  theme_minimal(base_size = 13)
```

Each blue line is one customer. The intercept (red dot at month 1) is their individual starting level; the slope is how steeply their line rises. The model estimates both factors' means, variances, and covariance.

```{r fit-lgc}
# Latent growth curve model in lavaan
lgc_model <- '
  # Intercept factor (loadings all fixed to 1)
  intercept =~ 1*savings_m1 + 1*savings_m2 + 1*savings_m3 + 1*savings_m4 + 1*savings_m5

  # Slope factor (loadings = time coding: 0, 1, 2, 3, 4)
  slope     =~ 0*savings_m1 + 1*savings_m2 + 2*savings_m3 + 3*savings_m4 + 4*savings_m5

  # Means of latent factors (the fixed effects)
  intercept ~ 1
  slope     ~ 1

  # Variances of latent factors (random effects variance)
  intercept ~~ intercept
  slope     ~~ slope

  # Covariance between intercept and slope
  intercept ~~ slope

  # Residual variances (constrained equal across time for simplicity)
  savings_m1 ~~ a*savings_m1
  savings_m2 ~~ a*savings_m2
  savings_m3 ~~ a*savings_m3
  savings_m4 ~~ a*savings_m4
  savings_m5 ~~ a*savings_m5
'

m_lgc <- growth(lgc_model, data = df_lgc)
summary(m_lgc, fit.measures = TRUE)
```

```{r lgc-interpret}
# Extract key parameters
parameterEstimates(m_lgc) |>
  filter(op %in% c("~1", "~~") & lhs %in% c("intercept", "slope")) |>
  select(lhs, op, rhs, est, se, pvalue) |>
  knitr::kable(digits = 3,
    caption = "LGC parameter estimates: mean baseline savings, mean monthly growth, and their variances")
```

## The Equivalence with MLM

```{r mlm-comparison}
# Fit equivalent MLM
df_long_lgc <- df_lgc |>
  mutate(id = 1:n) |>
  pivot_longer(-id, names_to = "wave", values_to = "savings") |>
  mutate(time = as.integer(str_extract(wave, "[0-9]")) - 1,
         id   = factor(id))

m_mlm <- lmer(savings ~ time + (time | id), data = df_long_lgc)

# Compare estimates
tibble(
  Parameter        = c("Mean intercept (baseline savings)", "Mean slope (monthly growth)",
                       "Var(intercept)", "Var(slope)"),
  `MLM estimate`   = c(fixef(m_mlm),
                        as.data.frame(VarCorr(m_mlm))$vcov[1:2]),
  `LGC estimate`   = c(
    lavInspect(m_lgc, "mean.lv")[c("intercept", "slope")],
    diag(lavInspect(m_lgc, "cov.lv"))[c("intercept", "slope")]
  )
) |>
  knitr::kable(digits = 3,
    caption = "MLM and LGC give equivalent results: loyalty program savings growth model")
```

The estimates match. The difference is in the language and the software ecosystem — not the math.

## When the SEM Framing Adds Value

::: {.callout-note}
## Advantage 1: Modeling Measurement Error

In the SEM framework, each observed score $Y_{ti}$ has a residual $\varepsilon_{ti}$ that can be interpreted as **measurement error** — the gap between the true latent trajectory and the imperfect observed score. This is explicit in the LGC model.

In standard MLM, the residual is treated as a mix of measurement error and genuine within-person fluctuation. If your outcome is a measured construct (scale scores, ratings, test scores), the SEM framing forces you to be explicit about measurement.
:::

::: {.callout-note}
## Advantage 2: Multiple Growth Processes

With SEM, you can model two interrelated growth curves simultaneously. A natural JDM application: how **app engagement** and **monthly savings** co-evolve over the first 5 months of a loyalty program. The covariance between the two intercept factors tells you whether customers who were already engaging heavily at enrollment also started with higher savings. The covariance between the two slope factors tells you whether customers whose engagement grew faster also showed steeper savings growth — that is, whether engagement and savings amplify each other over time.

This is difficult to do cleanly in standard MLM, which would require a multivariate response formulation or a separate bivariate outcome model.
:::

## Assumptions of LGC Models

::: {.callout-warning icon=false}
## Assumption: Correctly Specified Factor Loadings

**What it means:** The factor loadings for the slope factor encode the time metric. Linear loadings (0, 1, 2, 3...) assume linear growth. If growth is nonlinear, you can free some loadings to be estimated from the data — but the intercept must still be defined by fixing at least two loadings.

**How to check:** Test alternative loading patterns (free vs. fixed) and compare model fit using CFI, RMSEA, and SRMR.
:::

::: {.callout-warning icon=false}
## Assumption: Multivariate Normality

**What it means:** SEM estimation (default ML) assumes the observed variables are multivariate normal. Severely skewed outcomes or floor/ceiling effects can bias estimates.

**How to check:** Check univariate skewness and kurtosis. Use robust ML estimators (MLR in lavaan) if normality is questionable.
:::
