---
title: "Covariance Structures"
---

```{r setup, include=FALSE}
library(tidyverse)
library(nlme)
library(lme4)
library(ggplot2)
library(patchwork)
set.seed(77)
```

# Covariance Structures

::: {.callout-note icon=false}
## The Structure Question

> **Which of your repeated observations should look most alike?** Adjacent ones — a WTP rating taken immediately after a disclosure is more similar to a rating taken a week later than to a rating taken six months later. This is the starting intuition for choosing a covariance structure.
>
> **What does the covariance structure encode?** An assumption about the data-generating process — how the underlying psychological or financial state of a person persists, decays, or drifts over time.
>
> **What goes wrong if you choose wrong?** Standard errors and test statistics are computed using the assumed structure. An incorrect structure produces biased inference — usually overstated precision (too-small SEs) if you assume too little covariance.
:::

## The Covariance Matrix Is a Scientific Statement

When you measure the same person at multiple time points, those measurements co-vary. How they co-vary is not a nuisance parameter — it is a substantive statement about the DGP. The choice of covariance structure encodes an assumption about how the process unfolds over time.

**A useful starting question:** "If I knew Person A's WTP in week 1, how much does that tell me about their WTP in week 2 versus week 8?" The answer depends on the persistence of the underlying state being measured:

- **Short-lived state** (e.g., momentary risk perception after seeing a shocking news headline): week 1 barely predicts week 4. Adjacent observations are only a bit more similar than distant ones.
- **Persistent individual tendency** (e.g., stable financial risk tolerance): week 1 predicts week 8 almost as well as week 2. Observations are broadly similar regardless of lag.
- **Gradually decaying memory or habit** (e.g., recall of a disclosure or momentum of a savings habit): week 2 is more similar to week 1 than week 8 is, and the similarity decays smoothly with lag.

Each of these stories implies a different covariance structure.

---

## The Main Structures

### 1. Compound Symmetry (CS)

All variances equal. All covariances equal regardless of lag.

$$\Sigma_{CS} = \begin{pmatrix} \sigma^2 & \sigma^2\rho & \sigma^2\rho \\ \sigma^2\rho & \sigma^2 & \sigma^2\rho \\ \sigma^2\rho & \sigma^2\rho & \sigma^2 \end{pmatrix}$$

**The DGP story:** Each person has a stable individual level, and every observation is equally predictable from every other. The WTP at month 1 tells you as much about month 6 as it tells you about month 2.

**JDM setting:** Most appropriate when what you are measuring is a **fixed individual trait** that shows no dynamic persistence or decay — for example, a stable risk-preference score that does not drift meaningfully across sessions. In practice this is rare; most repeated JDM measurements show some decay structure.

**When it fails:** Whenever observations are more similar to adjacent ones than to distant ones — which is almost always true in longitudinal JDM studies.

---

### 2. Autoregressive Order 1 — AR(1)

Correlations decay exponentially with time lag. Observations one step apart correlate $\rho$; two steps apart, $\rho^2$; $k$ steps apart, $\rho^k$.

$$\Sigma_{AR1} = \sigma^2\begin{pmatrix} 1 & \rho & \rho^2 \\ \rho & 1 & \rho \\ \rho^2 & \rho & 1 \end{pmatrix}$$

**The DGP story:** Each observation partially inherits from the previous one. Yesterday's state predicts today's, but the influence decays at a constant rate. A financial decision made under a disclosure effect carries partial memory into the next measurement, but fades over longer gaps.

**JDM settings:**
- Daily spending diary: today's spending predicts tomorrow's more than next week's
- Weekly savings rate: this week's savings are correlated with last week's, but the correlation decays with the lag
- Risk perception during a market fluctuation: day-to-day persistence that gradually loses predictive power

**When it fits:** When your measurement intervals are equal and the process has smooth geometric decay in correlations.

---

### 3. Unstructured (UN)

No constraints. Every variance and covariance is estimated freely from the data.

**The DGP story:** The data speak for themselves. There is no assumed decay pattern or symmetry.

**JDM settings:** Reasonable when you have a small number of waves (3–4) and a large sample — the model has enough data to estimate every pairwise correlation without exploding the parameter count.

**When it fails:** With many waves or small samples, the unstructured model is severely overparameterized. With 5 waves and 50 people, you are estimating 15 covariance parameters from data where each person contributes only 5 observations. The estimates are unstable and the model often fails to converge.

---

### 4. Toeplitz

Like AR(1) but without the geometric constraint: each lag has its own correlation, but the correlation depends only on the lag size, not the absolute time point.

**The DGP story:** Lag-1 observations are more similar than lag-2, which are more similar than lag-3 — but the rate of decay does not have to be exponential. A financial decision's influence might fade quickly at first (week 1 to week 2) and then more slowly (week 3 to week 6), which AR(1) cannot capture but Toeplitz can.

---

## Visualizing the Structures

```{r vis-structures, fig.width=10, fig.height=5}
make_cov_plot <- function(Sigma, title, subtitle = NULL) {
  n <- nrow(Sigma)
  D <- diag(1 / sqrt(diag(Sigma)))
  Corr <- D %*% Sigma %*% D
  wlabs <- paste0("W", 1:n)

  data.frame(Corr) |>
    setNames(wlabs) |>
    mutate(row = wlabs) |>
    pivot_longer(-row, names_to = "col", values_to = "r") |>
    mutate(row = factor(row, levels = rev(wlabs)),
           col = factor(col, levels = wlabs)) |>
    ggplot(aes(x = col, y = row, fill = r)) +
    geom_tile(color = "white", linewidth = 1) +
    geom_text(aes(label = round(r, 2)), size = 3.5) +
    scale_fill_gradient2(low = "#f7fbff", mid = "#4292c6", high = "#08306b",
                         midpoint = 0.5, limits = c(-0.1, 1)) +
    labs(title = title, subtitle = subtitle, x = NULL, y = NULL) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none")
}

n <- 5
rho <- 0.6

CS   <- matrix(rho, n, n); diag(CS) <- 1; CS <- CS * 4
AR1  <- outer(0:(n-1), 0:(n-1), function(i, j) rho^abs(i-j)) * 4
Toep <- outer(0:(n-1), 0:(n-1), function(i, j) {
  c(1, 0.7, 0.5, 0.3, 0.2)[abs(i-j)+1]
}) * 4
UN <- matrix(c(4.0, 2.4, 1.8, 1.4, 1.0,
               2.4, 4.2, 2.2, 1.6, 1.1,
               1.8, 2.2, 4.5, 2.4, 1.5,
               1.4, 1.6, 2.4, 4.8, 2.2,
               1.0, 1.1, 1.5, 2.2, 5.0), n, n)

(make_cov_plot(CS,   "Compound Symmetry",
               "Flat correlations — any wave\npredicts any other equally") +
 make_cov_plot(AR1,  "AR(1)",
               "Geometric decay — adjacent\nwaves most similar")) /
(make_cov_plot(Toep, "Toeplitz",
               "Free decay by lag — flexible\nbut still lag-dependent") +
 make_cov_plot(UN,   "Unstructured",
               "No constraints — data-driven\nbut needs large N"))
```

The heatmaps show how correlation between waves (W1–W5) changes under each structure. AR(1) and Toeplitz capture the realistic pattern where similarity decreases as the time gap grows.

---

## Specifying Covariance Structures in R

`lme4` does not support custom covariance structures beyond the random effects structure. For explicit covariance modeling, use `nlme`:

```{r covariance-nlme}
# Simulate WTP data from a savings study: 5 waves with AR(1) correlation
n_people <- 50
n_times  <- 5
rho_true <- 0.7

Y_ar1 <- MASS::mvrnorm(n_people, rep(50, n_times),
                        outer(0:(n_times-1), 0:(n_times-1),
                              function(i, j) rho_true^abs(i-j)) * 9)

df_long <- as.data.frame(Y_ar1) |>
  setNames(paste0("wave", 1:n_times)) |>
  mutate(person = 1:n_people) |>
  pivot_longer(-person, names_to = "wave", values_to = "WTP") |>
  mutate(wave_num = as.integer(str_extract(wave, "[0-9]")),
         person   = factor(person))

# Fit four competing models
m_cs  <- gls(WTP ~ factor(wave_num),
             correlation = corCompSymm(form = ~1 | person),
             data = df_long)
m_ar1 <- gls(WTP ~ factor(wave_num),
             correlation = corAR1(form = ~wave_num | person),
             data = df_long)
m_un  <- gls(WTP ~ factor(wave_num),
             correlation = corSymm(form = ~wave_num | person),
             weights = varIdent(form = ~1 | factor(wave_num)),
             data = df_long)

tibble(
  Model = c("Compound Symmetry", "AR(1)", "Unstructured"),
  AIC   = c(AIC(m_cs), AIC(m_ar1), AIC(m_un)),
  BIC   = c(BIC(m_cs), BIC(m_ar1), BIC(m_un))
) |>
  arrange(AIC) |>
  knitr::kable(digits = 1,
    caption = "Model comparison: AR(1) fits best when data has geometric decay in correlations (lower AIC/BIC = better)")
```

When the true structure is AR(1), the AR(1) model is preferred by AIC. Compound symmetry underfits (too constrained). Unstructured overfits (too many parameters for this sample size).

---

## Assumptions

::: {.callout-warning icon=false}
## AR(1) Assumption: Stationarity

**Plain-language meaning:** The mean and variance of the outcome are constant across time, and the correlation between any two observations depends only on how far apart they are — not on *when* in the study they fall. An AR(1) with $|\phi| < 1$ is stationary by construction.

**What the model is assuming away:** Trending means (e.g., WTP systematically rising over time due to growing familiarity with the product) or trending variance (e.g., participants becoming more variable later in a long task when fatigue sets in).

**JDM example:** If participants in a repeated savings task become more financially confident across waves — so WTP increases for all participants regardless of condition — the mean is trending. The AR(1) model will confound this trend with the outcome trajectory unless time is included as a fixed effect.

**What a skeptical reviewer might say:** "You modeled repeated savings willingness with AR(1) — but if participants are becoming more engaged with the study instrument over time, the mean is trending and stationarity fails."

**Observable warning sign:** Plot wave-by-wave means. If the mean increases or decreases systematically across waves, stationarity is violated for the raw outcome. Including wave as a fixed effect typically restores stationarity for the residuals.

**What to check:** Plot means and variances by wave. Fit a model with time as a fixed effect and check whether residuals are stationary. Use the Augmented Dickey-Fuller test for stationarity on per-person residual series.

**What claim survives if the assumption is doubtful:** The direction of the condition effect is usually robust. The autoregressive coefficient estimate and its confidence interval may be biased if stationarity is violated.
:::

::: {.callout-warning icon=false}
## Unstructured Assumption: Sufficient Sample Size

**Plain-language meaning:** The unstructured model estimates $T(T+1)/2$ covariance parameters for $T$ waves. With 5 waves, that is 15 parameters — just for the covariance structure. These must be estimated from the data.

**What the model is assuming away:** Any parsimony. The unstructured model places no restrictions on how measurements co-vary.

**JDM example:** A 6-wave savings diary study with 60 participants requires estimating 21 covariance parameters from a sample that has only 360 observations. That is fewer than 17 observations per covariance parameter — estimates will be noisy and models may fail to converge.

**What to check:** Compare the number of observations to the number of covariance parameters (aim for at least 10:1). If the unstructured model fails to converge, start with AR(1) and use AIC to evaluate whether the more constrained model fits acceptably.
:::

---

## Model Selection Strategy

```{r model-selection-guide}
tibble(
  Structure          = c("Compound Symmetry", "AR(1)", "Toeplitz", "Unstructured"),
  `Params (T=5)`    = c("2", "2", "5", "15"),
  `Best for`         = c(
    "Stable trait; no decay; rarely appropriate",
    "Regular intervals; geometric decay; default starting point",
    "Irregular decay; flexible but tractable; good middle ground",
    "Small T; large N; no prior structure; fit check for simpler models"
  )
) |>
  knitr::kable(caption = "Covariance structure selection guide (T = number of waves)")
```

**Recommended workflow:**

1. Start with AR(1) as the default for regularly-spaced longitudinal JDM data
2. Compare to compound symmetry (more constrained) using AIC/BIC — if CS fits as well, use it
3. Try Toeplitz if the decay looks non-geometric from the raw correlation matrix
4. Try unstructured only if sample size allows ($T \leq 4$, large $N$) — use as a robustness check
5. Plot residual autocorrelations after fitting to verify that serial dependence has been removed

---

## Reviewer Challenge

::: {.callout-tip}
## What a Reviewer Would Ask

**Challenge 1:** "You fitted a random intercept model with `lme4`. What covariance structure does that imply for your repeated WTP measurements?"
A random intercept model in `lme4` implies compound symmetry — all waves correlate equally at $\rho = \sigma^2_u / (\sigma^2_u + \sigma^2_\varepsilon)$. If adjacent waves should correlate more than distant ones (almost always true), this model misspecifies the covariance structure. Switch to `nlme` with `corAR1()` for longitudinal data with meaningful time ordering.

**Challenge 2:** "Why did you choose AR(1) over unstructured?"
The data-driven answer: AIC/BIC favored AR(1) and the unstructured model showed convergence issues (too many parameters for the sample). The conceptual answer: WTP decay after a disclosure intervention is well described by geometric fading — the psychological mechanism supports AR(1) structure.

**Challenge 3:** "Your AR(1) estimate is $\hat{\rho} = 0.72$ — doesn't that suggest the waves are measuring a stable trait rather than a dynamic process?"
High AR(1) correlation does not necessarily mean it is a fixed trait. It means the process has strong short-term persistence. A Bayesian risk-tolerance parameter might show $\rho = 0.72$ across weeks while still responding to information interventions — the AR structure describes the baseline persistence, not the treatment effect.
:::
