---
title: "Applied Identification Strategies for Panel Data"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(plm)
library(AER)
library(broom)
library(broom.mixed)
library(ggplot2)
library(patchwork)
library(sandwich)
library(lmtest)
set.seed(42)
```

# Applied Identification Strategies for Panel Data

## Section 1: Introduction — Module 3 Meets Panel Data

Module 3 covered six sources of endogeneity and six identification strategies. The discussion there was largely cross-sectional: a researcher observes units at one point in time and tries to isolate causal effects despite confounding. This chapter applies each of those strategies to **longitudinal panel data** from a JDM perspective: consumer finance panels, policy rollout datasets, and platform-level behavioral data. The time dimension allows us to difference out stable individual differences (fixed effects) and exploit policy announcement dates (DiD, ITS) or randomized encouragement (IV) to recover causal estimates in settings where simple regression fails.

The table below maps Module 3 threats to their panel-data solutions.

| Module 3 Threat | Panel Solution | Key R Function |
|---|---|---|
| Time-invariant OVB (Part 1) | Fixed effects (Module 1) | `lmer(..., REML=F)` or `plm(model="within")` |
| Time-varying endogeneity (Part 2 IV) | Panel IV | `ivreg(y ~ x + factor(unit) | z + factor(unit))` |
| Policy adoption at known date (Part 5 DiD) | TWFE + event study | `plm(model="within", effect="twoways")` |
| Temporal threshold (Part 4 RDD) | Interrupted Time Series | `lmer(y ~ t + post + t_post + (1|unit))` |
| Non-random treatment adoption (Part 6 PS) | Time-varying PS matching | `matchit()` at each adoption wave |
| Non-random attrition (Part 3 Selection) | Available-case lmer + MI | Module 4 Part 9 |

: Module 3 identification strategies mapped to their panel-data analogs {tbl-colwidths="[30,35,35]"}

The sections that follow walk through hands-on implementations of the most commonly needed strategies: two-way fixed effects DiD with an event study, interrupted time series (the temporal RDD), and panel IV.

---

## Section 2: Panel DiD with Two-Way Fixed Effects

### Setup and simulation

The **two-way fixed effects (TWFE)** estimator is the workhorse of panel difference-in-differences. It absorbs unit fixed effects (removing stable confounders) and time fixed effects (removing common shocks), then estimates the treatment effect from within-unit, within-period variation.

We simulate 60 households observed over 8 months. Thirty households are enrolled in a default savings program at month 5 by their employer. The true average treatment effect on the treated (ATT) is 8 percentage points of additional monthly savings.

```{r twfe-simulate}
n_hh       <- 60
n_months   <- 8

hh_eff      <- rnorm(n_hh, 0, 4)         # stable household saving habits
month_eff   <- rnorm(n_months, 0, 2)     # macroeconomic month shocks

df_twfe <- expand_grid(household = 1:n_hh, month = 1:n_months) |>
  mutate(
    treated     = as.integer(household <= 30),
    post        = as.integer(month >= 5),
    D           = treated * post,
    h_eff       = hh_eff[household],
    m_eff       = month_eff[month],
    savings_pct = 12 + 8 * D + h_eff + m_eff + rnorm(n(), 0, 3),
    household   = factor(household),
    month       = factor(month)
  )
```

### Three equivalent estimators

```{r twfe-models}
# 1. Manual 2x2 DiD
did_manual <- with(
  df_twfe |>
    group_by(treated, post) |>
    summarise(y = mean(savings_pct), .groups = "drop"),
  (y[treated == 1 & post == 1] - y[treated == 1 & post == 0]) -
    (y[treated == 0 & post == 1] - y[treated == 0 & post == 0])
)
cat("Manual DiD estimate:", round(did_manual, 2), "\n")

# 2. Regression DiD with household-clustered SEs
m_reg_did  <- lm(savings_pct ~ treated * post, data = df_twfe)
se_cluster <- coeftest(m_reg_did, vcov = vcovCL(m_reg_did, cluster = ~household))

# 3. TWFE via plm (two-way within)
pdata_twfe <- pdata.frame(
  df_twfe |> mutate(hh_id = as.integer(household), mo = as.integer(month)),
  index = c("hh_id", "mo")
)
m_twfe <- plm(savings_pct ~ D, data = pdata_twfe,
              model = "within", effect = "twoways")
```

```{r twfe-compare, echo=FALSE}
naive_se <- sqrt(diag(vcov(m_reg_did)))["treated:post"]
clust_se <- se_cluster["treated:post", "Std. Error"]

tibble(
  Method           = c("Manual DiD", "Regression DiD (naive SE)",
                       "Regression DiD (clustered SE)", "TWFE via plm"),
  Estimate         = c(did_manual,
                       coef(m_reg_did)["treated:post"],
                       se_cluster["treated:post", "Estimate"],
                       coef(m_twfe)["D"]),
  `Std. Error`     = c(NA, naive_se, clust_se, NA)
) |>
  knitr::kable(digits = 3,
               caption = "All three approaches target the true ATT \u2248 8. Clustered SEs account for within-household correlation of savings shocks over time; they may be larger or smaller than naive SEs depending on the dependence structure.")
```

::: {.callout-warning icon=false}
## Why Cluster Standard Errors in Panel DiD?

Observations within the same household are not independent — savings behavior shocks persist over time. Treating them as independent (naive OLS SEs) understates uncertainty and inflates Type I error rates. The rule of thumb: cluster at the level of treatment assignment. Here, treatment varies at the household level, so we cluster on `household`. With fewer than ~30 clusters, consider wild cluster bootstrap (`fwildclusterboot` package).
:::

### Event study: testing parallel trends

The parallel trends assumption — that treated and control households would have followed the same savings trajectory absent the enrollment policy — is not directly testable, but we can look for violations by checking whether pre-treatment differences in trends are already present.

```{r event-study-plot, fig.width=8, fig.height=4.5}
m_event <- lm(savings_pct ~ treated * factor(month), data = df_twfe)

event_df <- broom::tidy(m_event, conf.int = TRUE) |>
  filter(str_detect(term, "treated:")) |>
  mutate(
    month = as.integer(str_extract(term, "[0-9]+")),
    pre   = month < 5
  )

ggplot(event_df, aes(x = month, y = estimate, color = pre)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 4.5, linetype = "dotted", color = "gray30") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("FALSE" = "#1e3a5f", "TRUE" = "#aec6e8"),
    labels = c("FALSE" = "Post-enrollment", "TRUE" = "Pre-enrollment (should \u2248 0)")
  ) +
  annotate("text", x = 4.7, y = max(event_df$conf.high) * 0.9,
           label = "Default\nenrollment", hjust = 0, size = 3.5, color = "gray30") +
  labs(
    title   = "Event study: pre-period near zero supports parallel trends in savings behavior",
    subtitle = "Post-period coefficients trace the materializing treatment effect",
    x       = "Month",
    y       = "Estimated treatment effect vs. Month 1",
    color   = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

Pre-period interaction coefficients (light blue) are scattered around zero — consistent with parallel pre-trends. Post-period coefficients (dark blue) step up after month 5 and hover near the true effect of 8. If the pre-period coefficients showed a systematic upward or downward slope, that would be evidence of pre-existing differential trends and would undermine the DiD identification assumption.

::: {.callout-tip}
## Reading an Event Study Plot

- **Pre-period coefficients near zero**: parallel trends plausible. Fail this check and DiD is suspect.
- **Sharp jump at treatment date**: consistent with an immediate treatment effect.
- **Gradual post-period increase**: treatment effect builds over time (dynamic effects).
- **Anticipation effects** (pre-period coefficients trending upward before treatment): households may be responding in advance; adjust the "post" indicator.
:::

---

## Section 3: Interrupted Time Series (Temporal RDD)

### Concept

**Interrupted Time Series (ITS)** applies the RDD logic to time rather than a continuous running variable. When a policy takes effect at a known calendar date $c$, we look for a discontinuity in the time series at that date — a sudden level jump and/or a change in slope. The key identification assumption mirrors RDD's continuity assumption: absent the policy, the time trend would have continued smoothly through $c$.

The ITS model decomposes the response into:

$$Y_{it} = \beta_0 + \beta_1 t + \underbrace{\beta_2 D_t}_{\text{level jump}} + \underbrace{\beta_3 (t - c)\, D_t}_{\text{slope change}} + u_{0i} + \varepsilon_{it}$$

where $D_t = \mathbf{1}[t \geq c]$, $t - c$ is time elapsed since the intervention (zero before), $u_{0i}$ is a unit random intercept (absorbing stable unit differences), and $\varepsilon_{it}$ is idiosyncratic error.

- $\beta_1$: pre-regulation monthly trend in customer retention
- $\beta_2$: immediate level change at month $c$ (jump in retention when regulation takes effect)
- $\beta_3$: additional slope change after regulation (change in rate of improvement)

### Simulation

We simulate 40 financial service branches observed for 30 months. A fee-transparency disclosure regulation takes effect at month 16. The true DGP has a $+4$ percentage point level jump in customer retention rate and a $+0.2$/month additional improvement in retention rate slope.

```{r its-simulate}
n_branches <- 40
n_months   <- 30
branch_eff <- rnorm(n_branches, 0, 4)

df_its <- expand_grid(branch = 1:n_branches, month = 1:n_months) |>
  mutate(
    post      = as.integer(month >= 16),
    t_post    = pmax(month - 15, 0),       # months elapsed since regulation; 0 before
    b_eff     = branch_eff[branch],
    retention = 72 + 0.3 * month + 4 * post + 0.2 * t_post + b_eff + rnorm(n(), 0, 4),
    branch    = factor(branch)
  )
```

```{r its-model}
m_its <- lmer(retention ~ month + post + t_post + (1 | branch), data = df_its)

broom.mixed::tidy(m_its, effects = "fixed") |>
  knitr::kable(digits = 3,
               caption = "ITS model: pre-trend (month), level jump (post), slope change (t_post). True values: 0.3, 4, 0.2.")
```

### Visualization

```{r its-plot, fig.width=8.5, fig.height=4.5}
branch_avg <- df_its |>
  group_by(month) |>
  summarise(mean_retention = mean(retention), .groups = "drop")

# Predicted values from lmer for an average branch
pred_df <- tibble(
  month  = 1:n_months,
  post   = as.integer(month >= 16),
  t_post = pmax(month - 15, 0),
  branch = factor(1)          # prediction for branch 1 (fixed effects only for plot)
)
pred_df$pred <- predict(m_its, newdata = pred_df, re.form = NA)

ggplot(branch_avg, aes(x = month, y = mean_retention)) +
  geom_point(color = "gray60", size = 2) +
  geom_line(data = pred_df, aes(y = pred), color = "#1e3a5f", linewidth = 1.2) +
  geom_vline(xintercept = 15.5, linetype = "dashed", color = "#c0392b") +
  annotate("segment",
           x = 15.5, xend = 15.5, y = 77, yend = 81,
           arrow = arrow(length = unit(0.25, "cm")), color = "#c0392b") +
  annotate("text", x = 16, y = 82,
           label = expression(beta[2]*": level jump"), hjust = 0,
           color = "#c0392b", size = 3.5) +
  annotate("text", x = 22, y = 90,
           label = expression(beta[3]*": slope change"),
           color = "#2980b9", size = 3.5) +
  labs(
    title    = "Interrupted Time Series: fee-transparency disclosure regulation at month 16",
    subtitle = "Level jump (\u03b22) and slope change (\u03b23) in customer retention at the regulatory cutoff",
    x        = "Month",
    y        = "Branch-average retention rate (%)"
  ) +
  theme_minimal(base_size = 13)
```

### Pre-trend test

The ITS equivalent of the RDD density (McCrary) test is to check whether the pre-intervention trend is genuinely linear, or whether a non-linear pre-trend exists that could masquerade as a level jump. If the pre-trend in retention was already curving before month 16, the jump at the cutoff could reflect a pre-existing trend rather than the regulation.

```{r its-pretrend}
df_pre <- df_its |> filter(month < 16)
m_pretrend <- lmer(retention ~ month + I(month^2) + (1 | branch), data = df_pre)
broom.mixed::tidy(m_pretrend, effects = "fixed") |>
  knitr::kable(digits = 4,
               caption = "Pre-period model with quadratic term. If I(month^2) is far from zero, the pre-trend is curved and the level-jump interpretation is suspect.")
```

A quadratic coefficient near zero in the pre-period supports the linearity assumption. A large quadratic coefficient means the trajectory was already bending at the cutoff — the ITS level jump would be confounded.

---

## Section 4: Panel IV

### When fixed effects are not enough

Unit fixed effects remove **time-invariant** confounders. But if a confounder varies both across units and across time — and is correlated with the predictor — FE alone is insufficient. Panel IV combines the within-unit differencing of FE with an instrument that shifts the endogenous predictor but has no direct path to the outcome.

### Simulation: budgeting tool use and savings

We simulate 200 platform customers over 6 months. Customers voluntarily use a budgeting tool (endogenous — customers with higher saving disposition use it more). Platform encouragement push (amount the platform pushes tool reminders) serves as the instrument: it is relevant (push increases tool use), excludes directly (platform push affects savings only through tool use), and is as-good-as-randomly assigned at the feature level.

```{r panel-iv-simulate}
n_c  <- 200
n_m  <- 6                               # months
base_saving    <- rnorm(n_c, 0, 5)     # unobserved saving disposition (quality analog)
platform_push  <- rnorm(10, 0, 2)[rep(1:10, each = n_c/10)]  # feature-level push intensity

df_piv <- expand_grid(customer = 1:n_c, month = 1:n_m) |>
  mutate(
    bs      = base_saving[customer],
    # Time-varying disposition shock (FE removes base_saving mean, not this shock)
    bs_tv   = bs + rnorm(n(), 0, 3),
    push    = platform_push[customer] + rnorm(n(), 0, 0.5),  # customer-month realization
    # Budgeting tool usage: endogenous — higher disposition + higher push -> more usage
    tool_use = 20 + 2 * bs_tv + 3 * push + rnorm(n(), 0, 5),
    # Savings rate: causal effect of tool_use = 0.4; also depends on disposition
    savings_rate = 5 + 0.4 * tool_use + 1.5 * bs_tv + rnorm(n(), 0, 3),
    customer = factor(customer),
    month    = factor(month)
  )
```

OLS + FE removes stable customer effects, but if saving disposition varies over time (a good-month disposition leads to more tool use AND more saving), fixed effects alone cannot isolate the causal effect of the tool.

```{r panel-iv-models}
# OLS + FE: still biased by time-varying disposition-tool correlation
m_ols_fe <- lm(savings_rate ~ tool_use + customer + month, data = df_piv)

# Panel IV: instrument = platform push
m_iv <- ivreg(savings_rate ~ tool_use + customer + month | push + customer + month,
              data = df_piv)

# First-stage F-statistic
fs <- lm(tool_use ~ push + customer + month, data = df_piv)
fs_F <- summary(fs)$fstatistic[1]
cat("First-stage F on instrument:", round(fs_F, 1), "\n")
```

```{r panel-iv-compare, echo=FALSE}
tibble(
  Model    = c("OLS + unit/month FE", "IV + unit/month FE"),
  Estimate = c(coef(m_ols_fe)["tool_use"], coef(m_iv)["tool_use"]),
  `Std. Error` = c(
    sqrt(diag(vcov(m_ols_fe)))["tool_use"],
    sqrt(diag(vcov(m_iv)))["tool_use"]
  ),
  `True value` = 0.4
) |>
  knitr::kable(digits = 4,
               caption = "Panel IV: platform push as instrument for budgeting tool usage. True causal effect = 0.4 per unit of tool use. OLS + FE is biased by time-varying saving disposition (bs_tv), which customer fixed effects cannot remove. First-stage: platform push strongly predicts tool usage (F > 10). Exclusion: push affects savings only through the tool (by design of the instrument).")
```

::: {.callout-warning icon=false}
## Weak Instruments in Panel Data

Panel fixed effects remove between-unit variation and retain only **within-unit** variation. This can substantially weaken an instrument that looked strong in cross-sectional data: the platform push intensity varies mostly across feature groups, and much of that cross-group variation is soaked up by the customer fixed effects. Always check the first-stage $F$-statistic using the **demeaned** (within-transformed) data, not the raw data. The conventional threshold is $F > 10$; for inference-robust methods (Anderson-Rubin test), the threshold matters less.

```r
# Safe check: first stage on demeaned data via plm
pdata_piv   <- pdata.frame(df_piv |> mutate(cust_id=as.integer(customer), mo=as.integer(month)),
                             index=c("cust_id","mo"))
fs_within   <- plm(tool_use ~ push, data=pdata_piv, model="within", effect="twoways")
summary(fs_within)
```
:::

---

## Section 5: Time-Varying Propensity Score Matching

When units adopt a treatment at **different waves** — a staggered rollout rather than a single cutoff — the two-by-two DiD logic breaks down because the "post" period differs across treated units. Two-way FE with staggered adoption can produce negatively-weighted treatment effect estimates when treatment effects are heterogeneous across cohorts (the "negative weighting" problem documented by Callaway & Sant'Anna, 2021).

One practical alternative is **time-varying propensity score matching**, which handles staggered adoption explicitly:

**Step 1**: For each adoption wave $w$, identify:
- Households (or customers) that just enrolled in wave $w$ (the "treated" at wave $w$).
- Households that have not yet enrolled by wave $w$ (the "not-yet-treated" controls).

**Step 2**: Compute the propensity score at wave $w$ among eligible units:

```r
glm(enrolled_now ~ lag_outcome + income + age + region,
    family = binomial,
    data   = eligible_at_wave_w)
```

**Step 3**: Match treated households to the most similar not-yet-treated households **at that wave** using nearest-neighbor matching within the wave-specific pool.

**Step 4**: Estimate the ATT on the matched sample, using the matched controls' post-enrollment outcomes as the counterfactual.

This approach produces a cohort-specific ATT for each adoption wave, which can be aggregated into an overall ATT. It is conceptually equivalent to the Callaway & Sant'Anna (2021) estimator (available in the `did` package) and is doubly robust when combined with outcome regression.

::: {.callout-tip}
## PS-Weighted DiD: Doubly Robust Panel Estimation

Combining time-varying PS weighting with DiD regression produces an estimator that is **doubly robust**: consistent if either (a) the propensity score model is correctly specified, or (b) parallel trends holds conditional on covariates. Neither assumption alone is sufficient for ordinary DiD; neither alone is sufficient for ordinary PSM. The combination of both provides extra insurance. See `drdid` package for implementation.
:::

---

## Section 6: Decision Tree — Which Strategy for Which Panel Problem?

::: {.callout-tip}
## Which Identification Strategy for Which Panel Problem?

| Problem | Method | Key assumption | Module reference |
|---------|---------|----------------|-----------------|
| Stable unit-level confounders | Unit fixed effects | Time-invariant endogeneity | M1 Parts 1–2 |
| Time-varying endogeneity | Panel IV | Instrument relevance + exclusion restriction | M3 Part 2 |
| Parallel policy rollout | TWFE DiD | Parallel trends | M3 Part 5 |
| Staggered adoption | Callaway-Sant'Anna / `drdid` | Parallel trends by cohort | M3 Part 5 |
| Temporal policy cutoff | ITS | Continuity of time trend at cutoff | M3 Part 4 |
| Non-random treatment adoption | Time-varying PS matching | Selection on observables | M3 Part 6 |
| Non-random attrition | Available-case lmer + MI | MAR | M4 Part 9 |

The first question is always: **what creates the identifying variation?** Time, a threshold, an instrument, or a policy rollout? The answer determines the method.
:::
