---
title: "From RM-ANOVA to Mixed Models"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(nlme)
library(ggplot2)
library(patchwork)
set.seed(2024)
```

# From RM-ANOVA to Mixed Models

::: {.callout-note icon=false}
## The Covariance Question

> **What does RM-ANOVA quietly assume about your repeated measurements?** That every pair of time points correlates equally — a measurement taken immediately after a disclosure intervention relates to a 6-month follow-up measurement exactly as much as it relates to a 1-week follow-up. This is the sphericity assumption, and it is almost always wrong.
>
> **What goes wrong when sphericity fails?** The F-test for the time effect — and the condition × time interaction — is anti-conservative. You reject the null more often than your stated alpha level actually permits.
>
> **What does the mixed model do differently?** It models the actual covariance structure of your repeated observations, rather than assuming a specific structure that data rarely satisfy.
:::

## The Standard Approach and Its Hidden Assumption

Repeated measures ANOVA (RM-ANOVA) has been the default tool for within-subjects designs for decades. It is built into every statistics textbook and available in SPSS, JASP, and R. And it makes an assumption about your data that is almost always wrong.

That assumption is **sphericity**.

## What Is Sphericity?

Sphericity constrains the variance-covariance matrix of your repeated measurements. Specifically, it requires that the variances of all pairwise *differences* between time points are equal.

Consider measuring willingness to pay (WTP) at three time points: **before** a financial disclosure intervention, **immediately after**, and at a **one-week follow-up**. Sphericity requires that:

$$\text{Var}(\text{WTP}_{\text{pre}} - \text{WTP}_{\text{post}}) = \text{Var}(\text{WTP}_{\text{pre}} - \text{WTP}_{\text{follow-up}}) = \text{Var}(\text{WTP}_{\text{post}} - \text{WTP}_{\text{follow-up}})$$

In plain language: the variability in how much people *change* must be the same regardless of which two time points you compare. This is equivalent to assuming **compound symmetry** — that all time points correlate equally with each other.

But this is implausible. The pre-to-post change (a few days) captures the immediate disclosure reaction. The pre-to-follow-up change (a week) is noisier because some people's attention has faded. The post-to-follow-up change (a few days) reflects short-term decay. These three differences have different variances — sphericity is already violated in principle before you collect any data.

```{r sphericity-diagram, fig.width=10, fig.height=3.5}
library(Matrix)

# Three covariance structures for 4 repeated WTP measurements
CS <- matrix(c(1.0, 0.5, 0.5, 0.5,
               0.5, 1.0, 0.5, 0.5,
               0.5, 0.5, 1.0, 0.5,
               0.5, 0.5, 0.5, 1.0), 4, 4)

rho <- 0.7
AR1 <- outer(1:4, 1:4, FUN = function(i, j) rho^abs(i - j))

US <- matrix(c(1.0, 0.8, 0.6, 0.3,
               0.8, 1.0, 0.7, 0.5,
               0.6, 0.7, 1.0, 0.8,
               0.3, 0.5, 0.8, 1.0), 4, 4)

plot_cov <- function(M, title, subtitle = NULL) {
  lbl <- c("Pre", "Post", "1wk", "1mo")
  as.data.frame(M) |>
    setNames(lbl) |>
    mutate(row = lbl) |>
    pivot_longer(-row, names_to = "col", values_to = "corr") |>
    mutate(row = factor(row, levels = rev(lbl)),
           col = factor(col, levels = lbl)) |>
    ggplot(aes(x = col, y = row, fill = corr)) +
    geom_tile(color = "white") +
    geom_text(aes(label = round(corr, 1)), size = 4) +
    scale_fill_gradient2(low = "#f7fbff", mid = "#6baed6", high = "#08306b",
                         midpoint = 0.5, limits = c(0, 1)) +
    labs(title = title, subtitle = subtitle,
         x = "WTP wave", y = "WTP wave", fill = "Corr") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none")
}

plot_cov(CS,  "Compound Symmetry",    "All pairs equally correlated.\nSphericity holds. Implausible.") +
plot_cov(AR1, "AR(1)",                "Adjacent waves more similar.\nDecaying with time lag. Common.") +
plot_cov(US,  "Unstructured",         "No constraints. Each pair\nestimated from data.") +
plot_layout(ncol = 3)
```

The compound symmetry structure (left) forces all off-diagonal entries to be equal — WTP pre-post relates to WTP post-follow-up the same as WTP pre-follow-up. This is the sphericity assumption. AR(1) (middle) lets correlations decay with the time gap. Unstructured (right) imposes no constraints.

Real longitudinal data almost always looks more like AR(1) or unstructured than compound symmetry.

---

## What Happens When Sphericity Is Violated?

When sphericity fails, RM-ANOVA F-tests are **anti-conservative** — Type I error rates inflate above your stated alpha level.

```{r sphericity-violation}
simulate_repeated <- function(n = 50, n_times = 4, spherical = TRUE) {
  Sigma <- if (spherical) CS * 4 else AR1 * 4
  MASS::mvrnorm(n, mu = rep(0, n_times), Sigma = Sigma)
}

n_sims <- 1000
type1_spherical <- numeric(n_sims)
type1_violated  <- numeric(n_sims)

for (i in 1:n_sims) {
  y_sph <- simulate_repeated(spherical = TRUE)
  y_vio <- simulate_repeated(spherical = FALSE)

  extract_pval <- function(y_mat) {
    df_aov <- data.frame(
      Y      = c(y_mat),
      time   = factor(rep(1:4, each = 50)),
      person = factor(rep(1:50, 4))
    )
    aov_fit <- aov(Y ~ time + Error(person/time), data = df_aov)
    s <- summary(aov_fit)
    tryCatch(s[["Error: person:time"]][[1]][["Pr(>F)"]][1],
             error = function(e) NA_real_)
  }

  type1_spherical[i] <- extract_pval(y_sph) < 0.05
  type1_violated[i]  <- extract_pval(y_vio) < 0.05
}

tibble(
  Condition = c("Sphericity holds (CS)", "Sphericity violated (AR1)"),
  `Type I Error Rate` = c(mean(type1_spherical, na.rm = TRUE),
                           mean(type1_violated,  na.rm = TRUE))
) |>
  knitr::kable(digits = 3,
    caption = "Type I error under sphericity vs. violation (nominal alpha = 0.05). With AR(1) data, RM-ANOVA rejects the null too often.")
```

With AR(1) data — where adjacent waves correlate more than distant waves — RM-ANOVA produces Type I error rates above 5%, even though the null is true.

---

## Assumption: Sphericity

::: {.callout-warning icon=false}
## Assumption: Sphericity (RM-ANOVA Only)

**Plain-language meaning:** All pairwise differences between time points have the same variance. Equivalent to compound symmetry: every pair of WTP waves correlates equally, regardless of the time gap separating them.

**What the model is assuming away:** The natural structure of most longitudinal data, where adjacent observations are more similar than distant ones. A WTP measurement taken immediately after a disclosure should correlate more strongly with a 1-week follow-up than with a 6-month follow-up.

**JDM example:** In a disclosure experiment with pre, immediate post, 1-week, and 1-month follow-ups, sphericity requires that the correlation structure is flat across all time gaps. But 1-week WTP should resemble immediate-post WTP more than it resembles baseline WTP.

**Experimental analog:** In an observational panel study of weekly savings behavior, sphericity would require that week 1 and week 4 savings correlate as strongly as week 1 and week 2 savings. This is never realistic.

**What a skeptical reviewer might say:** "You used RM-ANOVA with four waves and report a significant time × condition interaction — did you test sphericity? With decaying correlations across waves, RM-ANOVA F-tests will be anti-conservative."

**Observable warning sign:** Mauchly's W is significant (p < .05). Or simply: you have more than 2 repeated measurements and time points are separated by unequal gaps.

**What to check:** Run Mauchly's test as a preliminary check. But note: it is underpowered in small samples and overpowered in large ones. The better move is to switch to a mixed model — then the sphericity assumption is not required at all.

**What claim survives if the assumption is doubtful:** The direction of effects is usually robust. The p-value is too small under violation, so a conclusion that "just barely" passes a threshold is suspect.
:::

---

## Mixed-Effects Models: No Sphericity Required

A mixed-effects model makes no sphericity assumption. The random effects structure and residual covariance together determine what the model implies about how measurements co-vary — and you can specify the structure that fits the data rather than forcing compound symmetry.

```{r rm-anova-vs-mlm}
# Simulate a WTP study: 3 waves, AR(1) correlation (sphericity violated)
n_people <- 60
n_times  <- 3
Sigma_ar <- AR1[1:3, 1:3] * 4  # true structure is AR(1), not CS

# Null data: no true time effect
wtp_data <- MASS::mvrnorm(n_people, mu = c(50, 50, 50), Sigma = Sigma_ar) |>
  as.data.frame() |>
  setNames(c("pre", "post", "follow_up")) |>
  mutate(person = 1:n_people) |>
  pivot_longer(-person, names_to = "time", values_to = "WTP") |>
  mutate(
    time_num = case_when(time == "pre" ~ 1, time == "post" ~ 2, TRUE ~ 3),
    person   = factor(person)
  )

# RM-ANOVA
m_anova <- aov(WTP ~ factor(time_num) + Error(person / factor(time_num)),
               data = wtp_data)

# Mixed-effects model with AR(1) covariance
m_mlm <- lme(WTP ~ factor(time_num),
             random      = ~1 | person,
             correlation = corAR1(),
             data        = wtp_data)

cat("RM-ANOVA p (time):", round(
  summary(m_anova)$`Error: person:factor(time_num)`[[1]]["factor(time_num)", "Pr(>F)"], 4), "\n")
cat("MLM p (time):     ", round(anova(m_mlm)["factor(time_num)", "p-value"], 4), "\n")
```

When the true covariance structure is AR(1), the mixed-effects model gives more accurate p-values — because it uses the correct covariance structure rather than assuming an implausible flat one.

---

## The Key Advantages of Mixed Models Over RM-ANOVA

| Feature | RM-ANOVA | Mixed-Effects Model |
|---------|----------|---------------------|
| Assumes sphericity? | Yes — almost always violated | No — covariance is modeled |
| Handles missing data? | No — requires complete cases at every wave | Yes — uses all available observations |
| Allows continuous time? | No — time must be discrete categories | Yes |
| Models individual trajectories? | No | Yes (random slopes) |
| Handles unbalanced designs? | Poorly | Gracefully |
| Can include level-2 predictors? | No | Yes |
| Generalizes to non-normal outcomes? | No | Yes (GLMM) |

The only reason to prefer RM-ANOVA is familiarity. For most real longitudinal designs, mixed-effects models are more flexible, more transparent about assumptions, and better able to handle unequal spacing and missing data.

---

## Reviewer Challenge

::: {.callout-tip}
## What a Reviewer Would Ask

**Challenge 1:** "You measured WTP at 4 waves and analyzed with RM-ANOVA — did you test for sphericity? If it's violated, your interaction p-value is too small."
The correct response: switch to a mixed model (or at minimum apply Greenhouse-Geisser correction). A mixed model with AR(1) residuals is both more flexible and more honest about the covariance structure than corrected RM-ANOVA.

**Challenge 2:** "You have 12% missing data at wave 3 and excluded those cases. How does that affect the RM-ANOVA result?"
RM-ANOVA requires complete cases — excluding wave-3 dropouts creates a biased subsample. The mixed model with available-case estimation (all participants at their observed waves) is unbiased under MAR and should be the primary analysis.

**Challenge 3:** "Why do you need random effects for person in a within-subjects design where each person appears multiple times?"
The random intercept for person absorbs between-person variance in baseline WTP, reducing residual noise and improving power. Without it, that between-person variance inflates the residual, making the condition effect look less precise. The random intercept is not optional in a within-subjects mixed model — it is the direct translation of the design into the model.
:::
