---
title: "Autoregressive Structures"
---

```{r setup, include=FALSE}
library(tidyverse)
library(nlme)
library(ggplot2)
library(patchwork)
set.seed(42)
```

# Autoregressive Structures

::: {.callout-note icon=false}
## The Persistence Question

> **Does today's observation carry information about tomorrow's, beyond what the model's fixed effects already explain?** If yes — if yesterday's daily spending predicts today's after controlling for income and season — the residuals are serially correlated, and treating them as independent is a misspecification.
>
> **What goes wrong if you ignore it?** Standard errors are biased. More importantly, you are asserting a DGP where financial behavior has no memory — each day is a fresh draw. That assertion is almost never true.
>
> **What does AR(1) buy you?** A correct representation of the persistence structure. And a meaningful interpretation: the autoregressive coefficient tells you how much of today's deviation from the trend was already visible in yesterday's deviation.
:::

## The Memory of Time

Some outcomes carry their history. Today's daily spending predicts tomorrow's. This week's savings rate predicts next week's. A consumer's current risk perception reflects last week's perception, adjusted for new information. This temporal dependence — the "memory" of the process — is not just a nuisance. It is a feature of the DGP that your model must account for.

When you ignore serial dependence and treat residuals as independent:

1. Your standard errors are biased — usually too small if residuals are positively correlated
2. You are misrepresenting the process — claiming the outcome at each time point arises fresh, unconnected from what came before
3. You cannot distinguish **persistence** (genuine temporal carry-over) from **trend** (a steady directional shift)

## What AR(1) Means

In an **autoregressive order-1** (AR(1)) process, each observation is a linear function of the immediately preceding observation plus random noise:

$$Y_t = \phi \cdot Y_{t-1} + \varepsilon_t, \quad \varepsilon_t \sim \mathcal{N}(0, \sigma^2_\varepsilon)$$

The parameter $\phi$ is the **autoregressive coefficient** — how strongly the past predicts the present. When $\phi = 0$, observations are independent. When $\phi = 0.8$, yesterday's financial state strongly predicts today's.

**JDM examples where AR(1) is plausible:**

- **Daily risk perception**: a spike in risk perception triggered by a market event slowly decays back toward a person's baseline over subsequent days
- **Weekly savings rate**: savings behavior in one week partially inertially carries forward to the next
- **Daily spending**: expenditure patterns show week-to-week habit inertia
- **Price sensitivity across promotional cycles**: a consumer's sensitivity in one promotional period partially predicts their sensitivity in the next

```{r ar1-example, fig.width=9, fig.height=4}
# Simulate AR(1) process for one person: 30 days of daily risk perception
phi   <- 0.75
sigma <- 3
n_t   <- 30

Y <- numeric(n_t)
Y[1] <- rnorm(1, 50, 5)
for (t in 2:n_t) Y[t] <- phi * Y[t-1] + rnorm(1, 0, sigma)

p1 <- tibble(day = 1:n_t, risk_perception = Y) |>
  ggplot(aes(x = day, y = risk_perception)) +
  geom_line(color = "#1e3a5f", linewidth = 1) +
  geom_point(size = 2, color = "#1e3a5f", alpha = 0.7) +
  labs(
    title    = paste0("Daily risk perception: AR(1) with φ = ", phi),
    subtitle = "Each day's value depends on the previous — shows smooth persistence",
    x = "Day", y = "Risk perception"
  ) +
  theme_minimal(base_size = 13)

acf_vals <- acf(Y, plot = FALSE, lag.max = 10)
p2 <- tibble(lag = 0:10, acf = acf_vals$acf[,1,1]) |>
  filter(lag > 0) |>
  ggplot(aes(x = lag, y = acf)) +
  geom_col(fill = "#1e3a5f", alpha = 0.8) +
  geom_hline(yintercept = 0, color = "black") +
  geom_hline(yintercept = c(-1, 1) * 1.96 / sqrt(n_t),
             linetype = "dashed", color = "#e63946") +
  labs(
    title    = "Autocorrelation function (ACF)",
    subtitle = "Geometric decay is the signature of AR(1)",
    x = "Lag (days)", y = "ACF"
  ) +
  theme_minimal(base_size = 13)

p1 + p2
```

The geometric decay in the ACF — strong at lag 1, weaker at lag 2, gradually dying out — is the signature pattern of an AR(1) process.

---

## Trend Versus Persistence: A Critical Distinction

AR(1) is about **persistence** — today's deviation from the trend carries into tomorrow. This is fundamentally different from a **trend** — a systematic directional movement over time. Confusing the two leads to a common misspecification:

```{r trend-vs-persistence, fig.width=9, fig.height=4}
t_vals <- 1:30
# Trend: mean drifts upward, residuals are independent
y_trend <- 40 + 0.8 * t_vals + rnorm(30, 0, 3)

# Persistence: stationary mean, AR(1) residuals
phi_p <- 0.8
y_pers <- numeric(30)
y_pers[1] <- 50
for (t in 2:30) y_pers[t] <- 50 * (1 - phi_p) + phi_p * y_pers[t-1] + rnorm(1, 0, 3)

p_trend <- tibble(day = t_vals, y = y_trend) |>
  ggplot(aes(x = day, y = y)) +
  geom_line(color = "#e67e22", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "#c0392b", linetype = "dashed") +
  labs(
    title    = "Trend (not AR)",
    subtitle = "Mean drifts. Residuals are independent.",
    x = "Day", y = "Outcome"
  ) +
  theme_minimal(base_size = 12)

p_pers <- tibble(day = t_vals, y = y_pers) |>
  ggplot(aes(x = day, y = y)) +
  geom_line(color = "#1e3a5f", linewidth = 1) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "#888") +
  labs(
    title    = "Persistence (AR(1))",
    subtitle = "Stationary mean. Residuals carry memory.",
    x = "Day", y = "Outcome"
  ) +
  theme_minimal(base_size = 12)

p_trend + p_pers
```

A trend series looks like it has AR(1) structure — it autocorrelates — but the correct model includes time as a fixed effect, after which the residuals may be independent. AR(1) applies to the **residuals around the trend**, not to the raw series.

**JDM diagnostic question:** Is the serial correlation in your outcome driven by a steady directional shift (trend), or by short-term inertia around a stable mean (persistence)? The answer determines whether you need a trend term in your fixed effects, an AR(1) residual structure, or both.

---

## Detecting Autocorrelation in Your Residuals

The key diagnostic step: after fitting a model without an AR structure, check whether the residuals show serial dependence.

```{r detect-autocorr, fig.width=7, fig.height=4}
n_people <- 40
n_days   <- 20

# Simulate multilevel AR(1): daily spending over 20 days
df_ar <- expand_grid(person = 1:n_people, day = 1:n_days) |>
  arrange(person, day) |>
  group_by(person) |>
  mutate(
    baseline = rnorm(1, 50, 6),
    ar_resid = {
      r <- numeric(n_days); r[1] <- rnorm(1, 0, 3)
      for (t in 2:n_days) r[t] <- 0.7 * r[t-1] + rnorm(1, 0, 3)
      r
    },
    daily_spending = baseline + 2 * day / n_days + ar_resid
  ) |>
  ungroup() |>
  mutate(person = factor(person))

# Naive model without AR structure
m_naive <- lme(daily_spending ~ day,
               random = ~1 | person,
               data   = df_ar)

# Lag-1 autocorrelation of residuals per person
residuals_df <- data.frame(
  resid  = residuals(m_naive, type = "normalized"),
  person = df_ar$person,
  day    = df_ar$day
)

lag1_by_person <- residuals_df |>
  group_by(person) |>
  arrange(day) |>
  summarise(lag1 = cor(head(resid, -1), tail(resid, -1)))

ggplot(lag1_by_person, aes(x = lag1)) +
  geom_histogram(fill = "#1e3a5f", color = "white", bins = 15) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#e63946") +
  labs(
    title    = "Lag-1 autocorrelation of residuals (naive model)",
    subtitle = "Most values positive — AR structure in residuals that the model is ignoring",
    x        = "Lag-1 autocorrelation of residuals",
    y        = "Count"
  ) +
  theme_minimal(base_size = 13)
```

Most lag-1 residual autocorrelations are positive — strong evidence of serial dependence that the naive model is incorrectly treating as zero.

---

## Specifying AR(1) in nlme

```{r fit-ar1}
m_ar1 <- lme(daily_spending ~ day,
             random      = ~1 | person,
             correlation = corAR1(form = ~day | person),
             data        = df_ar)

tibble(
  Model = c("Naive (independent residuals)", "AR(1) residuals"),
  AIC   = c(AIC(m_naive), AIC(m_ar1)),
  BIC   = c(BIC(m_naive), BIC(m_ar1))
) |>
  knitr::kable(digits = 1,
    caption = "AR(1) substantially improves model fit (lower AIC/BIC = better)")
```

```{r ar1-estimate}
# What is the estimated autocorrelation?
intervals(m_ar1, which = "var-cov")$corStruct
```

---

## Assumptions

::: {.callout-warning icon=false}
## Assumption: Stationarity

**Plain-language meaning:** The mean and variance of the process are constant over time, and the autocorrelation depends only on the lag size — not on where you are in the study timeline. An AR(1) with $|\phi| < 1$ is stationary by construction.

**What the model is assuming away:** Any systematic drift in the mean (trend) or variance (heteroskedasticity over time). If consumers' risk perception is trending upward across a study because financial anxiety is increasing nationwide, the process is non-stationary.

**JDM example:** If daily risk perception increases during a financial crisis period within the study window, the mean is trending. AR(1) applied naively will have non-stationary residuals.

**Experimental analog:** In a daily diary experiment running across 8 weeks, if the pre-post intervention change creates a structural break in the mean, the AR(1) process applies separately within the pre- and post-intervention periods.

**What a skeptical reviewer might say:** "You applied AR(1) to a 12-week panel of financial diary data — but the data spans both the pre-policy and post-policy period. Did you account for the structural break in the mean at the policy change date?"

**Observable warning sign:** Wave-by-wave mean plots that show a systematic trend; ACF of raw residuals that does not decay to zero (persistent autocorrelation at long lags).

**What to check:** Include time as a fixed effect first. Examine whether residual ACF shows geometric decay (AR(1)) or long-run persistence (non-stationary). Use unit root tests if trend is suspected.

**What claim survives if the assumption is doubtful:** The sign and rough magnitude of the treatment effect are usually robust. The AR coefficient estimate and its confidence interval may be biased.
:::

::: {.callout-warning icon=false}
## Assumption: Lag-1 Dependence Only

**Plain-language meaning:** AR(1) assumes that $Y_{t-2}$ predicts $Y_t$ *only through* $Y_{t-1}$ — once you know yesterday's value, knowing the day before yesterday adds nothing.

**What the model is assuming away:** Longer-range memory. A consumer's spending behavior this week may depend on both last week *and* two weeks ago — a second-order AR process.

**JDM example:** Promotional spending behavior may have a two-week purchase cycle: today's spending is influenced by purchases from both last week (direct) and two weeks ago (restock timing), requiring AR(2).

**What to check:** Examine the partial autocorrelation function (PACF). If significant spikes appear at lag 2 or beyond after fitting AR(1), the model needs a higher order.

```{r pacf-check, fig.width=7, fig.height=3.5}
pacf_vals <- pacf(Y, plot = FALSE, lag.max = 10)
tibble(lag = 1:10, pacf = pacf_vals$acf[,1,1]) |>
  ggplot(aes(x = lag, y = pacf)) +
  geom_col(fill = "#1e3a5f", alpha = 0.8) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = c(-1, 1) * 1.96 / sqrt(n_t),
             linetype = "dashed", color = "#e63946") +
  labs(
    title    = "Partial ACF: only lag-1 significant = AR(1) is sufficient",
    subtitle = "Spikes at lag 2+ would indicate AR(2) or higher order",
    x = "Lag", y = "Partial ACF"
  ) +
  theme_minimal(base_size = 13)
```
:::

---

## Reviewer Challenge

::: {.callout-tip}
## What a Reviewer Would Ask

**Challenge 1:** "Your AR(1) coefficient is $\hat{\phi} = 0.72$. Does that mean consumers' financial decisions are mostly determined by their previous decision rather than your manipulation?"
A high AR(1) coefficient means the baseline level of the process is persistent — which is expected for daily financial behavior. The treatment effect is estimated *on top of* this persistence, as a departure from the AR trajectory. The AR coefficient describes the DGP's background noise structure, not the causal mechanism.

**Challenge 2:** "How do you distinguish AR(1) persistence from an unmodeled time trend that creates the appearance of autocorrelation?"
By including time as a fixed effect first. AR(1) persistence appears in residuals after removing trends. If the apparent AR disappears once time is included as a covariate, the "persistence" was actually a trend. Inspect the partial ACF after including time in the model — if lag-1 PACF is no longer significant, the trend was the source of apparent autocorrelation.

**Challenge 3:** "You estimated AR(1) for a daily 20-day series per person. With only 20 observations per person, how stable is the per-person autocorrelation estimate?"
Individual-level AR estimates are indeed noisy with short series. The `nlme` approach pools information across persons (assuming common AR structure), which is more stable than person-specific estimates. This is an advantage of the multilevel AR approach over fitting AR models separately per person.
:::
