---
title: "Autoregressive Structures"
---

```{r setup, include=FALSE}
library(tidyverse)
library(nlme)
library(ggplot2)
set.seed(42)
```

# Autoregressive Structures

## The Memory of Time

Some outcomes carry their history with them. Mood at 9am predicts mood at 10am. Anxiety on Monday affects anxiety on Tuesday. GDP in quarter 1 predicts GDP in quarter 2. This temporal dependency — the "memory" of the process — is not just nuisance variance. It is a feature of the DGP that your model must account for.

When you ignore serial dependence and treat residuals as independent, your standard errors are biased and your inference is wrong. More importantly, you are misrepresenting the process: you are claiming the outcome at each time point arises fresh, unconnected from what came before.

## What AR(1) Means

In an **autoregressive order 1** (AR(1)) process, each observation is a linear function of the immediately preceding observation plus random noise:

$$Y_t = \phi \cdot Y_{t-1} + \varepsilon_t, \quad \varepsilon_t \sim \mathcal{N}(0, \sigma^2_\varepsilon)$$

The parameter $\phi$ is the **autoregressive coefficient** — how strongly the past predicts the present. When $\phi = 0$, observations are independent. When $\phi = 0.9$, yesterday's value strongly predicts today's.

```{r ar1-example}
# Simulate AR(1) process for one person over 30 days
phi   <- 0.75
sigma <- 3
n_t   <- 30

Y <- numeric(n_t)
Y[1] <- rnorm(1, 50, 5)
for (t in 2:n_t) Y[t] <- phi * Y[t-1] + rnorm(1, 0, sigma)

p1 <- tibble(t = 1:n_t, Y = Y) |>
  ggplot(aes(x = t, y = Y)) +
  geom_line(color = "#1e3a5f", linewidth = 1) +
  geom_point(size = 2, color = "#1e3a5f", alpha = 0.7) +
  labs(title = paste0("AR(1) process: φ = ", phi),
       subtitle = "Each value depends on the previous",
       x = "Day", y = "Outcome") +
  theme_minimal(base_size = 13)

# ACF plot
acf_vals <- acf(Y, plot = FALSE, lag.max = 10)
p2 <- tibble(lag = 0:10, acf = acf_vals$acf[,1,1]) |>
  filter(lag > 0) |>
  ggplot(aes(x = lag, y = acf)) +
  geom_col(fill = "#1e3a5f", alpha = 0.8) +
  geom_hline(yintercept = 0, color = "black") +
  geom_hline(yintercept = c(-1, 1) * 1.96 / sqrt(n_t),
             linetype = "dashed", color = "#e63946") +
  labs(title = "Autocorrelation function (ACF)",
       subtitle = "Should decay exponentially for AR(1)",
       x = "Lag", y = "ACF") +
  theme_minimal(base_size = 13)

library(patchwork)
p1 + p2
```

## Detecting Autocorrelation in Your Residuals

The first diagnostic step: after fitting a model without an AR structure, check whether the residuals show serial dependence.

```{r detect-autocorr}
n_people <- 40
n_days   <- 20

# Simulate multilevel AR(1) data
df_ar <- expand_grid(person = 1:n_people, day = 1:n_days) |>
  arrange(person, day) |>
  group_by(person) |>
  mutate(
    baseline  = rnorm(1, 50, 6),
    ar_resid  = {
      r <- numeric(n_days); r[1] <- rnorm(1, 0, 3)
      for (t in 2:n_days) r[t] <- 0.7 * r[t-1] + rnorm(1, 0, 3)
      r
    },
    Y = baseline + 2 * day / n_days + ar_resid
  ) |>
  ungroup() |>
  mutate(person = factor(person))

# Fit naive model (ignores AR structure)
m_naive <- lme(Y ~ day, random = ~1 | person, data = df_ar)

# Check residual ACF
residuals_df <- data.frame(
  resid  = residuals(m_naive, type = "normalized"),
  person = df_ar$person,
  day    = df_ar$day
)

# Lag-1 autocorrelation per person
lag1_by_person <- residuals_df |>
  group_by(person) |>
  arrange(day) |>
  summarise(lag1 = cor(head(resid, -1), tail(resid, -1)))

ggplot(lag1_by_person, aes(x = lag1)) +
  geom_histogram(fill = "#1e3a5f", color = "white", bins = 15) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#e63946") +
  labs(
    title    = "Lag-1 autocorrelation of residuals (naive model)",
    subtitle = "Most values are positive — indicates AR structure in the residuals",
    x        = "Lag-1 autocorrelation",
    y        = "Count"
  ) +
  theme_minimal(base_size = 13)
```

Most lag-1 autocorrelations are positive — strong evidence of serial dependence that the naive model is ignoring.

## Specifying AR(1) in nlme

```{r fit-ar1}
# Correct model: AR(1) residual structure
m_ar1 <- lme(Y ~ day,
             random      = ~1 | person,
             correlation = corAR1(form = ~day | person),
             data        = df_ar)

# Compare AIC
tibble(
  Model = c("Naive (no AR)", "AR(1)"),
  AIC   = c(AIC(m_naive), AIC(m_ar1)),
  BIC   = c(BIC(m_naive), BIC(m_ar1))
) |>
  knitr::kable(digits = 1, caption = "AR(1) substantially improves model fit")
```

```{r ar1-estimate}
# What is the estimated autocorrelation?
intervals(m_ar1, which = "var-cov")$corStruct
```

## The Assumptions of AR(1) Models

::: {.callout-warning icon=false}
## Assumption 1: Stationarity

**What it means:** The mean and variance of the process are constant over time, and the autocorrelation depends only on the lag, not the absolute time point. An AR(1) process is stationary when $|\phi| < 1$.

**How to check:** Plot the mean and variance of your outcome at each time point. Systematic trends violate stationarity. Fit a detrended model (including time as a fixed effect) and check whether residuals are stationary.

**What to do if violated:** Include time trends as fixed effects. Consider differencing the outcome (analyzing change scores rather than raw scores).
:::

::: {.callout-warning icon=false}
## Assumption 2: Only Lag-1 Dependence

**What it means:** AR(1) assumes that $Y_{t-2}$ predicts $Y_t$ *only through* $Y_{t-1}$ — once you condition on yesterday, the day before yesterday adds no additional predictive power.

**How to check:** Examine the partial autocorrelation function (PACF). If the PACF shows significant values at lag 2 or beyond, you need AR(2) or a higher-order model.

```{r pacf-check}
pacf_vals <- pacf(Y, plot = FALSE, lag.max = 10)
tibble(lag = 1:10, pacf = pacf_vals$acf[,1,1]) |>
  ggplot(aes(x = lag, y = pacf)) +
  geom_col(fill = "#1e3a5f", alpha = 0.8) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = c(-1, 1) * 1.96 / sqrt(n_t),
             linetype = "dashed", color = "#e63946") +
  labs(title = "Partial ACF: only lag 1 significant = AR(1) is sufficient",
       x = "Lag", y = "Partial ACF") +
  theme_minimal(base_size = 13)
```
:::
