---
title: "Between-Unit vs. Within-Unit Effects"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
set.seed(202)
```

# Between-Unit vs. Within-Unit Effects

## The Same Variable, Two Stories

Consider studying whether a nation's GDP predicts its citizens' wellbeing. You have survey data from 60 countries over 10 years.

**The between-country question:** Do richer countries have happier citizens on average?
**The within-country question:** In years when a country's GDP grows, do its citizens report higher wellbeing?

These are completely different empirical questions. The between-country question picks up all stable differences between rich and poor countries (culture, institutions, climate, history). The within-country question is closer to a causal estimate of the effect of economic growth on wellbeing, net of these stable differences.

Research consistently finds that the answers are *different*: rich countries are happier than poor ones, but within a country, economic booms correlate with wellbeing only modestly (the "Easterlin Paradox").

## Separating the Effects: The Mundlak Device

The cleanest way to separate within- and between-unit effects is to include both in the same model:

1. Cluster-mean center the predictor (captures within-unit variation)
2. Include the cluster mean as a separate level-2 predictor (captures between-unit variation)

```{r mundlak-demo}
# Simulate country panel data
n_countries <- 40
n_years     <- 10

# Between: rich countries ARE happier (partly for other reasons)
country_gdp_avg  <- rnorm(n_countries, 50, 15)
country_wellbeing_base <- 50 + 0.6 * country_gdp_avg + rnorm(n_countries, 0, 8)

# Within: growth has only modest effect on wellbeing
df_gdp <- expand_grid(country = 1:n_countries, year = 1:n_years) |>
  mutate(
    gdp_raw       = country_gdp_avg[country] + rnorm(n(), 0, 5),
    wellbeing     = country_wellbeing_base[country] +
                    0.15 * (gdp_raw - country_gdp_avg[country]) +
                    rnorm(n(), 0, 4),
    country       = factor(country)
  ) |>
  group_by(country) |>
  mutate(
    gdp_mean = mean(gdp_raw),
    gdp_cmc  = gdp_raw - gdp_mean
  ) |>
  ungroup()

# Pooled (mixes effects)
m_pooled <- lmer(wellbeing ~ gdp_raw + (1 | country), data = df_gdp)

# Mundlak: both within and between
m_mundlak <- lmer(wellbeing ~ gdp_cmc + gdp_mean + (1 | country), data = df_gdp)

bind_rows(
  broom.mixed::tidy(m_pooled,  effects = "fixed") |> mutate(model = "Pooled"),
  broom.mixed::tidy(m_mundlak, effects = "fixed") |> mutate(model = "Mundlak")
) |>
  select(model, term, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "Within vs. between effects of GDP on wellbeing")
```

The Mundlak model reveals:
- **Within effect (gdp_cmc ≈ 0.15):** When a country's GDP is above its own average, wellbeing is modestly higher. Small effect.
- **Between effect (gdp_mean ≈ 0.6):** Countries with higher average GDP have substantially higher wellbeing — but this partly reflects other stable advantages of wealthy nations.

The pooled model blends these two stories into one coefficient that misrepresents both.

## The Assumption Test: Are Within and Between Effects Equal?

The formal test for whether within and between effects are equal is simple: if the CMC model's within coefficient ($\beta_{within}$) and between coefficient ($\beta_{between}$) are statistically indistinguishable, you can use the simpler pooled model. If they differ, they must be separated.

```{r wald-test}
library(car)
# Test: within effect = between effect?
linearHypothesis(m_mundlak, "gdp_cmc = gdp_mean")
```

A significant test (p < 0.05) means within and between effects differ — the pooled model is misspecified.
