---
title: "Simpson's Paradox"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(ggplot2)
library(patchwork)
set.seed(42)
```

# Simpson's Paradox

::: {.callout-note icon=false}
## When the Pooled Relationship Tells the Opposite Story

- **Question:** When consumers who receive more promotions are systematically from price-sensitive segments (lower baseline WTP), does pooling across segments tell you whether promotions work?
- **What goes wrong:** The pooled regression picks up the between-segment confound — price-sensitive segments receive more promotions AND have lower WTP — and reports a negative promotion effect. Within every segment, promotions increase purchase intent. The pooled estimate is wrong in both sign and magnitude.
- **The fix:** Decompose the analysis: compare the pooled slope to the within-cluster slope. If they differ in sign, you have a Simpson's Paradox. Use a mixed model or CMC decomposition (Module 2 Parts 1–2) that estimates the within-cluster effect separately from the between-cluster confound.
:::

## When the Aggregate Lies

Simpson's Paradox occurs when a relationship that exists in the aggregate data *reverses* (or dramatically changes) when you look within subgroups. It is not a statistical quirk — it is a direct consequence of ignoring nesting, and it is surprisingly common in social science data.

This chapter demonstrates Simpson's Paradox in three distinct research contexts:

1. **Cross-sectional nesting** — consumers nested in market segments
2. **Longitudinal panel data** — the within vs. between paradox: income and life satisfaction
3. **Within-subjects experiments** — the stimulus confound: WTP and fee transparency

Each context has a different structural cause and a different remedy. Together, they establish a single diagnostic principle: **always decompose before you report**.

---

## Context 1: Cross-Sectional Nesting — JDM Version

### The Consumer Segments × Promotional Messaging Example

Price-sensitive consumer segments receive more promotional messages AND have lower average purchase intent — not because promotions reduce intent, but because the segments are price-sensitive by nature (low brand loyalty, high price elasticity). Within every segment, consumers who receive more promotional messages show higher purchase intent. The pooled analysis reverses the sign.

```{r cross-sectional-sim}
n_consumers  <- 200
n_segments   <- 5

# Budget → Premium segments: high promo exposure AND low purchase intent → low promo AND high intent
seg_promo_avg    <- c(22, 17, 13, 9, 5)     # avg promo messages/month
seg_intent_base  <- c(28, 36, 46, 55, 62)   # baseline purchase intent (%) — loyalty proxy

df_consumers <- tibble(
  consumer = 1:n_consumers,
  segment  = rep(1:n_segments, each = n_consumers / n_segments)
) |>
  mutate(
    promo_msgs   = seg_promo_avg[segment] + rnorm(n_consumers, 0, 3),
    # Within-segment: more promo than usual → higher intent (promotions work)
    purch_intent = seg_intent_base[segment] +
                   1.8 * (promo_msgs - seg_promo_avg[segment]) +
                   rnorm(n_consumers, 0, 5),
    seg_label    = factor(segment,
                          labels = c("Budget", "Value", "Mixed", "Lifestyle", "Premium"))
  )
```

```{r cross-sectional-plot}
p_pooled <- ggplot(df_consumers, aes(x = promo_msgs, y = purch_intent)) +
  geom_point(aes(color = seg_label), alpha = 0.5) +
  geom_smooth(method = "lm", color = "black", linewidth = 1.5) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Pooled: more promotions → LOWER purchase intent?!",
       subtitle = "Pooled slope is negative — a sign reversal",
       x = "Promotional messages/month", y = "Purchase intent (%)", color = "Segment") +
  theme_minimal(base_size = 12)

p_within <- ggplot(df_consumers, aes(x = promo_msgs, y = purch_intent, color = seg_label)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Within segment: more promotions → HIGHER purchase intent",
       subtitle = "Within every segment, promotions increase intent",
       x = "Promotional messages/month", y = "Purchase intent (%)", color = "Segment") +
  theme_minimal(base_size = 12)

p_pooled / p_within
```

The pooled slope is negative: consumers who receive more promotional messages show lower purchase intent. The truth: budget segments receive the most promotions AND have the lowest baseline loyalty — not because promotions destroy intent, but because marketers target price-sensitive customers with promotions. Within any given segment, promotions work. The aggregate correlation is driven entirely by the confound between segment type and promotional targeting.

**The fix:** Use a mixed model that explicitly estimates the within-segment effect:

```{r cross-sectional-models}
m_pooled_cs <- lm(purch_intent ~ promo_msgs, data = df_consumers)
m_mixed_cs  <- lmer(purch_intent ~ promo_msgs + (1 | seg_label), data = df_consumers)

tibble(
  Model         = c("Pooled OLS", "Mixed model (within-segment)"),
  Promo_coeff   = c(coef(m_pooled_cs)["promo_msgs"],
                    fixef(m_mixed_cs)["promo_msgs"]),
  Direction     = c("Negative (wrong)", "Positive (correct)")
) |>
  knitr::kable(digits = 3,
               caption = "Promotional messaging coefficient: pooled vs. within-segment")
```

### Real-World Examples of Cross-Sectional Simpson's Paradox

**Berkeley admissions (1973):** Across departments, women were admitted at lower rates than men. Within departments, women were admitted at the same or higher rates. The paradox arose because women applied disproportionately to competitive departments.

**Bank loan defaults:** Branches in economically distressed areas have higher default rates AND also offer more financial coaching programs (targeted at at-risk customers). Pooled: coaching → higher defaults. Within branches: coaching → lower defaults. The confound is branch economic context.

**Wages across demographic groups:** Total wages can fall for every group while rising in the aggregate, if people are shifting from lower-paid to higher-paid occupations.

---

## Context 2: Longitudinal Panel Data — The Within vs. Between Paradox

### Income and Life Satisfaction: The Easterlin Paradox

Simpson's Paradox in panel data is so common it has its own name: the **Easterlin Paradox** — and it is the core motivation for the CMC/GMC decomposition covered in Module 2 Parts 1–2.

At any given point in time, richer people report higher life satisfaction. The cross-sectional (between-person) correlation is strongly positive. This has been used to argue that income causes happiness — that policies increasing income will increase well-being.

But when you follow the *same people* over time, earning more than usual barely predicts higher satisfaction. The within-person effect is near zero. This is the Easterlin Paradox: nations get richer over decades without becoming much happier; individuals earn more over their careers without becoming much happier.

The pooled OLS regression conflates these two very different stories into a single, misleading number.

### Simulation: 100 People × 8 Annual Waves

```{r panel-sim}
n_people <- 100
n_waves  <- 8

person_income_mean <- rnorm(n_people, 50, 15)  # stable between-person income level

df_panel_sp <- expand_grid(person = 1:n_people, wave = 1:n_waves) |>
  mutate(
    person_mean  = person_income_mean[person],
    income       = person_mean + rnorm(n(), 0, 5),       # small within-person variation
    income_gmc   = income - mean(income),                # grand-mean centered
    income_cmc   = income - person_mean,                 # within-person centered
    income_mean  = person_mean - mean(person_income_mean), # person mean centered on grand mean
    # True DGP: large between-person effect, small within-person effect
    satisfaction = 5 + 3.0 * income_mean + 0.4 * income_cmc + rnorm(n(), 0, 2),
    person       = factor(person)
  )
```

```{r panel-models}
# 1. Pooled OLS: conflates between and within
m_pooled <- lm(satisfaction ~ income_gmc, data = df_panel_sp)

# 2. Between-person: regress person means on person mean income
person_means <- df_panel_sp |>
  group_by(person) |>
  summarise(
    across(c(satisfaction, income_mean), mean),
    .groups = "drop"
  )

m_between <- lm(satisfaction ~ income_mean, data = person_means)

# 3. Within-person (CMC mixed model)
m_within  <- lmer(satisfaction ~ income_cmc + (1 | person), data = df_panel_sp)

# Summary table
tibble(
  Model          = c("Pooled OLS", "Between-person", "Within-person (CMC)"),
  Income_slope   = c(coef(m_pooled)["income_gmc"],
                     coef(m_between)["income_mean"],
                     fixef(m_within)["income_cmc"]),
  Interpretation = c("Blend of between and within (misleading)",
                     "Richer people are happier (selection effect)",
                     "Earning more than usual barely helps (causal estimate)")
) |>
  knitr::kable(digits = 3,
               caption = "Income–satisfaction slope: pooled vs. decomposed")
```

### Visualizing the Easterlin Paradox

```{r panel-plot}
# Sample 25 people for visual clarity
sampled_people <- sample(levels(df_panel_sp$person), 25)

p_spaghetti <- df_panel_sp |>
  filter(person %in% sampled_people) |>
  ggplot(aes(x = income, y = satisfaction, group = person)) +
  geom_line(alpha = 0.3, color = "#4393c3") +
  geom_point(alpha = 0.2, color = "#4393c3", size = 0.8) +
  geom_smooth(data = df_panel_sp |> filter(person %in% sampled_people),
              aes(group = NULL), method = "lm",
              color = "black", linewidth = 1.2, se = FALSE) +
  labs(title = "Panel: income and life satisfaction",
       subtitle = "Blue lines = within-person trajectories (nearly flat, \u03b2\u2248 0.4)\nBlack line = pooled OLS (follows between-person differences)",
       x = "Income", y = "Life satisfaction") +
  theme_minimal(base_size = 12)

p_between_scatter <- person_means |>
  ggplot(aes(x = income_mean, y = satisfaction)) +
  geom_point(alpha = 0.6, color = "#d73027") +
  geom_smooth(method = "lm", color = "#d73027", linewidth = 1.2) +
  labs(title = "Between-person: richer people report higher satisfaction",
       subtitle = paste0("Slope \u2248 ", round(coef(m_between)["income_mean"], 2),
                         " — large, but reflects WHO is rich, not what income does"),
       x = "Person-mean income (centered)", y = "Person-mean satisfaction") +
  theme_minimal(base_size = 12)

p_spaghetti / p_between_scatter
```

The spaghetti plot makes the paradox visible: individual trajectories over time are nearly flat (low within-person effect $\beta_W \approx 0.4$), but the scatter of person-level averages shows a strong positive slope ($\beta_B \approx 3.0$). The pooled OLS line follows the person-means, not the within-person trajectories.

### Why This Is Simpson's Paradox

The aggregate (pooled) relationship would suggest that policies raising income improve wellbeing substantially. But the within-person relationship reveals that once you account for who is rich versus poor, income gains do almost nothing for wellbeing.

**The policy implication:** A policy that increases everyone's income by 10% does not produce the 3.0-point wellbeing gain that cross-sectional data suggests — it produces the 0.4-point gain that within-person data reveals. Ignoring nesting leads to catastrophically wrong policy predictions.

### Connection to Module 2: The CMC/GMC Fix

The CMC/GMC decomposition is precisely the solution:

```r
m_full <- lmer(satisfaction ~ income_cmc + income_mean + (1 | person),
               data = df_panel_sp)
```

This model recovers both the within-person effect (`income_cmc` $\approx 0.4$) and the between-person effect (`income_mean` $\approx 3.0$) simultaneously, making the paradox explicit and the decomposition clean. This is exactly what Module 2 Parts 1 and 2 covered.

::: {.callout-important}
## The Panel Data Rule

In any longitudinal analysis, run both pooled OLS and a within-person model (fixed effects or CMC mixed model). If the coefficients differ substantially, you are in Simpson's Paradox territory. Report both. Explain which one answers your research question.
:::

---

## Context 3: Within-Subjects Experiments — The Stimulus Confound

### Fee Transparency, WTP, and Brand Familiarity

Simpson's Paradox can appear even in experiments when stimuli are not properly counterbalanced — when certain conditions happen to receive stimuli that systematically differ on a key attribute.

**The scenario:** A researcher tests whether *transparent fee disclosure* increases willingness to pay (WTP) for financial products. 50 consumers each evaluate 30 products: 15 displayed with full fee transparency and 15 with standard (opaque) disclosure.

**The flaw:** Products assigned to the transparent condition happen to be from *well-known brands* (higher brand familiarity). Products in the opaque condition are from lesser-known providers. Brand familiarity is a well-established driver of WTP — consumers pay more for brands they recognise, regardless of disclosure format.

**The result:** Pooled across consumers, transparent products attract higher WTP. But this is entirely because they are better-known brands. Within each level of brand familiarity, disclosure format has NO effect on WTP. This is Simpson's Paradox at the product level.

```{r experiment-sim}
n_par  <- 50
n_stim <- 30

par_effects  <- rnorm(n_par,  0, 5)
stim_effects <- rnorm(n_stim, 0, 4)
transparent  <- rep(c(0, 1), times = n_stim / 2)   # 0 = opaque, 1 = transparent
brand_fam    <- 3 + 2 * transparent + rnorm(n_stim, 0, 1)  # confound: transparent = better-known brands

df_exp_sp <- expand_grid(participant = 1:n_par, stimulus = 1:n_stim) |>
  mutate(
    par_eff   = par_effects[participant],
    stim_eff  = stim_effects[stimulus],
    transp_s  = transparent[stimulus],
    brand_s   = brand_fam[stimulus],
    # True DGP: brand familiarity drives WTP; transparency has NO direct effect
    wtp       = 50 + par_eff + stim_eff + 2 * brand_s + rnorm(n(), 0, 3),
    participant = factor(participant),
    stimulus    = factor(stimulus)
  )
```

### Fitting the Wrong and Correct Models

```{r experiment-models}
# Wrong model: ignores product-level confound and stimulus random effects
m_wrong   <- lmer(wtp ~ transp_s + (1 | participant),
                  data = df_exp_sp)

# Correct model: controls for brand familiarity + crossed stimulus RE
m_correct <- lmer(wtp ~ transp_s + brand_s + (1 | participant) + (1 | stimulus),
                  data = df_exp_sp)

# Compare transparency coefficients
tidy_wrong   <- broom.mixed::tidy(m_wrong,   effects = "fixed")
tidy_correct <- broom.mixed::tidy(m_correct, effects = "fixed")

bind_rows(
  tidy_wrong   |> filter(term == "transp_s") |>
    mutate(model = "Wrong (no brand familiarity control, no stimulus RE)"),
  tidy_correct |> filter(term == "transp_s") |>
    mutate(model = "Correct (brand familiarity + stimulus RE)")
) |>
  select(model, estimate, std.error, statistic) |>
  knitr::kable(digits = 3,
               caption = "Fee transparency coefficient: false positive vs. null result (statistic = t-value)")
```

The wrong model finds a statistically significant positive transparency effect — concluding that fee disclosure increases WTP. The correct model — which includes brand familiarity as a covariate and crossed product random effects — shows that transparency has no effect. The apparent WTP premium for transparent products disappears entirely once the brand familiarity confound is controlled.

### Visualizing the Stimulus Confound

```{r experiment-plot}
stimulus_summary <- df_exp_sp |>
  mutate(disclosure_label = ifelse(transp_s == 1, "transparent", "opaque")) |>
  group_by(stimulus, disclosure_label, brand_s) |>
  summarise(mean_wtp = mean(wtp), .groups = "drop")

p_disclosure_raw <- ggplot(stimulus_summary,
                            aes(x = disclosure_label, y = mean_wtp,
                                color = disclosure_label)) +
  geom_jitter(width = 0.1, alpha = 0.6, size = 2) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.3, linewidth = 0.8) +
  scale_color_manual(values = c("transparent" = "#d73027", "opaque" = "#4393c3")) +
  labs(title = "Apparent effect: transparent products attract higher WTP",
       subtitle = "This looks like a disclosure effect...",
       x = "Disclosure format", y = "Mean WTP (\u20ac)", color = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

p_brand <- ggplot(stimulus_summary,
                  aes(x = brand_s, y = mean_wtp,
                      color = disclosure_label)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(values = c("transparent" = "#d73027", "opaque" = "#4393c3")) +
  labs(title = "True structure: brand familiarity drives WTP",
       subtitle = "...but both lines have the same slope; disclosure just shifts brand familiarity",
       x = "Brand familiarity", y = "Mean WTP (\u20ac)",
       color = "Disclosure") +
  theme_minimal(base_size = 12)

p_disclosure_raw | p_brand
```

The right panel is diagnostic: the two regression lines (transparent and opaque) are nearly parallel — same slope for brand familiarity — and the gap between them narrows within each familiarity band. Transparent products appear higher overall only because they occupy higher positions on the brand familiarity axis.

### Why This Is Simpson's Paradox

Within each level of brand familiarity, transparent and opaque products attract the same WTP. But pooled across all products, transparent products look better because they happen to be well-known brands. This is precisely the Simpson's Paradox structure: the within-group relationship (null) and the aggregate relationship (positive) point in different directions.

### The Fix

1. **Design solution:** Properly counterbalance products across disclosure conditions — match transparent and opaque products on brand familiarity, product category, price range, and other relevant attributes. This removes the confound at the data collection stage.
2. **Analysis solution:** Include crossed product random effects (Module 1 Part 6) and known product attributes as covariates. This partials out the product-level confound after the fact — but requires that the confounding attribute was measured.

---

## Synthesis: The Diagnosis — Always Decompose Before You Report

Simpson's Paradox is not a special case. It is the predictable outcome whenever:

1. The predictor varies systematically across groups
2. The outcome varies systematically across groups
3. The group-level and unit-level relationships point in different directions

### Practical Checklist

**Cross-sectional nested data:**

- Compare the pooled coefficient to the within-cluster coefficient
- If they differ in sign or by more than 50%, you have a Simpson's Paradox
- Report the within-cluster coefficient; explain the between-cluster confound

**Panel data:**

- Always run both pooled OLS and a within-person model (FE or CMC)
- Compare $\hat{\beta}_W$ vs. $\hat{\beta}_B$: if they differ substantially, decompose and report both
- Ask: does the policy question target within-person change or between-person differences?

**Within-subjects experiments:**

- Include crossed stimulus random effects in all models
- Check whether stimuli are balanced on key attributes (frequency, concreteness, arousal, length) across conditions
- Include item-level covariates to control post-hoc stimulus confounds

::: {.callout-important}
## The Universal Check

Before reporting any correlation or regression coefficient from nested data, ask: **Does this relationship hold in the same direction within units as it does across units?**

If not, you are reporting a Simpson's Paradox. The CMC/GMC decomposition (Module 2) is the systematic fix.
:::

::: {.callout-tip icon=false}
## Reviewer Challenge: Three Questions About Aggregation Bias

**Challenge 1 — "Does your pooled result replicate within every subgroup?"**
The clearest diagnostic for Simpson's Paradox is to plot the relationship separately for each cluster (segment, branch, participant group) and compare to the pooled slope. If the within-cluster slopes all point opposite to the pooled slope, you have a paradox. A reviewer will ask: "Among all the consumer segments you studied, did promotions increase purchase intent within any of them? Or only in the aggregate?" Have a segment-by-segment breakdown ready.

**Challenge 2 — "What is the confounding variable at the cluster level?"**
Simpson's Paradox always has a structural cause — a between-cluster characteristic that correlates with both the predictor and the outcome. For the consumer segments example, it is segment price-sensitivity: budget segments receive more promotions AND have lower loyalty. For the Easterlin paradox, it is stable personality and circumstance. A reviewer will ask you to name the confound and justify why you believe it operates at the cluster level rather than the individual level. If you cannot name it, the reviewer may suspect unmeasured omitted variables.

**Challenge 3 — "Does controlling for the between-cluster confound change your conclusion?"**
If the within-cluster effect of promotions on purchase intent is positive (+1.8 per message), but the between-cluster relationship is negative (budget segments have high promo AND low intent), a reviewer will ask: when you control for segment-level characteristics (average income, price elasticity, brand loyalty) as fixed effects, does the within-cluster effect remain positive and significant? If it collapses to near zero after controlling for these characteristics, the within-cluster variation in promotions may itself be confounded by unmeasured segment-level factors.
:::


### Three-Panel Summary Visualization

```{r synthesis-plot, fig.width=12, fig.height=5}
# Panel 1: Cross-sectional Simpson (consumer segments)
p1 <- ggplot(df_consumers, aes(x = promo_msgs, y = purch_intent)) +
  geom_point(aes(color = seg_label), alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", color = "black", linewidth = 1.2) +
  geom_smooth(aes(color = seg_label), method = "lm",
              se = FALSE, linewidth = 0.8) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "1. Cross-sectional nesting",
       subtitle = "Pooled: negative\nWithin segment: positive",
       x = "Promo messages", y = "Purchase intent", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

# Panel 2: Panel data Simpson (person-level means vs within-person)
p2 <- ggplot(person_means, aes(x = income_mean, y = satisfaction)) +
  geom_point(alpha = 0.5, color = "#d73027", size = 1.5) +
  geom_smooth(method = "lm", color = "#d73027", linewidth = 1.2) +
  labs(title = "2. Panel data (Easterlin)",
       subtitle = paste0("Between persons: slope \u2248 ",
                         round(coef(m_between)["income_mean"], 1),
                         " (strong)\nWithin persons over time: slope \u2248 0.4 (weak)"),
       x = "Person-mean income (centered)", y = "Person-mean satisfaction") +
  theme_minimal(base_size = 11)

# Panel 3: Experiment Simpson
p3 <- ggplot(stimulus_summary,
             aes(x = brand_s, y = mean_wtp, color = disclosure_label)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("transparent" = "#d73027", "opaque" = "#4393c3")) +
  labs(title = "3. Within-subjects experiment",
       subtitle = "Marginal: transparent WTP higher\nWithin brand familiarity: no disclosure effect",
       x = "Brand familiarity", y = "Mean WTP (\u20ac)", color = "Disclosure") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p1 | p2 | p3
```

The table below summarizes the three contexts, their structural causes, and their remedies:

| Context | Mechanism | Aggregate relationship | Within-unit relationship | Remedy |
|---|---|---|---|---|
| Cross-sectional nesting | Segment price-sensitivity confounds promo–intent link | Negative (wrong) | Positive (correct) | Mixed model; within-segment slope |
| Panel data | Between-person selection confounds within-person change | Strongly positive (misleading) | Near zero (causal) | CMC decomposition; FE model |
| Experiment | Brand familiarity confounded with disclosure condition | Positive (false positive) | Null (correct) | Product RE; brand familiarity covariate |

::: {.callout-tip icon=false}
## Observational → Experimental Translation

**In observational panel data:**
The between-person and within-person slopes diverge when stable unit-level factors (personality, firm culture, socioeconomic status) drive both the predictor and the outcome. The "ecological fallacy" is comparing units when you should be comparing moments within a unit. CMC separates these two estimands.

**In within-subjects experiments:**
The between-stimulus and within-stimulus slopes diverge when stimuli vary systematically on a confound (e.g., positive words are also more concrete). Aggregating across stimuli without accounting for this confound produces a pooled slope that reflects the confound, not the experimental manipulation. The fix is the same: include the stimulus-level mean as a between-stimulus covariate alongside the within-stimulus deviation.

**Same assumption, different data collection context:**
In both cases, a single slope conflates two mechanisms. CMC/stimulus-mean centering is how you ask the data which mechanism is operating — and whether they point in the same direction.
:::
