---
title: "Centering as Decomposition"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(patchwork)
set.seed(101)
```

# Centering as Decomposition

## One Variable, Two Mechanisms

When you measure "hours of sleep" across multiple days from multiple people, that variable carries two completely different stories at the same time:

1. **The within-person story:** On days when *this person* sleeps more than they usually do, how does their mood change?
2. **The between-person story:** Do *people who sleep more overall* have higher mood than people who sleep less?

These are different empirical questions. They may have different answers. They probably operate through different *mechanisms* — the within-person effect may reflect acute physiological recovery, while the between-person effect may reflect stable personality traits (conscientious people both sleep more and have better mood, not because sleep causes mood but because conscientiousness causes both).

**Centering is not a cosmetic choice about where the intercept falls.** It is a decision about which mechanism you are modeling — and whether you want to see both.

---

## The Equations: From Conflation to Decomposition

### The Conflated Model

When you include an uncentered (or grand-mean-centered) predictor in a random-effects model, a single coefficient $\beta_1$ blends two mechanisms:

$$
\underbrace{Y_{ij}}_{\text{outcome}} = \underbrace{\gamma_{00}}_{\substack{\text{grand} \\ \text{mean}}} + \underbrace{\beta_1}_{\substack{\text{blended} \\ \text{slope}}} \underbrace{X_{ij}}_{\substack{\text{predictor}}} + \underbrace{u_{0j}}_{\substack{\text{person} \\ \text{RE}}} + \underbrace{\varepsilon_{ij}}_{\substack{\text{residual}}}
$$

$\beta_1$ absorbs variation from two sources simultaneously. Its value depends on which source dominates and how the two correlate.

### The Decomposition

CMC separates the predictor into its two components before entering the model:

$$
\underbrace{X_{ij}}_{\substack{\text{observed} \\ \text{predictor}}} = \underbrace{(X_{ij} - \bar{X}_{j})}_{\substack{\text{within-unit} \\ \text{deviation}}} + \underbrace{\bar{X}_{j}}_{\substack{\text{unit mean} \\ \text{(between)}}}
$$

Adding both as separate predictors gives the **decomposed model**. Notice how the single $\beta_1$ from before splits into two:

$$
\underbrace{Y_{ij}}_{\text{outcome}} = \underbrace{\gamma_{00}}_{\substack{\text{grand} \\ \text{mean}}} + \underbrace{\beta_W}_{\substack{\text{within-unit} \\ \text{slope}}} \underbrace{(X_{ij} - \bar{X}_j)}_{\substack{\text{deviation from} \\ \text{own mean}}} + \underbrace{\beta_B}_{\substack{\text{between-unit} \\ \text{slope}}} \underbrace{\bar{X}_j}_{\substack{\text{unit's} \\ \text{average level}}} + \underbrace{u_{0j}}_{\substack{\text{person} \\ \text{RE}}} + \underbrace{\varepsilon_{ij}}_{\substack{\text{residual}}}
$$

Now $\beta_W$ answers the within-person question and $\beta_B$ answers the between-person question. They are allowed to be completely different numbers with different interpretations.

::: {.callout-note}
## The DGP Interpretation

The decomposed model is a **hypothesis about the DGP**: it asserts that two distinct mechanisms are at work, one operating at the within-unit level and one at the between-unit level. If $\beta_W \neq \beta_B$, the DGP truly has two mechanisms. If $\beta_W \approx \beta_B$, a single mechanism may span both levels — but you only know this *after* estimating the decomposed model and testing whether they differ.
:::

---

## Three Archetypal Patterns

The most important practical question is: *how different are $\beta_W$ and $\beta_B$?*

### Pattern 1: Convergent Effects — Exercise and Wellbeing

Both mechanisms point the same direction with similar magnitude. The within-person mechanism (acute endorphins, reduced rumination) and the between-person mechanism (fit people have better overall routines and self-efficacy) both increase wellbeing. True: $\beta_W \approx \beta_B \approx 3$.

```{r exercise-wellbeing}
n_ppl <- 60; n_days <- 10

person_ex_avg <- rnorm(n_ppl, 45, 12)
person_wb_base <- 50 + 0.3 * person_ex_avg + rnorm(n_ppl, 0, 6)

df_ex <- expand_grid(person = 1:n_ppl, day = 1:n_days) |>
  mutate(
    exercise  = person_ex_avg[person] + rnorm(n(), 0, 8),
    wellbeing = person_wb_base[person] +
                3 * (exercise - person_ex_avg[person]) +
                rnorm(n(), 0, 5),
    person = factor(person)
  ) |>
  group_by(person) |>
  mutate(ex_mean = mean(exercise), ex_cmc = exercise - ex_mean) |>
  ungroup()

m_ex <- lmer(wellbeing ~ ex_cmc + ex_mean + (1 | person), data = df_ex)
broom.mixed::tidy(m_ex, effects = "fixed") |> filter(term != "(Intercept)") |>
  mutate(mechanism = c("Within-person", "Between-person")) |>
  select(mechanism, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "Pattern 1: within ≈ between ≈ 3. One coherent mechanism at two levels.")
```

Here, the pooled model would give you a decent approximation — but you only know that *after* running the decomposition. The convergence itself is substantively informative: the same process operates at both timescales.

### Pattern 2: Divergent Effects — Income and Life Satisfaction (Easterlin Paradox)

At the between-person level, wealthier people report substantially higher life satisfaction. But within individuals over time, income growth has a surprisingly modest effect — adaptation, shifting aspirations, and social comparison dampen the within-person response. True: $\beta_W \approx 0.3$, $\beta_B \approx 1.5$.

```{r income-satisfaction}
n_ppl2 <- 80; n_yrs <- 8

person_inc_avg  <- rnorm(n_ppl2, 50, 15)
person_sat_base <- 50 + 1.5 * person_inc_avg + rnorm(n_ppl2, 0, 8)

df_inc <- expand_grid(person = 1:n_ppl2, year = 1:n_yrs) |>
  mutate(
    income   = person_inc_avg[person] + rnorm(n(), 0, 6),
    life_sat = person_sat_base[person] +
               0.3 * (income - person_inc_avg[person]) + rnorm(n(), 0, 4),
    person   = factor(person)
  ) |>
  group_by(person) |>
  mutate(inc_mean = mean(income), inc_cmc = income - inc_mean) |>
  ungroup()

m_inc    <- lmer(life_sat ~ inc_cmc + inc_mean + (1 | person), data = df_inc)
m_pooled <- lmer(life_sat ~ income  + (1 | person), data = df_inc)

bind_rows(
  broom.mixed::tidy(m_pooled, effects = "fixed") |>
    filter(term == "income") |> mutate(model = "Pooled (blended)"),
  broom.mixed::tidy(m_inc, effects = "fixed") |>
    filter(term == "inc_cmc") |> mutate(model = "Within-person"),
  broom.mixed::tidy(m_inc, effects = "fixed") |>
    filter(term == "inc_mean") |> mutate(model = "Between-person")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Pattern 2: pooled ≈ 1.2, within ≈ 0.3, between ≈ 1.5. Pooled is wrong for both questions.")
```

The pooled model gives ~1.2 — it overstates the within-person effect by 4× and understates the between-person effect. A policy conclusion based on this ("raising incomes will substantially improve wellbeing") relies on the between-person relationship but must deliver on the within-person mechanism.

### Pattern 3: Sign Reversal — Study Hours and Grades

Within a student, studying more than their own average helps their grades. But between programs, the harder programs have students who study *more* and earn *lower* grades — because the programs are harder, not because studying hurts. True: $\beta_W \approx +2$, $\beta_B \approx -3$.

```{r study-grades}
n_students <- 200; n_programs <- 5
prog_difficulty <- c(2, 4, 6, 8, 10)
prog_study_avg  <- c(4, 5, 6, 7, 8)  # harder programs demand more study hours

df_study <- tibble(
  student = 1:n_students,
  program = rep(1:n_programs, each = n_students / n_programs)
) |>
  mutate(
    study = prog_study_avg[program] + rnorm(n_students, 0, 0.8),
    grade = 80 - 3 * prog_difficulty[program] +
            2 * (study - prog_study_avg[program]) + rnorm(n_students, 0, 3),
    program = factor(program)
  ) |>
  group_by(program) |>
  mutate(study_mean = mean(study), study_cmc = study - study_mean) |>
  ungroup()

m_study_decomp <- lmer(grade ~ study_cmc + study_mean + (1 | program), data = df_study)
m_study_pooled <- lm(grade ~ study, data = df_study)

bind_rows(
  broom::tidy(m_study_pooled) |>
    filter(term == "study") |> mutate(model = "Pooled OLS"),
  broom.mixed::tidy(m_study_decomp, effects = "fixed") |>
    filter(term == "study_cmc") |> mutate(model = "Within-program"),
  broom.mixed::tidy(m_study_decomp, effects = "fixed") |>
    filter(term == "study_mean") |> mutate(model = "Between-program")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Pattern 3: pooled is negative; within is positive; between is strongly negative. Sign reversal!")
```

```{r study-grades-plot}
p1 <- ggplot(df_study, aes(x = study, y = grade)) +
  geom_point(aes(color = program), alpha = 0.4, size = 1.2) +
  geom_smooth(method = "lm", color = "black", linewidth = 1.5) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Pooled: studying MORE → LOWER grades?",
       subtitle = "Slope is negative — a sign reversal",
       x = "Study hours/week", y = "Grade", color = "Program") +
  theme_minimal(base_size = 12)

p2 <- ggplot(df_study, aes(x = study, y = grade, color = program)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Within-program: studying MORE → HIGHER grades",
       subtitle = "Within every program, the slope is positive",
       x = "Study hours/week", y = "Grade", color = "Program") +
  theme_minimal(base_size = 12)

p1 + p2
```

The pooled model says studying hurts — which is wrong. The decomposition reveals the truth: within every program, studying helps, but harder programs demand more studying and are harder by definition. Ignoring the nesting produces a **Simpson's Paradox** (see Part 3).

---

## The Sleep-Mood Example: Applying Decomposition

```{r centering-demo}
n_people <- 50; n_days2 <- 7

person_sleep_avg  <- rnorm(n_people, 7, 1)
person_mood_base  <- 40 + 5 * person_sleep_avg + rnorm(n_people, 0, 5)

df <- expand_grid(person = 1:n_people, day = 1:n_days2) |>
  mutate(
    sleep_raw = person_sleep_avg[person] + rnorm(n(), 0, 1),
    mood      = person_mood_base[person] +
                2 * (sleep_raw - person_sleep_avg[person]) + rnorm(n(), 0, 4),
    person    = factor(person)
  ) |>
  group_by(person) |>
  mutate(sleep_mean = mean(sleep_raw), sleep_cmc = sleep_raw - sleep_mean) |>
  ungroup() |>
  mutate(sleep_gmc = sleep_raw - mean(sleep_raw))

m_uncentered <- lmer(mood ~ sleep_raw  + (1 | person), data = df)
m_gmc        <- lmer(mood ~ sleep_gmc  + (1 | person), data = df)
m_cmc        <- lmer(mood ~ sleep_cmc + sleep_mean + (1 | person), data = df)

bind_rows(
  broom.mixed::tidy(m_uncentered, effects = "fixed") |>
    filter(term == "sleep_raw") |>
    mutate(model = "Uncentered", mechanism = "Blended"),
  broom.mixed::tidy(m_gmc, effects = "fixed") |>
    filter(term == "sleep_gmc") |>
    mutate(model = "GMC", mechanism = "Blended"),
  broom.mixed::tidy(m_cmc, effects = "fixed") |>
    filter(term == "sleep_cmc") |>
    mutate(model = "CMC", mechanism = "Within-person"),
  broom.mixed::tidy(m_cmc, effects = "fixed") |>
    filter(term == "sleep_mean") |>
    mutate(model = "CMC", mechanism = "Between-person")
) |>
  select(model, mechanism, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Sleep → mood: blended ≈ 4.0; within ≈ 2.0; between ≈ 5.0. Neither blended estimate is correct.")
```

The uncentered and GMC models give ~4, which is wrong for both questions. CMC reveals: the within-person mechanism is modest (2 points per extra hour of sleep); the between-person correlation is much larger (5 points) because healthy lifestyle confounds it.

---

## Decision Guide

::: {.callout-tip}
## Which Centering to Use

| Goal | Centering | What you get |
|------|-----------|--------------|
| Interpret the intercept at a meaningful value | GMC | Intercept = E[Y] when X = grand mean |
| Estimate the within-unit effect | CMC | $\hat\beta_W$ — deviation from own mean |
| Separate within and between mechanisms | CMC + unit mean | $\hat\beta_W$ and $\hat\beta_B$ |
| Test if within = between | CMC + unit mean | Wald test on equality |
| Single-level data (no nesting) | Either | No within/between distinction |

**Default recommendation for nested data:** Always use CMC with the unit mean included. Run it, check whether $\hat\beta_W \approx \hat\beta_B$, and only then decide whether the pooled model is defensible.
:::

::: {.callout-warning icon=false}
## Assumption: No Conflation of Within and Between Effects

Including a predictor without CMC in a random-effects model implicitly assumes $\beta_W = \beta_B$. When this is wrong — as in Patterns 2 and 3 above — both estimates are biased.

**How to test:** Fit the CMC model. Compare $\hat\beta_W$ and $\hat\beta_B$. A Wald test (see Part 2) tests the null $H_0: \beta_W = \beta_B$.

**The stakes:** In the worst case (Pattern 3), conflation completely reverses the direction of the effect.
:::
