---
title: "Separating Within and Between Effects"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(car)
library(patchwork)
set.seed(202)
```

# Separating Within and Between Effects

::: {.callout-note icon=false}
## When the Blended Estimate Fails Both Questions at Once

- **Question:** You have a within-subjects observation (a customer's monthly savings) and a predictor (disclosure engagement). Your scientific question is: does engaging with the disclosure document cause savings to increase? But the pooled model also picks up the fact that financially engaged customers are generally better savers. Which question does your estimate actually answer?
- **What goes wrong:** A random-effects model with an uncentered predictor implicitly assumes that the within-customer effect of disclosure equals the between-customer correlation. When these differ — and in consumer financial behaviour, they almost always do — the pooled estimate is biased for both questions.
- **The fix:** The Mundlak device (cluster-mean centering + unit mean in the model) gives you two separate estimates: the within-customer causal question and the between-customer comparison. The Wald test tells you whether they are statistically distinguishable.
:::

## The Mundlak Device

Part 1 showed that a predictor in nested data carries two mechanisms. The formal statistical tool for separating them is known as the **Mundlak device** (Mundlak, 1978) or equivalently the **within-between (WB) model**.

The key equation — building on Part 1's decomposition — is:

$$
\underbrace{Y_{ij}}_{\text{outcome}} = \underbrace{\gamma_{00}}_{\substack{\text{grand} \\ \text{mean}}} + \underbrace{\beta_W (X_{ij} - \bar{X}_j)}_{\substack{\text{within-unit effect} \\ \text{(deviation from own mean)}}} + \underbrace{\beta_B \bar{X}_j}_{\substack{\text{between-unit effect} \\ \text{(unit's average level)}}} + \underbrace{u_{0j}}_{\substack{\text{unit random} \\ \text{intercept}}} + \underbrace{\varepsilon_{ij}}_{\substack{\text{residual}}}
$$

The **Mundlak insight**: this model is equivalent to including the unit mean $\bar{X}_j$ alongside the raw predictor $X_{ij}$ in a random-effects model. The coefficient on $X_{ij}$ then estimates the within-unit effect, and the coefficient on $\bar{X}_j$ estimates the *additional* between-unit effect beyond the within-unit effect.

Both are valid. The CMC parameterization (Part 1) gives $\beta_W$ and $\beta_B$ directly and makes interpretation cleaner. The Mundlak parameterization adds $\bar{X}_j$ alongside the raw $X_{ij}$ and gives a test of whether $\beta_W = \beta_B$.

---

## Example 1: The Easterlin Paradox — GDP and Wellbeing

The classic motivating case. Between countries, richer nations are happier. Within countries over time, economic growth has a modest and inconsistent effect on wellbeing. Two different mechanisms, same predictor.

```{r mundlak-demo}
n_countries <- 40; n_years <- 10

country_gdp_avg       <- rnorm(n_countries, 50, 15)
country_wellbeing_base <- 50 + 0.6 * country_gdp_avg + rnorm(n_countries, 0, 8)

df_gdp <- expand_grid(country = 1:n_countries, year = 1:n_years) |>
  mutate(
    gdp_raw   = country_gdp_avg[country] + rnorm(n(), 0, 5),
    wellbeing = country_wellbeing_base[country] +
                0.15 * (gdp_raw - country_gdp_avg[country]) + rnorm(n(), 0, 4),
    country   = factor(country)
  ) |>
  group_by(country) |>
  mutate(gdp_mean = mean(gdp_raw), gdp_cmc = gdp_raw - gdp_mean) |>
  ungroup()

m_pooled  <- lmer(wellbeing ~ gdp_raw + (1 | country), data = df_gdp)
m_mundlak <- lmer(wellbeing ~ gdp_cmc + gdp_mean + (1 | country), data = df_gdp)

bind_rows(
  broom.mixed::tidy(m_pooled,  effects = "fixed") |>
    filter(term == "gdp_raw") |> mutate(model = "Pooled"),
  broom.mixed::tidy(m_mundlak, effects = "fixed") |>
    filter(term == "gdp_cmc")  |> mutate(model = "Within-country"),
  broom.mixed::tidy(m_mundlak, effects = "fixed") |>
    filter(term == "gdp_mean") |> mutate(model = "Between-country")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "GDP → wellbeing: pooled ~0.45 blends within ~0.15 and between ~0.60")
```

**Interpretation:**
- **Within-country (gdp_cmc ≈ 0.15):** In years when GDP is above a country's own average, wellbeing ticks up slightly. Modest — consistent with hedonic adaptation and shifting aspirations.
- **Between-country (gdp_mean ≈ 0.60):** Countries with permanently higher GDP have substantially higher wellbeing. But this reflects institutions, infrastructure, healthcare, and social safety nets that co-occur with wealth — not just income itself.
- **Pooled (~0.45):** An average that describes neither mechanism accurately.

---

## Example 2: Within-Subjects Experiments — Trait vs. State Effects

In a within-subjects experiment, the same decomposition applies — but now the "cluster" is the participant and the nesting is observations within person.

**Design:** 80 participants each evaluate 10 charity appeals. Half feature statistical evidence; half feature personal stories. Outcome: whether they donate (continuous willingness-to-donate scale, 0–100).

**The within-person effect** ($\beta_W$): For a given participant, do story-based appeals get higher donations than statistical appeals — compared to that participant's *own average* donation?

**The between-person effect** ($\beta_B$): Do participants who are exposed to more story-based appeals (by random assignment variation) tend to donate more overall? This would pick up trait-level empathizers who respond to everything more generously.

With random assignment, $\beta_B$ should be near zero — randomization ensures that more story-exposed participants are not systematically more empathetic. A large $\beta_B$ is a red flag.

```{r experiment-demo}
n_p_exp <- 80; n_appeals <- 10

# Some participants are trait high-empathy: they donate more AND
# in this (imperfect) design, happen to see more story appeals
person_empathy     <- rnorm(n_p_exp, 50, 10)
person_story_prop  <- 0.5 + 0.01 * (person_empathy - 50) + rnorm(n_p_exp, 0, 0.05)
person_story_prop  <- pmin(pmax(person_story_prop, 0), 1)
person_base_donate <- 40 + 0.8 * person_empathy + rnorm(n_p_exp, 0, 8)

df_exp <- expand_grid(participant = 1:n_p_exp, appeal = 1:n_appeals) |>
  mutate(
    story = rbinom(n(), 1, person_story_prop[participant]),
    donate = person_base_donate[participant] +
             5 * (story - person_story_prop[participant]) +
             rnorm(n(), 0, 6),
    participant = factor(participant)
  ) |>
  group_by(participant) |>
  mutate(story_mean = mean(story), story_cmc = story - story_mean) |>
  ungroup()

m_exp <- lmer(donate ~ story_cmc + story_mean + (1 | participant), data = df_exp)
broom.mixed::tidy(m_exp, effects = "fixed") |> filter(term != "(Intercept)") |>
  mutate(mechanism = c("Within-person (causal effect)", "Between-person (trait confound)")) |>
  select(mechanism, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Story vs. stats: within ≈ 5 (treatment effect); between ≈ near zero (good randomization)")
```

With good randomization, $\beta_B \approx 0$ and $\beta_W$ recovers the true treatment effect. A significant $\beta_B$ would suggest a randomization failure or a design confound — worth always checking in within-subjects designs.

---

## Example 3: Ecological Fallacy — Education and Fertility

Between countries: more educated countries have lower fertility (strong negative relationship). Within countries over time: rising education predicts declining fertility (also negative). Here the signs align, but the magnitudes and mechanisms differ.

The *between-country* effect reflects structural factors: cultural norms, female labor force participation, urbanization, and healthcare access all co-vary with education across nations.

The *within-country* effect over time reflects a causal mechanism: as individual women gain education, they delay marriage, have fewer children, and invest more per child.

Same direction, but if you use the cross-national correlation to inform policy about what happens when you increase education within a country, you are using the wrong estimate — the between-country relationship is confounded by dozens of co-varying structural factors.

```{r education-fertility}
n_countries_ef <- 50; n_years_ef <- 8

country_edu_avg <- rnorm(n_countries_ef, 60, 12)
country_fert_base <- 4.5 - 0.05 * country_edu_avg + rnorm(n_countries_ef, 0, 0.4)

df_ef <- expand_grid(country = 1:n_countries_ef, year = 1:n_years_ef) |>
  mutate(
    education = country_edu_avg[country] + rnorm(n(), 0, 5),
    fertility = country_fert_base[country] -
                0.01 * (education - country_edu_avg[country]) + rnorm(n(), 0, 0.2),
    country   = factor(country)
  ) |>
  group_by(country) |>
  mutate(edu_mean = mean(education), edu_cmc = education - edu_mean) |>
  ungroup()

m_ef <- lmer(fertility ~ edu_cmc + edu_mean + (1 | country), data = df_ef)
broom.mixed::tidy(m_ef, effects = "fixed") |> filter(term != "(Intercept)") |>
  mutate(mechanism = c("Within-country over time", "Between-country cross-section")) |>
  select(mechanism, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Education → fertility: both negative, but between is 5× larger due to structural confounds")
```

The between-country estimate is 5× larger than the within-country estimate. Using the cross-sectional number to predict what happens when you raise education *within* a country would dramatically overstate the policy effect.

---

## Example 4: Convergent Effects — Consumer Financial Information Exposure and Savings

Not every decomposition reveals a divergence. When within and between effects are approximately equal, this is itself a substantive finding: the same mechanism operates at both levels.

**Monthly information exposure (coaching contact hours) → savings rate.** Within a customer: months where they have *more coaching contact than usual*, their savings rate increases (the coaching content is being absorbed and acted on). Between customers: customers with *higher average coaching contact* also have higher savings (both due to coaching AND because financially motivated customers seek more contact). Both effects are positive and of similar magnitude.

```{r coaching-savings}
n_cust  <- 100; n_months_c <- 8

cust_coach_avg  <- rnorm(n_cust, 3, 1)
cust_save_base  <- 20 + 3 * cust_coach_avg + rnorm(n_cust, 0, 5)

df_coach <- expand_grid(customer = 1:n_cust, month = 1:n_months_c) |>
  mutate(
    coaching = pmax(0, cust_coach_avg[customer] + rnorm(n(), 0, 0.8)),
    savings  = cust_save_base[customer] +
               2.5 * (coaching - cust_coach_avg[customer]) + rnorm(n(), 0, 4),
    customer = factor(customer)
  ) |>
  group_by(customer) |>
  mutate(coach_mean = mean(coaching), coach_cmc = coaching - coach_mean) |>
  ungroup()

m_coach <- lmer(savings ~ coach_cmc + coach_mean + (1 | customer), data = df_coach)
broom.mixed::tidy(m_coach, effects = "fixed") |> filter(term != "(Intercept)") |>
  mutate(mechanism = c("Within-customer (coaching → savings)", "Between-customer (engagement level)")) |>
  select(mechanism, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Coaching → savings: within ≈ between ≈ 3. Pooled model is approximately correct here.")
```

When $\beta_W \approx \beta_B$, the pooled model is defensible. But again: **you only know this after running the decomposition.** The convergence is informative — it suggests coaching contact itself (not just the trait of financial motivation that drives both coaching uptake and savings) is the active mechanism. This finding strengthens the case for scaling up the coaching program.

---

## Testing Whether Within and Between Effects Are Equal

The formal test is a **Wald test** on the null hypothesis $H_0: \beta_W = \beta_B$:

```{r wald-test}
# Test on the Easterlin model: are within and between GDP effects equal?
linearHypothesis(m_mundlak, "gdp_cmc = gdp_mean")
```

A significant result (p < 0.05) means you must report both effects separately — the pooled model is misspecified. A non-significant result means you *cannot rule out* that the effects are equal, and a pooled model may be defensible (though the CMC model is always more informative).

---

## A Decision Table

::: {.callout-tip}
## What to Report: Within, Between, or Both?

| Pattern | $\beta_W$ vs. $\beta_B$ | Wald test | What to report |
|---------|------------------------|-----------|----------------|
| Convergent | Similar magnitude and sign | p > .05 | Pooled estimate is OK, but note the decomposition confirmed equivalence |
| Divergent | Different magnitudes | p < .05 | Report both; discuss different mechanisms |
| Sign reversal | Opposite signs | p < .05 | Report both; warn about Simpson's Paradox |
| Experimental | $\beta_B \approx 0$, $\beta_W$ = treatment effect | p < .05 | Report within as the treatment effect; $\beta_B$ is a randomization check |

**Always run the decomposed model first.** Use the Wald test to decide whether to simplify. Do not start with the pooled model.
:::

::: {.callout-warning icon=false}
## Assumption: Correct Level Attribution

Including a variable as within-unit (CMC) or between-unit (unit mean) requires knowing which level it genuinely varies at. A variable that appears to vary within units but is actually stable should be treated as a level-2 predictor.

**How to check:** Compute the ICC of the predictor itself. High ICC (>0.70) means most variation is between units; CMC will give you noisy, low-variance within deviations. Low ICC means the variable genuinely varies at the within-unit level.
:::

::: {.callout-tip icon=false}
## Reviewer Challenge: Three Questions About Your Within–Between Decomposition

**Challenge 1 — "Is the between-customer estimate a selection effect or a dosage effect?"**
When $\hat\beta_B > \hat\beta_W$ for coaching contact, a reviewer will ask: does higher average coaching contact *cause* higher savings, or do financially motivated customers seek out more contact AND save more independently? The Mundlak device separates the estimates but does not resolve the causal interpretation of $\hat\beta_B$. If customers self-select into higher coaching tiers based on unobserved financial goals, the between-customer estimate is still a selection effect. Make clear what causal claim you are making about each component.

**Challenge 2 — "Did you check whether the Wald test has adequate power?"**
A non-significant Wald test ($H_0: \beta_W = \beta_B$) only tells you that the data are consistent with equal effects — it does not confirm they are equal. With small $n$ at the household level, the test has low power to detect moderate differences between within and between effects. Report the confidence interval for $\hat\beta_W - \hat\beta_B$ alongside the *p*-value. If the interval is wide, the data are uninformative about whether the mechanisms differ.

**Challenge 3 — "What is the policy-relevant quantity?"**
Reviewers will ask which estimate to use for intervention design. If a financial coaching program is rolled out at the branch level (all customers in a branch receive more coaching), the relevant estimate is the within-customer effect: does additional coaching, beyond a customer's own baseline, change their behaviour? The between-customer estimate is confounded by the selection process that determines who receives more coaching. Be explicit about which estimand matches the policy question.
:::

