---
title: "Centering Your Variables"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
set.seed(101)
```

# Centering Your Variables

## Why Centering Matters

When you include a predictor in a mixed-effects model, what exactly is it measuring? The answer depends entirely on how you center it. This is not a cosmetic choice — it changes the *meaning* of your coefficients.

## Grand-Mean Centering (GMC)

**Grand-mean centering** subtracts the overall mean from each observation:
$$X^{GMC}_{ij} = X_{ij} - \bar{X}_{..}$$

In a GMC model, the slope reflects a mix of within-unit and between-unit variation. The intercept now reflects the expected outcome when $X$ equals its grand mean — a meaningful value for continuous predictors.

GMC is appropriate when:
- You have a single-level dataset (no nesting)
- You are not trying to separate within- and between-person effects
- You want the intercept to be meaningful (grand mean of X rather than X = 0)

## Cluster-Mean Centering (CMC)

**Cluster-mean centering** (also called person-mean centering or within-person centering) subtracts each unit's mean from their own observations:
$$X^{CMC}_{ij} = X_{ij} - \bar{X}_{j.}$$

The centered variable now captures only within-unit deviations. For a daily diary study, $X^{CMC}_{ij}$ = "how much more/less did person $j$ sleep on day $i$ compared to their own average?"

The between-unit effect of X can be recovered separately by including the unit mean ($\bar{X}_{j.}$) as a level-2 predictor.

```{r centering-demo}
# Simulate: people differ in average sleep AND have within-person variation
n_people <- 50
n_days   <- 7

# Between-person: high sleepers have high mood (confound: healthy lifestyle)
person_sleep_avg <- rnorm(n_people, 7, 1)
person_mood_base <- 40 + 5 * person_sleep_avg + rnorm(n_people, 0, 5)

# Within-person: on high-sleep days, mood is slightly higher
df <- expand_grid(person = 1:n_people, day = 1:n_days) |>
  mutate(
    sleep_raw   = person_sleep_avg[person] + rnorm(n(), 0, 1),
    mood        = person_mood_base[person] + 2 * (sleep_raw - person_sleep_avg[person]) +
                  rnorm(n(), 0, 4),
    person      = factor(person)
  )

# Compute centering variables
df <- df |>
  group_by(person) |>
  mutate(
    sleep_person_mean = mean(sleep_raw),        # L2 variable (between)
    sleep_cmc         = sleep_raw - sleep_person_mean  # within-person deviation
  ) |>
  ungroup() |>
  mutate(
    sleep_gmc = sleep_raw - mean(sleep_raw)     # grand-mean centered
  )

# Three models
m_uncentered <- lmer(mood ~ sleep_raw + (1 | person), data = df)
m_gmc        <- lmer(mood ~ sleep_gmc + (1 | person), data = df)
m_cmc        <- lmer(mood ~ sleep_cmc + sleep_person_mean + (1 | person), data = df)

# Compare the sleep coefficient across models
results <- bind_rows(
  broom.mixed::tidy(m_uncentered, effects = "fixed") |> filter(term == "sleep_raw") |> mutate(model = "Uncentered"),
  broom.mixed::tidy(m_gmc, effects = "fixed")        |> filter(term == "sleep_gmc") |> mutate(model = "GMC"),
  broom.mixed::tidy(m_cmc, effects = "fixed")        |> filter(term == "sleep_cmc") |> mutate(model = "CMC (within)"),
  broom.mixed::tidy(m_cmc, effects = "fixed")        |> filter(term == "sleep_person_mean") |> mutate(model = "CMC (between)")
) |>
  select(model, term, estimate, std.error, p.value)

knitr::kable(results, digits = 3,
  caption = "Centering changes what the sleep coefficient means")
```

**Reading the table:**

- **Uncentered / GMC:** The slope is a blend of within- and between-person effects. Its magnitude is pulled by the larger between-person correlation.
- **CMC (within):** Close to the true within-person effect (~2). This answers "on days when you sleep more than usual, is your mood better?"
- **CMC (between):** The between-person effect (~5). This answers "do people who generally sleep more also have higher baseline mood?" — a question contaminated by lifestyle confounds.

The within-effect and the between-effect are different numbers with different interpretations. A model that conflates them gives you neither correctly.

::: {.callout-tip}
## When to Use Each

- **Grand-mean centering:** When you only care about the overall relationship, not within/between decomposition, and you want a meaningful intercept.
- **Cluster-mean centering:** Whenever you have a within-unit predictor and want to explicitly separate within-unit from between-unit effects. Default recommendation for longitudinal and panel data.
- **Uncentered:** When X = 0 is a meaningful value AND you do not need to separate within/between effects (rare in nested data).
:::

## Assumptions That Centering Affects

Centering is not just an interpretive choice — it affects which assumptions your model makes:

::: {.callout-warning icon=false}
## Assumption: No Conflation of Within and Between Effects

If you include a predictor without CMC in a random-effects model, the model assumes the within-unit and between-unit relationships are equal. If they are not equal — as in the example above — this is a misspecification that biases both estimates.

**How to test:** Fit both the uncentered and the CMC model. If the within-person and between-person coefficients from the CMC model differ substantially, the conflation assumption is violated.
:::
