---
title: "Centering as Decomposition"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(patchwork)
set.seed(101)
```

# Centering as Decomposition

::: {.callout-note icon=false}
## What Does Centering Reveal?

- **Question:** A single predictor in nested data carries two different relationships simultaneously — one operating within units (how does savings change when *this household* engages more with disclosure this month?) and one operating between units (do households that engage more overall save more?). Which one does your model estimate?
- **What goes wrong:** An uncentered predictor in a random-effects model blends both mechanisms into a single coefficient. If the within-unit and between-unit effects differ in size or sign, this blended estimate is wrong for both questions — and useless for both policy and theory.
- **The fix:** Cluster-mean centering (CMC) separates the predictor into a within-unit deviation (from the household's own average) and a between-unit mean. The model then estimates both effects simultaneously, allowing you to determine whether the same mechanism operates at both levels.
:::

## One Variable, Two Mechanisms

When you measure "hours of sleep" across multiple days from multiple people, that variable carries two completely different stories at the same time:

1. **The within-person story:** On days when *this person* sleeps more than they usually do, how does their mood change?
2. **The between-person story:** Do *people who sleep more overall* have higher mood than people who sleep less?

These are different empirical questions. They may have different answers. They probably operate through different *mechanisms* — the within-person effect may reflect acute physiological recovery, while the between-person effect may reflect stable personality traits (conscientious people both sleep more and have better mood, not because sleep causes mood but because conscientiousness causes both).

**Centering is not a cosmetic choice about where the intercept falls.** It is a decision about which mechanism you are modeling — and whether you want to see both.

---

## The Equations: From Conflation to Decomposition

### The Conflated Model

When you include an uncentered (or grand-mean-centered) predictor in a random-effects model, a single coefficient $\beta_1$ blends two mechanisms:

$$
\underbrace{Y_{ij}}_{\text{outcome}} = \underbrace{\gamma_{00}}_{\substack{\text{grand} \\ \text{mean}}} + \underbrace{\beta_1}_{\substack{\text{blended} \\ \text{slope}}} \underbrace{X_{ij}}_{\substack{\text{predictor}}} + \underbrace{u_{0j}}_{\substack{\text{person} \\ \text{RE}}} + \underbrace{\varepsilon_{ij}}_{\substack{\text{residual}}}
$$

$\beta_1$ absorbs variation from two sources simultaneously. Its value depends on which source dominates and how the two correlate.

### The Decomposition

CMC separates the predictor into its two components before entering the model:

$$
\underbrace{X_{ij}}_{\substack{\text{observed} \\ \text{predictor}}} = \underbrace{(X_{ij} - \bar{X}_{j})}_{\substack{\text{within-unit} \\ \text{deviation}}} + \underbrace{\bar{X}_{j}}_{\substack{\text{unit mean} \\ \text{(between)}}}
$$

Adding both as separate predictors gives the **decomposed model**. Notice how the single $\beta_1$ from before splits into two:

$$
\underbrace{Y_{ij}}_{\text{outcome}} = \underbrace{\gamma_{00}}_{\substack{\text{grand} \\ \text{mean}}} + \underbrace{\beta_W}_{\substack{\text{within-unit} \\ \text{slope}}} \underbrace{(X_{ij} - \bar{X}_j)}_{\substack{\text{deviation from} \\ \text{own mean}}} + \underbrace{\beta_B}_{\substack{\text{between-unit} \\ \text{slope}}} \underbrace{\bar{X}_j}_{\substack{\text{unit's} \\ \text{average level}}} + \underbrace{u_{0j}}_{\substack{\text{person} \\ \text{RE}}} + \underbrace{\varepsilon_{ij}}_{\substack{\text{residual}}}
$$

Now $\beta_W$ answers the within-person question and $\beta_B$ answers the between-person question. They are allowed to be completely different numbers with different interpretations.

::: {.callout-note}
## The DGP Interpretation

The decomposed model is a **hypothesis about the DGP**: it asserts that two distinct mechanisms are at work, one operating at the within-unit level and one at the between-unit level. If $\beta_W \neq \beta_B$, the DGP truly has two mechanisms. If $\beta_W \approx \beta_B$, a single mechanism may span both levels — but you only know this *after* estimating the decomposed model and testing whether they differ.
:::

---

## Why Not Just Use Pooled Regression?

::: {.callout-important icon=false}
## The Case Against Pooled OLS (and Uncentered Random Effects)

A natural question: what is wrong with running `lm(savings ~ disclosure, data = df)` or `lmer(savings ~ disclosure + (1 | household), data = df)` without centering? Both produce a single coefficient. Both are easy to fit. Neither requires any centering decisions.

The answer: **both conflate two mechanisms that may differ in sign and magnitude, making the coefficient uninterpretable for any specific question.**

| Model | What the coefficient estimates | When it goes wrong |
|---|---|---|
| Pooled OLS | Blend of within- and between-person effects | When between-person sorting confounds the within-person relationship (almost always in panel data) |
| Uncentered random-effects | Same blended coefficient | When the RE assumption (random effects ⊥ predictor) fails — which it usually does when the predictor varies between persons |
| CMC decomposition | $\beta_W$ (within) and $\beta_B$ (between) separately | It doesn't "go wrong" — it reveals both effects and makes the question explicit |

**The key question to ask before fitting any model with panel data:** *Does the predictor vary both within and between units?* If yes — almost always the case for continuous predictors like disclosure engagement, app usage, or spending — pooled OLS and uncentered RE give you a blended estimate that is wrong for both the causal within-unit question and the descriptive between-unit question.

Centering is not a technicality. It is the only way to know which question your model is answering.
:::

---

## Three Archetypal Patterns

The most important practical question is: *how different are $\beta_W$ and $\beta_B$?*

### Pattern 1: Convergent Effects — Budgeting App Engagement and Savings Rate

Both mechanisms point in the same direction with similar magnitude. On weeks when a customer opens the budgeting app more than their own typical frequency (within-person), they save more that week — the reflection and goal-salience triggered by app engagement directly improves saving behavior. And across customers, those who engage with the app more on average (between-person) also save more overall — they are both more financially motivated and get more cumulative benefit from the tool. True: $\beta_W \approx \beta_B \approx 1.5$.

```{r app-savings}
n_ppl <- 60; n_weeks <- 10

person_app_avg  <- rnorm(n_ppl, 5, 1.5)    # typical weekly app opens
person_sav_base <- 20 + 2 * person_app_avg + rnorm(n_ppl, 0, 4)  # motivated users save more

df_app <- expand_grid(person = 1:n_ppl, week = 1:n_weeks) |>
  mutate(
    app_opens = pmax(0, person_app_avg[person] + rnorm(n(), 0, 2)),
    savings   = person_sav_base[person] +
                1.5 * (app_opens - person_app_avg[person]) +
                rnorm(n(), 0, 4),
    person    = factor(person)
  ) |>
  group_by(person) |>
  mutate(app_mean = mean(app_opens), app_cmc = app_opens - app_mean) |>
  ungroup()

m_app <- lmer(savings ~ app_cmc + app_mean + (1 | person), data = df_app)
broom.mixed::tidy(m_app, effects = "fixed") |> filter(term != "(Intercept)") |>
  mutate(mechanism = c("Within-person (app_cmc)", "Between-person (app_mean)")) |>
  select(mechanism, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Pattern 1: within ≈ between ≈ 1.5. One coherent mechanism at two levels — pooled OLS approximates well here, but you only know that after checking.")
```

Here, the pooled model gives a reasonable approximation — but you only know that *after* running the decomposition and finding convergence. The convergence itself is substantively informative: the app appears to operate through the same mechanism at both timescales (goal activation and financial awareness).

### Pattern 2: Divergent Effects — Weekly Transactions and Monthly Savings Rate

A common misconception in financial data: customers who transact more frequently appear to save more (between-customer relationship). But within a customer, weeks when they make more transactions than usual correspond to *lower* savings that month — more spending activity drains current savings. The between-person relationship is positive because financially engaged customers are both more active AND more organized about saving. The within-person relationship is negative because extra transactions this week represent extra spending. True: $\beta_W \approx -0.8$ (within: more transactions → less saved), $\beta_B \approx +2.0$ (between: active customers save more).

```{r transactions-savings}
n_ppl2 <- 80; n_months <- 8

person_tx_avg   <- rnorm(n_ppl2, 20, 6)              # typical monthly transactions
person_sav_base <- 15 + 2.0 * person_tx_avg + rnorm(n_ppl2, 0, 5)  # engaged customers save more

df_tx <- expand_grid(person = 1:n_ppl2, month = 1:n_months) |>
  mutate(
    transactions = pmax(2, person_tx_avg[person] + rnorm(n(), 0, 5)),
    savings_rate = person_sav_base[person] +
                   (-0.8) * (transactions - person_tx_avg[person]) +   # more spending → less saving
                   rnorm(n(), 0, 4),
    person       = factor(person)
  ) |>
  group_by(person) |>
  mutate(tx_mean = mean(transactions), tx_cmc = transactions - tx_mean) |>
  ungroup()

m_tx     <- lmer(savings_rate ~ tx_cmc + tx_mean + (1 | person), data = df_tx)
m_tx_pool <- lmer(savings_rate ~ transactions + (1 | person), data = df_tx)

bind_rows(
  broom.mixed::tidy(m_tx_pool, effects = "fixed") |>
    filter(term == "transactions") |> mutate(model = "Pooled (blended)"),
  broom.mixed::tidy(m_tx, effects = "fixed") |>
    filter(term == "tx_cmc") |> mutate(model = "Within-person"),
  broom.mixed::tidy(m_tx, effects = "fixed") |>
    filter(term == "tx_mean") |> mutate(model = "Between-person")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Pattern 2: pooled ≈ +0.4 (near zero or positive); within ≈ −0.8 (spending crowds out saving); between ≈ +2.0 (engaged customers save more). Pooled is misleading for both questions.")
```

The pooled model gives a near-zero or weakly positive estimate — a blend of the negative within-customer spending-crowding effect and the positive between-customer engagement effect. A product team concluding "more transactions → more savings" (because the pooled slope is positive) would be wrong: within customers, more spending activity in a given month *reduces* that month's savings rate.

### Pattern 3: Sign Reversal — Price Level and Purchase Rate Across Market Segments

Within a market segment, promotional price reductions (prices below the segment's typical level) increase purchase rates — basic economics. But across segments, premium segments charge *higher* prices AND have *lower* purchase rates — because premium products are niche, not because high prices cause purchases. A pooled analysis finds that higher prices predict lower purchase rates. This is true between segments but **wrong as a within-segment statement**: within every segment, the price-demand relationship is negative (lower relative price → more purchases). True: $\beta_W \approx -0.4$ (demand elasticity), $\beta_B \approx -0.6$ (between-segment selection).

Wait — to get a *sign reversal* we need the between-segment effect to run opposite to within-segment. Here is the JDM version: **Discount depth and category conversion rates** — high-discount product categories (commodity goods: toilet paper, cleaning supplies) have both higher average discount depth *and* lower conversion rates (many competing substitutes). Low-discount categories (specialty foods, premium cosmetics) have lower discount depth *and* higher conversion rates (loyal buyer base). Pooled: deeper average discounts → lower conversion. Within category: deeper-than-average discounts → higher conversion. Sign reversal.

True: $\beta_W \approx +0.3$ (within-category: promotion → conversion up), $\beta_B \approx -0.5$ (between-category: commodity categories have high discounts but low loyalty).

```{r discount-conversion}
n_brands     <- 200
n_categories <- 5

# Commodity categories: high discount depth, low conversion (many substitutes)
# Premium categories: low discount depth, high conversion (loyal customers)
cat_discount_avg  <- c(22, 18, 14, 10, 6)    # avg discount depth (%) — high → low commodity→premium
cat_loyalty_base  <- c(18, 25, 35, 45, 52)   # baseline conversion rate (%)

df_discount <- tibble(
  brand    = 1:n_brands,
  category = rep(1:n_categories, each = n_brands / n_categories)
) |>
  mutate(
    discount   = cat_discount_avg[category] + rnorm(n_brands, 0, 4),
    conversion = cat_loyalty_base[category] +
                 0.3 * (discount - cat_discount_avg[category]) +   # within-cat: promotions help
                 rnorm(n_brands, 0, 5),
    category   = factor(category,
                        labels = c("Commodity", "Staple", "Mixed", "Specialty", "Premium"))
  ) |>
  group_by(category) |>
  mutate(disc_mean = mean(discount), disc_cmc = discount - disc_mean) |>
  ungroup()

m_disc_decomp <- lmer(conversion ~ disc_cmc + disc_mean + (1 | category), data = df_discount)
m_disc_pooled <- lm(conversion ~ discount, data = df_discount)

bind_rows(
  broom::tidy(m_disc_pooled) |>
    filter(term == "discount") |> mutate(model = "Pooled OLS"),
  broom.mixed::tidy(m_disc_decomp, effects = "fixed") |>
    filter(term == "disc_cmc") |> mutate(model = "Within-category"),
  broom.mixed::tidy(m_disc_decomp, effects = "fixed") |>
    filter(term == "disc_mean") |> mutate(model = "Between-category")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Pattern 3: pooled is negative; within is positive; between is strongly negative. Sign reversal!")
```

```{r discount-conversion-plot}
p1 <- ggplot(df_discount, aes(x = discount, y = conversion)) +
  geom_point(aes(color = category), alpha = 0.4, size = 1.2) +
  geom_smooth(method = "lm", color = "black", linewidth = 1.5) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Pooled: deeper discounts → LOWER conversion?",
       subtitle = "Between-category confound reverses the sign",
       x = "Discount depth (%)", y = "Conversion rate (%)", color = "Category") +
  theme_minimal(base_size = 12)

p2 <- ggplot(df_discount, aes(x = discount, y = conversion, color = category)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Within-category: deeper discounts → HIGHER conversion",
       subtitle = "Within every category, promotions work",
       x = "Discount depth (%)", y = "Conversion rate (%)", color = "Category") +
  theme_minimal(base_size = 12)

p1 + p2
```

The pooled model says promotions hurt conversion — which is wrong. The decomposition reveals the truth: within every category, discounts increase purchase rates; but commodity categories (with high average discounts) also have low inherent loyalty. Ignoring the nesting produces a **Simpson's Paradox** (see Part 3).

---

## The Disclosure–Savings Example: Applying Decomposition

50 households complete a monthly financial diary for 7 months. Each month, we record how many pages of the financial disclosure document they reviewed and their savings rate. Two questions are entangled:

1. **Within-household:** On months when a household reads *more than their own usual amount*, do they save more?
2. **Between-household:** Do households that engage more with disclosure *overall* have higher savings?

The between-household relationship is confounded by financial literacy: literate households both engage more with disclosure AND save more regardless of engagement. The within-household relationship is the one that matters for the causal question about disclosure.

```{r centering-demo}
n_hh   <- 50; n_months <- 7

hh_disc_avg    <- rnorm(n_hh, 6, 2)                   # household average disclosure
hh_savings_base <- 20 + 2 * hh_disc_avg + rnorm(n_hh, 0, 5)  # literacy confound

df_center <- expand_grid(household = 1:n_hh, month = 1:n_months) |>
  mutate(
    disc_raw = hh_disc_avg[household] + rnorm(n(), 0, 1.5),
    savings  = hh_savings_base[household] +
               1.5 * (disc_raw - hh_disc_avg[household]) + rnorm(n(), 0, 4),
    household = factor(household)
  ) |>
  group_by(household) |>
  mutate(disc_mean = mean(disc_raw), disc_cmc = disc_raw - disc_mean) |>
  ungroup() |>
  mutate(disc_gmc = disc_raw - mean(disc_raw))

m_uncentered <- lmer(savings ~ disc_raw  + (1 | household), data = df_center)
m_gmc        <- lmer(savings ~ disc_gmc  + (1 | household), data = df_center)
m_cmc        <- lmer(savings ~ disc_cmc + disc_mean + (1 | household), data = df_center)

bind_rows(
  broom.mixed::tidy(m_uncentered, effects = "fixed") |>
    filter(term == "disc_raw") |>
    mutate(model = "Uncentered", mechanism = "Blended"),
  broom.mixed::tidy(m_gmc, effects = "fixed") |>
    filter(term == "disc_gmc") |>
    mutate(model = "GMC", mechanism = "Blended"),
  broom.mixed::tidy(m_cmc, effects = "fixed") |>
    filter(term == "disc_cmc") |>
    mutate(model = "CMC", mechanism = "Within-household"),
  broom.mixed::tidy(m_cmc, effects = "fixed") |>
    filter(term == "disc_mean") |>
    mutate(model = "CMC", mechanism = "Between-household")
) |>
  select(model, mechanism, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Disclosure → savings: blended ≈ 3.5; within ≈ 1.5; between ≈ 4.0. Blended estimate is wrong for both questions.")
```

The uncentered and GMC models give ~3.5 — overstating the within-household effect by 2× because financial literacy inflates the between-household correlation. CMC reveals: the within-household causal effect (1.5 points per additional page) is meaningfully smaller than the between-household correlation (4.0 points) driven by literacy. A policy that relies on the blended estimate will project benefits of disclosure programmes that are unrealistically large.

---

## Decision Guide

::: {.callout-tip}
## Which Centering to Use

| Goal | Centering | What you get |
|------|-----------|--------------|
| Interpret the intercept at a meaningful value | GMC | Intercept = E[Y] when X = grand mean |
| Estimate the within-unit effect | CMC | $\hat\beta_W$ — deviation from own mean |
| Separate within and between mechanisms | CMC + unit mean | $\hat\beta_W$ and $\hat\beta_B$ |
| Test if within = between | CMC + unit mean | Wald test on equality |
| Single-level data (no nesting) | Either | No within/between distinction |

**Default recommendation for nested data:** Always use CMC with the unit mean included. Run it, check whether $\hat\beta_W \approx \hat\beta_B$, and only then decide whether the pooled model is defensible.
:::

::: {.callout-warning icon=false}
## Assumption: No Conflation of Within and Between Effects

Including a predictor without CMC in a random-effects model implicitly assumes $\beta_W = \beta_B$. When this is wrong — as in Patterns 2 and 3 above — both estimates are biased.

**How to test:** Fit the CMC model. Compare $\hat\beta_W$ and $\hat\beta_B$. A Wald test (see Part 2) tests the null $H_0: \beta_W = \beta_B$.

**The stakes:** In the worst case (Pattern 3), conflation completely reverses the direction of the effect.
:::

::: {.callout-tip icon=false}
## Reviewer Challenge: Three Questions About Your Centering Choice

**Challenge 1 — "Which question does your coefficient actually answer?"**
If you report a single coefficient for disclosure engagement without centering, a reviewer will ask: "Is this the within-household effect (what happens when a household reads more than usual) or the between-household effect (whether households with generally higher engagement save more)? Because those are different numbers with different policy implications." Run the CMC decomposition and report both — or justify why the blended estimate is the correct quantity of interest.

**Challenge 2 — "Why are the within- and between-household estimates different?"**
If $\hat\beta_W \neq \hat\beta_B$, reviewers will ask you to explain the mechanism. For the disclosure study: the between-household effect is inflated by financial literacy — literate households engage more AND save more regardless. The within-household effect is the pure causal question. Have a substantive explanation for the discrepancy, not just a statistical note that "random effects may be correlated with the predictor."

**Challenge 3 — "Does the sign reversal survive other robustness checks?"**
For the discount/conversion sign reversal example: a reviewer will ask whether the paradox holds when you control for other category-level characteristics (product shelf life, brand loyalty, advertising spend). If the within-category effect becomes near-zero once you add category-level controls, the pattern may reflect omitted variable bias rather than a genuine sign reversal. Run the model with additional controls at both levels.
:::

