---
title: "Survival Analysis"
---

```{r setup, include=FALSE}
library(tidyverse)
library(survival)
library(survminer)
set.seed(2024)
```

# Survival Analysis

## When Time-to-Event is the Outcome

Not all outcomes are measured on a continuous scale or as a snapshot at one moment. Some outcomes are events: the moment an employee quits, the day a patient relapses, the first purchase after acquiring a new customer. The outcome of interest is **when** the event occurs — or whether it has occurred at all by the time observation ends.

This is the domain of **survival analysis** (also called event history analysis or time-to-event analysis).

The defining feature: **censoring**. Some participants have not experienced the event by the end of the study — they were still employed, still in remission, still active users when the observation window closed. Censored observations contribute partial information: we know the event had not occurred up to that point, but we do not know when (or whether) it eventually will.

Standard regression discards this information. Survival analysis uses it.

## The Hazard Function

The central concept is the **hazard function** $h(t)$: the instantaneous rate at which events occur at time $t$, given that the event has not yet occurred by $t$.

$$h(t) = \lim_{\Delta t \to 0} \frac{P(t \le T < t + \Delta t \mid T \ge t)}{\Delta t}$$

Intuitively: if you have survived until time $t$, what is your risk of the event occurring in the next small interval?

## Kaplan-Meier Curves

The Kaplan-Meier estimator non-parametrically estimates the survival function $S(t) = P(T > t)$: the probability of surviving (not experiencing the event) past time $t$.

```{r km-curve}
# Simulate employee tenure data
n <- 200
group <- rep(c("High autonomy", "Low autonomy"), each = 100)

# High-autonomy employees stay longer
tenure_time  <- c(rexp(100, rate = 0.05),   # ~20 months average
                   rexp(100, rate = 0.10))   # ~10 months average
censored     <- rbinom(200, 1, 0.3)  # 30% are censored (still employed)
tenure_obs   <- pmin(tenure_time, 24)  # cap at 24 months observation window
status       <- ifelse(censored == 1, 0, 1)   # 0 = censored, 1 = event (quit)

df_surv <- tibble(time = tenure_obs, status, group)

# Kaplan-Meier
km_fit <- survfit(Surv(time, status) ~ group, data = df_surv)

ggsurvplot(
  km_fit,
  data        = df_surv,
  conf.int    = TRUE,
  pval        = TRUE,
  risk.table  = TRUE,
  palette     = c("#1e3a5f", "#aec6e8"),
  legend.labs = c("High autonomy", "Low autonomy"),
  xlab        = "Months",
  ylab        = "Probability of staying",
  title       = "Kaplan-Meier: employee tenure by autonomy condition"
)
```

## The Cox Proportional Hazards Model

The Cox model estimates the effect of predictors on the hazard rate without specifying the baseline hazard function — a semiparametric approach:

$$h(t \mid X) = h_0(t) \cdot \exp(\beta_1 X_1 + \beta_2 X_2 + \ldots)$$

Coefficients are interpreted as **hazard ratios** after exponentiation: $\exp(\beta) > 1$ means higher hazard (shorter survival); $\exp(\beta) < 1$ means lower hazard (longer survival).

```{r cox-model}
# Add predictors
df_surv <- df_surv |>
  mutate(
    autonomy     = ifelse(group == "High autonomy", 1, 0),
    age          = rnorm(200, 35, 8),
    satisfaction = rnorm(200, 60, 10)
  )

m_cox <- coxph(Surv(time, status) ~ autonomy + age + satisfaction,
               data = df_surv)

broom::tidy(m_cox, exponentiate = TRUE) |>
  knitr::kable(digits = 3,
    caption = "Cox hazard ratios: autonomy reduces hazard of quitting")
```

::: {.callout-warning icon=false}
## Assumption: Proportional Hazards

**What it means:** The hazard ratio between groups is constant over time. The effect of autonomy on the hazard of quitting is the same in month 1 as in month 20.

**Why it fails:** Some effects are strongest early (a drug that prevents relapse mainly in the first few weeks) and fade over time. This is a non-proportional hazard.

**How to check:** Schoenfeld residuals should be uncorrelated with time. The `cox.zph()` function tests this.

```{r ph-test}
ph_test <- cox.zph(m_cox)
print(ph_test)

# If significant: stratify by the variable or include time interaction
```
:::

## Frailty Models for Clustered Survival Data

When survival times are clustered (patients within hospitals, employees within firms), standard Cox models underestimate standard errors. The **frailty model** adds a random effect (the "frailty") to the Cox model — a direct analog of the random intercept in linear MLM.

```{r frailty-model}
# Simulate clustered survival data
n_firms    <- 20
n_per_firm <- 10

firm_effects <- rnorm(n_firms, 0, 0.5)

df_cluster <- tibble(
  firm     = rep(1:n_firms, each = n_per_firm),
  autonomy = rbinom(n_firms * n_per_firm, 1, 0.5),
  firm_eff = rep(firm_effects, each = n_per_firm),
  time_raw = rexp(n_firms * n_per_firm,
                   rate = exp(-1.5 - 0.4 * autonomy + firm_eff)),
  status   = rbinom(n_firms * n_per_firm, 1, 0.7),
  time     = pmin(time_raw, 24)
)

# Cox model with gamma frailty
m_frailty <- coxph(Surv(time, status) ~ autonomy + frailty(firm),
                   data = df_cluster)

broom::tidy(m_frailty) |>
  filter(term == "autonomy") |>
  knitr::kable(digits = 3,
    caption = "Frailty model: firm clustering accounted for in survival analysis")
```

::: {.callout-warning icon=false}
## Assumption: Non-Informative Censoring

**What it means:** The reason an observation is censored must be unrelated to the outcome. If sicker patients are more likely to drop out of a study (i.e., censoring is related to the hazard of relapse), standard survival estimates are biased.

**How to check:** Compare characteristics of censored and non-censored participants at their censoring time. Large differences suggest informative censoring.
:::
