---
title: "Binary Outcomes (Logistic MLM)"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(ggeffects)
library(broom.mixed)
set.seed(2024)
```

# Binary Outcomes: Logistic Mixed Models

## When Your Outcome is 0 or 1

Suppose you are studying whether a price promotion increases purchase probability, using transaction data from 5,000 customers across 50 stores. Your outcome is binary: purchased (1) or not (0). Customers are nested within stores.

A standard logistic regression ignores the store clustering. A logistic mixed model accounts for it — allowing stores to differ in their baseline purchase rates, and estimating the promotion effect while accounting for this.

## The Model

The logistic mixed model extends the standard logistic regression by adding a random intercept (and optionally a random slope):

$$\text{logit}(P(Y_{ij} = 1)) = \beta_0 + \beta_1 X_{ij} + u_{0j}$$

Where $u_{0j} \sim \mathcal{N}(0, \sigma^2_u)$ is the store-level random intercept.

```{r fit-logistic-mlm}
n_stores    <- 50
n_customers <- 100  # per store

store_effects <- rnorm(n_stores, 0, 1)  # store baseline differences

df_purchase <- expand_grid(
  store    = 1:n_stores,
  customer = 1:n_customers
) |>
  mutate(
    promotion   = rbinom(n(), 1, 0.5),
    store_eff   = store_effects[store],
    log_odds    = -1.5 + 0.8 * promotion + store_eff,
    purchased   = rbinom(n(), 1, plogis(log_odds)),
    store       = factor(store)
  )

# Logistic mixed model
m_logit <- glmer(purchased ~ promotion + (1 | store),
                 family  = binomial,
                 data    = df_purchase)

summary(m_logit)
```

```{r logit-fe}
broom.mixed::tidy(m_logit, effects = "fixed", exponentiate = TRUE) |>
  knitr::kable(digits = 3,
    caption = "Fixed effects as odds ratios: promotion increases purchase odds")
```

The promotion coefficient (as an odds ratio) tells us: customers who receive a promotion have `exp(0.8)` ≈ 2.2 times the odds of purchasing compared to those who don't, *after accounting for store-level differences*.

## The Marginal vs. Conditional Effects Problem

This is the single most important subtlety in logistic MLMs, and it is often missed.

**Conditional effect:** The effect of promotion *for a specific store at its mean random effect*. This is what the model directly estimates.

**Marginal effect:** The effect of promotion *averaged over the distribution of store effects* — the population-average effect. This is usually what you want to report.

In linear models, marginal and conditional effects are equal. In logistic models, they are not. The marginal effect is always attenuated relative to the conditional effect — the more the random effects vary, the larger the difference.

```{r marginal-effects}
# Conditional predictions
pred_conditional <- ggpredict(m_logit, terms = "promotion",
                               type = "random")   # conditional on random effects = 0

# Marginal predictions (averaged over store distribution)
pred_marginal <- ggpredict(m_logit, terms = "promotion",
                            type = "fixed")   # marginal (fixed effects only)

bind_rows(
  as.data.frame(pred_conditional) |> mutate(type = "Conditional (store RE = 0)"),
  as.data.frame(pred_marginal)    |> mutate(type = "Marginal (averaged over stores)")
) |>
  select(type, x, predicted, conf.low, conf.high) |>
  knitr::kable(digits = 3,
    caption = "Conditional vs. marginal predicted purchase probability")
```

For policy and generalization purposes, the marginal effect is usually more relevant. For understanding the mechanism within a specific store, the conditional effect is more relevant.

## Assumptions

::: {.callout-warning icon=false}
## Assumption 1: Correct Link Function

**What it means:** The logistic model assumes the log-odds of the outcome is linear in the predictors. This is the standard assumption for binary outcomes, but it may not hold at the extremes.

**How to check:** Examine whether predicted probabilities are reasonable at extreme predictor values. Consider complementary log-log (cloglog) as an alternative if your outcome is rare (P < 0.05).
:::

::: {.callout-warning icon=false}
## Assumption 2: Normality of Random Effects (on the Log-Odds Scale)

**What it means:** The store-level random intercepts are assumed normal on the log-odds scale. If store effects are heavily skewed (e.g., a few stores have dramatically different purchase rates), this assumption may fail.

**How to check:** Extract random effects with `ranef()` and examine their distribution. A Q-Q plot of the random intercepts is informative.

```{r re-normality}
re_vals <- ranef(m_logit)$store[, 1]
qqnorm(re_vals, main = "Q-Q plot of store random effects")
qqline(re_vals, col = "#e63946", lwd = 2)
```
:::

::: {.callout-warning icon=false}
## Assumption 3: No Overdispersion

**What it means:** For binary outcomes, overdispersion (more variance than the model predicts) can occur if there is residual clustering not accounted for by the random effects.

**How to check:** Compute the ratio of the Pearson chi-square statistic to its degrees of freedom. Values substantially greater than 1 indicate overdispersion.
:::
