---
title: "Count Outcomes (Poisson MLM)"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(glmmTMB)
library(broom.mixed)
set.seed(99)
```

# Count Outcomes: Poisson and Negative Binomial MLMs

## When Your Outcome is a Count

Count outcomes arise frequently in psychology and social science: the number of aggressive acts in a session, the number of days absent from work, the number of times a consumer repurchases, the number of words recalled.

Counts have two properties that make standard linear regression inappropriate:
1. They are non-negative integers (you cannot have −3 absences)
2. Their variance often increases with their mean (heteroskedasticity is built-in)

The **Poisson distribution** is the natural model for counts. The **Poisson mixed model** extends it to nested data.

## The Poisson Mixed Model

$$\log(\mu_{ij}) = \beta_0 + \beta_1 X_{ij} + u_{0j}, \quad Y_{ij} \sim \text{Poisson}(\mu_{ij})$$

The log link ensures predictions are always positive. Coefficients are interpreted as log rate ratios, or more intuitively as rate ratios after exponentiation.

```{r fit-poisson}
n_participants <- 60
n_days         <- 14

participant_effects <- rnorm(n_participants, 0, 0.5)

df_count <- expand_grid(person = 1:n_participants, day = 1:n_days) |>
  mutate(
    treatment = rep(rbinom(n_participants, 1, 0.5), each = n_days),
    p_eff     = rep(participant_effects, each = n_days),
    log_mu    = 1.5 - 0.5 * treatment + p_eff,  # true rate ratio = exp(-0.5) ≈ 0.61
    count     = rpois(n(), lambda = exp(log_mu)),
    person    = factor(person)
  )

m_pois <- glmer(count ~ treatment + (1 | person),
                family = poisson,
                data   = df_count)

broom.mixed::tidy(m_pois, effects = "fixed", exponentiate = TRUE) |>
  knitr::kable(digits = 3,
    caption = "Rate ratios: treatment reduces count outcome to 61% of control")
```

## Overdispersion: When Poisson is Not Enough

The Poisson distribution has a strict constraint: **mean = variance**. Real count data almost always has variance *exceeding* the mean — **overdispersion**. Sources of overdispersion include unobserved heterogeneity, clustering, and zero-inflation.

```{r check-overdispersion}
# Check for overdispersion
pearson_residuals <- residuals(m_pois, type = "pearson")
dispersion_stat   <- sum(pearson_residuals^2) / df.residual(m_pois)
cat("Dispersion statistic:", round(dispersion_stat, 3),
    "(should be ~1 for Poisson; > 1.5 indicates overdispersion)\n")
```

When overdispersion is present, use the **negative binomial** model, which adds a free dispersion parameter:

```{r negative-binomial}
# Simulate overdispersed count data
df_count_od <- df_count |>
  mutate(count_od = rnbinom(n(), mu = exp(log_mu), size = 2))  # overdispersed

m_nb <- glmmTMB(count_od ~ treatment + (1 | person),
                family = nbinom2,
                data   = df_count_od)

broom.mixed::tidy(m_nb, effects = "fixed", exponentiate = TRUE) |>
  knitr::kable(digits = 3,
    caption = "Negative binomial rate ratios (handles overdispersion)")
```

## Zero-Inflation

Some count outcomes have more zeros than any Poisson or negative binomial model can account for. This is common when zeros arise from two different processes: (1) the structural zero — a participant who never engages in the behavior regardless of conditions, and (2) the sampling zero — a participant who could engage but happened not to on this occasion.

```{r zero-inflation}
# Zero-inflated count
df_count_zi <- df_count |>
  mutate(
    # Some people are "structural zeros" (never engage)
    structural_zero = rbinom(n(), 1, 0.3),
    count_zi = ifelse(structural_zero == 1, 0, count)
  )

cat("Proportion of zeros (standard Poisson):", mean(df_count$count == 0), "\n")
cat("Proportion of zeros (zero-inflated):", mean(df_count_zi$count_zi == 0), "\n")

m_zi <- glmmTMB(count_zi ~ treatment + (1 | person),
                ziformula = ~1,   # intercept-only model for zero-inflation
                family    = poisson,
                data      = df_count_zi)

broom.mixed::tidy(m_zi, effects = "fixed", component = "cond", exponentiate = TRUE) |>
  knitr::kable(digits = 3,
    caption = "Zero-inflated Poisson: count model conditional on non-structural-zero")
```

## Assumptions

::: {.callout-warning icon=false}
## Assumption 1: Correct Distribution Family

| Distribution | When to use | Diagnosis |
|-------------|-------------|-----------|
| Poisson | Mean ≈ Variance; no excess zeros | Dispersion stat ≈ 1 |
| Negative binomial | Variance > Mean (overdispersion) | Dispersion stat > 1.5 |
| Zero-inflated Poisson | Poisson + excess zeros | More zeros than Poisson predicts |
| Zero-inflated NB | Overdispersion + excess zeros | Both overdispersion and excess zeros |
:::

::: {.callout-warning icon=false}
## Assumption 2: Log Link Linearity

**What it means:** The log of the expected count is a linear function of the predictors. This is usually reasonable but should be checked for continuous predictors.

**How to check:** Plot smoothed log-count against the continuous predictor to look for nonlinearity.
:::
