---
title: "Endogeneity: Why Regression Lies (and How to Catch It)"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(broom)
set.seed(1234)
```

# Endogeneity: Why Regression Lies (and How to Catch It)

## The Fundamental Problem of Causal Inference

Before introducing regression at all, consider the true object of scientific interest: the **causal effect** of X on Y for a specific unit $i$.

Under the **potential outcomes framework** (Rubin, 1974), every unit has two potential outcomes:

$$
\underbrace{\tau_i}_{\substack{\text{causal effect} \\ \text{for unit } i}} = \underbrace{Y_i(1)}_{\substack{\text{outcome if} \\ \text{treated}}} - \underbrace{Y_i(0)}_{\substack{\text{outcome if} \\ \text{not treated}}}
$$

The fundamental problem: **we can never observe both $Y_i(1)$ and $Y_i(0)$ for the same unit at the same time.** A person either received psychotherapy or they did not. A country either adopted a carbon tax or it did not. We observe one potential outcome and must infer the other.

Regression estimates the **average treatment effect** (ATE):

$$
\text{ATE} = E[\tau_i] = E[Y_i(1) - Y_i(0)]
$$

For regression to estimate the ATE correctly, the *observed* difference in outcomes between treated and untreated units must equal the *counterfactual* difference between what would have happened to those same units under both conditions. This requires that who gets treated is **independent** of the potential outcomes.

When this independence fails — when who gets X is related to Y through pathways other than X itself — the regression coefficient is **endogenous** and does not estimate the causal effect.

---

## What is Endogeneity?

**Endogeneity** means your predictor X is correlated with the error term $\varepsilon$ in your regression:

$$
Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i \quad \text{where } \text{Cov}(X_i, \varepsilon_i) \neq 0
$$

When this happens, $\hat{\beta}_1$ is biased. It estimates not just the effect of X on Y but a blend of that effect and all the other pathways through which X and Y are connected.

There are four main sources.

---

## Source 1: Omitted Variable Bias (OVB)

A variable Z influences both X and Y, but Z is not in your model. The path Z → Y gets attributed to X because X and Z are correlated.

### The Bias Formula

The expected value of the biased coefficient is:

$$
\underbrace{E[\hat{\beta}_1]}_{\substack{\text{what we} \\ \text{estimate}}} = \underbrace{\beta_1}_{\substack{\text{true} \\ \text{effect}}} + \underbrace{\delta_{ZY}}_{\substack{Z \text{'s effect} \\ \text{on } Y}} \cdot \underbrace{\tilde{\rho}_{XZ}}_{\substack{\text{correlation of} \\ X \text{ and } Z}}
$$

where $\delta_{ZY}$ is the effect of Z on Y (holding X constant) and $\tilde{\rho}_{XZ}$ is the correlation of Z with X after partialling out other covariates.

**The bias formula has a critical practical use:** even without data on Z, if you know the *sign* of both $\delta_{ZY}$ and $\tilde\rho_{XZ}$, you can predict the *direction* of the bias on your estimate.

```{r ovb-bias-table}
# Bias direction table
tibble(
  Example         = c("Budget tracker → savings", "Promotion → purchase", "Investment advice → returns"),
  Confounder_Z    = c("Platform quality", "Consumer wealth", "Financial sophistication"),
  `Sign Z→X`      = c("+", "+", "-"),
  `Sign Z→Y`      = c("+", "+", "+"),
  `Bias direction` = c("Positive (↑ est.)", "Positive (↑ est.)", "Negative (↓ est.)")
) |>
  knitr::kable(caption = "OVB bias direction: multiply signs to get bias direction")
```

::: {.callout-note}
## Reading the Bias Table

Sign(Z→X) × Sign(Z→Y) = sign of the bias:
- Both positive → positive bias (estimate is too large)
- Both negative → positive bias (two negatives multiply to positive)
- Mixed signs → negative bias (estimate is too small or reversed)

You can use this qualitatively even before collecting data: "If my estimate is already large, and the likely omitted confound would produce positive bias, my true effect is probably smaller than I think."
:::

### Three Demonstrations

```{r ovb-simulation}
n <- 500

# Example 1: Budget tracker / savings rate / platform quality
# Platform quality attracts more engaged users AND produces better financial outcomes
# True effect of using the tracker = 0; it is the platform quality that matters
Z_quality     <- rnorm(n, 50, 10)
X_tracker     <- 0.3 * Z_quality + rnorm(n, 0, 5)
Y_savings     <- 0.5 * Z_quality + rnorm(n, 0, 8)  # true X effect = 0

m_biased_1  <- lm(Y_savings ~ X_tracker)
m_correct_1 <- lm(Y_savings ~ X_tracker + Z_quality)

# Example 2: Premium promotional message exposure / purchase intention / consumer wealth
# Wealthy consumers are targeted more AND have higher purchase intentions for premium products
# True X effect = 0.15; wealth inflates this upward
Z_wealth      <- rnorm(n, 50, 10)
X_promo       <- 0.4 * Z_wealth + rnorm(n, 0, 8)
Y_intention   <- 0.6 * Z_wealth + 0.15 * X_promo + rnorm(n, 0, 10)

m_biased_2  <- lm(Y_intention ~ X_promo)
m_correct_2 <- lm(Y_intention ~ X_promo + Z_wealth)

# Example 3: Investment advice uptake / portfolio return / financial sophistication
# Sophisticated investors avoid advice (prefer self-directed) AND perform better
# True X effect = +0.5 (advice genuinely helps); sophistication biases estimate downward
Z_sophist     <- rnorm(n, 50, 10)
X_advice      <- -0.5 * Z_sophist + rnorm(n, 0, 6)   # sophisticated investors avoid advice
Y_returns     <- 0.4 * Z_sophist + 0.5 * X_advice + rnorm(n, 0, 8)

m_biased_3  <- lm(Y_returns ~ X_advice)
m_correct_3 <- lm(Y_returns ~ X_advice + Z_sophist)

bind_rows(
  tidy(m_biased_1)  |> filter(term == "X_tracker") |>
    mutate(example = "Budget tracker (true=0)",       model = "Omits Z"),
  tidy(m_correct_1) |> filter(term == "X_tracker") |>
    mutate(example = "Budget tracker (true=0)",       model = "Includes Z"),
  tidy(m_biased_2)  |> filter(term == "X_promo")   |>
    mutate(example = "Promotion (true=0.15)",          model = "Omits Z"),
  tidy(m_correct_2) |> filter(term == "X_promo")   |>
    mutate(example = "Promotion (true=0.15)",          model = "Includes Z"),
  tidy(m_biased_3)  |> filter(term == "X_advice")  |>
    mutate(example = "Investment advice (true=0.5)",   model = "Omits Z"),
  tidy(m_correct_3) |> filter(term == "X_advice")  |>
    mutate(example = "Investment advice (true=0.5)",   model = "Includes Z")
) |>
  select(example, model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "OVB across three examples: omitting Z biases estimates in predictable directions")
```

In every case, adding the confounder pulls the estimate toward the true value. The bias direction matches the prediction from the sign table.

The budget-tracking example illustrates a common error in consumer finance research: users of a platform feature (tracker) appear to save more, but the actual driver is platform quality — better platforms attract more financially engaged users who would have saved more regardless. The promotion example shows upward bias: promotions genuinely increase purchase intentions, but the naive estimate is inflated because wealth drives both targeting and intention. The investment advice example shows downward bias: advice genuinely improves portfolio returns, but financially sophisticated investors avoid advice while also outperforming — so the advice-returns correlation in cross-sectional data understates the true benefit.

::: {.callout-warning icon=false}
## Assumption: No Omitted Variables Correlated with Both X and Y

**How to check:** Draw a causal diagram. List every plausible cause of Y. Ask: is any of these also correlated with X? This requires substantive reasoning about the DGP — you cannot test it purely from data.

**In nested data:** OVB is especially insidious because the omitted variable often varies at a higher level. Studying whether neighborhood green space predicts depression while omitting neighborhood income biases your estimate; the omitted variable is a neighborhood-level characteristic that clustering creates.

**What to do:** Add the variable if measurable. Use fixed effects to absorb all stable unit-level confounders. Use IV, RDD, or DiD if the confound is unmeasurable (covered in Parts 2–6).
:::

---

## Source 2: Simultaneity / Reverse Causality

X causes Y, but Y also causes X. The causal arrow runs in both directions, creating a feedback loop that OLS cannot untangle.

```{r simultaneity-plot}
n_sim <- 300
set.seed(42)

# Social support → depression (negative), but depression → social withdrawal (negative)
# Simulate equilibrium: both processes are real
time_pts <- tibble(
  t    = 1:10,
  support = c(60, 55, 52, 50, 49, 48, 47, 47, 47, 47) + rnorm(10, 0, 2),
  depression = c(30, 35, 38, 40, 41, 42, 43, 43, 43, 43) + rnorm(10, 0, 2)
)

p_sim <- time_pts |>
  pivot_longer(-t) |>
  ggplot(aes(x = t, y = value, color = name, group = name)) +
  geom_line(linewidth = 1.2) + geom_point(size = 2.5) +
  scale_color_manual(values = c("#e63946", "#1e3a5f"),
                     labels = c("Depression", "Social Support")) +
  labs(title = "Simultaneity: support → depression AND depression → support",
       subtitle = "OLS on cross-sectional data picks up both arrows simultaneously",
       x = "Time", y = "Level", color = NULL) +
  theme_minimal(base_size = 13)

p_sim
```

**Psychology example:** Social support reduces depression — but depression also causes people to withdraw from social networks. A cross-sectional regression of depression on social support picks up both directions simultaneously and gives a confounded estimate.

The same dynamic is common in consumer markets: advertising drives sales (brand awareness → purchase), but sales revenue funds advertising (revenue → ad budget). A cross-sectional regression of sales on advertising spending cannot identify whether awareness causes purchases or revenue enables advertising. Longitudinal designs with lagged predictors are required.

::: {.callout-warning icon=false}
## Assumption: No Reverse Causality

**How to check:** Ask: could Y at time $t$ plausibly influence X at time $t$ or $t+1$? If yes, simultaneity is a concern.

**What to do:**
- Use **lagged predictors** ($X_{t-1}$ predicting $Y_t$) in panel data
- Use **instrumental variables** (Part 2) to isolate the causal direction
- Use **cross-lagged panel models** (Module 4, Part 6) which model both causal arrows explicitly
:::

---

## Source 3: Selection Effects

The observations in your dataset are not a random sample of the population. Who ends up in your data — and which treatment they receive — is systematically related to both X and Y.

**Self-selection:** People who choose to undergo therapy are systematically different from those who don't — they may be more motivated to change, have better social support, or have less severe disorders. Comparing therapy-goers to non-therapy-goers conflates the effect of therapy with all these pre-existing differences.

**Survivorship bias:** You only observe firms that survived long enough to be in your dataset. Failed firms are absent. Conclusions about firm strategy drawn from survivors are biased because you are conditioning on survival.

**Attrition in longitudinal studies:** Participants who drop out of a study are often systematically different from completers. If the sickest patients drop out of a clinical trial, the observed treatment effect among completers overstates the population-level effect.

This source of endogeneity is covered in depth in Part 3.

---

## Source 4: Collider Bias

This source is frequently overlooked but is just as dangerous as the others. **Conditioning on a collider** — a variable caused by *both* X and Y — creates a spurious association between X and Y even when none exists.

A **collider** is a variable C where two causal arrows "collide":

$$X \rightarrow C \leftarrow Y$$

When you condition on C (include it as a control, restrict your sample to a subset defined by C, or analyze a non-random sample selected on C), you open a spurious backdoor path between X and Y.

```{r collider-demo}
n_col <- 1000

# True DGP: X and Y are uncorrelated
X_col <- rnorm(n_col)
Y_col <- rnorm(n_col)  # truly independent of X

# Collider C is caused by both X and Y
C_col <- X_col + Y_col + rnorm(n_col, 0, 0.5)

# Full sample: no correlation
r_full <- cor(X_col, Y_col)

# Conditioning on C: restrict to top half of C
high_C <- C_col > median(C_col)
r_cond <- cor(X_col[high_C], Y_col[high_C])

tibble(
  Sample           = c("Full sample (no conditioning)", "Restricted to high-C (conditioning)"),
  `Cor(X, Y)`      = c(round(r_full, 3), round(r_cond, 3)),
  Interpretation   = c("True: X and Y are unrelated",
                        "Spurious: conditioning on C creates negative correlation")
) |>
  knitr::kable(caption = "Collider bias: conditioning on C creates a spurious X-Y association")
```

**The publish-or-perish example:** Prolificacy (X) and idea quality (Y) may be uncorrelated in the full population of researchers. But publication (C) is caused by both — prolific researchers with good ideas get published. If you study *only published papers* (conditioning on C), you'll observe that prolificacy and idea quality are *negatively correlated* within the published set, even though they're unrelated in the population.

```{r collider-viz}
df_col <- tibble(X = X_col, Y = Y_col, C_high = high_C)

p_full <- ggplot(df_col, aes(x = X, y = Y)) +
  geom_point(alpha = 0.15, color = "#1e3a5f") +
  geom_smooth(method = "lm", color = "#e63946", linewidth = 1.2) +
  labs(title = "Full sample", subtitle = paste0("r = ", round(r_full, 3)),
       x = "X", y = "Y") +
  theme_minimal(base_size = 12)

p_cond <- ggplot(df_col |> filter(C_high), aes(x = X, y = Y)) +
  geom_point(alpha = 0.2, color = "#2d6a4f") +
  geom_smooth(method = "lm", color = "#e63946", linewidth = 1.2) +
  labs(title = "Conditioned on high C", subtitle = paste0("r = ", round(r_cond, 3)),
       x = "X", y = "Y") +
  theme_minimal(base_size = 12)

p_full + p_cond
```

::: {.callout-tip icon=false}
## JDM Collider: Analyzing Only Platform Engagers

In a study of platform users, promotional sensitivity (X) and consumer wealth (Y) may be uncorrelated in the broader population. But both predict active platform engagement (C). If you analyze only high-engagement users — restricting your sample based on C — you introduce a spurious negative correlation between X and Y within your sample. Wealthier users who are also promo-sensitive are over-represented; the model sees this as a negative relationship between wealth and promo-sensitivity within the active user pool. This is collider bias from sample restriction, not from including C as a covariate.

The same logic applies any time you analyze a selected subset: only people who purchased, only participants who completed all waves, only firms that survived.
:::

::: {.callout-warning icon=false}
## Assumption: No Conditioning on Colliders

**How to check:** Draw a DAG. For every variable you are controlling for or selecting on, ask: is this caused by X, by Y, or by both? If caused by both, it is a collider — do not condition on it.

**Common mistake:** "Kitchen sink" regressions that add every available covariate. If any of those covariates is a collider, adding it *creates* bias where none existed.
:::

---

## Pearl's Causal Framework: Three Types of Paths

Judea Pearl's causal graph (DAG) framework provides a systematic way to identify endogeneity. Every causal graph has three fundamental path structures:

```{r dag-structures}
make_dag_panel <- function(nodes, edges_df, title, note = NULL) {
  p <- ggplot() +
    geom_point(data = nodes, aes(x = x, y = y), size = 18,
               shape = 21, fill = "#deebf7", color = "#1e3a5f", stroke = 1.5) +
    geom_text(data = nodes, aes(x = x, y = y, label = label), size = 3.8,
              fontface = "bold", color = "#1e3a5f")
  for (i in seq_len(nrow(edges_df))) {
    p <- p + annotate("segment",
                      x    = edges_df$x1[i], xend = edges_df$x2[i],
                      y    = edges_df$y1[i], yend = edges_df$y2[i],
                      arrow = arrow(length = unit(0.25, "cm")),
                      color = edges_df$color[i], linewidth = 1.2)
  }
  p + labs(title = title, subtitle = note) +
    xlim(0.3, 3.7) + ylim(0.4, 2.6) + theme_void(base_size = 11) +
    theme(plot.title = element_text(face = "bold", size = 11),
          plot.subtitle = element_text(size = 9, color = "gray40"))
}

nodes_fork <- tibble(label = c("Z", "X", "Y"), x = c(2, 1, 3), y = c(2, 1, 1))
edges_fork <- tibble(x1=c(1.8,2.2), x2=c(1.2,2.8), y1=c(1.9,1.9), y2=c(1.1,1.1),
                     color = c("#e63946","#e63946"))
p_fork <- make_dag_panel(nodes_fork, edges_fork, "Fork (Confounder)",
  "Z → X and Z → Y\nCreates spurious X-Y correlation\nExample: financial literacy (Z) → engagement\nwith disclosure format (X) and → risky choice (Y)\nFix: control for Z")

nodes_chain <- tibble(label = c("X", "M", "Y"), x = c(1, 2, 3), y = c(1, 1, 1))
edges_chain <- tibble(x1=c(1.2,2.2), x2=c(1.8,2.8), y1=c(1,1), y2=c(1,1),
                      color = c("#1e3a5f","#1e3a5f"))
p_chain <- make_dag_panel(nodes_chain, edges_chain, "Chain (Mediator)",
  "X → M → Y\nM is on the causal path\nExample: default enrollment (X) → savings habit (M)\n→ long-term savings (Y)\nFix: do NOT control for M")

nodes_coll <- tibble(label = c("X", "C", "Y"), x = c(1, 2, 3), y = c(1, 1, 1))
edges_coll <- tibble(x1=c(1.2,2.8), x2=c(1.8,2.2), y1=c(1,1), y2=c(1,1),
                     color = c("#1e3a5f","#1e3a5f"))
p_coll <- make_dag_panel(nodes_coll, edges_coll, "Collider",
  "X → C ← Y\nCondition on C = open backdoor\nExample: promotional sensitivity (X) →\nplatform activity (C) ← consumer wealth (Y)\nFix: do NOT control for C")

p_fork + p_chain + p_coll
```

| Path Type | Structure | What it means | What to do |
|-----------|-----------|---------------|------------|
| **Fork** | Z → X and Z → Y | Z confounds X-Y relationship | Block by controlling for Z |
| **Chain** | X → M → Y | M mediates X's effect on Y | Do NOT block (it carries the effect) |
| **Collider** | X → C ← Y | C is caused by both X and Y | Do NOT block (conditioning opens a backdoor) |

The goal of every identification strategy in the rest of this module is to **block all backdoor paths** (forks) without blocking causal paths (chains) and without conditioning on colliders.

---

## The Backdoor Criterion

A set of variables S satisfies Pearl's **backdoor criterion** for estimating the effect of X on Y if:

1. S blocks every path between X and Y that has an arrow *into* X (i.e., all backdoor paths)
2. S does not include any descendant of X on a causal path to Y (no colliders, no mediators on the X→Y path)

When you can find such a set S from measured variables, controlling for S gives you the causal effect. When you cannot — because the confounders are unmeasured — you need the identification strategies in Parts 2–6.

::: {.callout-important}
## Before Running Any Regression

Draw a causal diagram. Every arrow is a hypothesis about the DGP. Then ask:

1. Are there any **forks** (confounders) not in my model? → OVB
2. Could the outcome affect the predictor? → Simultaneity
3. Is my sample selected on something related to both X and Y? → Selection
4. Am I controlling for anything caused by X or Y? → Collider bias

Answering these four questions before looking at your results will save you from the most common errors in applied regression.
:::

::: {.callout-tip icon=false}
## Observational → Experimental Translation

**In observational nested data:**
The predictor X is almost never exogenous. Employees who participate in training programs are not randomly selected — they tend to be younger, more motivated, and employed at better-resourced firms. The OVB formula quantifies how much the naive estimate is inflated by this selection.

**In randomized experiments:**
Random assignment solves the endogeneity problem *for the assigned condition* — but not for everything downstream. Actual treatment receipt (compliance), session attendance, engagement level, and follow-up completion all reintroduce endogeneity after randomization. The residual endogeneity in experiments is the subject of Modules 5 Parts 8 and 9.

**Same diagnosis, different source:**
In both settings, the question is identical: *Is there a variable that simultaneously predicts X and Y, that I have not accounted for?* In observational data, this is about confounders in the sampling. In experiments, it is about post-randomization processes that reintroduce selection.
:::
