---
title: "Endogeneity: Why Regression Lies (and How to Catch It)"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(broom)
set.seed(1234)
```

# Endogeneity: Why Regression Lies (and How to Catch It)

## The Fundamental Problem of Causal Inference

Before introducing regression at all, consider the true object of scientific interest: the **causal effect** of X on Y for a specific unit $i$.

Under the **potential outcomes framework** (Rubin, 1974), every unit has two potential outcomes:

$$
\underbrace{\tau_i}_{\substack{\text{causal effect} \\ \text{for unit } i}} = \underbrace{Y_i(1)}_{\substack{\text{outcome if} \\ \text{treated}}} - \underbrace{Y_i(0)}_{\substack{\text{outcome if} \\ \text{not treated}}}
$$

The fundamental problem: **we can never observe both $Y_i(1)$ and $Y_i(0)$ for the same unit at the same time.** A household either received a financial coaching program or it did not. A country either adopted a mandatory disclosure policy or it did not. We observe one potential outcome and must infer the other.

Regression estimates the **average treatment effect** (ATE):

$$
\text{ATE} = E[\tau_i] = E[Y_i(1) - Y_i(0)]
$$

For regression to estimate the ATE correctly, the *observed* difference in outcomes between treated and untreated units must equal the *counterfactual* difference between what would have happened to those same units under both conditions. This requires that who gets treated is **independent** of the potential outcomes.

When this independence fails — when who gets X is related to Y through pathways other than X itself — the regression coefficient is **endogenous** and does not estimate the causal effect.

---

## What is Endogeneity?

**Endogeneity** means your predictor X is correlated with the error term $\varepsilon$ in your regression:

$$
Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i \quad \text{where } \text{Cov}(X_i, \varepsilon_i) \neq 0
$$

When this happens, $\hat{\beta}_1$ is biased. It estimates not just the effect of X on Y but a blend of that effect and all the other pathways through which X and Y are connected.

There are four main sources.

---

## Source 1: Omitted Variable Bias (OVB)

A variable Z influences both X and Y, but Z is not in your model. The path Z → Y gets attributed to X because X and Z are correlated.

### The Bias Formula

The expected value of the biased coefficient is:

$$
\underbrace{E[\hat{\beta}_1]}_{\substack{\text{what we} \\ \text{estimate}}} = \underbrace{\beta_1}_{\substack{\text{true} \\ \text{effect}}} + \underbrace{\delta_{ZY}}_{\substack{Z \text{'s effect} \\ \text{on } Y}} \cdot \underbrace{\tilde{\rho}_{XZ}}_{\substack{\text{correlation of} \\ X \text{ and } Z}}
$$

where $\delta_{ZY}$ is the effect of Z on Y (holding X constant) and $\tilde{\rho}_{XZ}$ is the correlation of Z with X after partialling out other covariates.

**The bias formula has a critical practical use:** even without data on Z, if you know the *sign* of both $\delta_{ZY}$ and $\tilde\rho_{XZ}$, you can predict the *direction* of the bias on your estimate.

```{r ovb-bias-table}
# Bias direction table
tibble(
  Example         = c("Budget tracker → savings", "Promotion → purchase", "Investment advice → returns"),
  Confounder_Z    = c("Platform quality", "Consumer wealth", "Financial sophistication"),
  `Sign Z→X`      = c("+", "+", "-"),
  `Sign Z→Y`      = c("+", "+", "+"),
  `Bias direction` = c("Positive (↑ est.)", "Positive (↑ est.)", "Negative (↓ est.)")
) |>
  knitr::kable(caption = "OVB bias direction: multiply signs to get bias direction")
```

::: {.callout-note}
## Reading the Bias Table

Sign(Z→X) × Sign(Z→Y) = sign of the bias:
- Both positive → positive bias (estimate is too large)
- Both negative → positive bias (two negatives multiply to positive)
- Mixed signs → negative bias (estimate is too small or reversed)

You can use this qualitatively even before collecting data: "If my estimate is already large, and the likely omitted confound would produce positive bias, my true effect is probably smaller than I think."
:::

### Same Observables, Two Stories

The same scatter of data points can support two completely different causal interpretations. The visualization below shows this directly: in panel (A), pooled data suggest that budget-tool use predicts higher savings. In panel (B), the same points are colored by financial literacy tertile — and within each literacy group, the relationship disappears. Financial literacy drives *both* budget-tool use and savings; the pooled slope is pure confounding.

When you have not measured the confounder — which is the typical case — you see only panel (A) and cannot make the comparison that panel (B) reveals.

```{r same-obs-two-stories, echo=FALSE, fig.width=10, fig.height=4.2, message=FALSE, warning=FALSE}
set.seed(99)
n_story <- 300

# Financial literacy (Z) drives both budget tool use (X) and savings (Y)
# True causal effect of X on Y = 0
Z_lit  <- rnorm(n_story, 50, 10)
X_tool <- 0.4 * Z_lit + rnorm(n_story, 0, 6)
Y_sav  <- 0.5 * Z_lit + rnorm(n_story, 0, 8)

df_story <- tibble(X_tool, Y_sav, Z_lit) |>
  mutate(Z_grp = cut(Z_lit,
                     breaks = quantile(Z_lit, probs = c(0, 1/3, 2/3, 1)),
                     labels = c("Low literacy", "Medium literacy", "High literacy"),
                     include.lowest = TRUE))

# Panel A: pooled — looks like tool causes savings
p_pooled <- ggplot(df_story, aes(x = X_tool, y = Y_sav)) +
  geom_point(alpha = 0.18, color = "#1e3a5f", size = 1.3) +
  geom_smooth(method = "lm", color = "#e63946", linewidth = 1.3, se = TRUE) +
  labs(
    title    = "(A) Pooled: budget tool predicts savings",
    subtitle = "Naive read: \u2018tool use causes higher savings\u2019\n(confounder not measured or modeled)",
    x = "Budget tool use (X)", y = "Monthly savings rate (Y)"
  ) +
  theme_minimal(base_size = 12)

# Panel B: colored by literacy — within-group slopes are flat
p_colored <- ggplot(df_story, aes(x = X_tool, y = Y_sav, color = Z_grp)) +
  geom_point(alpha = 0.38, size = 1.3) +
  geom_smooth(aes(group = Z_grp), method = "lm", se = FALSE, linewidth = 1.1) +
  scale_color_manual(
    values = c("Low literacy" = "#aec6e8",
               "Medium literacy" = "#4a90d9",
               "High literacy"   = "#1e3a5f"),
    name = "Literacy (Z)"
  ) +
  labs(
    title    = "(B) Within literacy groups: slope \u2248 0",
    subtitle = "Correct read: literacy (Z) drives both tool use (X) and savings (Y)\nTrue effect of tool use \u2248 0 within literacy groups",
    x = "Budget tool use (X)", y = "Monthly savings rate (Y)"
  ) +
  theme_minimal(base_size = 12)

p_pooled + p_colored
```

This is the OVB core: the observed X–Y correlation in the left panel is entirely explained by Z. When Z is unmeasured, the naive estimate is a biased blend of the (zero) true effect and the Z-driven confounding. The simulations below show this numerically across three financial services examples.

### Three Demonstrations

```{r ovb-simulation}
n <- 500

# Example 1: Budget tracker / savings rate / platform quality
# Platform quality attracts more engaged users AND produces better financial outcomes
# True effect of using the tracker = 0; it is the platform quality that matters
Z_quality     <- rnorm(n, 50, 10)
X_tracker     <- 0.3 * Z_quality + rnorm(n, 0, 5)
Y_savings     <- 0.5 * Z_quality + rnorm(n, 0, 8)  # true X effect = 0

m_biased_1  <- lm(Y_savings ~ X_tracker)
m_correct_1 <- lm(Y_savings ~ X_tracker + Z_quality)

# Example 2: Premium promotional message exposure / purchase intention / consumer wealth
# Wealthy consumers are targeted more AND have higher purchase intentions for premium products
# True X effect = 0.15; wealth inflates this upward
Z_wealth      <- rnorm(n, 50, 10)
X_promo       <- 0.4 * Z_wealth + rnorm(n, 0, 8)
Y_intention   <- 0.6 * Z_wealth + 0.15 * X_promo + rnorm(n, 0, 10)

m_biased_2  <- lm(Y_intention ~ X_promo)
m_correct_2 <- lm(Y_intention ~ X_promo + Z_wealth)

# Example 3: Investment advice uptake / portfolio return / financial sophistication
# Sophisticated investors avoid advice (prefer self-directed) AND perform better
# True X effect = +0.5 (advice genuinely helps); sophistication biases estimate downward
Z_sophist     <- rnorm(n, 50, 10)
X_advice      <- -0.5 * Z_sophist + rnorm(n, 0, 6)   # sophisticated investors avoid advice
Y_returns     <- 0.4 * Z_sophist + 0.5 * X_advice + rnorm(n, 0, 8)

m_biased_3  <- lm(Y_returns ~ X_advice)
m_correct_3 <- lm(Y_returns ~ X_advice + Z_sophist)

bind_rows(
  tidy(m_biased_1)  |> filter(term == "X_tracker") |>
    mutate(example = "Budget tracker (true=0)",       model = "Omits Z"),
  tidy(m_correct_1) |> filter(term == "X_tracker") |>
    mutate(example = "Budget tracker (true=0)",       model = "Includes Z"),
  tidy(m_biased_2)  |> filter(term == "X_promo")   |>
    mutate(example = "Promotion (true=0.15)",          model = "Omits Z"),
  tidy(m_correct_2) |> filter(term == "X_promo")   |>
    mutate(example = "Promotion (true=0.15)",          model = "Includes Z"),
  tidy(m_biased_3)  |> filter(term == "X_advice")  |>
    mutate(example = "Investment advice (true=0.5)",   model = "Omits Z"),
  tidy(m_correct_3) |> filter(term == "X_advice")  |>
    mutate(example = "Investment advice (true=0.5)",   model = "Includes Z")
) |>
  select(example, model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "OVB across three examples: omitting Z biases estimates in predictable directions")
```

In every case, adding the confounder pulls the estimate toward the true value. The bias direction matches the prediction from the sign table.

The budget-tracking example illustrates a common error in consumer finance research: users of a platform feature (tracker) appear to save more, but the actual driver is platform quality — better platforms attract more financially engaged users who would have saved more regardless. The promotion example shows upward bias: promotions genuinely increase purchase intentions, but the naive estimate is inflated because wealth drives both targeting and intention. The investment advice example shows downward bias: advice genuinely improves portfolio returns, but financially sophisticated investors avoid advice while also outperforming — so the advice-returns correlation in cross-sectional data understates the true benefit.

::: {.callout-warning icon=false}
## Assumption: No Omitted Variables Correlated with Both X and Y

**How to check:** Draw a causal diagram. List every plausible cause of Y. Ask: is any of these also correlated with X? This requires substantive reasoning about the DGP — you cannot test it purely from data.

**In nested data:** OVB is especially insidious because the omitted variable often varies at a higher level. Studying whether branch-level disclosure intensity predicts household savings while omitting branch-level financial literacy programming biases your estimate; the omitted variable is a branch-level characteristic that clustering creates — households within high-intensity-disclosure branches are also more likely to receive financial coaching.

**What to do:** Add the variable if measurable. Use fixed effects to absorb all stable unit-level confounders. Use IV, RDD, or DiD if the confound is unmeasurable (covered in Parts 2–6).
:::

::: {.callout-note icon=false}
## Reviewer Challenge: Omitted Variable Bias

> *"Your interpretation assumes you have controlled for all common causes of X and Y. What unmeasured variable could simultaneously predict your predictor and your outcome?"*

**Pattern that invites the challenge:** When X and Y are both desirable outcomes (higher savings AND higher engagement; better decisions AND more information), reviewers will ask what stable individual characteristic could explain both.

**What to show:** The OVB bias formula tells you the *direction* of the problem even without the data: Sign(Z→X) × Sign(Z→Y) = sign of your bias. Report this logic explicitly. If you have plausible confounders, bound the bias using coefficient stability (Oster, 2019) or sensitivity analysis (E-values).
:::

---

## Source 2: Simultaneity / Reverse Causality

X causes Y, but Y also causes X. The causal arrow runs in both directions, creating a feedback loop that OLS cannot untangle.

```{r simultaneity-plot}
n_sim <- 300
set.seed(42)

# Consumer trust → financial engagement (positive), but engagement → trust (positive)
# Simulate equilibrium: customer who engages more trusts the institution more,
# but the institution also targets more-trusting customers with higher engagement offers
time_pts <- tibble(
  t          = 1:10,
  engagement = c(30, 35, 38, 40, 41, 42, 43, 43, 43, 43) + rnorm(10, 0, 2),
  trust      = c(60, 55, 52, 50, 49, 48, 47, 47, 47, 47) + rnorm(10, 0, 2)
)

p_sim <- time_pts |>
  pivot_longer(-t) |>
  ggplot(aes(x = t, y = value, color = name, group = name)) +
  geom_line(linewidth = 1.2) + geom_point(size = 2.5) +
  scale_color_manual(values = c("#e63946", "#1e3a5f"),
                     labels = c("Platform Engagement", "Consumer Trust")) +
  labs(title = "Simultaneity: trust → engagement AND engagement → trust",
       subtitle = "OLS on cross-sectional data picks up both arrows simultaneously",
       x = "Time", y = "Level", color = NULL) +
  theme_minimal(base_size = 13)

p_sim
```

**JDM example:** Consumer trust in a financial institution increases platform engagement — but engagement also builds trust (repeated positive experiences reinforce confidence). A cross-sectional regression of engagement on trust picks up both directions simultaneously and gives a confounded estimate.

The same dynamic appears in advertising: promotional messaging drives purchase intent (awareness → intention), but purchase history drives the volume of targeted promotions a customer receives (purchase history → ad targeting). A cross-sectional regression of purchase intent on messaging frequency cannot identify whether messaging causes intent or prior purchasing causes targeting. Longitudinal designs with lagged predictors are required.

::: {.callout-warning icon=false}
## Assumption: No Reverse Causality

**How to check:** Ask: could Y at time $t$ plausibly influence X at time $t$ or $t+1$? If yes, simultaneity is a concern.

**What to do:**
- Use **lagged predictors** ($X_{t-1}$ predicting $Y_t$) in panel data
- Use **instrumental variables** (Part 2) to isolate the causal direction
- Use **cross-lagged panel models** (Module 4, Part 6) which model both causal arrows explicitly
:::

::: {.callout-note icon=false}
## Reviewer Challenge: Reverse Causality

> *"Your model assumes X causes Y, but couldn't Y cause X? What tells us the arrow runs in the direction you claim?"*

**Pattern that invites the challenge:** Cross-sectional data, outcomes that could plausibly influence the predictor, or settings where timing is ambiguous (e.g., both variables are measured at the same moment).

**What to show:** Lagged predictors ($X_{t-1}$ predicting $Y_t$) rule out same-period reverse causality. Natural experiments or randomized instruments that shift X but cannot respond to Y rule out the reverse path. The cross-lagged panel model (Module 4, Part 6) makes both causal arrows estimable.
:::

---

## Source 3: Selection Effects

The observations in your dataset are not a random sample of the population. Who ends up in your data — and which treatment they receive — is systematically related to both X and Y.

**Self-selection:** Customers who choose to open a premium bank account are systematically different from those who don't — they may be more financially literate, more savings-oriented, or already wealthier. Comparing premium account holders to standard customers conflates the effect of the account type with all these pre-existing differences.

**Survivorship bias:** You only observe the financial products that survived long enough to be in your dataset. Products that were discontinued due to poor uptake are absent. Conclusions about product design drawn from current survivors are biased because you are conditioning on market survival.

**Attrition in longitudinal studies:** Participants who drop out of a financial diary study are often systematically different from completers. If the most financially distressed households stop reporting because their situation is worsening, the observed trajectory among completers overstates average financial health improvement.

This source of endogeneity is covered in depth in Part 3.

::: {.callout-note icon=false}
## Reviewer Challenge: Selection Effects

> *"Who is in your sample, and why? Are the people you observed the same as the population your theory is about?"*

**Pattern that invites the challenge:** Studies of platform users, program participants, survey respondents, or experimental volunteers — any setting where participation is voluntary or observable only for a subset.

**What to show:** Compare baseline characteristics of participants to the eligible population (if known). Report participation rates. In longitudinal designs, compare baseline characteristics of completers to dropouts. If selective participation seems likely, use IPW (Module 3, Part 6) or bound the estimate under selection-on-observables.
:::

---

## Source 4: Collider Bias

This source is frequently overlooked but is just as dangerous as the others. **Conditioning on a collider** — a variable caused by *both* X and Y — creates a spurious association between X and Y even when none exists.

A **collider** is a variable C where two causal arrows "collide":

$$X \rightarrow C \leftarrow Y$$

When you condition on C (include it as a control, restrict your sample to a subset defined by C, or analyze a non-random sample selected on C), you open a spurious backdoor path between X and Y.

```{r collider-demo}
n_col <- 1000

# True DGP: X and Y are uncorrelated
X_col <- rnorm(n_col)
Y_col <- rnorm(n_col)  # truly independent of X

# Collider C is caused by both X and Y
C_col <- X_col + Y_col + rnorm(n_col, 0, 0.5)

# Full sample: no correlation
r_full <- cor(X_col, Y_col)

# Conditioning on C: restrict to top half of C
high_C <- C_col > median(C_col)
r_cond <- cor(X_col[high_C], Y_col[high_C])

tibble(
  Sample           = c("Full sample (no conditioning)", "Restricted to high-C (conditioning)"),
  `Cor(X, Y)`      = c(round(r_full, 3), round(r_cond, 3)),
  Interpretation   = c("True: X and Y are unrelated",
                        "Spurious: conditioning on C creates negative correlation")
) |>
  knitr::kable(caption = "Collider bias: conditioning on C creates a spurious X-Y association")
```

**The publish-or-perish example:** Prolificacy (X) and idea quality (Y) may be uncorrelated in the full population of researchers. But publication (C) is caused by both — prolific researchers with good ideas get published. If you study *only published papers* (conditioning on C), you'll observe that prolificacy and idea quality are *negatively correlated* within the published set, even though they're unrelated in the population.

```{r collider-viz}
df_col <- tibble(X = X_col, Y = Y_col, C_high = high_C)

p_full <- ggplot(df_col, aes(x = X, y = Y)) +
  geom_point(alpha = 0.15, color = "#1e3a5f") +
  geom_smooth(method = "lm", color = "#e63946", linewidth = 1.2) +
  labs(title = "Full sample", subtitle = paste0("r = ", round(r_full, 3)),
       x = "X", y = "Y") +
  theme_minimal(base_size = 12)

p_cond <- ggplot(df_col |> filter(C_high), aes(x = X, y = Y)) +
  geom_point(alpha = 0.2, color = "#2d6a4f") +
  geom_smooth(method = "lm", color = "#e63946", linewidth = 1.2) +
  labs(title = "Conditioned on high C", subtitle = paste0("r = ", round(r_cond, 3)),
       x = "X", y = "Y") +
  theme_minimal(base_size = 12)

p_full + p_cond
```

::: {.callout-tip icon=false}
## JDM Collider: Analyzing Only Platform Engagers

In a study of platform users, promotional sensitivity (X) and consumer wealth (Y) may be uncorrelated in the broader population. But both predict active platform engagement (C). If you analyze only high-engagement users — restricting your sample based on C — you introduce a spurious negative correlation between X and Y within your sample. Wealthier users who are also promo-sensitive are over-represented; the model sees this as a negative relationship between wealth and promo-sensitivity within the active user pool. This is collider bias from sample restriction, not from including C as a covariate.

The same logic applies any time you analyze a selected subset: only people who purchased, only participants who completed all waves, only firms that survived.
:::

::: {.callout-warning icon=false}
## Assumption: No Conditioning on Colliders

**How to check:** Draw a DAG. For every variable you are controlling for or selecting on, ask: is this caused by X, by Y, or by both? If caused by both, it is a collider — do not condition on it.

**Common mistake:** "Kitchen sink" regressions that add every available covariate. If any of those covariates is a collider, adding it *creates* bias where none existed.
:::

::: {.callout-note icon=false}
## Reviewer Challenge: Collider / Bad Control Bias

> *"The variable you added to your model — are you sure it's a confounder and not a collider? Could controlling for it have opened a backdoor path rather than closed one?"*

**Pattern that invites the challenge:** "Kitchen sink" regressions; controlling for behavioral outcomes that happen after X; restricting samples based on an outcome-related criterion (only analyzing purchasers, only active users, only completers).

**What to show:** Draw the causal diagram. For every covariate included, confirm it is caused by neither X nor Y (confounder), not caused by both (collider), and not on the causal path from X to Y (mediator). If uncertain, run the analysis with and without the suspect variable.
:::

---

## Pearl's Causal Framework: Three Types of Paths

Judea Pearl's causal graph (DAG) framework provides a systematic way to identify endogeneity. Every causal graph has three fundamental path structures:

```{r dag-structures}
#| fig-width: 10
#| fig-height: 4.5
make_dag_panel <- function(nodes, edges_df, title, note = NULL) {
  p <- ggplot() +
    geom_point(data = nodes, aes(x = x, y = y), size = 18,
               shape = 21, fill = "#deebf7", color = "#1e3a5f", stroke = 1.5) +
    geom_text(data = nodes, aes(x = x, y = y, label = label), size = 3.8,
              fontface = "bold", color = "#1e3a5f")
  for (i in seq_len(nrow(edges_df))) {
    p <- p + annotate("segment",
                      x    = edges_df$x1[i], xend = edges_df$x2[i],
                      y    = edges_df$y1[i], yend = edges_df$y2[i],
                      arrow = arrow(length = unit(0.25, "cm")),
                      color = edges_df$color[i], linewidth = 1.2)
  }
  # Place note text inside the plot area (below nodes) — avoids patchwork overflow
  if (!is.null(note)) {
    p <- p + annotate("text", x = 2, y = 1.1, label = note,
                      size = 2.55, color = "gray30", hjust = 0.5, vjust = 1,
                      lineheight = 1.3)
  }
  p + labs(title = title) +
    xlim(0.3, 3.7) + ylim(0.1, 2.8) + theme_void(base_size = 11) +
    theme(plot.title = element_text(face = "bold", size = 11, hjust = 0.5,
                                    margin = margin(b = 4)))
}

# Nodes shifted upward so text annotation sits cleanly below them
nodes_fork  <- tibble(label = c("Z", "X", "Y"), x = c(2, 1, 3), y = c(2.3, 1.6, 1.6))
edges_fork  <- tibble(x1 = c(1.85, 2.15), x2 = c(1.2,  2.8),
                      y1 = c(2.20, 2.20), y2 = c(1.70, 1.70),
                      color = c("#e63946", "#e63946"))
p_fork <- make_dag_panel(nodes_fork, edges_fork, "Fork (Confounder)",
  "Z → X  and  Z → Y\nCreates spurious X-Y correlation\nEx: literacy (Z) → disclosure (X)\n   and risky choice (Y)\nFix: control for Z")

nodes_chain <- tibble(label = c("X", "M", "Y"), x = c(1, 2, 3), y = c(1.7, 1.7, 1.7))
edges_chain <- tibble(x1 = c(1.2, 2.2), x2 = c(1.8, 2.8),
                      y1 = c(1.7, 1.7), y2 = c(1.7, 1.7),
                      color = c("#1e3a5f", "#1e3a5f"))
p_chain <- make_dag_panel(nodes_chain, edges_chain, "Chain (Mediator)",
  "X → M → Y\nM is on the causal path\nEx: auto-enroll (X) → savings habit (M)\n   → long-term savings (Y)\nFix: do NOT control for M")

nodes_coll  <- tibble(label = c("X", "C", "Y"), x = c(1, 2, 3), y = c(1.7, 1.7, 1.7))
edges_coll  <- tibble(x1 = c(1.2, 2.8), x2 = c(1.8, 2.2),
                      y1 = c(1.7, 1.7), y2 = c(1.7, 1.7),
                      color = c("#1e3a5f", "#1e3a5f"))
p_coll <- make_dag_panel(nodes_coll, edges_coll, "Collider",
  "X → C ← Y\nConditioning on C opens a backdoor\nEx: promo sensitivity (X) →\nplatform activity (C) ← wealth (Y)\nFix: do NOT control for C")

p_fork + p_chain + p_coll
```

| Path Type | Structure | What it means | What to do |
|-----------|-----------|---------------|------------|
| **Fork** | Z → X and Z → Y | Z confounds X-Y relationship | Block by controlling for Z |
| **Chain** | X → M → Y | M mediates X's effect on Y | Do NOT block (it carries the effect) |
| **Collider** | X → C ← Y | C is caused by both X and Y | Do NOT block (conditioning opens a backdoor) |

The goal of every identification strategy in the rest of this module is to **block all backdoor paths** (forks) without blocking causal paths (chains) and without conditioning on colliders.

---

## The Backdoor Criterion

A set of variables S satisfies Pearl's **backdoor criterion** for estimating the effect of X on Y if:

1. S blocks every path between X and Y that has an arrow *into* X (i.e., all backdoor paths)
2. S does not include any descendant of X on a causal path to Y (no colliders, no mediators on the X→Y path)

When you can find such a set S from measured variables, controlling for S gives you the causal effect. When you cannot — because the confounders are unmeasured — you need the identification strategies in Parts 2–6.

::: {.callout-important}
## Before Running Any Regression

Draw a causal diagram. Every arrow is a hypothesis about the DGP. Then ask:

1. Are there any **forks** (confounders) not in my model? → OVB
2. Could the outcome affect the predictor? → Simultaneity
3. Is my sample selected on something related to both X and Y? → Selection
4. Am I controlling for anything caused by X or Y? → Collider bias

Answering these four questions before looking at your results will save you from the most common errors in applied regression.
:::

::: {.callout-tip icon=false}
## Observational → Experimental Translation

**In observational nested data:**
The predictor X is almost never exogenous. Employees who participate in training programs are not randomly selected — they tend to be younger, more motivated, and employed at better-resourced firms. The OVB formula quantifies how much the naive estimate is inflated by this selection.

**In randomized experiments:**
Random assignment solves the endogeneity problem *for the assigned condition* — but not for everything downstream. Actual treatment receipt (compliance), session attendance, engagement level, and follow-up completion all reintroduce endogeneity after randomization. The residual endogeneity in experiments is the subject of Modules 5 Parts 8 and 9.

**Same diagnosis, different source:**
In both settings, the question is identical: *Is there a variable that simultaneously predicts X and Y, that I have not accounted for?* In observational data, this is about confounders in the sampling. In experiments, it is about post-randomization processes that reintroduce selection.
:::
