---
title: "Instrumental Variables"
---

```{r setup, include=FALSE}
library(tidyverse)
library(AER)
library(broom)
library(ggplot2)
library(patchwork)
set.seed(42)
```

# Instrumental Variables

## When Endogeneity Cannot Be Removed by Fixed Effects

Module 1's fixed effects model removes time-invariant confounders by exploiting within-unit variation over time. But fixed effects are not a universal cure for endogeneity. Two failure modes are especially common.

**In longitudinal panel research**, the endogeneity can be time-varying: bank branches that are on an upward savings trajectory are exactly the branches most likely to adopt new financial planning tools — they have the motivated staff, engaged customer base, and organizational energy to implement them. Branch fixed effects absorb stable branch quality, but they cannot absorb a time-varying engagement shock that simultaneously drives tool adoption and savings improvement. Even after demeaning, the estimate remains confounded.

**In experiments**, randomization solves endogeneity for assignment — but not always for treatment receipt. Participants randomly offered access to a financial planning tool may never log in. Customers sent a disclosure summary may never open it. When compliance is imperfect, the simple comparison of assigned groups underestimates the effect of actually using the tool.

In both cases, an **instrument** is a variable that provides exogenous variation in treatment uptake — variation that is not driven by the confounders affecting outcomes.

- **Longitudinal context:** The scheduled wave in which a branch received an IT infrastructure upgrade provides exogenous variation in financial planning tool adoption. Branches in Wave 1 adopted the tool at higher rates not because of their savings trajectory but because of administrative scheduling order. The rollout wave does not directly affect household savings — only the tool adoption it enables does.

- **Experimental context:** Random assignment to an encouragement condition (a structured outreach protocol with personalized prompts) provides exogenous variation in actual financial tool use. Whether you were randomized to encouragement does not directly improve your savings decisions — logging in and engaging with the tool does.

---

## The IV Structure: Permitted and Forbidden Pathways

The instrumental variables strategy works by tracing exactly which paths between Z, X, and Y are permitted and which must be blocked. The diagram below shows this structure for the branch savings example.

```{r iv-dag, fig.width=8, fig.height=4.2}
library(ggplot2)

# Node positions
nodes <- tibble(
  label = c("Z\n(IT rollout\nwave)", "X\n(Tool\nadoption)", "Y\n(Savings\nrate)", "U\n(Branch\nquality)"),
  x = c(0, 2.5, 5, 2.5),
  y = c(2,   2,  2,   0)
)

arrow_end <- 0.45   # shorten arrow ends to not overlap node circles

ggplot() +
  # --- permitted arrows (blue) ---
  # Z → X  (relevance)
  annotate("segment", x=0.45, xend=2.05, y=2, yend=2,
           arrow=arrow(length=unit(.25,"cm"), type="closed"),
           color="#1e3a5f", linewidth=1.3) +
  # X → Y  (causal effect — what we identify)
  annotate("segment", x=2.95, xend=4.55, y=2, yend=2,
           arrow=arrow(length=unit(.25,"cm"), type="closed"),
           color="#1e3a5f", linewidth=1.3) +
  # U → X  (confound)
  annotate("segment", x=2.5, xend=2.5, y=0.45, yend=1.55,
           arrow=arrow(length=unit(.25,"cm"), type="closed"),
           color="#888888", linewidth=1, linetype="dashed") +
  # U → Y  (confound)
  annotate("segment", x=2.95, xend=4.55, y=0.2, yend=1.65,
           arrow=arrow(length=unit(.25,"cm"), type="closed"),
           color="#888888", linewidth=1, linetype="dashed") +
  # --- FORBIDDEN arrow: Z → Y (dashed red, exclusion violation) ---
  annotate("segment", x=0.45, xend=4.55, y=1.7, yend=1.7,
           arrow=arrow(length=unit(.25,"cm"), type="closed"),
           color="#cc0000", linewidth=1, linetype="dotted") +
  annotate("text", x=2.5, y=1.45, label="FORBIDDEN\n(exclusion restriction)",
           size=3.2, color="#cc0000", fontface="bold") +
  # --- FORBIDDEN link: Z ↔ U (dashed orange, independence violation) ---
  annotate("segment", x=0.45, xend=2.1, y=1.65, yend=0.35,
           arrow=arrow(length=unit(.25,"cm"), type="closed"),
           color="#e67e00", linewidth=1, linetype="dotted") +
  annotate("text", x=0.7, y=0.85, label="FORBIDDEN\n(independence)", hjust=0,
           size=3.2, color="#e67e00", fontface="bold") +
  # --- node circles ---
  geom_point(data=nodes, aes(x=x, y=y), shape=21, size=14,
             fill=c("#d4e6f5","#d4e6f5","#d4e6f5","#fce8cc"),
             color=c("#1e3a5f","#1e3a5f","#1e3a5f","#a05000"),
             stroke=1.5) +
  geom_text(data=nodes, aes(x=x, y=y, label=label), size=3.3, lineheight=0.9) +
  # --- path labels ---
  annotate("text", x=1.25, y=2.18, label="Relevance\n(Z predicts X)",
           size=3.2, color="#1e3a5f") +
  annotate("text", x=3.75, y=2.18, label="Causal path\n(what IV identifies)",
           size=3.2, color="#1e3a5f") +
  annotate("text", x=3.5,  y=0.9,  label="Confounders\n(U affects X and Y)",
           size=3.2, color="#888888") +
  xlim(-0.5, 5.7) + ylim(-0.4, 2.8) +
  labs(
    title    = "IV causal diagram: permitted paths and forbidden pathways",
    subtitle = "Blue = valid paths (Z identifies X→Y effect)    Red/orange = paths that must be blocked for IV to be valid"
  ) +
  theme_void(base_size = 12) +
  theme(plot.title    = element_text(size=13, face="bold"),
        plot.subtitle = element_text(size=11, color="#555555"),
        plot.margin   = margin(5, 5, 5, 5))
```

The blue arrows show the structure IV exploits: Z creates exogenous variation in X (relevance), and X causally affects Y. The grey dashed arrows show the endogeneity problem: U confounds both X and Y, so naive OLS is biased. The **red dotted arrow** (Z → Y directly) is the **exclusion restriction violation** — if this path exists, Z affects Y through channels other than X, and the IV estimate is biased. The **orange dotted arrow** (Z → U) is the **independence violation** — if Z is correlated with the very confounders it is meant to bypass, it is not truly exogenous.

When all three forbidden pathways are blocked, the IV estimate recovers the causal effect of X on Y.

---

## The Three IV Assumptions

::: {.callout-warning icon=false}
## Assumption 1: Relevance

**What it means:** The instrument Z must actually cause the endogenous treatment X to change. An instrument that does not predict treatment is useless — it produces inflated standard errors and, when weak, can bias estimates in unpredictable directions.

**How to check:** Regress X on Z (the "first stage"). The F-statistic on Z must exceed 10. Below that threshold, the instrument is "weak."

**Longitudinal context:** The IT rollout wave must meaningfully predict whether branches adopt the financial planning tool. Wave 1 branches should adopt at substantially higher rates than Wave 2 branches during the observation window.

**Experimental context:** Being randomized to the encouragement condition must meaningfully predict actual tool use. If the encouragement protocol is so feeble that it does not increase log-ins and engagement, it cannot serve as an instrument.
:::

::: {.callout-warning icon=false}
## Assumption 2: Exclusion Restriction

**What it means:** Z affects Y *only through* X — not through any other pathway. This is the core untestable assumption. No statistical test can verify it; it requires theoretical argument and domain knowledge.

**How to scrutinize:** Trace every conceivable path from Z to Y that bypasses X. Each such path is a potential violation.

**Longitudinal context:** The IT rollout wave must not directly affect household savings rates. It should matter only because it determines which branches have the financial planning tool available. A potential violation: Wave 1 branches may have been selected because they were in larger, more urban markets — advantages that raise savings rates regardless of whether the tool is adopted.

**Experimental context:** Random assignment to the encouragement condition must not improve savings decisions except through actual tool use. A potential violation: receiving a personalized outreach message might itself prompt reflection on financial goals — a direct effect on savings behavior that bypasses the tool entirely, even if the participant never logs in.
:::

::: {.callout-warning icon=false}
## Assumption 3: Independence (Exogeneity of the Instrument)

**What it means:** Z is as-good-as-randomly assigned — it is not caused by the same confounders that affect X and Y.

**How to check:** Verify that Z is balanced across observable pre-treatment characteristics. In a natural experiment, check whether the instrument predicts baseline covariates.

**Longitudinal context:** The IT rollout wave should not be correlated with branch quality or growth trajectory. Branches did not self-select into Wave 1 based on their savings performance potential — the order was determined by infrastructure scheduling. In practice: check whether baseline savings rates and branch size predict wave assignment.

**Experimental context:** In a proper encouragement design, Z is genuinely randomized — independence holds by construction. The practical threat is post-randomization contamination (e.g., differential attrition by encouragement condition — less financially engaged customers more likely to drop out of the encouraged group) or a failed randomization procedure.
:::

---

## Section 3: Longitudinal IV — Budgeting Tool Adoption and Household Savings

### The Setup

200 bank branches are observed over 6 years. The research question is whether adopting a financial planning tool increases branch-level household savings rates. The identification problem: branches on an upward savings trajectory — engaged staff, motivated customer base — are exactly the ones most likely to implement new tools proactively. A naive OLS regression overestimates the tool's causal effect.

Even adding branch fixed effects does not fully resolve this: if branches experiencing a positive engagement shock (a new community outreach manager, an unusually successful financial literacy event) respond by deploying the tool, within-branch variation in adoption is still confounded by this time-varying engagement shock.

The instrument is rollout wave — a staggered IT infrastructure upgrade that determined which branches could deploy the tool first. The wave order was set by administrative scheduling (server installation logistics), not by branch savings performance. Branches in Wave 1 adopted the tool at much higher rates than Wave 2 branches during the early years.

```{r panel-iv-setup}
n_branches <- 200
n_years    <- 6

branch_quality <- rnorm(n_branches, 0, 5)   # unobserved branch engagement quality
# Wave 1 vs. Wave 2 (1 = early rollout, 0 = later rollout) — as-good-as-random
wave1          <- rbinom(n_branches, 1, 0.5)

df_panel <- expand_grid(branch = 1:n_branches, year = 1:n_years) |>
  mutate(
    bq       = branch_quality[branch],
    wave     = wave1[branch],
    # High-quality branches AND Wave 1 branches adopt the tool more
    p_adopt  = plogis(-0.5 + 0.4 * (bq / sd(branch_quality)) + 0.8 * wave +
                        rnorm(n(), 0, 0.3)),
    adopted  = rbinom(n(), 1, p_adopt),
    # True causal effect of tool adoption = 4 pp savings gain; quality also helps
    savings  = 50 + 4 * adopted + 2.5 * bq + rnorm(n(), 0, 4),
    branch   = factor(branch),
    year     = factor(year)
  )
```

### First Stage: Instrument Relevance

```{r panel-first-stage}
m_first <- lm(adopted ~ wave, data = df_panel)
f_stat  <- summary(m_first)$fstatistic[1]
cat("First-stage F-statistic:", round(f_stat, 1), "\n")
cat("(Rule of thumb: F > 10 indicates a relevant instrument)\n")
```

Wave 1 is a strong positive predictor of tool adoption (Wave 1 branches adopt at substantially higher rates). F > 10 confirms the instrument is not weak.

### OLS vs. IV

```{r panel-ols-vs-iv}
# Naive OLS: inflated by branch engagement confound
m_ols <- lm(savings ~ adopted, data = df_panel)

# IV via Two-Stage Least Squares: uses only wave-driven variation in adoption
m_iv  <- ivreg(savings ~ adopted | wave, data = df_panel)

bind_rows(
  tidy(m_ols) |> filter(term == "adopted") |>
    mutate(model = "OLS (confounded — true = 4)"),
  tidy(m_iv)  |> filter(term == "adopted") |>
    mutate(model = "IV / TSLS (consistent under exclusion restriction)")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Panel IV: effect of tool adoption on savings rate (true effect = 4). OLS is biased upward by branch quality; IV is consistent but with larger SE.")
```

OLS overestimates the tool's effect because branch engagement simultaneously drives adoption and savings improvement. IV isolates only the exogenous, wave-driven variation in adoption. IV is **consistent** — its bias shrinks to zero as $N \to \infty$ — but in any single finite sample the estimate can be above or below the truth. The larger standard error reflects this uncertainty honestly.

---

## Section 4: Two-Stage Least Squares — The Mechanics

TSLS makes the IV logic explicit in two regression steps.

**Stage 1:** Regress the endogenous treatment X on the instrument Z (and any controls W):

$$\hat{X}_i = \hat{\gamma}_0 + \hat{\gamma}_1 Z_i + \hat{\gamma}_2 W_i$$

This produces fitted values $\hat{X}$ that contain only the variation in X that is driven by Z — the "clean," exogenous part of treatment uptake.

**Stage 2:** Regress Y on $\hat{X}$ (and controls W):

$$Y_i = \beta_0 + \beta_1 \hat{X}_i + \beta_2 W_i + \varepsilon_i$$

Because $\hat{X}$ contains no endogenous variation, $\hat{\beta}_1$ is a consistent causal estimate.

```{r tsls-manual}
# Manual TSLS to show the mechanics
# Stage 1: predict tool adoption from rollout wave
stage1    <- lm(adopted ~ wave, data = df_panel)
df_panel  <- df_panel |> mutate(adopted_hat = fitted(stage1))

# Stage 2: regress savings on predicted adoption
stage2    <- lm(savings ~ adopted_hat, data = df_panel)

cat("Manual TSLS (Stage 2 coefficient on adopted_hat):",
    round(coef(stage2)["adopted_hat"], 3), "\n")
cat("ivreg() coefficient (should match):",
    round(coef(m_iv)["adopted"], 3), "\n")
```

::: {.callout-note icon=false}
## Note: Standard Errors in Manual TSLS

The coefficients from manual TSLS match `ivreg()`, but the standard errors do **not**. The second-stage regression uses $\hat{X}$ (a generated regressor), so residuals are computed against the wrong quantity. Always use `ivreg()` or an equivalent dedicated IV function to obtain correct standard errors.
:::

### Connection to Module 1 Fixed Effects

Panel fixed effects absorb time-invariant branch-level confounders. But if tool adoption responds to time-varying branch engagement shocks, fixed effects alone are insufficient. Adding IV on top of FE handles both:

$$\underbrace{\text{FE removes}}_{\substack{\text{time-invariant} \\ \text{confounders}}} \quad + \quad \underbrace{\text{IV removes}}_{\substack{\text{time-varying} \\ \text{endogeneity}}}$$

```{r panel-fe-iv}
# Panel FE-IV: year fixed effects as controls; rollout wave as instrument
m_fe_iv <- ivreg(savings ~ adopted + year | wave + year, data = df_panel)
cat("FE-IV estimate (adopted):", round(coef(m_fe_iv)["adopted"], 3), "\n")
```

By including year fixed effects in both stages, we account for common time trends while using rollout wave to instrument tool adoption within each year.

---

## Section 5: Experimental IV — Encouragement Designs and LATE

In experiments, the instrument is the **random assignment** Z itself. When all participants comply, the simple comparison of assigned groups gives an unbiased causal estimate. But when compliance is imperfect, Z becomes an instrument for actual treatment receipt D.

### The Wald Estimator

$$\text{LATE} = \frac{\underbrace{E[Y \mid Z=1] - E[Y \mid Z=0]}_{\text{Intent-to-treat effect (ITT)}}}{\underbrace{E[D \mid Z=1] - E[D \mid Z=0]}_{\text{Compliance rate (first stage)}}}$$

The **Intent-to-treat (ITT)** effect — the effect of assignment, regardless of compliance — is an unbiased estimate of the causal effect of *being assigned* to treatment. Dividing by the compliance rate scales the ITT up to recover the LATE: the causal effect of *actually using* the tool, among those who comply.

### Compliance Classes

When Z = random assignment and D = actual tool use, participants fall into four latent types:

| Type | D if Z = 0 | D if Z = 1 | Description |
|-------------|------------|------------|--------------------------------------|
| Complier | 0 | 1 | Uses tool iff encouraged — IV identifies this group |
| Always-taker | 1 | 1 | Uses tool regardless — uninformative |
| Never-taker | 0 | 0 | Never uses tool even when encouraged — uninformative |
| Defier | 1 | 0 | Does the opposite of assignment — assumed absent (monotonicity) |

Only **compliers** contribute to the IV estimate. LATE is the effect of using the tool, for compliers only.

### Simulation: Financial Planning Tool Encouragement Design

160 customers are randomized to encouragement (n = 80, receives personalized outreach and prompts to use the budgeting tool) or control (n = 80, standard materials). Of those encouraged, 70% actually engage with the planning tool; none in the control group have access. The true causal effect of *actually using* the tool is a 10-percentage-point improvement in savings rate.

```{r encouragement-design}
n <- 160
Z <- rep(c(0, 1), each = 80)  # random assignment to encouragement

# One-sided noncompliance: encouraged participants use tool at 70%
D <- ifelse(Z == 1, rbinom(80, 1, 0.70), 0)

# True effect of using the tool = +10 pp savings improvement
# Self-selected tool users (always-takers) are more financially motivated — biases as-treated
savings_rate <- 40 + 10 * D + rnorm(n, 0, 8)
df_exp       <- tibble(Z = factor(Z), D = factor(D), savings_rate, id = 1:n)

# Three estimators
m_itt  <- lm(savings_rate ~ Z, data = df_exp)         # ITT: effect of assignment
m_at   <- lm(savings_rate ~ D, data = df_exp)         # As-treated: biased by selection
m_late <- ivreg(savings_rate ~ D | Z, data = df_exp)  # LATE: IV estimate

bind_rows(
  tidy(m_itt)  |> filter(term == "Z1") |>
    mutate(model = "ITT — effect of assignment (diluted, unbiased)"),
  tidy(m_at)   |> filter(term == "D1") |>
    mutate(model = "As-treated — biased by self-selection"),
  tidy(m_late) |> filter(term == "D1") |>
    mutate(model = "IV / LATE — consistent estimate for compliers")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Encouragement design: three estimators (true tool effect = +10). ITT is diluted; as-treated overstates due to self-selection; LATE is consistent but individual samples can deviate from truth.")
```

```{r wald-manual}
# Manual Wald estimator: confirms IV equals ITT / compliance rate
itt_effect <- mean(savings_rate[Z == 1]) - mean(savings_rate[Z == 0])
compliance <- mean(D[Z == 1] == 1) - mean(D[Z == 0] == 1)
wald_late  <- itt_effect / compliance

cat("ITT (numerator)         :", round(itt_effect, 3), "\n")
cat("Compliance (denominator):", round(compliance, 3), "\n")
cat("Wald LATE               :", round(wald_late, 3), "\n")
cat("ivreg() LATE            :", round(coef(m_late)["D1"], 3), "\n")
```

The ITT is approximately +7: diluted because only 70% of encouraged customers actually used the tool. The as-treated estimate is biased upward — customers who chose to engage with the tool are likely more financially motivated than the average participant, so they would have saved more regardless. The IV/LATE is **consistent** under the exclusion restriction: on average across many samples it targets +10, though any single sample estimate can deviate from the truth due to sampling variability.

---

## Section 6: The LATE Interpretation

The IV estimand is always a LATE — the causal effect for compliers only. This has concrete implications in both research contexts.

**What LATE means in the longitudinal case:** The estimate applies specifically to branches that adopted the financial planning tool *because* they happened to be in Wave 1 of the rollout — not to branches that would have adopted regardless (always-takers: highly engaged branches that would find a way to deploy the tool in any scenario) or those that would never have adopted regardless (never-takers: branches with too little infrastructure or staff capacity). If Wave-1-compliant branches are mid-sized, moderately engaged branches, the LATE describes the effect for that marginal segment — it may not generalize to either the most or least engaged branches.

**What LATE means in the experimental case:** The estimate applies to customers who engaged with the financial planning tool *because* they were encouraged — not to always-takers who would have logged in and used it on their own initiative, and not to never-takers who would not engage even with personalized outreach. If compliers are ambivalent but malleable (responsive to prompts, but not intrinsically financially motivated), the LATE may understate how much the tool helps self-selected always-takers, while overstating how much it would help confirmed never-takers if somehow compelled to use it.

**Generalizability:** LATE is a valid causal estimate, but it is local. Policy conclusions drawn from LATE assume that the complier subpopulation is relevant to the policy question. When it is not — when you need to know the ATE for everyone — IV alone is insufficient.

---

## Section 7: Finding Good Instruments

The exclusion restriction is the bottleneck. Common instrument classes in social science:

| Instrument class | Longitudinal example | Experimental example | Why valid |
|-----------------------|------------------------------|-------------------------------|------------------------------|
| Administrative assignment | IT rollout wave for budgeting tool | Site randomization in multi-site trials | Order set by logistics, not branch quality |
| Natural experiment | Weather shock disrupting branch operations | Random lottery for waitlisted customers | Random by nature or design |
| Regulatory / eligibility threshold | Credit score cutoff for prime-rate access | Financial distress screening cutoff (RDD–IV) | Arbitrary administrative rule |
| Time-based shock | Policy change or fee structure revision | Seasonal or day-of-week variation in study timing | Exogenous timing |
| Geographic variation | Distance to nearest financial literacy center | Experimenter assignment in multi-advisor designs | Not chosen by customer or branch |

The mathematics of TSLS are straightforward. Finding a credible instrument — one whose exclusion restriction can be defended from theory and whose relevance can be demonstrated empirically — requires deep domain knowledge and careful thinking about the data-generating process. A weak but credible instrument is preferable to a strong instrument with a plausible exclusion-restriction violation. The first produces inflated standard errors; the second produces biased point estimates.
