---
title: "Instrumental Variables"
---

```{r setup, include=FALSE}
library(tidyverse)
library(AER)
library(broom)
set.seed(42)
```

# Instrumental Variables

## When You Cannot Randomize

The gold standard for causal inference is a randomized experiment. But often you cannot randomize. An **instrument** is a variable that gives you some of the benefits of randomization without actually randomizing. It shifts X exogenously — through a channel that does not itself affect Y.

## The Three Requirements

::: {.callout-warning icon=false}
## Assumption 1: Relevance

**What it means:** Z must actually cause X to change. An instrument that does not predict the predictor is useless (and produces infinite standard errors).

**How to check:** Regress X on Z (the "first stage"). The F-statistic on Z must be large — the standard rule of thumb is F > 10. Below 10, the instrument is "weak" and estimates become unreliable.
:::

::: {.callout-warning icon=false}
## Assumption 2: Exclusion Restriction

**What it means:** Z affects Y *only through* X — not through any other path. This is the key untestable assumption. You cannot verify it from data alone — you can only argue for it from theory and domain knowledge.

**How to scrutinize:** Trace every possible path from Z to Y. Is there any path that does not go through X?
:::

::: {.callout-warning icon=false}
## Assumption 3: Independence (Exogeneity of the Instrument)

**What it means:** Z is as-good-as-randomly assigned — it is not itself caused by the confounders that affect X and Y.

**How to check:** Verify that Z is balanced across observable pre-treatment covariates.
:::

## Two-Stage Least Squares (TSLS)

**Stage 1:** Regress X on Z. Extract fitted values $\hat{X}$ — the exogenous component of X.

**Stage 2:** Regress Y on $\hat{X}$. This is the IV causal estimate.

```{r tsls-example}
n <- 800

# Confounders: fatigue and commute stress
fatigue        <- rnorm(n, 50, 10)
commute_stress <- rnorm(n, 40, 8)

# Instrument: daylight saving time shift (exogenous clock reset)
DST_week <- rbinom(n, 1, 0.5)

# Time of day: influenced by DST AND by fatigue/stress
time_of_day <- 12 + 1.5 * DST_week - 0.1 * fatigue + rnorm(n, 0, 2)

# Variety seeking: affected by time of day AND by confounders
# True causal effect of time_of_day = 0.4
variety_seeking <- 3 + 0.4 * time_of_day - 0.2 * fatigue -
                   0.15 * commute_stress + rnorm(n, 0, 3)

df_iv <- tibble(variety_seeking, time_of_day, DST_week, fatigue, commute_stress)

# First stage: check instrument strength
m_first <- lm(time_of_day ~ DST_week, data = df_iv)
cat("First stage F-statistic:", round(summary(m_first)$fstatistic[1], 1), "\n")

# Naive OLS
m_ols <- lm(variety_seeking ~ time_of_day, data = df_iv)

# IV via TSLS
m_iv <- ivreg(variety_seeking ~ time_of_day | DST_week, data = df_iv)

bind_rows(
  tidy(m_ols) |> filter(term == "time_of_day") |> mutate(model = "OLS (biased)"),
  tidy(m_iv)  |> filter(term == "time_of_day") |> mutate(model = "IV / TSLS")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "OLS vs. IV: effect of time-of-day on variety seeking (true = 0.4)")
```

The OLS estimate is inflated by the confounders. The IV estimate recovers the true causal effect by using only the DST-driven variation in time-of-day.

## The LATE Interpretation

The IV estimate is not the population-average causal effect. It is the **Local Average Treatment Effect (LATE)**: the causal effect for **compliers** — units whose treatment status actually changed because of the instrument. People unaffected by DST (those who keep the same schedule regardless) are uninformative for the IV estimate.

This limits generalizability. Whether the LATE is the quantity you care about depends on your research question.

## Finding Good Instruments

The exclusion restriction is the bottleneck. Common instrument classes in social science:

| Instrument class | Example | Why valid |
|-----------------|---------|-----------|
| Natural experiment | Lottery assignment | Random by design |
| Regulatory threshold | Age cutoff for program | Arbitrary threshold |
| Geographic variation | Distance to hospital | Not chosen by individual |
| Time-based shock | Policy change, DST | Exogenous timing |
| Peer assignment | Random roommate assignment | Administrative randomization |

The math is easy. Finding a credible instrument — one whose exclusion restriction holds and can be argued for — requires deep domain knowledge and creative thinking about your DGP.
