---
title: "Endogeneity"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(broom)
set.seed(1234)
```

# Endogeneity

## What Does "Independent" Really Mean?

In regression, we call X the "independent variable." The name suggests that X is free of any influence — that it varies on its own, unrelated to the forces that also shape Y. In a well-executed randomized experiment, this is approximately true.

In observational data — and in most nested data — this assumption is almost never satisfied.

**Endogeneity** means that your predictor X is correlated with the error term in your regression. Some of the forces that drive X are the same forces that drive Y, beyond what your model accounts for. When this happens, your regression coefficient is not an estimate of the effect of X on Y. It is a blend of that effect and spurious associations created by shared forces.

There are three main sources of endogeneity.

---

## Source 1: Omitted Variable Bias (OVB)

A variable Z influences both X and Y, but you have not included Z in your model. The effect of Z on Y gets attributed to X.

**Classic example:** Ice cream sales (X) predict drowning deaths (Y). Not because ice cream causes drowning — but because summer heat (Z) increases both.

```{r ovb-simulation}
n <- 500
Z <- rnorm(n, 70, 10)       # Temperature (the confounder)
X <- 0.3 * Z + rnorm(n, 0, 5)  # Ice cream sales (caused by heat)
Y <- 0.5 * Z + rnorm(n, 0, 8)  # Drowning deaths (caused by heat, NOT ice cream)
# True effect of X on Y = 0

df_ovb <- tibble(X, Y, Z)

m_biased  <- lm(Y ~ X, data = df_ovb)
m_correct <- lm(Y ~ X + Z, data = df_ovb)

bind_rows(
  tidy(m_biased)  |> filter(term == "X") |> mutate(model = "Omits Z (biased)"),
  tidy(m_correct) |> filter(term == "X") |> mutate(model = "Includes Z (correct)")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "OVB: the ice cream effect disappears once we control for heat")
```

Without Z, the X coefficient is large and significant — even though the true effect is zero.

**In nested data:** OVB is especially insidious because the omitted variable often varies at a higher level. Studying whether neighborhood green space predicts depression while omitting neighborhood income (which predicts both) biases your estimate. The omitted variable is a neighborhood-level characteristic — exactly the kind of thing that clustering creates.

::: {.callout-warning icon=false}
## Assumption: No Omitted Variables Correlated with Both X and Y

**How to check:** Draw a causal diagram. List every plausible cause of Y and ask: is any of these also correlated with X? You cannot fully test this from data — it requires substantive reasoning about the DGP.

**What to do:** Add the variable if measurable. If not measurable, consider IV, fixed effects, or DiD.
:::

---

## Source 2: Simultaneity / Reverse Causality

X causes Y, but Y also causes X. The causal arrow runs in both directions, creating a feedback loop.

**Example:** Income and spending. Higher income causes more spending. But the desire to spend more can motivate people to seek higher income. Both directions are real. A cross-sectional regression picks up a mix of both.

Simultaneity is especially common in panel and longitudinal data, where today's outcome can become tomorrow's predictor — a structure naturally produced by the nested DGP of time.

::: {.callout-warning icon=false}
## Assumption: No Reverse Causality

**How to check:** Ask whether Y at time $t$ could plausibly influence X at time $t$ or $t+1$.

**What to do:** Use lagged predictors ($X_{t-1}$ predicting $Y_t$). Use IV. Use cross-lagged panel models (Module 4, Part 6).
:::

---

## Source 3: Selection Effects

The observations in your dataset are not a random sample. Something about who ends up in your data is related to both X and Y. This is covered in depth in Part 3.

---

## Pearl's Backdoor Criterion

Judea Pearl's framework gives us precise language for endogeneity. A **backdoor path** from X to Y is any path that flows through a common cause — a path that goes "backwards" into X from a confounder.

```{r backdoor-diagram}
dag_df <- tibble(
  node = c("Z\n(Confounder)", "X\n(Predictor)", "Y\n(Outcome)"),
  x    = c(2, 1, 3),
  y    = c(2, 1, 1)
)

ggplot(dag_df, aes(x = x, y = y, label = node)) +
  geom_point(size = 22, color = "#deebf7", shape = 21, fill = "#deebf7", stroke = 1.5) +
  geom_text(size = 3.5, fontface = "bold", color = "#1e3a5f") +
  annotate("segment", x = 1.8, xend = 1.2, y = 1.9, yend = 1.1,
           arrow = arrow(length = unit(0.3, "cm")), color = "#e63946", linewidth = 1.2) +
  annotate("segment", x = 2.2, xend = 2.8, y = 1.9, yend = 1.1,
           arrow = arrow(length = unit(0.3, "cm")), color = "#e63946", linewidth = 1.2) +
  annotate("segment", x = 1.2, xend = 2.8, y = 1.0, yend = 1.0,
           arrow = arrow(length = unit(0.3, "cm")), color = "#1e3a5f", linewidth = 1.5) +
  annotate("text", x = 2, y = 0.75, label = "Causal path of interest",
           color = "#1e3a5f", size = 3.2) +
  annotate("text", x = 1.5, y = 1.62, label = "Backdoor\npath", color = "#e63946", size = 3) +
  xlim(0.5, 3.5) + ylim(0.5, 2.5) +
  theme_void() +
  labs(title = "Backdoor path: Z creates a non-causal correlation between X and Y")
```

Blocking the backdoor is the goal of every identification strategy in this module — either by including Z in the model (when measurable), by using fixed effects to absorb all time-invariant confounds, or by finding an instrument that purges the contaminated variation in X.
