---
title: "Selection Effects & Attrition"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
set.seed(555)
```

# Selection Effects & Attrition

## When Your Sample is Not Your Population

Even with a perfect model, if your observations are not representative of the population you care about, your conclusions may be wrong. Selection effects are the family of processes by which observations non-randomly enter or leave your dataset.

Selection is especially dangerous in nested and longitudinal data because the same processes that structure your data also drive who stays in it over time.

---

## Type 1: Representativeness / Undercover Bias

Your sampling frame does not cover the full population you want to generalize to.

**Secondary data example:** Online review platform data. People who leave reviews are not a random sample — they felt strongly enough to write one (typically either very satisfied or very dissatisfied). Analyses generalize to "opinionated reviewers," not "all customers."

**Experimental example:** Convenience samples from university subject pools (WEIRD populations — Western, Educated, Industrialized, Rich, Democratic) may not generalize to the adults whose behavior you are studying.

::: {.callout-warning icon=false}
## Assumption: Sampling Frame Covers the Population of Interest

**How to check:** Compare your sample's distributions on key variables to known population statistics. Are any groups systematically over- or under-represented?

**What to do:** Acknowledge the limitation. Use stratified sampling. Apply post-stratification weights if population distributions are known.
:::

---

## Type 2: Self-Selection / Non-Response Bias

Participants choose whether to enter your study based on factors related to your variables of interest.

The landmark example: the 1936 *Literary Digest* poll predicted Landon would defeat Roosevelt by a landslide, based on 2.4 million responses. Their sampling frame (telephone directories and car registration lists) over-represented wealthy Republicans. Roosevelt won in a historic landslide.

```{r selection-sim}
n_total    <- 1000
health_con <- rnorm(n_total, 50, 10)  # latent health consciousness

# Health-conscious people more likely to participate
p_participate <- plogis(-2 + 0.06 * health_con)
participated  <- rbinom(n_total, 1, p_participate)

# Outcome influenced by health consciousness
health_score <- 40 + 0.5 * health_con + rnorm(n_total, 0, 8)

tibble(
  Group    = c("Full population", "Self-selected sample"),
  N        = c(n_total, sum(participated)),
  Mean_health = c(mean(health_score),
                   mean(health_score[participated == 1]))
) |>
  knitr::kable(digits = 2, caption = "Self-selection inflates estimated health level")
```

---

## Type 3: Survivorship Bias

You only observe units that survived some selection filter correlated with your outcome.

**WWII aircraft:** Analysts studied bullet-hole patterns on returning planes to decide where to add armor. Abraham Wald pointed out: planes hit in the zero-hole areas never returned. Those were the critical spots to armor.

**In longitudinal research:** Studying predictors of firm success using currently existing firms excludes all firms that failed. If failure is related to your predictors, estimates are systematically optimistic.

::: {.callout-warning icon=false}
## Assumption: No Survivorship Bias

**How to check:** Ask whether your outcome is systematically related to the probability of appearing in your dataset. If yes, you likely have survivorship bias.

**What to do:** Obtain historical data including non-survivors. Use survival analysis (Module 6, Part 3) to model time-to-exit explicitly.
:::

---

## Type 4: Attrition Bias in Longitudinal Data

Participants drop out of your study non-randomly over time. This is one of the most serious threats to longitudinal research — and a direct consequence of the nested DGP of time.

```{r attrition-sim}
n_people <- 200
n_waves  <- 5

baseline_dep <- rnorm(n_people, 50, 10)
treatment    <- rbinom(n_people, 1, 0.5)

df_long <- expand_grid(person = 1:n_people, wave = 1:n_waves) |>
  mutate(
    baseline  = baseline_dep[person],
    treat     = treatment[person],
    depression = baseline - 5 * treat * (wave / n_waves) + rnorm(n(), 0, 4)
  )

# People with HIGH depression are more likely to drop out
df_long <- df_long |>
  group_by(person) |>
  mutate(
    drop_risk   = plogis(-3 + 0.04 * mean(depression)),
    dropped_out = cummax(rbinom(n(), 1, drop_risk)),
    observed    = dropped_out == 0 | wave == 1
  ) |>
  ungroup()

complete_cases <- df_long |>
  group_by(person) |> filter(all(observed)) |> ungroup()

tibble(
  Sample = c("All (Wave 1)", "Completers only"),
  N      = c(n_distinct(df_long$person[df_long$wave == 1]),
              n_distinct(complete_cases$person)),
  Mean_baseline_depression = c(
    mean(baseline_dep),
    mean(unique(complete_cases$baseline))
  )
) |>
  knitr::kable(digits = 2,
    caption = "Attrition: completers have lower baseline depression (healthier)")
```

```{r attrition-effect}
# True treatment effect = -5
m_completers <- lm(depression ~ treat + wave + baseline, data = complete_cases)
m_all        <- lm(depression ~ treat + wave + baseline,
                   data = df_long |> filter(observed))

bind_rows(
  broom::tidy(m_completers) |> filter(term == "treat") |>
    mutate(model = "Complete cases (attrition bias)"),
  broom::tidy(m_all) |> filter(term == "treat") |>
    mutate(model = "All observed data")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Attrition biases the treatment effect toward zero (true = -5)")
```

::: {.callout-warning icon=false}
## Assumption: Dropout Mechanism is Ignorable

**MCAR:** Dropout is unrelated to any variable. Very rare.

**MAR:** Dropout depends only on observed variables. If MAR holds, mixed-effects models using all available data (not just completers) give valid estimates.

**MNAR:** Dropout depends on the unobserved outcome itself. Requires sensitivity analysis or selection models (Heckman).

**How to check:** Compare baseline characteristics of completers vs. dropouts. If they differ substantially, dropout is non-random.
:::

---

## Type 5: Exclusion Bias

The researcher removes observations non-randomly — typically based on criteria correlated with the outcome.

**The key principle:** Any data deletion that is non-random creates selection bias. This includes removing "outliers" based on outcome values, excluding protocol violators after seeing results, or dropping countries/firms that seem "too different."

::: {.callout-tip}
## Pre-Registration as a Defense

Pre-registering exclusion criteria before data collection prevents post-hoc rationalization of exclusions that benefit your hypothesis. It does not eliminate the statistical bias, but it prevents the researcher from inadvertently introducing it.
:::
