---
title: "Propensity Scores & Matching"
---

```{r setup, include=FALSE}
library(tidyverse)
library(MatchIt)
library(broom)
library(lme4)
library(ggplot2)
library(patchwork)
set.seed(42)
```

# Propensity Scores & Matching

::: {.callout-note icon=false}
## The Selection Threat

> **What is the key problem here?** The people who enrolled in the budgeting program are not like the people who did not. They are more financially motivated, more digitally engaged, and more likely to save regardless of the program. A naive comparison confounds the program effect with the self-selection effect.
>
> **What do propensity score methods do?** They balance the treated and untreated groups on *observed* confounders, making the comparison more credible — similar to what randomization would have achieved if you could have run an experiment.
>
> **What can they never do?** Balance on *unobserved* confounders. If something that predicts both enrollment and savings outcomes is not in your data, propensity score methods cannot help. This is the method's binding limit.
:::

## The Problem PSM Solves

You want to estimate whether enrolling in a digital budgeting program increases household savings, using observational data from a financial services platform. Households self-selected into the program — financially motivated, more tech-comfortable, higher-income households are more likely to enroll. A naive comparison of enrollees vs. non-enrollees will overestimate the program's benefit.

**Propensity score methods** reduce this bias by constructing a comparison group similar to the treated group on all *observed* confounders. They work by estimating the probability that each household enrolled — the **propensity score** — then matching or weighting on that score.

$$e(X) = P(\text{Enrolled} = 1 \mid X)$$

Key result (Rosenbaum & Rubin, 1983): conditioning on $e(X)$ is equivalent to conditioning on all elements of $X$ — you do not need to match on every covariate separately.

### The selection funnel

```{mermaid}
flowchart TD
    A["All platform households\nN = 2,000"] --> B["Eligible for program\n(active accounts)\nN = 1,400"]
    B --> C["Aware of program\n(received notification)\nN = 900"]
    C --> D["Enrolled\n(treated)\nN = 180"]
    C --> E["Did not enroll\n(control pool)\nN = 720"]
    style D fill:#e8f5e9,stroke:#2d6a4f
    style E fill:#f0f0f0,stroke:#888888
```

At each stage, **who disappears is not random**. Financially savvy households stay eligible; engaged households see the notification; motivated households enroll. The enrolled group is the product of multiple successive selection filters — and the naive comparison confounds all of those filters with the program effect itself.

---

## Part A: Propensity Scores in Observational Data

### Simulation: Digital Budgeting Program

```{r ps-panel-sim}
n <- 600

financial_literacy <- rnorm(n, 50, 12)
income_percentile  <- rnorm(n, 50, 15)
baseline_savings   <- rnorm(n, 40, 10)

# Enrollment probability driven by observed confounders
p_enroll  <- plogis(-4 + 0.06 * financial_literacy + 0.04 * income_percentile
                       + 0.03 * baseline_savings)
enrolled  <- rbinom(n, 1, p_enroll)

# True ATE = 8 (program increases monthly savings by 8 units)
savings_change <- 0.2 * financial_literacy + 0.15 * income_percentile
               + 8 * enrolled + rnorm(n, 0, 7)

df_ps <- tibble(financial_literacy, income_percentile, baseline_savings,
                enrolled, savings_change)

m_naive <- lm(savings_change ~ enrolled, data = df_ps)
cat("Naive estimate:", round(coef(m_naive)["enrolled"], 2),
    "  |  True effect = 8\n")
```

The naive estimate is substantially inflated because it conflates financial literacy and income with the program effect — financially literate, higher-income households save more regardless of the program.

---

### Step 1: Estimate Propensity Scores and Check Overlap

```{r fit-ps}
ps_model <- glm(enrolled ~ financial_literacy + income_percentile + baseline_savings,
                family = binomial, data = df_ps)
df_ps <- df_ps |> mutate(ps = fitted(ps_model))
```

The overlap plot is the most important diagnostic before any matching or weighting. If the two distributions do not overlap, there are no valid comparison units for some treated households — causal inference is impossible for them.

```{r overlap-plot, fig.width=8, fig.height=4}
p_hist <- ggplot(df_ps, aes(x = ps, fill = factor(enrolled))) +
  geom_histogram(alpha = 0.65, bins = 35, position = "identity") +
  scale_fill_manual(values = c("#aec6e8", "#1e3a5f"),
                    labels = c("Did not enroll", "Enrolled"),
                    name   = NULL) +
  labs(
    title    = "Overlap: propensity score distributions",
    subtitle = "Both groups must cover the same PS range. Non-overlapping tails = no valid comparison.",
    x        = "Propensity score (P(enrolled))",
    y        = "Count"
  ) +
  theme_minimal(base_size = 13)

# Mirrored version for cleaner comparison
p_mirror <- df_ps |>
  mutate(group   = ifelse(enrolled == 1, "Enrolled", "Not enrolled"),
         ps_plot = ifelse(enrolled == 0, -ps, ps)) |>
  ggplot(aes(x = ps_plot, fill = group)) +
  geom_histogram(bins = 35, alpha = 0.8) +
  scale_fill_manual(values = c("Enrolled" = "#1e3a5f", "Not enrolled" = "#aec6e8"),
                    name = NULL) +
  scale_x_continuous(
    breaks = seq(-0.6, 0.6, 0.2),
    labels = abs(seq(-0.6, 0.6, 0.2))
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.8) +
  labs(
    title    = "Mirrored propensity score distribution",
    subtitle = "Non-enrolled below center line; enrolled above. Overlap = similar PS ranges on both sides.",
    x        = "Propensity score (mirrored for not-enrolled)",
    y        = "Count"
  ) +
  theme_minimal(base_size = 13)

p_hist / p_mirror
```

::: {.callout-warning icon=false}
## Assumption: Overlap (Common Support)

**Plain-language meaning:** Every enrolled household must have a non-trivial probability of *not* enrolling, and every non-enrolled household must have a non-trivial probability of *enrolling*. Without overlap, there are no valid comparison units for the extreme treated households — you are extrapolating, not comparing.

**What the model is assuming away:** That the enrolled and non-enrolled groups are ever comparable on the propensity score scale. If all households with high financial literacy enrolled, there are no high-literacy non-enrollees to serve as controls.

**JDM example:** If only households with incomes above $100k ever enrolled in the budgeting program, there is no valid comparison group for those households. Any estimated "effect" for them is pure extrapolation from lower-income non-enrollees who differ in many ways that matter.

**What a skeptical reviewer might say:** "Where is the common support? Show the propensity score distributions for treated and control groups — if they barely overlap, the comparison is not credible."

**Observable warning sign:** Propensity score histograms with little overlap; many enrolled households with PS > 0.8 while the control group's PS peaks below 0.3.

**What to check:** Plot mirrored PS histograms. Trim households outside the common support region (PS < min treated PS, or PS > max control PS). Report how many observations are excluded by trimming and whether results change.
:::

### Step 2: Matching and Balance Check

```{r match-balance}
m_out <- matchit(enrolled ~ financial_literacy + income_percentile + baseline_savings,
                 data   = df_ps,
                 method = "nearest",
                 ratio  = 1)

summary(m_out)$sum.matched |>
  knitr::kable(digits = 3,
    caption = "Balance after nearest-neighbor matching. Standardized mean differences should be < 0.10 after matching.")
```

The balance table shows how much the treated and matched-control groups differ on each covariate before and after matching. The goal is to reduce all standardized mean differences below 0.10 — indicating that matched groups are comparable on observed confounders.

### Step 3: PSM Treatment Effect Estimate

```{r ate-psm}
matched_data <- match.data(m_out)

m_psm <- lm(savings_change ~ enrolled + financial_literacy + income_percentile + baseline_savings,
            data    = matched_data,
            weights = weights)

tidy(m_psm) |> filter(term == "enrolled") |>
  knitr::kable(digits = 3,
    caption = "PSM estimate of ATT (true ATE = 8)")
```

### Inverse Probability Weighting (IPW)

PSM discards unmatched households, losing information. **IPW** retains all observations by reweighting so the weighted sample resembles a random draw from the full population:

$$w_i = \frac{D_i}{\hat{e}(X_i)} + \frac{1-D_i}{1-\hat{e}(X_i)}$$

Enrolled households with a low enrollment probability are upweighted (they are surprising enrollees — useful controls for the counterfactual). Non-enrolled households with a high enrollment probability are upweighted (they are surprising non-enrollees — useful controls for what would happen without the program).

```{r ipw}
df_ps <- df_ps |>
  mutate(
    ipw = ifelse(enrolled == 1,
                 1 / fitted(ps_model),
                 1 / (1 - fitted(ps_model)))
  )

m_ipw <- lm(savings_change ~ enrolled + financial_literacy + income_percentile + baseline_savings,
            data    = df_ps,
            weights = ipw)

tidy(m_ipw) |> filter(term == "enrolled") |>
  knitr::kable(digits = 3,
    caption = "IPW estimate (true effect = 8)")
```

### Doubly Robust Estimation

Doubly robust estimation combines a propensity score model with an outcome regression. The estimate is consistent if *either* model is correctly specified — providing extra protection against one of the two models being wrong.

```{r doubly-robust}
m_dr <- lm(savings_change ~ enrolled + financial_literacy + income_percentile + baseline_savings,
           data    = df_ps,
           weights = ipw)

bind_rows(
  tidy(m_naive) |> filter(term == "enrolled") |> mutate(method = "Naive OLS"),
  tidy(m_psm)   |> filter(term == "enrolled") |> mutate(method = "PSM (ATT)"),
  tidy(m_ipw)   |> filter(term == "enrolled") |> mutate(method = "IPW (ATE)"),
  tidy(m_dr)    |> filter(term == "enrolled") |> mutate(method = "Doubly Robust")
) |>
  select(method, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Comparison: naive vs. PSM vs. IPW vs. doubly robust (true = 8). In this simulated DGP, all three PS-based approaches recover close to the true effect; naive OLS is inflated.")
```

### Rosenbaum Sensitivity Analysis

After matching, the critical question is: how large would an unmeasured confounder need to be to overturn the conclusion? The sensitivity parameter $\Gamma$ quantifies this:

- $\Gamma = 1$: no hidden bias assumed.
- $\Gamma = 2$: an unmeasured confounder could double the odds of enrollment for otherwise-identical households.

A small $\Gamma$ threshold at significance (say, $\Gamma = 1.2$) means the conclusion is fragile — even a modest omitted confounder overturns it. A large threshold (say, $\Gamma = 3$) means the conclusion survives substantial unmeasured confounding. The `rbounds` package implements Rosenbaum bounds for continuous and binary outcomes. Always report the $\Gamma$ at which your p-value crosses 0.05.

---

## Part B: Propensity Scores in Experimental Research

Random assignment guarantees unbiased intent-to-treat estimates *at the point of randomization*. But three situations arise in practice where PS methods are needed even within an experiment.

### Context A: IPW for Non-Random Attrition

This is the most important experimental application. When participants drop out at follow-up and dropout is related to the outcome, the remaining sample is no longer a random subsample. High-debt households may disengage from a financial diary study precisely because their financial situation is worsening — excluding them from the analysis biases the control group's follow-up average upward.

```{r attrition-sim}
n_att <- 200
treatment_att    <- rep(c(0, 1), each = 100)
baseline_debt    <- rnorm(n_att, 50, 10)

# High debt + control group → higher dropout
# (distressed households avoid follow-up measurement)
p_remain  <- plogis(2.5 - 0.07 * baseline_debt + 0.8 * treatment_att)
remained  <- rbinom(n_att, 1, p_remain)

# True treatment effect = +6 (financial coaching increases savings)
outcome_att <- 45 + 6 * treatment_att - 0.3 * baseline_debt + rnorm(n_att, 0, 6)

df_att <- tibble(treatment = treatment_att, baseline_debt, remained, outcome_att)

cat("Control retention:  ", round(mean(remained[treatment_att == 0]) * 100, 1), "%\n")
cat("Treatment retention:", round(mean(remained[treatment_att == 1]) * 100, 1), "%\n")
```

```{r attrition-ipw}
ps_att     <- glm(remained ~ baseline_debt + treatment, family = binomial, data = df_att)
df_att$ipw_att <- 1 / fitted(ps_att)

df_complete     <- df_att |> filter(remained == 1)
m_naive_att     <- lm(outcome_att ~ treatment, data = df_complete)
m_ipw_att       <- lm(outcome_att ~ treatment, data = df_complete,
                       weights = df_att$ipw_att[df_att$remained == 1])

bind_rows(
  tidy(m_naive_att) |> filter(term == "treatment") |> mutate(method = "Naive complete-case analysis"),
  tidy(m_ipw_att)   |> filter(term == "treatment") |> mutate(method = "IPW-corrected analysis")
) |>
  select(method, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Attrition bias: naive vs. IPW-weighted (true effect = +6). High-debt control households drop out more, biasing the naive estimate downward. IPW is consistent under MAR.")
```

::: {.callout-tip icon=false}
## Reporting Attrition IPW

Report **both** estimates side by side. The naive estimate describes the effect among completers — a potentially biased, surviving subsample. The IPW estimate approximates the intent-to-treat effect for the full randomized sample. A large discrepancy signals material attrition bias.
:::

### Context B: PSM for Self-Selected Sub-Treatment Engagement

Random assignment gives households *access* to the coaching program, but actual engagement is voluntary. Comparing high-engagement to low-engagement households within the treatment arm is an **observational comparison** — highly engaged participants likely differ systematically in motivation, baseline financial literacy, and financial goals. PSM on pre-randomization covariates makes this comparison more credible.

### Context C: Pre-Test Matching for Nonlinear Covariate Relationships

ANCOVA assumes a linear, homogeneous relationship between the pre-test and post-test. When this relationship is nonlinear — for instance, when very high or very low baseline savers change at a different rate — matching on the pre-test creates a comparison group that is directly comparable at baseline, allowing nonparametric change estimation without linear extrapolation.

---

## Assumptions

::: {.callout-warning icon=false}
## Assumption: No Unmeasured Confounders (Ignorability)

**Plain-language meaning:** Given the observed covariates in the PS model, enrollment is effectively independent of the potential outcomes. There are no unmeasured variables that both predict enrollment and directly affect savings outcomes.

**What the model is assuming away:** Any confounding that operates through variables not in your data. Financial self-efficacy, financial anxiety, life-stage savings goals, and partner financial behavior could all simultaneously predict enrollment and savings outcomes — but if they are unmeasured, the PS model cannot account for them.

**JDM example:** If financially anxious households are both less likely to enroll *and* have worse savings outcomes for anxiety-related reasons (not program reasons), the PS model is missing the mechanism. Anxiety looks like a confounder but is not measured.

**Experimental analog:** In a perfectly randomized experiment, this assumption is satisfied by design. In an observational study, it is always an untestable assertion about your data-generating process.

**What a skeptical reviewer might say:** "You controlled for literacy and income — but what about financial self-efficacy? Confidence in managing money predicts both enrollment and savings outcomes independently, and you don't measure it."

**Observable warning sign:** Sensitivity analysis ($\Gamma$) showing fragility at values close to 1 — meaning even small amounts of unmeasured confounding overturn the conclusion.

**What to check:** Report the Rosenbaum $\Gamma$ or E-value. Discuss which unmeasured confounders are most plausible given domain knowledge. Consider whether any proxy measure can be added to the PS model to reduce confounding exposure.

**What claim survives if the assumption is doubtful:** The estimate is a lower bound on confounding bias — it shows what you would find if all observed confounders are balanced. The causal interpretation depends on how credible ignorability is given the specific context.
:::

::: {.callout-warning icon=false}
## Assumption: Correctly Specified Propensity Score Model

**Plain-language meaning:** The logistic regression used to estimate PS must include all relevant confounders and model their effects correctly (linear effects, or appropriate transformations and interactions if effects are nonlinear).

**What the model is assuming away:** Nonlinear confounder effects, important interaction terms, and confounders not in the model.

**JDM example:** If financial literacy's effect on enrollment probability has a threshold (below a certain level, almost no one enrolls; above it, nearly everyone does), the logistic model with a linear literacy term will misestimate the PS near the threshold.

**What to check:** Check balance *after* matching — if the PS model is correct and the matching is effective, observed confounders should be balanced between matched groups (SMD < 0.10). Poor balance despite matching signals PS model misspecification. Consider non-parametric PS estimation (CBPS or generalized boosted models) as a robustness check.
:::

---

## What Propensity Score Methods Can and Cannot Do

::: {.callout-note icon=false}
## The Limits of Matching

**What you can say after PSM/IPW:**

- Among households with similar *observed* characteristics, the enrollment contrast is less confounded by those characteristics.
- The estimate approximates what you would find in a randomized study *if* all confounders were measured and correctly modeled.
- For ATT: the effect among enrolled households, relative to a comparison group matched to look like the enrolled group on observables.

**What you cannot say:**

- That enrollment is as good as random — unobserved confounders remain and can dominate the result.
- That a large $\Gamma$ proves confounding is absent — it means you have *quantified the threshold* for overturning, not ruled out confounding.
- That balance on observed covariates implies balance on unobserved ones.

Propensity score methods shift the burden from "I controlled for X" to "I controlled for all confounders" — which is a judgment about the DGP, not a statistical guarantee. Every PS analysis should be accompanied by an explicit argument for why the observed confounders are sufficient given the institutional context.
:::

---

## Reviewer Challenge

::: {.callout-tip}
## What a Reviewer Would Ask

**Challenge 1:** "You matched on financial literacy and income — but motivation to save is the most obvious confounder and you have no measure of it."
This is the ignorability challenge. The appropriate response is: (1) acknowledge the limitation explicitly, (2) argue why the measured proxies (literacy, baseline savings) partially capture motivation, (3) conduct a sensitivity analysis showing how large an unmeasured confounder would need to be ($\Gamma$ or E-value), and (4) discuss whether the residual confounding, if present, is likely to be large enough to materially change the conclusion.

**Challenge 2:** "Your pre-matching balance table shows a large SMD for baseline savings. Why is your matching not working?"
Poor post-matching balance indicates either that the PS model is misspecified (wrong functional form, missing interactions) or that the matching ratio is too restrictive. Investigate the caliper setting, consider exact matching on the most influential covariate, or use entropy balancing which targets direct covariate balance rather than going through the PS.

**Challenge 3:** "Your overlap plot shows that enrolled households have PS mostly above 0.5, while non-enrolled are mostly below 0.2. Is the comparison credible?"
This is a common support failure. Matches in the tails are between very different households — the comparisons are not credible there. The honest response is to trim to the common support region, show results on that restricted sample, and acknowledge that the ATT estimate applies only to enrolled households with close matches, not all enrolled households.
:::

---

## Reporting Checklist for PS Analyses

::: {.callout-tip}
## PS Reporting Checklist

- [ ] Specify all covariates in the PS model — pre-treatment only; no mediators or post-treatment variables
- [ ] Report overlap check: mirrored PS histogram or plot by treatment group; note common support violations
- [ ] Report balance statistics **before AND after** matching/weighting (standardized mean difference table; target SMD < 0.10)
- [ ] Specify estimand: ATE (all households), ATT (enrolled only), or ATC (non-enrolled only)? Justify the choice
- [ ] Report sensitivity analysis ($\Gamma$ or E-value) to quantify vulnerability to unmeasured confounding
- [ ] For attrition IPW: report both naive (completer-only) and IPW-corrected estimates side by side
- [ ] For doubly robust: note which model (PS or outcome) is the primary specification and which provides the robustness guarantee
- [ ] Discuss the most plausible unmeasured confounders by name and explain why the design reduces their influence
:::
