---
title: "Propensity Scores & Matching"
---

```{r setup, include=FALSE}
library(tidyverse)
library(MatchIt)
library(broom)
set.seed(321)
```

# Propensity Scores & Matching

## The Observational Study Problem

You want to estimate whether participating in a corporate wellness program improves employee health, using observational HR data. Employees self-selected into the program — healthier, more motivated employees are more likely to participate. A naive comparison of participants vs. non-participants will overestimate the benefit.

**Propensity score methods** reduce this bias by creating a comparison group similar to the treated group on all *observed* confounders. They work by estimating the probability that each unit received treatment — the **propensity score** — then matching or weighting on that score.

## The Propensity Score

$$e(X) = P(\text{Treatment} = 1 \mid X)$$

The probability of receiving treatment given observed covariates. Key result (Rosenbaum & Rubin, 1983): conditioning on $e(X)$ is equivalent to conditioning on all of $X$ — you do not need to match on each covariate separately.

```{r propensity-setup}
n <- 600
age           <- rnorm(n, 40, 10)
health_aware  <- rnorm(n, 50, 12)
job_stress    <- rnorm(n, 45, 8)

# Treatment probability driven by confounders
p_treat   <- plogis(-4 + 0.05 * age + 0.06 * health_aware - 0.03 * job_stress)
treatment <- rbinom(n, 1, p_treat)

# True ATE = 5
health_score <- 40 + 0.3 * age + 0.4 * health_aware - 0.2 * job_stress +
                5 * treatment + rnorm(n, 0, 6)

df_ps <- tibble(age, health_aware, job_stress, treatment, health_score)

m_naive <- lm(health_score ~ treatment, data = df_ps)
cat("Naive estimate:", round(coef(m_naive)["treatment"], 2),
    "— true effect = 5\n")
```

## Step 1: Estimate Propensity Scores and Check Overlap

```{r fit-ps}
ps_model <- glm(treatment ~ age + health_aware + job_stress,
                family = binomial, data = df_ps)
df_ps <- df_ps |> mutate(ps = fitted(ps_model))

ggplot(df_ps, aes(x = ps, fill = factor(treatment))) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  scale_fill_manual(values = c("#aec6e8", "#1e3a5f"),
                    labels = c("Control", "Treated")) +
  labs(title = "Overlap: propensity score distributions",
       subtitle = "Both groups must cover the same PS range for valid comparison",
       x = "Propensity Score", y = "Count", fill = NULL) +
  theme_minimal(base_size = 13)
```

::: {.callout-warning icon=false}
## Assumption 1: Overlap (Common Support)

**What it means:** Every unit must have a propensity score strictly between 0 and 1. Units with PS near 0 or 1 have no valid comparison units — causal inference is not possible for them.

**How to check:** Plot the PS distributions for treated and control groups. Trim units outside the common support region.
:::

## Step 2: Match and Check Balance

```{r match-balance}
m_out <- matchit(treatment ~ age + health_aware + job_stress,
                 data = df_ps, method = "nearest", ratio = 1)

# Balance summary
summary(m_out)$sum.matched |>
  knitr::kable(digits = 3,
    caption = "Balance after nearest-neighbor matching (std. diff. should be near 0)")
```

::: {.callout-warning icon=false}
## Assumption 2: No Unmeasured Confounders (Ignorability)

**What it means:** Given observed covariates, treatment assignment is independent of potential outcomes. There are no unobserved variables that both predict treatment and affect the outcome.

**This is the critical untestable assumption.** Propensity score methods can only control for variables you measured. Any unobserved confounder still biases the estimate.

**How to scrutinize:** Conduct a sensitivity analysis (Rosenbaum bounds) asking: how large would an unmeasured confounder need to be to overturn your conclusion?
:::

## Step 3: Estimate the Treatment Effect

```{r ate-estimate}
matched_data <- match.data(m_out)

m_matched <- lm(health_score ~ treatment + age + health_aware + job_stress,
                data = matched_data, weights = weights)

tidy(m_matched) |> filter(term == "treatment") |>
  knitr::kable(digits = 3,
    caption = "Matched estimate of ATE (true = 5)")
```

## PSM vs. IPW vs. Doubly Robust

| Method | How it works | Key advantage | Key risk |
|--------|-------------|--------------|---------|
| PSM | Match treated to similar controls | Intuitive; easy to check balance | Discards unmatched units |
| IPW | Weight by inverse of propensity score | Uses all units | Extreme weights for PS near 0 or 1 |
| Doubly robust | Combines PS model + outcome model | Consistent if *either* model is correct | Slightly more complex |

Doubly robust estimation provides an extra layer of protection: your estimate is consistent if *either* the propensity model or the outcome model is correctly specified — not necessarily both.
