---
title: "Propensity Scores & Matching"
---

```{r setup, include=FALSE}
library(tidyverse)
library(MatchIt)
library(broom)
library(lme4)
library(ggplot2)
library(patchwork)
set.seed(42)
```

# Propensity Scores & Matching

## The Problem PSM Solves

You want to estimate whether participating in a corporate wellness program improves employee health, using observational HR data. Employees self-selected into the program — healthier, more motivated employees are more likely to participate. A naive comparison of participants vs. non-participants will overestimate the benefit.

**Propensity score methods** reduce this bias by creating a comparison group similar to the treated group on all *observed* confounders. They work by estimating the probability that each unit received treatment — the **propensity score** — then matching or weighting on that score.

$$e(X) = P(\text{Treatment} = 1 \mid X)$$

Key result (Rosenbaum & Rubin, 1983): conditioning on $e(X)$ is equivalent to conditioning on all of $X$ — you do not need to match on each covariate separately.

Propensity score methods appear in both settings covered in this course:

- **Longitudinal/panel data**: to address non-random treatment adoption across units and time.
- **Experimental data**: to correct for non-random attrition at follow-up, self-selected sub-treatment usage, or nonlinear pre-test relationships.

---

## Part A: Propensity Scores in Panel / Longitudinal Data

### Simulation: Employee Wellness Program

```{r ps-panel-sim}
n <- 600
age          <- rnorm(n, 40, 10)
health_aware <- rnorm(n, 50, 12)
job_stress   <- rnorm(n, 45, 8)

# Treatment probability driven by confounders
p_treat   <- plogis(-4 + 0.05 * age + 0.06 * health_aware - 0.03 * job_stress)
treatment <- rbinom(n, 1, p_treat)

# True ATE = 5
health_score <- 40 + 0.3 * age + 0.4 * health_aware - 0.2 * job_stress +
                5 * treatment + rnorm(n, 0, 6)

df_ps <- tibble(age, health_aware, job_stress, treatment, health_score)

m_naive <- lm(health_score ~ treatment, data = df_ps)
cat("Naive estimate:", round(coef(m_naive)["treatment"], 2),
    "  |  True effect = 5\n")
```

### Step 1: Estimate Propensity Scores and Check Overlap

```{r fit-ps}
ps_model <- glm(treatment ~ age + health_aware + job_stress,
                family = binomial, data = df_ps)
df_ps <- df_ps |> mutate(ps = fitted(ps_model))

ggplot(df_ps, aes(x = ps, fill = factor(treatment))) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  scale_fill_manual(values = c("#aec6e8", "#1e3a5f"),
                    labels = c("Control", "Treated")) +
  labs(
    title = "Overlap: propensity score distributions",
    subtitle = "Both groups must cover the same PS range for valid comparison",
    x = "Propensity Score", y = "Count", fill = NULL
  ) +
  theme_minimal(base_size = 13)
```

::: {.callout-warning icon=false}
## Assumption 1: Overlap (Common Support)

**What it means:** Every unit must have a propensity score strictly between 0 and 1. Units with PS near 0 or 1 have no valid comparison units — causal inference is not possible for them.

**How to check:** Plot the PS distributions for treated and control groups. Trim units outside the common support region if overlap is poor at the extremes.
:::

::: {.callout-warning icon=false}
## Assumption 2: No Unmeasured Confounders (Ignorability)

**What it means:** Given observed covariates, treatment assignment is independent of potential outcomes. There are no unobserved variables that both predict treatment and affect the outcome.

**This is the critical untestable assumption.** Propensity score methods can only control for variables you measured. Any unobserved confounder still biases the estimate. Conduct a sensitivity analysis after matching to quantify your exposure to this risk.
:::

### Step 2: Nearest-Neighbor Matching and Balance Check

```{r match-balance}
m_out <- matchit(treatment ~ age + health_aware + job_stress,
                 data = df_ps, method = "nearest", ratio = 1)

summary(m_out)$sum.matched |>
  knitr::kable(digits = 3,
    caption = "Balance after nearest-neighbor matching (std. diff. should be < 0.10)")
```

### Step 3: PSM Treatment Effect Estimate

```{r ate-psm}
matched_data <- match.data(m_out)

m_psm <- lm(health_score ~ treatment + age + health_aware + job_stress,
            data = matched_data, weights = weights)

tidy(m_psm) |> filter(term == "treatment") |>
  knitr::kable(digits = 3,
    caption = "PSM estimate of ATT (true effect = 5)")
```

### Inverse Probability Weighting (IPW)

PSM discards unmatched units, which loses information. **IPW** uses all units by reweighting observations so that the weighted sample looks like a random sample from the full population:

$$w_i = \frac{D_i}{\hat{e}(X_i)} + \frac{1-D_i}{1-\hat{e}(X_i)}$$

Treated units are upweighted when their treatment probability was low (they are "surprising" treated units); control units are upweighted when their treatment probability was high (they are "surprising" controls). After reweighting, covariates should be balanced between groups.

```{r ipw}
df_ps <- df_ps |>
  mutate(
    ipw = ifelse(treatment == 1,
                 1 / fitted(ps_model),
                 1 / (1 - fitted(ps_model)))
  )

m_ipw <- lm(health_score ~ treatment + age + health_aware + job_stress,
            data = df_ps, weights = ipw)

tidy(m_ipw) |> filter(term == "treatment") |>
  knitr::kable(digits = 3,
    caption = "IPW estimate (true effect = 5)")
```

IPW is generally preferred over PSM in panel data settings where you need all time points for each unit. Dropping unmatched observations in PSM can create unbalanced panels — some units contribute to some waves but not others — which complicates longitudinal estimation.

### Dynamic Treatment in Panel Data

In panel data, treatment adoption timing varies across units. The propensity score should account for *when* treatment was adopted — not just whether it was adopted.

At each wave, among units not yet treated, compute the probability of adopting treatment in this wave. Match recently-adopting units to not-yet-treated units with similar adoption probabilities at the time of adoption.

```{r ps-dynamic, eval=FALSE}
# Time-varying propensity score: probability of adopting in wave w,
# conditional on not yet having adopted
ps_wave <- glm(
  adopted_this_wave ~ lag_outcome + age + job_stress,
  family = binomial,
  data   = df_panel |> filter(wave == w, !previously_treated)
)
```

This approach is the foundation of **marginal structural models**, which chain together wave-specific IPW weights to estimate the effect of a treatment *history* rather than a single binary treatment.

### Doubly Robust Estimation

Doubly robust combines a propensity score model with an outcome model. The estimate is consistent if *either* model is correctly specified — providing an extra layer of protection against misspecification.

```{r doubly-robust}
# Doubly robust: IPW weights + regression adjustment on covariates
m_dr <- lm(health_score ~ treatment + age + health_aware + job_stress,
           data = df_ps, weights = ipw)

tidy(m_dr) |> filter(term == "treatment") |>
  knitr::kable(digits = 3,
    caption = "Doubly robust estimate (true effect = 5)")
```

```{r comparison-table}
# Side-by-side comparison of all four approaches
bind_rows(
  tidy(m_naive) |> filter(term == "treatment") |> mutate(method = "Naive OLS"),
  tidy(m_psm)   |> filter(term == "treatment") |> mutate(method = "PSM (ATT)"),
  tidy(m_ipw)   |> filter(term == "treatment") |> mutate(method = "IPW (ATE)"),
  tidy(m_dr)    |> filter(term == "treatment") |> mutate(method = "Doubly Robust")
) |>
  select(method, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Comparison: naive OLS vs. PSM vs. IPW vs. doubly robust (true = 5)")
```

### Rosenbaum Sensitivity Analysis

After PSM, the critical question is: how large would an unmeasured confounder need to be to overturn the conclusion? The sensitivity parameter $\Gamma$ (gamma) quantifies this.

- $\Gamma = 1$: no hidden bias assumed — two units with the same observed covariates have the same probability of treatment.
- $\Gamma = 2$: an unmeasured confounder could double the odds of treatment for otherwise-identical units.

A small $\Gamma$ threshold (say, $\Gamma = 1.2$) means the conclusion is fragile; a large threshold (say, $\Gamma = 3$) means the conclusion is robust to substantial unobserved confounding. The `rbounds` package implements Rosenbaum bounds for continuous and binary outcomes. Report the $\Gamma$ threshold at which the p-value crosses 0.05.

---

## Part B: Propensity Scores in Experimental Research

Random assignment guarantees unbiased intent-to-treat estimates *at baseline*. But three situations arise in practice where propensity score methods are needed even within an experiment.

### Context A: IPW for Non-Random Attrition

This is the most important experimental application of IPW. Random assignment guarantees an unbiased ITT estimate at baseline — but when participants drop out at follow-up and dropout is related to the outcome, the remaining sample is no longer a random subsample. The naive estimate among completers is biased.

**IPW corrects for this** by upweighting participants who were similar to drop-outs (by observed baseline characteristics) but happened to remain. The attrition propensity score is modelled as a function of baseline covariates and treatment assignment; each completer is then weighted by the inverse of their probability of remaining.

```{r attrition-sim}
n_att <- 200
set.seed(42)

treatment_att   <- rep(c(0, 1), each = 100)
baseline_anxiety <- rnorm(n_att, 50, 10)

# High anxiety + being in control group → higher dropout
p_remain <- plogis(2 - 0.04 * baseline_anxiety + 0.4 * treatment_att)
remained  <- rbinom(n_att, 1, p_remain)

# True treatment effect = -5
outcome <- 40 - 5 * treatment_att + 0.3 * baseline_anxiety + rnorm(n_att, 0, 6)

df_att <- tibble(treatment = treatment_att, baseline_anxiety, remained, outcome)

cat("Overall retention rate:", round(mean(df_att$remained) * 100, 1), "%\n")
cat("Control retention:     ", round(mean(df_att$remained[df_att$treatment == 0]) * 100, 1), "%\n")
cat("Treatment retention:   ", round(mean(df_att$remained[df_att$treatment == 1]) * 100, 1), "%\n")
```

```{r attrition-ipw}
# Attrition propensity model
ps_att    <- glm(remained ~ baseline_anxiety + treatment,
                 family = binomial, data = df_att)
df_att$ipw_att <- 1 / fitted(ps_att)

# Complete the analysis on retained participants
df_complete <- df_att |> filter(remained == 1)

m_naive_att <- lm(outcome ~ treatment, data = df_complete)
m_ipw_att   <- lm(outcome ~ treatment, data = df_complete,
                  weights = df_att$ipw_att[df_att$remained == 1])

bind_rows(
  tidy(m_naive_att) |> filter(term == "treatment") |> mutate(method = "Naive (completers only)"),
  tidy(m_ipw_att)   |> filter(term == "treatment") |> mutate(method = "IPW-corrected")
) |>
  select(method, estimate, std.error, p.value) |>
  mutate(true_effect = -5) |>
  knitr::kable(digits = 3,
    caption = "Attrition bias: naive vs. IPW-corrected (true effect = -5)")
```

::: {.callout-tip icon=false}
## Reporting Attrition IPW

Report **both** the naive (completer-only) and IPW-corrected estimates. They answer different questions:

- The naive estimate describes the treatment effect **among completers** — potentially a biased, self-selected subsample.
- The IPW estimate approximates the intent-to-treat effect **for the full randomized sample**, including those who would have dropped out.

A large discrepancy between the two signals that differential attrition is materially affecting your conclusions.
:::

```{r attrition-plot}
# Visualize: who drops out by treatment group and anxiety level
p_dropout <- ggplot(df_att, aes(x = baseline_anxiety,
                                 y = remained,
                                 color = factor(treatment))) +
  geom_jitter(height = 0.03, alpha = 0.5, size = 1.5) +
  geom_smooth(method = "glm",
              method.args = list(family = binomial),
              se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = c("#aec6e8", "#1e3a5f"),
                     labels = c("Control", "Treatment")) +
  labs(
    title = "Retention probability by baseline anxiety and treatment",
    subtitle = "High-anxiety controls are most likely to drop out — biasing naive estimate upward",
    x = "Baseline Anxiety", y = "P(Retained at follow-up)", color = NULL
  ) +
  theme_minimal(base_size = 13)

p_dropout
```

### Context B: PSM for Self-Selected Sub-Treatments

In field experiments, random assignment gives participants *access* to an intervention, but actual usage is voluntary. Comparing high-users to non-users within the treatment arm of an experiment is an **observational comparison** — high-users likely differ systematically from non-users on motivation, baseline severity, and other factors.

PSM on baseline covariates makes this comparison more credible by creating a matched non-user group similar to high-users on observed characteristics. This provides an estimate of the effect of intensive engagement, net of observable self-selection.

```{r psm-subtreat, eval=FALSE}
# Within the intervention arm: match high-users to low-users
# on pre-randomization covariates
m_sub <- matchit(
  high_user ~ pre_score + age + motivation + baseline_severity,
  data   = intervention_arm,
  method = "nearest",
  ratio  = 1
)

# Estimate effect of high engagement vs. low engagement
m_usage <- lm(outcome ~ high_user + pre_score + age + motivation,
              data = match.data(m_sub), weights = weights)
```

This design is sometimes called a **broken randomization** analysis or a within-arm observational study. The critical caveat: ignorability holds only if all determinants of usage that also affect outcomes are measured and included in the PSM model. Unmeasured motivation or severity remains a threat.

### Context C: Matching on Pre-Test Score

When the goal is to compare the *change* from pre to post (rather than just the post-test score), matching on the pre-test score before estimating treatment effects can be more robust than ANCOVA when the pre-test — outcome relationship is nonlinear.

ANCOVA assumes a linear, homogeneous (equal across groups) relationship between the pre-test and post-test. When this assumption is violated — for instance, when very high or very low pre-test scorers change at a different rate — matching on the pre-test creates a comparison group that is directly comparable at the pre-test level, allowing nonparametric estimation of the change. This is sometimes called **pre-test matching** or **coarsened exact matching on the pre-test**.

---

## Reporting Checklist for PS Analyses

::: {.callout-tip}
## PS Reporting Checklist
- [ ] Specify all covariates included in the PS model and justify their inclusion (pre-treatment only; no mediators)
- [ ] Report overlap check (PS histogram by treatment group; note any violations of common support)
- [ ] Report balance statistics **before AND after** matching/weighting (standardised mean difference table; target SMD < 0.10)
- [ ] Specify estimand: ATE, ATT, or ATC? And why does it match the research question?
- [ ] Report sensitivity analysis (Rosenbaum $\Gamma$ or E-value) to quantify vulnerability to unmeasured confounding
- [ ] For experimental attrition IPW: report both the naive (completer-only) and IPW-corrected estimates and note any discrepancy
- [ ] For doubly robust estimation: note which model (PS or outcome) is the primary specification and which provides the robustness guarantee
:::
