---
title: "Difference-in-Differences"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(broom)
library(broom.mixed)
library(lme4)
library(lmerTest)
library(patchwork)
library(sandwich)
library(lmtest)
set.seed(42)
```

# Difference-in-Differences

::: {.callout-note icon=false}
## The Counterfactual Question

> **What would the treated group's outcome have been if treatment had never happened?** You can't observe it. DiD borrows the control group's change as a stand-in.
>
> **What needs to be true for that substitution to work?** The two groups must have been on parallel trajectories before treatment — parallel trends. If treated units were already diverging before the policy, the control group's path is a poor proxy.
>
> **Where does this estimator live?** In panel data (firms, households, branches tracked across time with staggered or simultaneous policy rollout) and in experimental pre-post designs (pre-test → treatment → post-test for randomized vs. control conditions). Same estimator, same critical assumption.
:::

## The Core Idea

A savings nudge rolls out at some bank branches; others serve as controls. You have monthly transaction data from before and after. Can you estimate the nudge's effect on savings behavior?

The naive approach — compare treated and control branches *after* the rollout — is biased if they were already different. **Difference-in-Differences (DiD)** subtracts the control group's change from the treated group's change:

$$\hat{\tau}_{DiD} = \underbrace{(\bar{Y}_{T,\text{post}} - \bar{Y}_{T,\text{pre}})}_{\text{treated group change}} - \underbrace{(\bar{Y}_{C,\text{post}} - \bar{Y}_{C,\text{pre}})}_{\text{control group change}}$$

The control group's change serves as the counterfactual: what would have happened to treated units in the absence of the nudge?

### The counterfactual timeline

```{mermaid}
flowchart LR
    A["Pre-period\n(baseline)"] --> B["Treatment onset\n(policy rollout)"]
    B --> C["Post-period\n(observed)"]
    B --> D["Counterfactual\n(unobserved)"]
    style D fill:#fff3cd,stroke:#f39c12,stroke-dasharray:5 5
    style B fill:#e8f5e9,stroke:#2d6a4f
```

The dashed box is the missing data problem. The treated group's counterfactual post-period outcome — what they would have done without the nudge — is never observed. DiD fills that gap by assuming the control group's trajectory is a valid proxy.

DiD appears in two distinct research contexts that share this same logic:

- **Longitudinal panel data**: households, consumers, or bank branches tracked across calendar time, with a policy affecting some units.
- **Repeated-measures experimental data**: participants randomized to treatment vs. control and measured both before and after.

---

## Part A: Panel DiD

### Setup and Simulation

We simulate a savings nudge rolled out at 40 of 80 bank branches over 6 waves (waves 1–3: pre; 4–6: post). The true nudge effect is a +10 unit increase in average monthly savings. Branches differ in their baseline savings rates and all follow a common upward time trend.

```{r did-panel-sim}
n_branches <- 80
n_waves    <- 6  # waves 1–3 pre; 4–6 post

branch_treat <- c(rep(1, 40), rep(0, 40))
branch_base  <- rnorm(n_branches, 50, 8)

df_did <- expand_grid(branch = 1:n_branches, wave = 1:n_waves) |>
  mutate(
    treat          = branch_treat[branch],
    post           = as.integer(wave > 3),
    baseline       = branch_base[branch],
    trend          = (wave - 3.5) * 1.5,
    te             = treat * post * 10,
    avg_savings    = baseline + trend + te + rnorm(n(), 0, 4)
  )
```

### The Core Assumption: Parallel Trends

::: {.callout-warning icon=false}
## Assumption: Parallel Trends

**Plain-language meaning:** In the absence of the nudge, treated branches would have followed the same savings trajectory as control branches. They can start at different levels — DiD accounts for that with fixed effects. But their *rate of change over time* must be the same.

**What the model is assuming away:** Any pre-existing divergence in trends — treated branches might have been on a steeper growth trajectory already, making the post-treatment gap look like a nudge effect when it would have appeared anyway.

**JDM example:** If a bank selectively rolled out the savings nudge at high-momentum branches (those already gaining customers and growing savings), treated branches were on a faster pre-trend. The post-period gap partly reflects that pre-existing momentum, not just the nudge.

**Experimental analog:** In a randomized pre-post experiment, parallel trends is guaranteed in expectation by random assignment. In panel observational data, it is an assumption you must argue and test.

**What a skeptical reviewer might say:** "Why should the treatment and control branches have had the same underlying savings growth trajectory if the rollout was not random? Were higher-growth branches systematically chosen for the nudge?"

**Observable warning sign:** Treated and control group trends visibly diverging *before* the treatment onset in a raw data plot; significant treatment-period interactions in pre-treatment-only regressions.

**What to check:**
1. Plot pre-treatment trends for both groups — they should look parallel.
2. Regress outcome on `treat × wave` using *only pre-treatment observations* — significant interactions indicate pre-trend differences.
3. Estimate an event study (period-specific treatment effects) — pre-period coefficients should be statistically indistinguishable from zero.

**What claim survives if the assumption is doubtful:** The sign of the effect may still be correct if the pre-trend divergence is modest. But the magnitude is unreliable, and the direction can reverse if treated units were on a faster upward trend that would have created a post-period gap regardless.
:::

```{r parallel-trends-illustration, echo=FALSE, fig.width=10, fig.height=4.2, message=FALSE, warning=FALSE}
# Conceptual illustration: parallel trends holds vs. fails
# Two separate plots combined with patchwork (avoids facet_wrap gtable issue in ggplot2 4.x)
t_pts <- 0:5
cut_t <- 2.5

col_vals <- c("Treated" = "#1e3a5f", "Control" = "#aec6e8", "Counterfactual" = "#e67e00")
lty_vals <- c("Treated" = "solid",   "Control" = "solid",   "Counterfactual" = "dashed")

# --- Left panel: parallel trends holds ---
df_h <- bind_rows(
  tibble(t = t_pts, group = "Treated",
         y = 50 + 2 * t_pts + ifelse(t_pts > cut_t, 10, 0)),
  tibble(t = t_pts, group = "Control",
         y = 40 + 2 * t_pts)
)
df_hc <- tibble(t = t_pts[t_pts > cut_t], group = "Counterfactual",
                y = 50 + 2 * t_pts[t_pts > cut_t])

p_holds <- ggplot() +
  geom_vline(xintercept = cut_t, linetype = "dotted", color = "gray50", linewidth = 1.0) +
  geom_line(data = df_h,  aes(x = t, y = y, color = group, linetype = group), linewidth = 1.3) +
  geom_line(data = df_hc, aes(x = t, y = y, color = group, linetype = group), linewidth = 1.0) +
  annotate("text", x = cut_t + 0.1, y = 43, label = "Treatment\nonset",
           hjust = 0, size = 2.9, color = "gray40") +
  scale_color_manual(values = col_vals) +
  scale_linetype_manual(values = lty_vals) +
  labs(title = "Parallel trends HOLDS (valid DiD)",
       subtitle = "Pre-slopes equal — DiD captures only the jump",
       x = "Time period", y = "Outcome", color = NULL, linetype = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", size = 10))

# --- Right panel: parallel trends fails ---
df_f <- bind_rows(
  tibble(t = t_pts, group = "Treated",
         y = 50 + 4 * t_pts + ifelse(t_pts > cut_t, 4, 0)),
  tibble(t = t_pts, group = "Control",
         y = 40 + 2 * t_pts)
)
df_fc <- tibble(t = t_pts[t_pts > cut_t], group = "Counterfactual",
                y = 50 + 4 * t_pts[t_pts > cut_t])

p_fails <- ggplot() +
  geom_vline(xintercept = cut_t, linetype = "dotted", color = "gray50", linewidth = 1.0) +
  geom_line(data = df_f,  aes(x = t, y = y, color = group, linetype = group), linewidth = 1.3) +
  geom_line(data = df_fc, aes(x = t, y = y, color = group, linetype = group), linewidth = 1.0) +
  annotate("text", x = cut_t + 0.1, y = 43, label = "Treatment\nonset",
           hjust = 0, size = 2.9, color = "gray40") +
  scale_color_manual(values = col_vals) +
  scale_linetype_manual(values = lty_vals) +
  labs(title = "Parallel trends FAILS (invalid DiD)",
       subtitle = "Treated group already on steeper trajectory — DiD conflates\npolicy effect with pre-existing momentum",
       x = "Time period", y = "Outcome", color = NULL, linetype = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", size = 10))

p_holds + p_fails + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

In the left panel, both groups rise at the same rate before treatment — the DiD estimate cleanly captures only the jump caused by the policy. In the right panel, treated units were already on a steeper trajectory. The DiD estimate would be inflated because part of the post-period gap would have materialised even without the nudge. The pre-trend test and event study distinguish these two situations.

```{r did-panel-plot, fig.width=8, fig.height=4.5}
group_means <- df_did |>
  group_by(wave, treat) |>
  summarise(mean_sav = mean(avg_savings), .groups = "drop") |>
  mutate(group = ifelse(treat == 1, "Nudge branches", "Control branches"))

ggplot(group_means, aes(x = wave, y = mean_sav, color = group, linetype = group)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3) +
  geom_vline(xintercept = 3.5, linetype = "dotted", color = "gray40") +
  annotate("text", x = 3.6, y = max(group_means$mean_sav) - 1,
           label = "Nudge rollout", hjust = 0, color = "gray40", size = 3.5) +
  scale_color_manual(values = c("#aec6e8", "#1e3a5f")) +
  labs(
    title    = "DiD: parallel pre-trends, divergence post-rollout",
    subtitle = "Parallel pre-period trends support the counterfactual substitution; post-period gap = DiD estimate",
    x        = "Wave",
    y        = "Mean monthly savings (branch average)",
    color    = NULL, linetype = NULL
  ) +
  theme_minimal(base_size = 13)
```

### Regression Implementation: Two-Way Fixed Effects

The regression implementation of DiD is the **Two-Way Fixed Effects (TWFE)** model:

$$Y_{it} = \underbrace{\alpha_i}_{\text{branch FE}} + \underbrace{\lambda_t}_{\text{time FE}} + \underbrace{\delta \cdot D_{it}}_{\text{nudge effect}} + \varepsilon_{it}$$

Branch fixed effects $\alpha_i$ absorb all time-invariant differences across branches (e.g., branch size, local income demographics). Time fixed effects $\lambda_t$ absorb all common temporal trends (e.g., macroeconomic changes affecting all branches). The treatment indicator $D_{it}$ captures the DiD estimate net of both.

In the 2×2 case (one treated group, one pre and one post period), `treat:post` in an OLS interaction model is algebraically identical to the DiD estimator. With multiple periods and multiple units, the TWFE formulation generalises naturally.

```{r did-panel-lm}
m_did_fe <- lm(avg_savings ~ treat * post + factor(branch) + factor(wave),
               data = df_did)

tidy(m_did_fe) |>
  filter(term == "treat:post") |>
  knitr::kable(digits = 3,
    caption = "`treat:post` = DiD estimate via TWFE (true effect = 10)")
```

### Clustered Standard Errors

In DiD with panel data, residuals within the same branch are correlated across waves — the same branch's errors co-vary over time. Ignoring this inflates precision. Cluster at the branch level:

```{r clustered-se}
m_did_basic <- lm(avg_savings ~ treat * post + baseline, data = df_did)

coeftest(m_did_basic, vcov = vcovCL(m_did_basic, cluster = ~branch)) |>
  tidy() |>
  filter(term == "treat:post") |>
  knitr::kable(digits = 3,
    caption = "DiD with cluster-robust SEs (clustered by branch)")
```

### Pre-Trend Test

```{r pretrend-test}
m_pre <- lm(avg_savings ~ treat * factor(wave) + baseline,
            data = df_did |> filter(post == 0))

tidy(m_pre) |>
  filter(str_detect(term, "treat:")) |>
  knitr::kable(digits = 3,
    caption = "Pre-period treat × wave interactions — should be near zero if parallel trends holds")
```

### Event Study Plot

An event study estimates a separate treatment effect for each period, omitting one reference pre-period. Pre-period coefficients near zero confirm parallel trends; post-period coefficients trace the dynamic effect of the nudge.

```{r event-study, fig.width=8, fig.height=4.5}
m_ev  <- lm(avg_savings ~ treat * factor(wave) + baseline, data = df_did)
ev_df <- tidy(m_ev, conf.int = TRUE) |>
  filter(str_detect(term, "treat:factor")) |>
  mutate(
    wave = as.integer(str_extract(term, "[0-9]+")),
    pre  = wave <= 3
  )

ggplot(ev_df, aes(x = wave, y = estimate, color = pre)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_vline(xintercept = 3.5, linetype = "dotted", color = "gray40") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("TRUE" = "#aec6e8", "FALSE" = "#1e3a5f"),
    labels = c("TRUE"  = "Pre-treatment (should ≈ 0)",
               "FALSE" = "Post-treatment (nudge effect)")
  ) +
  labs(
    title    = "Event study: pre-period near zero, post-period shows nudge effect",
    subtitle = "Pre-period balance confirms parallel trends; post-period estimates trace effect trajectory",
    x        = "Wave",
    y        = "Estimated treatment effect (vs. wave 1 reference)",
    color    = NULL
  ) +
  theme_minimal(base_size = 13)
```

### A Note on Staggered Adoption

When different branches (or households) adopt treatment at different calendar times, the standard TWFE estimator produces a weighted average of all pairwise DiD comparisons — including "forbidden comparisons" where already-treated units serve as controls for later-treated units. If nudge effects are heterogeneous across cohorts or build up over time, this weighting can produce a biased and even sign-reversed estimate.

::: {.callout-warning icon=false}
## Staggered DiD: A Critical Limitation of Plain TWFE

Callaway & Sant'Anna (2021) and related estimators address this by computing group-time average treatment effects $ATT(g, t)$ for each cohort $g$ (defined by adoption timing) at each post-treatment period $t$, then aggregating appropriately. If you have staggered adoption, check whether the `did` or `staggered` packages are appropriate before relying on plain TWFE. See Roth et al. (2023) for a comprehensive review.

**JDM context:** If a financial education program rolls out to different consumer cohorts at different times (e.g., some households enrolled in January, others in April), and the program's effect builds over time, plain TWFE will produce a distorted average. The Callaway–Sant'Anna approach estimates the cohort-specific effect at each relative time period, then aggregates.
:::

---

## Part B: Experimental DiD

### DiD Logic in Pre-Post Controlled Experiments

DiD logic appears directly in experimental designs when you have both a **between-subjects factor** (treatment vs. control) and a **within-subjects time factor** (baseline vs. follow-up). This is the standard pre-post controlled experiment — and it is structurally a DiD.

$$\hat{\tau}_{DiD} = \underbrace{(\bar{Y}_{T,\text{post}} - \bar{Y}_{T,\text{pre}})}_{\text{treatment group change}} - \underbrace{(\bar{Y}_{C,\text{post}} - \bar{Y}_{C,\text{pre}})}_{\text{control group change}}$$

In a randomized experiment, pre-test differences are zero **in expectation**. Random assignment supports parallel trends. The DiD estimate is therefore an unbiased ATE.

Adding the pre-test makes the estimate **more efficient** than a post-only comparison: the participant random intercept absorbs baseline heterogeneity, reducing residual variance. This matters most in smaller samples where chance baseline imbalance is nontrivial.

### The Experimental DiD as a Mixed Model

Pre-post data are nested: measurements within participants within conditions. The correct model includes a random intercept for participant:

$$Y_{ij} = \gamma_{00} + \gamma_{10} \cdot \text{time}_{ij} + \gamma_{01} \cdot \text{cond}_j + \underbrace{\gamma_{11} \cdot \text{time}_{ij} \times \text{cond}_j}_{\text{DiD estimate}} + u_{0j} + \varepsilon_{ij}$$

The random effect $u_{0j}$ absorbs between-person variance in the outcome level, just as unit fixed effects do in the panel TWFE model. Both absorb stable individual heterogeneity; the DiD interaction identifies the treatment effect net of that heterogeneity.

### Worked Example: Savings Disclosure Intervention

120 participants (60 disclosure, 60 standard) are measured at **baseline** and at an **8-week follow-up**. The outcome is `wtp` (willingness to invest in a recommended product, 0–100 scale). The disclosure condition receives an expected-value breakdown of investment options; the standard condition receives the usual product description. The true DiD effect is +8: disclosure increases WTP 8 points more than the standard condition.

```{r disclosure-sim}
n_per <- 60

df_disc <- tibble(
  participant = rep(1:(n_per * 2), each = 2),
  condition   = rep(c("standard", "disclosure"), each = n_per * 2),
  wave        = rep(c("pre", "post"), times = n_per * 2),
  u0          = rep(rnorm(n_per * 2, 0, 8), each = 2)
) |>
  mutate(
    time_num  = as.integer(wave == "post"),
    treat_num = as.integer(condition == "disclosure"),
    wtp       = 50 + u0 + 1 * time_num + 8 * time_num * treat_num + rnorm(n(), 0, 5),
    participant = factor(participant),
    wave        = factor(wave, levels = c("pre", "post")),
    condition   = factor(condition, levels = c("standard", "disclosure"))
  )
```

```{r disclosure-model}
m_exp_did <- lmer(wtp ~ wave * condition + (1 | participant),
                  data = df_disc)

tidy(m_exp_did, effects = "fixed") |>
  knitr::kable(digits = 3,
    caption = "`wavepost:conditiondisclosure` = DiD estimate (true = +8)")
```

```{r disclosure-plot, fig.width=9, fig.height=4.5}
group_means_exp <- df_disc |>
  group_by(wave, condition) |>
  summarise(mean_wtp = mean(wtp), se = sd(wtp) / sqrt(n()), .groups = "drop")

p_means <- ggplot(group_means_exp,
                  aes(x = wave, y = mean_wtp, color = condition, group = condition)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_wtp - se, ymax = mean_wtp + se), width = 0.1) +
  annotate("segment",
           x    = 2, xend = 2,
           y    = group_means_exp$mean_wtp[group_means_exp$wave == "post" &
                                            group_means_exp$condition == "standard"],
           yend = group_means_exp$mean_wtp[group_means_exp$wave == "post" &
                                            group_means_exp$condition == "disclosure"],
           arrow = arrow(ends = "both", length = unit(0.1, "inches")),
           color = "#c0392b", linewidth = 0.8) +
  annotate("text", x = 2.08,
           y = mean(c(
             group_means_exp$mean_wtp[group_means_exp$wave == "post" &
                                       group_means_exp$condition == "standard"],
             group_means_exp$mean_wtp[group_means_exp$wave == "post" &
                                       group_means_exp$condition == "disclosure"]
           )),
           label = "DiD ≈ +8", hjust = 0, color = "#c0392b", size = 3.5) +
  scale_color_manual(values = c("standard" = "#aec6e8", "disclosure" = "#1e3a5f")) +
  labs(
    title    = "(a) Group means: WTP by wave and condition",
    subtitle = "Parallel pre-period (randomization); divergent post-period (disclosure effect)",
    x        = "Wave", y = "Mean WTP (0–100)", color = "Condition"
  ) +
  theme_minimal(base_size = 12)

set.seed(42)
sampled_ids <- df_disc |>
  distinct(participant, condition) |>
  group_by(condition) |>
  slice_sample(n = 30) |>
  pull(participant)

p_indiv <- df_disc |>
  filter(participant %in% sampled_ids) |>
  ggplot(aes(x = wave, y = wtp, group = participant, color = condition)) +
  geom_line(alpha = 0.35, linewidth = 0.5) +
  stat_summary(aes(group = condition), fun = mean,
               geom = "line", linewidth = 1.5) +
  scale_color_manual(values = c("standard" = "#aec6e8", "disclosure" = "#1e3a5f")) +
  labs(
    title    = "(b) Individual trajectories (sample)",
    subtitle = "Thin lines = participants; thick = group mean",
    x = "Wave", y = "WTP", color = "Condition"
  ) +
  theme_minimal(base_size = 12)

p_means + p_indiv
```

### Comparing Estimation Strategies

```{r compare-approaches}
m_post_only <- lm(wtp ~ condition,
                  data = df_disc |> filter(wave == "post"))

df_wide <- df_disc |>
  select(participant, condition, wave, wtp) |>
  pivot_wider(names_from = wave, values_from = wtp) |>
  rename(wtp_pre = pre, wtp_post = post)

m_ancova <- lm(wtp_post ~ condition + wtp_pre, data = df_wide)

bind_rows(
  tidy(m_post_only)       |> filter(term == "conditiondisclosure") |>
    mutate(model = "Post-only"),
  tidy(m_ancova)          |> filter(term == "conditiondisclosure") |>
    mutate(model = "ANCOVA"),
  tidy(m_exp_did, effects = "fixed") |>
    filter(str_detect(term, "wave.*condition|condition.*wave")) |>
    mutate(model = "Mixed DiD")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Estimation strategies for pre-post experimental data (true effect = +8). ANCOVA and Mixed DiD converge; post-only is unbiased but wider.")
```

### Experimental Event Study: Multiple Waves

With three or more measurement waves, you can estimate separate wave-specific effects — the experimental analog of the panel event study. Pre-intervention coefficients should be near zero (confirming randomization succeeded); post-intervention estimates trace the trajectory of the effect and whether it persists or fades.

```{r exp-event-study, fig.width=8, fig.height=4}
df_long <- tibble(
  participant = rep(1:(n_per * 2), each = 4),
  condition   = rep(c("standard", "disclosure"), each = n_per * 2 * 2),
  wave        = rep(c("pre", "mid", "post", "followup"), times = n_per * 2),
  u0          = rep(rnorm(n_per * 2, 0, 8), each = 4)
) |>
  mutate(
    wave_num  = case_when(wave == "pre" ~ 0, wave == "mid" ~ 1,
                          wave == "post" ~ 2, wave == "followup" ~ 3),
    treat_num = as.integer(condition == "disclosure"),
    true_did  = case_when(wave == "pre" ~ 0, wave == "mid" ~ 4,
                          wave == "post" ~ 8, wave == "followup" ~ 5),
    wtp         = 50 + u0 + 1 * wave_num + true_did * treat_num + rnorm(n(), 0, 5),
    wave        = factor(wave, levels = c("pre", "mid", "post", "followup")),
    participant = factor(participant),
    condition   = factor(condition, levels = c("standard", "disclosure"))
  )

m_ev_exp <- lmer(wtp ~ wave * condition + (1 | participant), data = df_long)

ev_exp <- tidy(m_ev_exp, effects = "fixed", conf.int = TRUE) |>
  filter(str_detect(term, "wave.*condition")) |>
  mutate(
    wave = str_extract(term, "mid|post|followup"),
    wave = factor(wave, levels = c("mid", "post", "followup"))
  )

ggplot(ev_exp, aes(x = wave, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, color = "#1e3a5f") +
  geom_point(size = 3, color = "#1e3a5f") +
  labs(
    title    = "Experimental event study: disclosure effect at each post-wave",
    subtitle = "Pre-period reference = 0 by construction; wave 'pre' is the omitted category",
    x        = "Wave (relative to pre-period)",
    y        = "DiD estimate (differential change vs. pre)"
  ) +
  theme_minimal(base_size = 13)
```

The effect builds from mid-intervention, peaks at post, and partially fades at follow-up — consistent with an information effect that decays without reinforcement. This trajectory cannot be detected with a simple two-wave pre-post design.

---

## Shared Assumptions

::: {.callout-warning icon=false}
## Assumption: SUTVA — No Spillovers

**Plain-language meaning:** Treatment of unit A should not affect the outcomes of unit B. If nudged branches compete with control branches for the same customer base, improvements at nudged branches may come partly at control branches' expense — contaminating the control trend.

**JDM example:** A financial education program at some bank branches may cause customers to switch from non-program branches, raising the program-branch savings rate partly through selection rather than education.

**Experimental analog:** In within-subjects experiments, SUTVA violation is called **carryover effects**: completing the disclosure condition may make participants more financially attentive in later trials, including control trials. This creates an asymmetry that the DiD estimate cannot separate from the disclosure effect.

**What a skeptical reviewer might say:** "Is there any reason to believe that treated branches and control branches are drawing from the same local customer pool? If so, the control group trend is contaminated."

**How to manage (experimental):** Randomize condition order; include filler tasks between conditions; use between-subjects designs when carryover is likely.

**How to check (panel):** Test whether control units near treated units trend differently from geographically distant controls.
:::

::: {.callout-warning icon=false}
## Assumption: Stable Group Composition

**Plain-language meaning:** The composition of treated and control groups should not change differentially after treatment. If the savings nudge causes higher-balance customers to switch into nudged branches (because they value the service), the post-period treated group is a different population.

**JDM example:** After a fee waiver rolls out to some branches, financially sophisticated customers may actively seek out those branches. Post-period comparisons then partly capture customer self-selection, not just the fee waiver effect.

**Experimental analog:** Differential attrition at follow-up — if financially distressed participants in the disclosure condition drop out before the post-test, the remaining disclosure group is a non-representative survivor sample.

**What to check:** Test whether observable baseline characteristics predict differential attrition or switching behavior. If they do, consider IPW to correct for non-random compositional change (see Part 6).
:::

---

## Reviewer Challenge

::: {.callout-tip}
## What a Reviewer Would Ask

**Challenge 1:** "How do you know the branches would have followed parallel trends without the nudge?"
The parallel trends assumption is untestable for the post-period — you never observe the treated group's counterfactual. The pre-period test (event study coefficients near zero before treatment) provides necessary but not sufficient evidence. A reviewer will want to know whether the rollout was random or selective, and whether treated and control branches look systematically different in observable characteristics.

**Challenge 2:** "Aren't you using already-treated branches as controls for later-treated branches in your staggered TWFE?"
Yes, in a staggered adoption setting. This is the Callaway–Sant'Anna critique: with staggered timing and heterogeneous effects, the TWFE weights can be negative for some (g,t) cells, potentially sign-reversing the estimate. If your rollout was staggered, use cohort-time estimators and present the event study disaggregated by cohort.

**Challenge 3:** "Your experimental DiD treats missing-at-random dropout as acceptable — but what if high-WTP participants in the standard condition dropped out because they found the baseline session discouraging?"
Differential dropout by the outcome of interest is MNAR, not MAR. The DiD estimate from completers would be biased. Sensitivity analysis comparing the completer estimate to an IPW-weighted estimate (and reporting the two together) is the appropriate response.
:::
