---
title: "Regression Discontinuity Design"
---

```{r setup, include=FALSE}
library(tidyverse)
library(rdrobust)
library(ggplot2)
set.seed(42)
```

# Regression Discontinuity Design

## The Threshold as a Quasi-Experiment

Suppose college admission is partly determined by a test score cutoff: applicants above 70 are admitted, those below are not. People just below 70 and just above 70 are otherwise very similar — nearly identical test scores, similar backgrounds, similar demographics. The only thing separating them is whether they fell above or below an arbitrary threshold.

This **discontinuity in treatment assignment**, driven by a continuous **running variable**, is the foundation of Regression Discontinuity Design (RDD).

The key insight: **right at the threshold, treatment is as-good-as-randomly assigned**, because small differences in the running variable near the cutoff are essentially random noise in the DGP.

## The Core Assumption

::: {.callout-warning icon=false}
## Assumption: Continuity of Potential Outcomes at the Threshold

**What it means:** In the absence of treatment, the expected outcome must be a continuous function of the running variable at the cutoff. The only reason outcomes should jump at the threshold is the treatment.

**Why it fails:**
- **Sorting/manipulation:** If people can manipulate their running variable to cross the threshold (retaking tests, strategic reporting), those just above and just below are not comparable.
- **Other policies at the same cutoff:** If another treatment also kicks in at exactly the same threshold, you cannot separate the effects.
- **Anticipation:** If people change behavior in anticipation of crossing the threshold.

**How to check:** McCrary density test — a spike in the running variable just above the cutoff indicates manipulation.
:::

## A Worked Example

```{r rdd-simulation}
n <- 1000
# Running variable: test score, centered at cutoff = 0
test_score <- rnorm(n, 0, 1)
admitted   <- as.integer(test_score >= 0)
# True causal effect of admission on earnings = 8
earnings   <- 50 + 5 * test_score + 8 * admitted + rnorm(n, 0, 5)

df_rdd <- tibble(test_score, admitted, earnings)

ggplot(df_rdd, aes(x = test_score, y = earnings, color = factor(admitted))) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE, linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
  scale_color_manual(values = c("#aec6e8", "#1e3a5f"),
                     labels = c("Not admitted", "Admitted")) +
  labs(title = "RDD: earnings vs. test score",
       subtitle = "Jump at the dashed cutoff = causal effect of admission",
       x = "Test Score (centered at cutoff)", y = "Earnings",
       color = NULL) +
  theme_minimal(base_size = 13)
```

```{r rdd-estimate}
rdd_result <- rdrobust(y = df_rdd$earnings, x = df_rdd$test_score, c = 0)
summary(rdd_result)
```

`rdrobust` uses local polynomial regression with an automatically selected optimal bandwidth.

## The Density Test

```{r density-test}
ggplot(df_rdd, aes(x = test_score)) +
  geom_histogram(bins = 40, fill = "#1e3a5f", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, color = "#e63946", linewidth = 1.2, linetype = "dashed") +
  labs(title = "Density test: check for manipulation at the cutoff",
       subtitle = "A spike just right of the dashed line = sorting violation",
       x = "Test Score (centered at cutoff)", y = "Count") +
  theme_minimal(base_size = 13)
```

## Falsification Tests

Run the RDD on **pre-treatment covariates** (age, gender, family income). These cannot be caused by the treatment. A jump at the threshold in a covariate means something else is discontinuous there — a serious threat to validity.

```{r falsification}
df_rdd <- df_rdd |> mutate(age = 18 + rnorm(n, 0, 1))
fake   <- rdrobust(y = df_rdd$age, x = df_rdd$test_score, c = 0)
cat("Jump in age at cutoff (should be ~0):", round(fake$coef[1], 3),
    "| p =", round(fake$pv[1], 3), "\n")
```

## Bandwidth Sensitivity

The estimate should be stable as you vary the bandwidth around the default. Dramatic changes signal dependence on functional form.

```{r bandwidth-sensitivity}
bws  <- seq(0.3, 1.5, by = 0.2)
ests <- sapply(bws, \(bw)
  rdrobust(y = df_rdd$earnings, x = df_rdd$test_score, c = 0, h = bw)$coef[1])

tibble(Bandwidth = bws, Estimate = ests) |>
  ggplot(aes(x = Bandwidth, y = Estimate)) +
  geom_line(color = "#1e3a5f", linewidth = 1.2) +
  geom_point(size = 3, color = "#1e3a5f") +
  geom_hline(yintercept = 8, linetype = "dashed", color = "#e63946") +
  labs(title = "Bandwidth sensitivity (red = true effect)",
       x = "Bandwidth", y = "Estimated Discontinuity") +
  theme_minimal(base_size = 13)
```
