---
title: "Regression Discontinuity Design"
---

```{r setup, include=FALSE}
library(tidyverse)
library(rdrobust)
library(ggplot2)
library(patchwork)
library(lme4)
set.seed(42)
```

# Regression Discontinuity Design

## The Threshold as a Quasi-Experiment

Regression Discontinuity Design (RDD) exploits a rule: treatment is assigned when a continuous **running variable** crosses a known threshold. Observations just below and just above the threshold are nearly identical in every respect except treatment status — because small differences in the running variable near the cutoff are essentially random noise in the data-generating process.

**The key insight:** Right at the threshold, treatment is as-good-as-randomly assigned. The jump in outcomes at the cutoff is the causal effect of treatment for units near the threshold.

RDD appears in two distinct forms across the research contexts this course covers. In **longitudinal research**, the running variable is often time itself (Interrupted Time Series) or a cumulative score that triggers a policy intervention. In **experimental research**, the running variable is often a continuous eligibility screening score, and the cutoff determines who qualifies for a treatment arm.

### Why Local Comparison Is Valid: The Bandwidth Argument

The central claim of RDD is that units *near* the cutoff are comparable in a way that units *far* from the cutoff are not. A borrower with a credit score of 648 and a borrower with a credit score of 652 are separated by what amounts to random noise in the data-generating process — a single missed payment, a dispute that cleared, a month's credit utilization that ticked slightly in one direction. Their underlying creditworthiness is essentially identical. But a borrower with a score of 600 and a borrower with a score of 700 differ systematically in creditworthiness, payment history, income stability, and many other factors that independently predict financial outcomes.

This means comparing *all* borrowers above the cutoff to *all* borrowers below the cutoff conflates the treatment effect with all the real differences between high- and low-credit-score individuals. The RDD narrows the comparison to a **bandwidth** around the cutoff where assignment is effectively local and the confounders are approximately balanced.

```{r rdd-concept-diagram, echo=FALSE, fig.width=9, fig.height=4.2, message=FALSE, warning=FALSE}
set.seed(42)
n_pts <- 100
bw    <- 0.8

df_concept <- tibble(
  x     = c(runif(n_pts / 2, -2.5, -0.01), runif(n_pts / 2, 0.01, 2.5)),
  treat = rep(c("Below cutoff (untreated)", "Above cutoff (treated)"), each = n_pts / 2)
) |>
  mutate(
    in_bw    = abs(x) <= bw,
    y_jitter = rnorm(n(), 0, 0.06)
  )

ggplot(df_concept, aes(x = x, y = y_jitter)) +
  annotate("rect",
           xmin = -bw, xmax = bw, ymin = -0.30, ymax = 0.30,
           fill = "#fffde7", alpha = 0.85, color = "#e67e00", linewidth = 1.0,
           linetype = "dashed") +
  geom_point(aes(color = treat, alpha = in_bw, size = in_bw)) +
  geom_vline(xintercept = 0, linewidth = 1.3, linetype = "longdash",
             color = "#333333") +
  annotate("text", x = 0, y = 0.34,
           label = paste0("\u2190 Bandwidth (\u00b1", bw, ") \u2192"),
           hjust = 0.5, vjust = 0, size = 3.3, color = "#e67e00", fontface = "bold") +
  annotate("text", x = 0.04, y = 0.24,
           label = "Near-threshold units:\nsmall score difference \u2248 random noise\n\u21d2 treated and control are comparable",
           hjust = 0.5, size = 2.7, color = "#1e3a5f", fontface = "italic") +
  annotate("text", x = -1.85, y = -0.34,
           label = "Low-score units:\nsystematically different\nfrom treated group",
           hjust = 0.5, size = 2.6, color = "#999999") +
  annotate("text", x = 1.85, y = -0.34,
           label = "High-score units:\nsystematically different\nfrom control group",
           hjust = 0.5, size = 2.6, color = "#999999") +
  scale_color_manual(values = c("Below cutoff (untreated)" = "#aec6e8",
                                "Above cutoff (treated)"   = "#1e3a5f")) +
  scale_alpha_manual(values = c("FALSE" = 0.18, "TRUE" = 0.95), guide = "none") +
  scale_size_manual(values  = c("FALSE" = 1.1,  "TRUE" = 2.6),  guide = "none") +
  scale_x_continuous(
    breaks = c(-2, -1, 0, 1, 2),
    labels = c("\u22122", "\u22121", "Cutoff", "+1", "+2")
  ) +
  labs(
    title    = "Why RDD restricts the comparison to units near the threshold",
    subtitle = "Bright large dots (inside bandwidth) are locally comparable. Faint small dots (far from cutoff) are not.",
    x        = "Running variable (e.g., credit score, centered at cutoff)",
    y        = NULL,
    color    = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.y      = element_blank(),
    axis.ticks.y     = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position  = "top"
  )
```

The bandwidth is not arbitrary — `rdrobust` selects it optimally to balance the bias-variance trade-off: a wider bandwidth increases sample size (lower variance) but includes more units far from the cutoff where the "approximately random" claim is weaker (higher bias). The density test and falsification tests below check whether the local-randomization assumption holds.

---

## Part A: RDD in Longitudinal Research

### Sharp RDD: Credit Score Threshold and Prime-Rate Loan Access

A common financial services RDD uses a credit score threshold to assign access to a preferential loan product. Borrowers whose score meets or exceeds the cutoff (e.g., 650) qualify for the prime-rate product with lower interest; those below do not. Borrowers just below and just above an arbitrary administrative cutoff are nearly identical in their actual creditworthiness — the cutoff is a bureaucratic rule, not a true discontinuity in financial behavior.

The outcome of interest: 12-month improvement in financial health (a composite of savings rate and debt reduction). If access to the prime-rate product improves financial health, a jump should appear precisely at the threshold.

::: {.callout-warning icon=false}
## Assumption: Continuity of Potential Outcomes at the Threshold

**What it means:** In the absence of treatment, the expected outcome must be a continuous function of the running variable at the cutoff. The only reason outcomes should jump at the threshold is the treatment itself — prime-rate access.

**Why it fails:**

- **Sorting/manipulation:** If borrowers can manipulate their credit score to cross the threshold (paying down a specific account, disputing a single error), those just above and just below are no longer comparable. The density test catches this: a spike in the density just above 650 indicates strategic score management.
- **Other policies at the same cutoff:** If a second product (e.g., a fee waiver) also triggers at credit score 650, you cannot separate the effects.
- **Anticipation effects:** If borrowers anticipate the cutoff and change their financial behavior *before* crossing it (e.g., deferring large purchases until they qualify), the pre-threshold trend is distorted.

**What survives if this assumption fails:** You can still describe the average difference between borrowers just above and below the threshold — but you cannot interpret it as causal. If sorting is limited to a narrow band, the estimate may still be approximately valid with a tighter bandwidth.

**How to check:** McCrary density test — a spike in the density just above the cutoff indicates strategic score manipulation.
:::

```{r rdd-simulation}
n <- 1000
# Running variable: credit score, centered at cutoff = 0 (e.g., 650 in real units)
credit_score  <- rnorm(n, 0, 1)
prime_access  <- as.integer(credit_score >= 0)
# True causal effect of prime-rate access on 12-month financial health = 8 points
fin_health    <- 50 + 5 * credit_score + 8 * prime_access + rnorm(n, 0, 5)

df_rdd <- tibble(credit_score, prime_access, fin_health)

ggplot(df_rdd, aes(x = credit_score, y = fin_health, color = factor(prime_access))) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE,
              linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black",
             linewidth = 1) +
  scale_color_manual(values = c("#aec6e8", "#1e3a5f"),
                     labels = c("Standard rate (score < cutoff)",
                                "Prime rate (score ≥ cutoff)")) +
  labs(title = "RDD: financial health vs. credit score",
       subtitle = "Jump at the dashed cutoff = causal effect of prime-rate access",
       x = "Credit Score (centered at cutoff)", y = "12-month financial health score",
       color = NULL) +
  theme_minimal(base_size = 13)
```

```{r rdd-estimate}
rdd_result <- rdrobust(y = df_rdd$fin_health, x = df_rdd$credit_score, c = 0)
summary(rdd_result)
```

`rdrobust` uses local polynomial regression with an automatically selected optimal bandwidth. The estimated jump at the cutoff recovers the true causal effect of approximately 8 financial health units.

### The Density Test

```{r density-test}
ggplot(df_rdd, aes(x = credit_score)) +
  geom_histogram(bins = 40, fill = "#1e3a5f", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, color = "#e63946", linewidth = 1.2,
             linetype = "dashed") +
  labs(title = "Density test: check for manipulation at the credit score cutoff",
       subtitle = "A spike just right of the dashed line = borrowers gaming their score",
       x = "Credit Score (centered at cutoff)", y = "Count") +
  theme_minimal(base_size = 13)
```

The distribution is smooth through the cutoff — no evidence of strategic score manipulation. In real credit data, a spike just above the threshold would indicate that borrowers near the margin are paying down specific accounts, disputing errors, or taking other targeted actions to cross it.

### Falsification Tests

Run the RDD on **pre-treatment covariates** (account tenure, age). These cannot be caused by getting prime-rate access. A jump at the threshold in a covariate means something else is discontinuous there — a serious threat to validity.

```{r falsification}
df_rdd <- df_rdd |> mutate(acct_tenure = 5 + rnorm(n, 0, 2))
fake   <- rdrobust(y = df_rdd$acct_tenure, x = df_rdd$credit_score, c = 0)
cat("Jump in account tenure at cutoff (should be ~0):", round(fake$coef[1], 3),
    "| p =", round(fake$pv[1], 3), "\n")
```

### Bandwidth Sensitivity

The estimate should be stable as you vary the bandwidth around the default. Dramatic changes signal dependence on functional form assumptions near the threshold.

```{r bandwidth-sensitivity}
bws  <- seq(0.3, 1.5, by = 0.2)
ests <- sapply(bws, \(bw)
  rdrobust(y = df_rdd$fin_health, x = df_rdd$credit_score, c = 0, h = bw)$coef[1])

tibble(Bandwidth = bws, Estimate = ests) |>
  ggplot(aes(x = Bandwidth, y = Estimate)) +
  geom_line(color = "#1e3a5f", linewidth = 1.2) +
  geom_point(size = 3, color = "#1e3a5f") +
  geom_hline(yintercept = 8, linetype = "dashed", color = "#e63946") +
  labs(title = "Bandwidth sensitivity (red dashed = true effect = 8)",
       x = "Bandwidth", y = "Estimated Discontinuity") +
  theme_minimal(base_size = 13)
```

---

### Interrupted Time Series as Temporal RDD

A special form of RDD uses **time** as the running variable. When a policy takes effect at a known date, observations before that date serve as "control" and observations after as "treated." The continuity assumption becomes: absent the policy, the pre-existing time trend would have continued smoothly through the policy date.

This design is called an **Interrupted Time Series (ITS)** and is especially valuable in longitudinal panel data where a policy is applied to all units simultaneously.

The ITS model separates three distinct quantities:

$$Y_t = \underbrace{\beta_0 + \beta_1 t}_{\text{pre-policy trend}} + \underbrace{\beta_2 D_t}_{\text{level jump at policy date}} + \underbrace{\beta_3 (t - c) D_t}_{\text{post-policy slope change}} + \varepsilon_t$$

- $\beta_1$: the pre-policy trend (the counterfactual slope — what would have continued)
- $\beta_2$: the immediate level jump at the policy date (the "intercept shift")
- $\beta_3$: the change in slope after the policy (whether the policy altered the trajectory)

```{r its-simulation}
n_stores <- 30
n_months <- 30
store_eff <- rnorm(n_stores, 0, 6)  # store-level random intercepts

df_its <- expand_grid(store = 1:n_stores, month = 1:n_months) |>
  mutate(
    post    = as.integer(month >= 16),          # policy starts at month 16
    t_post  = pmax(month - 15, 0),              # time since policy, 0 pre-policy
    store_u = store_eff[store],
    # True: level jump = +5, slope change = +0.3/month
    sales   = 60 + 0.8 * month + 5 * post + 0.3 * t_post + store_u +
                rnorm(n(), 0, 5),
    store   = factor(store)
  )

m_its <- lmer(sales ~ month + post + t_post + (1 | store), data = df_its,
              REML = FALSE)
summary(m_its)$coefficients |>
  as.data.frame() |>
  rownames_to_column("term") |>
  filter(term %in% c("month", "post", "t_post")) |>
  rename(estimate = Estimate, std.error = `Std. Error`, t = `t value`) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "ITS model: pre-trend, level jump (post), and slope change (t_post)")
```

```{r its-plot}
# Aggregate to store-average for plotting
df_its_agg <- df_its |>
  group_by(month, post) |>
  summarise(sales = mean(sales), .groups = "drop")

# Predicted line from model (using population-level prediction, no store RE)
coefs <- fixef(m_its)
df_pred <- tibble(month = 1:30) |>
  mutate(
    post   = as.integer(month >= 16),
    t_post = pmax(month - 15, 0),
    pred   = coefs["(Intercept)"] + coefs["month"] * month +
               coefs["post"] * post + coefs["t_post"] * t_post
  )

ggplot() +
  geom_point(data = df_its_agg, aes(x = month, y = sales, color = factor(post)),
             size = 2, alpha = 0.8) +
  geom_line(data = df_pred, aes(x = month, y = pred), linewidth = 1.2,
            color = "#1e3a5f") +
  geom_vline(xintercept = 15.5, linetype = "dashed", color = "#e63946",
             linewidth = 1.1) +
  annotate("text", x = 16.5, y = 72, label = expression(beta[2] ~ "(level jump)"),
           hjust = 0, color = "#e63946", size = 4) +
  annotate("text", x = 22, y = 84, label = expression(beta[3] ~ "(slope change)"),
           hjust = 0, color = "#555555", size = 4) +
  scale_color_manual(values = c("#aec6e8", "#1e3a5f"),
                     labels = c("Pre-policy", "Post-policy")) +
  labs(title = "Interrupted Time Series: pricing policy at month 16",
       x = "Month", y = "Mean Sales", color = NULL) +
  theme_minimal(base_size = 13)
```

**Interpretation:**

- $\hat{\beta}_2 \approx 5$: The policy caused an immediate, step-change increase in sales at implementation.
- $\hat{\beta}_3 \approx 0.3$: The policy also increased the monthly sales growth rate by 0.3 units per month compared to the pre-policy trend.
- $\hat{\beta}_1 \approx 0.8$: The counterfactual — in the absence of the policy, sales would have continued rising by 0.8 units per month.

The ITS design cannot rule out other events that happened exactly at month 16. The strongest threats are contemporaneous shocks (another policy, a market event) that coincide precisely with the policy date.

**The density-test equivalent for ITS:** Check for unusual clustering of observations just before the policy implementation date. Firms that anticipate a pricing policy may accelerate purchases or delay inventory, creating a temporary spike in the outcome in periods 14–15. Plot the distribution of the outcome in the 3–4 periods immediately before implementation and compare to the pre-trend expectation.

---

## Part B: RDD in Experimental Research

RDD logic appears in applied studies when eligibility for an intervention is determined by crossing a continuous screening threshold. Unlike the observational RDD — where the cutoff is an exogenous administrative rule — this version uses the cutoff to create a natural comparison group within a structured study.

### Eligibility Threshold: Financial Distress Screening and Coaching Access

A financial services organization screens applicants for an intensive debt-counseling program using a Financial Vulnerability Index (FVI) — a composite of debt-to-income ratio, payment history, and account balances scored 0–30 (higher = more financially distressed). Applicants scoring **≥ 15** are offered access to the intensive coaching program; those scoring below 15 receive standard informational materials.

This creates an RDD: applicants just above and just below FVI = 15 are nearly identical in their actual financial situation — the cutoff is an administrative rule, not a true behavioral discontinuity. The outcome is savings rate improvement over six months.

The FVI RDD is **sharp** if the cutoff is strictly enforced. It becomes **fuzzy** if advisors exercise discretion — some below-threshold clients are enrolled based on advisor judgment, some above-threshold clients decline. In the fuzzy case, the cutoff becomes an instrument for actual program participation, and the estimand is a LATE.

::: {.callout-warning icon=false}
## Assumption: Sharp vs. Fuzzy Cutoff

**Sharp RDD:** Crossing the threshold deterministically assigns treatment. The discontinuity directly identifies the causal effect for units near the cutoff.

**Fuzzy RDD:** The cutoff changes the *probability* of treatment receipt — some above-threshold clients decline, some below-threshold clients are enrolled by advisor discretion. Fuzzy RDD gives a LATE via IV: the cutoff is an instrument for actual program participation. Use `rdrobust(..., fuzzy = enrolled)` for fuzzy designs.

**What survives if neither version holds cleanly:** If enforcement is very fuzzy (the threshold is essentially ignored), the first stage is weak and the RDD estimate is unreliable. Report the first-stage jump in participation probability alongside the outcome estimate. A first-stage F below 10 is a red flag.

The key diagnostic: does crossing the threshold produce a large, discontinuous jump in the probability of participation?
:::

```{r coaching-rdd}
n   <- 600
fvi <- round(runif(n, 0, 30))        # FVI scores from 0 to 30
enrolled <- as.integer(fvi >= 15)    # sharp: ≥15 enrolled in intensive coaching

# True causal effect for applicants near the threshold = 6 pp savings improvement
# Underlying trend: more distressed applicants have lower baseline savings gains
savings_gain <- 8 - 0.2 * fvi + 6 * enrolled + rnorm(n, 0, 4)

df_coaching <- tibble(fvi, enrolled, savings_gain)

rdd_coaching <- rdrobust(y = df_coaching$savings_gain, x = df_coaching$fvi, c = 15)
summary(rdd_coaching)
```

```{r coaching-rdd-plot}
ggplot(df_coaching, aes(x = fvi, y = savings_gain, color = factor(enrolled))) +
  geom_point(alpha = 0.3, size = 1.2) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, linewidth = 1.3) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "black",
             linewidth = 1.1) +
  scale_color_manual(values = c("#aec6e8", "#1e3a5f"),
                     labels = c("Standard materials (FVI < 15)",
                                "Intensive coaching (FVI ≥ 15)")) +
  labs(title = "RDD: savings gain at Financial Vulnerability Index cutoff = 15",
       subtitle = "Jump at the cutoff = causal effect of coaching for applicants near FVI = 15",
       x = "Financial Vulnerability Index (FVI)", y = "6-month savings rate improvement (%)",
       color = NULL) +
  theme_minimal(base_size = 13)
```

### Density Test: Did Applicants Game Their FVI Score?

If applicants believe that scoring ≥ 15 gives them access to a desirable coaching program, they might strategically misreport their financial situation (understating income, overstating debts) to cross the threshold. A spike in the density just above 15 indicates this gaming behavior.

```{r coaching-density}
ggplot(df_coaching, aes(x = fvi)) +
  geom_histogram(bins = 31, fill = "#1e3a5f", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 15, color = "#e63946", linewidth = 1.2,
             linetype = "dashed") +
  labs(title = "Density test: Financial Vulnerability Index distribution",
       subtitle = "A spike just above 15 (red dashed) = applicants gaming their score to qualify",
       x = "Financial Vulnerability Index (FVI)", y = "Count") +
  theme_minimal(base_size = 13)
```

In this simulation there is no spike — the distribution is smooth through the cutoff, consistent with the continuity assumption.

### Falsification Test: No Jump in Pre-Screening Covariates

Run the RDD on account tenure — a pre-screening covariate that cannot be caused by coaching enrollment. A discontinuity at FVI = 15 in account tenure would indicate that something other than financial distress drives the threshold assignment.

```{r coaching-falsification}
df_coaching <- df_coaching |>
  mutate(acct_tenure = 5 + rnorm(n, 0, 3))

fake_coaching <- rdrobust(y = df_coaching$acct_tenure, x = df_coaching$fvi, c = 15)
cat("Jump in account tenure at FVI = 15 (should be ~0):",
    round(fake_coaching$coef[1], 3),
    "| p =", round(fake_coaching$pv[1], 3), "\n")
```

### The LATE Interpretation for the Coaching RDD

Like all RDD estimates, this is a **Local Average Treatment Effect** for applicants near FVI = 15 — those at the margin of financial distress. The treatment effect for severely distressed applicants (FVI = 25+) may be very different: they face structural financial constraints that intensive coaching may not overcome, or alternatively, they may benefit most. But RDD cannot speak to them — they are far from the threshold.

**The practical implication:** The RDD estimate tells you whether coaching helps *marginally distressed* applicants — exactly the population where the program's benefit is most uncertain and contested. It cannot tell you whether to expand the program to the most severely distressed, who would need a different identification strategy (e.g., randomized waitlist).

---

## Summary: RDD Across Contexts

| Feature | Observational RDD | Interrupted Time Series | Screening-Threshold RDD |
|---|---|---|---|
| Running variable | Continuous score (credit, income, age) | Time (months, quarters) | Continuous distress/eligibility screener |
| Threshold | Administrative cutoff (e.g., credit score 650) | Policy implementation date | Program eligibility cutoff (e.g., FVI ≥ 15) |
| Treatment | Product / advisory access | Pre vs. post policy | Intensive coaching enrollment |
| Density test | Spike in score just above cutoff = gaming | Unusual clustering before policy date | Spike in screener score just above cutoff |
| Continuity assumption | No sorting; no other policy at same cutoff | No confounding events at policy date | No strategic misreporting of screener |
| LATE for... | Borrowers near the credit score threshold | All units at the policy date | Applicants near the FVI = 15 cutoff |

RDD is most credible when the threshold is truly arbitrary from the unit's perspective, when the running variable cannot be precisely manipulated, and when the bandwidth is narrow enough that treated and control units are genuinely comparable. The narrower the bandwidth, the more credible the local randomization — but also the smaller the sample and the larger the standard errors.
