---
title: "Effect Sizes, Reporting, and Reproducibility"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(performance)
library(patchwork)
set.seed(2024)
```

# Effect Sizes, Reporting, and Reproducibility

## Why This Chapter Is the Capstone for Module 5

This chapter does two things. First, it covers the statistical mechanics of effect sizes for mixed models — what they mean and how to compute them. Second, and more importantly, it provides **design-specific reporting templates** so that a reader who completed Module 5 can write up any nested experimental analysis in a way that is transparent, verifiable, and reviewer-ready.

The guiding question:

> Given the nesting structure, estimand, and assumption profile of your analysis, what does a reviewer need to see to evaluate whether the design and model are appropriate?

This chapter covers:

1. Effect size measures for mixed models — what they mean and how to compute them
2. Translating fixed effect estimates to interpretable metrics
3. Conditional vs. marginal $R^2$
4. Design-specific reporting templates (crossed RE, partially nested, multi-site, GLMM, noncompliance, quasi-experimental)
5. A minimum reproducible model report
6. Preregistration prompts for nested experimental designs
7. Claim calibration: how to narrow your claim when assumptions are weak
8. Reproducibility practices: seeds, version pinning, and transparent reporting

---

## Part 1: Effect Size in Mixed Models

### The Problem with Standard Effect Sizes

Cohen's $d$ is defined as the standardized mean difference:

$$d = \frac{\mu_1 - \mu_2}{\sigma}$$

But what is $\sigma$ in a mixed model? There are multiple variance components: the residual variance $\sigma^2_\varepsilon$, the random intercept variance $\tau^2_{00}$, and possibly additional random slope variances. Each gives a different denominator — and therefore a different $d$.

The answer depends on the question you are asking:

- **Conditional $d$:** standardizes by the within-cluster (residual) variance. Answers: "How large is the effect within a person/cluster, relative to how much observations vary within that cluster?"
- **Marginal $d$:** standardizes by the total variance (random effects + residual). Answers: "How large is the effect relative to all variability in the outcome?"

Neither is wrong. They answer different questions.

### Cohen's $d$ for Experimental Effects in Nested Data

In a typical within-subjects experiment, the effect of interest is the difference between conditions, and the relevant variability is the within-person trial-level variance. The residual SD from the mixed model (not including the random effect variance) is the appropriate standardizer for a *within-person* $d$.

```{r compute-d}
# Simulate a within-subjects experiment: 80 participants × 10 trials per condition
n_par  <- 80
n_trial <- 10

df_d <- expand_grid(
  participant = 1:n_par,
  condition   = c("control", "treatment"),
  trial       = 1:n_trial
) |>
  mutate(
    u0          = rnorm(n_par, 0, 8)[participant],  # between-person SD = 8
    residual_sd = 5,
    effect      = 3,  # true within-person effect
    response    = 50 + u0 + effect * (condition == "treatment") + rnorm(n(), 0, residual_sd),
    participant = factor(participant),
    condition   = factor(condition, levels = c("control", "treatment"))
  )

m_d <- lmer(response ~ condition + (1 | participant), data = df_d)

# Extract variance components
vc <- as.data.frame(VarCorr(m_d))
sigma_residual <- vc$sdcor[vc$grp == "Residual"]     # within-person SD
sigma_person   <- vc$sdcor[vc$grp == "participant"]   # between-person SD
sigma_total    <- sqrt(sigma_person^2 + sigma_residual^2)

# Fixed effect estimate
beta_treatment <- fixef(m_d)["conditiontreatment"]

# Two versions of d
d_conditional <- beta_treatment / sigma_residual
d_marginal    <- beta_treatment / sigma_total

tibble(
  metric         = c("Fixed effect (unstandardized)", "Conditional d (÷ residual SD)", "Marginal d (÷ total SD)"),
  value          = c(beta_treatment, d_conditional, d_marginal),
  denominator    = c("—", round(sigma_residual, 2), round(sigma_total, 2))
) |>
  knitr::kable(digits = 3, caption = "Three ways to express the same treatment effect")
```

**Interpreting the two $d$ values:**
- **Conditional $d$:** standardizes within the trial-level noise. A $d$ of 0.6 here means the treatment shifts the mean by 0.6 trial-level SDs — controlling for stable person differences.
- **Marginal $d$:** standardizes relative to all variability including between-person differences. This will always be smaller than conditional $d$ when there is substantial between-person variance, because the denominator is larger. It represents the treatment's effect size in the full distribution of responses.

The **conditional $d$ is usually more appropriate for experimental reporting**, because the design specifically removes between-person variance through random assignment and repeated measurement. Marginal $d$ deflates the effect by variance that the design was explicitly not trying to explain.

---

## Part 2: Conditional and Marginal $R^2$ for Mixed Models

### The Nakagawa-Schielzeth $R^2$

For linear mixed models, Nakagawa & Schielzeth (2013) define two $R^2$ values:

- **Marginal $R^2_m$:** Variance explained by **fixed effects only** — how much the predictors in the model account for, ignoring individual differences.
- **Conditional $R^2_c$:** Variance explained by **fixed effects plus random effects** — how much the whole model (predictors + clustering) accounts for.

The difference $R^2_c - R^2_m$ is the variance uniquely attributable to the random effects (the clustering).

```{r r2-computation}
# Compute R² using the performance package
r2_results <- performance::r2(m_d)
print(r2_results)
```

### Interpreting $R^2$ in Nested Designs

Consider what these values tell you:

- A large **marginal $R^2$** means your experimental manipulations account for substantial variance in the outcome.
- A large gap between **conditional and marginal $R^2$** means individual differences (the random effects) account for substantial variance — much of the outcome is determined by who the person is, not what condition they were in.

In a typical within-subjects experiment, conditional $R^2$ can be quite high (0.5–0.8) while marginal $R^2$ is low (0.03–0.15). This is not a sign that the experiment failed — it is a sign that the random effects structure is doing its job, absorbing the stable between-person variance and leaving the experimental effect clearly visible in the residuals.

```{r r2-decomposition}
# Show R² decomposition visually
r2_vals <- performance::r2(m_d)

r2_df <- tibble(
  component = c("Fixed effects (marginal R²)", "Random effects (conditional − marginal)", "Unexplained"),
  proportion = c(
    r2_vals$R2_marginal,
    r2_vals$R2_conditional - r2_vals$R2_marginal,
    1 - r2_vals$R2_conditional
  )
)

ggplot(r2_df, aes(x = "", y = proportion, fill = component)) +
  geom_col(width = 0.5) +
  scale_fill_manual(values = c("#1e3a5f", "#4a90d9", "#d0d0d0")) +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(
    title = "Variance decomposition in the mixed model",
    subtitle = "Fixed effects (marginal R²) + Random effects contribution + Residual unexplained",
    x = NULL, y = "Proportion of total variance", fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

## Part 3: ICC as an Effect Size

The **Intraclass Correlation Coefficient (ICC)** — covered in Module 1 — is itself an effect size: it quantifies how much of the outcome's variance is attributable to the clustering structure.

$$\text{ICC} = \frac{\tau^2_{00}}{\tau^2_{00} + \sigma^2_\varepsilon}$$

In experimental reporting, the ICC tells readers how much stable between-person differences matter:

- **High ICC (> 0.5):** Most variance is between people. Within-subjects designs are very efficient here (they remove the large between-person noise). Between-subjects designs require large samples to detect effects against this noise.
- **Low ICC (< 0.1):** Little between-person clustering. Within- and between-subjects designs are roughly equivalent in power.

Reporting the ICC in an experimental methods section tells readers how much person-to-person variability was present and why a particular design choice (within vs. between subjects) was efficient.

---

## Part 4: Translating Model Output to Interpretable Metrics

### From Log-Odds to Probabilities

For logistic mixed models, the fixed effect is in log-odds units. Always translate to probabilities for reporting:

```{r logit-to-prob}
# Example: framing experiment on donation probability
# Log-odds effect = 0.59 (OR = 1.8)

log_odds_effect    <- 0.59
baseline_log_odds  <- -0.5  # average person's log-odds in control

control_prob   <- plogis(baseline_log_odds)
treatment_prob <- plogis(baseline_log_odds + log_odds_effect)

tibble(
  metric = c("Log-odds effect", "Odds ratio", "Control probability", "Treatment probability", "Probability difference"),
  value  = c(log_odds_effect, exp(log_odds_effect), control_prob, treatment_prob, treatment_prob - control_prob)
) |>
  knitr::kable(digits = 3, caption = "Translating from log-odds to probabilities for reporting")
```

**Reporting template:** "Loss-framed appeals had higher donation rates than gain-framed appeals (OR = 1.81, 95% CI [1.42, 2.31], $z$ = 4.7, $p$ < .001). At the average baseline donation probability of 38%, loss framing increased donation probability by approximately 12 percentage points (to 50%)."

### From Log-Counts to Rate Ratios

For Poisson/negative binomial models, exponentiate to get rate ratios:

```{r count-interpretation}
# Example: scarcity on helping count
# Log effect = -0.43 (rate ratio = 0.65)

log_rate_effect    <- -0.43
baseline_log_count <- 1.8  # average log(expected count) in control

control_count   <- exp(baseline_log_count)
treatment_count <- exp(baseline_log_count + log_rate_effect)

tibble(
  metric = c("Log-count effect", "Rate ratio", "Control mean count", "Treatment mean count", "Count reduction"),
  value  = c(log_rate_effect, exp(log_rate_effect), control_count, treatment_count, control_count - treatment_count)
) |>
  knitr::kable(digits = 3, caption = "Translating from log-counts to counts for reporting")
```

**Reporting template:** "Scarcity reduced helping behavior relative to control (RR = 0.65, 95% CI [0.48, 0.88], $p$ = .005). In the control condition, participants helped on average 6.0 times; in the scarcity condition, 3.9 times — a reduction of approximately 2.1 helping episodes."

---

## Part 5: Reporting Standards for Nested Experimental Data

### What to Report

A complete report for a mixed model analysis includes:

::: {.callout-tip}
## Reporting Checklist for Mixed Models

**Design description:**
- [ ] Whether the design is within-subjects, between-subjects, or mixed factorial
- [ ] Number of participants, clusters, and observations per cell
- [ ] ICC of the outcome (how much between-person variance was present)

**Model specification:**
- [ ] Fixed effects included (and their coding: treatment, contrast, etc.)
- [ ] Random effects structure (intercept only? slopes? which predictors vary by person?)
- [ ] Software and package versions used

**Results for each effect:**
- [ ] Unstandardized estimate with units
- [ ] Standard error and 95% confidence interval
- [ ] Test statistic and degrees of freedom ($t$, $z$, or $F$ with Satterthwaite/Kenward-Roger df)
- [ ] $p$-value
- [ ] Effect size ($d$, $\eta^2$, OR, RR as appropriate)

**Model-level fit:**
- [ ] Marginal and conditional $R^2$
- [ ] Variance components ($\tau^2_{00}$, $\sigma^2_\varepsilon$, and any random slopes)

**For non-normal outcomes:**
- [ ] Overdispersion check (for count models)
- [ ] Predicted probabilities or counts at substantively meaningful covariate values
:::

---

## Design-Specific Reporting Templates

Different designs require different minimum disclosures. Use the template for your design.

::: {.callout-tip icon=false}
## Crossed Participant × Stimulus Design (Part 1)

> "We fitted a linear mixed model with participant and message as crossed random effects (`lmer(Y ~ condition + (1|participant) + (1|message)`). The participant ICC was [value], indicating [X]% of response variance was stable between-person. The message ICC was [value]. Treating messages as fixed effects (ignoring message variance) would have underestimated the SE of the framing effect by approximately [X]%. The framing effect was *b* = [estimate], SE = [SE], 95% CI [CI], *t*([df]) = [t], *p* = [p]."
:::

::: {.callout-tip icon=false}
## Partially Nested Design (Part 2)

> "Participants were randomly assigned to a group coaching condition (*n* = [N], [K] coaches, [m] per coach) or an independent self-serve condition (*n* = [N]). Because only the coaching arm had shared group context, we used the Bauer et al. partially nested model: coaching groups were modeled with a random effect, while control participants each received a unique singleton cluster ID. The coaching-group ICC was [value]. The coaching effect estimate was *b* = [estimate], SE = [SE], *p* = [p]. We verified that using a naive model without clustering inflated the coaching effect SE by [X]%."
:::

::: {.callout-tip icon=false}
## Multi-Site Experiment (Part 4)

> "The study was run across [K] sites ([describe recruitment]). Sites were modeled as random effects; the treatment effect was allowed to vary by site (random slope for treatment by site). The average treatment effect across sites was *b* = [estimate], SE = [SE], 95% CI [CI]. The SD of treatment effects across sites was [value], indicating that the 80% plausible range of site effects was approximately [[lower], [upper]]. A leave-one-site-out analysis confirmed no single site was driving the average estimate. Because sites were [convenience-sampled / systematically recruited], results generalize to [describe population]."
:::

::: {.callout-tip icon=false}
## GLMM with Binary or Bounded Count Outcome (Part 6)

> **Binary:** "We fitted a logistic mixed model with [random effects structure]. The condition effect was *b* = [log-odds], OR = [exp(b)], 95% CI [CI], *z* = [z], *p* = [p]. At the sample mean baseline probability of [X]%, the treatment increased acceptance probability by approximately [Y] percentage points (to [Z]%). The random intercept variance was [τ²], ICC = [value]."

> **Bounded count (binomial):** "Risky choices out of 10 gambles per block were modeled with a binomial GLMM using `cbind(risky_count, 10-risky_count)`. Using Poisson for this bounded outcome would have been incorrect — the DGP is binomial, not Poisson. The disclosure effect was OR = [estimate] (per-gamble acceptance odds ratio), 95% CI [CI], *p* = [p]."
:::

::: {.callout-tip icon=false}
## Non-Compliance / LATE Analysis (Part 8)

> "Compliance in the treatment arm was [X]% (one-sided noncompliance — controls could not access the tool). We report both the intent-to-treat effect (ITT = [estimate], SE = [SE], *p* = [p]) and the complier average causal effect (LATE = [estimate], SE = [SE], *p* = [p]), estimated via instrumental variables regression with assignment as the instrument. The first-stage *F*-statistic was *F* = [F], supporting instrument relevance. The exclusion restriction — that assignment affected savings only through tool activation — is supported by [design argument]."
:::

::: {.callout-tip icon=false}
## Attrition / IPW Analysis (Part 9)

> "At follow-up, [X]% of participants were retained. Retention was lower in the control arm ([X]%) than the treatment arm ([Y]%), and predicted by baseline [covariate]. We estimated inverse probability weights for attrition using a logistic model for retention on baseline covariates. The IPW-corrected ITT was *b* = [estimate], SE = [SE], *p* = [p], compared to the naive complete-case estimate of *b* = [biased estimate]. The IPW analysis assumes dropout was missing at random given observed baseline covariates; if dropout was driven by unobserved outcomes, this correction will not fully remove bias."
:::

---

## Minimum Reproducible Model Report

Every mixed model analysis should include enough information for a reader to exactly reproduce it.

::: {.callout-note icon=false}
## Minimum Reproducible Report Box

```r
# Report all of these:
# 1. Formula
Y ~ condition + (1 + condition | participant) + (1 | stimulus)

# 2. Contrast coding
# condition: treatment = 1, control = 0 (dummy coding)

# 3. Optimizer
lmerControl(optimizer = "bobyqa")

# 4. REML vs. ML
# Fixed effects: REML = TRUE (default); LRT comparisons: REML = FALSE

# 5. Convergence check
isSingular(model)             # TRUE = degenerate variance
performance::check_convergence(model)

# 6. Key diagnostics run
plot(model)                   # residuals vs. fitted
qqnorm(ranef(model)$participant[[1]])  # RE normality

# 7. Package versions
sessionInfo()
```
:::

---

## Preregistration Prompts for Nested Experimental Designs

These prompts cover the decisions that are most consequential for nested experimental analyses and are most commonly left unspecified in preregistrations.

::: {.callout-note icon=false}
## Preregistration Checklist

**What is randomized and at what level?**
- [ ] Unit of randomization (individual, dyad, classroom, site)
- [ ] Whether stimuli are counterbalanced or condition-specific

**What is the estimand?**
- [ ] ATE, ITT, LATE, ATT — be explicit
- [ ] Whether the estimand is population-level or complier-restricted

**What is the primary model formula?**
- [ ] Fixed effects and coding scheme
- [ ] Random effects structure (exactly as you plan to fit it)
- [ ] Model family and link function (for GLMMs)

**Random effects simplification sequence:**
- [ ] Starting structure (maximal theoretically justified)
- [ ] Step 1 if singular/non-convergent: remove correlation parameters
- [ ] Step 2: remove smallest variance component (with LRT criterion)
- [ ] Disclosure: deviations from registered structure will be reported

**Exclusions and attrition rules:**
- [ ] Criteria for excluding participants from analysis
- [ ] How attrition will be handled (complete case, IPW, FIML)
- [ ] Minimum follow-up rate for IPW

**Compliance and instrument:**
- [ ] Expected compliance rate and how compliance is measured
- [ ] Whether LATE will be reported and what the exclusion restriction argument is

**Sensitivity analyses:**
- [ ] Alternative random effects structures
- [ ] Alternative model families (e.g., negative binomial vs. Poisson)
- [ ] Sensitivity to outlier sites or clusters
:::

---

## Claim Calibration: Narrowing the Claim When Assumptions Are Weak

Not every analysis can sustain its intended claim. When assumptions are questionable, the right response is to narrow the claim — not to hide the limitation.

| If this is weak... | Standard claim | Narrowed claim |
|--------------------|----------------|----------------|
| Stimulus sample is small or non-representative | "This framing effect generalizes to all messages of this type" | "This effect held for the specific stimuli tested; generalizability to other messages is unknown" |
| Sites are convenience-sampled | "This effect generalizes across contexts" | "This effect was observed across [K] [convenience-sampled] sites; generalization to other contexts requires similar recruitment" |
| Compliance is partial and exclusion restriction is debatable | "Tool use causally improves savings (LATE)" | "Assignment to the tool offer increased savings rates (ITT); the as-treated comparison is biased by motivation; IV assumptions require further defense" |
| Attrition is potentially MNAR | "IPW-corrected ITT is unbiased" | "The IPW correction assumes MAR; if high-risk participants dropped out because of worsening outcomes (MNAR), residual bias is possible" |
| Parallel trends assumption is untestable with only two waves | "DiD estimate is causal" | "The DiD estimate is consistent with a causal interpretation under the parallel trends assumption, which cannot be directly tested with two waves" |
| Random effects structure was simplified from preregistered | "Results are confirmatory" | "Results are confirmatory for the primary model; the preregistered maximal model was simplified due to convergence failure (reported in Appendix)" |

---

### Example APA-Style Results Paragraph

> "We analyzed donation decisions using a logistic mixed model with participant random intercepts, framing condition (gain vs. loss) as a fixed effect, and trial nested within participant. Loss-framing significantly increased donation probability (b = 0.59, OR = 1.81, 95% CI [1.42, 2.31], $z$ = 4.7, $p$ < .001). At the sample's average baseline donation rate (38%), loss framing increased donation probability by approximately 12 percentage points. The ICC was .48, indicating substantial stable individual differences in donation tendency; the random intercept accounted for 31% of total variance (conditional $R^2_c$ = .35; marginal $R^2_m$ = .04)."

---

## Part 6: Reproducibility Practices

### Setting Seeds

Simulations, multiple imputation, bootstrapped CIs, and permutation tests all depend on random number generation. A fixed seed ensures the analysis is reproducible:

```{r seed-demo, eval=FALSE}
# Set seed at the top of every analysis script
set.seed(42)

# For parallel computing, set seeds for each worker explicitly
library(parallel)
RNGkind("L'Ecuyer-CMRG")  # safe for parallel streams
set.seed(42)
```

A single `set.seed()` call at the top of the script is sufficient for non-parallel analyses. Document the seed in the manuscript methods section.

### Package Version Pinning

Results can change between package versions (model fitting algorithms improve, bugs are fixed, defaults change). Record the environment at the time of analysis:

```{r session-info, eval=FALSE}
# At the end of every analysis script:
sessionInfo()

# Or more concisely:
writeLines(capture.output(sessionInfo()), "session_info.txt")
```

For fully reproducible environments, use `renv`:

```r
# Initialize renv in the project root
renv::init()

# After installing/updating packages:
renv::snapshot()  # saves renv.lock with exact package versions

# To restore the environment on a new machine:
renv::restore()
```

### Preregistration for Experimental Analyses

Preregistration specifies your analysis plan before data collection, separating confirmatory from exploratory findings. For mixed model experiments, the preregistration should specify:

1. The exact model formula (including random effects structure)
2. The primary outcome variable
3. The family and link function (for GLMMs)
4. How multiple comparisons will be handled
5. Any planned sensitivity analyses (e.g., random slopes model as robustness check)

Deviations from the preregistered plan should be disclosed and labeled as exploratory.

---

## Part 7: Common Reporting Mistakes in Nested Data

::: {.callout-warning icon=false}
## Mistake 1: Reporting Only the Pooled Estimate

If you ran a random-intercepts model but forgot to decompose within- and between-person effects (Module 2), the fixed effect coefficient is a blend of two mechanisms. Report both if they differ.
:::

::: {.callout-warning icon=false}
## Mistake 2: Using Total SD for Within-Subjects d

Reporting Cohen's $d$ using the total SD (including between-person variance) dramatically understates the within-person effect. Use the residual SD from the mixed model for within-subjects standardization.
:::

::: {.callout-warning icon=false}
## Mistake 3: Interpreting ICC-Corrected SEs Incorrectly

Standard errors from `lmer` already account for the nested structure. Do not additionally "correct" them for clustering (e.g., by applying cluster-robust SEs on top of a random intercept model). That would double-count the clustering adjustment.
:::

::: {.callout-warning icon=false}
## Mistake 4: Reporting Conditional R² as the "Effect Size" of a Fixed Effect

Conditional $R^2$ includes variance explained by the random effects. It is not an effect size for your experimental manipulation. Use marginal $R^2$ for the fixed-effects explanation, or report individual $\eta^2$ or $d$ for specific contrasts.
:::

::: {.callout-warning icon=false}
## Mistake 5: Omitting the Random Effects Structure from the Methods

Reporting "we used a mixed model" without specifying the random effects structure (intercept only? random slopes for condition?) makes the analysis non-reproducible. Another researcher cannot verify or extend your work without knowing the exact model formula.
:::

---

## Part 8: What Goes in the Paper vs. the Appendix

One of the most practical reporting questions: which parts of the multilevel analysis belong in the main text, and which belong in a supplementary appendix?

::: {.callout-tip icon=false}
## Paper vs. Appendix Decision Guide

**Main text (always include):**
- The model formula (written out, not just stated as "mixed model")
- Fixed effect estimates with standard errors and *p*-values for all tested hypotheses
- ICC of the primary outcome
- Whether random slopes were tested for the focal predictor
- Effect size (conditional *d*, OR, RR) with confidence interval
- Software and key package versions

**Main text (include when substantively important):**
- Variance components ($\tau^2_{00}$, $\sigma^2_\varepsilon$) when the research question concerns between-unit variability
- Marginal and conditional R²
- Hausman test result (when FE vs. RE choice was non-obvious)
- Predicted probabilities or counts at theoretically meaningful covariate values (for GLMM)

**Appendix (include as standard robustness evidence):**
- Full model comparison table (null model, main model, saturated model)
- Alternative random effects specifications (e.g., random slopes added or removed)
- Sensitivity analysis results (e.g., influential cluster removal, model with and without outlier branches)
- Multiple imputation details and pooled results when missing data is non-trivial
- CMC decomposition results when pooled and within estimates differ

**Appendix (include when facing potential reviewer objections):**
- Fixed effects equivalent model output alongside the random effects model (when exchangeability may be questioned)
- Model diagnostics: residual QQ plots, random effects normality plots, variance homogeneity checks
- Alternative distributional families (Poisson vs. negative binomial for count outcomes)
- Falsification tests (for IV, RDD, DiD contexts)
:::

---

## Part 9: Preempting Common Reviewer Misunderstandings

Reviewers who are not mixed-model specialists raise predictable objections. Here is a field guide for the most common ones.

::: {.callout-note icon=false}
## Reviewer: "Why not just use repeated-measures ANOVA?"

**What they mean:** RM-ANOVA is the familiar tool for within-subjects designs.

**Why MLM is better:** RM-ANOVA requires complete cases (it drops participants with any missing observations). It cannot include continuous time-varying covariates. It assumes compound symmetry or sphericity (homogeneous variance across all time-lag pairs) — an assumption that rarely holds in practice. Mixed models handle all of these naturally.

**How to respond:** "We used a mixed model rather than RM-ANOVA because (a) 18% of participants had incomplete data at one or more waves, (b) we included continuous within-person time-varying covariates, and (c) preliminary Mauchly tests indicated sphericity was violated."
:::

::: {.callout-note icon=false}
## Reviewer: "Your ICC is small — why does clustering matter?"

**What they mean:** An ICC of 0.05 looks negligible.

**Why it still matters:** The relevant quantity is the design effect: DEFF = 1 + (cluster size − 1) × ICC. With 20 observations per cluster and ICC = 0.05, DEFF = 1.95 — your effective N is roughly half your nominal N, and standard errors are inflated by √1.95 ≈ 40%. A "small" ICC with large clusters is a large problem.

**How to respond:** "Although the ICC was modest (0.05), clusters contained an average of *n* = 22 observations each, producing a design effect of 2.05. Ignoring clustering would have underestimated standard errors by approximately 43% and inflated Type I error for the focal comparison."
:::

::: {.callout-note icon=false}
## Reviewer: "Why random effects rather than fixed effects? Did you test for this?"

**What they mean:** They are invoking the RE vs. FE concern from the Hausman tradition.

**Why the question matters:** If cluster-level random effects are correlated with predictors, RE estimates are biased. FE removes this problem at the cost of between-cluster comparisons.

**How to respond:** "We assessed the random effects assumption with the Hausman test [χ²(*df*) = X.XX, *p* = .XX], which did not support systematic differences between FE and RE estimates. Additionally, our primary predictor was experimentally assigned within clusters, ensuring orthogonality between treatment and cluster random effects by design."
:::

::: {.callout-note icon=false}
## Reviewer: "Why don't you include random slopes? Barr et al. (2013) say you should."

**What they mean:** They have read the "keep it maximal" literature.

**When they are right:** For a within-subjects predictor (every participant experiences every condition), a random slope for that predictor is theoretically warranted. Omitting it increases Type I error.

**When to push back:** Maximal models frequently fail to converge with real data. If the random slope variance was estimated near zero and the LRT comparing models was non-significant (*p* > .10), the simpler model is defensible.

**How to respond:** "We tested a model with a by-participant random slope for condition; the slope variance was estimated near zero (σ² = 0.003) and the likelihood ratio test did not support the more complex structure (χ²(2) = 0.84, *p* = .66). We retained the simpler random-intercept model following Matuschek et al. (2017)."
:::

::: {.callout-note icon=false}
## Reviewer: "Your random effects are not normally distributed — doesn't that invalidate the model?"

**What they mean:** They checked a caterpillar plot and saw non-normality in the BLUP residuals.

**Why this is usually not fatal:** The normality assumption in mixed models is on the *population distribution of random effects*, not on the finite-sample BLUPs. Small-to-moderate deviations from normality have limited impact on fixed-effect estimates and standard errors, especially with moderate to large numbers of clusters.

**How to respond:** "The random effects normality assumption was examined via QQ plots of the conditional modes (BLUPs). We observed [mild left skew / a few outlier clusters], but simulation studies (e.g., Maas & Hox, 2004) show that fixed-effect estimates in MLM are robust to moderate violations of the random-effects normality assumption. Results were substantively unchanged when [the three outlier clusters were excluded / a robust estimation method was used]."
:::
