---
title: "A Fully Worked Multilevel Modeling Example"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
set.seed(20240610)
```

# A Fully Worked Multilevel Modeling Example

This chapter does not introduce a new method. It shows what it looks like to *use* the methods from Parts 1–7 on a single realistic dataset, from first principles.

The voice is intentionally that of a supervision meeting. Every major decision arrives as a question, gets answered, and then gets tested. But this is not a clean textbook example — the data are simulated to include the kinds of problems a real JDM dataset actually has: nonlinear covariate effects, trial-order fatigue, missing outcomes, and an observation process that is not independent of the treatment or outcome.

> We simulate the kind of dataset a JDM researcher might actually receive: an intended clean experiment filtered through participant fatigue, missing responses, platform differences, bounded ratings, and mild implementation imbalance. Because this is a simulation, we know the latent truth. Because the observed data are messy, we still have to behave like real analysts.

The key teaching distinction running through this chapter:

> **Modeling choices minimize concerns; they rarely erase them.** Adding covariates reduces imbalance bias; it does not prove there is no unmeasured confound. Adjusting for missingness predictors reduces selection bias; it does not solve missing-not-at-random. The goal is to document what was done, what it addressed, and what remains.

---

## 1. The Research Question and Estimand

> **Does transparent fee disclosure increase willingness to adopt a financial product, and does that effect generalize across participants, product offers, and research sites?**

The **target estimand** is the average total effect of *assigned* fee transparency on willingness to adopt, averaged across the population of participants, product offers, and contexts similar to those sampled. This is an intention-to-treat style estimand — we want the effect of the transparency display, not the effect of comprehension or perceived clarity, which may be mechanisms. Controlling for comprehension or perceived transparency would change the estimand and should not be done in the main model.

This question contains several distinct sub-questions a pooled regression cannot answer:

- Effect for *these* participants (descriptive)
- Effect across the *population of participants* (requires participant random effects)
- Effect across the *population of product offers* (requires stimulus random effects)
- Effect across *different research sites* (requires site random effects)
- Effect conditional on *financial literacy* (requires the interaction)

---

## 2. Design Audit Before Modeling

| Design question | Answer in this study | Modeling implication |
|---|---|---|
| What is one row? | One participant's rating of one product offer | Trial-level response; multiple rows per participant |
| What is randomized? | Fee transparency display (within-participant) | Fixed effect for transparency |
| Does treatment vary within participant? | Yes — each sees both conditions | Candidate by-participant random slope |
| Are stimuli sampled? | Yes — 36 product offers from a broader set | Stimulus random intercept |
| Are participants crossed with stimuli? | Partially — each rates 12 of 36 | Crossed (partial) random effects |
| Are participants nested in sites? | Yes — 240 participants across 4 sites | Site random intercept |
| Are there stimulus covariates? | Risk level, expected value | Check functional form; control for balance |
| Is there a participant-level moderator? | Financial literacy (pre-treatment) | Fixed effect; interaction candidate |
| Is the outcome continuous? | 0–100 rating | Linear mixed model; check residuals and bounds |
| What missingness is plausible? | Higher risk → more non-response; late trials → fatigue | Missingness audit before modeling |
| Which variables are post-treatment? | Comprehension, perceived transparency | Exclude from main effect model |

: Design audit: these answers determine every subsequent modeling decision. {tbl-colwidths="[28,30,42]"}

---

## 3. Simulate the Latent DGP

The true data-generating process is clean and known — but the *observed* dataset will not be.

**DGP features:** nonlinear expected-value effect (diminishing returns), risk threshold effect, trial-order fatigue, participant random intercepts and slopes, stimulus and site random intercepts, heteroskedastic residuals (more variance for high-risk products).

```{r true-dgp-params}
dgp <- list(
  intercept           = 50,
  effect_transparency = 4,      # +4 points for transparent-fee display
  effect_risk         = -2.5,   # linear risk penalty
  effect_risk_high    = -2.0,   # extra penalty for very high risk (threshold)
  effect_ev_linear    = 2.0,    # standardised EV (expected_value_z), linear
  effect_ev_quad      = -0.8,   # diminishing returns on EV
  effect_literacy     = 2,
  interaction_tx_lit  = -1.5,   # transparency helps low-literacy more
  effect_fatigue      = -0.20,  # per trial: late trials get lower ratings
  sd_participant      = 8,
  sd_stimulus         = 5,
  sd_site             = 4,
  sd_slope_part       = 2,
  sd_residual         = 7
)
```

```{r simulate-latent}
n_participants  <- 240
n_stimuli       <- 36
n_sites         <- 4
trials_per_part <- 12

participants <- tibble(
  participant_id     = 1:n_participants,
  site_id            = rep(1:n_sites, each = n_participants / n_sites),
  financial_literacy = rnorm(n_participants, 0, 1),
  u0_participant     = rnorm(n_participants, 0, dgp$sd_participant),
  u1_participant     = rnorm(n_participants, 0, dgp$sd_slope_part)
)

site_effects <- tibble(
  site_id = 1:n_sites,
  u0_site = c(3, -2, 1, -2)
)

stimuli <- tibble(
  stimulus_id    = 1:n_stimuli,
  risk_level     = rnorm(n_stimuli, 0, 1),
  expected_value = rnorm(n_stimuli, 50, 15),
  u0_stimulus    = rnorm(n_stimuli, 0, dgp$sd_stimulus)
) |>
  mutate(
    expected_value_z = (expected_value - 50) / 15,
    risk_high        = as.integer(risk_level > 0.75)  # threshold effect
  )

trials <- participants |>
  select(participant_id, site_id, financial_literacy, u0_participant, u1_participant) |>
  cross_join(tibble(draw = 1:trials_per_part)) |>
  group_by(participant_id) |>
  mutate(
    stimulus_id = sample(1:n_stimuli, trials_per_part, replace = FALSE),
    trial_order = 1:trials_per_part
  ) |>
  ungroup() |>
  select(-draw) |>
  left_join(stimuli, by = "stimulus_id") |>
  left_join(site_effects, by = "site_id") |>
  mutate(
    transparency = as.integer(runif(n()) > 0.5),
    mu = dgp$intercept +
         dgp$effect_transparency  * transparency +
         dgp$effect_risk          * risk_level +
         dgp$effect_risk_high     * risk_high +
         dgp$effect_ev_linear     * expected_value_z +
         dgp$effect_ev_quad       * expected_value_z^2 +
         dgp$effect_literacy      * financial_literacy +
         dgp$interaction_tx_lit   * transparency * financial_literacy +
         dgp$effect_fatigue       * trial_order +
         u0_participant + u1_participant * transparency +
         u0_stimulus + u0_site,
    resid_sd = dgp$sd_residual * exp(0.12 * pmax(risk_level, 0)),
    adoption_intent = pmin(100, pmax(0, rnorm(n(), mu, resid_sd)))
  ) |>
  select(participant_id, site_id, stimulus_id, trial_order,
         transparency, risk_level, risk_high, expected_value, expected_value_z,
         financial_literacy, adoption_intent)
```

---

## 4. The Observation Process: How Real Data Get Messy

A real experiment does not hand you `adoption_intent` for every intended trial. Between the latent DGP and the analyst's dataset, several things happen:

- Some participants are on mobile devices, producing noisier or faster responses.
- Some trials are skipped because participants are fatigued or confused.
- Missingness is not random: it is higher for high-risk products, later trials, low-literacy participants, and mobile users.
- Some trials have failed attention checks — and those failures are partly caused by the treatment (transparent-fee displays are more complex to process).

```{r simulate-observed}
set.seed(20240611)  # separate seed for observation process

trials <- trials |>
  mutate(
    # Platform/device: more common at sites 2–3 (online platforms)
    device_mobile = rbinom(n(), 1, plogis(-0.3 + 0.6 * as.integer(site_id %in% c(2, 3)))),

    # Missingness: higher for risky products, transparent condition, late trials,
    # low-literacy participants, and mobile users
    p_missing = plogis(-3.5 +
                  0.40 * risk_level +
                  0.30 * transparency +
                  0.25 * as.integer(trial_order > 9) -
                  0.35 * financial_literacy +
                  0.20 * device_mobile),
    missing_y           = rbinom(n(), 1, p_missing),
    adoption_intent_obs = if_else(missing_y == 1, NA_real_, adoption_intent),

    # Attention failures: related to late trials, low literacy, and — importantly —
    # transparency (more complex displays cause more failures)
    attention_fail = rbinom(n(), 1,
                      plogis(-3.8 +
                             0.40 * as.integer(trial_order > 9) -
                             0.45 * financial_literacy +
                             0.18 * transparency +
                             0.15 * risk_level))
  ) |>
  select(-p_missing)
```

::: {.callout-note icon=false}
## The Data Flow

```{mermaid}
flowchart LR
  D["Intended design<br/>240 × 12 = 2,880 trials"] --> A["Assignment<br/>transparency randomized"]
  A --> R["Response process<br/>fatigue, attention, device"]
  R --> M["Observed data<br/>missing outcomes, attention flags"]
  M --> Mod["Analysis<br/>mixed-effects + diagnostics"]
```

The analysis model sees only `adoption_intent_obs`. The latent `adoption_intent` is never observed. Diagnostics tell us whether the gap between intended and observed data threatens the estimand.
:::

---

## 5. Data Intake Audit

Before fitting any model, build a picture of what data you actually have.

### 5.1 Data retention table

```{r retention-table}
n_intended  <- nrow(trials)
n_observed  <- sum(!is.na(trials$adoption_intent_obs))
n_attfail   <- sum(trials$attention_fail == 1, na.rm = TRUE)
n_analytic  <- sum(!is.na(trials$adoption_intent_obs) & trials$attention_fail == 0)

tibble(
  Stage = c("Intended (all planned trials)",
            "Non-missing outcome",
            "Attention failures (in observed rows)",
            "Clean analytic sample"),
  n          = c(n_intended, n_observed, n_attfail, n_analytic),
  pct_of_intended = round(100 * c(n_intended, n_observed,
                                   sum(trials$attention_fail), n_analytic) /
                            n_intended, 1)
) |>
  knitr::kable(caption = "Data flow: how many rows are available at each stage?")
```

### 5.2 Missingness by design variables

**Student question:** The mixed model handles unequal observations per participant automatically. So why does missingness matter?

**Advisor response:** Mixed models handle *unequal* observations correctly — they do not require balanced data. What they do *not* handle is **informative** missingness: when the probability of observing a trial is related to what the outcome would have been, complete-case analysis produces biased estimates. That is different from just having different numbers of rows per person.

```{r missingness-audit}
# Missingness rate by transparency
miss_by_tx <- trials |>
  group_by(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  summarise(
    n_trials  = n(),
    n_missing = sum(missing_y),
    miss_rate = round(100 * mean(missing_y), 1)
  )
knitr::kable(miss_by_tx, caption = "Missingness rate by transparency condition")

# Missingness by site
miss_by_site <- trials |>
  group_by(site_id) |>
  summarise(miss_rate = round(100 * mean(missing_y), 1), n = n())
knitr::kable(miss_by_site, caption = "Missingness rate by site")

# Missingness by trial order
miss_by_order <- trials |>
  group_by(trial_order) |>
  summarise(miss_rate = round(100 * mean(missing_y), 1))
knitr::kable(miss_by_order, caption = "Missingness rate by trial order — late-trial fatigue?")
```

```{r missingness-heatmap, fig.width=9, fig.height=4}
heatmap_ids <- sort(sample(unique(trials$participant_id), 50))
trials |>
  filter(participant_id %in% heatmap_ids) |>
  mutate(status = ifelse(is.na(adoption_intent_obs), "Missing", "Observed")) |>
  ggplot(aes(x = trial_order, y = factor(participant_id), fill = status)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_manual(values = c("Observed" = "#1e3a5f", "Missing" = "#e63946")) +
  scale_x_continuous(breaks = 1:trials_per_part) +
  labs(
    title    = "Missingness pattern: 50 sampled participants × trial position",
    subtitle = "Red = missing outcome. Clustering by participant or trial order signals non-random missingness.",
    x        = "Trial position", y = "Participant (sample)", fill = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

```{r missingness-literacy}
trials |>
  mutate(literacy_tertile = ntile(financial_literacy, 3)) |>
  group_by(literacy_tertile) |>
  summarise(miss_rate = round(100 * mean(missing_y), 1), n = n()) |>
  knitr::kable(caption = "Missingness by financial literacy tertile — low-literacy participants more likely to drop out")
```

**What we learned:** Missingness is higher in the transparent condition, in later trials, and for low-literacy participants. This is not random. A complete-case analysis will systematically underrepresent high-risk/late-trial/low-literacy observations. We will return to this when checking robustness (Section 15).

### 5.3 Outcome checks

```{r outcome-checks, fig.width=9, fig.height=3.5}
p1 <- ggplot(trials, aes(x = adoption_intent_obs)) +
  geom_histogram(bins = 30, fill = "#1e3a5f", color = "white", na.rm = TRUE) +
  labs(title = "Distribution of adoption_intent_obs", x = "Adoption intent (0–100)") +
  theme_minimal(base_size = 10)

p2 <- trials |>
  filter(!is.na(adoption_intent_obs)) |>
  mutate(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  ggplot(aes(x = Condition, y = adoption_intent_obs, fill = Condition)) +
  geom_boxplot(show.legend = FALSE, width = 0.5) +
  scale_fill_manual(values = c("#cccccc", "#1e3a5f")) +
  labs(title = "By condition", y = "Adoption intent", x = NULL) +
  theme_minimal(base_size = 10)

p3 <- trials |>
  filter(!is.na(adoption_intent_obs)) |>
  ggplot(aes(x = trial_order, y = adoption_intent_obs, group = trial_order)) +
  geom_boxplot(fill = "#2d6a4f", alpha = 0.6, outlier.size = 0.5) +
  labs(title = "By trial order — fatigue?", y = NULL, x = "Trial") +
  theme_minimal(base_size = 10)

gridExtra::grid.arrange(p1, p2, p3, ncol = 3)

# Floor/ceiling check
cat("Proportion at floor (≤2):", round(mean(trials$adoption_intent_obs <= 2, na.rm=TRUE), 3), "\n")
cat("Proportion at ceiling (≥98):", round(mean(trials$adoption_intent_obs >= 98, na.rm=TRUE), 3), "\n")
```

---

## 6. Assignment and Post-Treatment Audit

### 6.1 Was assignment actually balanced?

Even with randomization, the *realized* assignment can be unbalanced. An assignment audit fits a logistic model predicting the treatment indicator from pre-treatment variables. A well-randomized assignment should produce no predictors with large, systematic coefficients.

```{r assignment-audit}
tx_audit <- glm(
  transparency ~ risk_level + expected_value_z + financial_literacy +
    factor(site_id) + trial_order + device_mobile,
  data  = trials,
  family = binomial()
)

broom::tidy(tx_audit) |>
  filter(term != "(Intercept)") |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = paste0(
    "Assignment audit: does any pre-treatment variable predict transparency assignment? ",
    "Large coefficients (|b| > 0.2) or small p-values suggest implementation imbalance."))
```

**What we learned:** The assignment audit tells us which variables, if any, are unbalanced and need to be controlled in the model. If the intercept-only model fits as well as the full model, randomization worked in the realized sample.

### 6.2 Which variables belong in the main model?

Not every variable should enter the main model. Some are pre-treatment design features (safe to include). Some are post-treatment measurements or mechanisms (including them changes the estimand).

```{r timing-table}
tibble(
  Variable = c("Risk level", "Expected value", "Financial literacy",
               "Trial order", "Device type",
               "Attention check score", "Perceived transparency",
               "Comprehension score"),
  Timing   = c("Pre-treatment stimulus attribute", "Pre-treatment stimulus attribute",
               "Pre-treatment participant trait",
               "Design/process variable", "Pre-treatment contextual",
               "Post-trial, ambiguous", "Post-treatment mediator",
               "Post-treatment mediator"),
  `Safe in main model?` = c("Yes", "Yes", "Yes",
                             "Yes if imbalance detected",
                             "Yes as robustness covariate",
                             "Caution",
                             "No — changes estimand",
                             "No — changes estimand"),
  Why = c(
    "Stimulus feature not caused by transparency display",
    "Stimulus feature not caused by transparency display",
    "Measured before any trial",
    "Could confound if transparency correlates with position",
    "Affects response quality and missingness; not caused by transparency",
    "May reflect difficulty of processing the transparency display",
    "Blocking a mechanism changes total-effect estimand",
    "Blocking a mechanism changes total-effect estimand"
  )
) |>
  knitr::kable(caption = "Variable timing audit: what should enter the main model?")
```

::: {.callout-warning icon=false}
## Post-Treatment Controls Change the Estimand

```{mermaid}
flowchart LR
  T["Fee transparency\n(assigned)"] --> C["Comprehension /\nperceived clarity"]
  C --> Y["Adoption intent"]
  T --> Y
  L["Financial literacy"] --> C
  L --> Y
```

Transparency may work *by* increasing comprehension of fees. If we control for comprehension, we estimate only the part of the transparency effect that does not flow through comprehension — a different and narrower claim than "does the display change adoption intent?" The main model should not include comprehension or perceived transparency.
:::

---

## 7. Start With the Wrong Model

```{r m0-naive}
m0 <- lm(adoption_intent_obs ~ transparency, data = trials)

cat("Rows in dataset:", nrow(trials), "\n")
cat("Rows used by m0:", nobs(m0), "\n")
cat("Rows silently dropped (missing outcome):", nrow(trials) - nobs(m0), "\n\n")

broom::tidy(m0) |>
  filter(term == "transparency") |>
  mutate(model = "m0: pooled OLS (complete cases)") |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "Naive pooled OLS: transparency estimate")
```

**Student question:** The model ran, it gave a significant result, and I can see the estimate. What's wrong with it?

**Advisor response:** Several things. First — did you notice it silently dropped rows? Every model using `adoption_intent_obs` drops missing rows without telling you. You need to check that explicitly. Second, the model treats all 2,500-odd remaining rows as independent. They are not: the same participants appear repeatedly, the same stimuli appear repeatedly, and participants are nested in sites. In this within-person design, ignoring participant baselines actually *hurts* precision — participant-level variance stays in the residual and dilutes the within-person signal. Third, the model assumes risk and expected value have linear effects — we know from the DGP that they do not. Fourth, it ignores trial-order fatigue. And fifth, it assumes missingness is irrelevant.

```{r m0-diagnosis}
tibble(
  Problem = c("Silent row dropping", "Independence assumption", "Participant baselines in residual",
              "Stimulus variance ignored", "Nonlinear covariate effects", "Trial-order fatigue",
              "Informative missingness"),
  `What m0 does` = c(
    "Drops missing rows without warning",
    "Treats ~2,500 rows as independent",
    "Person-to-person differences inflate residual SD",
    "Product-level attractiveness absorbed into residual",
    "Risk and EV treated as linear",
    "No adjustment for late-trial fatigue",
    "Complete cases treated as a random subsample"
  ),
  `What could go wrong` = c(
    "If dropped rows are not random, estimate may be biased",
    "SE is wrong for clustered data",
    "Less efficient estimation of within-person effect",
    "Cannot generalize to new product offers",
    "Covariate misspecification may bias transparency estimate",
    "If transparency is unbalanced across trial order, confound exists",
    "If missingness depends on treatment or outcome, bias follows"
  )
) |>
  knitr::kable(caption = "Why m0 misrepresents the design — a checklist of problems")
```

---

## 8. Add Participant Random Intercepts

```{r m1}
m1 <- lmer(adoption_intent_obs ~ transparency + (1 | participant_id), data = trials)
```

```{r m1-summary}
vc1 <- as.data.frame(VarCorr(m1))
var_part <- vc1$vcov[vc1$grp == "participant_id"]
var_res1 <- vc1$vcov[vc1$grp == "Residual"]
icc_part <- var_part / (var_part + var_res1)
cat("Participant ICC:", round(icc_part, 3), "\n")

broom.mixed::tidy(m1, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m1: + participant RI") |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)
```

**Student question:** The SE went *down* when I added random intercepts. I expected it to go up.

**Advisor response:** This is the within-person design working correctly. In a between-subjects design, clustering means the effective N is smaller than the raw N, and random intercepts inflate the SE. Here, transparency is randomized *within* participants — each person sees both conditions. The participant random intercept removes their stable adoption tendency from the residual. With less residual noise, the within-person comparison is estimated more precisely. The SE shrinks not because we got conservative, but because the model separated signal from noise correctly.

**What is still wrong:** Stimuli are still treated as fixed. Product attractiveness differences are sitting in the residual. We cannot generalize beyond these 36 offers.

---

## 9. Add Stimulus Random Intercepts

```{r m2}
m2 <- lmer(adoption_intent_obs ~ transparency +
             (1 | participant_id) + (1 | stimulus_id),
           data = trials)
```

```{r m2-summary-and-plot, fig.width=7, fig.height=4}
vc2       <- as.data.frame(VarCorr(m2))
var_stim2 <- vc2$vcov[vc2$grp == "stimulus_id"]
icc_stim  <- var_stim2 / (var_stim2 + vc2$vcov[vc2$grp == "Residual"])
cat("Stimulus ICC:", round(icc_stim, 3), "\n")
cat("Stimulus SD:", round(sqrt(var_stim2), 2),
    "(true =", dgp$sd_stimulus, "; transparency effect is", dgp$effect_transparency, ")\n")

broom.mixed::tidy(m2, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m2: + stimulus RI") |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)

# Caterpillar plot
ranef(m2)$stimulus_id |>
  as.data.frame() |>
  rownames_to_column("id") |>
  rename(blup = `(Intercept)`) |>
  arrange(blup) |>
  ggplot(aes(x = reorder(id, blup), y = blup)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#888888") +
  geom_segment(aes(xend = id, y = 0, yend = blup), color = "#1e3a5f", linewidth = 0.5, alpha = 0.7) +
  geom_point(color = "#1e3a5f", size = 1.8) +
  coord_flip() +
  labs(title = "Stimulus random intercepts: some products are inherently more attractive",
       subtitle = paste0("SD ≈ ", round(sqrt(var_stim2), 1),
                         " points — larger than the transparency effect itself."),
       x = NULL, y = "Estimated baseline deviation (BLUP)") +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

::: {.callout-note icon=false}
## m2 Is the Aha Moment

The stimulus SD is larger than the transparency effect. This means that if we had used a different set of 36 product offers, we could have seen a meaningfully different average transparency effect — not because the psychology changed, but because our particular sample happened to skew toward naturally attractive or unattractive products.

The crossed model says: given this kind of between-product variance, the transparency effect is estimated as `r round(fixef(m2)["transparency"], 2)` on average across the population of such products. That is a different and more honest claim than "the effect is X for these 36 products."
:::

---

## 10. Add Site Structure

```{r m3}
m3 <- lmer(adoption_intent_obs ~ transparency +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)
```

```{r m3-summary}
vc3 <- as.data.frame(VarCorr(m3))
cat("Site SD:", round(sqrt(vc3$vcov[vc3$grp == "site_id"]), 2),
    "(true =", dgp$sd_site, ")\n")

trials |>
  group_by(site_id) |>
  summarise(mean_adoption = round(mean(adoption_intent_obs, na.rm = TRUE), 1),
            n_part = n_distinct(participant_id)) |>
  knitr::kable(caption = "Site-specific means confirm real between-site differences")

broom.mixed::tidy(m3, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m3: + site RI") |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)
```

**Student question:** Adding site barely changed the estimate or the SE. Did it do anything?

**Advisor response:** Yes — but not by changing the estimate. It did two things: (1) it correctly partitioned between-site baseline differences out of the participant random intercept, which was absorbing site-level noise before; and (2) it shifted the estimand. Before, the claim was "effect across these participants in these four sites." After, the claim is "average effect across a population of sites similar to these four." Site differences here are baseline differences, not transparency confounds — that is why the estimate did not change. If the transparency effect itself varied across sites, we would need a random slope for transparency by site. With only four sites, that slope would be unreliably estimated.

---

## 11. Add Design Covariates

Risk level, expected value, and financial literacy are pre-treatment variables. Including them adjusts for any realized imbalance in the randomization, even when that imbalance is small.

```{r m4}
m4 <- lmer(adoption_intent_obs ~ transparency + risk_level + expected_value_z +
             financial_literacy +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)

broom.mixed::tidy(m4, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m4: + pre-treatment covariates (linear)") |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)
```

::: {.callout-note icon=false}
## Stable Estimates Are a Scientific Result

The transparency estimate has barely moved from m0 through m4. That is not a failure — it is evidence that the design was clean: transparency was not confounded with participant characteristics, product features, or site. The modeling sequence still matters because it answers different questions about precision, generalization, and estimand.

If the estimate *had* changed when covariates were added, it would tell us there was realized imbalance and the covariate model is correcting for it.
:::

---

## 12. Functional-Form Check

**Student question:** We added risk level and expected value as linear terms. But the DGP has a threshold risk effect and diminishing EV returns. Does getting the functional form wrong matter?

**Advisor response:** It can. If the linear covariate model misspecifies the risk or EV relationship, it may absorb some of that misspecification into the transparency estimate. The diagnostic is to plot residuals against the covariates and compare a flexible model to the linear one.

```{r functional-form-diag, fig.width=9, fig.height=3.5}
resid_m4 <- tibble(
  residual   = residuals(m4),
  risk_level = model.frame(m4)$risk_level,
  ev_z       = model.frame(m4)$expected_value_z,
  trial_ord  = trials$trial_order[!is.na(trials$adoption_intent_obs)]
)

p_r <- ggplot(resid_m4, aes(x = risk_level, y = residual)) +
  geom_point(alpha = 0.1, color = "#1e3a5f") +
  geom_smooth(method = "loess", se = FALSE, color = "#e63946", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs risk level", subtitle = "Curve = loess smoother",
       x = "Risk level", y = "Residual") +
  theme_minimal(base_size = 10)

p_e <- ggplot(resid_m4, aes(x = ev_z, y = residual)) +
  geom_point(alpha = 0.1, color = "#1e3a5f") +
  geom_smooth(method = "loess", se = FALSE, color = "#e63946", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs expected value (std.)", x = "Expected value (z)", y = "Residual") +
  theme_minimal(base_size = 10)

p_t <- ggplot(resid_m4, aes(x = trial_ord, y = residual, group = trial_ord)) +
  geom_boxplot(fill = "#1e3a5f", alpha = 0.5, outlier.size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs trial order", x = "Trial", y = "Residual") +
  theme_minimal(base_size = 10)

gridExtra::grid.arrange(p_r, p_e, p_t, ncol = 3)
```

```{r m4b}
# Nonlinear model: quadratic EV, threshold risk indicator, trial order
m4b <- lmer(adoption_intent_obs ~ transparency +
              risk_level + risk_high +
              expected_value_z + I(expected_value_z^2) +
              financial_literacy + trial_order +
              (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
            data = trials)

bind_rows(
  broom.mixed::tidy(m4,  effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m4: linear risk/EV, no trial order"),
  broom.mixed::tidy(m4b, effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m4b: nonlinear risk/EV + trial order")
) |>
  select(model, estimate, std.error, p.value) |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = paste0("Does nonlinear functional form + trial order change the transparency estimate?",
    " True effect = ", dgp$effect_transparency))
```

**What we learned:** If the transparency estimate shifts between m4 and m4b, the linear covariate model was absorbing covariate misspecification into the treatment estimate. The nonlinear model is more honest about the product-valuation process.

**A substantive note on functional form:** A linear expected-value term says each additional dollar of EV matters equally. In most JDM settings, that is implausible — people have diminishing sensitivity to expected value, especially at higher magnitudes. The quadratic term captures this psychologically motivated nonlinearity. Choosing a functional form is not only a statistical decision; it is a claim about the decision process.

---

## 13. Add the Theoretically Motivated Interaction

The hypothesis: fee transparency helps low-literacy participants more because they struggle to infer fees from product descriptions without explicit disclosure.

```{r m5}
m5 <- lmer(adoption_intent_obs ~ transparency * financial_literacy +
             risk_level + risk_high +
             expected_value_z + I(expected_value_z^2) +
             trial_order +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)

broom.mixed::tidy(m5, effects = "fixed") |>
  select(term, estimate, std.error, statistic, p.value) |>
  mutate(across(c(estimate, std.error, statistic, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = "m5: transparency × financial literacy interaction")
```

```{r m5-simple-slopes}
b_tx  <- fixef(m5)["transparency"]
b_int <- fixef(m5)["transparency:financial_literacy"]
tibble(
  `Financial literacy` = c("Low (−1 SD)", "Average (0)", "High (+1 SD)"),
  `Transparency effect` = round(c(b_tx + b_int * -1,
                                   b_tx + b_int *  0,
                                   b_tx + b_int *  1), 2)
) |>
  knitr::kable(caption = paste0("Simple slopes. True: low = ",
    dgp$effect_transparency + dgp$interaction_tx_lit * -1, ", avg = ",
    dgp$effect_transparency, ", high = ",
    dgp$effect_transparency + dgp$interaction_tx_lit * 1))
```

```{r m5-plot, fig.width=6.5, fig.height=4}
lit_vals <- seq(-2, 2, length.out = 100)
pred_df  <- tibble(
  financial_literacy = rep(lit_vals, 2),
  transparency       = rep(c(0L, 1L), each = 100),
  risk_level = 0, risk_high = 0L, expected_value_z = 0,
  trial_order = 6,
  participant_id = NA_integer_, site_id = NA_integer_, stimulus_id = NA_integer_
)
pred_df$predicted <- predict(m5, newdata = pred_df, re.form = NA)

pred_df |>
  mutate(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  ggplot(aes(x = financial_literacy, y = predicted,
             color = Condition, linetype = Condition)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("#cccccc", "#1e3a5f")) +
  labs(title = "Transparency effect moderated by financial literacy",
       x = "Financial literacy (standardised)", y = "Predicted adoption intent",
       color = NULL, linetype = NULL) +
  theme_minimal(base_size = 11)
```

::: {.callout-note icon=false}
## m5 Is the First Model That Changes the Substantive Claim

Models m0 through m4b answered: "Is there a transparency effect, and is the estimate credible?"

m5 answers: "For whom does the effect hold?" After m5, the transparency "main effect" is the effect at average financial literacy, not a universal constant. This changes what researchers should say in their discussion — and which populations policy-makers should target.
:::

---

## 14. Consider Random Slopes

```{r m6}
m6 <- lmer(adoption_intent_obs ~ transparency * financial_literacy +
             risk_level + risk_high +
             expected_value_z + I(expected_value_z^2) +
             trial_order +
             (1 + transparency | participant_id) +
             (1 | stimulus_id) + (1 | site_id),
           data = trials,
           control = lmerControl(optimizer = "bobyqa"))

cat("Singular (correlated slope):", isSingular(m6), "\n")
```

```{r m6-uncorrelated}
m6b <- lmer(adoption_intent_obs ~ transparency * financial_literacy +
              risk_level + risk_high +
              expected_value_z + I(expected_value_z^2) +
              trial_order +
              (1 + transparency || participant_id) +
              (1 | stimulus_id) + (1 | site_id),
            data = trials,
            control = lmerControl(optimizer = "bobyqa"))

cat("Singular (uncorrelated slope):", isSingular(m6b), "\n")
```

```{r m6-slope-viz, fig.width=6.5, fig.height=3.5}
# Empirical per-participant transparency effects
emp_slopes <- trials |>
  filter(!is.na(adoption_intent_obs)) |>
  group_by(participant_id) |>
  summarise(
    mean_tx   = mean(adoption_intent_obs[transparency == 1], na.rm = TRUE),
    mean_ntx  = mean(adoption_intent_obs[transparency == 0], na.rm = TRUE),
    n_tx = sum(transparency == 1), n_ntx = sum(transparency == 0)
  ) |>
  filter(n_tx > 0, n_ntx > 0) |>
  mutate(emp_effect = mean_tx - mean_ntx)

vc6b <- as.data.frame(VarCorr(m6b))
slope_row <- vc6b[!is.na(vc6b$var1) & vc6b$var1 == "transparency" & is.na(vc6b$var2), ]
slope_sd_val <- if (nrow(slope_row) > 0) sqrt(slope_row$vcov[1]) else NA_real_

ggplot(emp_slopes, aes(x = emp_effect)) +
  geom_histogram(bins = 25, fill = "#1e3a5f", color = "white", alpha = 0.8) +
  geom_vline(xintercept = fixef(m6b)["transparency"],
             color = "#e63946", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of participant-level transparency effects",
       subtitle = paste0("Red = model average. Estimated slope SD ≈ ",
                         ifelse(is.na(slope_sd_val), "NA", round(slope_sd_val, 1)),
                         " points."),
       x = "Empirical transparency effect (transparent − non-transparent)",
       y = "Count") +
  theme_minimal(base_size = 11)
```

::: {.callout-warning icon=false}
## If the Random Slope Is Singular

1. Remove the correlation parameter first (`||`). This is less demanding and often resolves boundary issues.
2. If still singular, the slope variance may be genuinely negligible — report it and retain m5.
3. Never silently drop the random slope. Report the simplification path.

**Decision rule:** Keep the slope if (a) non-singular, (b) slope SD is substantively non-trivial, and (c) it does not destabilize fixed-effect inference. Use `isTRUE()` guards in code to avoid `if(NA)` errors.
:::

```{r m6-comparison}
bind_rows(
  broom.mixed::tidy(m5,  effects = "fixed") |>
    filter(term %in% c("transparency", "transparency:financial_literacy")) |>
    mutate(model = "m5: no slope"),
  broom.mixed::tidy(m6b, effects = "fixed") |>
    filter(term %in% c("transparency", "transparency:financial_literacy")) |>
    mutate(model = "m6b: uncorrelated slope")
) |>
  select(model, term, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) |>
  knitr::kable(caption = "Does the random slope change fixed-effect inference?")
```

---

## 15. Missingness Sensitivity

**Student question:** We've been running complete-case models throughout. Is that OK?

**Advisor response:** It depends on why data are missing. The intake audit showed that missingness is higher in the transparent condition and for low-literacy participants. If the transparency effect is smaller for the kinds of trials that tend to go missing — and those trials are more likely to be transparent — the complete-case estimate could be biased upward.

**Diagnostic move:** Compare the complete-case model to a model that includes the missingness-related predictors (trial order, device, risk). If including them changes the transparency estimate, the complete-case analysis was affected by differential missingness.

```{r m-miss-adj}
# Missingness-adjusted: add all missingness predictors to the model
# (same complete cases, but missingness-correlated variables are now controlled)
m_miss_adj <- lmer(
  adoption_intent_obs ~ transparency * financial_literacy +
    risk_level + risk_high +
    expected_value_z + I(expected_value_z^2) +
    trial_order + device_mobile +
    (1 + transparency || participant_id) +
    (1 | stimulus_id) + (1 | site_id),
  data   = trials,
  control = lmerControl(optimizer = "bobyqa")
)

bind_rows(
  broom.mixed::tidy(m6b,       effects = "fixed") |>
    filter(term == "transparency") |>
    mutate(model = "m6b: complete-case, without device_mobile",
           n_obs = nobs(m6b)),
  broom.mixed::tidy(m_miss_adj, effects = "fixed") |>
    filter(term == "transparency") |>
    mutate(model = "m_miss_adj: + device_mobile (missingness predictor)",
           n_obs = nobs(m_miss_adj))
) |>
  select(model, n_obs, estimate, std.error, p.value) |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = paste0(
    "Missingness sensitivity: does adjusting for missingness-related predictors ",
    "change the transparency estimate? True effect = ", dgp$effect_transparency))
```

**What we learned:** If the estimate changed when missingness predictors were added, differential dropout was biasing the complete-case estimate. If it is stable, complete-case analysis is more defensible — though not proven ignorable. We report both estimates and the direction of any change.

::: {.callout-warning icon=false}
## What This Sensitivity Analysis Can and Cannot Do

Adjusting for observed missingness predictors reduces bias *if* missingness is explainable by those variables (missing at random conditional on covariates). It does **not** solve missing-not-at-random: if participants with low latent adoption intent are especially likely to skip high-risk transparent trials — and that tendency is unobservable — no covariate adjustment reaches it.

The honest statement: "We tested whether missingness-related variables changed our estimate. [Result]. The remaining concern is that missingness may depend on latent adoption intent, which we cannot fully rule out."
:::

---

## 16. Estimator and Inference Decisions

::: {.callout-note icon=false}
## Key Estimator Choices

| Decision | Default | When to change |
|---|---|---|
| REML vs ML | REML for final model | Use ML (REML=FALSE) to compare *fixed-effect* structures |
| df method | Satterthwaite (lmerTest) | Kenward-Roger for small or unbalanced samples |
| Optimizer | nloptwrap (lme4 default) | Try `bobyqa` for convergence warnings |
| LRT for random slopes | Boundary-aware (one-sided test, df=1) | Parametric bootstrap if LRT misbehaves |
:::

---

## 17. Select and Diagnose the Final Model

```{r final-model-selection}
slope_sd_check <- tryCatch({
  vc <- as.data.frame(VarCorr(m6b))
  tx <- vc[!is.na(vc$var1) & vc$var1 == "transparency" & is.na(vc$var2), ]
  if (nrow(tx) > 0) sqrt(tx$vcov[1]) else NA_real_
}, error = function(e) NA_real_)

keep_slope <- isTRUE(!isSingular(m6b)) &&
              isTRUE(!is.na(slope_sd_check)) &&
              isTRUE(slope_sd_check > 0.5)

m_final <- if (keep_slope) m_miss_adj else m5

cat("Slope SD:", if (is.na(slope_sd_check)) "NA" else round(slope_sd_check, 2), "\n")
cat("Singular (m6b):", isSingular(m6b), "\n")
cat("Final model selected:", if (keep_slope) "m_miss_adj (missingness-adjusted + slope)" else "m5 (slope singular, no missingness adjustment)", "\n")
```

```{r residual-diagnostics, fig.width=10, fig.height=3.5}
resid_df <- tibble(
  fitted   = fitted(m_final),
  residual = residuals(m_final),
  site_id  = trials$site_id[!is.na(trials$adoption_intent_obs)],
  tx       = trials$transparency[!is.na(trials$adoption_intent_obs)]
)

p_rv <- ggplot(resid_df, aes(x = fitted, y = residual)) +
  geom_point(alpha = 0.15, color = "#1e3a5f") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#e63946") +
  labs(title = "Residuals vs Fitted", x = "Fitted", y = "Residual") +
  theme_minimal(base_size = 10)

p_qq <- ggplot(resid_df, aes(sample = residual)) +
  stat_qq(alpha = 0.2, color = "#1e3a5f") +
  stat_qq_line(color = "#e63946") +
  labs(title = "QQ Plot", x = "Theoretical", y = "Sample") +
  theme_minimal(base_size = 10)

p_site <- ggplot(resid_df, aes(x = factor(site_id), y = residual)) +
  geom_boxplot(fill = "#2d6a4f", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals by site", x = "Site", y = "Residual") +
  theme_minimal(base_size = 10)

gridExtra::grid.arrange(p_rv, p_qq, p_site, ncol = 3)
```

---

## 18. Model Comparison and Robustness Table

```{r what-changed-table}
tibble(
  Model    = c("m0 → m1", "m1 → m2", "m2 → m3", "m3 → m4", "m4 → m4b", "m4b → m5", "m5 → m6b", "m6b → m_miss_adj"),
  `What changed?` = c(
    "SE decreased; estimate stable",
    "SE decreased again; estimate stable",
    "Estimand changed; estimate stable",
    "Estimate checked for imbalance",
    "Functional form corrected",
    "Substantive interpretation changed",
    "SE increased modestly",
    "Checked for missingness influence"
  ),
  `Main insight` = c(
    "Within-person design: removing stable baselines sharpens contrast",
    "Product attractiveness is large; generalization needs stimulus modeling",
    "Site baseline differences partitioned; estimand now population-level",
    "Clean randomization: covariates document robustness, not fix bias",
    "Nonlinear risk and EV — linear model misfit the product-valuation process",
    "Transparency effect is larger for low-literacy participants",
    "Participants differ in responsiveness; slope SD ≈ true DGP value",
    "Missingness predictors checked; stable estimate = less missingness concern"
  ),
  `Changed the claim?` = c("No", "No", "Estimand", "No", "Functional form", "Yes", "Heterogeneity", "Robustness check")
) |>
  knitr::kable(caption = "What changed at each modeling step and why it matters")
```

```{r robustness-table}
extract_tx <- function(m, label, is_lm = FALSE) {
  if (is_lm) {
    broom::tidy(m) |> filter(term == "transparency") |>
      select(estimate, std.error) |> mutate(model = label, n_obs = nobs(m))
  } else {
    broom.mixed::tidy(m, effects = "fixed") |> filter(term == "transparency") |>
      select(estimate, std.error) |> mutate(model = label, n_obs = nobs(m))
  }
}

bind_rows(
  extract_tx(m0,        "m0: pooled OLS",                             is_lm = TRUE),
  extract_tx(m2,        "m2: crossed RE baseline"),
  extract_tx(m4,        "m4: + design covariates (linear)"),
  extract_tx(m4b,       "m4b: + nonlinear risk/EV + trial_order"),
  extract_tx(m5,        "m5: + literacy interaction"),
  extract_tx(m6b,       "m6b: + random slope"),
  extract_tx(m_miss_adj,"m_miss_adj: + device (missingness sensitivity)")
) |>
  mutate(
    across(c(estimate, std.error), \(x) round(x, 3)),
    `Threat addressed` = c(
      "None",
      "Participant + stimulus clustering",
      "Realized covariate imbalance",
      "Functional-form misspecification + fatigue",
      "Treatment-effect heterogeneity by literacy",
      "Residual participant-level slope heterogeneity",
      "Differential missingness by condition/device"
    ),
    `What remains?` = c(
      "Everything",
      "Site, covariates, nonlinearity, missingness",
      "Nonlinear effects, missingness",
      "Missingness, slope heterogeneity",
      "Slope heterogeneity, missingness",
      "Missingness sensitivity untested",
      "Unobserved missingness drivers (MNAR)"
    )
  ) |>
  select(model, n_obs, estimate, std.error, `Threat addressed`, `What remains?`) |>
  knitr::kable(caption = paste0(
    "Robustness table: each model is tied to the threat it addresses. ",
    "True effect = ", dgp$effect_transparency, " at average literacy."))
```

---

## 19. Final Interpretation

```{r final-interp}
fe_final <- broom.mixed::tidy(m_final, effects = "fixed") |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3)))
knitr::kable(fe_final |> select(term, estimate, std.error, statistic, p.value),
             caption = "Final model: fixed effects")

vc_final <- as.data.frame(VarCorr(m_final))
knitr::kable(vc_final |> select(grp, var1, vcov) |>
               mutate(SD = round(sqrt(vcov), 2), vcov = round(vcov, 2)),
             caption = "Final model: variance components")
```

**Plain-language interpretation:**

- **Transparency effect at average literacy:** Fee transparency increased willingness to adopt by approximately `r round(fixef(m_final)["transparency"], 1)` points (true DGP: `r dgp$effect_transparency`). The model recovered the effect accurately despite the nonlinear covariates and missing data.
- **Literacy moderation:** The interaction confirms that low-literacy participants benefit more from transparency. After m5, the "main effect" is the effect at average literacy — not a universal constant.
- **Participant variance:** Stable individual differences in adoption tendency are large. Random intercepts for participants correctly separate this from the within-person transparency contrast.
- **Stimulus variance:** Between-product attractiveness differences are larger than the transparency effect. This confirms that the stimulus set matters for generalization.
- **Missingness sensitivity:** When device platform (a missingness predictor) was added, the transparency estimate shifted from `r round(fixef(m6b)["transparency"], 2)` to `r round(fixef(m_miss_adj)["transparency"], 2)` points (difference: `r round(abs(fixef(m_miss_adj)["transparency"] - fixef(m6b)["transparency"]), 2)` points). A small shift suggests complete-case analysis captured the main effect with minimal differential-dropout bias; unobserved MNAR drivers remain a caveat.

---

## 20. What a Reviewer Would Ask

**1. "How much data were dropped, and why?"**

Show the data retention table from Section 5.1. State explicitly: missing outcomes were more common in the transparent condition and later trials. The missingness sensitivity check in Section 15 addresses this.

---

**2. "Was missingness related to treatment?"**

Yes — show the missingness rate by condition table. The transparent condition had higher dropout, likely because fee-transparent displays were more cognitively demanding. The missingness-adjusted model addresses this partially; unobserved MNAR cannot be fully ruled out.

---

**3. "Did exclusions preserve randomization?"**

Attention failures were slightly more common in the transparent condition (`attention_fail` rates differed by condition). Excluding these rows conditions on a post-treatment variable and may narrow the effective sample toward easily-processed stimuli. We report results both with and without attention-flagged rows as a sensitivity check.

---

**4. "Why is expected value linear in your model?"**

It is not — m4b includes a quadratic term because diminishing sensitivity to expected value is psychologically motivated. We checked whether the linear model distorted the transparency estimate (it did/did not by X points).

---

**5. "Could trial order or fatigue explain the effect?"**

The assignment audit checked whether transparency was balanced across trial positions. Including trial order as a fixed effect (m4b onwards) adjusted for any systematic fatigue. The transparency estimate [was stable / shifted], suggesting [fatigue is not / may be] a confound.

---

**6. "Are you conditioning on a post-treatment variable?"**

No. The timing table in Section 6.2 documents which variables are pre-treatment, design-process, and post-treatment. Comprehension and perceived transparency were excluded from the main model precisely because they are potential mechanisms.

---

**7. "Did you treat stimuli as random?"**

Yes. The product offers are a sample from a broader population of financial product descriptions. Treating them as fixed would have claimed that these 36 offers are the only relevant stimuli — a claim we did not intend to make. The stimulus caterpillar plot shows real between-product variance larger than the transparency effect itself.

---

**8. "Are results sensitive to one site or a few stimuli?"**

A leave-one-site-out sensitivity (as in Part 4) would test site influence. With only 4 sites, no site should drive the result dramatically; the assignment audit confirmed that sites did not strongly predict transparency assignment. For stimulus influence: the large stimulus random-effect SD suggests individual products matter, but the crossed model averages over them appropriately.

---

**9. "What is the estimand?"**

The average total effect of *assigned* fee transparency on willingness to adopt, averaged across the population of participants, product offers, and sites similar to those sampled. It is not the effect for these specific 36 products or these exact four sites. It is not the direct effect holding comprehension constant.

---

**10. "Did the model converge cleanly?"**

The final model used the `bobyqa` optimizer and was non-singular. We confirmed convergence by restarting from multiple initial values and obtaining equivalent estimates.

---

## 21. Reporting Template

> We analyzed adoption-intent ratings (0–100) using linear mixed-effects models (lme4; Bates et al., 2015) with Satterthwaite degrees of freedom (lmerTest).
>
> **Observed data.** Of 2,880 intended trial observations (240 participants × 12 trials), [X]% had missing outcomes. Missingness was higher in the transparent condition and later trials (see Supplementary Table 1). We used complete-case analysis and checked sensitivity by adding missingness-related predictors (trial order, device type) to the main model; the transparency estimate changed by [amount], suggesting [minimal / moderate] impact of differential missingness.
>
> **Pre-analysis audits.** We inspected transparency assignment across pre-treatment variables (Section 6.1). [No / Modest] imbalance was detected. We excluded comprehension and perceived transparency from the main model because they are potential mechanisms from assignment to outcome.
>
> **Model specification.** Fixed effects: transparency, financial literacy, their interaction, risk level, a high-risk indicator, standardised expected value (linear + quadratic), and trial order. Random intercepts for participants, stimuli, and sites. By-participant random slope for transparency (justified by within-participant design; [correlated / uncorrelated / dropped] after singularity check).
>
> **Model comparison.** We compared: pooled OLS → participant RI → crossed RI → site RI → design covariates → nonlinear covariates + trial order → literacy interaction → random slope. The transparency estimate was stable across m0–m4b ([X] to [X]), confirming no large covariate confound. m5 added the literacy interaction, changing the substantive interpretation.
>
> **Results.** Fee transparency increased adoption intent by *b* = [X] (95% CI [X, X]), *t*([df]) = [X], *p* = [.X], at average financial literacy. Financial literacy moderated the effect (*b* = [X]), with larger transparency benefits for lower-literacy participants. Participant, stimulus, and site SDs were [X], [X], and [X].
>
> **Generalization.** The transparency effect generalizes, within the limits of these samples, to participants, product offers, and contexts similar to those studied.

::: {.callout-tip icon=false}
## Minimum Reproducible Model Report

- [ ] Rows intended, rows used, rows dropped — and why
- [ ] Missingness audit (rate by condition, site, trial order)
- [ ] Assignment audit (pre-treatment balance check)
- [ ] Post-treatment variable timing table
- [ ] Model formula (verbatim)
- [ ] Optimizer and convergence status; isSingular() result
- [ ] Estimator (REML/ML) and df method
- [ ] All fixed-effect estimates with SEs and CIs
- [ ] All random-effect variance components (SDs)
- [ ] Functional-form check (residuals vs covariates)
- [ ] Trial-order adjustment sensitivity
- [ ] Missingness sensitivity comparison
- [ ] Explicit estimand statement
:::

---

## Appendix: Stress Tests — When Threats Are Severe

The main worked example showed that a well-designed experiment with mild violations can still produce interpretable results. The stress tests below show what happens when the same threats are severe.

### Stress Test A: Severe Stimulus Imbalance

```{r stress-a}
set.seed(101)
trials_stress_a <- trials |>
  left_join(stimuli  |> select(stimulus_id, u0_stimulus),  by = "stimulus_id") |>
  left_join(participants |> select(participant_id, u0_participant, u1_participant), by = "participant_id") |>
  left_join(site_effects, by = "site_id") |>
  mutate(
    p_tx_imb = plogis(qlogis(0.50) - 0.55 * risk_level),
    tx_imb   = rbinom(n(), 1, p_tx_imb),
    mu_imb   = dgp$intercept +
               dgp$effect_transparency * tx_imb +
               dgp$effect_risk         * risk_level +
               dgp$effect_risk_high    * risk_high +
               dgp$effect_ev_linear    * expected_value_z +
               dgp$effect_ev_quad      * expected_value_z^2 +
               dgp$effect_literacy     * financial_literacy +
               dgp$interaction_tx_lit  * tx_imb * financial_literacy +
               u0_participant + u1_participant * tx_imb +
               u0_stimulus + u0_site,
    y_imb = pmin(100, pmax(0, rnorm(n(), mu_imb, dgp$sd_residual)))
  )

m_imb_raw <- lmer(y_imb ~ tx_imb +
                    (1 | participant_id) + (1 | stimulus_id),
                  data = trials_stress_a)

m_imb_adj <- lmer(y_imb ~ tx_imb + risk_level + risk_high + expected_value_z +
                    (1 | participant_id) + (1 | stimulus_id),
                  data = trials_stress_a)

bind_rows(
  broom.mixed::tidy(m_imb_raw, effects = "fixed") |>
    filter(term == "tx_imb") |> mutate(model = "Unadjusted"),
  broom.mixed::tidy(m_imb_adj, effects = "fixed") |>
    filter(term == "tx_imb") |> mutate(model = "Risk + EV adjusted")
) |>
  select(model, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) |>
  knitr::kable(caption = paste0("Stress test A: severe risk imbalance. True effect = ",
    dgp$effect_transparency, ". Unadjusted is biased; adjusted recovers it."))
```

**Teaching point:** In the clean design, covariates barely moved the estimate. Here, they recover it. The balance check from Section 5 would have flagged this imbalance before modeling.

---

### Stress Test B: Site-Level Treatment Heterogeneity

```{r stress-b}
set.seed(102)
site_effects_b <- site_effects |> mutate(u1_site = c(1.5, -1.0, 0.5, -1.0))

trials_stress_b <- trials |>
  left_join(stimuli  |> select(stimulus_id, u0_stimulus),  by = "stimulus_id") |>
  left_join(participants |> select(participant_id, u0_participant, u1_participant), by = "participant_id") |>
  left_join(site_effects_b, by = "site_id") |>
  mutate(
    mu_b = dgp$intercept +
           (dgp$effect_transparency + u1_site) * transparency +
           dgp$effect_risk      * risk_level +
           dgp$effect_risk_high * risk_high +
           dgp$effect_ev_linear * expected_value_z +
           dgp$effect_ev_quad   * expected_value_z^2 +
           dgp$effect_literacy  * financial_literacy +
           u0_participant + u1_participant * transparency +
           u0_stimulus + u0_site,
    y_b = pmin(100, pmax(0, rnorm(n(), mu_b, dgp$sd_residual)))
  )

m_site_ri <- lmer(y_b ~ transparency +
                    (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
                  data = trials_stress_b)

m_site_ix <- lmer(y_b ~ transparency * factor(site_id) +
                    (1 | participant_id) + (1 | stimulus_id),
                  data = trials_stress_b)

broom.mixed::tidy(m_site_ix, effects = "fixed") |>
  filter(str_detect(term, "transparency")) |>
  select(term, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) |>
  knitr::kable(caption = "Stress test B: transparency effect varies across sites. A site-only random intercept would not reveal this heterogeneity.")
```

**Teaching point:** A site random intercept captures baseline differences. A site interaction captures whether the treatment itself works differently across contexts. These answer different reviewer questions.

---

### Stress Test C: Sparse Within-Person Data

```{r stress-c}
set.seed(103)
trials_sparse <- trials |>
  group_by(participant_id) |>
  slice_sample(n = 4) |>
  ungroup()

m_sparse <- lmer(adoption_intent_obs ~ transparency * financial_literacy +
                   risk_level + expected_value_z +
                   (1 + transparency | participant_id) +
                   (1 | stimulus_id) + (1 | site_id),
                 data  = trials_sparse,
                 control = lmerControl(optimizer = "bobyqa"))

cat("Singular (sparse, 4 trials per person):", isSingular(m_sparse), "\n")
```

**Teaching point:** With 4 trials per participant (average 2 per condition), the data cannot reliably distinguish one person's transparency sensitivity from another's. The design justifies the random slope; the data do not support it. Drop the slope, report the attempt, and note that the sparse design is a limitation.
