---
title: "Crossed Random Effects in Experiments"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
set.seed(2025)
```

# Crossed Random Effects in Experiments

::: {.callout-note icon=false}
## The Stimulus Sampling Question

> **What are you actually generalizing over?** Not just participants — also over the stimuli (messages, gambles, products, scenarios) you used to operationalize your manipulation.
>
> **What goes wrong when you treat stimuli as fixed?** Stimulus-level variance is absorbed into the residual, making the treatment effect look more precise than it is. Standard errors for the manipulation effect are too small. Type I error is inflated.
>
> **What is the fix?** Add stimuli as a second random factor. Both participants *and* stimuli are sampled from populations of interest — both belong in the model.
:::

## Stimulus Audit Before Analysis

Before fitting a crossed random-effects model, audit the stimulus structure. Many of the most common errors in JDM experiments trace back to stimulus decisions made before analysis — not statistical choices.

::: {.callout-note icon=false}
## Stimulus Audit Checklist

| Question | Why it matters | What to check |
|----------|---------------|---------------|
| Were stimuli sampled, constructed, or selected after piloting? | Piloted-then-retained stimuli are not a random sample | Describe the stimulus selection process |
| Can each stimulus appear in each condition? | Enables full counterbalancing and crossed RE | Check design matrix |
| Are stimulus features (valence, concreteness, EV) balanced across conditions? | Unbalanced features confound the manipulation effect | Compare mean stimulus features by condition |
| How many stimuli per condition? | Drives stimulus-side power | Count N_stimuli by condition |
| Is the stimulus population the target of generalization? | Determines whether stimulus RE is appropriate | State what you want to generalize to |
| Are stimuli the same across participants, or are they sampled per participant? | Determines whether design is crossed or nested | Check whether each participant sees the same stimuli |

:::

::: {.callout-warning icon=false}
## When Stimuli Are Condition-Specific

If each stimulus belongs to only one condition — gain-framed messages were written for gain framing, loss-framed messages were written for loss framing — the condition effect and the stimulus-quality difference are **perfectly confounded**. The crossed random-effects model cannot disentangle them.

**What claim survives:** The treatment effect estimate for these specific stimuli is still valid. Generalization requires that the condition-specific stimuli are representative of the broader classes, which must be argued conceptually — it cannot be verified statistically.

**What to check:** Report the stimulus-level ICC. If it is substantially above zero, stimulus quality differences exist and should be acknowledged. If you have coded stimulus features (concreteness, familiarity, expected value), include them as covariates to rule out feature confounds.
:::

---

## The Stimuli-Are-Random Problem

Every JDM experiment that presents multiple stimuli — gain- and loss-framed messages, product descriptions, gamble pairs, disclosure formats, price offers, policy options — has the same structural feature: the stimuli you used are a *sample* from a larger population of possible stimuli.

A gain-frame study uses, say, 12 gain-framed messages. Those 12 were chosen (or written) by the researcher. Some turned out to be unusually compelling. Others fell flat. If you happened to write gain-framed messages that were particularly vivid and loss-framed messages that were blandly written, your "framing effect" captures both the actual psychological effect of framing *and* the luck of your specific stimulus draw.

**The problem:** If you treat your stimuli as fixed — as if these 12 messages were the only 12 that could ever represent gain framing — then stimulus-level variability is absorbed into the residual error term. The residual looks small and the treatment effect looks precise. But that precision is false.

**The fix:** Treat stimuli as a random factor, just like participants. Model both sources of random variability explicitly: participants vary in their baseline tendencies; stimuli vary in their baseline effectiveness. The treatment effect is estimated net of both.

---

## Crossed, Not Nested

In most within-subjects experiments, participants and stimuli are **crossed**: every participant sees multiple stimuli, and every stimulus is seen by multiple participants. Neither factor is nested within the other.

This is different from the nested structures in Modules 1–2. In a nested design, each participant belongs to *one* group and each observation belongs to *one* participant. In a crossed design, participants and stimuli form a matrix — each cell is a (participant × stimulus) observation:

|              | Message 1 | Message 2 | Message 3 | ... | Message 24 |
|:-------------|:---------:|:---------:|:---------:|:---:|:----------:|
| Participant 1| ✓ | ✓ | — | ... | ✓ |
| Participant 2| — | ✓ | ✓ | ... | ✓ |
| Participant 3| ✓ | — | ✓ | ... | — |
| ...          | ... | ... | ... | ... | ... |

Each checkmark is one observation. Participant random effects capture stable differences in how people respond (e.g., generally higher WTP). Stimulus random effects capture stable differences in how stimuli perform (e.g., a particularly vivid message consistently gets higher responses). The treatment effect — gain vs. loss framing — is what remains after removing both sources of variance.

When participants only see a *subset* of stimuli, the design is **partially crossed**. The analysis is the same; `lme4` handles incomplete designs automatically.

---

## Simulation: Gain vs. Loss Framing and WTP

We simulate a framing experiment: 60 participants each see 12 of 24 gain/loss-framed messages describing financial product benefits. They rate their willingness to adopt the product. The true framing effect is +3 points (loss framing increases adoption intent). Both participants and messages have their own stable variance.

```{r simulate-experiment}
n_participants <- 60
n_messages     <- 24  # 12 gain-framed, 12 loss-framed

participant_effects <- rnorm(n_participants, 0, 8)   # SD = 8 (large between-person spread)
message_effects     <- rnorm(n_messages, 0, 6)        # SD = 6 (meaningful between-message spread)
message_condition   <- rep(c("gain", "loss"), each = 12)

# Partial crossing: each participant sees approximately half the messages
set.seed(2025)
df_exp <- expand_grid(
  participant = 1:n_participants,
  message     = 1:n_messages
) |>
  slice_sample(prop = 0.5) |>   # each participant sees ~12 messages
  mutate(
    condition   = message_condition[message],
    p_eff       = participant_effects[participant],
    m_eff       = message_effects[message],
    WTP         = 50 + 3 * (condition == "loss") + p_eff + m_eff + rnorm(n(), 0, 5),
    participant = factor(participant),
    message     = factor(message)
  )
```

### Two models, two answers

```{r compare-models}
# Incorrect: ignores stimulus variance
m_wrong <- lmer(WTP ~ condition + (1 | participant), data = df_exp)

# Correct: both participant and message as crossed random effects
m_correct <- lmer(WTP ~ condition + (1 | participant) + (1 | message), data = df_exp)

bind_rows(
  broom.mixed::tidy(m_wrong,   effects = "fixed") |>
    filter(term == "conditionloss") |>
    mutate(model = "Incorrect: participant RE only\n(ignores message variance)"),
  broom.mixed::tidy(m_correct, effects = "fixed") |>
    filter(term == "conditionloss") |>
    mutate(model = "Correct: participant + message RE\n(crossed random effects)")
) |>
  select(model, estimate, std.error, statistic, p.value) |>
  knitr::kable(digits = 3,
    caption = "Ignoring message variance produces an overconfident SE. The correct crossed model reveals the true uncertainty about the framing effect across all possible messages of this type.")
```

The incorrect model looks more precise — but its SE is too small because it attributes stimulus-level variance to residual noise instead of a proper separate component. This is a false precision that will not replicate when a different message set is used.

### Visualizing the two variance components

```{r variance-plot, fig.width=8, fig.height=4}
# Extract variance components from the correct model
vc_df <- as.data.frame(VarCorr(m_correct)) |>
  filter(grp != "Residual") |>
  mutate(
    component = case_when(
      grp == "participant" ~ "Between-participant variance\n(stable individual tendencies)",
      grp == "message"     ~ "Between-message variance\n(stimulus quality differences)",
      TRUE                 ~ grp
    ),
    sd = sdcor
  )

vc_res <- as.data.frame(VarCorr(m_correct)) |>
  filter(grp == "Residual") |>
  mutate(component = "Within-cell residual\n(unexplained individual × message variation)", sd = sdcor)

vc_all <- bind_rows(vc_df, vc_res) |>
  mutate(component = factor(component, levels = c(
    "Between-participant variance\n(stable individual tendencies)",
    "Between-message variance\n(stimulus quality differences)",
    "Within-cell residual\n(unexplained individual × message variation)"
  )))

ggplot(vc_all, aes(x = component, y = sd, fill = component)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  scale_fill_manual(values = c("#1e3a5f", "#2d6a4f", "#cccccc")) +
  labs(
    title    = "Variance components: participant, message, and residual",
    subtitle = "The incorrect model collapses the message component into residual — shrinking the SE artificially",
    x        = NULL,
    y        = "SD (variance component)"
  ) +
  theme_minimal(base_size = 12)
```

---

### Post-fit: report stimulus ICC and compare models

After fitting, always report the stimulus ICC and show what the SE would have been if stimulus variance had been ignored. This is the evidence that the crossed model was necessary.

```{r stimulus-diagnostics}
# Stimulus ICC from the correct model
vc <- as.data.frame(VarCorr(m_correct))
var_message  <- vc$vcov[vc$grp == "message"]
var_residual <- vc$vcov[vc$grp == "Residual"]
icc_message  <- var_message / (var_message + var_residual)

cat("Message ICC:", round(icc_message, 3), "\n")
cat("Interpretation: ", round(icc_message * 100, 1),
    "% of residual variance (after participant effects) is due to stable message differences.\n\n")

# SE comparison
se_wrong   <- broom.mixed::tidy(m_wrong,   effects = "fixed") |> filter(term == "conditionloss") |> pull(std.error)
se_correct <- broom.mixed::tidy(m_correct, effects = "fixed") |> filter(term == "conditionloss") |> pull(std.error)
cat("SE ignoring messages: ", round(se_wrong, 3), "\n")
cat("SE with message RE:   ", round(se_correct, 3), "\n")
cat("SE inflation from ignoring messages:", round((se_wrong / se_correct - 1) * 100, 1), "% smaller\n")
```

---

## The Judd–Westfall–Kenny (2012) Argument

Judd, Westfall & Kenny (2012) documented a systematic error in social psychology experiments: researchers treat stimuli (words, faces, messages, images) as *fixed* rather than random. They labeled this the **"stimuli-as-fixed-effects fallacy."** The same problem applies directly to JDM research:

- A study of risk communication using 8 gamble descriptions treats those 8 as the only possible gambles — when they are clearly a sample
- A study of disclosure effects on WTP using 6 product descriptions generalizes implicitly to all product descriptions — but the model does not account for stimulus-level variance
- A study of default effects using 4 enrollment forms treats those 4 forms as exhaustive — underestimating between-form variance

The consequence is systematic Type I error inflation that is *not* corrected by increasing sample size alone. Adding more participants does not help if the binding constraint is stimulus-level variance.

---

## Power Depends on Both N and Number of Stimuli

Westfall, Kenny & Judd (2014) showed that in crossed designs, statistical power is jointly determined by the number of participants *and* the number of stimuli. With only a few stimuli, adding more participants eventually yields diminishing returns because stimulus-level variance remains the binding constraint.

```{r power-schematic, fig.width=8, fig.height=4.5}
# Schematic: how SE changes with N_stimuli vs. N_participants
# (simplified illustration, not simulation-calibrated)
se_grid <- expand_grid(
  n_stim = c(4, 8, 16, 32),
  n_part = seq(20, 200, by = 20)
) |>
  mutate(
    # Approximate SE ∝ 1/sqrt(n_part) + 1/sqrt(n_stim), simplified
    approx_se = sqrt(64 / n_part + 36 / n_stim),
    n_stim    = factor(paste(n_stim, "stimuli"), levels = paste(c(4, 8, 16, 32), "stimuli"))
  )

ggplot(se_grid, aes(x = n_part, y = approx_se, color = n_stim, group = n_stim)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(
    values = c("#c0392b", "#e67e22", "#2d6a4f", "#1e3a5f"),
    name   = NULL
  ) +
  labs(
    title    = "SE of treatment effect as function of N participants and N stimuli",
    subtitle = "With few stimuli, adding participants has limited impact. Expand the stimulus set.",
    x        = "Number of participants",
    y        = "Approximate SE of treatment effect"
  ) +
  theme_minimal(base_size = 12)
```

**Practical implication:** If you are planning a framing experiment and you have only 6 message pairs, consider doubling the message set before doubling the participant count. The precision of your framing-effect estimate depends heavily on how many distinct stimuli you have sampled.

---

## Assumptions

::: {.callout-warning icon=false}
## Assumption: Stimuli Are a Representative Sample From the Population of Interest

**Plain-language meaning:** The crossed model generalizes the treatment effect to new participants *and* to new stimuli drawn from the same population. This is only valid if your stimuli were sampled in a way that represents the full range of relevant stimuli — not cherry-picked to maximize or demonstrate an effect.

**What the model is assuming away:** The possibility that you selected stimuli precisely because they produced the pattern you expected — in which case the stimulus "sample" is not a random draw from the population of all possible stimuli.

**JDM example:** A framing researcher writes 12 gain messages and 12 loss messages. If the loss messages were unconsciously written to be more concrete and vivid (because the researcher was more practiced at loss language), the between-condition difference reflects both framing and writing quality.

**Experimental analog in observational work:** In observational stimulus-response data (e.g., consumer responses to real advertisements), stimuli are not randomly sampled — they are produced by advertisers with incentives to optimize. The crossed model would still be appropriate for inference about the sample, but generalization to "all possible ads" requires caution.

**What a skeptical reviewer might say:** "How were your stimuli selected? Were they piloted and retained based on effect-size criteria? If so, the crossed-effects estimate may overstate how well the manipulation generalizes."

**Observable warning sign:** Between-stimulus variance that is unusually low (suggesting pre-selection of comparable stimuli) or unusually high (suggesting highly heterogeneous stimuli from different populations).

**What to check:** Report stimulus-level ICC: $\sigma^2_{\text{message}} / (\sigma^2_{\text{message}} + \sigma^2_{\text{residual}})$. If it is effectively zero, message variance is negligible and the concern is less pressing. If it is large, the stimulus sample matters a lot and should be defended.

**What claim survives if the assumption is doubtful:** The treatment effect estimate for *these specific stimuli* is still valid. Generalization to the population of stimuli depends on representativeness.
:::

::: {.callout-warning icon=false}
## Assumption: Random Effects for Participants and Stimuli Are Independent of the Manipulation

**Plain-language meaning:** The model assumes that which stimuli end up in the gain condition versus the loss condition is unrelated to their "natural strength" as stimuli. In other words, the assignment of stimuli to conditions was not influenced by stimulus quality.

**What the model is assuming away:** Any correlation between stimulus random intercepts and condition assignment — which would exist if, for example, researchers assigned their best-quality messages to the manipulation they expected to produce the larger effect.

**JDM example:** If a researcher piloted messages and assigned the most persuasive ones to the loss-frame condition (expecting loss aversion to need less help), the loss condition starts with a built-in quality advantage.

**What a skeptical reviewer might say:** "Were stimuli randomly assigned to conditions, or were they constructed for their assigned condition from the start? If constructed, the random-effects independence assumption requires that condition-specific writing quality is negligible."

**What to check:** In a fully counterbalanced design (where each message can appear in either condition across participants), this assumption is guaranteed by design. In a design where stimuli were created for a specific condition, the assumption must be argued conceptually.
:::

---

## Reviewer Challenge

::: {.callout-tip}
## What a Reviewer Would Ask

**Challenge 1:** "You have only 8 messages per condition. Isn't that too few to treat messages as random?"
The answer is: too few for high power, yes — but that is an argument for adding more stimuli, not for treating the existing stimuli as fixed. Treating 8 messages as fixed asserts that those exact messages are the population of interest, which makes the result entirely non-generalizable. A small number of stimuli produces low power with correct random effects; it produces false precision with fixed effects.

**Challenge 2:** "Your message random effect variance is larger than your participant variance. Does that mean messages matter more than participants?"
Not necessarily — it means there is more heterogeneity across your message set than across your participant sample. This could reflect genuine message quality differences, or it could reflect that your message set was less homogeneous than your participant pool. Either way, the correct model accounts for both.

**Challenge 3:** "Why not just average across messages and analyze at the participant level?"
Averaging loses information and implicitly weights all messages equally, which is rarely justified. More importantly, it only corrects participant-level inference — it does not account for the possibility that the results are specific to this particular message set rather than generalizing to the population of possible messages. Averaging kills the second random effect.
:::
