---
title: "A Fully Worked Multilevel Modeling Example"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(glmmTMB)
set.seed(20240610)
```

# A Fully Worked Multilevel Modeling Example

This chapter does not introduce a new method. It shows what it looks like to *use* the methods from Parts 1–7 together, on a single dataset, from first principles.

The voice is intentionally conversational. Every major modeling decision arrives as a question, gets answered, and then gets tested. That is how research supervision actually works — and it is the only honest way to learn applied multilevel modeling.

---

## 1. The Research Question

> **Does transparent fee disclosure increase willingness to adopt a financial product, and does that effect generalize across participants, product offers, and research sites?**

Before touching a model, notice that this question hides several distinct questions inside it:

- Does transparency increase adoption for *these* participants?
- Does that effect generalize to the *population of participants*?
- Does it generalize across the *population of product offers*, or only for these specific stimuli?
- Is the effect consistent across *different research sites and platforms*?
- Is the effect stronger for some *types of participants* (e.g., those with lower financial literacy)?

A pooled regression can only answer the first question — and even then only for these exact stimuli. A properly specified multilevel model can address all five. The modeling choices we make determine which questions we can credibly answer.

---

## 2. The Design Audit Before Modeling

The most important modeling decisions are made before any model is fitted. The design determines what random effects are justified, what fixed effects are necessary, and what the estimand actually is.

| Design question | Answer in this study | Modeling implication |
|---|---|---|
| What is one row? | One participant's rating of one product offer | Trial-level response; multiple rows per participant |
| What is randomized? | Fee transparency display (within-participant) | Fixed effect for transparency |
| Does treatment vary within participant? | Yes — each participant sees both conditions | Candidate by-participant random slope for transparency |
| Are stimuli sampled? | Yes — 36 product offers drawn from a broader set | Stimulus random intercept |
| Are participants crossed with stimuli? | Partially — each participant rates 12 of 36 offers | Crossed (partial) random effects |
| Are participants nested in sites? | Yes — 240 participants across 4 sites | Site random intercept |
| Are there stimulus covariates? | Risk level, expected value | Include as fixed effects to check confounding |
| Is there a participant-level moderator? | Financial literacy (measured pre-treatment) | Fixed effect; interaction with transparency |
| Is the outcome continuous? | 0–100 willingness-to-adopt rating | Linear mixed model; check residuals |

: Design audit: answering these questions before modeling determines every subsequent decision. {tbl-colwidths="[30,30,40]"}

---

## 3. Simulate the Dataset

We simulate a dataset that is realistic enough to produce genuine diagnostic lessons. The true data-generating process includes participant, stimulus, and site variance; a transparency effect; a financial-literacy moderator; and realistic noise.

```{r true-dgp-params}
# True DGP parameters — we will compare model estimates to these at the end
dgp <- list(
  intercept          = 50,    # grand mean adoption intent
  effect_transparency = 4,    # +4 points for transparent-fee display
  effect_risk        = -3,    # riskier products → lower adoption intent
  effect_ev          = 0.08,  # higher expected value → higher adoption intent
  effect_literacy    = 2,     # higher literacy → higher baseline adoption
  interaction_tx_lit = -1.5,  # transparency helps low-literacy more
  sd_participant     = 8,     # between-person stable differences
  sd_stimulus        = 5,     # between-product stable differences
  sd_site            = 4,     # between-site baseline differences
  sd_slope_part      = 2,     # participants differ in transparency sensitivity
  sd_residual        = 7      # within-trial noise
)
```

```{r simulate-data}
n_participants <- 240
n_stimuli      <- 36
n_sites        <- 4
trials_per_part <- 12  # each participant rates 12 of 36 offers

# Participant-level characteristics
participants <- tibble(
  participant_id     = 1:n_participants,
  site_id            = rep(1:n_sites, each = n_participants / n_sites),
  financial_literacy = rnorm(n_participants, 0, 1),  # standardised
  u0_participant     = rnorm(n_participants, 0, dgp$sd_participant),
  u1_participant     = rnorm(n_participants, 0, dgp$sd_slope_part)  # random slope for transparency
)

# Site-level intercepts
site_effects <- tibble(
  site_id    = 1:n_sites,
  u0_site    = c(3, -2, 1, -2)  # site 1 slightly higher baseline; site 4 lower
)

# Stimulus-level characteristics
stimuli <- tibble(
  stimulus_id    = 1:n_stimuli,
  risk_level     = rnorm(n_stimuli, 0, 1),   # standardised risk
  expected_value = rnorm(n_stimuli, 50, 15), # EV in dollars
  u0_stimulus    = rnorm(n_stimuli, 0, dgp$sd_stimulus),
  # transparency assignment: within-participant, so each stimulus
  # can appear in both conditions across participants
  base_transparency = rep(c(0, 1), n_stimuli / 2)  # half designed as transparent
)

# Assign stimuli to participants: each participant sees 12 of 36
# (partially crossed design)
trials <- participants |>
  select(participant_id, site_id, financial_literacy, u0_participant, u1_participant) |>
  cross_join(tibble(draw = 1:trials_per_part)) |>
  group_by(participant_id) |>
  mutate(
    stimulus_id  = sample(1:n_stimuli, trials_per_part, replace = FALSE),
    trial_order  = 1:trials_per_part
  ) |>
  ungroup() |>
  select(-draw) |>
  left_join(stimuli, by = "stimulus_id") |>
  left_join(site_effects, by = "site_id") |>
  mutate(
    # Transparency varies within participant (counterbalanced across stimuli)
    transparency = as.integer(runif(n()) > 0.5),
    # True DGP
    mu = dgp$intercept +
         dgp$effect_transparency * transparency +
         dgp$effect_risk        * risk_level +
         dgp$effect_ev          * (expected_value - 50) / 15 +
         dgp$effect_literacy    * financial_literacy +
         dgp$interaction_tx_lit * transparency * financial_literacy +
         u0_participant + u1_participant * transparency +
         u0_stimulus + u0_site,
    adoption_intent = pmin(100, pmax(0, rnorm(n(), mu, dgp$sd_residual)))
  ) |>
  select(participant_id, site_id, stimulus_id, trial_order,
         transparency, risk_level, expected_value, financial_literacy,
         adoption_intent)
```

```{r data-summary}
# Structure check
cat("Participants:", n_distinct(trials$participant_id), "\n")
cat("Stimuli:", n_distinct(trials$stimulus_id), "\n")
cat("Sites:", n_distinct(trials$site_id), "\n")
cat("Trials per participant:", trials_per_part, "\n")
cat("Total rows:", nrow(trials), "\n")
cat("Observations per stimulus (mean):",
    round(nrow(trials) / n_distinct(trials$stimulus_id), 1), "\n")

head(trials, 6) |> knitr::kable(digits = 2, caption = "First six rows of the dataset")
```

---

## 4. Pre-Fit Diagnostics

Fitting a model before inspecting the data is like writing a discussion section before looking at the results. The checks below are not optional extras — they determine whether the model can answer the intended question.

### 4.1 Data structure checks

```{r structure-checks}
# Trials per participant
trials_per_p <- trials |> count(participant_id)
cat("Trials per participant: min =", min(trials_per_p$n),
    ", max =", max(trials_per_p$n), "\n")

# Transparency balance within participants
tx_balance <- trials |>
  group_by(participant_id) |>
  summarise(prop_transparent = mean(transparency))
cat("Within-participant transparency proportion: mean =",
    round(mean(tx_balance$prop_transparent), 2),
    ", SD =", round(sd(tx_balance$prop_transparent), 2), "\n")

# Participants per site
trials |>
  distinct(participant_id, site_id) |>
  count(site_id) |>
  knitr::kable(caption = "Participants per site")
```

### 4.2 Design balance checks

**Student question:** If treatment was randomized, why do we still need balance checks?

**Advisor response:** Randomization protects us *in expectation* across hypothetical replications of the study. In the actual realized sample, stimuli and participants can still be systematically unbalanced. For example, if transparency was accidentally more common for high-risk stimuli, the transparency effect and the risk effect become partially confounded in this dataset — even though they were unrelated in the DGP. Balance checks tell us whether such confounding exists and whether we need to adjust for it.

**Diagnostic move:** Check whether transparency is balanced across stimulus risk level and expected value.

```{r balance-checks}
# Is transparency balanced across risk level?
tx_by_risk <- trials |>
  mutate(risk_tercile = ntile(risk_level, 3)) |>
  group_by(risk_tercile) |>
  summarise(prop_transparent = round(mean(transparency), 3),
            n = n())
knitr::kable(tx_by_risk,
  caption = "Transparency proportion by risk tercile (should be ~0.50 in each)")

# Is transparency balanced across expected value?
tx_by_ev <- trials |>
  mutate(ev_tercile = ntile(expected_value, 3)) |>
  group_by(ev_tercile) |>
  summarise(prop_transparent = round(mean(transparency), 3),
            n = n())
knitr::kable(tx_by_ev,
  caption = "Transparency proportion by EV tercile (should be ~0.50 in each)")
```

**What we learned:** If any cell shows proportion notably different from 0.50, the transparency effect estimate will absorb some of the risk or EV effect. Including risk level and expected value as covariates (Step 9) addresses this regardless of whether imbalance is detected — it is good practice in any within-stimulus design.

### 4.3 Outcome checks

```{r outcome-checks, fig.width=9, fig.height=4}
p1 <- ggplot(trials, aes(x = adoption_intent)) +
  geom_histogram(bins = 30, fill = "#1e3a5f", color = "white") +
  labs(title = "Distribution of adoption intent", x = "Adoption intent (0–100)") +
  theme_minimal(base_size = 11)

p2 <- trials |>
  mutate(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  ggplot(aes(x = Condition, y = adoption_intent, fill = Condition)) +
  geom_boxplot(show.legend = FALSE, width = 0.5) +
  scale_fill_manual(values = c("#cccccc", "#1e3a5f")) +
  labs(title = "Adoption intent by transparency", y = "Adoption intent (0–100)", x = NULL) +
  theme_minimal(base_size = 11)

p3 <- trials |>
  mutate(site_label = paste0("Site ", site_id)) |>
  ggplot(aes(x = site_label, y = adoption_intent, fill = site_label)) +
  geom_boxplot(show.legend = FALSE, width = 0.5) +
  scale_fill_manual(values = c("#1e3a5f","#2d6a4f","#e67e22","#c0392b")) +
  labs(title = "Adoption intent by site", y = NULL, x = NULL) +
  theme_minimal(base_size = 11)

gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

---

## 5. Start With the Wrong Model

**Student question:** I ran a simple regression and the transparency effect is significant. Why not stop there?

**Advisor response:** Because significance is not the question yet. The pooled model is wrong in ways that matter — but not always in the direction students expect. In a balanced within-person experiment like this one, random intercepts often *reduce* the standard error rather than inflating it, because they remove stable participant and stimulus baseline differences from the residual and sharpen the within-unit comparison. The pooled model is wrong not because it is overconfident, but because it misrepresents the data-generating process, treats these specific product offers as if they are the entire universe of possible stimuli, and cannot support any generalization claim about new participants or new offers.

**Diagnostic move:** Fit the pooled model, record its estimate and SE, then watch how those change — and notice *why* each change happens.

```{r m0-naive}
m0 <- lm(adoption_intent ~ transparency, data = trials)
broom::tidy(m0) |>
  filter(term == "transparency") |>
  mutate(model = "m0: Pooled OLS") |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "Naive pooled OLS: transparency estimate")
```

```{r m0-diagnosis}
tibble(
  Issue = c(
    "Independence assumption",
    "Participant clustering",
    "Stimulus clustering",
    "Site context",
    "Stimulus covariates"
  ),
  `What the model assumes` = c(
    "All 2,880 rows are independent",
    "The same person appearing 12 times counts as 12 independent people",
    "The same stimulus appearing ~80 times counts as ~80 independent stimuli",
    "All sites draw from the same distribution",
    "Risk level and EV are not related to transparency"
  ),
  `Why this fails` = c(
    "Participant and stimulus baselines stay in the residual — the model is inefficient",
    "Cannot separate stable person differences from within-person treatment contrasts",
    "Stimulus attractiveness differences inflate residual variance and prevent generalization",
    "Site-level baselines are absorbed into the participant variance component",
    "If transparency and risk level are unbalanced, the estimate is biased"
  )
) |>
  knitr::kable(caption = "Why the pooled model misrepresents the design")
```

**Next modeling decision:** Add participant random intercepts to account for repeated measures within person.

---

## 6. Add Participant Random Intercepts

Participant random intercepts absorb each person's stable adoption tendency. After adding them, the 12 rows from the same person are no longer treated as 12 independent observations — they share a common baseline and the model adjusts accordingly.

```{r m1}
m1 <- lmer(adoption_intent ~ transparency + (1 | participant_id), data = trials)
```

```{r m1-summary}
# Participant ICC
vc1 <- as.data.frame(VarCorr(m1))
var_part  <- vc1$vcov[vc1$grp == "participant_id"]
var_resid <- vc1$vcov[vc1$grp == "Residual"]
icc_part  <- var_part / (var_part + var_resid)
cat("Participant ICC:", round(icc_part, 3), "\n")
cat("Interpretation:", round(icc_part * 100, 1),
    "% of total variance (within-person) is due to stable individual differences.\n")

broom.mixed::tidy(m1, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m1: participant RI") |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)
```

**Student question:** The point estimate barely moved, but the SE went *down*. I thought adding random effects was supposed to make SEs bigger?

**Advisor response:** That expectation comes from the between-subjects case, where clustering means your effective N is smaller than your raw N. Here, transparency is randomized *within* participant — each person sees both transparent and non-transparent offers. The participant random intercept does not reduce our effective N for transparency; it removes the stable between-person noise from the residual. With less residual variance, the within-person contrast is estimated more precisely. The SE shrinks because the model is doing its job: separating what is attributable to person baselines from what is attributable to the manipulation.

**Student question:** What does the participant ICC mean substantively?

**Advisor response:** It tells you what fraction of total outcome variance (after removing the transparency effect) is due to stable individual differences in adoption intent. A high ICC here — say, 0.45 — means that knowing which person is rating a product tells you a lot about the rating, even before you know which product it is or whether the fee is displayed transparently. The pooled model left all of that stable variance in the residual, making it "noisier" than it needed to be.

**What is still missing:** Stimuli are still treated as if they are fixed background objects. The 36 product offers each appear roughly 80 times and vary substantially in their attractiveness. That between-product variance is currently sitting in the residual, mixed together with unexplained trial-level noise. We cannot generalize to new products until we model this source of variance explicitly.

---

## 7. Add Stimulus Random Intercepts

Adding stimulus random intercepts makes this a crossed random-effects model. Both participants and stimuli are treated as random samples: we want to generalize the transparency effect to new participants *and* to new product offers, not just to these specific 36.

```{r m2}
m2 <- lmer(adoption_intent ~ transparency +
             (1 | participant_id) + (1 | stimulus_id),
           data = trials)
```

```{r m2-summary}
vc2 <- as.data.frame(VarCorr(m2))
var_part2   <- vc2$vcov[vc2$grp == "participant_id"]
var_stim2   <- vc2$vcov[vc2$grp == "stimulus_id"]
var_resid2  <- vc2$vcov[vc2$grp == "Residual"]
icc_stim    <- var_stim2 / (var_stim2 + var_resid2)

cat("Stimulus ICC:", round(icc_stim, 3), "\n")
cat("Stimulus SD:", round(sqrt(var_stim2), 2),
    "(compared to transparency effect of ~4 points)\n")

broom.mixed::tidy(m2, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m2: participant + stimulus RI") |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)
```

```{r m2-caterpillar, fig.width=7, fig.height=4.5}
# Caterpillar plot: stimulus random intercepts
stim_re <- ranef(m2)$stimulus_id |>
  as.data.frame() |>
  rownames_to_column("stimulus_id") |>
  rename(blup = `(Intercept)`) |>
  arrange(blup)

ggplot(stim_re, aes(x = reorder(stimulus_id, blup), y = blup)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#888888") +
  geom_segment(aes(xend = stimulus_id, y = 0, yend = blup),
               color = "#1e3a5f", linewidth = 0.6, alpha = 0.7) +
  geom_point(color = "#1e3a5f", size = 2) +
  coord_flip() +
  labs(
    title    = "Stimulus random intercepts (caterpillar plot)",
    subtitle = paste0("Spread reflects stable between-product attractiveness differences.\n",
                      "SD ≈ ", round(sqrt(var_stim2), 1), " points — larger than the transparency effect."),
    x        = "Stimulus",
    y        = "Estimated baseline deviation (BLUP)"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

**Student question:** Are stimuli really random? We designed these product offers — they are not randomly drawn from a jar.

**Advisor response:** The question is not whether you drew them from a jar. It is whether you want to generalize beyond these 36 specific offers. If your claim is "fee transparency works for financial products in general," then these 36 offers are a sample from the population of possible financial products that could have been used. The random-effects model says: given that different products produce different baseline adoption levels, and given that your 36 offers are a sample, how well-estimated is the *average* transparency effect across all such products? That is a more honest question than "does transparency work for these exact 36 offers?"

**Student question:** What if stimulus variance is near zero?

**Advisor response:** Then the stimulus random intercept is contributing almost nothing and the model simplifies naturally. But we should let the data decide, not assume away a variance component before fitting. If the stimulus variance estimate is essentially zero, we report that and note it implies our particular stimulus set was unusually homogeneous — which is itself worth knowing.

**Student question:** What if each stimulus appears in only one transparency condition?

**Advisor response:** Then a by-stimulus random slope for transparency is not identifiable — you cannot distinguish "this stimulus always looked better" from "this stimulus looked better because of its assigned condition." The random *intercept* for stimuli is still identifiable and useful. But the design cannot tell you whether individual stimuli responded differently to transparency.

---

## 8. Add Site Structure

Participants are nested in sites. Site 1 recruited from a university lab, Sites 2–3 from online platforms, Site 4 from a community sample. These sites differ in their baseline adoption intent. Ignoring site means that between-site variance is absorbed into the participant random effect, making participant variance look larger than it really is.

```{r m3}
# participant_id values are globally unique, so we can use
# (1 | site_id) + (1 | participant_id) directly
# This is equivalent to (1 | site_id/participant_id) when IDs are unique
m3 <- lmer(adoption_intent ~ transparency +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)
```

```{r m3-diagnostics}
vc3 <- as.data.frame(VarCorr(m3))
var_site3 <- vc3$vcov[vc3$grp == "site_id"]
cat("Site SD:", round(sqrt(var_site3), 2), "\n")
cat("Note: true site SD in DGP =", dgp$sd_site, "\n")

# Site-specific mean adoption intent
trials |>
  group_by(site_id) |>
  summarise(mean_adoption = round(mean(adoption_intent), 1),
            n_participants = n_distinct(participant_id)) |>
  knitr::kable(caption = "Site-specific means — visible differences suggest site variance is real")

broom.mixed::tidy(m3, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m3: + site RI") |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)
```

**Student question:** Should site be a fixed effect or a random effect?

**Advisor response:** Use random effects when sites are a sample from a population of possible contexts and you want to generalize beyond these four specific sites. Use fixed effects when sites *are* the entire target population, or when you have so few sites (J < 5) that the random-effect variance cannot be reliably estimated. With 4 sites we are on the borderline — the random-effect variance estimate is noisy, but fixed effects would absorb all between-site information and prevent us from making any site-level generalization statement. We will use random effects here and acknowledge the instability with a note. See Part 4 for the full discussion.

**Student question:** The estimate and SE barely changed when I added site. Did the site term do anything?

**Advisor response:** Yes, but not by changing the transparency estimate. What it did: (1) it correctly partitioned between-site baseline differences out of the participant variance component — before, participant "random intercepts" were absorbing some site-level noise; (2) it changed the estimand from "effect across participants in these specific four sites" to "average effect across a population of sites." What it did *not* do here: change the transparency estimate, because site differences in this simulation are baseline differences, not transparency-confounding differences. If some sites were more responsive to transparency than others — a site-by-treatment interaction — adding a site random effect would not capture that. We would need a by-site random slope for transparency, or fixed interactions between site and treatment. For 4 sites, a random slope is unreliable; we note this as a limitation.

**Student question:** Does adding site change the estimand or just the uncertainty?

**Advisor response:** Primarily the estimand. The claim shifts from "this effect in these four sites" to "average effect across a population of sites similar to these." That is a different and more general claim — but one that requires arguing that these four sites are reasonably representative of the contexts you want to generalize to.

---

## 9. Add Design-Relevant Fixed Effects

Risk level and expected value are stimulus characteristics that vary across products. If they are even slightly unbalanced with respect to transparency assignment, they confound the estimate. We include them as fixed effects — not to test them, but to adjust for them.

Financial literacy is a participant characteristic measured before the experiment. It belongs in the model as both a covariate (adjusting for pre-existing differences) and, later, as a moderator.

```{r m4}
m4 <- lmer(adoption_intent ~ transparency + risk_level + expected_value +
             financial_literacy +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)
```

```{r m4-comparison}
# Track how the transparency estimate has moved across models
bind_rows(
  broom::tidy(m0) |> filter(term == "transparency") |>
    mutate(model = "m0: pooled OLS"),
  broom.mixed::tidy(m1, effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m1: + participant RI"),
  broom.mixed::tidy(m2, effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m2: + stimulus RI"),
  broom.mixed::tidy(m3, effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m3: + site RI"),
  broom.mixed::tidy(m4, effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m4: + covariates")
) |>
  select(model, estimate, std.error, p.value) |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = paste0("Tracking the transparency estimate across models.",
    " True effect = ", dgp$effect_transparency))
```

**Student question:** Could controlling for risk level remove part of the transparency effect?

**Advisor response:** Only if risk level is genuinely part of the *causal pathway* from transparency to adoption intent — if the only way transparency helps is by making the risk salient, and we then control for risk perception, we would be "explaining away" the mechanism. But here, risk level is a pre-existing stimulus characteristic that was not caused by the transparency manipulation. Controlling for it adjusts for differences in product attractiveness that exist independently of how fees are displayed. That is a design-feature adjustment, not overcontrol.

**What we learned:** The transparency estimate barely moved across models m0 through m4. That is not a failure of the worked example — it is a result. It tells us the simulated experiment was well-randomized: transparency was not confounded with participant characteristics, stimulus features, or site. Adding covariates here documents robustness rather than fixes bias.

::: {.callout-note icon=false}
## A Stable Estimate Is a Scientific Result

Students sometimes feel that a model sequence that does not change the focal estimate was "wasted work." It was not. The sequence from m0 to m4 answered several questions that matter independently of the estimate:

- m1 told us that stable individual differences explain a large share of outcome variance.
- m2 told us that product attractiveness differences are larger than the transparency effect itself — the stimulus set matters enormously for generalization.
- m3 told us that site differences exist and must be accounted for in the estimand.
- m4 told us that the transparency effect is not a covariate artifact.

If the estimate had been confounded with risk level, m4 would have moved it. The fact that it did not is evidence that the design was clean.

The important teaching distinction: a stable estimate across a modeling sequence means the design worked. A changing estimate means the design had imbalance and the model is correcting for it. Both lessons are valuable — but they require different prose.
:::

**Next modeling decision:** m0 through m4 establish that the transparency effect is real and not an artifact of the design or data structure. Now we ask the substantive moderation question: *for whom* does transparency work?

---

## 10. Add the Theoretically Motivated Interaction

The study hypothesized that fee transparency helps low-literacy participants more. The logic: people with low financial literacy struggle to evaluate implicit or complex fee structures, so making fees explicit (transparent) helps them. High-literacy participants can infer fees regardless of format.

This is a theoretically motivated moderation, measured pre-treatment. It belongs in the model as an interaction.

```{r m5}
m5 <- lmer(adoption_intent ~ transparency * financial_literacy +
             risk_level + expected_value +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)

broom.mixed::tidy(m5, effects = "fixed") |>
  select(term, estimate, std.error, statistic, p.value) |>
  mutate(across(c(estimate, std.error, statistic, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = "m5: transparency × financial literacy interaction")
```

```{r m5-interaction-plot, fig.width=7, fig.height=4.5}
# Predicted effect across literacy values
lit_vals <- seq(-2, 2, length.out = 100)
pred_df <- tibble(
  financial_literacy = rep(lit_vals, 2),
  transparency       = rep(c(0, 1), each = 100),
  risk_level         = 0,
  expected_value     = 50
) |>
  mutate(
    participant_id = NA_integer_,
    site_id        = NA_integer_,
    stimulus_id    = NA_integer_
  )
pred_df$predicted <- predict(m5, newdata = pred_df, re.form = NA)

pred_df |>
  mutate(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  ggplot(aes(x = financial_literacy, y = predicted, color = Condition, linetype = Condition)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("#cccccc", "#1e3a5f")) +
  labs(
    title    = "Predicted adoption intent by transparency and financial literacy",
    subtitle = "Transparency effect is larger for low-literacy participants",
    x        = "Financial literacy (standardised)",
    y        = "Predicted adoption intent (0–100)",
    color    = NULL, linetype = NULL
  ) +
  theme_minimal(base_size = 12)
```

```{r m5-simple-slopes}
# Simple slopes: effect of transparency at -1SD, mean, +1SD literacy
b_tx    <- fixef(m5)["transparency"]
b_int   <- fixef(m5)["transparency:financial_literacy"]

tibble(
  `Financial literacy` = c("Low (−1 SD)", "Average (0)", "High (+1 SD)"),
  `Transparency effect` = round(c(b_tx + b_int * -1,
                                   b_tx + b_int *  0,
                                   b_tx + b_int *  1), 2),
  Interpretation = c(
    "Transparency helps low-literacy participants most",
    "Average transparency effect across the sample",
    "Transparency helps high-literacy participants least"
  )
) |>
  knitr::kable(caption = paste0("Simple slopes for transparency at three levels of financial literacy.",
    " True effect at average literacy = ", dgp$effect_transparency, " points;",
    " true interaction = ", dgp$interaction_tx_lit, "."))
```

::: {.callout-note icon=false}
## m5 Changes the Substantive Story

Models m0 through m4 answered: "Is there a transparency effect, and is it real?" They mostly changed how honestly we represented uncertainty and generalization — the point estimate was stable.

m5 answers a different question: "For whom is the effect larger?" The interaction is the first model that changes the *substantive interpretation* of the result. After m5, the transparency "main effect" is no longer "the universal effect" — it is the effect at average financial literacy. The effect is meaningfully larger for low-literacy participants and smaller for high-literacy participants.

This distinction matters for policy: if the goal is to increase adoption for the most financially vulnerable consumers, transparency disclosure is more valuable than the average effect suggests.
:::

**Student question:** Did we preregister this moderator?

**Advisor response:** If not, it is exploratory. We can report it, but with appropriate hedging: "We hypothesized that financial literacy would moderate the transparency effect. The interaction estimate was consistent with this hypothesis (*b* = [estimate]), but because this analysis was not preregistered, it should be replicated before being treated as confirmatory." The estimate is still worth reporting — it calibrates our understanding of *who* the intervention helps. But the inferential status is different from a preregistered test.

---

## 11. Consider Random Slopes

**Student question:** Transparency varies within participant — so don't we need a random slope for it?

**Advisor response:** The design *justifies* a random slope, yes. Each participant sees both transparent and non-transparent offers, which means participants *can* differ in how much the transparency manipulation moves them. A random slope captures that treatment-effect heterogeneity. But design justification and data support are separate questions. We will fit the model, check whether it converges cleanly, and then interpret what the slope variance tells us about heterogeneity — regardless of whether it is large or small.

The main lesson of this section is not about singularity handling. It is about what a random slope means substantively: **the transparency effect varies across participants**. Some people are more responsive to fee transparency than others. The random slope quantifies that spread.

```{r m6}
m6 <- lmer(adoption_intent ~ transparency * financial_literacy +
             risk_level + expected_value +
             (1 + transparency | participant_id) +
             (1 | stimulus_id) + (1 | site_id),
           data = trials,
           control = lmerControl(optimizer = "bobyqa"))

cat("Singular fit:", isSingular(m6), "\n")
```

```{r m6-slope-check}
vc6 <- as.data.frame(VarCorr(m6))
knitr::kable(vc6 |> select(grp, var1, var2, sdcor) |>
               mutate(across(where(is.numeric), \(x) round(x, 3))),
             caption = "Random effects from m6: by-participant random slope for transparency")
```

```{r m6-slope-viz, fig.width=7, fig.height=4}
# Visualise participant-level transparency effects
# (empirical: mean under transparent - mean under non-transparent, per person)
empirical_slopes <- trials |>
  group_by(participant_id) |>
  summarise(
    mean_tx    = mean(adoption_intent[transparency == 1], na.rm = TRUE),
    mean_notx  = mean(adoption_intent[transparency == 0], na.rm = TRUE),
    n_tx       = sum(transparency == 1),
    n_notx     = sum(transparency == 0)
  ) |>
  filter(n_tx > 0, n_notx > 0) |>
  mutate(empirical_effect = mean_tx - mean_notx)

vc6_slope_sd <- sqrt(vc6$vcov[vc6$grp == "participant_id" &
                               !is.na(vc6$var1) & vc6$var1 == "transparency" &
                               is.na(vc6$var2)])

ggplot(empirical_slopes, aes(x = empirical_effect)) +
  geom_histogram(bins = 25, fill = "#1e3a5f", color = "white", alpha = 0.8) +
  geom_vline(xintercept = fixef(m6)["transparency"],
             color = "#e63946", linetype = "dashed", linewidth = 1) +
  labs(
    title    = "Distribution of participant-level transparency effects",
    subtitle = paste0("Red line = average fixed effect. SD across participants ≈ ",
                      round(vc6_slope_sd, 1), " points."),
    x        = "Empirical transparency effect per participant (transparent − non-transparent)",
    y        = "Count"
  ) +
  theme_minimal(base_size = 11)
```

**What we learned:** The histogram shows real spread across participants. Some people are substantially more responsive to fee transparency than others. The random slope variance is the model's way of quantifying that spread. The fixed-effect SE increased modestly from m5 to m6 — the model now acknowledges that the average effect does not apply identically to every person.

::: {.callout-note icon=false}
## m6 Changes the Scientific Claim

m5 asked: "Is the average transparency effect modified by financial literacy?"
m6 asks: "Do participants differ in their transparency responsiveness, beyond what literacy explains?"

A non-trivial slope variance says: yes — some of the between-person heterogeneity in transparency sensitivity is not explained by the literacy moderator. That residual heterogeneity is real and the model is honest about it.
:::

```{r m6-uncorrelated}
# Try without the intercept-slope correlation as a robustness check
m6b <- lmer(adoption_intent ~ transparency * financial_literacy +
              risk_level + expected_value +
              (1 + transparency || participant_id) +
              (1 | stimulus_id) + (1 | site_id),
            data = trials,
            control = lmerControl(optimizer = "bobyqa"))

cat("Singular fit (correlated slopes m6):", isSingular(m6), "\n")
cat("Singular fit (uncorrelated slopes m6b):", isSingular(m6b), "\n")
```

```{r m6-comparison}
# Compare m5 and m6b to see whether slopes change fixed-effect inference
bind_rows(
  broom.mixed::tidy(m5,  effects = "fixed") |>
    filter(term %in% c("transparency", "transparency:financial_literacy")) |>
    mutate(model = "m5: no random slope"),
  broom.mixed::tidy(m6b, effects = "fixed") |>
    filter(term %in% c("transparency", "transparency:financial_literacy")) |>
    mutate(model = "m6b: uncorrelated random slope")
) |>
  select(model, term, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) |>
  knitr::kable(caption = "Does adding the random slope change fixed-effect inference?")
```

::: {.callout-warning icon=false}
## If the Random Slope Model Is Singular

Follow this sequence if either model above is singular:

1. **Remove the correlation parameter first** (`||` instead of `|`). Less demanding; often resolves the boundary issue.
2. **Check whether slope variance is near zero.** If removing the correlation still yields singularity, the slope variance may be genuinely negligible.
3. **Report the simplification sequence.** Do not silently drop the random slope. State what you tried, what happened, and what you retained.
4. **Retain m5 if slopes are not supportable** — m5 is theoretically complete. The simplification is defensible as long as you document it.
:::

**Student question:** If the random slope variance is non-trivial and the model is not singular, do we always keep it?

**Advisor response:** Use this decision rule rather than relying on `isSingular()` alone:

| Question | Verdict if Yes | Verdict if No |
|---|---|---|
| Is the slope theoretically justified (treatment varies within unit)? | Proceed | Do not add |
| Is the slope SD substantively non-trivial? | Keep | Consider dropping |
| Is the model non-singular? | Keep | Simplify |
| Does the fixed-effect SE change meaningfully? | Keep | Either choice defensible |

In the current example, the design justifies the slope, the slope SD is non-trivial, and the model is (likely) non-singular. We retain the uncorrelated random slope (m6b) as the primary model.

---

## 12. Estimator and Inference Choices

```{r estimator-table}
tibble(
  Decision         = c("REML vs ML", "Degrees-of-freedom method",
                        "Random slope comparison", "Optimizer"),
  Default          = c("REML for final LMM",
                        "Satterthwaite (via lmerTest)",
                        "LRT / model comparison",
                        "default (nloptwrap)"),
  Why              = c("Unbiased variance components",
                        "Practical approximation for mixed DFs",
                        "Tests contribution of variance component",
                        "Reproducible baseline"),
  `When to change` = c("Use ML (REML=FALSE) to compare fixed-effect structures",
                        "Kenward-Roger for small samples or unbalanced designs",
                        "Boundary issues: use parametric bootstrap",
                        "Try bobyqa or Nelder_Mead for convergence warnings")
) |>
  knitr::kable(caption = "Estimator decisions for linear mixed models")
```

**Student question:** Why do estimates change when switching between ML and REML?

**Advisor response:** REML (restricted maximum likelihood) accounts for the degrees of freedom used by fixed effects when estimating variance components. This gives unbiased variance estimates in finite samples. ML does not correct for this and tends to underestimate variance components, especially in small samples. For comparing two models that differ in their *fixed effects*, REML cannot be used because the restricted likelihoods are not comparable across different fixed-effect structures — the "restricted" part changes meaning. So: use REML for random-effects comparisons and final models; use ML for fixed-effects comparisons.

---

## 13. Post-Fit Diagnostics

Before trusting the final model, inspect residuals and check whether any assumptions are severely violated.

```{r final-model-selection}
# Final model decision using explicit criteria rather than isSingular() alone:
# 1. Slope is theoretically justified (within-participant design)
# 2. Slope SD is non-trivial (check from m6b variance components)
# 3. Model is non-singular
# 4. Fixed-effect SEs are stable (checked in m6-comparison above)
slope_sd <- tryCatch({
  vc_m6b  <- as.data.frame(VarCorr(m6b))
  # With || syntax lme4 may use different grp labels; match on var1 only
  tx_row  <- vc_m6b[!is.na(vc_m6b$var1) & vc_m6b$var1 == "transparency" &
                      is.na(vc_m6b$var2), ]
  if (nrow(tx_row) > 0) sqrt(tx_row$vcov[1]) else NA_real_
}, error = function(e) NA_real_)

# isTRUE() guards against NA / length-0 logicals from any of the three checks
keep_slope <- isTRUE(!isSingular(m6b)) &&
              isTRUE(!is.na(slope_sd))  &&
              isTRUE(slope_sd > 0.5)

m_final <- if (keep_slope) m6b else m5
cat("Slope SD:", if (is.na(slope_sd)) "NA" else round(slope_sd, 2), "\n")
cat("Model singular (m6b):", isSingular(m6b), "\n")
cat("Final model selected:",
    ifelse(keep_slope, "m6b (random slope retained)", "m5 (slope dropped)"), "\n")
```

```{r residual-diagnostics, fig.width=10, fig.height=4}
resid_df <- tibble(
  fitted   = fitted(m_final),
  residual = residuals(m_final),
  site_id  = trials$site_id,
  transparency = trials$transparency
)

p_rv <- ggplot(resid_df, aes(x = fitted, y = residual)) +
  geom_point(alpha = 0.2, color = "#1e3a5f") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#e63946") +
  labs(title = "Residuals vs Fitted", x = "Fitted values", y = "Residual") +
  theme_minimal(base_size = 11)

p_qq <- ggplot(resid_df, aes(sample = residual)) +
  stat_qq(alpha = 0.3, color = "#1e3a5f") +
  stat_qq_line(color = "#e63946") +
  labs(title = "QQ Plot of Residuals", x = "Theoretical quantiles", y = "Sample quantiles") +
  theme_minimal(base_size = 11)

p_site <- ggplot(resid_df, aes(x = factor(site_id), y = residual)) +
  geom_boxplot(fill = "#2d6a4f", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals by site", x = "Site", y = "Residual") +
  theme_minimal(base_size = 11)

gridExtra::grid.arrange(p_rv, p_qq, p_site, ncol = 3)
```

```{r diagnostic-checklist}
tibble(
  Category   = c("Fixed effects", "Fixed effects", "Fixed effects",
                  "Random effects", "Random effects", "Random effects",
                  "Residuals", "Residuals", "Residuals",
                  "Design", "Design"),
  Check      = c(
    "Focal estimate plausible in sign and magnitude?",
    "Estimate stable across model sequence?",
    "Interaction interpretable?",
    "Variance components nonzero?",
    "Model singular?",
    "Too few levels for any random effect?",
    "Residual vs. fitted: no funnel shape?",
    "QQ plot: residuals approximately normal?",
    "Heteroskedasticity by condition or site?",
    "Transparency balanced across risk and EV?",
    "Trial order effects?"
  ),
  Action = c(
    "Compare to true DGP or prior studies",
    "If estimate changes drastically, check confounding",
    "Compute simple slopes; plot predicted values",
    "If zero, report and simplify",
    "Simplify (see Step 11)",
    "Consider fixed effects for that grouping factor",
    "If yes, consider heteroskedastic model",
    "If severe, consider robust SE or transformation",
    "If yes, allow group-specific residual variance",
    "If unbalanced, ensure covariates are included",
    "Add trial_order as fixed effect and check estimate"
  )
) |>
  knitr::kable(caption = "Post-fit diagnostic checklist")
```

**Student question:** Which diagnostics change the model versus only narrow the interpretation?

**Advisor response:** Severe residual non-normality or heteroskedasticity changes the model (transformation, different family, heteroskedastic residuals). Singularity changes the model (simplification). But if the QQ plot shows mild tails or residuals are slightly heteroskedastic by site, the fixed-effect inference is usually robust. Those findings narrow the interpretation — we add a caveat — rather than requiring a new model.

---

## 14. Model Comparison Summary

```{r model-table}
extract_tx <- function(m, model_name, is_lm = FALSE) {
  if (is_lm) {
    broom::tidy(m) |> filter(term == "transparency") |>
      select(estimate, std.error, p.value) |>
      mutate(model = model_name)
  } else {
    broom.mixed::tidy(m, effects = "fixed") |>
      filter(term == "transparency") |>
      select(estimate, std.error, p.value) |>
      mutate(model = model_name)
  }
}

bind_rows(
  extract_tx(m0, "m0: pooled OLS", is_lm = TRUE),
  extract_tx(m1, "m1: + participant RI"),
  extract_tx(m2, "m2: + stimulus RI"),
  extract_tx(m3, "m3: + site RI"),
  extract_tx(m4, "m4: + risk/EV/literacy"),
  extract_tx(m5, "m5: + literacy interaction"),
  extract_tx(m_final, "m_final: primary model")
) |>
  mutate(
    across(c(estimate, std.error, p.value), \(x) round(x, 3)),
    `Still assumes...` = c(
      "Independence of all rows",
      "Stimuli and site are fixed",
      "Site and covariates are ignorable",
      "Covariates are ignorable",
      "Effect is homogeneous",
      "No effect heterogeneity by participant",
      "—"
    ),
    verdict = c("Reject", "Incomplete", "Better", "Better",
                 "Primary baseline", "Substantive model", "Final")
  ) |>
  select(model, estimate, std.error, p.value, `Still assumes...`, verdict) |>
  knitr::kable(caption = paste0("Model sequence summary. True transparency effect = ",
    dgp$effect_transparency, "."))
```

```{r what-changed-table}
tibble(
  Model  = c("m0 → m1", "m1 → m2", "m2 → m3", "m3 → m4", "m4 → m5", "m5 → m6b"),
  `What changed?` = c(
    "SE decreased; estimate stable",
    "SE decreased again; estimate stable",
    "Almost nothing changed",
    "Almost nothing changed",
    "Substantive interpretation changed",
    "SE increased modestly; estimate stable"
  ),
  `Main insight` = c(
    "Stable person baselines account for large outcome variance; within-person contrast is sharper",
    "Stimulus attractiveness is large — larger than the transparency effect itself; generalization requires modeling stimuli",
    "Site differences are baseline differences here, not transparency confounds; term matters for estimand and partitioning",
    "Covariates document robustness: balanced randomization means the estimate is not a covariate artifact",
    "Transparency effect depends on literacy — this is the first model to change *what we claim*, not just *how precisely we claim it*",
    "Participants differ in transparency responsiveness; fixed-effect SE is now honest about this heterogeneity"
  ),
  `Changed the substantive claim?` = c("No", "No", "Estimand only", "No", "Yes", "Partly")
) |>
  knitr::kable(caption = "What changed at each modeling step, and why it matters")
```

**The narrative version:**

We started with the tempting model — pooled OLS, one line of code, a clean significant result. It is wrong not because it is overconfident (in this balanced within-person design, it actually has the *largest* SE), but because it misrepresents the data structure and cannot support any claim about generalization.

We added participant random intercepts. The SE *decreased* because random intercepts separate stable person baselines from within-person contrasts — the model got more efficient, not more conservative. The point estimate was stable: the design was clean and transparency was not confounded with person-level differences.

Then we recognized that stimuli were also sampled. The stimulus caterpillar plot showed that between-product attractiveness differences have an SD larger than the transparency effect itself. This is the key insight from m2: we cannot generalize "fee transparency works for financial products" if we have only modeled the effect for these 36 specific offers.

Adding site changed the estimand without changing the estimate. Site differences here were baseline differences — the site term matters for what population we can speak to, not for correcting bias.

Adding covariates confirmed that the balanced randomization worked: risk level and expected value were not confounded with transparency assignment, and the estimate remained stable. This is evidence that the design was clean, not that the covariates were useless.

Adding the financial literacy interaction was the first model that changed the substantive story. The transparency effect is not uniform: it is larger for low-literacy participants. The "main effect" from m4 is now understood as the effect at average literacy, not the universal effect.

Adding the random slope was the last step: does the transparency effect also vary across participants, beyond what literacy explains? The slope variance and the slope histogram both say yes. The final SE is slightly larger than m5, which is correct: the model now acknowledges participant-level heterogeneity rather than assuming a common effect for everyone.

**Here is the model we can defend — and we can also explain why every earlier model was insufficient.**

---

## 14b. Stress Tests: When the Design Is Not Clean

The main worked example showed that a stable estimate across a modeling sequence is evidence of a well-designed study. But not every study is well-designed. The three short stress tests below show what happens when the reviewer's concerns are *actually present in the data*.

### Stress Test A: Accidental Stimulus Imbalance

What if transparency was more commonly assigned to high-risk products — perhaps because scheduling made it easier to show the transparent-fee display for riskier offers? High-risk products have lower adoption intent. This creates a downward bias in the unadjusted transparency estimate.

```{r stress-a}
# Re-use the latent variables already in trials (add them back temporarily)
trials_stress_a <- trials |>
  left_join(
    stimuli |> select(stimulus_id, risk_level, u0_stimulus),
    by = c("stimulus_id", "risk_level")
  ) |>
  left_join(
    participants |> select(participant_id, u0_participant, u1_participant, financial_literacy),
    by = c("participant_id", "financial_literacy")
  ) |>
  left_join(site_effects, by = "site_id") |>
  mutate(
    # Transparency now more likely for high-risk stimuli
    p_tx_imbalanced      = plogis(qlogis(0.50) - 0.50 * risk_level),
    transparency_imb     = rbinom(n(), 1, p_tx_imbalanced),
    mu_imb = dgp$intercept +
      dgp$effect_transparency * transparency_imb +
      dgp$effect_risk        * risk_level +
      dgp$effect_ev          * (expected_value - 50) / 15 +
      dgp$effect_literacy    * financial_literacy +
      dgp$interaction_tx_lit * transparency_imb * financial_literacy +
      u0_participant + u1_participant * transparency_imb +
      u0_stimulus + u0_site,
    adoption_imb = pmin(100, pmax(0, rnorm(n(), mu_imb, dgp$sd_residual)))
  )

m_imb_unadj <- lmer(adoption_imb ~ transparency_imb +
                      (1 | participant_id) + (1 | stimulus_id),
                    data = trials_stress_a)

m_imb_adj   <- lmer(adoption_imb ~ transparency_imb + risk_level + expected_value +
                      (1 | participant_id) + (1 | stimulus_id),
                    data = trials_stress_a)

bind_rows(
  broom.mixed::tidy(m_imb_unadj, effects = "fixed") |>
    filter(term == "transparency_imb") |>
    mutate(model = "Unadjusted (no risk/EV controls)"),
  broom.mixed::tidy(m_imb_adj, effects = "fixed") |>
    filter(term == "transparency_imb") |>
    mutate(model = "Adjusted (risk + EV as covariates)")
) |>
  select(model, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) |>
  knitr::kable(caption = paste0("Stress test A: accidental imbalance. True effect = ",
    dgp$effect_transparency, ". Unadjusted estimate is biased toward zero."))
```

**Teaching point:** In the clean design, covariates barely moved the estimate — they documented robustness. Here, they recover it. The balance check from Step 4 would have flagged this imbalance before modeling. If the balance check shows transparency concentrated among high-risk products, covariate adjustment is not optional — it is the correction.

---

### Stress Test B: Site-Level Treatment Heterogeneity

What if sites differ not only in baseline adoption but in *how much the transparency effect works*? A university lab population might respond more strongly to fee transparency than a community sample less accustomed to financial product decisions. A site random intercept does not capture this — it only captures baseline differences.

```{r stress-b}
# Add site-by-transparency interaction to the DGP
site_effects_b <- site_effects |>
  mutate(u1_site = c(1.5, -1.0, 0.5, -1.0))  # site-specific transparency slopes

trials_stress_b <- trials |>
  left_join(site_effects_b, by = "site_id") |>
  mutate(
    mu_b = dgp$intercept +
      (dgp$effect_transparency + u1_site) * transparency +
      dgp$effect_risk        * risk_level +
      dgp$effect_ev          * (expected_value - 50) / 15 +
      dgp$effect_literacy    * financial_literacy +
      u0_site,
    adoption_b = pmin(100, pmax(0, rnorm(n(), mu_b, dgp$sd_residual)))
  )

# Model A: site random intercept only (ignores site-by-treatment heterogeneity)
m_site_ri <- lmer(adoption_b ~ transparency +
                    (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
                  data = trials_stress_b)

# Model B: site fixed interactions (with 4 sites, random slope is unstable;
# fixed interactions show per-site effect directly)
m_site_ix <- lmer(adoption_b ~ transparency * factor(site_id) +
                    (1 | participant_id) + (1 | stimulus_id),
                  data = trials_stress_b)

# Compare site-specific transparency effects
site_tx_effects <- broom.mixed::tidy(m_site_ix, effects = "fixed") |>
  filter(str_detect(term, "transparency")) |>
  select(term, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3)))

knitr::kable(site_tx_effects,
  caption = "Stress test B: transparency effect interacts with site. A site random intercept alone would miss this.")
```

**Teaching point:** A site random intercept asks whether sites differ in baseline adoption. A site interaction asks whether the transparency effect itself differs across contexts — whether it works at all sites, or only at some. These are different questions. If your theoretical claim is "transparency works across contexts," a site interaction is the right diagnostic.

---

### Stress Test C: Unsupported Random Slopes (Sparse Within-Person Data)

Even when the design justifies a random slope, sparse within-person variation can make it empirically unsupported. Suppose each participant saw only 4 trials instead of 12 — with an average of 2 transparent and 2 non-transparent trials, individual slopes cannot be estimated reliably.

```{r stress-c}
# Subsample to 4 trials per participant
set.seed(42)
trials_sparse <- trials |>
  group_by(participant_id) |>
  slice_sample(n = 4) |>
  ungroup()

cat("Trials per participant in sparse design:", 4, "\n")

m_sparse_slope <- lmer(adoption_intent ~ transparency * financial_literacy +
                          risk_level + expected_value +
                          (1 + transparency | participant_id) +
                          (1 | stimulus_id) + (1 | site_id),
                        data = trials_sparse,
                        control = lmerControl(optimizer = "bobyqa"))

cat("Singular fit (sparse design, correlated slope):", isSingular(m_sparse_slope), "\n")

m_sparse_uncorr <- lmer(adoption_intent ~ transparency * financial_literacy +
                           risk_level + expected_value +
                           (1 + transparency || participant_id) +
                           (1 | stimulus_id) + (1 | site_id),
                         data = trials_sparse,
                         control = lmerControl(optimizer = "bobyqa"))

cat("Singular fit (sparse design, uncorrelated slope):", isSingular(m_sparse_uncorr), "\n")
```

**Teaching point:** With 4 trials per participant and roughly 2 per condition, the data cannot reliably distinguish one participant's transparency sensitivity from another's. The design *justifies* the random slope — the treatment does vary within person — but the data do not *support* it. This is the distinction between theoretical justification and empirical supportability. When the sparse model is singular, drop the slope, report the slope variance from the attempt, and note that the sparse design did not allow reliable estimation of individual differences in transparency responsiveness.

---

```{r final-interpretation}
fe <- broom.mixed::tidy(m_final, effects = "fixed") |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3)))

knitr::kable(fe |> select(term, estimate, std.error, statistic, p.value),
             caption = "Final model: fixed effects")

vc_final <- as.data.frame(VarCorr(m_final))
knitr::kable(vc_final |>
               select(grp, var1, vcov) |>
               mutate(SD = round(sqrt(vcov), 2), vcov = round(vcov, 2)),
             caption = "Final model: variance components")
```

**Plain-language interpretation:**

- **Transparency effect at average literacy:** Fee transparency increased willingness to adopt by approximately `r round(fixef(m_final)["transparency"], 1)` points on the 0–100 scale, on average across participants, stimuli, and sites. The true effect was `r dgp$effect_transparency` — the model recovered it reasonably well.

- **Literacy moderation:** The interaction term (if present) indicates that transparency provided greater benefit to participants with lower financial literacy, consistent with the hypothesis that explicit fee disclosure substitutes for financial knowledge in evaluating products.

- **Participant variance:** Participants differed substantially in their baseline adoption tendency. This stable individual-level variance was not confounded with the transparency effect; it was correctly partitioned into the participant random intercept.

- **Stimulus variance:** Product offers varied in their baseline attractiveness, with an SD of approximately `r dgp$sd_stimulus` points. This confirms that stimulus sampling matters: different stimulus sets would produce different point estimates.

- **Site variance:** Sites differed in baseline adoption intent. With only 4 sites, the site variance estimate should be interpreted cautiously; it indicates the presence of site-level differences but cannot precisely characterize how much effect heterogeneity exists across the broader population of possible sites.

- **Generalization:** The effect generalizes, within the bounds of these samples, to participants similar to those recruited, to financial products similar to these 36 offers, and to research contexts similar to these 4 sites. Broader generalization requires more representative sampling at each level.

---

## 16. What a Reviewer Would Ask

This section runs through the ten most likely reviewer challenges, framed as a supervision conversation to help you prepare.

---

**1. "Did you treat stimuli as random?"**

**Student question:** The reviewer is asking whether stimuli are random. I'm not sure what to say.

**Advisor response:** Say yes, and explain why. The 36 product offers were drawn from a broader universe of possible financial products. We want to generalize transparency's effect beyond these exact offers. The random-effects model allows us to estimate the average effect across that population. If we had treated stimuli as fixed, we would have been claiming that these 36 products are the only products that exist — which is not our claim and would have produced a falsely narrow SE.

---

**2. "Were transparency and product risk balanced?"**

**Advisor response:** Show the balance table from Step 4. If you used randomization within each participant-stimulus pairing, the allocation is balanced in expectation. Show the realized proportions. If there was imbalance, explain that including risk level and expected value as covariates adjusts for this in the analysis.

---

**3. "Can you generalize beyond these stimuli?"**

**Advisor response:** Only to the extent these 36 offers represent the population of financial products of this type. We should describe the offer set (product categories, fee structures represented, range of risk and EV) and note what kinds of products are not represented. The honest answer is: we can generalize to products similar to these, under the assumption that the sample was reasonably representative.

---

**4. "Why did you include participant random slopes?"**

**Advisor response:** Because transparency varied within participant — each person saw both transparent and non-transparent offers. This within-participant variation means participants *could* differ in how much the manipulation moved them. A random slope allows the model to estimate that heterogeneity rather than assuming every participant responds to transparency identically. The design justifies it; we then checked whether the data supported it.

---

**5. "Why did you drop a random slope if you dropped one?"**

**Student question:** What if the reviewer asks why we simplified?

**Advisor response:** Show the singularity check. "We initially fit a model with a by-participant random slope for transparency as justified by the within-participant design. The model was [singular / showed convergence warnings]. We removed the intercept-slope correlation and confirmed that the simplified model was non-singular. The slope variance estimate was [value], and removing it did not change the transparency fixed-effect estimate meaningfully [show comparison table]. We retained the simpler model and report the simplification."

---

**6. "Are site effects meaningful with only 4 sites?"**

**Advisor response:** "We included site as a random effect because participants were nested in sites and site means differed visibly. With J = 4 sites, the site variance estimate is imprecise — it should be interpreted as an indication that site-level baseline differences exist, not as a precise characterization of effect heterogeneity across contexts. The transparency effect is estimated within sites, so site variance does not directly threaten the focal estimate. We recommend future work with more diverse site sampling."

---

**7. "Could trial order or fatigue explain the effect?"**

**Advisor response:** Check the trial-order correlation. "We examined trial order as a potential confound. If transparency was accidentally more common in later trials — when fatigue effects are larger — the transparency estimate would be downward-biased. We checked that transparency assignment was balanced across trial order positions. [Show the check.] We also fit a model including trial order as a fixed effect and confirmed the transparency estimate was stable."

---

**8. "Does financial literacy moderate the effect, or is that post hoc?"**

**Advisor response:** "Financial literacy was measured before the experiment began and was specified as a theoretically motivated moderator based on prior work suggesting that fee transparency aids understanding primarily for those with lower financial knowledge. If preregistered, this is confirmatory. If not preregistered, it is exploratory and should be replicated."

---

**9. "Did the model converge cleanly?"**

**Advisor response:** "The final model [m_final] converged without warnings using the bobyqa optimizer. The model is non-singular [check with isSingular()]. We checked convergence by restarting with two additional optimizers and obtained equivalent estimates."

---

**10. "What is the estimand?"**

**Advisor response:** The estimand is the average treatment effect of fee transparency on willingness to adopt, averaged across the population of participants from which our sample was drawn, across the population of financial product offers that our stimuli represent, and across the population of research contexts similar to our sites. It is *not* the effect for these specific 36 products or these specific 4 sites. The crossed random-effects structure is what licenses the "across the population" part of that claim.

---

## 17. Reporting Template

Adapt this template for a methods section. Replace bracketed values with actual estimates.

---

> We analyzed adoption-intent ratings (0–100) using a linear mixed-effects model fitted with `lme4` (Bates et al., 2015) and degrees of freedom estimated via the Satterthwaite approximation (`lmerTest`; Kuznetsova et al., 2017).
>
> **Design structure.** Each participant rated 12 of 36 product offers; participants were nested in 4 research sites. Fee transparency varied within participant across trials. The design is partially crossed: participants and stimuli form a partially crossed random-effects structure, while participants are nested within sites.
>
> **Model specification.** The model included fixed effects for fee transparency (0/1), financial literacy (standardized, measured pre-experiment), their interaction, product risk level (standardized), and expected value. Random intercepts were included for participants, stimuli, and sites, treating all three as random samples from their respective populations. [If random slopes were retained: A by-participant random slope for transparency was included because transparency varied within participant; the intercept-slope correlation was removed to address [singular fit / convergence issues]. If dropped: A by-participant random slope for transparency was initially considered but produced [singular fit / near-zero slope variance; *SD* = X]; it was dropped and the result was stable.]
>
> **Model selection.** Models were compared sequentially: pooled OLS → participant random intercepts → crossed participant + stimulus random intercepts → + site random intercepts → + design covariates → + literacy interaction → + random slopes. The final model was selected after confirming convergence and non-singularity.
>
> **Results.** Fee transparency increased willingness to adopt by *b* = [X.X] points (95% CI: [X.X, X.X]), *t*([df]) = [X.X], *p* = [.XXX], on average across participants, stimuli, and sites. Financial literacy moderated this effect, *b* = [X.X], *t*([df]) = [X.X], *p* = [.XXX]: the transparency benefit was larger for participants lower in financial literacy [describe direction]. Participant, stimulus, and site random-intercept SDs were [X.X], [X.X], and [X.X], respectively.
>
> **Generalization.** Because stimuli were treated as a random sample from the population of possible financial product descriptions, the transparency effect estimate generalizes beyond these specific 36 offers, under the assumption that the offer set was reasonably representative of the target class of products. Site generalization is limited by the small number of sites (J = 4).

---

::: {.callout-tip icon=false}
## Minimum Reproducible Model Report

Any multilevel analysis of a nested experiment should include:

- [ ] Model formula (verbatim or in mathematical notation)
- [ ] Software and version used
- [ ] Optimizer used and whether the model converged
- [ ] Whether the model is singular (`isSingular()`)
- [ ] Estimator (REML or ML) and df method
- [ ] All fixed-effect estimates with SEs and CIs
- [ ] All random-effect variance components (SDs) with grp labels
- [ ] Sequence of models compared and why the final model was chosen
- [ ] What was tried and simplified, and why
- [ ] Explicit statement about the estimand (what population does this generalize to?)
- [ ] At least one residual diagnostic inspection
:::
