---
title: "Maximal vs. Parsimonious Random Effects"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
set.seed(77)
```

# Maximal vs. Parsimonious Random Effects

## The Debate

In 2013, Barr, Levy, Scheepers, & Tily published an influential paper titled "Random Effects Structure for Confirmatory Hypothesis Testing: Keep It Maximal." Their argument:

> To control Type I error at the nominal level, researchers should include the *maximal* random effects structure — every random slope that is justified by the experimental design.

In a within-subjects design where every participant sees every condition, you must include a random slope for condition by participants. Omitting this random slope will produce anti-conservative p-values.

Two years later, Matuschek, Kliegl, Vasishth, Baayen, & Bates (2017) pushed back in a paper titled "Balancing Type I Error and Power in Linear Mixed Models." Their argument:

> Maximal models frequently fail to converge or produce degenerate variance components. Over-parameterized models can actually have *worse* power than well-specified parsimonious models. Use likelihood ratio tests to decide what to include.

Both papers are correct. The question is when each applies.

## Why Maximal Models Can Fail

```{r convergence-example}
# Simulate a within-subjects design
# Each participant sees each condition 4 times (trials)
# 40 participants × 3 conditions × 4 trials = 480 observations
# Random effects needed: 40 × 3 = 120 — well below 480
n_participants <- 40
n_conditions   <- 3
n_trials       <- 4

participant_intercepts <- rnorm(n_participants, 50, 8)
# Small (but nonzero) random slopes
participant_slopes     <- rnorm(n_participants, 0, 0.5)

df_within <- expand_grid(
  participant = 1:n_participants,
  condition   = 1:n_conditions,
  trial       = 1:n_trials
) |>
  mutate(
    condition_eff = c(-2, 0, 2)[condition],  # condition effect
    p_slope       = participant_slopes[participant],
    Y = participant_intercepts[participant] + condition_eff +
        p_slope * condition_eff + rnorm(n(), 0, 5),
    participant = factor(participant),
    condition   = factor(condition)
  )

# Maximal model
m_max <- lmer(Y ~ condition + (condition | participant),
              data    = df_within,
              control = lmerControl(optimizer = "bobyqa"))

# Does it converge?
library(performance)
check_convergence(m_max)

# Parsimonious model (random slope variance is tiny — can drop it)
m_pars <- lmer(Y ~ condition + (1 | participant), data = df_within)
```

```{r compare-models}
# Compare: do the p-values differ substantially?
anova(m_pars, m_max)
```

## The Decision Framework

The current consensus among methodologists:

::: {.callout-note}
## When to Use Maximal Random Effects

1. **Within-subjects experimental manipulations:** Always include random slopes for within-subjects factors. This is the critical case. Barr et al. (2013) are correct: omitting random slopes for within-subject conditions inflates Type I error.

2. **When theory or pilot data suggests heterogeneous effects:** If you have prior reason to believe the treatment effect varies substantially across participants, include the random slope.

3. **Large samples:** With enough data, complex models converge and the efficiency loss from over-parameterization is small.
:::

::: {.callout-note}
## When to Simplify (Parsimonious)

1. **Convergence failures:** If the maximal model does not converge, you must simplify. Start by removing the correlation between random intercept and slope (`(1 | participant) + (0 + condition | participant)`). Then consider removing the random slope itself.

2. **Degenerate variance components:** If the random slope variance is estimated at exactly zero (or very near), it is contributing nothing and can be dropped.

3. **Very small variance components:** Use an LRT to test whether the random slope significantly improves fit. If not, the parsimonious model is appropriate.
:::

## A Practical Simplification Protocol

```{r simplification-protocol}
# Step 1: Try the maximal model
m_max_2 <- tryCatch(
  lmer(Y ~ condition + (condition | participant),
       data = df_within, control = lmerControl(optimizer = "bobyqa")),
  warning = function(w) { cat("Warning:", conditionMessage(w), "\n"); NULL }
)

# Step 2: If it fails/has issues, remove correlation first
m_no_corr <- lmer(Y ~ condition + (1 | participant) + (0 + condition | participant),
                  data = df_within)

# Step 3: Test: does the random slope improve fit?
anova(m_pars, m_no_corr)

# If p > 0.05 on the LRT, the random slope does not help — use m_pars
```

::: {.callout-warning icon=false}
## Assumption: The Random Effects Structure Matches the Design

**What it means:** The random effects you include should reflect the actual nesting structure of your design. A random slope for a between-subjects factor makes no statistical sense (between-subjects factors cannot vary within a subject).

**Common mistake:** Including a random slope for a between-subjects condition. This will either produce a degenerate model or absorb all variance in a nonsensical way.

**How to check:** For each predictor, ask: does this variable vary *within* the grouping unit (participant, stimulus)? If yes, it can have a random slope for that unit. If no (between-subjects), it cannot.
:::
