---
title: "Multi-Site Experiments"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
set.seed(2024)
```

# Multi-Site Experiments

::: {.callout-note icon=false}
## The Multi-Site Design Question

> **Are sites a sample from a population of possible contexts?** If yes, model site as a random effect. If your sites are the entire target population (e.g., every branch of one specific bank), fixed site effects may be more appropriate.
>
> **How much does the treatment effect vary across sites?** This is the key scientific question in a many-labs study. A large random-slope variance for treatment means the effect is heterogeneous — it works well in some contexts and not others.
>
> **What claim can you make if sites are not representative?** An average effect across the convenience sample of sites you recruited — not a generalizable estimate for all contexts of the type.
:::

---

### Multi-Site Design Audit

Before fitting the multi-site model, audit the site structure the same way you audit participants and stimuli.

::: {.callout-note icon=false}
## Site Audit Before Analysis

| Question | Why it matters | What to check |
|----------|---------------|---------------|
| How were sites recruited? | Determines generalizability | Opportunistic, systematic, or stratified recruitment |
| Are sites the target population or a convenience sample? | Limits scope of inference | Compare site characteristics to the intended population |
| Was treatment randomized within site or by site? | Determines assignment level | If by site: site IS the unit; need many sites |
| Are sample sizes balanced across sites? | Affects estimation of site-level variance | Plot N by site; check range |
| Was the protocol identical across sites? | Confounds implementation vs. site heterogeneity | Protocol fidelity documentation |
| Were experimenters trained and monitored? | Experimenter effects are a form of site heterogeneity | Training records; fidelity checks |
| How many sites are there? | Determines precision of random-slope variance | J < 10: interpret heterogeneity estimates cautiously |
:::

---

## Generalizability is a Nesting Problem

The replication crisis in psychology revealed that many findings were fragile — they held at one university, with one sample, run by one experimenter, but failed to replicate elsewhere. The problem is that **site** is a nesting variable that standard analysis ignores.

A **multi-site experiment** (also called a many-labs design) runs the same study across multiple locations: different universities, countries, online platforms, or time periods. Each site contributes data. Sites are treated as a random sample from a larger population of sites.

This design answers a richer question: not "does this effect occur at location X?" but "does this effect occur, on average, across the population of contexts?" And it allows you to ask: **how much does the effect size vary across sites?**

## The Model

Sites are random. The treatment effect may vary across sites (cross-level interaction between treatment and site).

$$Y_{ijk} = \gamma_{00} + \gamma_{10} \text{Treat}_{ij} + u_{0k} + u_{1k} \text{Treat}_{ij} + \varepsilon_{ijk}$$

Where:
- $k$ = site; $j$ = participant; $i$ = observation
- $u_{0k}$: site-level random intercept (baseline differences across sites)
- $u_{1k}$: site-level random slope for treatment (heterogeneity in treatment effect)
- $\gamma_{10}$: average treatment effect across sites

```{r many-labs-sim}
n_sites       <- 20
n_per_site    <- 40  # 20 treatment, 20 control

site_intercepts <- rnorm(n_sites, 50, 5)   # sites differ in baseline
site_effects    <- rnorm(n_sites, 5, 3)    # treatment effect varies across sites
# Average effect = 5; some sites show 8, some show 2

df_ml <- expand_grid(
  site = 1:n_sites,
  participant = 1:n_per_site
) |>
  mutate(
    treatment   = rep(c(rep(0, 20), rep(1, 20)), n_sites),
    site_base   = site_intercepts[site],
    site_effect = site_effects[site],
    outcome     = site_base + site_effect * treatment + rnorm(n(), 0, 6),
    site        = factor(site)
  )
```

```{r fit-many-labs}
# Random intercepts and slopes for sites
m_ml <- lmer(outcome ~ treatment + (treatment | site), data = df_ml)

broom.mixed::tidy(m_ml, effects = "fixed") |>
  knitr::kable(digits = 3,
    caption = "Average treatment effect across sites")

broom.mixed::tidy(m_ml, effects = "ran_pars") |>
  knitr::kable(digits = 3,
    caption = "Variance in treatment effect across sites (how heterogeneous is the effect?)")
```

```{r forest-plot}
# Forest plot with uncertainty intervals — not just point estimates

# Extract site-specific estimates: BLUP + shrinkage
site_coefs <- coef(m_ml)$site |>
  as.data.frame() |>
  rownames_to_column("site") |>
  rename(intercept = `(Intercept)`, effect = treatment)

# Approximate SE for site estimates via simulation from the fitted model
# (for a clean forest plot we use the lmer variance + residual to compute a naive SE)
n_per_site_val  <- n_per_site / 2   # ~20 per arm per site
resid_se        <- sigma(m_ml)
slope_se_approx <- resid_se / sqrt(n_per_site_val)   # rough per-site SE

site_estimates <- site_coefs |>
  mutate(
    se_approx = slope_se_approx,
    ci_lo     = effect - 1.96 * se_approx,
    ci_hi     = effect + 1.96 * se_approx
  ) |>
  arrange(effect)

avg_effect <- fixef(m_ml)["treatment"]

ggplot(site_estimates, aes(x = reorder(site, effect), y = effect)) +
  geom_hline(yintercept = avg_effect,
             color = "#e63946", linetype = "dashed", linewidth = 1) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0.25, color = "#1e3a5f", linewidth = 0.7) +
  geom_point(size = 3, color = "#1e3a5f") +
  annotate("text", x = 1, y = avg_effect + 0.5,
           label = paste0("Average: ", round(avg_effect, 1)),
           color = "#e63946", hjust = 0, size = 3.2) +
  coord_flip() +
  labs(
    title    = "Site-specific treatment effects with uncertainty intervals",
    subtitle = "Red dashed line = average effect across sites. Overlap with zero = this site's estimate is not significant alone.",
    x        = "Site",
    y        = "Treatment Effect (with 95% CI)"
  ) +
  theme_minimal(base_size = 12)
```

**Why uncertainty intervals matter in a forest plot:** Site-specific point estimates from a random-effects model are *shrunken* toward the average — they understate the true variability across sites. Showing the uncertainty around each estimate reminds the reader that no individual site's deviation from the average is precisely estimated. A site that appears to show a null effect may simply have wide uncertainty, not a true zero.

```{r leave-one-out-sensitivity}
# Leave-one-site-out sensitivity: how much does the average effect change?
loo_results <- map_dfr(unique(df_ml$site), function(s) {
  df_sub <- df_ml |> filter(site != s)
  m_sub  <- lmer(outcome ~ treatment + (treatment | site),
                 data = df_sub, control = lmerControl(optimizer = "bobyqa"))
  tibble(
    dropped_site = s,
    avg_effect   = fixef(m_sub)["treatment"]
  )
})

ggplot(loo_results, aes(x = reorder(dropped_site, avg_effect), y = avg_effect)) +
  geom_hline(yintercept = fixef(m_ml)["treatment"],
             color = "#e63946", linetype = "dashed") +
  geom_point(color = "#1e3a5f", size = 2.5) +
  coord_flip() +
  labs(
    title    = "Leave-one-site-out sensitivity",
    subtitle = "How much does the average effect change when each site is removed?\nLarge swings suggest one site is influential.",
    x        = "Site dropped",
    y        = "Average effect (leave-one-out)"
  ) +
  theme_minimal(base_size = 12)
```

---

## Heterogeneity Diagnostics

```{r heterogeneity-table}
# Post-fit summary: what does the site-level variance tell us?
var_te  <- as.data.frame(VarCorr(m_ml))$vcov[2]
sd_te   <- sqrt(var_te)
avg_te  <- fixef(m_ml)["treatment"]

tibble(
  Quantity             = c("Average treatment effect", "SD of effect across sites",
                           "80% plausible interval for site effects",
                           "Proportion of sites where effect > 0 (approx)"),
  Value = c(
    round(avg_te, 2),
    round(sd_te, 2),
    paste0("[", round(avg_te - 1.28 * sd_te, 1), ", ", round(avg_te + 1.28 * sd_te, 1), "]"),
    round(pnorm(avg_te / sd_te), 2)
  )
) |>
  knitr::kable(caption = "Treatment effect heterogeneity across sites. The 80% plausible interval shows the range of effects you would expect across a new sample of similar sites.")
```

::: {.callout-warning icon=false}
## Small-J Warning: Few Sites, Unreliable Heterogeneity Estimates

The random-slope variance for treatment effects — the quantity that tells you how heterogeneous the effect is across sites — is itself estimated with substantial uncertainty when J (number of sites) is small. With J < 10 sites, the random-slope variance estimate has wide confidence intervals that overlap zero even when true heterogeneity is moderate.

**Practical consequences of few sites:**
- The average treatment effect remains interpretable even with few sites
- The site heterogeneity estimate (random slope variance) is unreliable
- Do not over-interpret whether the effect "varies meaningfully" across sites with J < 10
- Forest plots with few sites should show uncertainty intervals — point-only plots hide how imprecise each site estimate is

**What claim survives with few sites:**
- If sites are convenience-sampled: "The effect averaged *b* = [estimate] across the [J] sites studied, which represent [describe recruitment]. Generalization to other contexts requires similar sites."
- If implementation differed across sites: "Observed site heterogeneity may reflect implementation differences rather than genuine psychological heterogeneity across populations."
:::

---

## The ICC for Treatment Effects

The random slope for treatment tells you how much the effect varies across sites. The **I²** statistic (from meta-analysis) provides a related measure: what proportion of the total variation in effect size estimates is due to genuine heterogeneity vs. sampling error?

```{r heterogeneity}
# Variance of the treatment effect across sites
var_treatment_effect <- as.data.frame(VarCorr(m_ml))$vcov[2]  # slope variance
cat("Standard deviation of treatment effect across sites:",
    round(sqrt(var_treatment_effect), 2), "\n")
cat("Average effect:", round(fixef(m_ml)["treatment"], 2), "\n")
cat("Effect ranges approximately from",
    round(fixef(m_ml)["treatment"] - 2*sqrt(var_treatment_effect), 1),
    "to",
    round(fixef(m_ml)["treatment"] + 2*sqrt(var_treatment_effect), 1),
    "across sites\n")
```

::: {.callout-warning icon=false}
## Assumption: Sites are a Representative Sample

**What it means:** The multi-site model generalizes to the population of sites from which your sites were drawn. If your 20 labs are all top-10 research universities, the results generalize to that population — not to community colleges, online platforms, or international contexts.

**How to check:** Describe your site-selection process. Were sites systematically sampled? Opportunistically recruited? The inference should match the selection.
:::

::: {.callout-warning icon=false}
## Assumption: Treatment Procedure is Comparable Across Sites

**What it means:** The "treatment" must be operationally equivalent across sites. If different labs implemented the procedure differently — different materials, different timing, different experimenter instructions — site-to-site variability reflects procedure differences, not true effect heterogeneity.

**How to check:** Pre-specify and enforce standardized protocols. Measure implementation fidelity. Include implementation fidelity as a moderator.
:::

::: {.callout-warning icon=false}
## Assumption: Enough Sites to Estimate Heterogeneity

**What it means:** Estimating random-slope variance requires a sufficient number of level-2 units. With J = 5 sites, the random-slope variance estimate can be zero even when true heterogeneity is substantial. The average effect remains estimable, but the heterogeneity estimate is unreliable.

**What to do:** With J < 10, use the average effect as the primary claim. Treat heterogeneity estimates as exploratory. Consider presenting a prediction interval for the average effect rather than a confidence interval.
:::

---

## Reviewer Challenge

::: {.callout-tip}
## What a Reviewer Would Ask

**Challenge 1:** "Your sites are all US universities. How does this generalize?"
It generalizes to contexts similar to US university labs — convenience-sampled student populations in controlled laboratory settings. This is a limitation that should be stated, not dismissed. If generalizability is important, report it as a scope limitation and suggest what a more representative site sample would look like.

**Challenge 2:** "Your random-slope variance for treatment is large. Does that mean the intervention doesn't work in some sites?"
Possibly — or the site heterogeneity reflects implementation differences, sample composition differences, or imprecise estimation from small within-site samples. Inspect the forest plot for which sites show negative effects and whether those sites share identifiable features (e.g., smaller N, different recruitment, known implementation problems).

**Challenge 3:** "Why model sites as random effects rather than including site fixed effects?"
Site random effects allow you to estimate the average treatment effect across sites and to quantify how much the effect varies. Site fixed effects absorb all between-site variance but cannot produce a generalization-weighted average effect or a heterogeneity estimate. Random effects are appropriate when sites are a sample from a population; fixed effects are appropriate when sites are the entire target population or when there is concern about confounding between site characteristics and treatment assignment.
:::

---

## Reporting Checklist for Multi-Site Experiments

::: {.callout-note icon=false}
## Minimum Reporting for Multi-Site Studies

- [ ] Number of sites, how recruited, and description of site characteristics
- [ ] Whether treatment was randomized within sites or by site
- [ ] Protocol fidelity description
- [ ] Sample sizes per site (range and mean)
- [ ] Average treatment effect with SE and CI
- [ ] Random intercept and random slope variance estimates
- [ ] Forest plot with site-specific estimates (with uncertainty intervals)
- [ ] Leave-one-site-out sensitivity check
- [ ] Explicit statement about generalizability given the site sample
:::

