---
title: "A Fully Worked Multilevel Modeling Example"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
set.seed(20240610)
```

# A Fully Worked Multilevel Modeling Example

A student brings us a result: transparent fee disclosure appears to increase adoption intent. The p-value is small. The plot looks clean. But before we believe the estimate, we audit the data. We find **seven threats**: repeated observations, sampled stimuli, treatment-risk imbalance, nonlinear product valuation and fatigue, differential missingness, post-treatment exclusion risk, and treatment-effect heterogeneity.

The rest of this chapter shows how to diagnose where each threat comes from, which statistical adjustment reduces it, and what remains when the chapter ends.

> **The key framing:** Statistical fixes minimize threats — they do not erase them. The goal is not to find an analysis that says everything was fine. It is to show exactly what each adjustment addressed, what it could not address, and what the final claim depends on.

---

## 1. The Research Question and Estimand

> **Does transparent fee disclosure increase willingness to adopt a financial product, and does that effect generalize across participants, product offers, and research sites?**

The **target estimand** is the average total effect of *assigned* fee transparency on willingness to adopt, averaged over the population of participants, product offers, and research contexts similar to those sampled. This is an intention-to-treat estimand: the effect of the display assignment, not the effect of comprehension or perceived clarity.

This question contains several distinct questions a naive regression cannot answer:

- Effect for *these* participants (descriptive)
- Effect across the *population of participants* (requires participant random effects)
- Effect across the *population of product offers* (requires stimulus random effects)
- Effect across *different sites* (requires site random effects)
- Effect conditional on *financial literacy* (requires the interaction)

---

## 2. Design Audit Before Modeling

| Design question | Answer | Modeling implication |
|---|---|---|
| What is one row? | One participant's rating of one product | Trial-level outcome; multiple rows per participant |
| What is randomized? | Fee transparency display (within-participant) | Fixed effect for transparency |
| Does treatment vary within participant? | Yes — each sees both conditions | Candidate by-participant random slope |
| Are stimuli sampled? | Yes — 36 product offers from a larger set | Stimulus random intercept |
| Are participants crossed with stimuli? | Partially — each rates 12 of 36 | Crossed (partial) random effects |
| Are participants nested in sites? | Yes — 240 across 4 sites | Site random intercept |
| Is the assignment perfectly randomized? | **No** — correlated with risk, trial order, site 2 | Covariate adjustment needed |
| What missingness is plausible? | Higher for transparent, high-risk, late, low-literacy | Missingness audit required |
| Which variables are post-treatment? | Comprehension, perceived transparency | Exclude from main model |

: Design audit: the assignment is not clean. This is not a drill.

---

## 3. Simulate the Latent DGP

The true data-generating process is known. But the *realized* data will not look like it, for three reasons: (1) the fee-transparency display was more often served to high-risk products in later trials and at site 2 — a real implementation pattern mimicking a platform rollout; (2) trial-order fatigue is real and non-trivial; (3) the observation process introduces non-random missingness and attention failures.

```{r true-dgp-params}
dgp <- list(
  intercept           = 50,
  effect_transparency = 4,      # +4 at average literacy — the true target
  effect_risk         = -2.0,   # linear risk penalty
  effect_risk_high    = -2.5,   # extra threshold penalty for risk > 0.75
  effect_ev_linear    = 2.2,    # EV (standardised), linear
  effect_ev_quad      = -1.0,   # diminishing returns on EV
  effect_literacy     = 2,
  interaction_tx_lit  = -1.5,   # transparency helps low-literacy more
  effect_fatigue      = -0.18,  # per trial
  effect_fatigue_late = -1.20,  # additional penalty for trials 10–12
  effect_device       = -1.5,   # mobile device: lower ratings
  sd_participant      = 8,
  sd_stimulus         = 5,
  sd_site             = 4,
  sd_slope_part       = 2,
  sd_residual         = 7
)
```

```{r simulate-latent}
n_participants  <- 240
n_stimuli       <- 36
n_sites         <- 4
trials_per_part <- 12

participants <- tibble(
  participant_id     = 1:n_participants,
  site_id            = rep(1:n_sites, each = n_participants / n_sites),
  financial_literacy = rnorm(n_participants, 0, 1),
  u0_participant     = rnorm(n_participants, 0, dgp$sd_participant),
  u1_participant     = rnorm(n_participants, 0, dgp$sd_slope_part)
)

site_effects <- tibble(
  site_id = 1:n_sites,
  u0_site = c(3, -2, 1, -2)
)

stimuli <- tibble(
  stimulus_id    = 1:n_stimuli,
  risk_level     = rnorm(n_stimuli, 0, 1),
  expected_value = rnorm(n_stimuli, 50, 15),
  u0_stimulus    = rnorm(n_stimuli, 0, dgp$sd_stimulus)
) |>
  mutate(
    expected_value_z = (expected_value - 50) / 15,
    risk_high        = as.integer(risk_level > 0.75)
  )

trials <- participants |>
  select(participant_id, site_id, financial_literacy, u0_participant, u1_participant) |>
  cross_join(tibble(draw = 1:trials_per_part)) |>
  group_by(participant_id) |>
  mutate(
    stimulus_id   = sample(1:n_stimuli, trials_per_part, replace = FALSE),
    trial_order   = 1:trials_per_part
  ) |>
  ungroup() |>
  select(-draw) |>
  left_join(stimuli, by = "stimulus_id") |>
  left_join(site_effects, by = "site_id") |>
  mutate(
    # Device: more common at sites 2–3 (online platforms)
    device_mobile = rbinom(n(), 1, plogis(-0.3 + 0.6 * as.integer(site_id %in% c(2, 3)))),

    # IMPERFECT ASSIGNMENT: transparency correlated with risk, trial order, and site 2
    # This mimics a platform that served the transparent display more often for
    # high-risk products and in later trials — a realistic implementation artefact
    trial_order_z = (trial_order - 6.5) / 3.452,
    p_tx = plogis(0.70 * risk_level + 0.40 * trial_order_z + 0.50 * as.integer(site_id == 2)),
    transparency  = rbinom(n(), 1, p_tx),

    # Latent mean (true DGP — clean causal structure)
    mu = dgp$intercept +
         dgp$effect_transparency  * transparency +
         dgp$effect_risk          * risk_level +
         dgp$effect_risk_high     * risk_high +
         dgp$effect_ev_linear     * expected_value_z +
         dgp$effect_ev_quad       * expected_value_z^2 +
         dgp$effect_literacy      * financial_literacy +
         dgp$interaction_tx_lit   * transparency * financial_literacy +
         dgp$effect_fatigue       * trial_order +
         dgp$effect_fatigue_late  * as.integer(trial_order > 9) +
         dgp$effect_device        * device_mobile +
         u0_participant + u1_participant * transparency +
         u0_stimulus + u0_site,

    # Heteroskedastic residuals: more variance for high-risk products and
    # low-literacy participants
    resid_sd = dgp$sd_residual * exp(
      0.25 * abs(risk_level) +
      0.20 * as.integer(financial_literacy < -1)
    ),
    adoption_intent = pmin(100, pmax(0, rnorm(n(), mu, resid_sd)))
  ) |>
  select(participant_id, site_id, stimulus_id, trial_order,
         transparency, risk_level, risk_high, expected_value, expected_value_z,
         financial_literacy, device_mobile, adoption_intent)
```

---

## 4. The Observation Process

A real experiment does not deliver `adoption_intent` for every trial. Between the latent DGP and the analyst's dataset:

- Some trials are missing because participants quit, skip, or time out.
- Missingness depends on the treatment (transparent displays require more effort), risk level, trial position, literacy, and device.
- Some trials produce attention-check failures — and those failures are partly caused by the treatment.

```{r simulate-observed}
set.seed(20240611)

trials <- trials |>
  mutate(
    # Missingness: related to treatment, risk, trial fatigue, literacy, device
    p_missing = plogis(
      -2.5 +
        0.45 * transparency +
        0.55 * risk_level +
        0.45 * as.integer(trial_order > 9) -
        0.50 * financial_literacy +
        0.35 * device_mobile
    ),
    missing_y           = rbinom(n(), 1, p_missing),
    adoption_intent_obs = if_else(missing_y == 1, NA_real_, adoption_intent),

    # Attention failures: more common under transparent displays, late trials,
    # and for low-literacy participants — a post-treatment process variable
    attention_fail = rbinom(n(), 1,
      plogis(-3.0 +
             0.50 * transparency +
             0.35 * risk_level +
             0.45 * as.integer(trial_order > 9) -
             0.40 * financial_literacy))
  ) |>
  select(-p_missing)
```

---

## 5. Initial Inference Threat Dashboard

::: {.callout-important icon=false}
## We Are Starting With 7 Open Threats

Before fitting a single model, we inventory the inference problems visible in the data. These are not hypothetical — we can see each one in tables and plots.

1. **Participant dependence** — same participants appear in multiple rows.
2. **Stimulus generalization** — product-level attractiveness dominates responses; stimuli are sampled, not exhaustive.
3. **Treatment-risk imbalance** — transparent trials disproportionately involve high-risk products and later trial positions.
4. **Nonlinear product valuation and fatigue** — risk and EV have threshold/diminishing effects; late-trial fatigue is non-trivial.
5. **Differential missingness** — missing outcomes are more common for transparent, high-risk, late-trial, low-literacy, and mobile trials.
6. **Post-treatment exclusion risk** — attention failures are partially caused by the treatment; excluding them narrows the sample in a non-neutral way.
7. **Treatment-effect heterogeneity** — the transparency effect varies across participants and by financial literacy.

Each section below addresses one or more of these. We keep a running count.
:::

```{r threat-dashboard}
tibble(
  `#` = 1:7,
  Threat = c(
    "Participant dependence",
    "Stimulus generalization",
    "Treatment–risk imbalance",
    "Nonlinear valuation/fatigue",
    "Differential missingness",
    "Post-treatment exclusions",
    "Effect heterogeneity"
  ),
  `Diagnostic evidence` = c(
    "Same participant rated 12 products (ICC ≈ large)",
    "Product attractiveness SD > transparency effect",
    "Transparent trials have higher mean risk; more common late/at site 2",
    "Residuals curve against risk, EV, and trial order",
    "Missing rate: transparent > non-transparent; late > early; mobile > non-mobile",
    "Attention failure rate: transparent > non-transparent",
    "Participant-level effect distribution is wide"
  ),
  Severity = c("High", "High", "High", "Medium–High", "High", "Medium", "Medium"),
  `First response` = c(
    "Participant random intercept",
    "Stimulus random intercept",
    "Add risk terms + check imbalance",
    "Nonlinear terms + trial order in model",
    "Missingness predictors / IPW sensitivity",
    "Report both including and excluding failures",
    "Random slope; literacy interaction"
  )
) |>
  knitr::kable(caption = "Initial threat dashboard — built from data diagnostics before any model is fit")
```

---

## 6. Diagnose the Threats: Empirical Evidence

We now examine each threat with actual data. These diagnostics should happen before the first model, not after.

### 6.1 Assignment imbalance

```{r assignment-audit}
# Check whether transparency correlates with pre-treatment variables
tx_audit <- glm(
  transparency ~ risk_level + expected_value_z + financial_literacy +
    factor(site_id) + trial_order_z,
  data   = trials |> mutate(trial_order_z = (trial_order - 6.5) / 3.452),
  family = binomial()
)

broom::tidy(tx_audit) |>
  filter(term != "(Intercept)") |>
  mutate(
    `Magnitude` = case_when(
      abs(estimate) > 0.30 ~ "Large — imbalance present",
      abs(estimate) > 0.10 ~ "Moderate",
      TRUE                  ~ "Small"
    ),
    across(c(estimate, std.error, p.value), \(x) round(x, 3))
  ) |>
  select(term, estimate, std.error, p.value, Magnitude) |>
  knitr::kable(caption = paste0(
    "Assignment audit: logistic regression predicting transparency from pre-treatment variables.",
    " Coefficients near 0 = clean assignment; large coefficients = implementation imbalance."))
```

**What we see:** Risk level and trial order strongly predict assignment. Site 2 shows elevated transparency. This is *not* a minor issue: the risk imbalance will pull the naive transparency estimate downward because transparent trials tend to involve products that are harder to evaluate (higher risk → lower adoption).

### 6.2 Missingness by condition and design variables

```{r missingness-audit}
cat("Overall missing rate:", round(100 * mean(trials$missing_y), 1), "%\n\n")

# By transparency
miss_by_tx <- trials |>
  group_by(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  summarise(n_trials = n(), n_missing = sum(missing_y),
            miss_rate = round(100 * mean(missing_y), 1))
knitr::kable(miss_by_tx, caption = "Missingness by condition — differential dropout is visible")

# By trial order
miss_by_order <- trials |>
  group_by(trial_order) |>
  summarise(miss_rate = round(100 * mean(missing_y), 1))
knitr::kable(miss_by_order, caption = "Missingness by trial order — late-trial dropout")

# By financial literacy tertile
trials |>
  mutate(lit_tertile = ntile(financial_literacy, 3)) |>
  group_by(lit_tertile) |>
  summarise(miss_rate = round(100 * mean(missing_y), 1), n = n()) |>
  knitr::kable(caption = "Missingness by literacy tertile")
```

```{r attention-by-condition}
trials |>
  group_by(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  summarise(
    n = n(),
    n_fail = sum(attention_fail),
    fail_rate = round(100 * mean(attention_fail), 1)
  ) |>
  knitr::kable(caption = "Attention failure rate by condition — failures are post-treatment and condition-dependent")
```

### 6.3 Data retention

```{r retention-table}
n_intended <- nrow(trials)
n_observed <- sum(!is.na(trials$adoption_intent_obs))
n_attfail  <- sum(trials$attention_fail == 1)
n_analytic <- sum(!is.na(trials$adoption_intent_obs) & trials$attention_fail == 0)

tibble(
  Stage = c("Intended (all planned trials)",
            "Observed (non-missing outcome)",
            "Attention failures (in all trials)",
            "Clean analytic sample (no miss, no attn fail)"),
  n   = c(n_intended, n_observed, n_attfail, n_analytic),
  pct = round(100 * c(n_intended, n_observed, n_attfail, n_analytic) / n_intended, 1)
) |>
  knitr::kable(caption = "Data flow: how many rows are available at each stage?")
```

### 6.4 Outcome overview

```{r outcome-checks, fig.width=9, fig.height=3.5}
p1 <- ggplot(trials, aes(x = adoption_intent_obs)) +
  geom_histogram(bins = 30, fill = "#1e3a5f", color = "white", na.rm = TRUE) +
  labs(title = "Observed outcome distribution", x = "Adoption intent (0–100)") +
  theme_minimal(base_size = 10)

p2 <- trials |>
  filter(!is.na(adoption_intent_obs)) |>
  mutate(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  ggplot(aes(x = Condition, y = adoption_intent_obs, fill = Condition)) +
  geom_boxplot(show.legend = FALSE, width = 0.5) +
  scale_fill_manual(values = c("#cccccc", "#1e3a5f")) +
  labs(title = "By condition", y = "Adoption intent", x = NULL) +
  theme_minimal(base_size = 10)

p3 <- trials |>
  filter(!is.na(adoption_intent_obs)) |>
  ggplot(aes(x = trial_order, y = adoption_intent_obs, group = trial_order)) +
  geom_boxplot(fill = "#2d6a4f", alpha = 0.6, outlier.size = 0.5) +
  labs(title = "By trial order — fatigue", y = NULL, x = "Trial") +
  theme_minimal(base_size = 10)

gridExtra::grid.arrange(p1, p2, p3, ncol = 3)

cat("Proportion at floor (≤2):", round(mean(trials$adoption_intent_obs <= 2, na.rm=TRUE), 3), "\n")
cat("Proportion at ceiling (≥98):", round(mean(trials$adoption_intent_obs >= 98, na.rm=TRUE), 3), "\n")
```

### 6.5 Post-treatment variable timing

Not every variable belongs in the main model. Including post-treatment variables changes the estimand.

```{r timing-table}
tibble(
  Variable = c("Risk level", "Expected value", "Financial literacy",
               "Trial order", "Device type",
               "Attention check", "Perceived transparency", "Comprehension"),
  Timing   = c("Pre-treatment stimulus", "Pre-treatment stimulus",
               "Pre-treatment participant",
               "Design/process variable", "Pre-treatment contextual",
               "Post-trial — ambiguous", "Post-treatment mediator",
               "Post-treatment mediator"),
  `Safe in main model?` = c("Yes", "Yes", "Yes", "Yes — controls fatigue confound",
                             "Yes — controls missingness", "Caution",
                             "No — changes estimand", "No — changes estimand"),
  Rationale = c(
    "Stimulus attribute not affected by the display",
    "Stimulus attribute not affected by the display",
    "Measured before any trial",
    "Controls trial-order fatigue; removes a source of assignment imbalance",
    "Predicts missingness; adding it makes the estimand closer to the intended sample",
    "Failure may reflect difficulty of processing the transparent display",
    "Blocks the mechanism; estimates only non-comprehension pathway",
    "Same"
  )
) |>
  knitr::kable(caption = "Post-treatment timing audit: what enters the main model?")
```

::: {.callout-warning icon=false}
## Why Comprehension Must Stay Out of the Model

```{mermaid}
flowchart LR
  T["Transparency\n(assigned)"] --> C["Comprehension /\nclarity"]
  C --> Y["Adoption intent"]
  T --> Y
  L["Literacy"] --> C
  L --> Y
```

Controlling for comprehension estimates only the direct transparency effect that does not work through understanding fees — a narrower and different claim. The main model should not include comprehension or perceived transparency.
:::

---

## 7. Start With the Wrong Model

```{r m0-naive}
m0 <- lm(adoption_intent_obs ~ transparency, data = trials)

cat("Rows in dataset:", nrow(trials), "\n")
cat("Rows used (non-missing):", nobs(m0), "\n")
cat("Rows silently dropped:", nrow(trials) - nobs(m0), "\n\n")

broom::tidy(m0) |>
  filter(term == "transparency") |>
  mutate(model = "m0: pooled OLS",
         `True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "Naive pooled OLS — estimate is substantially below the true effect")
```

```{r m0-diagnosis}
tibble(
  Problem = c("Silent row dropping", "Independence assumption",
              "Treatment-risk imbalance", "Participant baselines in residual",
              "Stimulus heterogeneity in residual", "Nonlinear covariates ignored",
              "Trial-order fatigue ignored", "Informative missingness ignored"),
  `What m0 does` = c(
    "Drops missing rows without warning",
    paste0("Treats ~", round(nobs(m0)/1000, 1), "k rows as independent"),
    "Transparent trials have higher mean risk → transparency gets blamed for risk penalty",
    "Person-to-person differences dilute within-person signal",
    "Product attractiveness absorbed into residual; no generalization",
    "Risk and EV assumed linear — misspecified functional form",
    "No adjustment; transparent trials are later on average",
    "Complete cases treated as a random subsample"
  ),
  `Direction of bias` = c(
    "Unknown — depends on missingness structure",
    "SE is wrong",
    "Downward — transparency confounded with high-risk products",
    "SE inflated",
    "SE inflated; no generalization claim",
    "Partial confound with transparency estimate",
    "Downward — transparency confounded with late-trial position",
    "Upward — missing transparent/high-risk/late trials excluded"
  )
) |>
  knitr::kable(caption = "Why m0 is not credible — a problem inventory")
```

**Student question:** The estimate is significant. Why isn't it believable?

**Advisor response:** Because we know exactly what this model confounds. Transparent trials in this dataset are disproportionately high-risk products (from the assignment audit) and disproportionately late in the trial sequence (from the trial-order plot). Both risk and lateness reduce adoption intent. The naive model attributes all of that to the transparency condition. The estimate reflects "what happened to be true in this particular realized sample" — not the causal effect of fee transparency.

::: {.callout-important icon=false}
## Open Threats After m0: 7 of 7

All seven threats are still active. m0 does not address any of them. The estimate is substantially below the true effect due to treatment-risk imbalance and fatigue confounding.

**Still open:** participant dependence, stimulus generalization, treatment–risk imbalance, nonlinear valuation/fatigue, differential missingness, post-treatment exclusions, effect heterogeneity.
:::

---

## 8. Add Participant Random Intercepts

**Student question:** Can I fix the bias by adding participant random intercepts? They should account for individual differences in adoption tendency.

**Advisor response:** Participant random intercepts improve the model in one specific way: they remove stable person-level adoption baselines from the residual, which sharpens the within-person estimate of transparency. But they do not fix the treatment-risk imbalance. That imbalance is at the *stimulus* level — transparent trials go to riskier products regardless of who the participant is. So the estimate barely moves; what changes is the SE.

```{r m1}
m1 <- lmer(adoption_intent_obs ~ transparency + (1 | participant_id), data = trials)

vc1 <- as.data.frame(VarCorr(m1))
icc_part <- with(vc1, vcov[grp == "participant_id"] /
                   (vcov[grp == "participant_id"] + vcov[grp == "Residual"]))
cat("Participant ICC:", round(icc_part, 3), "— substantial person-level clustering\n")

broom.mixed::tidy(m1, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m1: + participant RI", `True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "m1 — estimate unchanged, as expected: participant RI fixes dependence, not imbalance")
```

::: {.callout-important icon=false}
## Open Threats After m1: 6 of 7

**Reduced:** ~~Participant dependence (rows not independent)~~

**Still open:** stimulus generalization, treatment–risk imbalance, nonlinear valuation/fatigue, differential missingness, post-treatment exclusions, effect heterogeneity.

The estimate did not move because participant RI does not address the treatment-risk imbalance. That will require covariate adjustment.
:::

---

## 9. Add Stimulus Random Intercepts

```{r m2}
m2 <- lmer(adoption_intent_obs ~ transparency +
             (1 | participant_id) + (1 | stimulus_id),
           data = trials)
```

```{r m2-summary-and-plot, fig.width=7, fig.height=4}
vc2 <- as.data.frame(VarCorr(m2))
var_stim2 <- vc2$vcov[vc2$grp == "stimulus_id"]
cat("Stimulus SD:", round(sqrt(var_stim2), 2),
    "(true =", dgp$sd_stimulus, "; transparency effect is", dgp$effect_transparency, ")\n")
cat("Stimulus SD > transparency effect: product attractiveness dominates responses.\n")

broom.mixed::tidy(m2, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m2: + stimulus RI", `True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "m2 — generalization claim changed; estimate still biased")

# Caterpillar plot
ranef(m2)$stimulus_id |>
  as.data.frame() |>
  rownames_to_column("id") |>
  rename(blup = `(Intercept)`) |>
  ggplot(aes(x = reorder(id, blup), y = blup)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#888888") +
  geom_segment(aes(xend = id, y = 0, yend = blup),
               color = "#1e3a5f", linewidth = 0.5, alpha = 0.7) +
  geom_point(color = "#1e3a5f", size = 1.8) +
  coord_flip() +
  labs(title = paste0("Stimulus BLUPs: product-level SD ≈ ",
                      round(sqrt(var_stim2), 1),
                      " — larger than the transparency effect"),
       subtitle = "Products vary enormously in baseline attractiveness. This must be modeled.",
       x = NULL, y = "Estimated baseline deviation (BLUP)") +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

**What changed:** Not the point estimate — still biased. What changed is the *generalization claim*. Before m2, the result was specific to these 36 products. After m2, it is a claim about the population of products like these. That is a meaningfully different scientific assertion.

**What is still wrong:** Stimulus random intercepts tell us about product baseline variance, but they do not remove the treatment-risk imbalance. The transparency effect is still confounded with product risk.

::: {.callout-important icon=false}
## Open Threats After m2: 5 of 7

**Reduced:** ~~Participant dependence~~, ~~Stimulus generalization~~

**Still open:** treatment–risk imbalance, nonlinear valuation/fatigue, differential missingness, post-treatment exclusions, effect heterogeneity.

The estimate has not moved. The crossing of participants and stimuli is now correctly handled, but the confounding from risk imbalance is still in the transparency coefficient.
:::

---

## 10. Add Site Structure

```{r m3}
m3 <- lmer(adoption_intent_obs ~ transparency +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)
```

```{r m3-summary}
vc3 <- as.data.frame(VarCorr(m3))
cat("Site SD:", round(sqrt(vc3$vcov[vc3$grp == "site_id"]), 2),
    "(true =", dgp$sd_site, ")\n")

trials |>
  filter(!is.na(adoption_intent_obs)) |>
  group_by(site_id) |>
  summarise(mean_adoption = round(mean(adoption_intent_obs, na.rm = TRUE), 1),
            pct_transparent = round(100 * mean(transparency), 1),
            n = n_distinct(participant_id)) |>
  knitr::kable(caption = "Site-level summaries: site 2 has lower baseline AND higher transparency rate — a confound")

broom.mixed::tidy(m3, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m3: + site RI", `True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "m3 — small upward movement as site 2's confounding is partially removed")
```

**Student question:** The estimate moved up a little. Is that from the site structure?

**Advisor response:** Yes. Site 2 has a below-average adoption baseline AND an above-average transparency rate (from the assignment imbalance). In m2, those site-level differences were in the residual, pulling the transparent-condition mean down. Adding a site random intercept partitions out that baseline difference, removing a small part of the downward bias. But the larger bias — from risk level being positively correlated with transparency assignment — is still entirely unaddressed. That requires adding risk as a covariate.

::: {.callout-important icon=false}
## Open Threats After m3: 4 of 7

**Reduced:** ~~Participant dependence~~, ~~Stimulus generalization~~, ~~Site confounding (partially)~~

**Still open:** treatment–risk imbalance, nonlinear valuation/fatigue, differential missingness, post-treatment exclusions, effect heterogeneity.

The estimate moved modestly. The main remaining bias is from treatment-risk imbalance.
:::

---

## 11. Add Pre-Treatment Covariates

**Student question:** Now I add risk level, expected value, and financial literacy as covariates. Should I expect the estimate to change?

**Advisor response:** Yes — substantially. The assignment audit showed that risk level strongly predicts transparency (coefficient ~0.70). Risk level also reduces adoption intent. Covariate adjustment for risk should remove most of that downward bias. Watch the estimate move.

```{r m4}
m4 <- lmer(adoption_intent_obs ~ transparency + risk_level + expected_value_z +
             financial_literacy +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)

broom.mixed::tidy(m4, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m4: + linear risk/EV/literacy", `True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  knitr::kable(digits = 3, caption = "m4 — estimate jumps substantially upward after risk adjustment")
```

**Student question:** The estimate moved up by more than a point. What happened?

**Advisor response:** Risk adjustment happened. Before, the model was comparing high-risk/transparent products to lower-risk/non-transparent products. Now it's comparing within risk level — transparent vs non-transparent for products at the same risk. The upward movement is the risk-imbalance bias being removed. The estimate is still not at 4.0 because: (a) we haven't adjusted for the nonlinear risk threshold (risk > 0.75 has an extra penalty the linear term doesn't capture), and (b) we haven't controlled for trial-order fatigue. Both of those continue to bias transparency downward.

::: {.callout-warning icon=false}
## Open Threats After m4: 3 of 7

**Reduced:** ~~Participant dependence~~, ~~Stimulus generalization~~, ~~Site confounding~~, ~~Linear risk/covariate imbalance~~

**Still open:** nonlinear valuation/fatigue (estimate still below truth), differential missingness, post-treatment exclusions, effect heterogeneity.

**This is the pedagogical moment:** The estimate moved substantially when risk was added, not before. Participant and stimulus random intercepts fixed the data structure. Covariate adjustment fixed the bias. These are different problems requiring different solutions.
:::

---

## 12. Functional-Form and Fatigue Check

We added a linear risk term. But the DGP includes a threshold penalty for high-risk products (risk_level > 0.75). A linear term partially adjusts for this, but it is misspecified. Residual plots should show this.

```{r functional-form-diag, fig.width=9, fig.height=3.5}
resid_m4 <- tibble(
  residual    = residuals(m4),
  risk_level  = model.frame(m4)$risk_level,
  ev_z        = model.frame(m4)$expected_value_z,
  trial_order = trials$trial_order[!is.na(trials$adoption_intent_obs)]
)

p_r <- ggplot(resid_m4, aes(x = risk_level, y = residual)) +
  geom_point(alpha = 0.08, color = "#1e3a5f") +
  geom_smooth(method = "loess", se = FALSE, color = "#e63946", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs risk", subtitle = "Curve at high risk = threshold effect", x = "Risk level", y = "Residual") +
  theme_minimal(base_size = 10)

p_e <- ggplot(resid_m4, aes(x = ev_z, y = residual)) +
  geom_point(alpha = 0.08, color = "#1e3a5f") +
  geom_smooth(method = "loess", se = FALSE, color = "#e63946", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs EV (std.)", subtitle = "Curve = diminishing returns", x = "Expected value (z)", y = "Residual") +
  theme_minimal(base_size = 10)

p_t <- ggplot(resid_m4, aes(x = trial_order, y = residual, group = trial_order)) +
  geom_boxplot(fill = "#1e3a5f", alpha = 0.5, outlier.size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs trial order", subtitle = "Downward trend = unmodeled fatigue", x = "Trial", y = "Residual") +
  theme_minimal(base_size = 10)

gridExtra::grid.arrange(p_r, p_e, p_t, ncol = 3)
```

**What we see:** The loess curve over risk shows a downward dip at high risk that the linear term does not capture. The trial-order residual plot shows a downward trend — later trials have systematically negative residuals, meaning the model over-predicts their adoption intent. We need both a risk threshold indicator and a trial-order term.

```{r m4b}
m4b <- lmer(adoption_intent_obs ~ transparency +
              risk_level + risk_high +
              expected_value_z + I(expected_value_z^2) +
              financial_literacy + trial_order +
              (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
            data = trials)

bind_rows(
  broom.mixed::tidy(m4,  effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m4: linear risk/EV, no trial_order"),
  broom.mixed::tidy(m4b, effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m4b: + risk_high + I(ev_z^2) + trial_order")
) |>
  mutate(`True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = paste0("m4b: estimate continues moving toward the true effect of ",
    dgp$effect_transparency, " as the remaining confounds are removed"))
```

**What happened:** Adding the risk threshold indicator removed the residual confounding from high-risk products getting transparent displays disproportionately. Adding trial order removed the fatigue confounding from transparent trials occurring later in the task. The estimate moved again — now close to the true effect.

::: {.callout-warning icon=false}
## Open Threats After m4b: 2 of 7 (plus 1 pending)

**Reduced:** ~~Participant dependence~~, ~~Stimulus generalization~~, ~~Site confounding~~, ~~Risk/covariate imbalance~~, ~~Nonlinear valuation/fatigue~~

**Still open:** differential missingness, post-treatment exclusions (pending), effect heterogeneity.

We are now close to the truth, but the remaining concerns are about *which data were observed* (missingness) and *who the analysis is for* (attention exclusions, heterogeneity).
:::

---

## 13. Missingness Sensitivity

**Student question:** We've been running complete-case models. Is that OK?

**Advisor response:** The missingness audit showed that transparent trials are more likely to go missing (about `r round(100 * mean(trials$missing_y[trials$transparency == 1]), 0)`% vs `r round(100 * mean(trials$missing_y[trials$transparency == 0]), 0)`% for non-transparent). The m4b model already controls for most missingness predictors (risk, trial order, literacy) in the outcome model. But one predictor — `device_mobile` — is not yet in the model. Device is correlated with missingness (coefficient 0.35) AND with adoption intent (the true DGP has a -1.5 effect of mobile). Not adjusting for it means complete-case analysis excludes a non-random set of mobile/high-missingness trials.

**Diagnostic move:** Compare the m4b estimate to a model that adds `device_mobile` and then to a model that uses only non-missing complete cases after filtering.

```{r m4c}
m4c <- lmer(
  adoption_intent_obs ~ transparency +
    risk_level + risk_high +
    expected_value_z + I(expected_value_z^2) +
    financial_literacy + trial_order + device_mobile +
    (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
  data = trials
)

bind_rows(
  broom.mixed::tidy(m4b, effects = "fixed") |>
    filter(term == "transparency") |>
    mutate(model = "m4b: complete-case, without device_mobile",
           n_obs = nobs(m4b)),
  broom.mixed::tidy(m4c, effects = "fixed") |>
    filter(term == "transparency") |>
    mutate(model = "m4c: + device_mobile (missingness/DGP predictor)",
           n_obs = nobs(m4c))
) |>
  mutate(`True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, n_obs, estimate, std.error, p.value) |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = paste0(
    "Missingness sensitivity: the shift from adding device_mobile tells us whether ",
    "device-related differential dropout was influencing the estimate. ",
    "True effect = ", dgp$effect_transparency))
```

```{r miss-sensitivity-commentary}
est_m4b  <- round(fixef(m4b)["transparency"], 3)
est_m4c  <- round(fixef(m4c)["transparency"], 3)
est_diff <- round(abs(est_m4c - est_m4b), 3)
```

**What we learned:** The estimate shifted by `r est_diff` points when device was added. If the shift is small, the device-related missingness was not substantially biasing the complete-case estimate. If it is larger, device-correlated selection was influencing results.

::: {.callout-warning icon=false}
## Limits of This Sensitivity Analysis

Adding observed missingness predictors adjusts for **MAR** (missing at random conditional on covariates). It does **not** address **MNAR** (missing not at random): if participants with low latent adoption intent are the ones most likely to skip transparent/high-risk trials — over and above what their literacy, trial order, and device predict — no covariate adjustment reaches it.

The honest framing: "After adding missingness-related covariates, the estimate changed by `r est_diff` points. The remaining concern is that missingness may depend on latent adoption intent — a bias direction we cannot test from the observed data."
:::

::: {.callout-warning icon=false}
## Open Threats After m4c: 1 of 7 (plus 1 pending)

**Reduced:** ~~Participant dependence~~, ~~Stimulus generalization~~, ~~Site confounding~~, ~~Risk/covariate imbalance~~, ~~Nonlinear valuation/fatigue~~, ~~Observed differential missingness~~

**Still open:** post-treatment exclusions (now we check), effect heterogeneity (m5, m6).

The treatment estimate is now near the true DGP value. The remaining work is to check whether *who* we included in the analysis changes the claim.
:::

---

## 14. Attention Exclusion Sensitivity

**Student question:** Standard practice is to exclude participants who fail attention checks. Shouldn't I do that?

**Advisor response:** Not automatically. Look at the numbers. Attention failures are more common in the transparent condition (`r round(100 * mean(trials$attention_fail[trials$transparency == 1]), 1)`% vs `r round(100 * mean(trials$attention_fail[trials$transparency == 0]), 1)`% for non-transparent). If we exclude attention-fail trials, we are removing a set of transparent trials that was especially likely to fail — perhaps because the fee-transparency display was complex or cognitively demanding. Excluding them conditions on a post-treatment variable and narrows the sample to "trials where the participant processed the display without difficulty." That is a different population than "all participants assigned to the transparent display."

The right move is to show *both* estimates and discuss what the difference means for the estimand.

```{r m4d}
# Model on all observed (non-missing) trials — our main analytic sample
m_all_obs <- m4c  # already fitted above

# Model excluding attention failures from observed trials
trials_no_fail <- trials |> filter(attention_fail == 0)
m4d <- lmer(
  adoption_intent_obs ~ transparency +
    risk_level + risk_high +
    expected_value_z + I(expected_value_z^2) +
    financial_literacy + trial_order + device_mobile +
    (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
  data = trials_no_fail
)

bind_rows(
  broom.mixed::tidy(m_all_obs, effects = "fixed") |>
    filter(term == "transparency") |>
    mutate(model = "m4c: all observed rows (incl. attention failures)",
           n_obs = nobs(m_all_obs)),
  broom.mixed::tidy(m4d, effects = "fixed") |>
    filter(term == "transparency") |>
    mutate(model = "m4d: excluding attention failures",
           n_obs = nobs(m4d))
) |>
  mutate(`True effect` = dgp$effect_transparency,
         `Estimand change` = c("Intention-to-treat (all assigned)", "Complier-like (easy-to-process trials)")) |>
  select(model, `Estimand change`, `True effect`, n_obs, estimate, std.error, p.value) |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = paste0(
    "Attention exclusion sensitivity: does excluding post-treatment attention failures ",
    "change the estimate or narrow the estimand? True effect = ", dgp$effect_transparency))
```

```{r attention-exclusion-notes}
est_all  <- round(fixef(m_all_obs)["transparency"], 3)
est_excl <- round(fixef(m4d)["transparency"], 3)
diff_attn <- round(est_excl - est_all, 3)
```

**What we see:** Excluding attention failures `r ifelse(diff_attn > 0, "inflates", "deflates")` the transparency estimate by `r abs(diff_attn)` points. The direction is consistent with post-treatment selection: we removed trials where the transparent display was most likely to overwhelm the participant. Those trials disproportionately came from the transparent condition and may have had lower adoption intent. Removing them selects toward participants/trials where transparency was most easily processed — which may be exactly the population for whom it is most beneficial.

**Recommended practice:** Report m4c (all observed) as the primary estimate. Report m4d (excluding attention failures) as a sensitivity check and flag the estimand difference explicitly.

---

## 15. Add the Theoretically Motivated Interaction

```{r m5}
m5 <- lmer(adoption_intent_obs ~ transparency * financial_literacy +
             risk_level + risk_high +
             expected_value_z + I(expected_value_z^2) +
             trial_order + device_mobile +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)

broom.mixed::tidy(m5, effects = "fixed") |>
  select(term, estimate, std.error, statistic, p.value) |>
  mutate(across(c(estimate, std.error, statistic, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = "m5: transparency × financial literacy — the claim changes from 'average effect' to 'effect for whom'")
```

```{r m5-simple-slopes}
b_tx  <- fixef(m5)["transparency"]
b_int <- fixef(m5)["transparency:financial_literacy"]
tibble(
  `Financial literacy` = c("Low (−1 SD)", "Average (0)", "High (+1 SD)"),
  `Transparency effect` = round(c(b_tx + b_int * -1, b_tx, b_tx + b_int * 1), 2),
  `True DGP`           = c(dgp$effect_transparency + dgp$interaction_tx_lit * -1,
                             dgp$effect_transparency,
                             dgp$effect_transparency + dgp$interaction_tx_lit * 1)
) |>
  knitr::kable(caption = "Simple slopes: transparency effect at three literacy levels")
```

```{r m5-plot, fig.width=6.5, fig.height=4}
lit_vals <- seq(-2, 2, length.out = 100)
pred_df  <- tibble(
  financial_literacy = rep(lit_vals, 2),
  transparency       = rep(c(0L, 1L), each = 100),
  risk_level = 0, risk_high = 0L, expected_value_z = 0,
  trial_order = 6, device_mobile = 0L,
  participant_id = NA_integer_, site_id = NA_integer_, stimulus_id = NA_integer_
)
pred_df$predicted <- predict(m5, newdata = pred_df, re.form = NA)

pred_df |>
  mutate(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  ggplot(aes(x = financial_literacy, y = predicted,
             color = Condition, linetype = Condition)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("#cccccc", "#1e3a5f")) +
  labs(title = "Transparency effect moderated by financial literacy",
       subtitle = "Larger benefit for lower-literacy participants",
       x = "Financial literacy (standardised)", y = "Predicted adoption intent",
       color = NULL, linetype = NULL) +
  theme_minimal(base_size = 11)
```

::: {.callout-note icon=false}
## m5 Changes the Substantive Claim

Before m5, the analysis answered: "Is there a transparency effect, and is it credible?"

m5 answers: "For whom does the effect hold?" After m5, the transparency coefficient is the effect at *average* financial literacy — not a universal constant. The chapter's headline claim changes from "transparency increases adoption by X points" to "transparency benefits lower-literacy participants more, with an average effect of X at average literacy."

This is a different scientific claim. It should be stated explicitly in the Discussion and should inform policy targeting.
:::

---

## 16. Consider Random Slopes

```{r m6}
m6 <- lmer(adoption_intent_obs ~ transparency * financial_literacy +
             risk_level + risk_high +
             expected_value_z + I(expected_value_z^2) +
             trial_order + device_mobile +
             (1 + transparency | participant_id) +
             (1 | stimulus_id) + (1 | site_id),
           data = trials,
           control = lmerControl(optimizer = "bobyqa"))
cat("Singular (correlated slope):", isSingular(m6), "\n")
```

```{r m6b}
m6b <- lmer(adoption_intent_obs ~ transparency * financial_literacy +
              risk_level + risk_high +
              expected_value_z + I(expected_value_z^2) +
              trial_order + device_mobile +
              (1 + transparency || participant_id) +
              (1 | stimulus_id) + (1 | site_id),
            data = trials,
            control = lmerControl(optimizer = "bobyqa"))
cat("Singular (uncorrelated slope):", isSingular(m6b), "\n")
```

```{r m6-slope-viz, fig.width=6.5, fig.height=3.5}
emp_slopes <- trials |>
  filter(!is.na(adoption_intent_obs)) |>
  group_by(participant_id) |>
  summarise(
    mean_tx  = mean(adoption_intent_obs[transparency == 1], na.rm = TRUE),
    mean_ntx = mean(adoption_intent_obs[transparency == 0], na.rm = TRUE),
    n_tx = sum(transparency == 1), n_ntx = sum(transparency == 0)
  ) |>
  filter(n_tx > 0, n_ntx > 0) |>
  mutate(emp_effect = mean_tx - mean_ntx)

vc6b <- as.data.frame(VarCorr(m6b))
slope_row <- vc6b[!is.na(vc6b$var1) & vc6b$var1 == "transparency" & is.na(vc6b$var2), ]
slope_sd_val <- if (nrow(slope_row) > 0) sqrt(slope_row$vcov[1]) else NA_real_

ggplot(emp_slopes, aes(x = emp_effect)) +
  geom_histogram(bins = 25, fill = "#1e3a5f", color = "white", alpha = 0.8) +
  geom_vline(xintercept = fixef(m6b)["transparency"],
             color = "#e63946", linetype = "dashed", linewidth = 1) +
  labs(title = "Participant-level transparency effects: the spread is real",
       subtitle = paste0("Red = model average. Slope SD ≈ ",
                         ifelse(is.na(slope_sd_val), "NA", round(slope_sd_val, 1)),
                         " points. Some participants benefit much more than others."),
       x = "Empirical transparency effect (transparent − non-transparent)", y = "Count") +
  theme_minimal(base_size = 11)
```

```{r m6-comparison}
bind_rows(
  broom.mixed::tidy(m5,  effects = "fixed") |>
    filter(term %in% c("transparency", "transparency:financial_literacy")) |>
    mutate(model = "m5: fixed slope"),
  broom.mixed::tidy(m6b, effects = "fixed") |>
    filter(term %in% c("transparency", "transparency:financial_literacy")) |>
    mutate(model = "m6b: random slope (uncorrelated)")
) |>
  select(model, term, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) |>
  knitr::kable(caption = "Does allowing slope heterogeneity change fixed-effect estimates?")
```

::: {.callout-warning icon=false}
## If the Random Slope Is Singular

1. Remove the correlation parameter first (`||`). This resolves most boundary issues.
2. If still singular, slope variance may be genuinely negligible: report the attempt and retain m5.
3. Never silently drop the random slope without documenting the decision.
:::

---

## 17. Estimator and Inference Decisions

::: {.callout-note icon=false}
## Key Estimator Choices

| Decision | Default | When to change |
|---|---|---|
| REML vs ML | REML for final model | ML (REML=FALSE) to compare fixed-effect structures |
| df method | Satterthwaite (lmerTest) | Kenward-Roger for small/unbalanced samples |
| Optimizer | nloptwrap (lme4 default) | `bobyqa` for convergence warnings |
| LRT for random slopes | Boundary-aware one-sided test | Parametric bootstrap if LRT misbehaves |
:::

---

## 18. Select and Diagnose the Final Model

```{r final-model-selection}
slope_sd_check <- tryCatch({
  vc <- as.data.frame(VarCorr(m6b))
  tx <- vc[!is.na(vc$var1) & vc$var1 == "transparency" & is.na(vc$var2), ]
  if (nrow(tx) > 0) sqrt(tx$vcov[1]) else NA_real_
}, error = function(e) NA_real_)

keep_slope <- isTRUE(!isSingular(m6b)) &&
              isTRUE(!is.na(slope_sd_check)) &&
              isTRUE(slope_sd_check > 0.5)

m_final <- if (keep_slope) m6b else m5

cat("Slope SD:", if (is.na(slope_sd_check)) "NA" else round(slope_sd_check, 2), "\n")
cat("Singular (m6b):", isSingular(m6b), "\n")
cat("Final model selected:", if (keep_slope) "m6b (random slope)" else "m5 (no slope)", "\n")
```

```{r residual-diagnostics, fig.width=10, fig.height=3.5}
# Align residuals to observed-only rows in trials
trials_obs <- trials |> filter(!is.na(adoption_intent_obs))

resid_df <- tibble(
  fitted   = fitted(m_final),
  residual = residuals(m_final),
  site_id  = trials_obs$site_id,
  risk     = trials_obs$risk_level
)

p_rv <- ggplot(resid_df, aes(x = fitted, y = residual)) +
  geom_point(alpha = 0.12, color = "#1e3a5f") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#e63946") +
  labs(title = "Residuals vs Fitted", x = "Fitted", y = "Residual") +
  theme_minimal(base_size = 10)

p_qq <- ggplot(resid_df, aes(sample = residual)) +
  stat_qq(alpha = 0.15, color = "#1e3a5f") +
  stat_qq_line(color = "#e63946") +
  labs(title = "QQ Plot", x = "Theoretical", y = "Sample") +
  theme_minimal(base_size = 10)

p_risk <- ggplot(resid_df, aes(x = risk, y = residual)) +
  geom_point(alpha = 0.08, color = "#1e3a5f") +
  geom_smooth(method = "loess", se = FALSE, color = "#e63946", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Risk (heteroskedasticity check)",
       x = "Risk level", y = "Residual") +
  theme_minimal(base_size = 10)

gridExtra::grid.arrange(p_rv, p_qq, p_risk, ncol = 3)
```

::: {.callout-note icon=false}
## A Note on Heteroskedasticity

The residual fan visible in the risk plot reflects genuine heteroskedasticity: the DGP has higher residual variance for high-risk products. The LMM assumes constant residual variance. For the transparency main effect, this is unlikely to substantially bias estimates, but it does mean confidence intervals may be too narrow for high-risk stimuli. A GLMM or heteroskedastic LMM (via glmmTMB) would address this if it is judged important.
:::

---

## 19. Model Comparison: What Moved and Why

```{r movement-table}
tibble(
  Step         = c("m0 → m1", "m1 → m2", "m2 → m3", "m3 → m4",
                   "m4 → m4b", "m4b → m4c", "m4c → m4d", "m4d → m5"),
  `Threat addressed` = c(
    "Participant dependence",
    "Stimulus generalization",
    "Site baseline confounding",
    "Treatment–risk imbalance (linear risk/EV)",
    "Nonlinear risk threshold + trial-order fatigue",
    "Device-related differential missingness",
    "Post-treatment attention exclusion check",
    "Treatment-effect heterogeneity by literacy"
  ),
  `Expected estimate movement` = c(
    "Little — dependence ≠ bias in within-person design",
    "Little — stimulus RI changes generalization claim, not point estimate",
    "Small upward — site 2 confound removed",
    "**Large upward** — main source of downward bias",
    "**Another upward movement** — residual confounds removed",
    "Small — tests device-driven differential dropout",
    "Can go either way — post-treatment selection",
    "Changes interpretation, not just point estimate"
  ),
  `Teaching point` = c(
    "RE fixes structure; covariate adjustment fixes bias",
    "Generalization requires modeling sampling; estimates change later",
    "Sites are contexts, not just noise; imbalanced transparency rates matter",
    "This is the diagnostic payoff: the naive estimate was wrong because of this",
    "Functional form matters when covariates are both nonlinear and imbalanced",
    "Observed MAR sensitivity; MNAR caveat remains",
    "Exclusions change the estimand; report both",
    "Policy relevance: effect is not uniform across the literacy distribution"
  )
) |>
  knitr::kable(caption = "What moved, why, and what each step teaches")
```

```{r robustness-table}
extract_tx <- function(m, label, is_lm = FALSE) {
  if (is_lm) {
    broom::tidy(m) |> filter(term == "transparency") |>
      select(estimate, std.error) |>
      mutate(model = label, n_obs = nobs(m))
  } else {
    broom.mixed::tidy(m, effects = "fixed") |> filter(term == "transparency") |>
      select(estimate, std.error) |>
      mutate(model = label, n_obs = nobs(m))
  }
}

bind_rows(
  extract_tx(m0,        "m0: pooled OLS (biased baseline)",               is_lm = TRUE),
  extract_tx(m2,        "m2: crossed RE (structure fixed, still biased)"),
  extract_tx(m3,        "m3: + site RI (small site adjustment)"),
  extract_tx(m4,        "m4: + linear risk/EV (large upward jump)"),
  extract_tx(m4b,       "m4b: + nonlinear risk + trial_order (recovered)"),
  extract_tx(m4c,       "m4c: + device_mobile (missingness sensitivity)"),
  extract_tx(m4d,       "m4d: excluding attention failures (estimand check)"),
  extract_tx(m5,        "m5: + literacy interaction (moderation)")
) |>
  mutate(
    across(c(estimate, std.error), \(x) round(x, 3)),
    `True effect` = dgp$effect_transparency,
    `Remaining concerns` = c(
      "All seven threats open",
      "Risk imbalance, fatigue, missingness",
      "Risk imbalance, fatigue, missingness",
      "Nonlinear risk, fatigue, missingness",
      "Missingness, attention exclusions, heterogeneity",
      "MNAR; attention exclusions; heterogeneity",
      "Narrower estimand; MNAR",
      "MNAR; post-treatment selection; 4-site limit"
    )
  ) |>
  select(model, `True effect`, n_obs, estimate, std.error, `Remaining concerns`) |>
  knitr::kable(caption = paste0(
    "Full robustness table: estimate movement traces the diagnostic sequence. ",
    "True effect at average literacy = ", dgp$effect_transparency))
```

---

## 20. Final Interpretation

```{r final-interp}
fe_final <- broom.mixed::tidy(m_final, effects = "fixed") |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3)))
knitr::kable(fe_final |> select(term, estimate, std.error, statistic, p.value),
             caption = "Final model: fixed effects")

vc_final <- as.data.frame(VarCorr(m_final))
knitr::kable(vc_final |> select(grp, var1, vcov) |>
               mutate(SD = round(sqrt(vcov), 2), vcov = round(vcov, 2)),
             caption = "Final model: variance components")
```

**Transparency effect at average literacy:** `r round(fixef(m_final)["transparency"], 1)` points (true DGP: `r dgp$effect_transparency`). The naive model gave approximately `r round(coef(m0)["transparency"], 1)` — a difference of `r round(fixef(m_final)["transparency"] - coef(m0)["transparency"], 1)` points. That gap was earned by diagnosing and removing three sources of downward bias: treatment-risk imbalance, nonlinear risk threshold effects, and trial-order fatigue.

---

## 21. What We Fixed vs What We Did Not Fix

### What the analysis substantially reduced

- **Row dependence** — participant and stimulus random intercepts correctly separate within-person/within-product estimates from between-unit baselines.
- **Treatment-risk imbalance** — adding risk level (linear + threshold) removed the largest source of bias. The naive estimate was ~`r round(dgp$effect_transparency - coef(m0)["transparency"], 1)` points too low.
- **Nonlinear product valuation** — the quadratic EV term and risk_high indicator removed residual misfit from functional-form assumptions.
- **Trial-order fatigue** — adding trial_order as a fixed effect removed the downward confounding from transparent trials occurring later in the task.
- **Observed differential missingness** — adding device_mobile adjusted for one source of differential dropout.
- **Post-treatment exclusion ambiguity** — we reported both the intent-to-treat and the attention-excluded estimates and described the estimand difference.

### What the analysis did not fully solve

- **MNAR** — if participants with low latent adoption intent are disproportionately the ones skipping transparent high-risk late-trial observations, no covariate adjustment addresses it. The sensitivity test is informative but not conclusive.
- **Post-treatment selection** — excluding attention failures restricts the sample to trials processed without difficulty. That restriction is not neutral, and the "correct" choice depends on the intended estimand, not on data quality alone.
- **Limited site generalization** — with four sites, the site random intercept is estimated from very little information. Site heterogeneity in the transparency effect cannot be reliably detected at this design.
- **Heteroskedasticity** — the LMM's constant-variance assumption is violated for high-risk stimuli. For policy claims targeted at high-risk products, confidence intervals may be too narrow.
- **Unobserved product features** — risk level and expected value capture some product characteristics, but other attractiveness dimensions (brand familiarity, temporal framing, product complexity) are absorbed into the stimulus random effect and may not be adequately controlled.

### The honest final claim

> After diagnosing and removing the main observed threats to inference — treatment-risk imbalance, fatigue confounding, nonlinear covariate effects, and partial missingness adjustment — fee transparency appears to increase adoption intent by approximately `r round(fixef(m_final)["transparency"], 1)` points at average financial literacy (true DGP: `r dgp$effect_transparency`). The effect is larger for lower-literacy participants. This estimate is meaningfully more credible than the naive result, but it still depends on assumptions about whether missingness is MAR conditional on observed covariates, whether attention exclusions are neutral for the estimand, and whether the four-site sample represents the intended population of deployment contexts.

---

## 22. What a Reviewer Would Ask

**1. "The naive estimate was substantially below the true effect. Was the assignment actually randomized?"**

No — not in this dataset. The assignment audit showed that the fee-transparency display was more commonly served to high-risk products, later trial positions, and participants at site 2. This is a known implementation pattern in platform-based studies where display logic interacts with product metadata. The bias was removed by covariate adjustment, but a reviewer is right to ask this question.

---

**2. "Your main models are complete-case. How do you handle missing outcomes?"**

Missingness is non-trivial (~`r round(100 * mean(trials$missing_y), 0)`%) and differential: transparent trials have higher dropout. We documented this in the missingness audit and compared a complete-case model to one including the main missingness predictors. The shift was `r est_diff` points. The remaining concern is MNAR — if latent adoption intent predicts dropout over and above the observed covariates — which we cannot directly test.

---

**3. "You excluded attention-check failures. That's a post-treatment variable."**

We did not exclude them in the primary analysis. We ran both an all-observed estimate and an attention-excluded estimate and reported both. Attention failures were more common in the transparent condition, consistent with the display being more cognitively demanding. Excluding them increases the estimate, which is the expected direction if removed trials had lower adoption intent. We flag this as an estimand issue, not a data quality fix.

---

**4. "Why is expected value quadratic? That seems arbitrary."**

Diminishing sensitivity to expected value is psychologically motivated: people are not linearly sensitive to financial value at all magnitudes. We confirmed this choice with a residual diagnostic — the loess curve over expected value showed visible nonlinearity after the linear adjustment. Forcing linearity leaves systematic residual misfit.

---

**5. "Trial order is in your model. How do you know it's a fatigue effect and not a learning effect?"**

The residual plot after m4 shows that later trials have lower adoption intent on average, which is more consistent with fatigue than learning (learning would increase ratings over time). The fatigue interpretation is also supported by the DGP — but in a real dataset, we cannot distinguish cleanly. We would recommend noting this ambiguity and checking whether the trial-order effect differs by condition.

---

**6. "The literacy interaction is theoretically motivated. Did you pre-register it?"**

This worked example is a tutorial, not a report. In a real study, the interaction between financial literacy and fee transparency is a plausible and theoretically grounded moderation hypothesis. Pre-registration is the appropriate way to control family-wise error for exploratory moderation tests.

---

**7. "Four sites is a small sample for a random effect."**

Correct. With four sites, the site random-effect variance is estimated from very few data points. The site RI is still the right specification — it correctly represents that sites are a sample from a population of contexts — but the variance estimate is unreliable. A sensitivity analysis treating site as fixed (using `factor(site_id)`) would show whether site-level differences substantially change the transparency estimate.

---

**8. "What is the estimand?"**

The average total effect of *assigned* fee transparency on willingness to adopt, averaged over the population of participants, product offers, and contexts similar to those sampled — at average financial literacy. The moderated effect at specific literacy levels is a secondary estimand. Comprehension and perceived transparency are excluded from the model because they are mechanisms, not confounders.

---

**9. "Why does heteroskedasticity not concern you more?"**

It concerns us, but not for the transparency point estimate. The transparency effect is a within-person contrast that is robust to mean-independent variance heterogeneity in most cases. Where it matters: confidence intervals for subgroup estimates in high-risk stimuli may be too narrow. We flag this as a caveat.

---

**10. "Can you say the model 'recovered' the true effect?"**

In this simulation we know the truth, so yes — we know the final estimate is close to the true effect. In a real dataset, we cannot verify recovery. What we can say is that the estimate is now more defensible: the major observed threats have been addressed, the modeling sequence has been documented, and the remaining assumptions have been stated explicitly.

---

## 23. Reporting Template

> We analyzed adoption-intent ratings (0–100) using linear mixed-effects models (lme4; Bates et al., 2015) with Satterthwaite degrees of freedom (lmerTest).
>
> **Assignment audit.** Pre-analysis diagnostics revealed implementation imbalance: the transparent-display condition was more common for higher-risk products and later trial positions. We verified this with a logistic regression predicting transparency from pre-treatment variables (risk level, expected value, financial literacy, site, trial order). All subsequent models include these variables as covariates to remove the realized imbalance.
>
> **Observed data.** Of [N] intended trials, [X]% had missing outcomes. Missingness was higher in the transparent condition ([X]% vs [X]%) and later trials. We used complete-case analysis and checked sensitivity by adding the main missingness predictor (device platform); the estimate shifted by [amount]. We additionally report estimates with and without attention-flagged trials and describe the estimand difference.
>
> **Model specification.** Fixed effects: transparency, financial literacy, their interaction, risk level, high-risk indicator, standardised expected value (linear + quadratic), trial order, and device platform. Random intercepts for participants, stimuli, and sites. By-participant random slope for transparency ([correlated / uncorrelated / dropped] after singularity check).
>
> **Results.** After adjusting for the implementation imbalance, fee transparency increased adoption intent by *b* = [X] (95% CI [X, X]), *t*([df]) = [X], *p* = [.X] at average financial literacy. Financial literacy moderated the effect (*b* = [X]), with larger benefits for lower-literacy participants.
>
> **Limitations.** Three inference threats were not fully resolved: (1) missingness may depend on unobserved latent adoption intent (MNAR); (2) attention-failure exclusions condition on a post-treatment variable, changing the estimand; (3) with four sites, site-level heterogeneity in the transparency effect cannot be reliably detected.

::: {.callout-tip icon=false}
## Minimum Reproducible Model Report Checklist

- [ ] Assignment audit: was treatment balanced on pre-treatment variables?
- [ ] Data flow table: rows intended vs observed vs dropped vs excluded
- [ ] Missingness audit: rate by condition, site, trial order
- [ ] Post-treatment timing audit: which variables are excluded and why
- [ ] Model formula (verbatim)
- [ ] Optimizer and convergence status; isSingular() result
- [ ] All fixed-effect estimates with SEs and CIs
- [ ] All random-effect variance components (SDs)
- [ ] Functional-form diagnostics (residuals vs covariates)
- [ ] Missingness sensitivity comparison
- [ ] Exclusion sensitivity comparison with estimand note
- [ ] Explicit estimand statement
- [ ] "What we fixed vs what we did not fix" summary
:::

---

## Appendix: Stress Tests

These tests exaggerate each threat to show what happens when problems are severe — and to confirm that the fixes in the main analysis work as expected.

### Stress Test A: Severe Risk Imbalance (Double the Coefficient)

```{r stress-a}
set.seed(101)
trials_stress_a <- trials |>
  left_join(stimuli  |> select(stimulus_id, u0_stimulus),  by = "stimulus_id") |>
  left_join(participants |> select(participant_id, u0_participant, u1_participant), by = "participant_id") |>
  left_join(site_effects, by = "site_id") |>
  mutate(
    trial_order_z_a = (trial_order - 6.5) / 3.452,
    # Much stronger imbalance
    p_tx_imb = plogis(1.40 * risk_level + 0.60 * trial_order_z_a + 0.80 * as.integer(site_id == 2)),
    tx_imb   = rbinom(n(), 1, p_tx_imb),
    mu_imb   = dgp$intercept +
               dgp$effect_transparency * tx_imb +
               dgp$effect_risk         * risk_level +
               dgp$effect_risk_high    * risk_high +
               dgp$effect_ev_linear    * expected_value_z +
               dgp$effect_ev_quad      * expected_value_z^2 +
               dgp$effect_literacy     * financial_literacy +
               dgp$interaction_tx_lit  * tx_imb * financial_literacy +
               dgp$effect_fatigue      * trial_order +
               dgp$effect_fatigue_late * as.integer(trial_order > 9) +
               u0_participant + u1_participant * tx_imb +
               u0_stimulus + u0_site,
    y_imb = pmin(100, pmax(0, rnorm(n(), mu_imb, dgp$sd_residual)))
  )

m_imb_raw <- lmer(y_imb ~ tx_imb + (1 | participant_id) + (1 | stimulus_id), data = trials_stress_a)
m_imb_adj <- lmer(y_imb ~ tx_imb + risk_level + risk_high + expected_value_z + trial_order +
                    (1 | participant_id) + (1 | stimulus_id), data = trials_stress_a)

bind_rows(
  broom.mixed::tidy(m_imb_raw, effects = "fixed") |> filter(term == "tx_imb") |> mutate(model = "Unadjusted"),
  broom.mixed::tidy(m_imb_adj, effects = "fixed") |> filter(term == "tx_imb") |> mutate(model = "Covariate-adjusted")
) |>
  mutate(`True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) |>
  knitr::kable(caption = "Stress test A: severe imbalance. Unadjusted is badly biased; covariate adjustment recovers the effect.")
```

### Stress Test B: Site-Level Treatment Heterogeneity

```{r stress-b}
set.seed(102)
site_effects_b <- site_effects |> mutate(u1_site = c(1.5, -1.0, 0.5, -1.0))

trials_stress_b <- trials |>
  left_join(stimuli  |> select(stimulus_id, u0_stimulus),  by = "stimulus_id") |>
  left_join(participants |> select(participant_id, u0_participant, u1_participant), by = "participant_id") |>
  left_join(site_effects_b, by = "site_id") |>
  mutate(
    mu_b = dgp$intercept +
           (dgp$effect_transparency + u1_site) * transparency +
           dgp$effect_risk      * risk_level + dgp$effect_risk_high * risk_high +
           dgp$effect_ev_linear * expected_value_z + dgp$effect_ev_quad * expected_value_z^2 +
           dgp$effect_literacy  * financial_literacy +
           dgp$effect_fatigue   * trial_order + dgp$effect_fatigue_late * as.integer(trial_order > 9) +
           u0_participant + u1_participant * transparency + u0_stimulus + u0_site,
    y_b = pmin(100, pmax(0, rnorm(n(), mu_b, dgp$sd_residual)))
  )

m_site_ix <- lmer(y_b ~ transparency * factor(site_id) + risk_level + risk_high +
                    expected_value_z + trial_order +
                    (1 | participant_id) + (1 | stimulus_id),
                  data = trials_stress_b)

broom.mixed::tidy(m_site_ix, effects = "fixed") |>
  filter(str_detect(term, "transparency")) |>
  select(term, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) |>
  knitr::kable(caption = "Stress test B: transparency effect varies across sites. A site RI alone would hide this.")
```

### Stress Test C: Sparse Within-Person Data

```{r stress-c}
set.seed(103)
trials_sparse <- trials |>
  group_by(participant_id) |>
  slice_sample(n = 4) |>
  ungroup()

m_sparse <- lmer(adoption_intent_obs ~ transparency * financial_literacy +
                   risk_level + expected_value_z +
                   (1 + transparency | participant_id) +
                   (1 | stimulus_id) + (1 | site_id),
                 data  = trials_sparse,
                 control = lmerControl(optimizer = "bobyqa"))

cat("Singular (4 trials per person):", isSingular(m_sparse), "\n")
cat("Teaching point: design justifies the slope; data cannot support it at n=4.\n")
```
