---
title: "Quasi-Experimental Methods for Applied Researchers"
---

```{r setup, include=FALSE}
library(tidyverse)
library(MatchIt)
library(broom)
library(broom.mixed)
library(lme4)
library(lmerTest)
library(ggplot2)
library(patchwork)
library(sandwich)
library(lmtest)
set.seed(42)
```

# Quasi-Experimental Methods for Applied Researchers

::: {.callout-note icon=false}
## The Experimental Recovery Question

> **If this experiment had gone perfectly — full compliance, no attrition, balanced randomization — what would each comparison identify?** Unbiased ITT effects for all contrasts.
>
> **What complications arise in practice?** Attrition creates a non-random survivor sample; voluntary sub-treatment use creates observational contrasts within arms; pre-post designs raise parallel trends questions.
>
> **What are the next-best estimands?** This chapter works through four cases: DiD logic for pre-post designs, IPW-corrected ITT for attrition, ATT for within-arm matching, and RDD-based LATE for threshold eligibility.
:::

## Section 1: When Experiments Have Observational Features

A fully randomized experiment with perfect compliance and no attrition is the gold standard for causal inference — but it rarely survives contact with the field intact. In practice, experiments accumulate observational features: participants drop out for reasons related to the treatment, some assigned to treatment never engage, ethical or logistical constraints limit who can be randomized, and field implementations vary substantially across sites.

These complications do not invalidate the experiment. They do require the researcher to reach into the quasi-experimental toolkit that Module 3 introduced for observational studies. The good news: the same tools that handle observational confounding also handle experimental imperfections — often in a cleaner setting, because the original randomization still provides some traction.

The table below maps common experimental complications to their methodological remedies and their Module 3 antecedents:

| Situation | Method | Module 3 Connection |
|-----------|--------|---------------------|
| Perfect randomization, no attrition | Standard mixed model | — (Modules 1–2) |
| Non-random attrition at follow-up | IPW for attrition | Part 6 (propensity scores) |
| Noncompliance with assignment | ITT or LATE via IV | Part 2 (IV) → Module 5 Part 8 |
| Eligibility determined by a threshold | Regression discontinuity | Part 4 (RDD) |
| Pre-post design with comparison group | DiD mixed model | Part 5 (DiD) |
| Self-selected sub-treatment within experiment | PSM within treatment arm | Part 6 (PS) |

: Experimental complications and their quasi-experimental remedies {tbl-colwidths="[30,30,40]"}

This chapter walks through implementations for four of these situations: DiD as the logic of pre-post designs, IPW for non-random attrition, PSM for voluntary sub-treatment participation, and threshold-based eligibility (RDD). The noncompliance case is covered in detail in Module 5 Part 8.

---

## Section 2: DiD as the Logic of Pre-Post Designs

### The structural equivalence

The standard randomized pre-post design **is** a difference-in-differences design. It has:

- Two groups: intervention and control (the "treated vs. untreated" dimension of DiD)
- Two time points: baseline and follow-up (the "pre vs. post" dimension of DiD)
- The DiD estimand: $(\bar{Y}^{\text{post}}_{\text{int}} - \bar{Y}^{\text{pre}}_{\text{int}}) - (\bar{Y}^{\text{post}}_{\text{ctrl}} - \bar{Y}^{\text{pre}}_{\text{ctrl}})$

Recognizing this equivalence is valuable for two reasons. First, it makes explicit what the pre-post mixed model assumes (parallel trends in the absence of treatment — which randomization supports in expectation across repeated samples, though any realized sample can show chance pre-trend imbalance). Second, it gives you the full DiD toolkit: you can add multiple waves, test pre-trend balance, and use the event study framework to trace the treatment trajectory over time.

### Simulation: information disclosure intervention

We simulate 120 participants (60 per arm) measured at baseline (pre) and 8-week follow-up (post). An information disclosure intervention (gain-framed vs. standard disclosure) is randomized. The disclosure intervention is expected to increase `risk_willingness` (gain framing increases confidence in recommended investments). True DiD effect = +8 units.

```{r did-simulate}
n_per <- 60

df_pp <- expand_grid(
  participant = 1:(n_per * 2),
  wave        = c("pre", "post")
) |>
  mutate(
    condition = ifelse(participant <= n_per, "standard", "gain_frame"),
    time_num  = as.integer(wave == "post"),
    treat_num = as.integer(condition == "gain_frame"),
    u0        = rnorm(n_per * 2, 0, 6)[participant],
    risk_willingness = 45 + u0 + 1.5 * time_num + 8 * time_num * treat_num + rnorm(n(), 0, 5),
    participant = factor(participant),
    wave      = factor(wave, levels = c("pre", "post")),
    condition = factor(condition, levels = c("standard", "gain_frame"))
  )
```

### DiD mixed model

```{r did-model}
m_did_exp <- lmer(risk_willingness ~ wave * condition + (1 | participant),
                  data = df_pp, REML = FALSE)

broom.mixed::tidy(m_did_exp, effects = "fixed") |>
  knitr::kable(digits = 3,
               caption = "DiD mixed model. The wave:condition interaction is the DiD estimate (target: +8).")
```

The `wavepost:conditiongain_frame` interaction coefficient is the DiD estimand — the differential change from pre to post in the gain-frame arm relative to the standard disclosure arm.

### Comparison to ANCOVA

ANCOVA on the wide-format outcome, controlling for the pre-test, is a common alternative:

```{r ancova}
df_wide <- df_pp |>
  select(participant, condition, wave, risk_willingness) |>
  pivot_wider(names_from = wave, values_from = risk_willingness,
              names_prefix = "rw_") |>
  mutate(rw_pre_c = rw_pre - mean(rw_pre))

m_ancova <- lm(rw_post ~ condition + rw_pre_c, data = df_wide)

broom::tidy(m_ancova) |>
  filter(term == "conditiongain_frame") |>
  knitr::kable(digits = 3,
               caption = "ANCOVA treatment estimate. Numerically close to DiD mixed model when pre-test is mean-centered.")
```

The two approaches converge when the pre-test is mean-centered and the random-effect variance is estimated well. The DiD mixed model generalizes more naturally to three or more waves and produces an interpretable random-effects structure, making it preferred for longitudinal experimental designs.

Recognizing this equivalence is valuable: the pre-test provides a baseline for the DiD logic, and randomization supports parallel trends in expectation.

### Event study visualization

```{r did-plot, fig.width=8, fig.height=4.5}
summ_pp <- df_pp |>
  group_by(wave, condition) |>
  summarise(mean_rw = mean(risk_willingness), se_rw = sd(risk_willingness) / sqrt(n()),
            .groups = "drop")

ggplot(summ_pp, aes(x = wave, y = mean_rw,
                     color = condition, group = condition)) +
  geom_line(linewidth = 1.1) +
  geom_errorbar(aes(ymin = mean_rw - 1.96 * se_rw,
                    ymax = mean_rw + 1.96 * se_rw), width = 0.1) +
  geom_point(size = 3.5) +
  scale_color_manual(values = c("standard" = "#aec6e8", "gain_frame" = "#1e3a5f")) +
  annotate("segment",
           x = 2.05, xend = 2.05, y = 47, yend = 55,
           arrow = arrow(length = unit(0.25, "cm"), ends = "both"),
           color = "#c0392b") +
  annotate("text", x = 2.12, y = 51,
           label = "DiD \u2248 +8", color = "#c0392b", hjust = 0, size = 3.5) +
  labs(
    title    = "Pre-post risk willingness by condition",
    subtitle = "Parallel pre-period (randomization), divergent post-period (treatment)",
    x        = "Wave",
    y        = "Mean risk willingness (0\u2013100)",
    color    = "Condition"
  ) +
  theme_minimal(base_size = 13)
```

### Three-wave extension

With baseline, mid-point, and follow-up, the time × condition interaction becomes a trajectory. Pre-baseline balance checks verify randomization; post-baseline coefficients trace how the treatment effect builds over time. This is the experimental event study:

```{r three-wave-sketch, eval=FALSE}
# Three-wave data with weeks 0, 4, 8
m_3wave <- lmer(risk_willingness ~ factor(week) * condition + (1 | participant),
                data = df_3wave)
# Plot estimated coefficients for each week x condition cell
```

Pre-period interaction coefficients near zero confirm that randomization produced balanced groups. Post-period coefficients show whether treatment effects are immediate, delayed, or accumulating.

---

## Section 3: IPW for Non-Random Attrition

### The problem

Even in a fully randomized trial, attrition at follow-up can reintroduce bias. If high-risk-averse control-group participants are more likely to drop out — because the study involves making financial decisions that feel threatening and they disengage — then the observed control-group follow-up average is deflated (it excludes the most risk-averse participants). The treatment appears less effective than it truly is for the full intention-to-treat sample, biasing the naive estimate downward.

### Simulation

```{r ipw-simulate}
n_att <- 200
set.seed(42)

treatment_att    <- rep(c(0, 1), each = 100)
baseline_risk_av <- rnorm(n_att, 50, 10)   # baseline risk aversion

# High risk aversion + control group → higher dropout
# (risk-averse participants avoid making financial decisions in the study)
p_remain   <- plogis(2.5 - 0.07 * baseline_risk_av + 0.8 * treatment_att)
remained   <- rbinom(n_att, 1, p_remain)

# True treatment effect = +6 (disclosure increases willingness to take recommended risk)
outcome_att <- 45 + 6 * treatment_att - 0.3 * baseline_risk_av + rnorm(n_att, 0, 6)

df_att <- tibble(treatment = treatment_att, baseline_risk_av, remained, outcome_att)

cat("Retention rate — control:  ", mean(remained[treatment_att == 0]), "\n")
cat("Retention rate — treatment:", mean(remained[treatment_att == 1]), "\n")
```

### IPW correction

We model the probability of remaining in the study as a function of baseline risk aversion and treatment, then weight complete cases by the inverse of their retention probability:

```{r ipw-model}
ps_att     <- glm(remained ~ baseline_risk_av + treatment, family = binomial, data = df_att)
df_att$ipw_att <- 1 / fitted(ps_att)

df_complete     <- df_att |> filter(remained == 1)
m_naive_att     <- lm(outcome_att ~ treatment, data = df_complete)
m_ipw_att       <- lm(outcome_att ~ treatment, data = df_complete,
                       weights = df_att$ipw_att[df_att$remained == 1])
```

```{r ipw-compare, echo=FALSE}
bind_rows(
  broom::tidy(m_naive_att) |> filter(term == "treatment") |>
    mutate(method = "Naive complete-case analysis"),
  broom::tidy(m_ipw_att)   |> filter(term == "treatment") |>
    mutate(method = "IPW-corrected analysis")
) |>
  select(method, estimate, std.error, p.value) |>
  knitr::kable(
    digits  = 3,
    caption = "Attrition bias: naive vs. IPW-weighted (true treatment effect = +6). High-risk-averse participants in the control group drop out more, biasing the naive estimate downward. IPW is consistent under MAR."
  )
```

::: {.callout-warning icon=false}
## IPW Assumptions for Attrition Correction

IPW corrects for attrition under **Missing at Random (MAR)**: the probability of dropout depends on observed pre-treatment variables (here, baseline risk aversion and treatment condition), but not on the unobserved post-treatment outcome. If dropout is driven by the unmeasured outcome itself — e.g., people drop out because their financial situation has worsened and they do not want to be measured — IPW will not fully correct the bias. Consider collecting data on dropouts via brief follow-up contacts to test the MAR assumption empirically.
:::

### Visualization

```{r ipw-plot, fig.width=7, fig.height=4}
# Retention probability plot
df_plot_ret <- df_att |>
  mutate(
    p_ret    = fitted(ps_att),
    trt_lab  = ifelse(treatment == 1, "Treatment", "Control")
  )

ggplot(df_plot_ret, aes(x = baseline_risk_av, y = p_ret, color = trt_lab)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("Control" = "#e67e22", "Treatment" = "#1e3a5f"),
                     name = "Group") +
  labs(
    title    = "Retention probability by baseline risk aversion and treatment",
    subtitle = "High-risk-averse controls are most likely to drop out — biasing naive estimate downward",
    x        = "Baseline Risk Aversion",
    y        = "P(Retained at follow-up)"
  ) +
  theme_minimal(base_size = 13)
```

---

## Section 4: PSM for Voluntary Sub-Treatment Participation

### The within-experiment observational problem

Random assignment creates treatment and control arms, but within the treatment arm, **engagement is voluntary**. In a financial coaching field experiment, random assignment gives participants *access* to an AI-powered savings advisor. Actual usage is voluntary — high-usage participants likely differ systematically from low-usage participants in motivation, baseline financial literacy, and financial goals.

Propensity score matching on pre-treatment covariates makes this within-arm comparison more credible by constructing a control group of low-users who resembled high-users at baseline.

### Simulation and matching

We work within the treatment arm of the attrition simulation and create a high-use indicator that is related to baseline risk aversion:

```{r psm-usage}
df_treat_only <- df_att |>
  filter(treatment == 1) |>
  mutate(
    high_user = rbinom(n(), 1, plogis(-0.5 + 0.04 * baseline_risk_av))
  )

# Within the coaching arm: match high-users to low-users
# on pre-randomization covariates
m_sub <- matchit(
  high_user ~ baseline_risk_av,
  data   = df_treat_only,
  method = "nearest",
  ratio  = 1
)

summary(m_sub)$sum.matched |>
  knitr::kable(digits = 3,
               caption = "Covariate balance after 1:1 nearest-neighbor matching on baseline risk aversion.")
```

```{r psm-estimate}
matched      <- match.data(m_sub)
m_usage_eff  <- lm(outcome_att ~ high_user + baseline_risk_av,
                   data = matched, weights = weights)

broom::tidy(m_usage_eff) |>
  filter(term == "high_user") |>
  select(estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
               caption = "Matched estimate of high-use vs. low-use effect on savings outcome, among participants with similar baseline risk aversion.")
```

::: {.callout-warning icon=false}
## What PSM Within an Experiment Identifies

Matching on pre-treatment covariates removes **observed** confounders of the high-use vs. low-use comparison. But unobserved motivation, financial self-efficacy, and expectancy effects remain. The matched estimate is closer to causal than the naive comparison, but it is not an RCT-quality estimate.

Formally, this identifies the **Average Treatment effect on the Treated (ATT)** for the sub-population of high-users — the effect of high engagement among people with observable characteristics similar to those who actually engaged heavily. This is not a LATE (which is an IV estimand defined by instrument-induced compliers) and is not the ATE for all participants. Interpret as "the effect of high engagement, for people with similar observable characteristics to high engagers," not "the effect of high engagement for everyone." Crucially, this rests on a selection-on-observables assumption: all confounders of the usage–outcome relationship are measured and included in the PS model.
:::

---

## Section 5: Threshold-Based Eligibility (RDD in Applied Research)

When access to a program is determined by whether a continuous running variable crosses a threshold, **regression discontinuity design (RDD)** is the appropriate analysis. The continuity assumption — that potential outcomes vary smoothly across the cutoff — is often more credible in financial-decision settings than the selection-on-observables assumption required by propensity score methods.

A common example: a platform introduces a fee waiver for customers whose account balance exceeds $1,000. Customers just below and just above the threshold are otherwise similar — the threshold creates local quasi-random variation. Similarly, subsidy eligibility cutoffs based on income or credit score thresholds produce the same discontinuity logic.

```{r rdd-sketch, eval=FALSE}
# Requires rdrobust package
library(rdrobust)

# score_threshold: running variable (e.g., account balance in dollars)
# eligible: indicator for whether balance >= 1000
# outcome_rdd: e.g., monthly savings rate or financial product adoption

rdd_out <- rdrobust(y = df$outcome_rdd, x = df$score_threshold, c = 1000)
summary(rdd_out)
rdplot(y = df$outcome_rdd, x = df$score_threshold, c = 1000,
       title = "RDD: account balance fee-waiver threshold at $1,000")
```

The `rdrobust` package implements optimal bandwidth selection, bias-corrected confidence intervals, and the McCrary density test (for manipulation of the running variable). The full treatment of RDD mechanics, bandwidth selection, and robustness checks is in Module 3 Part 4; that framework applies directly here.

::: {.callout-tip}
## RDD in Financial and Consumer Behavior Research

RDD is underused in consumer behavior and financial decision research relative to economics, despite being highly credible when a threshold rule genuinely determines assignment. Common examples: credit score thresholds for product access, account balance thresholds for fee waivers or tier upgrades, subsidy eligibility cutoffs based on income, and performance review thresholds that trigger intervention. If your research setting uses a cutoff rule, seriously consider RDD before reaching for PSM.
:::

---

## Section 6: Practical Decision Guide

::: {.callout-tip}
## Choosing Your Method When Randomization Is Imperfect

Work through this sequence when your experimental data does not satisfy the simple randomization ideal:

1. **Fully randomized, no attrition, full compliance?**
   Standard mixed model (Modules 1–2). No quasi-experimental adjustment needed.

2. **Randomized + differential attrition?**
   IPW for attrition (this chapter, Section 3). Requires MAR assumption.

3. **Randomized + noncompliance?**
   Report ITT (always). If complier effect is of interest, add LATE via IV (Module 5 Part 8).

4. **Eligibility determined by a continuous threshold?**
   Regression discontinuity (Module 3 Part 4). Local to the threshold — do not generalize far from cutoff.

5. **Pre-post design with randomized comparison group?**
   DiD mixed model (this chapter, Section 2). Generalizes to multiple waves as event study.

6. **Self-selection into sub-treatment within an experiment?**
   PSM within treatment arm (this chapter, Section 4). Residual unobserved confounding remains.

7. **Observational study with good observed confounders?**
   PSM or IPW on covariates (Module 3 Part 6). Validity depends entirely on selection-on-observables.

8. **Observational study with a credible instrument?**
   IV or LATE (Module 3 Part 2, Module 5 Part 8). Validity depends on exclusion restriction.

No method is universally best. The right method depends on the data-generating process and which assumption is most defensible given your study design.
:::

---

## Section 7: The Full Course Arc

This workshop has built from the ground up.

**Module 1** established the mixed-effects framework: how to model data where observations are nested within units, how to decompose variance across levels, and why treating clustered data as independent inflates false positive rates. The random intercept model is not a statistical correction tacked on after the fact — it is a direct translation of the sampling structure into the model.

**Module 2** showed that *how* you enter predictors matters as much as *which* predictors you include. Within-unit centering decomposes predictors into their between-unit and within-unit components, corresponding to different mechanisms operating at different levels of the hierarchy. Getting the centering wrong means estimating the wrong estimand, even with a correctly specified error structure.

**Module 3** showed where the well-specified mixed model can still go wrong: omitted variables (time-invariant and time-varying), endogenous regressors, selection into the sample, regression discontinuities at thresholds, treatment effect heterogeneity, and non-random treatment assignment. Six threats; six identification strategies. Each strategy requires a different source of identifying variation and a different set of untestable assumptions.

**Module 4** addressed the practicalities of longitudinal data: handling missing values without complete-case bias, and applying the Module 3 identification strategies to the panel setting where time creates both new threats and new opportunities.

**Module 5** closed the loop for applied experimental researchers: what to do when the experiment is imperfect, and how to connect the quasi-experimental toolkit back to the experimental design.

The unifying insight across all five modules:

> Understanding your **data generating process** — who is nested in whom, what varies within units versus across units, what mechanisms connect your predictors to your outcome, and what determines who ends up in each condition — is the prerequisite for every methodological choice covered in this workshop.

The nesting structure is not a nuisance to correct for with a random effect and move on. It is the scientific structure of your data. The causal question is always about a mechanism operating at a particular level of that structure, through a particular pathway, for a particular population. The method is in service of making that mechanism legible.
