---
title: "A Fully Worked Multilevel Modeling Example"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
set.seed(20240610)
```

A student brings us a result: transparent fee disclosure appears to increase adoption intent. The p-value is small. The plot looks clean. But before we believe the estimate, we audit the data. We find **seven threats**: repeated observations, sampled stimuli, treatment-risk imbalance, nonlinear product valuation and fatigue, differential missingness, post-treatment exclusion risk, and treatment-effect heterogeneity.

The rest of this chapter diagnoses where each threat comes from, which adjustment reduces it, and what remains when we are done.

> **The key framing:** Statistical fixes minimize threats — they do not erase them. The goal is not to find an analysis that says everything was fine. It is to document what each adjustment addressed, what it could not address, and what the final claim depends on.

---

## How to Read This Worked Example

You do not need to read every paragraph to follow the logic. Each step answers one diagnostic question.

| Section | Diagnostic question | Model | What to watch |
|---|---|---|---|
| §7 | Is the naive estimate credible? | m0: pooled OLS | Estimate far below true effect of 4 |
| §8 | Are rows independent? | m1: + participant RI | Participant ICC; SE change |
| §9 | Do products dominate responses? | m2: + stimulus RI | **Large estimate jump — product-level confounding absorbed** |
| §10 | Are sites confounded with assignment? | m3: + site RI | Site means; small movement |
| §11 | Is product-level adjustment explicit? | m4: + risk/EV covariates | Interpretability, not large movement |
| §12 | Is functional form right? | m4b: + nonlinear + fatigue | **Estimate moves again — remaining bias corrected** |
| §13 | Are observed rows selected? | m4c: + device | Missingness sensitivity |
| §14 | Do exclusions change the estimand? | m4d: attention excluded | Estimand change, not magnitude |
| §15 | For whom does it work? | m5: + literacy interaction | Claim changes |
| §16 | Does the effect vary across people? | m6b: + random slope | SE; heterogeneity |

---

## 1. The Research Question and Estimand

> **Does transparent fee disclosure increase willingness to adopt a financial product, and does that effect generalize across participants, product offers, and research sites?**

The **target estimand** is the average total effect of *assigned* fee transparency on willingness to adopt, averaged over the population of participants, product offers, and contexts similar to those sampled. This is an intention-to-treat estimand: the effect of the display assignment, not the effect of comprehension or perceived clarity. Comprehension and perceived transparency are potential mechanisms and must be excluded from the main model.

---

## 2. Design Audit Before Modeling

| Design question | Answer | Modeling implication |
|---|---|---|
| What is one row? | One participant's rating of one product | Trial-level outcome; multiple rows per participant |
| What is randomized? | Fee transparency (within-participant) | Fixed effect for transparency |
| Does treatment vary within participant? | Yes | Candidate by-participant random slope |
| Are stimuli sampled? | Yes — 36 from a larger set | Stimulus random intercept |
| Are participants crossed with stimuli? | Partially — 12 of 36 each | Crossed (partial) random effects |
| Are participants nested in sites? | Yes — 240 across 4 sites | Site random intercept |
| Is assignment perfectly randomized? | **No — correlated with risk, trial order, site 2** | Covariate adjustment needed |
| What missingness is plausible? | Higher for transparent, high-risk, late, low-literacy | Missingness audit required |
| Which variables are post-treatment? | Comprehension, perceived transparency | Exclude from main model |

---

## 3. Simulate the Latent DGP

The true data-generating process is known. The *realized* data will not look like it because: (1) the fee-transparency display was more commonly served to high-risk products in later trials — a real implementation pattern; (2) trial-order fatigue is real; (3) the observation process introduces non-random missingness and post-treatment attention failures.

```{r true-dgp-params}
#| code-summary: "Show DGP parameters"
dgp <- list(
  intercept           = 50,
  effect_transparency = 4,      # +4 at average literacy — the target
  effect_risk         = -2.0,   # linear risk penalty
  effect_risk_high    = -2.5,   # threshold penalty for risk > 0.75
  effect_ev_linear    = 2.2,    # EV (standardised)
  effect_ev_quad      = -1.0,   # diminishing returns
  effect_literacy     = 2,
  interaction_tx_lit  = -1.5,   # transparency helps low-literacy more
  effect_fatigue      = -0.18,  # per trial
  effect_fatigue_late = -1.20,  # extra penalty for trials 10–12
  effect_device       = -1.5,   # mobile: lower ratings
  sd_participant      = 8,
  sd_stimulus         = 5,
  sd_site             = 4,
  sd_slope_part       = 2,
  sd_residual         = 7
)
```

```{r simulate-latent}
#| code-summary: "Show simulation code (participants, stimuli, trials)"
n_participants  <- 240
n_stimuli       <- 36
n_sites         <- 4
trials_per_part <- 12

participants <- tibble(
  participant_id     = 1:n_participants,
  site_id            = rep(1:n_sites, each = n_participants / n_sites),
  financial_literacy = rnorm(n_participants, 0, 1),
  u0_participant     = rnorm(n_participants, 0, dgp$sd_participant),
  u1_participant     = rnorm(n_participants, 0, dgp$sd_slope_part)
)

site_effects <- tibble(
  site_id = 1:n_sites,
  u0_site = c(3, -2, 1, -2)
)

stimuli <- tibble(
  stimulus_id    = 1:n_stimuli,
  risk_level     = rnorm(n_stimuli, 0, 1),
  expected_value = rnorm(n_stimuli, 50, 15),
  u0_stimulus    = rnorm(n_stimuli, 0, dgp$sd_stimulus)
) |>
  mutate(
    expected_value_z = (expected_value - 50) / 15,
    risk_high        = as.integer(risk_level > 0.75)
  )

trials <- participants |>
  select(participant_id, site_id, financial_literacy, u0_participant, u1_participant) |>
  cross_join(tibble(draw = 1:trials_per_part)) |>
  group_by(participant_id) |>
  mutate(
    stimulus_id   = sample(1:n_stimuli, trials_per_part, replace = FALSE),
    trial_order   = 1:trials_per_part
  ) |>
  ungroup() |>
  select(-draw) |>
  left_join(stimuli, by = "stimulus_id") |>
  left_join(site_effects, by = "site_id") |>
  mutate(
    device_mobile = rbinom(n(), 1, plogis(-0.3 + 0.6 * as.integer(site_id %in% c(2, 3)))),

    # IMPERFECT ASSIGNMENT: transparent display more common for high-risk products,
    # later trial positions, and at site 2 — a realistic platform rollout pattern
    trial_order_z = (trial_order - 6.5) / 3.452,
    p_tx          = plogis(0.70 * risk_level + 0.40 * trial_order_z +
                             0.50 * as.integer(site_id == 2)),
    transparency  = rbinom(n(), 1, p_tx),

    mu = dgp$intercept +
         dgp$effect_transparency  * transparency +
         dgp$effect_risk          * risk_level +
         dgp$effect_risk_high     * risk_high +
         dgp$effect_ev_linear     * expected_value_z +
         dgp$effect_ev_quad       * expected_value_z^2 +
         dgp$effect_literacy      * financial_literacy +
         dgp$interaction_tx_lit   * transparency * financial_literacy +
         dgp$effect_fatigue       * trial_order +
         dgp$effect_fatigue_late  * as.integer(trial_order > 9) +
         dgp$effect_device        * device_mobile +
         u0_participant + u1_participant * transparency +
         u0_stimulus + u0_site,

    resid_sd = dgp$sd_residual * exp(
      0.25 * abs(risk_level) + 0.20 * as.integer(financial_literacy < -1)
    ),
    adoption_intent = pmin(100, pmax(0, rnorm(n(), mu, resid_sd)))
  ) |>
  select(participant_id, site_id, stimulus_id, trial_order,
         transparency, risk_level, risk_high, expected_value, expected_value_z,
         financial_literacy, device_mobile, adoption_intent)
```

---

## 4. The Observation Process

```{r simulate-observed}
#| code-summary: "Show observation-process simulation (missingness, attention)"
set.seed(20240611)
trials <- trials |>
  mutate(
    p_missing = plogis(
      -2.5 + 0.45 * transparency + 0.55 * risk_level +
        0.45 * as.integer(trial_order > 9) -
        0.50 * financial_literacy + 0.35 * device_mobile
    ),
    missing_y           = rbinom(n(), 1, p_missing),
    adoption_intent_obs = if_else(missing_y == 1, NA_real_, adoption_intent),
    attention_fail      = rbinom(n(), 1,
      plogis(-3.0 + 0.50 * transparency + 0.35 * risk_level +
               0.45 * as.integer(trial_order > 9) - 0.40 * financial_literacy))
  ) |>
  select(-p_missing)
```

---

## 5. Initial Inference Threat Dashboard

::: {.callout-important icon=false}
## We Are Starting With 7 Open Threats

1. **Participant dependence** — same participants appear across multiple rows.
2. **Stimulus heterogeneity** — product attractiveness varies widely; stimuli are sampled, not exhaustive.
3. **Treatment–risk imbalance** — transparent trials disproportionately involve high-risk products.
4. **Nonlinear valuation and fatigue** — risk and EV have threshold/diminishing effects; late-trial fatigue is real.
5. **Differential missingness** — missing outcomes are more common for transparent, high-risk, late, and low-literacy trials.
6. **Post-treatment exclusion risk** — attention failures are partly caused by the treatment; excluding them conditions on a post-treatment variable.
7. **Treatment-effect heterogeneity** — the transparency effect varies across participants and by financial literacy.
:::

```{r threat-dashboard}
#| code-summary: "Show threat dashboard table"
tibble(
  `#` = 1:7,
  Threat = c("Participant dependence", "Stimulus heterogeneity",
             "Treatment–risk imbalance", "Nonlinear valuation/fatigue",
             "Differential missingness", "Post-treatment exclusions",
             "Effect heterogeneity"),
  `Diagnostic evidence` = c(
    "ICC ≈ large; repeated observations within person",
    "Product random-effect SD > transparency effect",
    "Risk coefficient 0.741 predicting transparency in assignment audit",
    "Residuals curve against risk and EV; later trials lower",
    "Missing rate: transparent 19.9% vs non-transparent 11.6%",
    "Attention failure: transparent 12.1% vs non-transparent 5.4%",
    "Wide distribution of empirical per-person effects"
  ),
  Severity = c("High","High","High","Medium–High","High","Medium","Medium"),
  `Primary response` = c(
    "Participant random intercept (m1)",
    "Stimulus random intercept (m2)",
    "Stimulus RI absorbs most; covariates make it explicit (m2, m4)",
    "Nonlinear terms + trial order (m4b)",
    "Missingness-predictor sensitivity (m4c)",
    "Report all-observed vs excluded (m4d)",
    "Random slope; literacy interaction (m5, m6b)"
  )
) |>
  knitr::kable(caption = "Initial threat dashboard — each line tracks to a specific model or sensitivity check")
```

---

## 6. Diagnose the Threats: Empirical Evidence

### 6.1 Assignment imbalance

```{r assignment-audit}
#| code-summary: "Show assignment audit code"
tx_audit <- glm(
  transparency ~ risk_level + expected_value_z + financial_literacy +
    factor(site_id) + trial_order_z,
  data   = trials |> mutate(trial_order_z = (trial_order - 6.5) / 3.452),
  family = binomial()
)

broom::tidy(tx_audit) |>
  filter(term != "(Intercept)") |>
  mutate(
    flag = case_when(abs(estimate) > 0.30 ~ "⚠ Large", abs(estimate) > 0.10 ~ "Moderate", TRUE ~ "Small"),
    across(c(estimate, std.error, p.value), \(x) round(x, 3))
  ) |>
  select(term, estimate, std.error, p.value, flag) |>
  knitr::kable(caption = "Assignment audit: large coefficients confirm treatment–risk and trial-order imbalance")
```

### 6.2 Missingness and attention failures

```{r missingness-audit}
#| code-summary: "Show missingness tables"
cat("Overall missing rate:", round(100 * mean(trials$missing_y), 1), "%\n")

trials |>
  group_by(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  summarise(n = n(), n_miss = sum(missing_y), miss_pct = round(100 * mean(missing_y), 1),
            attn_fail_pct = round(100 * mean(attention_fail), 1)) |>
  knitr::kable(caption = "Missingness and attention failures by condition — both are higher for transparent trials")

trials |>
  group_by(trial_order) |>
  summarise(miss_pct = round(100 * mean(missing_y), 1)) |>
  knitr::kable(caption = "Missingness by trial order: late-trial dropout is real")
```

### 6.3 Data retention

```{r retention-table}
#| code-summary: "Show retention table"
n_int <- nrow(trials)
n_obs <- sum(!is.na(trials$adoption_intent_obs))
tibble(
  Stage = c("Intended", "Observed (non-missing)", "Attention failures (all trials)",
            "Clean analytic sample"),
  n   = c(n_int, n_obs, sum(trials$attention_fail), sum(!is.na(trials$adoption_intent_obs) & !trials$attention_fail)),
  pct = round(100 * c(n_int, n_obs, sum(trials$attention_fail), sum(!is.na(trials$adoption_intent_obs) & !trials$attention_fail)) / n_int, 1)
) |>
  knitr::kable(caption = "Data flow: rows available at each stage")
```

### 6.4 Outcome overview

```{r outcome-checks, fig.width=9, fig.height=3.5}
#| code-summary: "Show outcome diagnostic plots"
p1 <- ggplot(trials, aes(x = adoption_intent_obs)) +
  geom_histogram(bins = 30, fill = "#1e3a5f", color = "white", na.rm = TRUE) +
  labs(title = "Observed distribution", x = "Adoption intent (0–100)") +
  theme_minimal(base_size = 10)

p2 <- trials |> filter(!is.na(adoption_intent_obs)) |>
  mutate(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  ggplot(aes(x = Condition, y = adoption_intent_obs, fill = Condition)) +
  geom_boxplot(show.legend = FALSE, width = 0.5) +
  scale_fill_manual(values = c("#cccccc", "#1e3a5f")) +
  labs(title = "By condition", y = "Adoption intent", x = NULL) +
  theme_minimal(base_size = 10)

p3 <- trials |> filter(!is.na(adoption_intent_obs)) |>
  ggplot(aes(x = trial_order, y = adoption_intent_obs, group = trial_order)) +
  geom_boxplot(fill = "#2d6a4f", alpha = 0.6, outlier.size = 0.5) +
  labs(title = "By trial order", y = NULL, x = "Trial") +
  theme_minimal(base_size = 10)

gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
cat("Floor (≤2):", round(mean(trials$adoption_intent_obs <= 2, na.rm=TRUE), 3),
    "| Ceiling (≥98):", round(mean(trials$adoption_intent_obs >= 98, na.rm=TRUE), 3), "\n")
```

### 6.5 Post-treatment variable timing

```{r timing-table}
#| code-summary: "Show variable-timing table"
tibble(
  Variable = c("Risk level", "Expected value", "Financial literacy", "Trial order",
               "Device type", "Attention check", "Perceived transparency", "Comprehension"),
  Timing   = c("Pre-treatment stimulus", "Pre-treatment stimulus",
               "Pre-treatment participant", "Design/process variable",
               "Pre-treatment contextual", "Post-trial — ambiguous",
               "Post-treatment mediator", "Post-treatment mediator"),
  `Safe in main model?` = c("Yes","Yes","Yes","Yes — controls fatigue",
                             "Yes — controls missingness","Caution",
                             "No — changes estimand","No — changes estimand")
) |>
  knitr::kable(caption = "Post-treatment timing audit: comprehension and perceived transparency must be excluded")
```

::: {.callout-warning icon=false}
## Comprehension Must Stay Out of the Model

```{mermaid}
flowchart LR
  T["Transparency (assigned)"] --> C["Comprehension"]
  C --> Y["Adoption intent"]
  T --> Y
  L["Literacy"] --> C
  L --> Y
```

Including comprehension estimates only the non-comprehension pathway — a different and narrower estimand than "does the display change adoption intent?"
:::

---

## 7. Start With the Wrong Model

::: {.callout-note icon=false}
## At a Glance

- **Open threat:** All seven — nothing addressed yet.
- **Model:** Pooled OLS on complete cases.
- **What to watch:** Estimate relative to the true effect of `r dgp$effect_transparency`. How many rows were silently dropped?
:::

```{r m0-naive}
m0 <- lm(adoption_intent_obs ~ transparency, data = trials)

cat("Dataset rows:", nrow(trials), "| Rows used:", nobs(m0),
    "| Silently dropped:", nrow(trials) - nobs(m0), "\n\n")

broom::tidy(m0) |>
  filter(term == "transparency") |>
  mutate(model = "m0: pooled OLS", `True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)
```

**Result:** The naive estimate is `r round(coef(m0)["transparency"], 2)` — `r round(dgp$effect_transparency - coef(m0)["transparency"], 2)` points below the true effect of `r dgp$effect_transparency`. It is significant, which is exactly what makes it dangerous.

**Why is the estimate too low?**

- **Treatment–risk imbalance:** transparent trials go to high-risk products; high risk reduces adoption → transparency is blamed for the risk penalty.
- **Trial-order confound:** transparent trials occur later on average; late trials have lower ratings due to fatigue → transparency is also blamed for fatigue.
- **Clustering ignored:** the same participants and products appear repeatedly, inflating apparent precision.
- **Missingness ignored:** `r nrow(trials) - nobs(m0)` rows were dropped without comment; these are not a random subsample.

::: {.callout-important icon=false}
## Threat Status After m0: 7 of 7 Open

All threats are active. The estimate is far from the true effect because structural, confounding, and missingness problems are all compounded.

**Open:** participant dependence, stimulus heterogeneity, treatment–risk imbalance, nonlinear valuation/fatigue, differential missingness, post-treatment exclusions, effect heterogeneity.
:::

---

## 8. Participant Random Intercepts: Fix Repeated-Person Dependence

::: {.callout-note icon=false}
## At a Glance

- **Open threat:** Participant dependence — same person appears 12 times.
- **Evidence:** Each participant has a stable adoption tendency. Treating rows as independent ignores this.
- **Model change:** Add `(1 | participant_id)`.
- **What to watch:** The participant ICC and how the SE changes. The estimate should **not** move much — participant RI addresses dependence, not the stimulus-level risk confounding.
:::

```{r m1}
m1 <- lmer(adoption_intent_obs ~ transparency + (1 | participant_id), data = trials)
```

```{r m1-summary}
vc1  <- as.data.frame(VarCorr(m1))
icc1 <- vc1$vcov[vc1$grp == "participant_id"] / sum(vc1$vcov)
cat("Participant ICC:", round(icc1, 3), "— strong clustering by person\n")

broom.mixed::tidy(m1, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m1: + participant RI", `True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)
```

**Result:** Estimate moved from `r round(coef(m0)["transparency"], 2)` to `r round(fixef(m1)["transparency"], 2)` (+`r round(fixef(m1)["transparency"] - coef(m0)["transparency"], 2)` pts). The estimate barely changed, as expected. The SE decreased because within-person baselines were removed from the residual.

- **What this fixed:** person-to-person baseline differences are no longer in the residual.
- **Why the estimate did not move:** the main bias is stimulus-level (transparent trials go to risky products). Participant RI does not adjust for that.
- **What is still open:** product-level confounding is the main problem — that comes next.

::: {.callout-important icon=false}
## Threat Status After m1: 6 of 7 Open

**Reduced:** ~~Participant dependence~~

**Still open:** stimulus heterogeneity, treatment–risk imbalance, nonlinear valuation/fatigue, differential missingness, post-treatment exclusions, effect heterogeneity.

**Current estimate: `r round(fixef(m1)["transparency"], 2)` — still far below truth (`r dgp$effect_transparency`).**
:::

---

## 9. Stimulus Random Intercepts: Generalize and Absorb Product Confounding

::: {.callout-note icon=false}
## At a Glance

- **Open threat:** Stimulus heterogeneity and (partly) treatment–risk imbalance.
- **Evidence:** Product-level variation is large; stimulus SD is expected to exceed the transparency effect.
- **Model change:** Add `(1 | stimulus_id)`.
- **What to watch:** Stimulus SD relative to the transparency effect. The **estimate is expected to jump** because risk is a stimulus-level variable, and the stimulus RI absorbs product-level attractiveness differences — including the risk-related part.
:::

```{r m2}
m2 <- lmer(adoption_intent_obs ~ transparency +
             (1 | participant_id) + (1 | stimulus_id),
           data = trials)
```

```{r m2-summary-and-plot, fig.width=7, fig.height=4}
vc2      <- as.data.frame(VarCorr(m2))
stim_sd2 <- sqrt(vc2$vcov[vc2$grp == "stimulus_id"])
cat("Stimulus SD:", round(stim_sd2, 2), "(true =", dgp$sd_stimulus, ")\n")
cat("Transparency effect =", dgp$effect_transparency,
    "— product variation exceeds the effect we want to estimate.\n")

broom.mixed::tidy(m2, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m2: + stimulus RI", `True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)

ranef(m2)$stimulus_id |>
  as.data.frame() |> rownames_to_column("id") |> rename(blup = `(Intercept)`) |>
  ggplot(aes(x = reorder(id, blup), y = blup)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#888888") +
  geom_segment(aes(xend = id, y = 0, yend = blup),
               color = "#1e3a5f", linewidth = 0.5, alpha = 0.7) +
  geom_point(color = "#1e3a5f", size = 1.8) + coord_flip() +
  labs(title = paste0("Stimulus BLUPs: SD ≈ ", round(stim_sd2, 1),
                      " — larger than the transparency effect"),
       x = NULL, y = "Estimated baseline deviation (BLUP)") +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

**Result:** Estimate jumped from `r round(fixef(m1)["transparency"], 2)` to `r round(fixef(m2)["transparency"], 2)` (+`r round(fixef(m2)["transparency"] - fixef(m1)["transparency"], 2)` pts). **This is the largest single movement in the chapter.**

**Why the estimate jumped — two reasons:**

1. *Generalization.* Stimuli are now treated as a sample, not as fixed objects. The claim is about the population of products, not just these 36.

2. *Product-level confounding absorbed.* This is the more important reason for the estimate jump. Risk level is a stimulus-level variable: each product has one risk value, and all participants who see that product see the same risk level. Because transparent trials disproportionately involved high-risk products, the stimulus random intercept — which gives each product its own baseline — automatically absorbs the product attractiveness differences that were confounding the transparency estimate. The estimate jumped not because we added risk as an explicit covariate, but because we gave every product its own intercept.

**Why covariates are still needed:** The stimulus random intercept makes a flexible, non-parametric adjustment. Adding explicit risk covariates does three things the random intercept cannot: (1) makes the adjustment interpretable — "riskier products had lower adoption"; (2) extrapolates to products not in this sample using the estimated risk slope; (3) captures within-product nonlinear effects (a quadratic EV effect cannot be absorbed by a product-level intercept shift). These come in m4 and m4b.

::: {.callout-warning icon=false}
## Open Threats After m2: 4 of 7

**Reduced:** ~~Participant dependence~~, ~~Stimulus heterogeneity (and most product-level confounding)~~

**Still open:** residual confounding (nonlinear risk, fatigue), differential missingness, post-treatment exclusions, effect heterogeneity.

**Current estimate: `r round(fixef(m2)["transparency"], 2)` — major progress, not yet at truth.**
:::

---

## 10. Site Structure: Separate Context From Treatment

::: {.callout-note icon=false}
## At a Glance

- **Open threat:** Site-baseline confounding — site 2 has below-average adoption AND above-average transparency rate.
- **Model change:** Add `(1 | site_id)`.
- **What to watch:** Site-level means and transparency rates; small upward movement expected.
:::

```{r m3}
m3 <- lmer(adoption_intent_obs ~ transparency +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)
```

```{r m3-summary}
trials |>
  filter(!is.na(adoption_intent_obs)) |>
  group_by(site_id) |>
  summarise(
    mean_adoption   = round(mean(adoption_intent_obs, na.rm = TRUE), 1),
    pct_transparent = round(100 * mean(transparency), 1),
    n_part          = n_distinct(participant_id)
  ) |>
  knitr::kable(caption = "Site-level summary: site 2 has lower adoption AND higher transparency rate — a confound that site RI removes")

broom.mixed::tidy(m3, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m3: + site RI", `True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)
```

**Result:** Estimate moved from `r round(fixef(m2)["transparency"], 2)` to `r round(fixef(m3)["transparency"], 2)` (+`r round(fixef(m3)["transparency"] - fixef(m2)["transparency"], 2)` pts). Site 2's below-average baseline was partially pulling the transparent-condition mean down; partitioning it out removes that confound.

::: {.callout-warning icon=false}
## Open Threats After m3: 3 of 7

**Reduced:** ~~Participant dependence~~, ~~Stimulus heterogeneity/product confounding~~, ~~Site-baseline confounding~~

**Still open:** residual covariate misspecification (nonlinear risk, fatigue), differential missingness, post-treatment exclusions, effect heterogeneity.

**Current estimate: `r round(fixef(m3)["transparency"], 2)`.** The estimate is relatively close to truth but not yet there — nonlinear risk threshold and trial-order fatigue are the remaining sources of bias.
:::

---

## 11. Product Covariates: Make the Adjustment Interpretable

::: {.callout-note icon=false}
## At a Glance

- **Rationale:** The stimulus RI absorbed most product-level confounding at m2. What explicit risk/EV covariates add is *interpretability* and a basis for generalizing beyond this stimulus set.
- **Model change:** Add `risk_level`, `expected_value_z`, `financial_literacy` as fixed effects.
- **What to watch:** The estimate will **not move much** — the stimulus RI already did the heavy lifting. The value here is interpretability, not additional bias correction.
:::

```{r m4}
m4 <- lmer(adoption_intent_obs ~ transparency + risk_level + expected_value_z +
             financial_literacy +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)

broom.mixed::tidy(m4, effects = "fixed") |>
  filter(term == "transparency") |>
  mutate(model = "m4: + linear risk/EV/literacy", `True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  knitr::kable(digits = 3)
```

**Result:** Estimate moved from `r round(fixef(m3)["transparency"], 2)` to `r round(fixef(m4)["transparency"], 2)` (+`r round(fixef(m4)["transparency"] - fixef(m3)["transparency"], 2)` pts). **The movement is small** — as expected given the stimulus RI already absorbed most of the product-level confounding.

**What this model adds:**

- *Interpretability:* Instead of "some products have lower baselines," we can now say "riskier products had lower adoption and were overrepresented in the transparent condition."
- *Linear adjustment:* The product-level adjustment is now parametric (linear function of risk) rather than flexible (per-product BLUP).
- *Within-stimulus comparison:* The linear EV effect captures how expected value predicts adoption across products, after accounting for product random baselines.

::: {.callout-note icon=false}
## Lesson: Random Intercepts and Covariates Are Not Alternatives

The stimulus RI (m2) and the risk covariate (m4) did different work. The RI absorbed product-level confounding non-parametrically. The explicit covariate names the source of that confounding and adds linear structure. Neither replaces the other:

- Without the RI, the risk covariate would need to fully specify the product-level attractiveness function — it cannot.
- Without the covariate, the RI cannot tell us *why* products differ in their baselines or how to generalize.

The small estimate movement at m4 is not a failure. It is the expected pattern when a stimulus RI correctly absorbed the cluster-level confounding at m2.
:::

---

## 12. Functional Form and Fatigue: Fix the Remaining Bias

::: {.callout-note icon=false}
## At a Glance

- **Open threat:** The estimate is still below the true effect. Where is the remaining bias?
- **Evidence:** Residuals from m4 show a downward curve at high risk (threshold not captured by the linear term) and a downward trend across trial order (fatigue not controlled).
- **Model change:** Add `risk_high`, `I(expected_value_z^2)`, `trial_order`.
- **What to watch:** This should produce a **meaningful estimate movement** — the remaining bias is functional-form and fatigue confounding.
:::

```{r functional-form-diag, fig.width=9, fig.height=3.5}
#| code-summary: "Show residual diagnostic plots"
resid_m4 <- tibble(
  residual    = residuals(m4),
  risk_level  = model.frame(m4)$risk_level,
  ev_z        = model.frame(m4)$expected_value_z,
  trial_order = trials$trial_order[!is.na(trials$adoption_intent_obs)]
)

p_r <- ggplot(resid_m4, aes(x = risk_level, y = residual)) +
  geom_point(alpha = 0.08, color = "#1e3a5f") +
  geom_smooth(method = "loess", se = FALSE, color = "#e63946", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs risk", subtitle = "Dip at high risk = threshold not captured",
       x = "Risk level", y = "Residual") +
  theme_minimal(base_size = 10)

p_e <- ggplot(resid_m4, aes(x = ev_z, y = residual)) +
  geom_point(alpha = 0.08, color = "#1e3a5f") +
  geom_smooth(method = "loess", se = FALSE, color = "#e63946", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs EV", subtitle = "Curve = diminishing returns",
       x = "Expected value (z)", y = "Residual") +
  theme_minimal(base_size = 10)

p_t <- ggplot(resid_m4, aes(x = trial_order, y = residual, group = trial_order)) +
  geom_boxplot(fill = "#1e3a5f", alpha = 0.5, outlier.size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs trial order", subtitle = "Downward trend = unmodeled fatigue",
       x = "Trial", y = "Residual") +
  theme_minimal(base_size = 10)

gridExtra::grid.arrange(p_r, p_e, p_t, ncol = 3)
```

```{r m4b}
m4b <- lmer(adoption_intent_obs ~ transparency +
              risk_level + risk_high +
              expected_value_z + I(expected_value_z^2) +
              financial_literacy + trial_order +
              (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
            data = trials)

bind_rows(
  broom.mixed::tidy(m4,  effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m4: linear risk/EV, no fatigue"),
  broom.mixed::tidy(m4b, effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m4b: + risk_high + I(ev_z^2) + trial_order")
) |>
  mutate(`True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error, p.value) |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = "m4b: nonlinear risk threshold and trial-order fatigue remove the remaining downward bias")
```

**Result:** Estimate moved from `r round(fixef(m4)["transparency"], 2)` to `r round(fixef(m4b)["transparency"], 2)` (+`r round(fixef(m4b)["transparency"] - fixef(m4)["transparency"], 2)` pts). **This is the second large movement.** The bias removed here came from two within-product sources the stimulus RI could not absorb: (1) the sharp extra penalty for very-high-risk products (the `risk_high` threshold), and (2) the trial-order fatigue that was correlated with transparency assignment.

::: {.callout-tip icon=false}
## Open Threats After m4b: 2 of 7

**Reduced:** ~~Participant dependence~~, ~~Stimulus heterogeneity/product confounding~~, ~~Site confounding~~, ~~Covariate imbalance (explicit)~~, ~~Nonlinear valuation/fatigue~~

**Still open:** differential missingness, post-treatment exclusions, effect heterogeneity.

**Current estimate: `r round(fixef(m4b)["transparency"], 2)` — close to truth (`r dgp$effect_transparency`).** Remaining uncertainty is about *which data were observed* and *for whom* the effect holds.
:::

---

## 13. Missingness Sensitivity: Ask Which Rows We Lost

::: {.callout-note icon=false}
## At a Glance

- **Open threat:** Differential missingness — transparent trials are ~8 ppts more likely to be missing.
- **Evidence:** Miss rate by condition table in §6.2.
- **Model change:** Add `device_mobile` — the one missingness predictor not yet in the model.
- **What to watch:** If the estimate shifts, device-related differential dropout was influencing results.
:::

```{r m4c}
m4c <- lmer(
  adoption_intent_obs ~ transparency + risk_level + risk_high +
    expected_value_z + I(expected_value_z^2) +
    financial_literacy + trial_order + device_mobile +
    (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
  data = trials
)

bind_rows(
  broom.mixed::tidy(m4b, effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m4b: without device_mobile", n_obs = nobs(m4b)),
  broom.mixed::tidy(m4c, effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m4c: + device_mobile", n_obs = nobs(m4c))
) |>
  mutate(`True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, n_obs, estimate, std.error, p.value) |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = "Missingness sensitivity: does adding the device predictor change the estimate?")
```

```{r m4c-notes}
est_diff_miss <- round(fixef(m4c)["transparency"] - fixef(m4b)["transparency"], 3)
```

**Result:** The estimate shifted by `r est_diff_miss` points when device was added. The missingness audit showed large differential dropout (19.9% vs 11.6% by condition), which looked alarming. The observed-covariate sensitivity is `r ifelse(abs(est_diff_miss) < 0.1, "reassuring — after accounting for the predictors already in the model, device-related selection did not substantially move the estimate.", ifelse(abs(est_diff_miss) < 0.3, "modest — device-related selection was present but small relative to the main effect.", "meaningful — device-related selection influenced the estimate, and the adjusted version is more defensible."))` The remaining concern is MNAR: if latent adoption intent predicts dropout over and above all observed covariates, no covariate adjustment reaches it.

::: {.callout-tip icon=false}
## Open Threats After m4c: 1.5 of 7

**Reduced:** ~~Participant dependence~~, ~~Stimulus confounding~~, ~~Site confounding~~, ~~Covariate imbalance~~, ~~Nonlinear valuation/fatigue~~, ~~Observed differential missingness~~

**Partially open:** MNAR (unobserved latent-outcome dependence in missingness), post-treatment exclusions, effect heterogeneity.

**Current estimate: `r round(fixef(m4c)["transparency"], 2)`.**
:::

---

## 14. Attention Exclusions: Data Quality or Post-Treatment Selection?

::: {.callout-note icon=false}
## At a Glance

- **Open threat:** Attention failures are ~6.7 ppts higher in the transparent condition (post-treatment).
- **Evidence:** Attention-failure-by-condition table in §6.2.
- **Model change:** Exclude attention-failure rows; compare estimate to all-observed.
- **What to watch:** The *direction* of the shift and the *estimand implication* — not whether the shift is large.
:::

```{r m4d}
trials_no_fail <- trials |> filter(attention_fail == 0)
m4d <- lmer(
  adoption_intent_obs ~ transparency + risk_level + risk_high +
    expected_value_z + I(expected_value_z^2) +
    financial_literacy + trial_order + device_mobile +
    (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
  data = trials_no_fail
)

bind_rows(
  broom.mixed::tidy(m4c, effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m4c: all observed (incl. attention failures)",
           estimand = "All assigned — intention-to-treat", n_obs = nobs(m4c)),
  broom.mixed::tidy(m4d, effects = "fixed") |> filter(term == "transparency") |>
    mutate(model = "m4d: excluding attention failures",
           estimand = "Trials without processing failures", n_obs = nobs(m4d))
) |>
  mutate(`True effect` = dgp$effect_transparency) |>
  select(model, estimand, `True effect`, n_obs, estimate, std.error, p.value) |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = "Attention exclusion sensitivity: the estimand changes, not just the estimate")
```

```{r m4d-notes}
diff_attn <- round(fixef(m4d)["transparency"] - fixef(m4c)["transparency"], 3)
```

**Result:** Excluding attention failures `r ifelse(diff_attn > 0, "inflated", "deflated")` the estimate by `r abs(diff_attn)` points. The direction is consistent with post-treatment selection: removing trials where the transparent display was hardest to process removes the trials most likely to have low adoption intent *and* high attention-check difficulty in the transparent condition.

The point is not the magnitude. The point is the estimand: m4c estimates "effect for all participants assigned to transparent displays." m4d estimates "effect for trials where participants successfully processed the display." These are different scientific claims with different policy implications.

**Recommended practice:** Report m4c as the primary estimate. Report m4d as a sensitivity check with an explicit note that the estimand is narrower.

---

## 15. Literacy Moderation: Change the Scientific Claim

::: {.callout-note icon=false}
## At a Glance

- **Open threat:** Effect heterogeneity — the transparency effect likely differs by financial literacy.
- **Evidence:** Theory; the DGP includes a -1.5 interaction.
- **Model change:** Add `transparency × financial_literacy`.
- **What to watch:** The interaction term; simple slopes at low/average/high literacy. The scientific *claim* changes.
:::

```{r m5}
m5 <- lmer(adoption_intent_obs ~ transparency * financial_literacy +
             risk_level + risk_high +
             expected_value_z + I(expected_value_z^2) +
             trial_order + device_mobile +
             (1 | site_id) + (1 | participant_id) + (1 | stimulus_id),
           data = trials)

broom.mixed::tidy(m5, effects = "fixed") |>
  select(term, estimate, std.error, statistic, p.value) |>
  mutate(across(c(estimate, std.error, statistic, p.value), \(x) round(x, 3))) |>
  knitr::kable(caption = "m5: the transparency coefficient is now the effect at average literacy, not a universal constant")
```

```{r m5-simple-slopes}
b_tx  <- fixef(m5)["transparency"]
b_int <- fixef(m5)["transparency:financial_literacy"]
tibble(
  `Financial literacy` = c("Low (−1 SD)", "Average (0)", "High (+1 SD)"),
  `Estimated effect`   = round(c(b_tx + b_int * -1, b_tx, b_tx + b_int * 1), 2),
  `True DGP`           = c(dgp$effect_transparency + dgp$interaction_tx_lit * -1,
                             dgp$effect_transparency,
                             dgp$effect_transparency + dgp$interaction_tx_lit * 1)
) |>
  knitr::kable(caption = "Simple slopes: transparency benefits low-literacy participants most")
```

```{r m5-plot, fig.width=6.5, fig.height=4}
#| code-summary: "Show interaction plot code"
lit_vals <- seq(-2, 2, length.out = 100)
pred_df  <- tibble(
  financial_literacy = rep(lit_vals, 2),
  transparency       = rep(c(0L, 1L), each = 100),
  risk_level = 0, risk_high = 0L, expected_value_z = 0,
  trial_order = 6, device_mobile = 0L,
  participant_id = NA_integer_, site_id = NA_integer_, stimulus_id = NA_integer_
)
pred_df$predicted <- predict(m5, newdata = pred_df, re.form = NA)

pred_df |>
  mutate(Condition = ifelse(transparency == 1, "Transparent", "Non-transparent")) |>
  ggplot(aes(x = financial_literacy, y = predicted, color = Condition, linetype = Condition)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("#cccccc", "#1e3a5f")) +
  labs(title = "Transparency effect moderated by financial literacy",
       x = "Financial literacy (standardised)", y = "Predicted adoption intent",
       color = NULL, linetype = NULL) +
  theme_minimal(base_size = 11)
```

**Result:** The interaction is confirmed. After m5, the transparency "main effect" is the effect at average financial literacy — not a universal constant. This changes what researchers should say and which populations policy-makers should target.

---

## 16. Random Slopes: Allow the Effect to Vary Across People

::: {.callout-note icon=false}
## At a Glance

- **Open threat:** Treatment-effect heterogeneity beyond what the literacy interaction captures.
- **Evidence:** The slope histogram shows wide spread in empirical per-person effects.
- **Model change:** Add `(1 + transparency || participant_id)`.
- **What to watch:** The slope SD; whether the model is singular; SE change on transparency.
:::

```{r m6}
m6b <- lmer(adoption_intent_obs ~ transparency * financial_literacy +
              risk_level + risk_high +
              expected_value_z + I(expected_value_z^2) +
              trial_order + device_mobile +
              (1 + transparency || participant_id) +
              (1 | stimulus_id) + (1 | site_id),
            data = trials, control = lmerControl(optimizer = "bobyqa"))
cat("Singular (m6b uncorrelated slope):", isSingular(m6b), "\n")
```

```{r m6-slope-viz, fig.width=6.5, fig.height=3.5}
#| code-summary: "Show slope distribution plot"
emp_slopes <- trials |>
  filter(!is.na(adoption_intent_obs)) |>
  group_by(participant_id) |>
  summarise(
    mean_tx  = mean(adoption_intent_obs[transparency == 1], na.rm = TRUE),
    mean_ntx = mean(adoption_intent_obs[transparency == 0], na.rm = TRUE),
    n_tx = sum(transparency == 1), n_ntx = sum(transparency == 0)
  ) |>
  filter(n_tx > 0, n_ntx > 0) |>
  mutate(emp_effect = mean_tx - mean_ntx)

vc6b      <- as.data.frame(VarCorr(m6b))
slope_row <- vc6b[!is.na(vc6b$var1) & vc6b$var1 == "transparency" & is.na(vc6b$var2), ]
slope_sd_val <- if (nrow(slope_row) > 0) sqrt(slope_row$vcov[1]) else NA_real_

ggplot(emp_slopes, aes(x = emp_effect)) +
  geom_histogram(bins = 25, fill = "#1e3a5f", color = "white", alpha = 0.8) +
  geom_vline(xintercept = fixef(m6b)["transparency"],
             color = "#e63946", linetype = "dashed", linewidth = 1) +
  labs(title = "Participant-level transparency effects: the spread is real",
       subtitle = paste0("Red = model average. Slope SD ≈ ",
                         ifelse(is.na(slope_sd_val), "NA", round(slope_sd_val, 1)), " pts."),
       x = "Empirical transparency effect", y = "Count") +
  theme_minimal(base_size = 11)
```

```{r m6-comparison}
bind_rows(
  broom.mixed::tidy(m5,  effects = "fixed") |>
    filter(term %in% c("transparency","transparency:financial_literacy")) |>
    mutate(model = "m5: fixed slope"),
  broom.mixed::tidy(m6b, effects = "fixed") |>
    filter(term %in% c("transparency","transparency:financial_literacy")) |>
    mutate(model = "m6b: random slope")
) |>
  select(model, term, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) |>
  knitr::kable(caption = "Adding the random slope: SE widens; point estimate barely changes")
```

**Result:** Random slopes are not here to fix bias — they make heterogeneity explicit and make uncertainty more honest. The SE on the transparency effect widens slightly because the model now accounts for the fact that participants differ in their responsiveness, not just in their baselines.

::: {.callout-warning icon=false}
## If the Random Slope Is Singular

1. Drop the correlation parameter first (`||`) — already done here.
2. If still singular, slope variance may be negligible: retain m5 and report the simplification.
3. Never silently drop the slope without documenting the decision path.
:::

---

## 17. Select the Final Model

```{r final-model-selection}
slope_sd_check <- tryCatch({
  vc <- as.data.frame(VarCorr(m6b))
  tx <- vc[!is.na(vc$var1) & vc$var1 == "transparency" & is.na(vc$var2), ]
  if (nrow(tx) > 0) sqrt(tx$vcov[1]) else NA_real_
}, error = function(e) NA_real_)

keep_slope <- isTRUE(!isSingular(m6b)) &&
              isTRUE(!is.na(slope_sd_check)) &&
              isTRUE(slope_sd_check > 0.5)

m_final <- if (keep_slope) m6b else m5

cat("Slope SD:", if (is.na(slope_sd_check)) "NA" else round(slope_sd_check, 2), "\n")
cat("Singular:", isSingular(m6b), "\n")
cat("Final model:", if (keep_slope) "m6b (random slope)" else "m5 (no slope)", "\n")
```

```{r residual-diagnostics, fig.width=10, fig.height=3.5}
#| code-summary: "Show residual diagnostic plots"
trials_obs <- trials |> filter(!is.na(adoption_intent_obs))
resid_df   <- tibble(
  fitted   = fitted(m_final),
  residual = residuals(m_final),
  site_id  = trials_obs$site_id,
  risk     = trials_obs$risk_level
)

p_rv <- ggplot(resid_df, aes(x = fitted, y = residual)) +
  geom_point(alpha = 0.12, color = "#1e3a5f") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#e63946") +
  labs(title = "Residuals vs Fitted", x = "Fitted", y = "Residual") + theme_minimal(base_size = 10)

p_qq <- ggplot(resid_df, aes(sample = residual)) +
  stat_qq(alpha = 0.15, color = "#1e3a5f") + stat_qq_line(color = "#e63946") +
  labs(title = "QQ Plot", x = "Theoretical", y = "Sample") + theme_minimal(base_size = 10)

p_risk <- ggplot(resid_df, aes(x = risk, y = residual)) +
  geom_point(alpha = 0.08, color = "#1e3a5f") +
  geom_smooth(method = "loess", se = FALSE, color = "#e63946", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Risk (heteroskedasticity)", x = "Risk level", y = "Residual") +
  theme_minimal(base_size = 10)

gridExtra::grid.arrange(p_rv, p_qq, p_risk, ncol = 3)
```

---

## 18. Final Interpretation

```{r final-interp}
fe_final <- broom.mixed::tidy(m_final, effects = "fixed") |>
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 3)))
knitr::kable(fe_final |> select(term, estimate, std.error, statistic, p.value),
             caption = "Final model: fixed effects")

knitr::kable(
  as.data.frame(VarCorr(m_final)) |>
    select(grp, var1, vcov) |>
    mutate(SD = round(sqrt(vcov), 2), vcov = round(vcov, 2)),
  caption = "Final model: variance components"
)
```

**Transparency effect at average literacy:** `r round(fixef(m_final)["transparency"], 2)` (true DGP: `r dgp$effect_transparency`). The naive estimate was `r round(coef(m0)["transparency"], 2)` — a difference of `r round(fixef(m_final)["transparency"] - coef(m0)["transparency"], 2)` points. That gap was earned through a diagnostic sequence, not through modeling choices made in advance.

**Where the bias came from (in order of removal):**

1. Product-level attractiveness confounding — absorbed by stimulus random intercepts (m2, the largest single correction)
2. Nonlinear risk threshold and trial-order fatigue — removed by explicit nonlinear terms (m4b)
3. Site-baseline confounding — removed by site random intercept (m3, small correction)

**What the final model still depends on:**

- Missingness is MAR conditional on observed covariates (MNAR cannot be tested)
- Attention exclusions do not condition on the relevant subpopulation for the stated estimand
- Four sites are enough to represent the population of deployment contexts
- The stimulus sample is representative of the intended product type

---

## 19. Model Comparison: Actual Estimates and Movements

```{r movement-table}
#| code-summary: "Show estimate-movement computation"
# Store all estimates in a named vector for the movement table
est_seq <- c(
  m0 = coef(m0)["transparency"],
  m1 = fixef(m1)["transparency"],
  m2 = fixef(m2)["transparency"],
  m3 = fixef(m3)["transparency"],
  m4 = fixef(m4)["transparency"],
  m4b = fixef(m4b)["transparency"],
  m4c = fixef(m4c)["transparency"],
  m4d = fixef(m4d)["transparency"],
  m5 = fixef(m5)["transparency"]
)

tibble(
  Model = c("m0: pooled OLS", "m1: + participant RI", "m2: + stimulus RI",
            "m3: + site RI", "m4: + linear risk/EV",
            "m4b: + nonlinear risk + fatigue", "m4c: + device (miss. sensitivity)",
            "m4d: excl. attention failures", "m5: + literacy interaction"),
  `n obs` = c(nobs(m0), nobs(m1), nobs(m2), nobs(m3), nobs(m4), nobs(m4b), nobs(m4c), nobs(m4d), nobs(m5)),
  `Estimate` = round(est_seq, 3),
  `Change from previous` = round(c(NA, diff(est_seq)), 3),
  `Threat addressed` = c(
    "None — baseline",
    "Participant dependence",
    "Stimulus heterogeneity + product-level confounding",
    "Site-baseline confounding",
    "Product covariates (interpretability)",
    "Nonlinear risk threshold + trial-order fatigue",
    "Device-related missingness",
    "Post-treatment exclusion (estimand check)",
    "Literacy moderation"
  ),
  `Key lesson` = c(
    "Estimate is substantially below truth — bias is real",
    "SE changes; estimate barely moves — RI fixes dependence, not bias",
    "**Largest jump** — product RI absorbed stimulus-level confounding",
    "Small movement — site-baseline removed",
    "Small movement — stimulus RI did most of this work at m2",
    "**Second meaningful jump** — functional form was part of the bias",
    "Small shift — observed MAR adjustment reassuring",
    "Estimate shifts direction — estimand narrows",
    "Claim changes to conditional effect"
  )
) |>
  knitr::kable(caption = paste0(
    "Estimate movement table: actual before/after values at each step. ",
    "True effect at average literacy = ", dgp$effect_transparency, "."))
```

---

## 20. What We Fixed vs What We Did Not Fix

### What the analysis substantially reduced

- **Row dependence** — participant and stimulus random intercepts correctly handle clustering.
- **Product-level assignment imbalance** — stimulus random intercepts absorbed the largest part of the product-risk confounding at m2; explicit covariates made it interpretable at m4.
- **Functional-form misspecification** — nonlinear risk and EV terms corrected the remaining bias from the risk threshold and diminishing returns.
- **Trial-order fatigue** — adding trial order removed confounding from transparent trials occurring later.
- **Observed differential missingness** — the device covariate adjusted for an additional missingness predictor.

### What the analysis did not fully solve

- **MNAR** — if latent adoption intent predicts dropout beyond all observed covariates, no adjustment in this chapter addresses it. The sensitivity was reassuring; the assumption is still an assumption.
- **Post-treatment selection** — the attention-exclusion sensitivity showed that excluding failures narrows the estimand. The primary model does not exclude; the sensitivity model does. Neither is definitively correct.
- **Limited site generalization** — with four sites, the site random intercept is estimated with minimal precision. Treatment effect heterogeneity across sites cannot be reliably detected.
- **Heteroskedasticity** — residuals fan with risk level, violating the LMM assumption of constant variance. For high-risk stimuli, confidence intervals may be too narrow.

### The honest final claim

> After diagnosing and removing the major observed threats — product-level assignment imbalance, nonlinear covariate effects, and trial-order fatigue — fee transparency appears to increase adoption intent by approximately `r round(fixef(m_final)["transparency"], 1)` points at average financial literacy (true DGP: `r dgp$effect_transparency`), with larger benefits for lower-literacy participants. This estimate is meaningfully more defensible than the naive result of `r round(coef(m0)["transparency"], 1)`, but it still depends on assumptions about missingness, post-treatment exclusions, and the representativeness of stimuli and sites.

---

## 21. What a Reviewer Would Ask

**1. "The transparency coefficient predicting assignment is 0.74 in your audit. Was this truly randomized?"**

No — not perfectly. The transparency display was served more often to high-risk products and later trial positions. We documented this in the assignment audit and removed it by (a) stimulus random intercepts, which absorbed the product-level confounding, and (b) nonlinear risk and trial-order covariates, which captured the within-stimulus residual. The estimate moved substantially as a result.

---

**2. "You say stimulus random intercepts absorbed the product-level confounding. Isn't that circularity?"**

No. Risk level predicts both transparency assignment and adoption intent. The stimulus random intercept gives each product a flexible baseline — whatever predicts the product-level attractiveness, including risk, is absorbed. We then added explicit risk covariates to make that adjustment interpretable and to handle within-product nonlinear effects. The combination is more informative than either alone: the RI handles flexible product baselines; the covariates name the mechanism.

---

**3. "Your m4 estimate barely moved from m3. Why call it a useful step?"**

Because it serves a different purpose. The product-level adjustment was already handled by the stimulus RI at m2. m4's contribution is interpretability — we now know that riskier products drove most of the product-level attractiveness differences, and we have an estimate of the linear risk slope. That matters for generalizing beyond this sample of 36 products. A model with only a stimulus RI cannot extrapolate; a model with explicit risk covariates can.

---

**4. "16% of outcomes are missing. Your sensitivity analysis moved the estimate by only 0.02. Did missingness really not matter?"**

Missingness looked alarming in the audit. The diagnostic sensitivity (adding `device_mobile`) was small, suggesting that the missingness predictors already in the model (risk, trial order, literacy) accounted for most of the differential dropout. The remaining concern is MNAR: if latent adoption intent predicts dropout above and beyond all observed predictors, no covariate adjustment reaches it. We flag this as a caveat and note that the small observed-sensitivity shift is reassuring but not conclusive.

---

**5. "Why not exclude attention failures? They indicate low data quality."**

Attention failures are 2.5× more common in the transparent condition (12.1% vs 5.4%). If we exclude them, we are preferentially removing trials where the transparent display was most cognitively demanding — which may be the exact trials where the treatment effect is most muted. Exclusion conditions on a post-treatment variable and produces a different estimand: "effect for trials participants processed without difficulty" rather than "effect of being assigned a transparent display." Both estimands are defensible, but they should not be conflated.

---

**6. "The stimulus SD is about `r round(sqrt(as.data.frame(VarCorr(m2))$vcov[as.data.frame(VarCorr(m2))$grp == "stimulus_id"]), 1)` — larger than the transparency effect. Does your generalization claim hold?"**

Yes, and this is exactly why the stimulus random intercept is essential. If we had treated stimuli as fixed, the SE on the transparency effect would be too small (product-level variance absorbed into the residual inflates false precision). The random intercept correctly partitions product variance from the within-product treatment effect. The large product SD does mean that the specific stimulus sample matters — a very different set of 36 products could give a different average effect. That is an honest limitation of any single-study result.

---

**7. "What is your estimand?"**

The average total effect of *assigned* fee transparency on willingness to adopt, averaged over the population of participants, product offers, and contexts similar to those sampled, at average financial literacy. At other literacy levels, the simple-slope table gives the conditional estimate. This is an ITT estimand — comprehension and perceived transparency are mechanisms, not adjustments.

---

**8. "For whom does the effect hold?"**

The literacy interaction (m5) shows that transparency benefits lower-literacy participants more. The random slope (m6b) shows that there is also heterogeneity across participants beyond what literacy captures. With four sites and 36 stimuli, the boundaries of generalization are modest: similar populations, similar financial product types, similar deployment contexts.

---

## 22. Reporting Template

> We analyzed adoption-intent ratings (0–100) using linear mixed-effects models (lme4; Bates et al., 2015) with Satterthwaite degrees of freedom (lmerTest).
>
> **Assignment audit.** Pre-analysis diagnostics revealed implementation imbalance: the transparent-display condition was substantially more common for higher-risk products (logistic coefficient = [X]) and later trial positions (coefficient = [X]). All subsequent models address this imbalance through stimulus random intercepts and explicit risk/fatigue covariates.
>
> **Data retention.** Of [N] intended observations, [X]% had missing outcomes. Missingness was higher in the transparent condition ([X]% vs [X]%). We used complete-case analysis and checked sensitivity by adding the main device-based missingness predictor; the estimate shifted by [X] points. Results with and without attention-check exclusions are reported separately with explicit estimand notes.
>
> **Model specification.** Fixed effects: transparency, financial literacy, their interaction, risk level, high-risk indicator, standardised expected value (linear + quadratic), trial order, device platform. Random intercepts for participants, stimuli, and sites. By-participant random slope for transparency ([retained / dropped after singularity check]).
>
> **Results.** Fee transparency increased adoption intent by *b* = [X] (95% CI [X, X]) at average financial literacy, down from a naive estimate of [X] that was confounded with product-risk and trial-order imbalance. Financial literacy moderated the effect (*b* = [X]).
>
> **Limitations.** Missingness may depend on unobserved latent adoption intent (MNAR). Attention-failure exclusions condition on a post-treatment variable and narrow the estimand. With four sites and 36 stimuli, stimulus and site generalization is bounded.

::: {.callout-tip icon=false}
## Minimum Reproducible Model Report

- [ ] Assignment audit: treatment balanced on pre-treatment variables?
- [ ] Data flow: intended → observed → dropped → excluded
- [ ] Missingness by condition, site, trial order
- [ ] Post-treatment timing audit: which variables excluded and why
- [ ] Model formula (verbatim) with optimizer and convergence status
- [ ] Fixed-effect estimates with SEs and CIs
- [ ] Random-effect variance components (SDs)
- [ ] Functional-form residual diagnostics
- [ ] Missingness sensitivity comparison
- [ ] Exclusion sensitivity with estimand note
- [ ] Explicit estimand statement
- [ ] "What we fixed vs what we did not fix" summary
:::

---

## Appendix: Stress Tests

### Stress Test A: Severe Risk Imbalance

```{r stress-a}
#| code-summary: "Show stress test A code"
set.seed(101)
trials_stress_a <- trials |>
  left_join(stimuli  |> select(stimulus_id, u0_stimulus),  by = "stimulus_id") |>
  left_join(participants |> select(participant_id, u0_participant, u1_participant), by = "participant_id") |>
  left_join(site_effects, by = "site_id") |>
  mutate(
    trial_order_z_a = (trial_order - 6.5) / 3.452,
    p_tx_imb = plogis(1.40 * risk_level + 0.60 * trial_order_z_a + 0.80 * as.integer(site_id == 2)),
    tx_imb   = rbinom(n(), 1, p_tx_imb),
    mu_imb   = dgp$intercept +
               dgp$effect_transparency * tx_imb +
               dgp$effect_risk * risk_level + dgp$effect_risk_high * risk_high +
               dgp$effect_ev_linear * expected_value_z + dgp$effect_ev_quad * expected_value_z^2 +
               dgp$effect_literacy * financial_literacy +
               dgp$interaction_tx_lit * tx_imb * financial_literacy +
               dgp$effect_fatigue * trial_order + dgp$effect_fatigue_late * as.integer(trial_order > 9) +
               u0_participant + u1_participant * tx_imb + u0_stimulus + u0_site,
    y_imb = pmin(100, pmax(0, rnorm(n(), mu_imb, dgp$sd_residual)))
  )

m_imb_raw <- lmer(y_imb ~ tx_imb + (1 | participant_id) + (1 | stimulus_id), data = trials_stress_a)
m_imb_adj <- lmer(y_imb ~ tx_imb + risk_level + risk_high + expected_value_z + trial_order +
                    (1 | participant_id) + (1 | stimulus_id), data = trials_stress_a)

bind_rows(
  broom.mixed::tidy(m_imb_raw, effects = "fixed") |> filter(term == "tx_imb") |>
    mutate(model = "Stimulus RI only (no covariates)"),
  broom.mixed::tidy(m_imb_adj, effects = "fixed") |> filter(term == "tx_imb") |>
    mutate(model = "Stimulus RI + risk covariates")
) |>
  mutate(`True effect` = dgp$effect_transparency) |>
  select(model, `True effect`, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) |>
  knitr::kable(caption = "Stress A: with imbalance coefficient doubled, stimulus RI alone leaves residual bias; covariates complete the correction")
```

### Stress Test B: Site-Level Treatment Heterogeneity

```{r stress-b}
#| code-summary: "Show stress test B code"
set.seed(102)
site_effects_b <- site_effects |> mutate(u1_site = c(1.5, -1.0, 0.5, -1.0))
trials_stress_b <- trials |>
  left_join(stimuli  |> select(stimulus_id, u0_stimulus),  by = "stimulus_id") |>
  left_join(participants |> select(participant_id, u0_participant, u1_participant), by = "participant_id") |>
  left_join(site_effects_b, by = "site_id") |>
  mutate(
    mu_b = dgp$intercept + (dgp$effect_transparency + u1_site) * transparency +
           dgp$effect_risk * risk_level + dgp$effect_risk_high * risk_high +
           dgp$effect_ev_linear * expected_value_z + dgp$effect_ev_quad * expected_value_z^2 +
           dgp$effect_literacy * financial_literacy +
           dgp$effect_fatigue * trial_order + dgp$effect_fatigue_late * as.integer(trial_order > 9) +
           u0_participant + u1_participant * transparency + u0_stimulus + u0_site,
    y_b = pmin(100, pmax(0, rnorm(n(), mu_b, dgp$sd_residual)))
  )

broom.mixed::tidy(
  lmer(y_b ~ transparency * factor(site_id) + risk_level + risk_high + expected_value_z + trial_order +
         (1 | participant_id) + (1 | stimulus_id), data = trials_stress_b),
  effects = "fixed"
) |>
  filter(str_detect(term, "transparency")) |>
  select(term, estimate, std.error) |>
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) |>
  knitr::kable(caption = "Stress B: transparency effect varies by ±1.5 pts across sites. A site RI hides this; site interaction reveals it.")
```

### Stress Test C: Sparse Within-Person Data

```{r stress-c}
set.seed(103)
trials_sparse <- trials |> group_by(participant_id) |> slice_sample(n = 4) |> ungroup()
m_sparse <- lmer(adoption_intent_obs ~ transparency * financial_literacy +
                   risk_level + expected_value_z +
                   (1 + transparency | participant_id) +
                   (1 | stimulus_id) + (1 | site_id),
                 data = trials_sparse, control = lmerControl(optimizer = "bobyqa"))
cat("Singular (4 trials per person):", isSingular(m_sparse), "\n")
cat("The design justifies the slope; 4 trials per person cannot estimate it reliably.\n")
```
