---
title: "Partially Nested Designs"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
set.seed(2025)
```

# Partially Nested Designs

## When Only One Condition Has Clustering

Consider a financial coaching field experiment. The treatment group receives group financial coaching: 8 participants per coach, 15 coaches. The control group receives a self-serve information module — there is no coach and no group structure. Participants in the control condition are independent of each other.

This design is **partially nested**: the treatment condition has a natural cluster structure (participants within coaching groups), but the control condition does not.

If you fit a standard two-level model treating all participants the same way, you either:
1. Ignore clustering in the treatment group — deflating standard errors and inflating Type I error for the treatment effect
2. Apply random effects to all participants — misspecifying the control group, which has no actual clustering

The **partially nested model** (Bauer, Sterba & Hallfors, 2008) handles this correctly by allowing different variance structures in different conditions.

## Simulating a Partially Nested Design

```{r pn-simulation}
n_coaches    <- 15
n_per_group  <- 8
n_control    <- n_coaches * n_per_group  # same total N = 120

# Treatment group: coach clustering
coach_effects <- rnorm(n_coaches, 0, 3)  # ICC ≈ 0.15 from coach variance

df_treatment <- tibble(
  condition   = "coaching",
  coach       = rep(1:n_coaches, each = n_per_group),
  participant = 1:(n_coaches * n_per_group),
  coach_eff   = rep(coach_effects, each = n_per_group),
  savings_increase = 5 + 5 + coach_eff + rnorm(n_coaches * n_per_group, 0, 6)
  # baseline mean = 5, treatment adds 5, coach effect varies
)

# Control group: independent participants (no coach clustering)
df_control <- tibble(
  condition   = "self_serve",
  coach       = NA_integer_,
  participant = 1:n_control + max(df_treatment$participant),
  coach_eff   = 0,
  savings_increase = 5 + rnorm(n_control, 0, 6)
)

df_pn <- bind_rows(df_treatment, df_control)
```

## The Partially Nested Model in lme4

The trick: create a dummy coach variable for control participants (unique ID for each, indicating no shared clustering) and fit a model where the random effect only applies to the treatment group.

```{r fit-pn}
# Assign unique cluster IDs: coaching groups for treatment, individual IDs for control
df_pn <- df_pn |>
  mutate(
    cluster_id = ifelse(condition == "coaching",
                        paste0("Coach", coach),
                        paste0("Ctrl", participant))
  )

# Partially nested: cluster RE only in coaching condition
df_pn <- df_pn |>
  mutate(treat_dummy = as.integer(condition == "coaching"))

# Fit: allow random effect only within coaching clusters
m_pn    <- lmer(savings_increase ~ condition + (1 | cluster_id), data = df_pn)

# Compare to naive model (ignores all clustering)
m_naive <- lm(savings_increase ~ condition, data = df_pn)

bind_rows(
  broom::tidy(m_naive) |> filter(term == "conditioncoaching") |>
    mutate(model = "Naive (no clustering)"),
  broom.mixed::tidy(m_pn, effects = "fixed") |>
    filter(term == "conditioncoaching") |>
    mutate(model = "Partially nested model")
) |>
  select(model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Partially nested model: correct SE for coaching-group clustering")
```

Participants in the coaching condition share a coach. Their savings outcomes are not independent: a particularly effective (or ineffective) coach shifts all their clients' outcomes together. Participants in the control condition have no such shared structure — each receives the same information independently.

The naive model ignores this clustering in the coaching arm, producing standard errors that are too small and p-values that are too liberal. The partially nested model correctly inflates the SE for the coaching condition, where observations are non-independent, while leaving the control group's independent structure intact.

::: {.callout-warning icon=false}
## Assumption: Correct Identification of Which Condition is Clustered

**What it means:** You must correctly specify which condition has clustering and which does not. If clustering is misattributed (e.g., you apply cluster random effects to the self-serve control group when there is none), variance components will be estimated incorrectly.

**How to check:** Draw the design. Which participants share a natural group, coach, or classroom? That condition has the nested structure. In this experiment, only participants in the group coaching condition share a coach — control participants each received the same self-serve materials independently.
:::

::: {.callout-warning icon=false}
## Assumption: Homogeneity of Within-Cluster Variance

**What it means:** The standard partially nested model assumes the residual variance is the same in the coaching and self-serve groups (only the random effect variance differs).

**How to check:** Compare within-group variance across conditions. If the coaching group has substantially more or less residual variance than the self-serve control, allow heteroskedastic residuals.
:::
