---
title: "Power Analysis for Nested Designs"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(ggplot2)
set.seed(42)
```

# Power Analysis for Nested Designs

## Why Standard Power Formulas Are Wrong

Standard power formulas (Cohen's power tables, G\*Power) assume independent observations. They compute: given my expected effect size and desired power, how many participants do I need?

In nested data, this question is incomplete. You also need to know:
- **How many clusters?** (classrooms, firms, sites, therapists)
- **How many observations per cluster?**
- **What is the ICC?** (how similar are observations within clusters)

The ICC and cluster size together determine the **design effect** (DEFF), which shrinks your effective sample size:

$$N_{effective} = \frac{N}{1 + (n_{cluster} - 1) \times \text{ICC}}$$

Ignoring this leads to severely underpowered studies.

## The ICC-Power Relationship

```{r icc-power-curve}
# Power as a function of ICC, for a fixed N = 300 in clusters of 15
compute_power <- function(icc, n_total = 300, cluster_size = 15, effect = 0.3, alpha = 0.05) {
  deff    <- 1 + (cluster_size - 1) * icc
  n_eff   <- n_total / deff
  # Two-sample t-test power with effective N
  power.t.test(n = n_eff / 2, delta = effect, sd = 1, sig.level = alpha)$power
}

icc_vals <- seq(0, 0.5, by = 0.01)
power_vals <- sapply(icc_vals, compute_power)

tibble(ICC = icc_vals, Power = power_vals) |>
  ggplot(aes(x = ICC, y = Power)) +
  geom_line(color = "#1e3a5f", linewidth = 1.3) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "#e63946") +
  annotate("text", x = 0.35, y = 0.82, label = "80% power threshold",
           color = "#e63946", size = 3.5) +
  labs(
    title    = "Power declines as ICC increases (N = 300, cluster size = 15)",
    subtitle = "Even modest ICC substantially reduces effective power",
    x        = "ICC",
    y        = "Statistical Power"
  ) +
  theme_minimal(base_size = 13)
```

At ICC = 0, you have the expected power for 300 independent observations. At ICC = 0.2, power drops below 80% — you would need more observations to compensate.

## Simulation-Based Power Analysis

For complex nested designs, analytic formulas do not always exist. The gold standard is **simulation-based power analysis**: generate data from the assumed DGP, fit the model, check whether the effect is significant, repeat 1000 times.

```{r sim-power}
simulate_power <- function(n_clusters, n_per_cluster, icc,
                            true_effect = 0.3, n_sims = 500) {
  sigma_between <- sqrt(icc)
  sigma_within  <- sqrt(1 - icc)

  p_values <- replicate(n_sims, {
    # Simulate two groups of clusters
    n_c <- ceiling(n_clusters / 2)
    cluster_effects <- rnorm(n_clusters, 0, sigma_between)

    df <- tibble(
      cluster = rep(1:n_clusters, each = n_per_cluster),
      group   = ifelse(cluster <= n_c, 0, 1)
    ) |>
      mutate(
        Y = true_effect * group + cluster_effects[cluster] +
            rnorm(n(), 0, sigma_within)
      )

    m <- lmer(Y ~ group + (1 | cluster), data = df)
    coef(summary(m))["group", "Pr(>|t|)"]
  })

  mean(p_values < 0.05)
}

# Power across different designs
designs <- expand_grid(
  n_clusters      = c(10, 20, 30, 40),
  n_per_cluster   = c(10, 20),
  icc             = c(0.05, 0.15)
) |>
  mutate(
    power = pmap_dbl(
      list(n_clusters, n_per_cluster, icc),
      ~simulate_power(..1, ..2, ..3, n_sims = 200)  # reduce for speed
    ),
    total_N = n_clusters * n_per_cluster,
    design  = paste0("ICC=", icc, ", n/cluster=", n_per_cluster)
  )

ggplot(designs, aes(x = n_clusters, y = power, color = design, linetype = design)) +
  geom_line(linewidth = 1.1) + geom_point(size = 2.5) +
  geom_hline(yintercept = 0.80, linetype = "dotted", color = "gray50") +
  scale_color_brewer(palette = "Set2") +
  labs(
    title    = "Power by number of clusters (simulated)",
    subtitle = "Effect size d = 0.3; n_sims = 200 per design",
    x        = "Number of Clusters",
    y        = "Power",
    color    = "Design", linetype = "Design"
  ) +
  theme_minimal(base_size = 12)
```

## The `simr` Package for Complex Designs

For multilevel models with random slopes, crossed random effects, or other complex structures, the `simr` package provides a convenient simulation-based power analysis directly from a fitted model:

```{r simr-demo, eval=FALSE}
library(simr)

# Fit a model with your expected parameters
m_pilot <- lmer(Y ~ group + (1 | cluster),
                data = your_pilot_data)

# Change the effect size to your expected value
fixef(m_pilot)["group"] <- 0.3

# Simulate power for different sample sizes
power_curve <- powerCurve(m_pilot,
                           along = "cluster",   # vary number of clusters
                           nsim  = 200)
plot(power_curve)
```

## Key Takeaways for Power Planning

::: {.callout-note}
## Rules of Thumb

1. **More clusters > more per cluster** when ICC is high. If ICC = 0.20, doubling the number of clusters gains much more power than doubling the size of each cluster.

2. **The ICC threshold:** If ICC > 0.10 and cluster size > 10, standard power formulas underestimate required N by 50–100%.

3. **For crossed designs (participants × stimuli):** Power depends on both the number of participants *and* the number of stimuli. Westfall, Kenny & Judd (2014) showed that many published studies are severely underpowered on the stimulus side.

4. **Collect pilot ICC estimates:** Run a small pilot study and compute the ICC before committing to a full design.
:::
