---
title: "Power Analysis for Nested Designs"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(ggplot2)
set.seed(42)
```

# Power Analysis for Nested Designs

## A Practical Power Planning Workflow

Power analysis is not a single calculation — it is a design exploration. The goal is to understand how power changes across the design space so you can make an informed choice, not to produce a single "magic number."

::: {.callout-note icon=false}
## Power Planning Workflow

1. **Define the estimand** — what is the effect you are testing? (average treatment effect, ITT, LATE, etc.)
2. **Identify the assignment level** — is treatment assigned to individuals, clusters, or sites?
3. **Identify the analysis model** — the power simulation must fit the *exact* planned analysis model
4. **Estimate plausible ICCs** from pilot data, the literature, or a range of assumptions
5. **Specify plausible effect sizes** — express uncertainty in the assumed effect; simulate a range
6. **Simulate several designs** — vary N, clusters, stimuli, and compliance rate
7. **Report uncertainty** — power is a range across plausible assumptions, not a single number
8. **Document inputs** for the preregistration: assumed ICC, assumed effect size, attrition rate, compliance rate, simulation code or package, number of replications
:::

| Design lever | What it controls | When to prioritize it |
|-------------|-----------------|----------------------|
| Number of participants | Within-cluster precision | When ICC is low and clusters are large |
| Number of clusters | Between-cluster variance | When ICC is high — adding clusters > adding per-cluster observations |
| Number of stimuli | Stimulus-level variance | When stimulus RE variance is large (Part 1 designs) |
| Number of sites | Site heterogeneity estimation | When multi-site generalization is the goal |
| Waves / time points | Longitudinal power | For growth curves, more waves help more than more participants |
| Compliance rate | First-stage strength | Affects LATE precision; low compliance → wide LATE SE |
| Attrition rate | Effective N | Even 20% attrition substantially reduces power |

::: {.callout-warning icon=false}
## Simulation Quality Warning

Simulation-based power is only as good as the assumed DGP. If your ICC assumption is off by a factor of 2, your power estimate can be dramatically wrong. Best practice:

- Run sensitivity analyses across a plausible ICC range (e.g., 0.05, 0.10, 0.20)
- Run sensitivity analyses across a plausible effect size range (small, medium, expected)
- Report a **power curve** or **power surface**, not a single number
- State explicitly which assumptions are uncertain and how sensitive power is to them
:::

---

## Why Standard Power Formulas Are Wrong

Standard power formulas (Cohen's power tables, G\*Power) assume independent observations. They compute: given my expected effect size and desired power, how many participants do I need?

In nested data, this question is incomplete. You also need to know:
- **How many clusters?** (classrooms, firms, sites, therapists)
- **How many observations per cluster?**
- **What is the ICC?** (how similar are observations within clusters)

The ICC and cluster size together determine the **design effect** (DEFF), which shrinks your effective sample size:

$$N_{effective} = \frac{N}{1 + (n_{cluster} - 1) \times \text{ICC}}$$

Ignoring this leads to severely underpowered studies.

## The ICC-Power Relationship

```{r icc-power-curve}
# Power as a function of ICC, for a fixed N = 300 in clusters of 15
compute_power <- function(icc, n_total = 300, cluster_size = 15, effect = 0.3, alpha = 0.05) {
  deff    <- 1 + (cluster_size - 1) * icc
  n_eff   <- n_total / deff
  # Two-sample t-test power with effective N
  power.t.test(n = n_eff / 2, delta = effect, sd = 1, sig.level = alpha)$power
}

icc_vals <- seq(0, 0.5, by = 0.01)
power_vals <- sapply(icc_vals, compute_power)

tibble(ICC = icc_vals, Power = power_vals) |>
  ggplot(aes(x = ICC, y = Power)) +
  geom_line(color = "#1e3a5f", linewidth = 1.3) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "#e63946") +
  annotate("text", x = 0.35, y = 0.82, label = "80% power threshold",
           color = "#e63946", size = 3.5) +
  labs(
    title    = "Power declines as ICC increases (N = 300, cluster size = 15)",
    subtitle = "Even modest ICC substantially reduces effective power",
    x        = "ICC",
    y        = "Statistical Power"
  ) +
  theme_minimal(base_size = 13)
```

At ICC = 0, you have the expected power for 300 independent observations. At ICC = 0.2, power drops below 80% — you would need more observations to compensate.

::: {.callout-warning icon=false}
## Design-Effect Conversion Is an Approximation

The design-effect formula `DEFF = 1 + (m − 1) × ICC` assumes equal cluster sizes and that assignment is at the cluster level. It is a useful heuristic for back-of-envelope planning, but it can mislead in several common situations:

- **Unequal cluster sizes:** Power is lower than the equal-size formula predicts when cluster sizes vary substantially.
- **Assignment below the cluster level:** In within-subjects designs (participants × stimuli), DEFF does not straightforwardly apply — you need to account for the stimulus-level random effect separately.
- **Multiple random effects:** Three-level designs have two ICCs; collapsing to a single design effect loses information.

**For decision-quality power planning** — grant applications, pre-registrations, trial designs — use simulation under your exact assignment unit, cluster structure, and analysis model (see the simulation section below). The design-effect shortcut is a starting point, not a final answer.
:::

## Simulation-Based Power Analysis

For complex nested designs, analytic formulas do not always exist. The gold standard is **simulation-based power analysis**: generate data from the assumed DGP, fit the model, check whether the effect is significant, repeat 1000 times.

```{r sim-power}
simulate_power <- function(n_clusters, n_per_cluster, icc,
                            true_effect = 0.3, n_sims = 500) {
  sigma_between <- sqrt(icc)
  sigma_within  <- sqrt(1 - icc)

  p_values <- replicate(n_sims, {
    # Simulate two groups of clusters
    n_c <- ceiling(n_clusters / 2)
    cluster_effects <- rnorm(n_clusters, 0, sigma_between)

    df <- tibble(
      cluster = rep(1:n_clusters, each = n_per_cluster),
      group   = ifelse(cluster <= n_c, 0, 1)
    ) |>
      mutate(
        Y = true_effect * group + cluster_effects[cluster] +
            rnorm(n(), 0, sigma_within)
      )

    tryCatch({
      m <- lmer(Y ~ group + (1 | cluster), data = df, REML = FALSE)
      coef(summary(m))["group", "Pr(>|t|)"]
    }, error = function(e) NA_real_,
       warning = function(w) {
         m <- lmer(Y ~ group + (1 | cluster), data = df, REML = FALSE)
         tryCatch(coef(summary(m))["group", "Pr(>|t|)"],
                  error = function(e) NA_real_)
       })
  })
  p_values <- p_values[!is.na(p_values)]

  mean(p_values < 0.05)
}

# Power across different designs
designs <- expand_grid(
  n_clusters      = c(10, 20, 30, 40),
  n_per_cluster   = c(10, 20),
  icc             = c(0.05, 0.15)
) |>
  mutate(
    power = pmap_dbl(
      list(n_clusters, n_per_cluster, icc),
      ~simulate_power(..1, ..2, ..3, n_sims = 200)  # reduce for speed
    ),
    total_N = n_clusters * n_per_cluster,
    design  = paste0("ICC=", icc, ", n/cluster=", n_per_cluster)
  )

ggplot(designs, aes(x = n_clusters, y = power, color = design, linetype = design)) +
  geom_line(linewidth = 1.1) + geom_point(size = 2.5) +
  geom_hline(yintercept = 0.80, linetype = "dotted", color = "gray50") +
  scale_color_brewer(palette = "Set2") +
  labs(
    title    = "Power by number of clusters (simulated)",
    subtitle = "Effect size d = 0.3; n_sims = 200 per design",
    x        = "Number of Clusters",
    y        = "Power",
    color    = "Design", linetype = "Design"
  ) +
  theme_minimal(base_size = 12)
```

## The `simr` Package for Complex Designs

For multilevel models with random slopes, crossed random effects, or other complex structures, the `simr` package provides a convenient simulation-based power analysis directly from a fitted model:

```{r simr-demo, eval=FALSE}
library(simr)

# Fit a model with your expected parameters
m_pilot <- lmer(Y ~ group + (1 | cluster),
                data = your_pilot_data)

# Change the effect size to your expected value
fixef(m_pilot)["group"] <- 0.3

# Simulate power for different sample sizes
power_curve <- powerCurve(m_pilot,
                           along = "cluster",   # vary number of clusters
                           nsim  = 200)
plot(power_curve)
```

## Power for Crossed Designs: Participants × Stimuli

In a crossed participant-by-stimulus design (Part 1), power depends on **both** the number of participants and the number of stimuli. This is the most commonly misunderstood case in JDM power planning.

```{r crossed-power-sim}
# Power as a function of N_participants AND N_stimuli simultaneously
simulate_crossed_power <- function(n_part, n_stim, true_effect = 3,
                                   sigma_part = 8, sigma_stim = 6, sigma_e = 5,
                                   n_sims = 300) {
  p_vals <- replicate(n_sims, {
    p_eff <- rnorm(n_part, 0, sigma_part)
    s_eff <- rnorm(n_stim, 0, sigma_stim)
    cond  <- rep(c(0, 1), each = ceiling(n_stim / 2))[1:n_stim]

    df_c <- expand_grid(p = 1:n_part, s = 1:n_stim) |>
      slice_sample(prop = 0.5) |>
      mutate(
        Y   = 50 + true_effect * cond[s] + p_eff[p] + s_eff[s] + rnorm(n(), 0, sigma_e),
        p   = factor(p), s = factor(s), cond = factor(cond[s])
      )

    tryCatch({
      m <- lmer(Y ~ cond + (1 | p) + (1 | s), data = df_c, REML = FALSE,
                control = lmerControl(optimizer = "bobyqa"))
      coef(summary(m))["cond1", "Pr(>|t|)"]
    }, error = function(e) NA_real_, warning = function(w) NA_real_)
  })
  mean(p_vals < 0.05, na.rm = TRUE)
}

# Power grid: varies N_stimuli at fixed N_participants
crossed_grid <- expand_grid(
  n_stim = c(8, 16, 24, 32),
  n_part = c(40, 80)
) |>
  mutate(
    power  = map2_dbl(n_part, n_stim, ~simulate_crossed_power(.x, .y, n_sims = 150)),
    design = paste0(n_part, " participants")
  )

ggplot(crossed_grid, aes(x = n_stim, y = power, color = design, group = design)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("#1e3a5f", "#e63946"), name = NULL) +
  labs(
    title    = "Power in a crossed design: stimuli matter as much as participants",
    subtitle = "Effect size = 3 units; σ_participant = 8; σ_stimulus = 6. Adding stimuli gains more power than adding participants once N_participants > 40.",
    x        = "Number of stimuli",
    y        = "Power"
  ) +
  theme_minimal(base_size = 12)
```

**The take-home:** Increasing participants cannot fully compensate for too few stimuli when stimulus variance is substantial. If you have 8 stimuli and σ_stimulus = 6, no amount of participant recruitment will push power to 80% for a small effect — the stimulus-level variance is the binding constraint. Westfall, Kenny & Judd (2014) showed this is the reason many JDM replications fail even with seemingly large samples.

---

## Key Takeaways for Power Planning

::: {.callout-note}
## Rules of Thumb

1. **More clusters > more per cluster** when ICC is high. If ICC = 0.20, doubling the number of clusters gains much more power than doubling the size of each cluster.

2. **The ICC threshold:** If ICC > 0.10 and cluster size > 10, standard power formulas underestimate required N by 50–100%.

3. **For crossed designs (participants × stimuli):** Power depends on both the number of participants *and* the number of stimuli. Westfall, Kenny & Judd (2014) showed that many published studies are severely underpowered on the stimulus side.

4. **Collect pilot ICC estimates:** Run a small pilot study and compute the ICC before committing to a full design.

5. **Simulate uncertainty:** Report a range of power values across plausible ICC and effect size assumptions, not a single number.
:::

---

## Minimum Reporting Checklist for Power

::: {.callout-tip icon=false}
## What to Report in a Power Section or Preregistration

- [ ] Assumed ICC (with source: pilot data, literature, or sensitivity range)
- [ ] Assumed random slope variance (if applicable)
- [ ] Assumed effect size (with justification)
- [ ] Assumed attrition rate and compliance rate (if non-trivial)
- [ ] Analysis model used in the simulation (exact formula)
- [ ] Simulation code or package name and version
- [ ] Number of simulation replications
- [ ] Power curve or sensitivity table across plausible ICC × effect size combinations
- [ ] Final design choice and rationale (why this N over other options)
:::

