---
title: "Data Structure and Format"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
set.seed(77)
```

# Data Structure and Format

## The Shape of Nested Data

Before fitting any model, you need your data in the right shape. Nested data analysis requires **long format**: one row per observation at the lowest level of the hierarchy, with identifier variables that track higher-level units.

This sounds simple. In practice, it trips up a surprising number of researchers.

## Long Format vs. Wide Format

Most people who have worked with repeated measures data in SPSS or Excel have encountered data in **wide format**: one row per person, with multiple columns for each time point or condition.

```{r wide-format}
# Wide format example
df_wide <- tibble(
  person_id = 1:6,
  sex       = c("F", "M", "F", "F", "M", "M"),
  age       = c(23, 31, 27, 25, 29, 34),
  mood_day1 = c(65, 72, 58, 70, 63, 55),
  mood_day2 = c(68, 69, 61, 74, 60, 58),
  mood_day3 = c(71, 67, 64, 72, 65, 62),
  mood_day4 = c(66, 74, 59, 75, 61, 57),
  mood_day5 = c(70, 70, 63, 78, 64, 60)
)

knitr::kable(df_wide, caption = "Wide format: 6 people × 5 days = 6 rows")
```

This format feels intuitive — it keeps all information about one person on one row. But it is not suitable for multilevel modeling. Why?

Because the model needs to see the nesting explicitly. It needs to know that `mood_day1`, `mood_day2`, and `mood_day3` belong to the same person — that they are three realizations of the same level-1 unit nested inside one level-2 unit.

**Long format** makes the nesting explicit:

```{r long-format}
df_long <- df_wide |>
  pivot_longer(
    cols      = starts_with("mood_"),
    names_to  = "day",
    values_to = "mood"
  ) |>
  mutate(
    day = as.integer(str_remove(day, "mood_day"))
  ) |>
  select(person_id, sex, age, day, mood)

knitr::kable(df_long, caption = "Long format: 6 people × 5 days = 30 rows")
```

Now each row is one person-day observation. The nesting is explicit: `person_id` is the level-2 unit, `day` is the level-1 index within each person.

## Converting Wide to Long in R

```{r pivot-longer-demo, eval=FALSE}
# The pivot_longer() function does the conversion
df_long <- df_wide |>
  pivot_longer(
    cols      = starts_with("mood_"),   # which columns to collapse
    names_to  = "day",                  # new column for what the columns were about
    values_to = "mood"                  # new column for the values
  ) |>
  mutate(
    day = as.integer(str_remove(day, "mood_day"))  # clean up the day variable
  )
```

## The Essential Identifier Variables

Every nested dataset needs at least two types of identifier variables:

1. **Level-2 identifier:** A unique code for each higher-level unit (person, firm, classroom, country). This is what tells the model which observations belong together.
2. **Level-1 identifier or index:** A variable that distinguishes observations *within* the same level-2 unit (day number, condition, trial number).

In three-level data, you need identifiers at all three levels.

::: {.callout-important icon=false}
## Checklist: Preparing Nested Data

Before fitting any model, verify:

- [ ] **One row per level-1 observation** (not one row per person, unless you have only cross-sectional nesting)
- [ ] **A unique, clean level-2 ID** (numeric or factor, no missing values, consistent coding)
- [ ] **A level-1 index within each level-2 unit** (day number, trial number, condition)
- [ ] **Level-2 variables appear on every row for that unit** (age, sex, and other person-level variables are repeated across all a person's rows)
- [ ] **No structural missingness** (if some observations are genuinely missing vs. structurally absent, treat them differently)
- [ ] **Balanced or unbalanced — you know which** (lme4 handles both, but you should know what you have)
:::

## Level-1 vs. Level-2 Variables

A common source of confusion: variables that describe the *person* versus variables that describe the *observation*.

```{r level-check}
df_long |>
  head(10) |>
  knitr::kable(caption = "Identifying level in the dataset")
```

In this dataset:
- `person_id`, `sex`, `age` are **level-2 variables** — they describe the person and are constant across all of that person's rows
- `day` and `mood` are **level-1 variables** — they vary across observations within a person

This distinction matters enormously for model building (centering, interpretation of effects) and will come up repeatedly in Module 2.

## A Note on Balanced vs. Unbalanced Data

**Balanced** data means every level-2 unit has the same number of level-1 observations. All 30 people have exactly 5 days of data. All 25 classrooms have exactly 20 students.

**Unbalanced** data means the number of observations varies across level-2 units. Some people completed 7 days of the diary; others dropped out after 2. Some classrooms have 15 students; others have 28.

Standard ANOVA-based approaches for repeated measures struggle with unbalanced data — they were designed for balanced designs. Mixed-effects models handle unbalanced data gracefully, because they use the information available for each unit without requiring complete cases.

```{r unbalanced-demo}
# Show what unbalanced data looks like
df_unbalanced <- df_long |>
  # randomly drop some observations to simulate dropout
  filter(!(person_id %in% c(1, 4) & day >= 4),
         !(person_id == 6 & day >= 3))

df_unbalanced |>
  group_by(person_id) |>
  summarise(n_obs = n(), days_observed = paste(day, collapse = ", ")) |>
  knitr::kable(caption = "Unbalanced data: some people have fewer observations")
```

## Diagnosing Your Data Structure

Before fitting any model, it is good practice to run a quick structure check:

```{r structure-check}
# How many observations per person?
df_long |>
  count(person_id, name = "n_obs") |>
  summary()

# Is the data truly in long format? Check that level-2 variables don't vary within person
df_long |>
  group_by(person_id) |>
  summarise(
    n_unique_sex = n_distinct(sex),   # should be 1
    n_unique_age = n_distinct(age)    # should be 1
  ) |>
  filter(n_unique_sex > 1 | n_unique_age > 1)  # should return 0 rows
```

Zero rows returned means no person has inconsistent sex or age coding across their observations. Good.

```{r structure-check2}
# How many level-2 units?
n_distinct(df_long$person_id)

# What is the range of level-1 observations per unit?
df_long |>
  count(person_id) |>
  pull(n) |>
  range()
```

This takes two minutes and saves hours of debugging later.

## Looking Ahead

With the conceptual foundations in place — DGP thinking, types of nesting, ICC, and data format — we are ready to start fitting models. Module 1 builds the core toolkit: fixed effects, random intercepts, and random slopes. At each step, we will be explicit about what assumption the model is making and what happens when that assumption is wrong.
