---
title: "Types of Nested Structures"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(lme4)
set.seed(123)
```

# Types of Nested Structures

## A Taxonomy of Nesting

Nesting takes several forms in social science data. Getting them confused leads to misspecified models and wrong answers. This section maps out the main types, shows what they look like in a dataset, and explains how the DGP perspective helps you identify which structure you have.

The key distinction is between **pure hierarchies** (units nested strictly within higher-level units) and **crossed structures** (units belonging to multiple grouping factors simultaneously). Real data often combines both.

---

## Pure Hierarchy: One Nesting Level

The simplest nested structure: lower-level units are grouped into higher-level units, and each lower-level unit belongs to exactly one group.

**Example:** Consumers nested within retail stores.

- Hundreds of consumers, each shopping at one of several stores
- A consumer's purchase amount depends partly on their own spending habits and partly on the store environment (location, product mix, ambient cues)
- Purchases within the same store are more similar to each other than to purchases from other stores

Think of a rectangle divided into colored columns — each column is a store, and each dot within a column is a consumer. The columns (stores) explain part of why purchases cluster.

```{r hierarchy-diagram}
# Simulate a two-level structure
n_stores             <- 6
n_consumers_per_store <- 8

store_effects <- rnorm(n_stores, 0, 15)

df_stores <- expand_grid(
  store      = 1:n_stores,
  consumer   = 1:n_consumers_per_store
) |>
  mutate(
    consumer_id     = row_number(),
    store_effect    = store_effects[store],
    purchase_amount = 85 + store_effect + rnorm(n(), 0, 8),
    store           = factor(store, labels = paste("Store", 1:n_stores))
  )

ggplot(df_stores, aes(x = store, y = purchase_amount, color = store)) +
  geom_jitter(width = 0.15, size = 3, alpha = 0.8) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5,
               color = "black", linewidth = 0.8) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title    = "Consumers nested within retail stores",
    subtitle = "Black bars = store means; dots = individual consumers",
    x = "Store", y = "Purchase Amount ($)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

Notice that purchase amounts cluster within stores. Some stores attract higher-spending consumers due to location, product mix, or ambient cues. Purchases within the same store are more similar to each other than to purchases from other stores. A model that ignores stores treats these 48 observations as 48 independent data points. They are not — they are 6 groups of 8.

**Data format:** One row per consumer visit. You need a consumer ID and a store ID.

```{r hierarchy-data, echo=FALSE}
df_stores |>
  select(store, consumer, purchase_amount) |>
  head(12) |>
  knitr::kable(digits = 1, caption = "First 12 rows of a two-level dataset")
```

::: {.callout-tip}
## Exercise

Look at the store-level means in the plot above. Which store appears to have the highest average purchase amount? Now consider: what store-level factors (location quality, product assortment, store ambiance) might explain those differences? This is the between-store variance your model needs to account for.
:::

---

## Three Levels: Nesting Within Nesting

The hierarchy can extend further. Consumers are nested within stores, which are nested within retail chains or geographic regions.

**Example:** Purchase occasions nested within customers nested within banking platforms.

Three nested boxes: the outer box is the platform, within it are customers, and within each customer are multiple occasions.

Each level adds a new source of systematic variance:

- Occasions differ from each other within a customer (day of week, mood, context)
- Customers differ in their baseline spending and financial habits
- Banks and platforms differ in the population they serve, fees, and product mix

```{r three-level}
n_banks                 <- 3
n_customers_per_bank    <- 4
n_occasions_per_customer <- 5

bank_effects     <- rnorm(n_banks, 0, 8)
customer_effects <- rnorm(n_banks * n_customers_per_bank, 0, 5)

df_three <- expand_grid(
  bank     = 1:n_banks,
  customer = 1:n_customers_per_bank,
  occasion = 1:n_occasions_per_customer
) |>
  mutate(
    customer_id       = (bank - 1) * n_customers_per_bank + customer,
    bank_eff          = bank_effects[bank],
    customer_eff      = customer_effects[customer_id],
    transaction_amount = 150 + bank_eff + customer_eff + rnorm(n(), 0, 4),
    bank_label        = paste("Platform", bank),
    customer_label    = paste0("C", customer_id)
  )

ggplot(df_three, aes(x = customer_label, y = transaction_amount,
                      color = bank_label)) +
  geom_jitter(width = 0.1, alpha = 0.7, size = 2.5) +
  stat_summary(fun = mean, geom = "point", shape = 18,
               size = 4, color = "black") +
  facet_wrap(~ bank_label, scales = "free_x") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title  = "Three-level structure: Occasions within Customers within Platforms",
    x      = "Customer",
    y      = "Transaction Amount ($)",
    color  = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

The key diagnostic question for three-level data: **how much variance lives at each level?** Later, the ICC will give us a principled answer.

---

## Panel Data: Cross-Sectional and Longitudinal

**Panel data** combines the cross-sectional and time-series structures. You observe multiple units (people, firms, households) at multiple time points.

**Example:** Monthly spending by 50 households measured over 24 months.

```{r panel-data}
n_households <- 50
n_months     <- 24

household_baseline <- rnorm(n_households, 1000, 200)

df_panel <- expand_grid(
  household = 1:n_households,
  month     = 1:n_months
) |>
  mutate(
    baseline   = household_baseline[household],
    spending   = baseline + 20 * (month - 12.5) / 12 + rnorm(n(), 0, 80),
    time_trend = month
  )

# Show 6 households
sample_hh <- sample(1:n_households, 6)

ggplot(df_panel |> filter(household %in% sample_hh),
       aes(x = month, y = spending, color = factor(household))) +
  geom_line(alpha = 0.8) +
  geom_point(size = 1.5, alpha = 0.6) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title    = "Panel data: 6 households over 24 months",
    subtitle = "Each line is one household's spending trajectory",
    x        = "Month",
    y        = "Monthly Spending ($)",
    color    = "Household"
  ) +
  theme_minimal(base_size = 13)
```

Panel data has **two sources of nesting** simultaneously:
1. Observations are nested within households (same household at different times)
2. Observations might also be nested within time periods (all households in month 1 face the same economic conditions)

This means you potentially need fixed or random effects for both dimensions — a crossed structure, not a pure hierarchy.

---

## Crossed Random Effects

In a **crossed** structure, observations belong to multiple grouping factors that are not nested within each other.

**Example from JDM research:** Participants evaluate product messages in a pricing study.

In a JDM experiment, 40 participants each evaluate 80 product messages. Each message is a pricing or promotional offer sampled from a larger population. The outcome is willingness to pay (WTP) in dollars.

A grid: rows are participants, columns are messages. Every cell is one WTP judgment. No row is nested inside any column — they cross.

- Each observation belongs to one participant AND one message
- Participants are not nested within messages, and messages are not nested within participants
- Both participant differences (baseline WTP, income, brand affinity) and message differences (quality of framing, scarcity cues) contribute to WTP

```{r crossed-diagram}
# Small illustration: 4 participants × 4 messages
crossed_df <- expand_grid(
  participant = paste("P", 1:4),
  message     = paste("M", 1:4)
) |>
  mutate(wtp = 50 + rnorm(16, 0, 15))

ggplot(crossed_df, aes(x = message, y = participant, fill = wtp)) +
  geom_tile(color = "white", linewidth = 2) +
  geom_text(aes(label = round(wtp)), size = 4) +
  scale_fill_gradient(low = "#deebf7", high = "#08519c") +
  labs(
    title    = "Crossed structure: each participant evaluates each message",
    subtitle = "Cell = WTP for that participant-message combination",
    x        = "Message",
    y        = "Participant",
    fill     = "WTP ($)"
  ) +
  theme_minimal(base_size = 13)
```

The key point: participant 1's WTP for message 3 is not in the same "group" as participant 2's WTP for message 3. Both observations share a message, but not a participant. This crossed structure requires a different model than a pure hierarchy.

Now a full simulation showing the consequences of getting this wrong:

```{r crossed-full-sim}
n_participants <- 40
n_messages     <- 80

participant_effects <- rnorm(n_participants, 0, 12)   # people differ in baseline WTP
message_effects     <- rnorm(n_messages, 0, 8)        # messages differ in baseline WTP
message_quality     <- runif(n_messages, 1, 10)       # message quality predictor

df_msgs <- expand_grid(participant = 1:n_participants, message = 1:n_messages) |>
  mutate(
    participant_eff = participant_effects[participant],
    message_eff     = message_effects[message],
    quality         = message_quality[message],
    wtp             = 50 + 4 * quality + participant_eff + message_eff + rnorm(n(), 0, 10),
    participant     = factor(participant),
    message         = factor(message)
  )
```

```{r crossed-models}
# Without message random effects (wrong)
model_no_msgs <- lmer(wtp ~ quality + (1 | participant),
                      data = df_msgs)

# With crossed random effects (correct)
model_crossed <- lmer(wtp ~ quality + (1 | participant) + (1 | message),
                      data = df_msgs)
```

```{r crossed-se-comparison}
wrong_se   <- broom.mixed::tidy(model_no_msgs)[2, "std.error"][[1]]
correct_se <- broom.mixed::tidy(model_crossed)[2, "std.error"][[1]]

tibble(
  Model          = c("Without message RE", "With message RE (crossed)"),
  Quality_SE     = c(wrong_se, correct_se)
) |>
  knitr::kable(digits = 3,
    caption = "SE of message quality effect: ignoring vs. modeling message-level variance")
```

Some participants have higher baseline WTP due to income, brand affinity, or general deal-seeking tendencies. Some messages are more compelling due to framing quality, scarcity cues, or discount depth. Both sources of variance must be modeled — ignoring message-level variance deflates the standard error on the quality effect, producing anti-conservative inference.

---

## Repeated Measures and Longitudinal Data

**Repeated measures** means the same unit is measured multiple times — on the same variable, under different conditions, at different time points, or both.

The critical distinction:

- **Repeated measures (experimental):** Each person experiences multiple conditions. The conditions are fixed (e.g., gain frame vs. loss frame). The person is the random unit.
- **Longitudinal data:** Each person is measured at multiple time points, and time is part of the model. Change over time is the phenomenon of interest.

Both are cases of **observations nested within persons**. But they ask different questions. Repeated-measures experiments ask: "does the framing condition affect the outcome?" Longitudinal models ask: "how does the outcome change over time, and does it change differently for different people?"

```{r rm-vs-long}
# Repeated measures: gain vs. loss framing
df_rm <- tibble(
  person    = rep(1:20, each = 2),
  condition = rep(c("loss_frame", "gain_frame"), 20),
  purchase_intention = 55 + ifelse(condition == "gain_frame", 8, 0) +
                       rnorm(40, 0, 20)
)

p_rm <- ggplot(df_rm, aes(x = condition, y = purchase_intention,
                           group = person, color = factor(person))) +
  geom_line(alpha = 0.5) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_viridis_d() +
  labs(title = "Repeated measures: 2 framing conditions per person",
       x = NULL, y = "Purchase Intention (0-100)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# Longitudinal
df_long <- tibble(
  person = rep(1:10, each = 5),
  time   = rep(1:5, 10),
  score  = rep(rnorm(10, 50, 8), each = 5) +
           rep(rnorm(10, 2, 1), each = 5) * (rep(1:5, 10) - 3) +
           rnorm(50, 0, 3)
)

p_long <- ggplot(df_long, aes(x = time, y = score,
                               group = person, color = factor(person))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_viridis_d() +
  labs(title = "Longitudinal:\n5 time points per person",
       x = "Time", y = "Score") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

library(patchwork)
p_rm + p_long
```

---

## A Map of Common Data Structures

| Structure | Example | Nesting | DGP implication |
|-----------|---------|---------|-----------------|
| Pure hierarchy | Consumers within retail stores | Strict: each unit in one group | Between-group variance is real |
| Three-level | Purchase occasions within customers within platforms | Nested levels | Variance at each level |
| Panel | Households across months (panel) | Cross-sectional + time | Unit effects and time effects |
| Crossed | Participants × product messages | Each unit in multiple groups | Two sources of random variance |
| Repeated measures | Person × framing condition | Each person measured in each condition | Within-person correlation |
| Longitudinal | Person measured over time | Observations in person | Individual trajectories |

The single most common mistake is treating one of these as another. The second most common mistake is failing to ask the question at all.

::: {.callout-tip}
## Practical Exercise

Look at the last dataset you analyzed. Ask:

1. What is the row in this dataset?
2. Can I draw a diagram of what contains what?
3. Are there two dimensions of grouping (e.g., people AND stimuli) that are crossed rather than nested?
4. What does a standard regression assume about the structure I just identified?
:::
