---
title: "Data Structure and Format"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
set.seed(77)
```

# Data Structure and Format

## The Shape of Nested Data

Before fitting any model, you need your data in the right shape. Nested data analysis requires **long format**: one row per observation at the lowest level of the hierarchy, with identifier variables that track higher-level units.

This sounds simple. In practice, it trips up a surprising number of researchers.

## Long Format vs. Wide Format

Most people who have worked with repeated measures data in SPSS or Excel have encountered data in **wide format**: one row per person, with multiple columns for each time point or condition.

```{r wide-format}
# Wide format example: household spending diary
df_wide <- tibble(
  household_id  = 1:6,
  region        = c("North", "South", "North", "West", "South", "West"),
  income_k      = c(55, 82, 63, 48, 91, 70),   # household income in thousands
  spending_wk1  = c(420, 680, 510, 390, 740, 580),
  spending_wk2  = c(445, 655, 525, 405, 710, 595),
  spending_wk3  = c(435, 700, 498, 385, 760, 570),
  spending_wk4  = c(460, 670, 515, 420, 725, 610),
  spending_wk5  = c(450, 690, 505, 410, 745, 590)
)

knitr::kable(df_wide, caption = "Wide format: 6 households × 5 weeks = 6 rows")
```

This format feels intuitive — it keeps all information about one household on one row. But it is not suitable for multilevel modeling because the model needs to see the nesting explicitly: `spending_wk1` through `spending_wk5` are five realizations of the same level-1 unit (week) nested inside one level-2 unit (household).

**Long format** makes the nesting explicit:

```{r long-format}
df_long <- df_wide |>
  pivot_longer(
    cols      = starts_with("spending_"),
    names_to  = "week",
    values_to = "spending"
  ) |>
  mutate(
    week = as.integer(str_remove(week, "spending_wk"))
  ) |>
  select(household_id, region, income_k, week, spending)

knitr::kable(df_long, caption = "Long format: 6 households × 5 weeks = 30 rows")
```

Now each row is one household-week observation. The nesting is explicit: `household_id` is the level-2 unit, `week` is the level-1 index within each household.

## Converting Wide to Long in R

```{r pivot-longer-demo, eval=FALSE}
# The pivot_longer() function does the conversion
df_long <- df_wide |>
  pivot_longer(
    cols      = starts_with("spending_"),   # which columns to collapse
    names_to  = "week",                     # new column for the time index
    values_to = "spending"                  # new column for the values
  ) |>
  mutate(
    week = as.integer(str_remove(week, "spending_wk"))  # clean up the week variable
  )
```

## The Essential Identifier Variables

Every nested dataset needs at least two types of identifier variables:

1. **Level-2 identifier:** A unique code for each higher-level unit (household, branch, customer, country). This is what tells the model which observations belong together.
2. **Level-1 identifier or index:** A variable that distinguishes observations *within* the same level-2 unit (week number, transaction, product evaluated).

In three-level data, you need identifiers at all three levels.

::: {.callout-important icon=false}
## Checklist: Preparing Nested Data

Before fitting any model, verify:

- [ ] **One row per level-1 observation** (not one row per person, unless you have only cross-sectional nesting)
- [ ] **A unique, clean level-2 ID** (numeric or factor, no missing values, consistent coding)
- [ ] **A level-1 index within each level-2 unit** (day number, trial number, condition)
- [ ] **Level-2 variables appear on every row for that unit** (age, sex, and other person-level variables are repeated across all a person's rows)
- [ ] **No structural missingness** (if some observations are genuinely missing vs. structurally absent, treat them differently)
- [ ] **Balanced or unbalanced — you know which** (lme4 handles both, but you should know what you have)
:::

## Level-1 vs. Level-2 Variables

A common source of confusion: variables that describe the *person* versus variables that describe the *observation*.

```{r level-check}
df_long |>
  head(10) |>
  knitr::kable(caption = "Identifying level in the dataset")
```

In this dataset:
- `household_id`, `region`, `income_k` are **level-2 variables** — they describe the household and are constant across all of that household's rows
- `week` and `spending` are **level-1 variables** — they vary across observations within a household

This distinction matters enormously for model building (centering, interpretation of effects) and will come up repeatedly in Module 2.

## A Note on Balanced vs. Unbalanced Data

**Balanced** data means every level-2 unit has the same number of level-1 observations. All 6 households have exactly 5 weeks of spending data. All 20 bank branches have exactly 15 customer records.

**Unbalanced** data means the number of observations varies across level-2 units. Some households completed the full 5-week spending diary; others dropped out after 2 weeks. Some branches have 10 customers in the dataset; others have 28.

Standard ANOVA-based approaches for repeated measures struggle with unbalanced data — they were designed for balanced designs. Mixed-effects models handle unbalanced data gracefully, because they use the information available for each unit without requiring complete cases.

```{r unbalanced-demo}
# Show what unbalanced data looks like
df_unbalanced <- df_long |>
  # randomly drop some observations to simulate dropout / incomplete diary
  filter(!(household_id %in% c(1, 4) & week >= 4),
         !(household_id == 6 & week >= 3))

df_unbalanced |>
  group_by(household_id) |>
  summarise(n_obs = n(), weeks_observed = paste(week, collapse = ", ")) |>
  knitr::kable(caption = "Unbalanced data: some households have fewer weekly observations")
```

## Diagnosing Your Data Structure

Before fitting any model, it is good practice to run a quick structure check:

```{r structure-check}
# How many observations per household?
df_long |>
  count(household_id, name = "n_obs") |>
  summary()

# Is the data truly in long format? Check that level-2 variables don't vary within household
df_long |>
  group_by(household_id) |>
  summarise(
    n_unique_region    = n_distinct(region),    # should be 1
    n_unique_income    = n_distinct(income_k)   # should be 1
  ) |>
  filter(n_unique_region > 1 | n_unique_income > 1)  # should return 0 rows
```

Zero rows returned means no household has inconsistent region or income coding across its weekly observations. Good.

```{r structure-check2}
# How many level-2 units?
n_distinct(df_long$household_id)

# What is the range of level-1 observations per unit?
df_long |>
  count(household_id) |>
  pull(n) |>
  range()
```

This takes two minutes and saves hours of debugging later.

## Experimental Long Format: Participants × Products

The household-week structure above is the observational prototype. In experiments, the analogous structure is **participants × products (or stimuli)**: one row per participant-product combination, with participant and product identifiers at every row.

```{r rm-vs-long}
# Wide format: one row per participant, columns for each product evaluated
df_exp_wide <- tibble(
  participant = 1:4,
  condition   = c("control", "treatment", "control", "treatment"),
  prod_1_wtp  = c(42, 58, 39, 61),
  prod_2_wtp  = c(35, 52, 33, 55),
  prod_3_wtp  = c(48, 63, 44, 67),
  prod_4_wtp  = c(38, 54, 36, 57)
)

# Long format: one row per participant-product observation
df_exp_long <- df_exp_wide |>
  pivot_longer(
    cols      = starts_with("prod_"),
    names_to  = "product",
    values_to = "wtp"
  ) |>
  mutate(product = str_remove(product, "_wtp"))

knitr::kable(df_exp_long, caption = "Long format: 4 participants × 4 products = 16 rows. Both participant and product IDs appear on every row.")
```

The two identifier columns — `participant` and `product` — are what allow `lme4` to fit crossed random effects: a random intercept for participants *and* a random intercept for products, simultaneously. Without the long format, you cannot express this structure.

::: {.callout-note icon=false}
## Observational vs. Experimental Long Format

| Feature | Observational (household-week) | Experimental (participant-product) |
|---------|-------------------------------|-------------------------------------|
| Level-1 unit | spending week | product / stimulus |
| Level-2 unit | household | participant |
| Additional clustering | — | product (crossed) |
| Level-1 index | week number | product ID |
| Key model formula | `(1 \| household)` | `(1 \| participant) + (1 \| product)` |

Both formats follow the same rule: **one row per level-1 observation**. The difference is that experimental data often has *two* grouping factors (participants and products), creating a crossed rather than nested structure.
:::

## Looking Ahead

With the conceptual foundations in place — DGP thinking, types of nesting, ICC, and data format — we are ready to start fitting models. Module 1 builds the core toolkit: fixed effects, random intercepts, and random slopes. At each step, we will be explicit about what assumption the model is making and what happens when that assumption is wrong.
