---
title: "Measuring Non-Independence"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(ggplot2)
set.seed(999)
```

# Measuring Non-Independence

## Why We Need a Number

Nested structure is not binary — it is not just "present" or "absent." Nesting can be strong (observations within a group are nearly identical) or weak (observations within a group are barely more similar than observations from different groups). The strength of nesting determines how badly standard analysis fails.

We need a number that quantifies this strength. That number is the **Intraclass Correlation Coefficient (ICC)**.

## The Intraclass Correlation Coefficient

The ICC answers a deceptively simple question:

> **Of all the variance in my outcome, what fraction is due to group membership?**

If you pick two observations at random from the same group, how correlated will they be? That correlation is the ICC.

$$
\text{ICC} = \frac{\sigma^2_{\text{between}}}{\sigma^2_{\text{between}} + \sigma^2_{\text{within}}}
$$

where $\sigma^2_{\text{between}}$ is the variance in group means, and $\sigma^2_{\text{within}}$ is the variance of individual observations around their group mean.

::: {.callout-note}
## Interpreting ICC Values

- **ICC ≈ 0:** Knowing which group an observation is in tells you almost nothing. Group membership barely predicts the outcome. Nesting is present but weak.
- **ICC ≈ 0.05–0.15:** Typical for many psychological studies. Modest clustering. Looks small but has real consequences for inference.
- **ICC ≈ 0.20–0.40:** Strong clustering. Common in organizational data (employees within firms), educational data (students within schools), or within-person repeated measures.
- **ICC ≈ 1.0:** All variance is between groups. Observations within a group are identical. Knowing the group tells you everything.
:::

Let's see three cases:

```{r icc-illustration}
# Generate data with three different levels of ICC
make_grouped_data <- function(icc, n_groups = 10, n_per_group = 20) {
  sigma_between <- sqrt(icc)
  sigma_within  <- sqrt(1 - icc)

  group_means <- rnorm(n_groups, 0, sigma_between)

  expand_grid(group = 1:n_groups, obs = 1:n_per_group) |>
    mutate(
      y   = group_means[group] + rnorm(n(), 0, sigma_within),
      icc_label = paste0("ICC = ", icc)
    )
}

df_icc <- bind_rows(
  make_grouped_data(0.02),
  make_grouped_data(0.20),
  make_grouped_data(0.60)
)

ggplot(df_icc, aes(x = factor(group), y = y, color = factor(group))) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1.5) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5,
               color = "black", linewidth = 0.8) +
  facet_wrap(~ icc_label, ncol = 3) +
  scale_color_viridis_d() +
  labs(
    title    = "Same data, different ICCs",
    subtitle = "Black bars = group means; dots = individual observations",
    x        = "Group",
    y        = "Outcome"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

When ICC = 0.02, the group means barely differ — the groups look almost interchangeable. When ICC = 0.60, the group means dominate: knowing which group you are in tells you most of what you need to know about someone's score.

## How to Compute the ICC in R

The ICC comes directly from a null (intercept-only) multilevel model — one with no predictors, just a random intercept for the grouping variable. The model partitions total variance into between-group and within-group components.

```{r compute-icc}
# Simulate a dataset: employees nested within firms
n_firms     <- 30
n_employees <- 15

firm_effects <- rnorm(n_firms, 50, 6)   # between-firm variance

df_firms <- expand_grid(
  firm     = 1:n_firms,
  employee = 1:n_employees
) |>
  mutate(
    satisfaction = firm_effects[firm] + rnorm(n(), 0, 8)  # within-firm variance
  )

# Fit the null model
null_model <- lmer(satisfaction ~ 1 + (1 | firm), data = df_firms)
summary(null_model)
```

```{r icc-from-model}
# Extract variance components
vc <- as.data.frame(VarCorr(null_model))
var_between <- vc$vcov[1]   # firm-level variance
var_within  <- vc$vcov[2]   # residual variance

icc_value <- var_between / (var_between + var_within)
cat("ICC =", round(icc_value, 3))
```

The `performance` package provides a cleaner interface:

```{r icc-performance}
library(performance)
icc(null_model)
```

## Why Even a Small ICC is a Big Problem

This is one of the most important things to understand before you run any analysis on nested data: **even a modest ICC has large consequences for inference.**

The reason is the **design effect** (sometimes called DEFF): a multiplier that tells you how much your effective sample size shrinks due to clustering.

$$
\text{DEFF} = 1 + (n_{\text{cluster}} - 1) \times \text{ICC}
$$

where $n_{\text{cluster}}$ is the number of observations per cluster.

```{r deff-calculation}
# Design effect as a function of ICC and cluster size
deff_df <- expand_grid(
  icc          = c(0.01, 0.05, 0.10, 0.20, 0.30),
  cluster_size = c(5, 10, 20, 50)
) |>
  mutate(
    deff           = 1 + (cluster_size - 1) * icc,
    effective_n    = 1000 / deff,  # if nominal N = 1000
    label          = paste0("n = ", cluster_size)
  )

ggplot(deff_df, aes(x = icc, y = deff, color = factor(cluster_size),
                     linetype = factor(cluster_size))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Set2", name = "Cluster size") +
  scale_linetype_discrete(name = "Cluster size") +
  labs(
    title    = "Design Effect as a function of ICC and cluster size",
    subtitle = "DEFF = 1 means no clustering penalty; DEFF = 5 means your 1000 obs act like 200",
    x        = "ICC",
    y        = "Design Effect (DEFF)",
    caption  = "DEFF = 1 + (cluster_size - 1) × ICC"
  ) +
  theme_minimal(base_size = 13)
```

Let's make this concrete. Suppose you have 200 students in 10 classrooms of 20 students each, and the ICC on your outcome is 0.10 — a seemingly small value often found in educational research.

$$
\text{DEFF} = 1 + (20 - 1) \times 0.10 = 1 + 1.9 = 2.9
$$

Your 200 observations act like only $200 / 2.9 \approx 69$ independent observations. You have nearly three times less statistical power than you thought. If you treat 200 clustered observations as 200 independent ones, your standard errors are too small, your p-values are too small, and your confidence intervals are too narrow.

```{r type1-inflation}
# Simulate Type I error inflation from ignoring clustering
simulate_inflation <- function(icc, n_groups = 20, n_per_group = 10, n_sims = 2000) {
  sigma_between <- sqrt(icc)
  sigma_within  <- sqrt(1 - icc)

  p_values <- replicate(n_sims, {
    group_effects <- rnorm(n_groups, 0, sigma_between)
    y <- rep(group_effects, each = n_per_group) +
         rnorm(n_groups * n_per_group, 0, sigma_within)
    x <- rnorm(n_groups * n_per_group)
    t.test(y[x > 0], y[x <= 0])$p.value
  })

  mean(p_values < 0.05)
}

icc_grid <- c(0, 0.05, 0.10, 0.20, 0.30)
type1_rates <- sapply(icc_grid, simulate_inflation)

tibble(ICC = icc_grid, `False Positive Rate` = type1_rates) |>
  knitr::kable(digits = 3,
               caption = "Type I error rate when ignoring clustering (nominal alpha = 0.05)")
```

When ICC = 0, ignoring clustering is fine — you get the expected 5% false positive rate. But as ICC grows, so does your false positive rate. At ICC = 0.20, you are rejecting the null hypothesis incorrectly nearly twice as often as you should be.

## Assumptions of the ICC

The ICC is not just a number — it embodies assumptions. Checking those assumptions is part of using it correctly.

::: {.callout-warning icon=false}
## Assumption 1: Groups are exchangeable

The ICC assumes that groups are drawn from a common population, and that the relationship between observations within a group reflects a common process. If your groups are wildly different in kind — not just in degree — the ICC may not characterize the nesting correctly.

**How to check:** Look at group means and within-group variances. Are any groups wildly different from the rest? Could the nesting structure be non-stationary (e.g., the within-group correlation is strong in some groups, absent in others)?
:::

::: {.callout-warning icon=false}
## Assumption 2: The grouping variable captures the relevant nesting

A single ICC assumes there is one meaningful level of nesting. If data is actually three-level (observations in therapists in clinics), a single ICC from a two-level model will be misspecified.

**How to check:** Fit null models at each level and compute ICCs for each. Compare the variance components.
:::

::: {.callout-warning icon=false}
## Assumption 3: The ICC is stable across groups

A standard ICC treats within-group variance as homogeneous across groups. In reality, some groups may be much more variable than others (heteroskedasticity at level 2).

**How to check:** Plot within-group standard deviations against group means. A strong relationship suggests heteroskedasticity.
:::

```{r icc-assumptions-check}
# Check: is within-group variance stable?
df_firms |>
  group_by(firm) |>
  summarise(
    group_mean = mean(satisfaction),
    group_sd   = sd(satisfaction),
    n          = n()
  ) |>
  ggplot(aes(x = group_mean, y = group_sd)) +
  geom_point(size = 3, color = "#1e3a5f", alpha = 0.7) +
  geom_smooth(method = "loess", se = TRUE, color = "#e63946") +
  labs(
    title    = "Within-group SD vs. group mean",
    subtitle = "A strong relationship suggests heteroskedastic nesting",
    x        = "Group Mean (Satisfaction)",
    y        = "Within-Group SD"
  ) +
  theme_minimal(base_size = 13)
```

In this simulated dataset, there is no strong relationship — within-group variance is roughly constant across firms. That is reassuring, but not always the case in real data.

## ICC in Experimental Research: Two Sources of Non-Independence

The ICC applies equally to experimental designs — but the sources of non-independence are different, and there are typically *two* of them at once.

In a **within-subjects experiment**, each participant responds to multiple stimuli or conditions. Two responses from the *same participant* will be more similar than responses from different participants: this is the participant-level ICC. Two responses to the *same stimulus* (rated by different participants) will also be more similar than responses to different stimuli: this is the stimulus-level ICC.

::: {.callout-note icon=false}
## Two ICCs in a Crossed Design

Consider a word recall experiment: 40 participants each see 80 words, half positive and half neutral. The outcome (recalled = 1 / not = 0) has clustering at two levels simultaneously:

- **Participant ICC:** People differ in overall recall ability. A high-ability participant's responses are positively correlated with each other.
- **Stimulus ICC:** Some words are simply more memorable. Responses to the same word across participants share variance.

Ignoring *either* ICC inflates the Type I error rate for the focal treatment comparison. This is why Module 1 Part 6 introduces **crossed random effects**: `(1 | participant) + (1 | stimulus)` models both ICCs simultaneously, and the design-effect penalty compounds across both levels.

```r
# Crossed design: two ICCs
null_crossed <- lmer(recalled ~ 1 + (1 | participant) + (1 | stimulus), data = df_exp)
icc(null_crossed)  # returns ICCs for both grouping levels
```

In typical psycholinguistic and judgment studies, stimulus ICCs of 0.05–0.20 are common — small enough to ignore in any single study, large enough to inflate false positives systematically across a literature.
:::

## The ICC as a Scientific Finding

Throughout this workshop, we treat the ICC primarily as a diagnostic tool — something you check before deciding what model to use. But it is also a finding in its own right.

If your outcome is job satisfaction and the ICC across firms is 0.30, that means 30% of the variance in job satisfaction is explained simply by *which firm* someone works for. That is a substantive result. It tells you that organizational-level factors — culture, management, pay structure — explain nearly a third of the variation in how people feel about their work. That is worth reporting, and worth investigating.

The decomposition of variance across levels is one of the most underutilized findings in social science research.
