---
title: "Crossed Random Effects"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(ggplot2)
library(patchwork)
set.seed(42)
```

# Crossed Random Effects

Pure hierarchies assume each lower-level unit belongs to exactly one higher-level unit. But many real data structures — in both experimental consumer research and longitudinal panel research — have units that belong to *multiple* grouping factors simultaneously. This chapter covers two major instances of crossed random effects:

1. **Within-subjects experiments**: participants × product messages (the consumer WTP problem)
2. **Longitudinal panel data**: customers × calendar months (the spending panel problem)

Both cases share the same formal structure. Both produce badly wrong standard errors when one of the crossed grouping factors is omitted.

---

## Part A: Crossed Random Effects in Experimental Designs

### The Problem with Treating Messages as Fixed

A consumer research experiment: 40 participants each evaluate 80 pricing messages. Each observation belongs to one participant AND one message. Participants are not nested within messages. Messages are not nested within participants. Both grouping factors are fully **crossed**.

If you only include a random effect for participants, you are assuming that all variance due to messages is captured by the residual. But messages differ systematically — some promotional offers are more compelling due to framing quality, discount depth, or scarcity cues. This message-level variance inflates your residuals and biases your standard errors.

Judd, Westfall, and Kenny (2012, 2017) showed that failing to include random effects for stimuli is one of the most widespread sources of Type I error inflation in social and cognitive psychology. The same logic applies directly to consumer research: when messages (or products, or offers) are the stimuli, they must be treated as a random factor.

### The Model

For observation $y_{ps}$ from participant $p$ evaluating message $s$:

$$
Y_{ps} = \underbrace{\gamma_{00}}_{\text{grand mean}} + \underbrace{\beta_1 X_{ps}}_{\text{predictor}} + \underbrace{u_{0p}}_{\text{participant RE}} + \underbrace{w_{0s}}_{\text{message RE}} + \underbrace{\varepsilon_{ps}}_{\text{residual}}
$$

Where:

- $u_{0p} \sim \mathcal{N}(0, \sigma^2_u)$: participant random intercepts (some people have higher baseline WTP)
- $w_{0s} \sim \mathcal{N}(0, \sigma^2_w)$: message random intercepts (some offers are more compelling overall)
- $\varepsilon_{ps}$: trial-level residual noise

```{r experiment-sim}
# Simulate a consumer WTP study: participants × product messages
n_participants <- 40
n_messages     <- 80

participant_effects <- rnorm(n_participants, 0, 15)   # people differ in baseline WTP
message_effects     <- rnorm(n_messages, 0, 12)        # messages differ in baseline WTP
discount_depth      <- runif(n_messages, 0, 40)        # discount depth predictor (0-40%)

df_promo <- expand_grid(
  participant = 1:n_participants,
  message     = 1:n_messages
) |>
  mutate(
    participant_eff = participant_effects[participant],
    message_eff     = message_effects[message],
    discount_depth  = discount_depth[message],
    wtp             = 80 + 0.8 * discount_depth +
                      participant_eff + message_eff +
                      rnorm(n(), 0, 15),
    participant     = factor(participant),
    message         = factor(message)
  )
```

```{r experiment-models}
# Without message random effects (wrong)
model_no_msgs <- lmer(wtp ~ discount_depth + (1 | participant),
                      data = df_promo)

# With message random effects (correct: crossed)
model_crossed <- lmer(wtp ~ discount_depth + (1 | participant) + (1 | message),
                      data = df_promo)

summary(model_crossed)
```

```{r experiment-variance}
broom.mixed::tidy(model_crossed, effects = "ran_pars") |>
  knitr::kable(digits = 2,
               caption = "Variance components: participants, messages, and residual")
```

**Reading the output:** The model decomposes variance into three sources: participant-level differences in WTP, message-level differences in appeal, and residual trial-level noise. The discount effect is estimated after accounting for *both* sources of clustering.

### Comparing: With and Without Message Random Effects

```{r experiment-se-comparison}
wrong_se   <- broom.mixed::tidy(model_no_msgs)[2, "std.error"][[1]]
correct_se <- broom.mixed::tidy(model_crossed)[2, "std.error"][[1]]

tibble(
  Model          = c("Without message RE", "With message RE (crossed)"),
  Discount_SE    = c(wrong_se, correct_se)
) |>
  knitr::kable(digits = 3,
    caption = "SE of discount effect: ignoring vs. modeling message-level variance")
```

When message-level variance is ignored, the standard error for the discount depth effect is smaller — the model treats message-level variance as residual noise, artificially deflating uncertainty. This produces anti-conservative inference: *p*-values are too small.

### The Assumptions of Crossed Random-Effects Models

::: {.callout-warning icon=false}
## Assumption 1: Independence of Participant and Message Random Effects

**What it means:** The random effect for participant $p$ and the random effect for message $s$ are assumed to be independent. Person 7 having an unusually high WTP is not related to whether message 42 is unusually compelling.

**How to check:** In a balanced fully-crossed design, this is usually reasonable. In unbalanced designs (not all participants see all messages), check whether missingness is systematic — for example, whether higher-WTP participants were more likely to complete high-discount evaluations.
:::

::: {.callout-warning icon=false}
## Assumption 2: Random Effects are Independent of the Predictor

**What it means:** The random intercepts for participants and messages should be uncorrelated with discount depth. Discount level should not predict which messages were seen by which participants — in a balanced design, it cannot.

**Why this is usually satisfied in experiments:** In a fully-crossed, counterbalanced design, every participant evaluates every message, so assignment is orthogonal to message characteristics. This is one reason well-designed experiments support random-effects modeling.
:::

::: {.callout-warning icon=false}
## Assumption 3: The Same Model Applies to All Participant–Message Pairs

**What it means:** The discount depth effect is assumed to be the same for all participants unless a random slope for discount depth by participants is included. If some participants are more deal-sensitive than others — for example, high price-consciousness individuals respond more strongly to deeper discounts — a random slope is needed.

**How to check:** Use likelihood ratio tests and theoretical reasoning.

```{r experiment-random-slopes}
model_maximal <- lmer(wtp ~ discount_depth + (discount_depth | participant) + (1 | message),
                      data    = df_promo,
                      control = lmerControl(optimizer = "bobyqa"))

anova(model_crossed, model_maximal)
```
:::

### The "Keep It Maximal" Debate

Barr et al. (2013) argued that you should always include the **maximal random effects structure** — every random slope that is theoretically justified by the design. This minimizes Type I error.

Bates et al. (2015) and Matuschek et al. (2017) pushed back, arguing that maximal models frequently fail to converge, and that parsimonious models often provide better estimates.

The current consensus: include random slopes for all within-unit manipulations (this is critical), but simplify the correlation structure or remove slopes not supported by the data if the model fails to converge.

In the context of the discount experiment: if every participant evaluates every message, discount depth is a within-participant factor, and a random slope for discount depth by participant is theoretically justified. Whether participants actually vary in their discount sensitivity is an empirical question answered by the LRT above.

This debate is covered in depth in Module 5.

---

## Part B: Crossed Random Effects in Longitudinal Panel Data

### Handling Month Effects in Customer Spending Panels

In a customer spending panel, every customer is observed across the same calendar months. Months are **not** nested within customers — instead, months and customers are **crossed**. Each observation belongs simultaneously to one customer AND one month.

The data-generating process has two sources of clustering:

1. **Customer random effects**: stable between-customer differences (some customers are always higher spenders — due to income, habits, or financial behavior)
2. **Month random effects**: seasonal patterns and economic conditions that affect **all** customers in a given month equally (year-end bonuses, holiday spending periods, macroeconomic shocks)

**Why this matters — if month random effects are ignored:**

- Month variance is absorbed into the residual
- Standard errors for predictors are too small (anti-conservative)
- Seasonal and economic patterns appear as noise when they are actually systematic, shared variance

### The Model

For observation $Y_{ct}$ from customer $c$ in month $t$:

$$
Y_{ct} = \underbrace{\gamma_{00}}_{\text{grand mean}} + \underbrace{\beta_1 \text{Promo}_{ct}}_{\text{predictor effect}} + \underbrace{u_{0c}}_{\text{customer RE}} + \underbrace{w_{0t}}_{\text{month RE}} + \underbrace{\varepsilon_{ct}}_{\text{residual}}
$$

Where:

- $u_{0c} \sim \mathcal{N}(0, \sigma^2_{\text{customer}})$: customer random intercepts (persistent spending differences)
- $w_{0t} \sim \mathcal{N}(0, \sigma^2_{\text{month}})$: month random intercepts (seasonal and economic month effects)
- $\varepsilon_{ct}$: customer-month idiosyncratic noise

The formal structure is identical to the product message model — only the substantive labels change.

### Simulation: 30 Customers × 15 Monthly Waves

```{r panel-sim}
n_customers <- 30
n_months    <- 15

customer_effects <- rnorm(n_customers, 0, 80)   # between-customer spending variance
month_effects    <- rnorm(n_months, 0, 40)       # seasonal/economic month shocks

df_panel <- expand_grid(customer = 1:n_customers, month = 1:n_months) |>
  mutate(
    promo_exposure = rnorm(n(), 20, 8),
    spending       = 600 + 0.4 * promo_exposure +
                     customer_effects[customer] +
                     month_effects[month] +
                     rnorm(n(), 0, 30),
    customer       = factor(customer),
    month_fac      = factor(month)
  )

glimpse(df_panel)
```

Notice: `month_effects` — the month-level seasonal and economic variation — is **not included as a predictor** in any model below. It is the hidden source of month-level clustering that the month random effect must absorb.

### Fitting Three Models

```{r panel-models}
# Model 1: Wrong — customer RE only; month variance goes into residual
m_customer_only <- lmer(spending ~ promo_exposure + (1 | customer),
                        data = df_panel)

# Model 2: Correct — crossed RE for customer and month
m_crossed       <- lmer(spending ~ promo_exposure + (1 | customer) + (1 | month_fac),
                        data = df_panel)

# Model 3: Conservative alternative — month fixed effects (month as dummy variables)
m_month_fe      <- lmer(spending ~ promo_exposure + (1 | customer) + month_fac,
                        data = df_panel)
```

### Variance Components: The Smoking Gun

```{r panel-variance-table}
vc_wrong   <- broom.mixed::tidy(m_customer_only, effects = "ran_pars")
vc_correct <- broom.mixed::tidy(m_crossed,       effects = "ran_pars")

bind_rows(
  vc_wrong   |> mutate(model = "Customer RE only (wrong)"),
  vc_correct |> mutate(model = "Crossed RE (correct)")
) |>
  select(model, group, term, estimate) |>
  knitr::kable(digits = 2,
               caption = "Variance components: customer-only vs. crossed specification")
```

In the crossed model, variance is correctly partitioned into customer-level variance, month-level variance, and residual. In the customer-only model, the residual variance is inflated — it has absorbed both the true residual and the month-level seasonal and economic shocks.

### The Key Diagnostic: Standard Error Comparison

```{r panel-se-comparison}
se_wrong   <- broom.mixed::tidy(m_customer_only)[2, "std.error"][[1]]
se_crossed <- broom.mixed::tidy(m_crossed)[2,     "std.error"][[1]]
se_fe      <- broom.mixed::tidy(m_month_fe)[2,    "std.error"][[1]]

tibble(
  Model             = c("Customer RE only (wrong)",
                         "Customer + Month RE (correct)",
                         "Customer RE + Month FE"),
  Promo_SE          = c(se_wrong, se_crossed, se_fe),
  Anti_conservative = c("Yes — too small", "Correct", "Correct")
) |>
  knitr::kable(digits = 4,
               caption = "SE of promotional exposure coefficient: month effects ignored vs. modeled")
```

The customer-only model understates uncertainty for the promotional exposure coefficient. When seasonal and economic month variance is left in the residual, the model mistakes correlated month-level noise for independent observations, producing an artificially small standard error.

### Visualizing the Problem: Residual Heatmaps

A powerful diagnostic: plot model residuals as a heatmap over the customer × month grid. Under the wrong model, seasonal and economic patterns create **vertical stripes** — in a high-spending month like December, all customers have positive residuals simultaneously. Under the correct model, residuals are patternless.

```{r panel-heatmap}
df_diag <- df_panel |>
  mutate(
    resid_wrong   = residuals(m_customer_only),
    resid_correct = residuals(m_crossed)
  )

p_wrong <- ggplot(df_diag, aes(x = month_fac, y = customer, fill = resid_wrong)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#d73027",
                       midpoint = 0, name = "Residual") +
  labs(title = "Wrong model (customer RE only)",
       subtitle = "Vertical stripes = month shocks pooled into residuals",
       x = "Month", y = "Customer") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_correct <- ggplot(df_diag, aes(x = month_fac, y = customer, fill = resid_correct)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#d73027",
                       midpoint = 0, name = "Residual") +
  labs(title = "Correct model (customer + month RE)",
       subtitle = "No vertical stripes — month shocks absorbed by month RE",
       x = "Month", y = "Customer") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_wrong / p_correct
```

The vertical stripes in the wrong model are the visual signature of an omitted crossed random effect. Each stripe represents a month in which seasonal conditions or economic factors drove spending unusually high or low — variance that is genuinely month-level but has been misattributed to the residual.

### Month Fixed Effects vs. Month Random Effects

In panel data, analysts often absorb month effects using **month dummy variables** (month fixed effects) rather than a random intercept. The choice has a precise statistical interpretation:

| Approach | When to use | Key assumption |
|---|---|---|
| Month random effects | Months are a random sample from a broader population of time periods | Normality of month effects; exchangeability |
| Month fixed effects | These specific months are the complete population of interest | No distributional assumption on month effects |
| Neither | Months have no systematic shared component | All month variance is idiosyncratic noise |

**Rule of thumb:** With $\geq 10$ months, random effects and fixed effects give similar coefficient estimates. With $< 5$ months, fixed effects are more conservative and generally preferred because there is too little information to estimate the month-level variance reliably.

**Connection to Module 1 Part 1 (Fixed Effects):** Month dummies in a fixed-effects model ARE month fixed effects. The choice between fixed effects and random effects for the month dimension mirrors the general FE vs. RE choice from Part 4 — it depends on whether you want to generalize to months outside your sample.

### The Assumptions of Crossed Panel Random-Effects Models

::: {.callout-warning icon=false}
## Assumption: Month Effects are Exchangeable

**What it means:** The month random intercepts $w_{0t}$ are assumed to be i.i.d. draws from $\mathcal{N}(0, \sigma^2_{\text{month}})$. This implies no autocorrelation across months — a high-spending month does not predict the next month will also be high.

**When it fails:** In the presence of multi-month trends (sustained seasonal spending cycles, economic expansions or contractions), month effects are correlated over time. The remedy is to include a linear time trend as a fixed effect, or to model AR(1) residual structure (covered in Module 4).
:::

::: {.callout-warning icon=false}
## Assumption: Customer and Month Effects are Independent

**What it means:** A customer being a high spender should not predict whether they were observed during high-spending or low-spending months. In panels with attrition (customers entering and exiting), this can fail — highly engaged customers may be more likely to transact consistently across all months, while low-engagement customers may drop out during low-promotion periods.

**How to check:** Examine whether customer-level characteristics (account age, income tier, product type) are systematically different across months of entry and exit. If customers disengage preferentially during certain seasonal periods, month random effects will be estimated from a selected sample.
:::

::: {.callout-warning icon=false}
## Assumption: Crossed Effects are Independent

**What it means:** The customer random effect and month random effect are assumed to be independent. This fails if certain types of customers are disproportionately observed in certain months (non-random panel composition over time).

**How to check:** Does the distribution of customer types change substantially over months in your panel? For example, if a platform promotion in month 8 attracts a new cohort of high-spending customers, the composition of the sample changes. The customer and month random effects are no longer independent.
:::

### Experimental vs. Panel: Side-by-Side Summary

| Dimension | Participant × Message (WTP Study) | Customer × Month (Spending Panel) |
|---|---|---|
| **Observation** | Participant $\times$ Message evaluation | Customer $\times$ Month observation |
| **Crossed factor 1** | Participant | Customer |
| **Crossed factor 2** | Message (stimulus) | Calendar month |
| **RE 1 absorbs** | Baseline WTP differences across people | Persistent spending differences across customers |
| **RE 2 absorbs** | Appeal differences across messages | Seasonal and economic patterns shared by all customers in a month |
| **If RE 2 omitted** | Message variance inflates residual; discount SE too small | Month variance inflates residual; promo exposure SE too small |
| **Predictor** | Discount depth (%) | Promotional exposure ($) |
| **Fix** | `(1\|participant) + (1\|message)` | `(1\|customer) + (1\|month_fac)` |

---

## Recommended Reading

| Topic | Citation |
|---|---|
| Subjects and items as crossed random effects | Judd, C. M., Westfall, J., & Kenny, D. A. (2012). Treating stimuli as a random factor in social psychology. *Journal of Personality and Social Psychology, 103*(1), 54–69. |
| Keep it maximal | Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language, 68*(3), 255–278. |
| Parsimonious random effects | Matuschek, H., Kliegl, R., Vasishth, S., Baayen, H., & Bates, D. (2017). Balancing Type I error and power in linear mixed models. *Journal of Memory and Language, 94*, 305–315. |
| Two-way fixed effects in panel data | Wooldridge, J. M. (2010). *Econometric Analysis of Cross Section and Panel Data* (2nd ed.). MIT Press. Chapter 10. |
