---
title: "Fixed Effects Models"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(plm)
library(ggplot2)
library(patchwork)
library(broom)
set.seed(2024)
```

# Fixed Effects Models

## Two Settings, One Problem

Fixed effects models are usually introduced in the context of panel data — the same firms tracked over multiple years, the same countries tracked across decades. But the exact same logic applies whenever observations are nested within units, *regardless of whether time is involved*.

Consider two studies:

**Study A (Longitudinal):** You survey 500 employees across 25 firms about autonomy and job satisfaction — and you do it again the following year. Each employee is measured twice.

**Study B (Experimental):** You run a within-subjects experiment. Each of 80 participants evaluates 12 product descriptions — 6 under a "scarcity" framing ("only 3 left!") and 6 under a neutral framing. Each participant sees both conditions.

Both studies share the same fundamental structure: **multiple observations nested within the same unit** (employee, participant). And both share the same fundamental problem: unit-level differences confound your estimate of the effect you care about.

In Study A, high-autonomy firms also happen to have better cultures — so the autonomy effect is confounded with culture.

In Study B, some participants are just more easily persuaded than others — so the framing effect is confounded with individual persuadability.

Fixed effects solve both problems in exactly the same way.

---

## The Model

Let $Y_{ij}$ be the outcome for observation $i$ within unit $j$ (employee $i$ in firm $j$; rating $i$ by participant $j$). The fixed-effects model is:

$$
\underbrace{Y_{ij}}_{\text{outcome}} = \underbrace{\alpha_j}_{\substack{\text{unit fixed} \\ \text{effect}}} + \underbrace{\beta_1}_{\substack{\text{within-unit} \\ \text{slope}}} \underbrace{X_{ij}}_{\substack{\text{within-unit} \\ \text{predictor}}} + \underbrace{\varepsilon_{ij}}_{\substack{\text{residual}}}
$$

The $\alpha_j$ is a **separate intercept for every unit** — a dummy variable for firm $j$, or a dummy variable for participant $j$. It absorbs *everything* that is constant within that unit: the firm's culture, the participant's baseline persuadability, the country's institutions.

The slope $\beta_1$ is estimated purely from **within-unit variation**: how does $Y$ change when $X$ changes *within the same unit*?

### The Within-Demeaning Transformation

An equivalent way to see what FE does is through demeaning. Let $\bar{Y}_j = \frac{1}{n_j}\sum_i Y_{ij}$ be the unit mean. Subtracting the unit mean from every observation:

$$
\underbrace{(Y_{ij} - \bar{Y}_j)}_{\substack{\text{outcome} \\ \text{demeaned}}} = \beta_1 \underbrace{(X_{ij} - \bar{X}_j)}_{\substack{\text{predictor} \\ \text{demeaned}}} + \underbrace{(\varepsilon_{ij} - \bar{\varepsilon}_j)}_{\substack{\text{residual} \\ \text{demeaned}}}
$$

After demeaning, every trace of between-unit variation is gone. The $\alpha_j$ terms vanish completely — they cancel out when you subtract the unit mean from both sides. What remains is the pure **within-unit relationship** between $X$ and $Y$.

In the firm study: we are comparing each employee to the *average* employee in their own firm.

In the experiment: we are comparing each participant's rating under scarcity to their *own average* rating — which is precisely what within-subjects experimental designs intend.

---

## Two Examples: Longitudinal and Experimental

### Example 1: Firms (Longitudinal Cross-Section)

```{r pooled-problem}
# --- STUDY A: Employees nested within firms ---
n_firms     <- 25
n_employees <- 20

# The confound: high-autonomy firms also have better culture
firm_autonomy_mean  <- rnorm(n_firms, 50, 10)
firm_culture_effect <- firm_autonomy_mean * 0.4 + rnorm(n_firms, 0, 5)

df_firms <- expand_grid(
  firm     = 1:n_firms,
  employee = 1:n_employees
) |>
  mutate(
    firm_autonomy = firm_autonomy_mean[firm],
    firm_culture  = firm_culture_effect[firm],
    autonomy      = firm_autonomy + rnorm(n(), 0, 5),
    # True WITHIN-firm effect = 0.2; but between-firm confound inflates it
    satisfaction  = 30 + 0.2 * autonomy + firm_culture + rnorm(n(), 0, 8),
    firm_label    = factor(firm)
  )

# Naive pooled regression
naive_firms <- lm(satisfaction ~ autonomy, data = df_firms)
broom::tidy(naive_firms) |>
  filter(term == "autonomy") |>
  knitr::kable(digits = 3,
    caption = "Pooled OLS: estimated effect of autonomy (confounded)")
```

### Example 2: Within-Subjects Experiment

```{r within-subjects-demo}
# --- STUDY B: Ratings nested within participants ---
n_participants <- 80
n_ratings      <- 12   # 6 scarcity, 6 neutral

# The confound: participants differ in baseline persuadability
persuadability <- rnorm(n_participants, 50, 10)

df_exp <- expand_grid(
  participant = 1:n_participants,
  rating_num  = 1:n_ratings
) |>
  mutate(
    condition    = rep(c("scarcity", "neutral"), n_ratings / 2)[rating_num],
    scarcity_bin = as.integer(condition == "scarcity"),
    # True WITHIN-person effect of scarcity = 4 points
    # But persuadability (a person-level trait) also drives ratings
    purchase_intent = persuadability[participant] + 4 * scarcity_bin +
                      rnorm(n(), 0, 8),
    participant_f   = factor(participant)
  )

# Naive pooled regression: mixes within-person and between-person variation
naive_exp <- lm(purchase_intent ~ scarcity_bin, data = df_exp)
broom::tidy(naive_exp) |>
  filter(term == "scarcity_bin") |>
  knitr::kable(digits = 3,
    caption = "Pooled OLS: estimated scarcity effect (ignores person differences)")
```

In the experiment, the pooled estimate is closer to the truth than in the firm study — because condition was randomly assigned, persuadability does not systematically correlate with condition. But random assignment only guarantees this *on average*; individual differences still add noise and can distort estimates in small samples or unbalanced designs.

```{r pooled-vs-within}
# Show both confound patterns side by side
firm_summary <- df_firms |>
  group_by(firm) |>
  summarise(mean_auto = mean(autonomy), mean_sat = mean(satisfaction))

p1 <- ggplot(df_firms, aes(x = autonomy, y = satisfaction)) +
  geom_point(alpha = 0.15, color = "#1e3a5f") +
  geom_smooth(method = "lm", color = "#e63946", linewidth = 1.3) +
  geom_point(data = firm_summary, aes(x = mean_auto, y = mean_sat),
             color = "#e63946", size = 3) +
  labs(title = "Study A: Firms",
       subtitle = "Red dots = firm means; slope is inflated by culture confound",
       x = "Autonomy", y = "Satisfaction") +
  theme_minimal(base_size = 12)

person_summary <- df_exp |>
  group_by(participant, condition) |>
  summarise(mean_intent = mean(purchase_intent), .groups = "drop")

p2 <- ggplot(df_exp, aes(x = scarcity_bin, y = purchase_intent,
                          color = participant_f, group = participant_f)) +
  geom_line(alpha = 0.3, linewidth = 0.5) +
  geom_point(alpha = 0.2, size = 0.8) +
  stat_summary(aes(group = 1), fun = mean, geom = "line",
               color = "#e63946", linewidth = 1.5) +
  scale_x_continuous(breaks = c(0, 1),
                     labels = c("Neutral", "Scarcity")) +
  labs(title = "Study B: Experiment",
       subtitle = "Each line = one participant across conditions",
       x = "Condition", y = "Purchase Intent") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  scale_color_viridis_d()

p1 + p2
```

## How Fixed Effects Solve This

Fixed effects recover the within-unit slope by including a dummy for every unit. In the firm study that is 25 firm dummies. In the experiment it is 80 participant dummies. Each dummy absorbs everything that is constant within that unit.

```{r fe-model}
# Study A: FE via plm
df_firms_pdata <- pdata.frame(df_firms, index = c("firm", "employee"))
fe_firms <- plm(satisfaction ~ autonomy,
                data   = df_firms_pdata,
                model  = "within",
                effect = "individual")

# Study B: FE via lm with participant dummies (equivalent)
fe_exp <- lm(purchase_intent ~ scarcity_bin + participant_f, data = df_exp)
```

```{r fe-compare}
# Compare naive vs. FE for both studies
bind_rows(
  broom::tidy(naive_firms)  |> filter(term == "autonomy") |>
    mutate(study = "Study A", model = "Pooled OLS", term = "Predictor"),
  broom::tidy(lm(satisfaction ~ autonomy + firm_label, data = df_firms)) |>
    filter(term == "autonomy") |>
    mutate(study = "Study A", model = "Fixed Effects", term = "Predictor"),
  broom::tidy(naive_exp)  |> filter(term == "scarcity_bin") |>
    mutate(study = "Study B", model = "Pooled OLS", term = "Predictor"),
  broom::tidy(fe_exp) |> filter(term == "scarcity_bin") |>
    mutate(study = "Study B", model = "Fixed Effects", term = "Predictor")
) |>
  select(study, model, estimate, std.error, p.value) |>
  knitr::kable(digits = 3,
    caption = "Pooled vs. Fixed Effects: Study A (true effect = 0.2), Study B (true effect = 4)")
```

::: {.callout-note}
## The DGP Connection

In Study A, the pooled estimate is badly confounded — it reflects both the true autonomy effect (0.2) *and* the between-firm culture effect. Fixed effects remove the latter, recovering the within-unit DGP.

In Study B, the pooled estimate is closer to the truth because random assignment broke the correlation between condition and persuadability. But fixed effects still improve efficiency and explicitly answer the within-person question: for the same participant, does scarcity increase intent?

Both uses are valid. The experimental use is often overlooked — but within-subjects ANOVA is, mathematically, a fixed-effects model.
:::

---

## The Assumptions of Fixed Effects Models

Each assumption is framed for *both* a longitudinal/observational setting and an experimental/repeated-measures setting.

::: {.callout-warning icon=false}
## Assumption 1: Strict Exogeneity

**The equation implication:** For the within-demeaned estimator to be unbiased, we require:

$$E[\varepsilon_{ij} \mid X_{i1}, X_{i2}, \ldots, X_{in_j}, \alpha_j] = 0$$

In words: the residuals must be uncorrelated with the predictor **at every observation point** — past, present, and future — after conditioning on the fixed effect.

**In longitudinal data:** This fails if the outcome feeds back into the predictor. If a firm's declining satisfaction causes it to *increase* autonomy in the next period, then autonomy at time $t+1$ is correlated with the residual at time $t$. The predictor is no longer exogenous.

**In experiments:** Strict exogeneity is typically satisfied by design — random assignment ensures that condition is not caused by the outcome. But it can fail in carryover designs where the treatment in one block affects the outcome in a later block (i.e., order effects or fatigue). Counterbalancing and washout periods address this.

**How to check:** In longitudinal data, plot residuals against *future* values of the predictor. In experiments, check for order effects by including trial number as a covariate.
:::

::: {.callout-warning icon=false}
## Assumption 2: No Time-Varying (or Trial-Varying) Confounds

**What it means:** Fixed effects only remove *stable* unit-level confounds. If a confound changes across observations within the same unit, the fixed effect cannot absorb it.

$$\alpha_j = \text{constant over all } i \text{ within unit } j$$

If the unit's "true level" shifts — a firm's culture changes after a merger; a participant becomes fatigued midway through a session — the single $\alpha_j$ is no longer a valid representation.

**In longitudinal data:** Any policy, economic shock, or personal event that co-occurs with your predictor is a time-varying confound. Example: if firms that grant more autonomy simultaneously adopt flexible hours, and flexible hours boost satisfaction, FE does not remove the flexible hours effect.

**In experiments:** Practice effects, fatigue, and demand characteristics that change systematically across trials. If participants become better at evaluating products as the experiment progresses (learning), and if scarcity items happen to appear later in the session, this is a trial-varying confound.

**How to check:** Include trial order or time as a covariate. Check whether your substantive estimates change when you control for order.
:::

::: {.callout-warning icon=false}
## Assumption 3: Sufficient Within-Unit Variation

**What it means:** Fixed effects can only estimate $\beta_1$ from within-unit variation. If the predictor barely changes within units, the estimator is imprecise:

$$\widehat{\beta}_1 = \frac{\sum_j \sum_i (X_{ij} - \bar{X}_j)(Y_{ij} - \bar{Y}_j)}{\sum_j \sum_i (X_{ij} - \bar{X}_j)^2}$$

The denominator — total within-unit variation in $X$ — appears in the denominator of the variance of $\hat\beta_1$. Small within-unit variation → large standard errors.

**In longitudinal data:** A predictor like firm size barely changes within a firm from year to year. FE is poorly suited to estimating the effect of stable characteristics.

**In experiments:** This is usually not an issue because condition is deliberately manipulated. But if the manipulation is very small (e.g., a subtle framing tweak that most participants do not notice), within-unit variation in the *effective* treatment is low.

```{r variation-check}
# Study A: how much within-firm variation does autonomy have?
df_firms |>
  group_by(firm) |>
  summarise(within_sd = sd(autonomy)) |>
  summarise(
    mean_within_sd   = mean(within_sd),
    total_sd         = sd(df_firms$autonomy),
    within_fraction  = mean(within_sd) / sd(df_firms$autonomy)
  ) |>
  knitr::kable(digits = 3,
    caption = "Study A: fraction of autonomy variance that is within-firm")
```
:::

::: {.callout-warning icon=false}
## Assumption 4: Fixed Effects Are Stable (One Intercept Per Unit)

**What it means:** The model assigns each unit exactly one intercept $\alpha_j$. This assumes the unit's baseline is constant across all its observations. The model equation is:

$$Y_{ij} = \alpha_j + \beta_1 X_{ij} + \varepsilon_{ij}$$

and $\alpha_j$ is the **same** regardless of which observation $i$ we are looking at within unit $j$.

**In longitudinal data:** If a firm's culture changes dramatically after a merger halfway through the study, the single $\alpha_j$ no longer accurately represents either period. The early-period and late-period intercepts are different, but the model conflates them.

**In experiments:** If a participant's attention or mood changes substantially mid-session (and this change is not modeled), the stable-intercept assumption is violated. Split-session designs or attention checks help detect this.

**How to check:** For longitudinal data, split the sample at the midpoint and compare unit intercepts across periods. For experiments, split the trials into first-half and second-half blocks and test whether the effect estimate shifts.
:::

::: {.callout-warning icon=false}
## Assumption 5: Homoskedasticity and Independence of Residuals Within Units

**What it means:** After conditioning on fixed effects, the residuals $\varepsilon_{ij}$ are assumed to be independent and identically distributed:

$$\varepsilon_{ij} \overset{iid}{\sim} \mathcal{N}(0,\ \sigma^2_\varepsilon)$$

Two components: (1) constant variance across units and observations (homoskedasticity), and (2) no correlation between residuals of the same unit across observations (independence).

**In longitudinal data:** Serial correlation is nearly universal. Today's mood affects tomorrow's mood, so residuals at $t$ and $t+1$ are correlated. Heteroskedasticity also appears if variance changes over time (e.g., outcomes become more variable in later periods).

**In experiments:** Autocorrelation is less severe for blocked designs (participants evaluate each product once), but can appear in trial-level data with carryover. Heteroskedasticity is possible if some conditions produce more variable responses than others.

**What to do:** Use cluster-robust standard errors, which are valid even under heteroskedasticity and within-unit correlation.

```{r exogeneity-check}
fe_lm_firms <- lm(satisfaction ~ autonomy + firm_label, data = df_firms)
df_firms <- df_firms |> mutate(fe_resid = residuals(fe_lm_firms))

ggplot(df_firms, aes(x = autonomy, y = fe_resid)) +
  geom_point(alpha = 0.25, color = "#1e3a5f") +
  geom_smooth(method = "loess", color = "#e63946", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title    = "Study A: FE residuals vs. autonomy",
    subtitle = "No pattern expected if exogeneity holds",
    x = "Autonomy", y = "FE Residual"
  ) +
  theme_minimal(base_size = 13)
```

```{r robust-se}
library(sandwich)
library(lmtest)

# Cluster-robust standard errors for Study A
coeftest(fe_lm_firms,
         vcov = vcovCL(fe_lm_firms, cluster = ~firm_label))["autonomy", , drop = FALSE] |>
  knitr::kable(digits = 3,
    caption = "Study A autonomy coefficient with cluster-robust SEs")

# For Study B, cluster by participant
coeftest(fe_exp,
         vcov = vcovCL(fe_exp, cluster = ~participant_f))["scarcity_bin", , drop = FALSE] |>
  knitr::kable(digits = 3,
    caption = "Study B scarcity coefficient with cluster-robust SEs")
```
:::

---

## What Fixed Effects Cannot Do

Fixed effects are powerful but have important limits that apply equally in both settings:

1. **They cannot estimate unit-level predictors.** In Study A, you cannot estimate the effect of firm industry (constant within firm). In Study B, you cannot estimate the effect of participant gender (constant within participant). All between-unit variation is absorbed by the $\alpha_j$ terms.

2. **They discard between-unit information.** If the between-unit relationship is scientifically interesting — whether more autonomous firms differ from less autonomous ones — FE discards that comparison. This is the price of controlling for all stable unit-level confounds.

3. **They assume a common slope.** The model estimates one $\beta_1$ for everyone. If the effect of autonomy on satisfaction differs across firms, or if the scarcity effect differs across participants, FE averages over this heterogeneity. Random slopes (Module 1, Part 3) address this.

4. **They are not well suited to purely between-subjects designs.** If every participant sees only one condition (between-subjects), there is no within-person variation in the predictor — FE collapses entirely to OLS. The fixed-effects logic requires that the predictor varies within units.

---

## Recommended Reading

| Topic | Citation |
|-------|----------|
| Fixed effects and panel data | Wooldridge, J. M. (2010). *Econometric Analysis of Cross Section and Panel Data* (2nd ed.). MIT Press. (Chapter 10) |
| FE in experimental psychology | Judd, C. M., Westfall, J., & Kenny, D. A. (2012). Treating stimuli as a random factor in social psychology. *Journal of Personality and Social Psychology, 103*(1), 54–69. |
| Two-way fixed effects | Imai, K., & Kim, I. S. (2021). On the use of two-way fixed effects regression models for causal inference with panel data. *Political Analysis, 29*(3), 405–415. |
| Cluster-robust standard errors | Cameron, A. C., & Miller, D. L. (2015). A practitioner's guide to cluster-robust inference. *Journal of Human Resources, 50*(2), 317–372. |
