---
title: "Random Slopes"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(ggplot2)
set.seed(42)
```

# Random Slopes

## When the Effect Varies Across Units

The random-intercept model assumes that the *slope* — the relationship between the predictor and the outcome — is identical across all units. Everyone's mood responds to sleep in exactly the same way; only their baseline mood differs.

But is that realistic?

Some people are highly sensitive to sleep: one bad night and they feel terrible. Others are remarkably resilient: they function almost the same regardless of how much they sleep. The *magnitude* of the sleep-mood relationship varies across people. If we force a single slope for everyone, we are misrepresenting the data-generating process.

The **random-slope model** allows the slope to vary across units, treating individual slopes as draws from a distribution.

## The Model

**Level 1:**
$$Y_{ij} = \beta_{0j} + \beta_{1j} X_{ij} + \varepsilon_{ij}$$

**Level 2:**
$$\beta_{0j} = \gamma_{00} + u_{0j}$$
$$\beta_{1j} = \gamma_{10} + u_{1j}$$

Combined:
$$Y_{ij} = \gamma_{00} + \gamma_{10} X_{ij} + u_{0j} + u_{1j} X_{ij} + \varepsilon_{ij}$$

The new component is $u_{1j}$: each unit's deviation from the average slope. We estimate $\gamma_{10}$ (the average slope), $\sigma^2_{u_0}$ (variance in intercepts), $\sigma^2_{u_1}$ (variance in slopes), and crucially — $\sigma_{u_0 u_1}$ (the correlation between intercepts and slopes).

```{r fit-random-slopes}
n_people <- 40
n_days   <- 7

person_intercepts <- rnorm(n_people, mean = 50, sd = 8)
person_slopes     <- rnorm(n_people, mean = 3, sd = 2.5)  # slopes vary!

df <- expand_grid(
  person = 1:n_people,
  day    = 1:n_days
) |>
  mutate(
    sleep  = rnorm(n(), mean = 7, sd = 1.5),
    mood   = person_intercepts[person] +
             person_slopes[person] * (sleep - 7) +
             rnorm(n(), 0, 5),
    person = factor(person)
  )

# Random intercept only
model_ri <- lmer(mood ~ sleep + (1 | person), data = df)

# Random intercept + slope
model_rs <- lmer(mood ~ sleep + (sleep | person), data = df)

# Compare
broom.mixed::tidy(model_rs, effects = "ran_pars") |>
  knitr::kable(digits = 3, caption = "Variance components: random intercepts and slopes")
```

```{r visualize-slopes}
# Show individual slopes
person_lines <- df |>
  group_by(person) |>
  summarise(
    intercept = coef(lm(mood ~ sleep, data = cur_data()))[[1]],
    slope     = coef(lm(mood ~ sleep, data = cur_data()))[[2]],
    .groups   = "drop"
  )

# Plot individual regression lines
sleep_range <- seq(4, 10, by = 0.1)

lines_df <- person_lines |>
  mutate(
    x = list(sleep_range),
    y = map2(intercept, slope, ~ .x + .y * (sleep_range - 7))
  ) |>
  unnest(c(x, y)) |>
  rename(sleep = x, mood = y)

ggplot() +
  geom_line(data = lines_df,
            aes(x = sleep, y = mood, group = person),
            color = "#1e3a5f", alpha = 0.3, linewidth = 0.5) +
  geom_abline(
    slope     = fixef(model_rs)["sleep"],
    intercept = fixef(model_rs)["(Intercept)"] - fixef(model_rs)["sleep"] * 7,
    color     = "#e63946", linewidth = 2
  ) +
  labs(
    title    = "Random slopes: each person's sleep-mood relationship",
    subtitle = "Red line = average slope (fixed effect); grey lines = individual slopes",
    x        = "Hours of sleep",
    y        = "Mood"
  ) +
  xlim(4.5, 9.5) +
  theme_minimal(base_size = 13)
```

The grey lines show enormous variation in how strongly sleep predicts mood across people. The red line — the fixed effect — is the average story. But for some people, the relationship is strong; for others, nearly flat. A model that forces a single slope misses this heterogeneity entirely.

## The Intercept-Slope Correlation

The random-slope model also estimates the correlation between intercepts and slopes: $\rho_{u_0, u_1}$.

This is substantively meaningful. A positive correlation would mean people with higher baseline mood also benefit more from sleep. A negative correlation would mean people with low baseline mood are more sensitive to sleep deprivation.

```{r slope-intercept-corr}
# Plot the intercept-slope relationship
ggplot(person_lines, aes(x = intercept, y = slope)) +
  geom_point(size = 3, color = "#1e3a5f", alpha = 0.7) +
  geom_smooth(method = "lm", color = "#e63946", se = TRUE) +
  labs(
    title    = "Intercept-slope correlation",
    subtitle = "Is the sleep effect stronger for people with higher or lower baseline mood?",
    x        = "Individual intercept (baseline mood)",
    y        = "Individual slope (sleep effect)"
  ) +
  theme_minimal(base_size = 13)
```

## The Assumptions of Random-Slope Models

::: {.callout-warning icon=false}
## Assumption 1: Multivariate Normality of Random Effects

**What it means:** The random intercepts and slopes are jointly drawn from a bivariate normal distribution:
$$\begin{pmatrix} u_{0j} \\ u_{1j} \end{pmatrix} \sim \mathcal{N}\left(\begin{pmatrix} 0 \\ 0 \end{pmatrix}, \begin{pmatrix} \sigma^2_{u_0} & \sigma_{u_0 u_1} \\ \sigma_{u_0 u_1} & \sigma^2_{u_1} \end{pmatrix}\right)$$

**How to check:** Plot the estimated random effects (from `ranef()`) as a scatterplot. Look for elliptical distributions, not banana-shaped or clustered patterns.

```{r ranef-scatter}
re_df <- as.data.frame(ranef(model_rs)$person)
names(re_df) <- c("intercept", "slope")

ggplot(re_df, aes(x = intercept, y = slope)) +
  geom_point(size = 3, color = "#1e3a5f", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60") +
  labs(
    title    = "Bivariate distribution of random effects",
    subtitle = "Should look elliptical for bivariate normality",
    x        = "Random intercept",
    y        = "Random slope"
  ) +
  theme_minimal(base_size = 13)
```
:::

::: {.callout-warning icon=false}
## Assumption 2: The Correct Random Effects Structure is Specified

**What it means:** The model must include random slopes for every predictor whose effect varies meaningfully across units. Failing to include a necessary random slope biases standard errors downward — producing anti-conservatively small p-values.

**Judd, Westfall, & Kenny (2012, 2017)** showed that omitting random slopes that should be in the model is one of the most common sources of Type I error inflation in experimental psychology, particularly when stimuli are treated as fixed rather than random.

**How to check:** Likelihood ratio tests (LRT) comparing models with and without the random slope. If the LRT is significant, the random slope is needed.

```{r lrt-slopes}
# Test: does adding the random slope significantly improve fit?
model_ri_refit <- lmer(mood ~ sleep + (1 | person), data = df, REML = FALSE)
model_rs_refit <- lmer(mood ~ sleep + (sleep | person), data = df, REML = FALSE)

anova(model_ri_refit, model_rs_refit)
```

Note: use `REML = FALSE` when comparing models with likelihood ratio tests. Switch back to `REML = TRUE` for final parameter estimation.
:::

::: {.callout-warning icon=false}
## Assumption 3: Convergence

**What it means:** Complex random-slope models — especially with many predictors and crossed random effects — frequently fail to converge. A convergence warning from `lme4` means the optimization algorithm did not find a stable minimum. Estimates from non-converged models should not be trusted.

**How to check:** Look for convergence warnings in `lmer()` output. Use `check_convergence()` from the `performance` package.

**What to do if it fails:**
- Rescale continuous predictors (center and divide by SD)
- Simplify the random effects structure (remove the correlation between intercept and slope: use `(1 | person) + (0 + sleep | person)`)
- Try alternative optimizers: `control = lmerControl(optimizer = "bobyqa")`
- Switch to Bayesian estimation (`brms`), which is more robust for complex models

```{r convergence-check}
library(performance)
check_convergence(model_rs)
```
:::

## Model Comparison: Do You Need Random Slopes?

Adding random slopes adds parameters and complexity. You should include them when:

1. **Theory predicts heterogeneous effects:** Do you have a reason to believe the relationship varies across units?
2. **The LRT supports them:** The model with random slopes fits significantly better than the model without.
3. **The design requires them:** In within-subject experimental designs, random slopes for experimental conditions are almost always needed (see Module 5).

## Recommended Reading

| Topic | Citation |
|-------|----------|
| Random slopes in experimental designs | Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language, 68*(3), 255–278. |
| When random slopes matter | Heisig, J. P., & Schaeffer, M. (2019). Why you should always include a random slope for the lower-level variable involved in a cross-level interaction. *European Sociological Review, 35*(2), 258–279. |
