---
title: "Random Slopes"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(performance)
library(ggplot2)
library(patchwork)
set.seed(42)
```

# Random Slopes

::: {.callout-note icon=false}
## When the Average Effect Hides the Most Interesting Story

- **Question:** If the random-intercept model only lets units start at different levels, what happens when the *effect* of a predictor also varies across units — when disclosure motivates some households strongly but leaves others cold?
- **What goes wrong:** A model with random intercepts but a fixed slope estimates one disclosure effect for everyone. That single number is the *average*, but the average may describe no individual household accurately. Omitting the random slope also pushes treatment heterogeneity into the residual, deflating standard errors for the average effect.
- **The fix:** Random slopes add a unit-specific deviation to the slope — estimating the distribution of effects across units, not just the mean. The result is correct standard errors, richer interpretation, and a direct measure of treatment effect heterogeneity.
:::

## When Effects Vary Across Units

The random-intercept model assumes that the *slope* — the relationship between a predictor and the outcome — is identical across all units. Only the baseline level differs from unit to unit. This is often an unrealistic assumption.

Consider two research settings:

**Longitudinal panel data.** A financial coaching program tracks households across monthly waves. The average effect of disclosure engagement on savings may be positive — households that read more of the disclosure tend to save more that month. But some households are highly responsive: reading one extra page pushes them to act on the advice. Others are dismissive: they read the same pages but do not change their behaviour. The slope of disclosure → savings *varies across households*.

**Within-subjects experiments.** A price-framing study asks participants to complete WTP tasks under either a "high anchor" or "neutral" framing. On average, the anchor condition increases WTP. But individual responses to the manipulation vary considerably — some participants are strongly anchored, others barely register a difference. The *treatment effect itself* is heterogeneous across participants.

The **random-slope model** addresses this directly. Rather than fixing one slope for everyone, we treat each unit's slope as a draw from a distribution, allowing us to estimate:

- The **average** effect across units (the fixed slope, $\gamma_{10}$)
- The **variability** of that effect across units (the random slope variance, $\sigma^2_{u_1}$)
- The **correlation** between where units start and how strongly they respond ($\rho_{u_0, u_1}$)

## The Model

**Level 1** (observations within units):
$$Y_{ij} = \beta_{0j} + \beta_{1j} X_{ij} + \varepsilon_{ij}$$

**Level 2** (units):
$$\beta_{0j} = \gamma_{00} + u_{0j}$$
$$\beta_{1j} = \gamma_{10} + u_{1j}$$

**Combined form:**
$$Y_{ij} = \underbrace{\gamma_{00}}_{\text{grand mean}} + \underbrace{\gamma_{10} X_{ij}}_{\text{avg slope}} + \underbrace{u_{0j}}_{\text{unit intercept}} + \underbrace{u_{1j} X_{ij}}_{\text{unit slope deviation}} + \varepsilon_{ij}$$

The random effects are assumed to follow a bivariate normal distribution:
$$\begin{pmatrix} u_{0j} \\ u_{1j} \end{pmatrix} \sim \mathcal{N}\left(\begin{pmatrix} 0 \\ 0 \end{pmatrix}, \begin{pmatrix} \sigma^2_{u_0} & \sigma_{u_0 u_1} \\ \sigma_{u_0 u_1} & \sigma^2_{u_1} \end{pmatrix}\right)$$

**What does $u_{1j}$ mean in each context?**

- **Longitudinal:** $u_{1j}$ is household $j$'s deviation from the average disclosure sensitivity. A large positive $u_{1j}$ means this household responds unusually strongly to disclosure — each page read produces a steep increase in savings. A near-zero $u_{1j}$ means disclosure has its typical (average) effect on this household.

- **Experimental:** $u_{1j}$ is participant $j$'s deviation from the average anchoring effect. A large $u_{1j}$ signals that this individual is strongly anchored — the framing shifts their WTP far more than average. A near-zero $u_{1j}$ means the manipulation barely affected them.

## Longitudinal Example — Heterogeneous Response to Financial Disclosure

**Design:** 50 households measured across 8 monthly waves. The predictor is monthly disclosure engagement (pages read, centered at its grand mean of 6). The outcome is monthly savings rate (%). Each household has their own disclosure sensitivity — modeled as a random slope.

The simulation imposes a correlation of −0.3 between intercepts and slopes: households that start with high baseline savings are somewhat less sensitive to disclosure (perhaps because they are already well-informed and the marginal effect of additional reading is smaller).

```{r longitudinal-data}
n_households <- 50
n_waves      <- 8

# Correlated random intercepts and slopes: SD_int = 8, SD_slope = 0.8, rho = -0.3
Sigma_long <- matrix(c(64, -1.92, -1.92, 0.64), nrow = 2)
re_mat     <- MASS::mvrnorm(n_households, mu = c(0, 0), Sigma = Sigma_long)

df_diary <- expand_grid(
  household = 1:n_households,
  wave      = 1:n_waves
) |>
  mutate(
    disclosure   = rnorm(n(), mean = 6, sd = 2),
    disclosure_c = disclosure - 6,
    savings      = (25 + re_mat[household, 1]) +
                   (2 + re_mat[household, 2]) * disclosure_c +
                   rnorm(n(), 0, 5),
    household    = factor(household)
  )

# Random intercept only
model_ri_diary <- lmer(savings ~ disclosure_c + (1 | household),
                       data = df_diary, REML = TRUE)

# Random intercept + slope
model_rs_diary <- lmer(savings ~ disclosure_c + (disclosure_c | household),
                       data = df_diary, REML = TRUE)

broom.mixed::tidy(model_rs_diary, effects = "ran_pars") |>
  knitr::kable(digits = 3,
               caption = "Variance components: disclosure → savings random slope model")
```

The variance component for `disclosure_c` at the household level ($\hat\sigma^2_{u_1}$) captures how much individual sensitivity to disclosure varies. The correlation term estimates the intercept-slope association.

```{r longitudinal-slopes-plot}
# Extract OLS estimates per household for visualization
household_lines <- df_diary |>
  group_by(household) |>
  summarise(
    intercept = coef(lm(savings ~ disclosure_c))[[1]],
    slope     = coef(lm(savings ~ disclosure_c))[[2]],
    .groups   = "drop"
  )

disc_range <- seq(-4, 4, by = 0.5)

lines_df_long <- household_lines |>
  mutate(
    x = list(disc_range),
    y = map2(intercept, slope, ~ .x + .y * disc_range)
  ) |>
  unnest(c(x, y)) |>
  rename(disclosure_c = x, savings = y)

p_long_slopes <- ggplot() +
  geom_line(data = lines_df_long,
            aes(x = disclosure_c, y = savings, group = household),
            color = "#1e3a5f", alpha = 0.3, linewidth = 0.4) +
  geom_abline(
    slope     = fixef(model_rs_diary)["disclosure_c"],
    intercept = fixef(model_rs_diary)["(Intercept)"],
    color     = "#e63946", linewidth = 1.8
  ) +
  labs(
    title    = "Longitudinal: disclosure → savings",
    subtitle = "Each line = one household; red = average fixed slope",
    x        = "Disclosure engagement (centered)",
    y        = "Monthly savings rate (%)"
  ) +
  theme_minimal(base_size = 12)

p_long_slopes
```

The grey lines reveal substantial variation in disclosure sensitivity. Most households show the expected positive slope, but the *steepness* varies considerably. A few households show near-flat slopes — for them, disclosure engagement may reflect motivation that exists regardless of the specific advice read.

**Intercept-slope correlation:** A negative correlation means households with higher baseline savings tend to show a *weaker* positive disclosure effect. High-saving households may already be well-informed, so additional reading yields diminishing returns. Conversely, low-saving households may have the most to gain from each additional page.

## Experimental Example — Anchoring and WTP

**Design:** 60 participants complete 12 WTP tasks each (6 in the high-anchor condition, 6 in the neutral condition, counterbalanced). The predictor is framing condition (anchor = 1, neutral = 0). The outcome is WTP (0–100 scale). Each participant is allowed their own anchoring sensitivity.

The simulation imposes a positive correlation of 0.2 between intercepts and slopes: participants with higher baseline WTP show a slightly stronger anchoring effect (the anchor has more "room" to pull their estimates upward).

```{r experimental-data}
n_participants <- 60
n_trials_each  <- 6   # per condition

# Correlated intercepts and slopes: SD_int = 10, SD_slope = 3, rho = 0.2
Sigma_exp  <- matrix(c(100, 6, 6, 9), nrow = 2)
re_mat_exp <- MASS::mvrnorm(n_participants, mu = c(0, 0), Sigma = Sigma_exp)

df_exp <- expand_grid(
  participant = 1:n_participants,
  condition   = c(0, 1),
  trial       = 1:n_trials_each
) |>
  mutate(
    wtp             = (55 + re_mat_exp[participant, 1]) +
                      (8 + re_mat_exp[participant, 2]) * condition +
                      rnorm(n(), 0, 6),
    participant     = factor(participant),
    condition_label = factor(condition, labels = c("Neutral", "Anchor"))
  )

# Random intercept only
model_ri_exp <- lmer(wtp ~ condition + (1 | participant),
                     data = df_exp, REML = TRUE)

# Random intercept + slope for condition
model_rs_exp <- lmer(wtp ~ condition + (condition | participant),
                     data = df_exp, REML = TRUE)

broom.mixed::tidy(model_rs_exp, effects = "ran_pars") |>
  knitr::kable(digits = 3,
               caption = "Variance components: anchor framing condition random slope model")
```

```{r experimental-ranef-histogram}
# Extract per-participant anchoring effects (random slope BLUPs + fixed effect)
re_exp     <- ranef(model_rs_exp)$participant
names(re_exp) <- c("u0", "u1")
avg_effect <- fixef(model_rs_exp)["condition"]

re_exp <- re_exp |>
  mutate(person_anchor_effect = avg_effect + u1)

p_hist <- ggplot(re_exp, aes(x = person_anchor_effect)) +
  geom_histogram(bins = 20, fill = "#457b9d", color = "white", alpha = 0.85) +
  geom_vline(xintercept = avg_effect, color = "#e63946",
             linewidth = 1.5, linetype = "dashed") +
  annotate("text", x = avg_effect + 0.4, y = Inf,
           label = "avg effect", color = "#e63946",
           vjust = 2, hjust = 0, size = 3.5) +
  labs(
    title    = "Distribution of individual anchoring effects on WTP",
    subtitle = "Each bar = how many participants experienced that effect size",
    x        = "Person-level anchoring effect on WTP",
    y        = "Count"
  ) +
  theme_minimal(base_size = 12)

p_hist
```

Some participants show almost no anchoring effect; others show a WTP increase exceeding +20 points. The random slope captures this theoretically important heterogeneity — heterogeneity that is invisible in a model that forces the same effect on everyone.

```{r experimental-slopes-plot}
participant_lines <- df_exp |>
  group_by(participant) |>
  summarise(
    intercept = coef(lm(wtp ~ condition))[[1]],
    slope     = coef(lm(wtp ~ condition))[[2]],
    .groups   = "drop"
  )

p_exp_slopes <- ggplot() +
  geom_segment(
    data = participant_lines,
    aes(x = 0, xend = 1,
        y = intercept,
        yend = intercept + slope,
        group = participant),
    color = "#457b9d", alpha = 0.4, linewidth = 0.5
  ) +
  geom_segment(
    aes(x = 0, xend = 1,
        y = fixef(model_rs_exp)["(Intercept)"],
        yend = fixef(model_rs_exp)["(Intercept)"] + fixef(model_rs_exp)["condition"]),
    color = "#e63946", linewidth = 2
  ) +
  scale_x_continuous(breaks = c(0, 1), labels = c("Neutral", "Anchor")) +
  labs(
    title    = "Experimental: anchor framing → WTP",
    subtitle = "Each line = one participant's framing effect; red = average",
    x        = "Condition",
    y        = "WTP (0–100)"
  ) +
  theme_minimal(base_size = 12)

p_exp_slopes
```

::: {.callout-warning icon=false}
## Key Point: Random Slopes and Type I Error in Experiments

**Barr, Levy, Scheepers, & Tily (2013)** demonstrated that omitting random slopes for within-subjects conditions inflates Type I error rates substantially. The mechanism: if you do not model variation in the treatment effect across participants, that variation is absorbed into the residual — deflating standard errors and making the average effect appear more precise than it is.

**The recommendation:** In within-subjects designs, include random slopes for all within-subjects factors by default (the "maximal" random effects structure). The random slope is not merely a nuisance parameter — it *is* the treatment effect heterogeneity, which is often scientifically interesting in its own right.
:::

## Comparing the Two Examples Side by Side

```{r side-by-side}
p_long_slopes + p_exp_slopes +
  plot_annotation(
    title   = "Random slopes: heterogeneous effects in two research designs",
    caption = "Red lines/segments = fixed (average) effects; grey lines = individual unit slopes"
  )
```

## The Intercept-Slope Correlation

The random-slope model estimates not just variance in intercepts and variance in slopes, but also their **covariance** — equivalently, the correlation $\rho_{u_0, u_1}$.

This correlation is substantively meaningful in both contexts.

**Longitudinal interpretation.** A negative intercept-slope correlation means: households with *higher* baseline savings tend to show a *weaker* positive disclosure effect. High-saving households may already implement advice regardless of how much they read — diminishing returns to disclosure. Conversely, low-saving households may have the most to gain from each additional page.

**Experimental interpretation.** A positive intercept-slope correlation means: participants with *higher* baseline WTP show a *stronger* anchoring effect. One interpretation: the anchor provides a reference that is proportionally more relevant when a participant already has a high willingness to pay — more room to be pulled upward.

```{r intercept-slope-corr}
p_corr_long <- ggplot(household_lines, aes(x = intercept, y = slope)) +
  geom_point(size = 2.5, color = "#1e3a5f", alpha = 0.7) +
  geom_smooth(method = "lm", color = "#e63946", se = TRUE, linewidth = 1) +
  labs(
    title    = "Longitudinal",
    subtitle = "Negative: high-baseline households less disclosure-sensitive",
    x        = "Intercept (baseline savings rate)",
    y        = "Slope (disclosure effect)"
  ) +
  theme_minimal(base_size = 12)

p_corr_exp <- ggplot(participant_lines, aes(x = intercept, y = slope)) +
  geom_point(size = 2.5, color = "#457b9d", alpha = 0.7) +
  geom_smooth(method = "lm", color = "#e63946", se = TRUE, linewidth = 1) +
  labs(
    title    = "Experimental",
    subtitle = "Positive: high-baseline WTP participants more anchor-sensitive",
    x        = "Intercept (neutral condition WTP)",
    y        = "Slope (anchoring effect)"
  ) +
  theme_minimal(base_size = 12)

p_corr_long + p_corr_exp +
  plot_annotation(title = "Intercept-slope correlations in each design")
```

## The Zero-Correlation Parameterization

By default, `(predictor | group)` estimates random slope variance *and* the intercept-slope correlation. When the correlation is near zero, theoretically uninteresting, or causing convergence problems, the **zero-correlation parameterization** separates the two:

```r
# Full: estimates intercept variance, slope variance, AND their correlation
lmer(savings ~ disclosure_c + (disclosure_c | household), data = df_diary)

# Zero-correlation: estimates both variances, forces correlation to zero
lmer(savings ~ disclosure_c + (1 | household) + (0 + disclosure_c | household), data = df_diary)
```

Applied to the experimental data:

```{r zero-corr-example}
model_zcp_exp <- lmer(wtp ~ condition +
                        (1 | participant) + (0 + condition | participant),
                      data = df_exp, REML = FALSE)

# Staged LRT: (1) does the random slope variance matter?
#             (2) does the correlation additionally matter?
model_ri_only_ml <- lmer(wtp ~ condition + (1 | participant),
                         data = df_exp, REML = FALSE)
model_zcp_ml     <- lmer(wtp ~ condition +
                           (1 | participant) + (0 + condition | participant),
                         data = df_exp, REML = FALSE)
model_full_rs_ml <- lmer(wtp ~ condition + (condition | participant),
                         data = df_exp, REML = FALSE)

anova(model_ri_only_ml, model_zcp_ml, model_full_rs_ml)
```

The sequence tests: (1) does the random slope variance matter at all? (RI vs. ZCP) and (2) does the intercept-slope correlation additionally matter? (ZCP vs. full). This staged approach identifies the minimum sufficient random effects structure.

**When to prefer zero-correlation parameterization:**

| Situation | Recommendation |
|-----------|---------------|
| Small sample (< 20 units) | ZCP; estimating $\rho$ is unreliable |
| Model fails to converge | Try ZCP as a first simplification step |
| Correlation is theoretically uninteresting | ZCP avoids unnecessary parameters |
| Testing whether the correlation matters | Compare ZCP vs. full with LRT |

## Assumptions of Random-Slope Models

::: {.callout-warning icon=false}
## Assumption 1: Multivariate Normality of Random Effects

**What it means:** The random intercepts and slopes are jointly drawn from a bivariate normal distribution. Severe departures — banana-shaped clouds, truncation, discrete clusters — suggest the normality assumption is violated, and estimates may be biased.

**JDM examples:** If households fall into two discrete types (highly financially engaged vs. completely disengaged), the random slopes will form two clusters rather than an elliptical cloud. Similarly, if participants in an anchoring study come from two distinct pools (experienced negotiators vs. novices), the slope distribution may be bimodal.

**How to check:** Plot the BLUP estimates from `ranef()` as a scatterplot. A well-specified model produces an elliptical cloud centered at the origin.

```{r ranef-check}
re_df_exp <- as.data.frame(ranef(model_rs_exp)$participant)
names(re_df_exp) <- c("u0_intercept", "u1_slope")

ggplot(re_df_exp, aes(x = u0_intercept, y = u1_slope)) +
  geom_point(size = 2.5, color = "#457b9d", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title    = "BLUP random effects: experimental model",
    subtitle = "Elliptical cloud centered at (0,0) supports bivariate normality",
    x        = "Random intercept (u0j)",
    y        = "Random slope for condition (u1j)"
  ) +
  theme_minimal(base_size = 13)
```
:::

::: {.callout-warning icon=false}
## Assumption 2: Correct Random Effects Structure Is Specified

**What it means:** The model must include random slopes for every predictor whose effect varies meaningfully across units. Omitting a necessary random slope pushes that variability into the residual, deflating standard errors for the fixed effect and producing anti-conservatively small *p*-values.

**How to check:** Use likelihood ratio tests (LRT) comparing models with and without the random slope. Models must be fitted with `REML = FALSE` for valid LRT comparisons.

```{r lrt-slopes}
model_ri_diary_ml <- lmer(savings ~ disclosure_c + (1 | household),
                          data = df_diary, REML = FALSE)
model_rs_diary_ml <- lmer(savings ~ disclosure_c + (disclosure_c | household),
                          data = df_diary, REML = FALSE)

cat("=== Longitudinal: does disclosure slope vary across households? ===\n")
anova(model_ri_diary_ml, model_rs_diary_ml)

model_ri_exp_ml  <- lmer(wtp ~ condition + (1 | participant),
                          data = df_exp, REML = FALSE)
model_rs_exp_ml  <- lmer(wtp ~ condition + (condition | participant),
                          data = df_exp, REML = FALSE)

cat("\n=== Experimental: does anchoring effect vary across participants? ===\n")
anova(model_ri_exp_ml, model_rs_exp_ml)
```

A significant LRT ($p < .05$) indicates the random slope accounts for meaningful variance and should be retained. Use `REML = TRUE` for final parameter estimation once the model structure is settled.
:::

::: {.callout-warning icon=false}
## Assumption 3: Convergence

**What it means:** Complex random-slope models — especially with multiple predictors or crossed random effects — frequently fail to converge. A convergence warning from `lme4` signals the optimization algorithm did not find a stable solution. Estimates from non-converged models should not be trusted or reported.

**How to check:** Use `check_convergence()` from the `performance` package.

```{r convergence-check}
check_convergence(model_rs_diary)
check_convergence(model_rs_exp)
```

**What to do if convergence fails:**

- Center and scale continuous predictors
- Remove the intercept-slope correlation (see zero-correlation parameterization above)
- Try alternative optimizers: `control = lmerControl(optimizer = "bobyqa")`
- Simplify the random effects structure if the LRT supports it (non-significant random slope variance)
- Consider Bayesian estimation via `brms`, which handles complex random-effects structures more robustly
:::

## Model Comparison — Do You Need Random Slopes?

Add random slopes when:

1. **Theory predicts heterogeneous effects.** In financial coaching research, household-level differences in disclosure sensitivity are plausible and theoretically interesting — some households act on advice, others do not. In anchoring studies, individual differences in susceptibility to reference points are well-documented.

2. **The LRT supports them.** The model with random slopes fits significantly better than the model without (at `REML = FALSE`).

3. **The design mandates them.** In within-subjects experimental designs, random slopes for within-subjects conditions are structurally required to account for heterogeneous treatment effects. Omitting them is not conservative — it inflates Type I error (Barr et al., 2013).

4. **Sample size is adequate.** Estimating a random slope requires sufficient variation in the predictor *within each unit*. Longitudinally: each household must have multiple waves with varying disclosure engagement. Experimentally: each participant must complete trials in multiple conditions.

```{r model-comparison-table}
bind_rows(
  broom.mixed::glance(model_ri_diary_ml) |> mutate(model = "RI only",    setting = "Longitudinal"),
  broom.mixed::glance(model_rs_diary_ml) |> mutate(model = "RI + Slope", setting = "Longitudinal"),
  broom.mixed::glance(model_ri_exp_ml)   |> mutate(model = "RI only",    setting = "Experimental"),
  broom.mixed::glance(model_rs_exp_ml)   |> mutate(model = "RI + Slope", setting = "Experimental")
) |>
  select(setting, model, AIC, BIC, logLik) |>
  knitr::kable(digits = 1,
               caption = "Model fit: random intercept vs. random intercept + slope")
```

::: {.callout-tip icon=false}
## Reviewer Challenge: Three Questions About Your Random Slopes

**Challenge 1 — "Is the slope variance scientifically meaningful or just noise?"**
A significant LRT for the random slope variance tells you that slopes vary across units more than expected by chance. But variance could reflect true heterogeneity in disclosure sensitivity *or* measurement noise in within-household disclosure estimates. Reviewers will ask: is the $\hat\sigma_{u_1}$ estimate large enough to be practically meaningful? If the SD of individual slopes is only 0.1 when the average effect is 2.0, the heterogeneity is probably not theoretically important, even if statistically significant. Report $\hat\sigma_{u_1}$ alongside the fixed slope.

**Challenge 2 — "What does the intercept-slope correlation mean for your theory?"**
A reviewer will ask whether the negative correlation between household baseline savings and disclosure sensitivity is consistent with a theoretical mechanism (diminishing returns) or whether it is an artefact of regression to the mean. Present the correlation with a scatterplot of BLUPs and a substantive interpretation — do not just report the number from the variance-covariance matrix.

**Challenge 3 — "Did you test the maximal structure?"**
If you included a random slope for disclosure engagement but not for other within-household predictors (e.g., season, promotion activity), a reviewer may ask why. The Barr et al. (2013) recommendation is to include random slopes for all within-unit factors. Be ready to justify each slope inclusion or exclusion with a combination of theoretical reasoning and LRT results.
:::

## Recommended Reading

| Topic | Citation |
|-------|----------|
| Maximal random effects in experiments | Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language, 68*(3), 255–278. |
| Random slopes and cross-level interactions | Heisig, J. P., & Schaeffer, M. (2019). Why you should always include a random slope for the lower-level variable involved in a cross-level interaction. *European Sociological Review, 35*(2), 258–279. |
| Random slopes in experimental psychology | Judd, C. M., Westfall, J., & Kenny, D. A. (2012). Treating stimuli as a random factor in social psychology: A new and compelling case for mixed models. *Journal of Personality and Social Psychology, 103*(1), 54–69. |
| Practical guide to `lme4` | Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting linear mixed-effects models using lme4. *Journal of Statistical Software, 67*(1), 1–48. |
