---
title: "Three-Level Structures"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(performance)
library(ggplot2)
library(patchwork)
set.seed(42)
```

# Three-Level Structures

::: {.callout-note icon=false}
## What Does a Third Level Add?

- **Question:** When observations are nested in customers, and customers are nested in bank branches (or stores, regions, clinics), collapsing to two levels misattributes branch-level variance to customers — producing a biased ICC and wrong standard errors for any branch-level predictor.
- **What goes wrong:** A two-level model treats all customer-level residuals as independent within the sample, ignoring the fact that customers at the same branch share their local economic environment, advisor quality, and program availability. Any branch-level variation ends up in the wrong variance component.
- **The fix:** Add a third random intercept for branches — `(1 | branch/customer)` — so the model partitions variance correctly across all three levels. The branch ICC ($\text{ICC}_3$) becomes a scientific finding in its own right: it tells you how much branch-level factors shape customer outcomes.
:::

## When Two Levels Aren't Enough

Real data rarely comes in neat two-level packages. The nesting structures encountered in organizational research and experimental science are often deeper, and collapsing them to two levels misattributes variance in ways that affect both inference and interpretation.

Three-level structures arise in two distinct research contexts:

**Consumer panel data nested within service branches.** Monthly observations (L1) are nested within customers (L2), who are nested within bank branches (L3). The ICC at L3 tells you how much branches differ in customer financial outcomes — a direct measure of branch-level influence. Local economic conditions, advisor quality, branch program availability, and community effects all contribute to this branch-level variance. Ignoring the branch level means assuming that all customers within the same branch are no more similar to each other than customers at different branches — an implausible assumption when branches shape the advice received, the peers observed, and the local financial environment.

**Multi-site experimental research.** Trials (L1) are nested within participants (L2), who are nested within labs or research sites (L3). The ICC at L3 tells you how much results vary across labs — a direct measure of the replication generalizability problem. A multi-site replication study — increasingly common in psychological science as a response to the replication crisis — is structurally a three-level model. Ignoring the lab level treats site-to-site variation as random noise attributable to participants or trials, misattributing variance and producing overconfident estimates.

## The Model

**Level 1** (observations within level-2 units):
$$Y_{ijk} = \beta_{0jk} + \beta_1 X_{ijk} + e_{ijk}$$

**Level 2** (level-2 units within level-3 groups):
$$\beta_{0jk} = \gamma_{00k} + u_{0jk}$$

**Level 3** (groups):
$$\gamma_{00k} = \delta_{000} + v_{0k}$$

**Combined form:**
$$Y_{ijk} = \underbrace{\delta_{000}}_{\text{grand mean}} + \underbrace{\beta_1 X_{ijk}}_{\text{L1 predictor}} + \underbrace{v_{0k}}_{\text{org/lab RE}} + \underbrace{u_{0jk}}_{\text{employee/participant RE}} + \underbrace{e_{ijk}}_{\text{residual}}$$

**Three variance components:**

- $\sigma^2_v$: variance at level 3 (between organizations or labs)
- $\sigma^2_u$: variance at level 2 (between employees or participants, within groups)
- $\sigma^2_e$: variance at level 1 (within employees or participants, across occasions or trials)

The total variance in $Y$ is $\sigma^2_v + \sigma^2_u + \sigma^2_e$, and each component can be expressed as a proportion of this total to obtain the ICCs.

## Longitudinal Example — Consumer Savings Panel in Bank Branches

**Design:** 8 bank branches × 15 customers × 6 monthly waves = 720 observations. The three levels are:

- **Level 1:** monthly observation (savings rate, disclosure engagement)
- **Level 2:** customer (random intercept per customer, nested within branch)
- **Level 3:** branch (random intercept per branch)

**Research question:** How much of total savings variance is attributable to branch-level differences vs. stable customer-level differences vs. month-to-month fluctuation within customers? And after partitioning variance, is within-customer disclosure engagement associated with higher savings?

```{r longitudinal-3level}
n_branches  <- 8
n_customers <- 15   # customers per branch
n_waves     <- 6    # monthly observations per customer

# Level-3 random effects (branch)
branch_effects <- rnorm(n_branches, 0, 5)          # SD = 5 between branches
                                                    # (local economy, advisor quality)

# Level-2 random effects (customer within branch)
cust_effects <- rnorm(n_branches * n_customers, 0, 4)  # SD = 4 between customers within branch

df_3l <- expand_grid(
  branch   = 1:n_branches,
  customer = 1:n_customers,
  wave     = 1:n_waves
) |>
  mutate(
    cust_id    = (branch - 1) * n_customers + customer,
    disclosure = rnorm(n(), mean = 5, sd = 2),
    disc_c     = disclosure - 5,
    savings    = 25 +
                 branch_effects[branch] +
                 cust_effects[cust_id] +
                 0.8 * disc_c +
                 rnorm(n(), 0, 3),            # SD = 3 within-customer monthly noise
    branch  = factor(branch,  labels = paste0("Branch",  1:n_branches)),
    cust_id = factor(cust_id)
  )

# Three-level model: waves within customers within branches
model_3l <- lmer(savings ~ disc_c + (1 | branch/cust_id),
                 data = df_3l)

summary(model_3l)
```

The `(1 | branch/cust_id)` shorthand is equivalent to `(1 | branch) + (1 | cust_id:branch)`. Both tell `lme4` to estimate a random intercept for each branch *and* for each customer *within* their branch.

```{r longitudinal-variance-decomp}
vc_long <- as.data.frame(VarCorr(model_3l))

vc_long |>
  select(grp, vcov) |>
  mutate(
    pct_variance = round(100 * vcov / sum(vcov), 1)
  ) |>
  knitr::kable(
    digits  = 2,
    col.names = c("Level", "Variance", "% of Total"),
    caption = "Variance decomposition: customer savings panel within branches"
  )
```

**Reading the variance components:**

- `branch`: How much of total savings variance is explained by *which branch* a customer belongs to? This is the branch-level ICC (ICC$_3$). It captures local economic conditions, advisor quality, branch program availability, and community savings norms.
- `cust_id:branch`: How much is explained by stable differences *between customers* within the same branch — persistent individual differences in income, financial habits, risk tolerance.
- `Residual`: How much is month-to-month fluctuation *within the same customer* — the portion that varies over time and could be affected by disclosure engagement.

A large branch-level ICC (ICC$_3 > 0.10$) means branches matter substantially for customer savings — suggesting branch-level interventions (coaching quality, program design, community outreach) have real impact beyond individual-level characteristics.

## Experimental Example — Multi-Lab Replication Study

**Design:** 5 labs × 20 participants × 8 trials = 800 observations. The three levels are:

- **Level 1:** trial (decision quality score, condition: scarcity vs. control)
- **Level 2:** participant (random intercept per participant, nested within lab)
- **Level 3:** lab (random intercept per lab)

**Research question:** How much variance in decision quality is attributable to lab differences, participant differences within labs, and trial-to-trial fluctuation? The lab-level ICC quantifies how much the results depend on *where* the study was run.

```{r experimental-3level}
n_labs    <- 5
n_par     <- 20   # participants per lab
n_trials  <- 8    # trials per participant

# Level-3 random effects (lab)
lab_effects <- rnorm(n_labs, 0, 3)             # SD = 3 between labs

# Level-2 random effects (participant within lab)
par_effects <- rnorm(n_labs * n_par, 0, 5)    # SD = 5 between participants within lab

df_multilab <- expand_grid(
  lab         = 1:n_labs,
  participant = 1:n_par,
  trial       = 1:n_trials
) |>
  mutate(
    par_id    = (lab - 1) * n_par + participant,
    condition = rep(c(0, 1), times = n_trials / 2)[trial],
    dq        = 65 +
                lab_effects[lab] +
                par_effects[par_id] +
                -4 * condition +
                rnorm(n(), 0, 4),            # SD = 4 within-participant trial noise
    lab        = factor(lab,  labels = paste0("Lab", 1:n_labs)),
    par_id     = factor(par_id),
    condition  = factor(condition, labels = c("control", "scarcity"))
  )

# Three-level model: trials within participants within labs
model_multilab <- lmer(dq ~ condition + (1 | lab/par_id),
                       data = df_multilab)

summary(model_multilab)
```

```{r experimental-variance-decomp}
vc_exp <- as.data.frame(VarCorr(model_multilab))

vc_exp |>
  select(grp, vcov) |>
  mutate(
    pct_variance = round(100 * vcov / sum(vcov), 1)
  ) |>
  knitr::kable(
    digits  = 2,
    col.names = c("Level", "Variance", "% of Total"),
    caption = "Variance decomposition: multi-lab scarcity replication"
  )
```

**Reading the variance components in the experimental case:**

- `lab`: Lab-to-lab variability. How much does it matter *which lab* ran the study? If this is high, the same nominal procedure produces meaningfully different baseline decision quality across sites — suggesting lab culture, participant pools, or local implementation differences are consequential.
- `par_id:lab`: Individual differences between participants within the same lab. In a within-subjects design this is typically the largest component.
- `Residual`: Trial-to-trial noise within a participant — random fluctuation from one decision to the next.

**The connection to the replication crisis.** If ICC$_3 = 0.15$, then 15% of variance in decision quality is due to which lab a participant was tested in — independent of the scarcity manipulation. A single-lab study that happens to be run in a high-performing lab will systematically overestimate baseline scores, and if the manipulation interacts with lab characteristics, will also misestimate the effect size. Many replication failures occur because the original study had high lab-level variance: the effect was real in lab A but not lab B. Three-level models quantify this risk directly.

## ICC Decomposition for Three-Level Models

In a three-level model, there are two ICCs to report and interpret:

**Level-3 ICC** (variance attributable to the highest-level unit alone):
$$\text{ICC}_3 = \frac{\sigma^2_v}{\sigma^2_v + \sigma^2_u + \sigma^2_e}$$

**Level-2 ICC** (variance attributable to level-2 *and above*):
$$\text{ICC}_2 = \frac{\sigma^2_v + \sigma^2_u}{\sigma^2_v + \sigma^2_u + \sigma^2_e}$$

The difference $\text{ICC}_2 - \text{ICC}_3$ is the variance *uniquely* attributable to level-2 units after removing level-3 variance.

```{r icc-both}
icc_long <- icc(model_3l)
icc_exp  <- icc(model_multilab)

cat("=== Longitudinal consumer panel (savings) ===\n")
print(icc_long)

cat("\n=== Experimental multi-lab replication ===\n")
print(icc_exp)
```

```{r icc-viz}
# Build a side-by-side variance partition bar chart for both models
make_varpart_df <- function(model, labels, design) {
  vc  <- as.data.frame(VarCorr(model))
  tot <- sum(vc$vcov)
  tibble(
    level  = labels,
    pct    = vc$vcov / tot,
    design = design
  )
}

vp_long <- make_varpart_df(
  model_3l,
  labels = c("Branch (L3)", "Customer (L2)", "Month (L1)"),
  design = "Longitudinal (Consumer Panel)"
)

vp_exp <- make_varpart_df(
  model_multilab,
  labels = c("Lab (L3)", "Participant (L2)", "Trial (L1)"),
  design = "Experimental"
)

vp_all <- bind_rows(vp_long, vp_exp)

p_vp_long <- ggplot(vp_long, aes(x = "", y = pct, fill = factor(level, levels = rev(level)))) +
  geom_col(width = 0.5, color = "white") +
  scale_fill_manual(values = c("#d4e6f1", "#457b9d", "#1e3a5f")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Consumer Panel (branches)", y = "% of variance", x = NULL, fill = "Level") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p_vp_exp <- ggplot(vp_exp, aes(x = "", y = pct, fill = factor(level, levels = rev(level)))) +
  geom_col(width = 0.5, color = "white") +
  scale_fill_manual(values = c("#fde8e8", "#e63946", "#9b1d20")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Experimental (Multi-Lab)", y = "% of variance", x = NULL, fill = "Level") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p_vp_long + p_vp_exp +
  plot_annotation(title = "Variance decomposition: where does the outcome variance live?")
```

## Cross-Level Interactions

Three-level models allow testing whether higher-level variables moderate lower-level effects — some of the most theoretically interesting questions in multilevel research.

### Longitudinal: Does branch coaching intensity moderate disclosure → savings?

**Hypothesis:** In branches with intensive financial coaching programs (high-touch advisors, regular follow-ups), the positive effect of disclosure engagement on savings may be stronger — because the advice is reinforced by personal contact. In low-touch branches, disclosure alone may not be enough to change behaviour.

```{r cross-level-long}
# Add L3 predictor: branch coaching intensity (1 = high-touch, 0 = standard)
df_3l <- df_3l |>
  group_by(branch) |>
  mutate(high_touch = if_else(branch %in% c("Branch1", "Branch2", "Branch3", "Branch4"),
                              1L, 0L)) |>
  ungroup()

model_cli <- lmer(savings ~ disc_c * high_touch + (1 | branch/cust_id),
                  data = df_3l)

broom.mixed::tidy(model_cli, effects = "fixed") |>
  knitr::kable(
    digits  = 3,
    caption = "Cross-level interaction: branch coaching intensity × disclosure → savings"
  )
```

The interaction term `disc_c:high_touch` tests whether the disclosure slope differs between high-touch and standard branches. A significant positive coefficient would indicate that disclosure is more effective in high-touch branches — the behavioural change from reading the document is amplified by concurrent personal coaching.

### Experimental: Does lab context moderate the scarcity effect?

**Hypothesis:** Labs that routinely use deception may have participant pools that are more suspicious or alert, potentially dampening or amplifying responses to the scarcity manipulation.

```{r cross-level-exp}
# Add L3 predictor: lab type (1 = uses deception, 0 = does not)
df_multilab <- df_multilab |>
  group_by(lab) |>
  mutate(lab_type = if_else(lab %in% c("Lab1", "Lab2"), 1L, 0L)) |>
  ungroup()

model_cli_exp <- lmer(dq ~ condition * lab_type + (1 | lab/par_id),
                      data = df_multilab)

broom.mixed::tidy(model_cli_exp, effects = "fixed") |>
  knitr::kable(
    digits  = 3,
    caption = "Cross-level interaction: lab type × scarcity condition"
  )
```

```{r cli-viz}
# Visualize the experimental cross-level interaction
pred_df <- expand_grid(
  condition = c("control", "scarcity"),
  lab_type  = c(0L, 1L)
) |>
  mutate(
    predicted = predict(model_cli_exp,
                        newdata = tibble(condition, lab_type),
                        re.form = NA),
    lab_label = factor(lab_type, labels = c("No deception", "Uses deception"))
  )

ggplot(pred_df, aes(x = condition, y = predicted,
                    group = lab_label, color = lab_label)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 3.5) +
  scale_color_manual(values = c("#1e3a5f", "#e63946")) +
  labs(
    title    = "Cross-level interaction: lab type moderates scarcity effect",
    subtitle = "Predicted decision quality by condition and lab context",
    x        = "Condition",
    y        = "Predicted decision quality",
    color    = "Lab type"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

::: {.callout-warning icon=false}
## Including Random Slopes in Cross-Level Interactions

**Heisig & Schaeffer (2019)** showed that when a cross-level interaction involves a level-1 predictor, you must include a random slope for that predictor to obtain unbiased estimates of the interaction. Without the random slope, the cross-level interaction estimate absorbs variance that should be attributed to individual-level heterogeneity.

In practice: if you include `disc_c * high_touch`, you should also include `(disc_c | branch/cust_id)` or at minimum allow the disclosure slope to vary across branches. This model requires more data to estimate reliably, but it is the correctly specified model.
:::

## Assumptions of Three-Level Models

All two-level model assumptions carry over — normality of random effects at each level, correct specification of fixed effects, independence of residuals conditional on random effects. Three additional concerns become prominent with three levels.

::: {.callout-warning icon=false}
## Assumption 1: Correct Level Specification

**What it means:** The model must correctly identify which units belong to which level. If customers are incorrectly assigned to branches (data linkage errors, customers who switch branches mid-study), variance components will be misattributed between levels.

**In experimental designs:** Participants must be uniquely and correctly assigned to labs. Cross-site contamination — participants completing a study at one lab but attributed to another — corrupts the level-3 variance estimate.

**How to check:** Cross-tabulate your level-2 and level-3 identifiers. In a pure nested hierarchy, each level-2 unit belongs to exactly one level-3 unit. Any rows returned below signal a cross-classified or misassigned structure.

```{r level-check}
# Longitudinal: each customer should belong to exactly one branch
df_3l |>
  distinct(cust_id, branch) |>
  group_by(cust_id) |>
  filter(n() > 1)   # should return 0 rows

# Experimental: each participant should belong to exactly one lab
df_multilab |>
  distinct(par_id, lab) |>
  group_by(par_id) |>
  filter(n() > 1)   # should return 0 rows
```

Zero rows confirms a clean nested hierarchy. If rows appear, you have a cross-classified structure requiring a different model specification (e.g., `(1 | branch) + (1 | cust_id)` instead of `(1 | branch/cust_id)`).
:::

::: {.callout-warning icon=false}
## Assumption 2: Sufficient Sample Size at Each Level

**What it means:** Variance components at higher levels require sufficient numbers of higher-level units to be estimated reliably. With only 5 labs or 6 organizations, the level-3 variance estimate is extremely noisy — confidence intervals for $\sigma^2_v$ will be very wide.

**Why it matters:** Type I error rates for fixed effects are also affected. With very few level-3 units, the degrees of freedom for branch- or lab-level effects are essentially the number of level-3 units minus the number of predictors at that level.

**How to check:** Count units at each level explicitly.

```{r sample-size-check}
cat("=== Longitudinal consumer panel design ===\n")
cat("Branches (L3):", n_distinct(df_3l$branch), "\n")
cat("Customers per branch (avg):",
    round(nrow(distinct(df_3l, branch, cust_id)) / n_distinct(df_3l$branch), 1), "\n")
cat("Waves per customer (avg):",
    round(nrow(df_3l) / nrow(distinct(df_3l, cust_id)), 1), "\n")

cat("\n=== Experimental multi-lab design ===\n")
cat("Labs (L3):", n_distinct(df_multilab$lab), "\n")
cat("Participants per lab (avg):",
    round(nrow(distinct(df_multilab, lab, par_id)) / n_distinct(df_multilab$lab), 1), "\n")
cat("Trials per participant (avg):",
    round(nrow(df_multilab) / nrow(distinct(df_multilab, par_id)), 1), "\n")
```
:::

::: {.callout-warning icon=false}
## Assumption 3: Minimum Clusters at L3

**Rule of thumb:** ≥ 20 units at L3 for reliable variance estimation at the top level. Below this threshold, the level-3 ICC estimate has very wide confidence intervals, and Wald-type standard errors for level-3 predictors can be severely anti-conservative.

| Level | Minimum recommended | Note |
|-------|---------------------|------|
| Level 3 (top-level groups) | ≥ 20 units | Below this, L3 variance is highly unstable |
| Level 2 per L3 unit | ≥ 10 units | Needed for stable L2 variance within groups |
| Level 1 per L2 unit | ≥ 5 observations | Needed for within-unit estimates |

**Practical implication for multi-lab research:** With only 5 labs (as in our example), the lab-level ICC is estimated with very low precision. This is not a reason to avoid three-level models — a biased two-level model that ignores labs is worse — but it is a reason to:

- Report wide confidence intervals for the lab ICC
- Avoid strong claims about the magnitude of lab-level effects
- Plan future multi-site studies with more labs rather than more participants per lab, if cross-site generalizability is the scientific goal

For Bayesian approaches with informative priors on variance components, smaller level-3 samples can be handled more gracefully. The `brms` package provides a natural extension.
:::

## Adding Level-2 and Level-3 Predictors

One major advantage of three-level models is the ability to include predictors at each level simultaneously, each partialling out the others.

### Longitudinal Setting

```{r longitudinal-predictors}
# Add customer-level predictor (L2): account tenure (years with bank)
# Add branch-level predictor (L3): branch urban vs. rural location
set.seed(42)

df_3l <- df_3l |>
  group_by(branch) |>
  mutate(urban = if_else(branch %in% c("Branch1", "Branch2", "Branch3", "Branch4"),
                         1L, 0L)) |>
  ungroup()

df_3l <- df_3l |>
  group_by(branch, cust_id) |>
  mutate(acct_tenure = round(runif(1, 1, 15), 0)) |>   # years with bank
  ungroup()

model_3l_pred <- lmer(
  savings ~ disc_c + acct_tenure + urban + (1 | branch/cust_id),
  data = df_3l
)

broom.mixed::tidy(model_3l_pred, effects = "fixed") |>
  knitr::kable(
    digits  = 3,
    caption = "Three-level panel model with L1 (disclosure), L2 (account tenure), L3 (urban/rural) predictors"
  )
```

Each fixed effect is estimated controlling for the others: the disclosure effect is within-customer, the account-tenure effect is between-customers within the same branch, and the urban/rural effect is between branches.

### Experimental Setting

```{r experimental-predictors}
# Add participant-level predictor (L2): cognitive ability score
# Add lab-level predictor (L3): lab prestige (high vs. standard)
df_multilab <- df_multilab |>
  group_by(lab) |>
  mutate(lab_prestige = if_else(lab %in% c("Lab1", "Lab2"), 1L, 0L)) |>
  ungroup()

df_multilab <- df_multilab |>
  group_by(lab, par_id) |>
  mutate(cog_ability = round(rnorm(1, 100, 15), 0)) |>
  ungroup()

model_multilab_pred <- lmer(
  dq ~ condition + cog_ability + lab_prestige + (1 | lab/par_id),
  data = df_multilab
)

broom.mixed::tidy(model_multilab_pred, effects = "fixed") |>
  knitr::kable(
    digits  = 3,
    caption = "Three-level experimental model with condition (L1), cognitive ability (L2), and lab prestige (L3)"
  )
```

The model simultaneously estimates the scarcity effect (within participants across trials), the cognitive ability effect (between participants within labs), and the lab prestige effect (between labs) — each adjusted for the other two levels.

::: {.callout-tip icon=false}
## Reviewer Challenge: Three Questions About Your Three-Level Model

**Challenge 1 — "How precise is your branch-level ICC?"**
With only 8 branches, the branch-level variance ($\hat\sigma_v^2$) is estimated from 8 data points. Confidence intervals for the ICC at level 3 will be very wide. A reviewer will note this and may ask for a sensitivity analysis — reporting how your fixed-effect estimates change if you instead treat branches as fixed effects (branch dummies) rather than a random effect. If the coefficient on your focal predictor (disclosure engagement) shifts substantially, the branch-level random effect is doing a lot of heavy lifting and deserves scrutiny.

**Challenge 2 — "Is the cross-level interaction identified?"**
The interaction between disclosure engagement (L1) and branch coaching intensity (L3) claims that the disclosure slope is moderated by a branch-level characteristic. But there are only 8 branches — 4 high-touch and 4 standard. With 4 data points per group at L3, detecting a genuine interaction requires strong effect sizes. Reviewers will ask whether the interaction is robust to removing any single branch from the analysis (leave-one-branch-out sensitivity), and whether the pattern is consistent with the individual branch means rather than driven by one outlier branch.

**Challenge 3 — "Did you include the L2 random slope for the cross-level interaction?"**
Heisig & Schaeffer (2019) showed that cross-level interactions involving a L1 predictor must include a random slope for that predictor at L2, otherwise the interaction estimate is biased. If you include `disc_c * high_touch` but not `(disc_c | branch/cust_id)`, you are assuming that every customer's disclosure sensitivity is identical within branches — which contradicts the premise of the interaction. A reviewer who knows this literature will ask why the full random slope was not included.
:::

## Recommended Reading

| Topic | Citation |
|-------|----------|
| Three-level models: foundations | Raudenbush, S. W., & Bryk, A. S. (2002). *Hierarchical linear models: Applications and data analysis methods* (2nd ed.). Sage. |
| Multi-site replication and lab effects | Westfall, J., Kenny, D. A., & Judd, C. M. (2014). Statistical power and optimal design in experiments in which samples of participants respond to samples of stimuli. *Journal of Experimental Psychology: General, 143*(5), 2020–2045. |
| Cross-level interactions and random slopes | Heisig, J. P., & Schaeffer, M. (2019). Why you should always include a random slope for the lower-level variable involved in a cross-level interaction. *European Sociological Review, 35*(2), 258–279. |
| Sample size at higher levels | Maas, C. J. M., & Hox, J. J. (2005). Sufficient sample sizes for multilevel modeling. *Methodology, 1*(3), 86–92. |
| Replication crisis and multilevel structure | Nosek, B. A., et al. (2015). Estimating the reproducibility of psychological science. *Science, 349*(6251), aac4716. |
