---
title: "Random Slopes"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(performance)
library(ggplot2)
library(patchwork)
set.seed(42)
```

# Random Slopes

## When Effects Vary Across Units

The random-intercept model assumes that the *slope* — the relationship between a predictor and the outcome — is identical across all units. Only the baseline level differs from unit to unit. This is often an unrealistic assumption.

Consider two research settings:

**Longitudinal panel data.** An employee well-being study tracks workers across weekly waves. The average effect of workload on well-being may be negative — heavier workloads are associated with lower well-being. But some employees are highly sensitive: a small increase in workload produces a steep drop in their scores. Others are remarkably resilient: their well-being stays high regardless of how demanding their job gets. The slope of workload → well-being *varies across employees*.

**Within-subjects experiments.** A scarcity priming study asks participants to complete decision tasks under either resource-scarcity or control conditions. On average, the scarcity condition reduces decision quality. But individual responses to the manipulation vary considerably — some participants are strongly affected, others barely register a difference. The *treatment effect itself* is heterogeneous across participants.

The **random-slope model** addresses this directly. Rather than fixing one slope for everyone, we treat each unit's slope as a draw from a distribution, allowing us to estimate:

- The **average** effect across units (the fixed slope, $\gamma_{10}$)
- The **variability** of that effect across units (the random slope variance, $\sigma^2_{u_1}$)
- The **correlation** between where units start and how strongly they respond ($\rho_{u_0, u_1}$)

## The Model

**Level 1** (observations within units):
$$Y_{ij} = \beta_{0j} + \beta_{1j} X_{ij} + \varepsilon_{ij}$$

**Level 2** (units):
$$\beta_{0j} = \gamma_{00} + u_{0j}$$
$$\beta_{1j} = \gamma_{10} + u_{1j}$$

**Combined form:**
$$Y_{ij} = \underbrace{\gamma_{00}}_{\text{grand mean}} + \underbrace{\gamma_{10} X_{ij}}_{\text{avg slope}} + \underbrace{u_{0j}}_{\text{person intercept}} + \underbrace{u_{1j} X_{ij}}_{\text{person slope deviation}} + \varepsilon_{ij}$$

The random effects are assumed to follow a bivariate normal distribution:
$$\begin{pmatrix} u_{0j} \\ u_{1j} \end{pmatrix} \sim \mathcal{N}\left(\begin{pmatrix} 0 \\ 0 \end{pmatrix}, \begin{pmatrix} \sigma^2_{u_0} & \sigma_{u_0 u_1} \\ \sigma_{u_0 u_1} & \sigma^2_{u_1} \end{pmatrix}\right)$$

**What does $u_{1j}$ mean in each context?**

- **Longitudinal:** $u_{1j}$ is employee $j$'s deviation from the average workload sensitivity. A large negative $u_{1j}$ means this employee is especially sensitive to workload demands — their well-being drops steeply when workload increases. A near-zero $u_{1j}$ means workload has its typical effect on this person.

- **Experimental:** $u_{1j}$ is participant $j$'s deviation from the average scarcity effect on decision quality. A large positive or negative $u_{1j}$ signals heterogeneous treatment response — the manipulation was much stronger or weaker for this individual than average.

## Longitudinal Example — Employee Wellbeing and Workload

**Design:** 50 employees measured at 8 weekly time points. The predictor is weekly workload (centered at its grand mean of 50). The outcome is well-being (0–100 scale). Each employee has their own workload sensitivity — modeled as a random slope.

The simulation imposes a correlation of −0.3 between intercepts and slopes, so employees who start with high baseline well-being are somewhat less sensitive to workload fluctuations.

```{r longitudinal-data}
n_employees <- 50
n_waves     <- 8

# Correlated random intercepts and slopes: SD_int = 8, SD_slope = 1.5, rho = -0.3
Sigma_long <- matrix(c(64, -3.6, -3.6, 2.25), nrow = 2)
re_mat     <- MASS::mvrnorm(n_employees, mu = c(0, 0), Sigma = Sigma_long)

df_long <- expand_grid(
  employee = 1:n_employees,
  wave     = 1:n_waves
) |>
  mutate(
    workload   = rnorm(n(), mean = 50, sd = 10),
    workload_c = workload - 50,
    well_being = (55 + re_mat[employee, 1]) +
                 (-2 + re_mat[employee, 2]) * workload_c +
                 rnorm(n(), 0, 5),
    employee   = factor(employee)
  )

# Random intercept only
model_ri_long <- lmer(well_being ~ workload_c + (1 | employee),
                      data = df_long, REML = TRUE)

# Random intercept + slope
model_rs_long <- lmer(well_being ~ workload_c + (workload_c | employee),
                      data = df_long, REML = TRUE)

broom.mixed::tidy(model_rs_long, effects = "ran_pars") |>
  knitr::kable(digits = 3,
               caption = "Variance components: workload → well-being random slope model")
```

The variance component for `workload_c` at the employee level ($\hat\sigma^2_{u_1}$) captures how much individual sensitivity varies. The correlation term estimates the intercept-slope association.

```{r longitudinal-slopes-plot}
# Extract OLS estimates per employee for visualization
employee_lines <- df_long |>
  group_by(employee) |>
  summarise(
    intercept = coef(lm(well_being ~ workload_c))[[1]],
    slope     = coef(lm(well_being ~ workload_c))[[2]],
    .groups   = "drop"
  )

workload_range <- seq(-20, 20, by = 0.5)

lines_df_long <- employee_lines |>
  mutate(
    x = list(workload_range),
    y = map2(intercept, slope, ~ .x + .y * workload_range)
  ) |>
  unnest(c(x, y)) |>
  rename(workload_c = x, well_being = y)

p_long_slopes <- ggplot() +
  geom_line(data = lines_df_long,
            aes(x = workload_c, y = well_being, group = employee),
            color = "#1e3a5f", alpha = 0.3, linewidth = 0.4) +
  geom_abline(
    slope     = fixef(model_rs_long)["workload_c"],
    intercept = fixef(model_rs_long)["(Intercept)"],
    color     = "#e63946", linewidth = 1.8
  ) +
  labs(
    title    = "Longitudinal: workload → well-being",
    subtitle = "Each line = one employee; red = average fixed slope",
    x        = "Workload (centered)",
    y        = "Well-being"
  ) +
  theme_minimal(base_size = 12)

p_long_slopes
```

The grey lines reveal substantial variation in workload sensitivity. Most employees show the expected negative slope, but the *steepness* varies considerably. A few employees show near-flat or even slightly positive slopes — for them, workload may not be experienced as burdensome during this observation period.

**Intercept-slope correlation:** A negative correlation (as simulated here) means employees who start with high baseline well-being tend to show a *weaker* negative workload effect. High-well-being employees may have resources — social support, autonomy, resilience — that buffer against workload demands. Conversely, employees already struggling (low baseline) may be most vulnerable to workload spikes.

## Experimental Example — Scarcity and Decision Quality

**Design:** 60 participants complete 12 decision trials each (6 in the scarcity condition, 6 in the control condition, counterbalanced). The predictor is condition (scarcity = 1, control = 0). The outcome is decision quality (0–100 scale). Each participant is allowed their own scarcity sensitivity.

The simulation imposes a positive correlation of 0.2 between intercepts and slopes: participants with higher baseline decision quality show a slightly stronger scarcity effect.

```{r experimental-data}
n_participants <- 60
n_trials_each  <- 6   # per condition

# Correlated intercepts and slopes: SD_int = 10, SD_slope = 3, rho = 0.2
Sigma_exp  <- matrix(c(100, 6, 6, 9), nrow = 2)
re_mat_exp <- MASS::mvrnorm(n_participants, mu = c(0, 0), Sigma = Sigma_exp)

df_exp <- expand_grid(
  participant = 1:n_participants,
  condition   = c(0, 1),
  trial       = 1:n_trials_each
) |>
  mutate(
    decision_quality = (65 + re_mat_exp[participant, 1]) +
                       (-5 + re_mat_exp[participant, 2]) * condition +
                       rnorm(n(), 0, 6),
    participant     = factor(participant),
    condition_label = factor(condition, labels = c("Control", "Scarcity"))
  )

# Random intercept only
model_ri_exp <- lmer(decision_quality ~ condition + (1 | participant),
                     data = df_exp, REML = TRUE)

# Random intercept + slope for condition
model_rs_exp <- lmer(decision_quality ~ condition + (condition | participant),
                     data = df_exp, REML = TRUE)

broom.mixed::tidy(model_rs_exp, effects = "ran_pars") |>
  knitr::kable(digits = 3,
               caption = "Variance components: scarcity condition random slope model")
```

```{r experimental-ranef-histogram}
# Extract per-participant scarcity effects (random slope BLUPs + fixed effect)
re_exp     <- ranef(model_rs_exp)$participant
names(re_exp) <- c("u0", "u1")
avg_effect <- fixef(model_rs_exp)["condition"]

re_exp <- re_exp |>
  mutate(person_scarcity_effect = avg_effect + u1)

p_hist <- ggplot(re_exp, aes(x = person_scarcity_effect)) +
  geom_histogram(bins = 20, fill = "#457b9d", color = "white", alpha = 0.85) +
  geom_vline(xintercept = avg_effect, color = "#e63946",
             linewidth = 1.5, linetype = "dashed") +
  annotate("text", x = avg_effect + 0.4, y = Inf,
           label = "avg effect", color = "#e63946",
           vjust = 2, hjust = 0, size = 3.5) +
  labs(
    title    = "Distribution of individual scarcity effects",
    subtitle = "Each bar = how many participants experienced that effect size",
    x        = "Person-level scarcity effect on decision quality",
    y        = "Count"
  ) +
  theme_minimal(base_size = 12)

p_hist
```

Some participants show almost no scarcity effect; others show a drop exceeding −10 points. The random slope captures this theoretically important heterogeneity — heterogeneity that is invisible in a model that forces the same effect on everyone.

```{r experimental-slopes-plot}
participant_lines <- df_exp |>
  group_by(participant) |>
  summarise(
    intercept = coef(lm(decision_quality ~ condition))[[1]],
    slope     = coef(lm(decision_quality ~ condition))[[2]],
    .groups   = "drop"
  )

p_exp_slopes <- ggplot() +
  geom_segment(
    data = participant_lines,
    aes(x = 0, xend = 1,
        y = intercept,
        yend = intercept + slope,
        group = participant),
    color = "#457b9d", alpha = 0.4, linewidth = 0.5
  ) +
  geom_segment(
    aes(x = 0, xend = 1,
        y = fixef(model_rs_exp)["(Intercept)"],
        yend = fixef(model_rs_exp)["(Intercept)"] + fixef(model_rs_exp)["condition"]),
    color = "#e63946", linewidth = 2
  ) +
  scale_x_continuous(breaks = c(0, 1), labels = c("Control", "Scarcity")) +
  labs(
    title    = "Experimental: scarcity → decision quality",
    subtitle = "Each line = one participant's treatment effect; red = average",
    x        = "Condition",
    y        = "Decision quality"
  ) +
  theme_minimal(base_size = 12)

p_exp_slopes
```

::: {.callout-warning icon=false}
## Key Point: Random Slopes and Type I Error in Experiments

**Barr, Levy, Scheepers, & Tily (2013)** demonstrated that omitting random slopes for within-subjects conditions inflates Type I error rates substantially. The mechanism: if you do not model variation in the treatment effect across participants, that variation is absorbed into the residual — deflating standard errors and making the average effect appear more precise than it is.

**The recommendation:** In within-subjects designs, include random slopes for all within-subjects factors by default (the "maximal" random effects structure). The random slope is not merely a nuisance parameter — it *is* the treatment effect heterogeneity, which is often scientifically interesting in its own right.
:::

## Comparing the Two Examples Side by Side

```{r side-by-side}
p_long_slopes + p_exp_slopes +
  plot_annotation(
    title   = "Random slopes: heterogeneous effects in two research designs",
    caption = "Red lines/segments = fixed (average) effects; colored lines = individual unit slopes"
  )
```

## The Intercept-Slope Correlation

The random-slope model estimates not just variance in intercepts and variance in slopes, but also their **covariance** — equivalently, the correlation $\rho_{u_0, u_1}$.

This correlation is substantively meaningful in both contexts.

**Longitudinal interpretation.** A negative intercept-slope correlation means: employees with *higher* baseline well-being tend to show a *weaker* negative workload effect. This is a buffering pattern — high-well-being employees are more resilient to workload fluctuations than their struggling colleagues.

**Experimental interpretation.** A positive intercept-slope correlation means: participants with *higher* baseline decision quality (in the control condition) show a *stronger* drop under scarcity. One interpretation: the scarcity manipulation particularly disrupts the elaborate, careful strategies that high-ability participants rely on — a "more to lose" pattern.

```{r intercept-slope-corr}
p_corr_long <- ggplot(employee_lines, aes(x = intercept, y = slope)) +
  geom_point(size = 2.5, color = "#1e3a5f", alpha = 0.7) +
  geom_smooth(method = "lm", color = "#e63946", se = TRUE, linewidth = 1) +
  labs(
    title    = "Longitudinal",
    subtitle = "Negative: high-baseline employees less workload-sensitive",
    x        = "Intercept (baseline well-being)",
    y        = "Slope (workload effect)"
  ) +
  theme_minimal(base_size = 12)

p_corr_exp <- ggplot(participant_lines, aes(x = intercept, y = slope)) +
  geom_point(size = 2.5, color = "#457b9d", alpha = 0.7) +
  geom_smooth(method = "lm", color = "#e63946", se = TRUE, linewidth = 1) +
  labs(
    title    = "Experimental",
    subtitle = "Positive: high-baseline participants more scarcity-sensitive",
    x        = "Intercept (control condition quality)",
    y        = "Slope (scarcity effect)"
  ) +
  theme_minimal(base_size = 12)

p_corr_long + p_corr_exp +
  plot_annotation(title = "Intercept-slope correlations in each design")
```

## The Zero-Correlation Parameterization

By default, `(predictor | group)` estimates random slope variance *and* the intercept-slope correlation. When the correlation is near zero, theoretically uninteresting, or causing convergence problems, the **zero-correlation parameterization** separates the two:

```r
# Full: estimates intercept variance, slope variance, AND their correlation
lmer(well_being ~ workload_c + (workload_c | employee), data = df_long)

# Zero-correlation: estimates both variances, forces correlation to zero
lmer(well_being ~ workload_c + (1 | employee) + (0 + workload_c | employee), data = df_long)
```

Applied to the experimental data:

```{r zero-corr-example}
model_zcp_exp <- lmer(decision_quality ~ condition +
                        (1 | participant) + (0 + condition | participant),
                      data = df_exp, REML = FALSE)

# Staged LRT: (1) does the random slope variance matter?
#             (2) does the correlation additionally matter?
model_ri_only_ml <- lmer(decision_quality ~ condition + (1 | participant),
                         data = df_exp, REML = FALSE)
model_zcp_ml     <- lmer(decision_quality ~ condition +
                           (1 | participant) + (0 + condition | participant),
                         data = df_exp, REML = FALSE)
model_full_rs_ml <- lmer(decision_quality ~ condition + (condition | participant),
                         data = df_exp, REML = FALSE)

anova(model_ri_only_ml, model_zcp_ml, model_full_rs_ml)
```

The sequence tests: (1) does the random slope variance matter at all? (RI vs. ZCP) and (2) does the intercept-slope correlation additionally matter? (ZCP vs. full). This staged approach identifies the minimum sufficient random effects structure.

**When to prefer zero-correlation parameterization:**

| Situation | Recommendation |
|-----------|---------------|
| Small sample (< 20 units) | ZCP; estimating $\rho$ is unreliable |
| Model fails to converge | Try ZCP as a first simplification step |
| Correlation is theoretically uninteresting | ZCP avoids unnecessary parameters |
| Testing whether the correlation matters | Compare ZCP vs. full with LRT |

## Assumptions of Random-Slope Models

::: {.callout-warning icon=false}
## Assumption 1: Multivariate Normality of Random Effects

**What it means:** The random intercepts and slopes are jointly drawn from a bivariate normal distribution. Severe departures — banana-shaped clouds, truncation, discrete clusters — suggest the normality assumption is violated, and estimates may be biased.

**How to check:** Plot the BLUP estimates from `ranef()` as a scatterplot. A well-specified model produces an elliptical cloud centered at the origin.

```{r ranef-check}
re_df_exp <- as.data.frame(ranef(model_rs_exp)$participant)
names(re_df_exp) <- c("u0_intercept", "u1_slope")

ggplot(re_df_exp, aes(x = u0_intercept, y = u1_slope)) +
  geom_point(size = 2.5, color = "#457b9d", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title    = "BLUP random effects: experimental model",
    subtitle = "Elliptical cloud centered at (0,0) supports bivariate normality",
    x        = "Random intercept (u0j)",
    y        = "Random slope for condition (u1j)"
  ) +
  theme_minimal(base_size = 13)
```
:::

::: {.callout-warning icon=false}
## Assumption 2: Correct Random Effects Structure Is Specified

**What it means:** The model must include random slopes for every predictor whose effect varies meaningfully across units. Omitting a necessary random slope pushes that variability into the residual, deflating standard errors for the fixed effect and producing anti-conservatively small *p*-values.

**How to check:** Use likelihood ratio tests (LRT) comparing models with and without the random slope. Models must be fitted with `REML = FALSE` for valid LRT comparisons.

```{r lrt-slopes}
model_ri_long_ml <- lmer(well_being ~ workload_c + (1 | employee),
                         data = df_long, REML = FALSE)
model_rs_long_ml <- lmer(well_being ~ workload_c + (workload_c | employee),
                         data = df_long, REML = FALSE)

cat("=== Longitudinal: does workload slope vary across employees? ===\n")
anova(model_ri_long_ml, model_rs_long_ml)

model_ri_exp_ml  <- lmer(decision_quality ~ condition + (1 | participant),
                          data = df_exp, REML = FALSE)
model_rs_exp_ml  <- lmer(decision_quality ~ condition + (condition | participant),
                          data = df_exp, REML = FALSE)

cat("\n=== Experimental: does scarcity effect vary across participants? ===\n")
anova(model_ri_exp_ml, model_rs_exp_ml)
```

A significant LRT ($p < .05$) indicates the random slope accounts for meaningful variance and should be retained. Use `REML = TRUE` for final parameter estimation once the model structure is settled.
:::

::: {.callout-warning icon=false}
## Assumption 3: Convergence

**What it means:** Complex random-slope models — especially with multiple predictors or crossed random effects — frequently fail to converge. A convergence warning from `lme4` signals the optimization algorithm did not find a stable solution. Estimates from non-converged models should not be trusted or reported.

**How to check:** Use `check_convergence()` from the `performance` package.

```{r convergence-check}
check_convergence(model_rs_long)
check_convergence(model_rs_exp)
```

**What to do if convergence fails:**

- Center and scale continuous predictors
- Remove the intercept-slope correlation (see zero-correlation parameterization above)
- Try alternative optimizers: `control = lmerControl(optimizer = "bobyqa")`
- Simplify the random effects structure if the LRT supports it (non-significant random slope variance)
- Consider Bayesian estimation via `brms`, which handles complex random-effects structures more robustly
:::

## Model Comparison — Do You Need Random Slopes?

Add random slopes when:

1. **Theory predicts heterogeneous effects.** In organizational research, employee-level differences in workload sensitivity are plausible and interesting. In experiments, individual differences in response to priming manipulations are well-documented.

2. **The LRT supports them.** The model with random slopes fits significantly better than the model without (at `REML = FALSE`).

3. **The design mandates them.** In within-subjects experimental designs, random slopes for within-subjects conditions are structurally required to account for heterogeneous treatment effects. Omitting them is not conservative — it inflates Type I error (Barr et al., 2013).

4. **Sample size is adequate.** Estimating a random slope requires sufficient variation in the predictor *within each unit*. Longitudinally: each employee must have multiple waves with varying workload. Experimentally: each participant must complete trials in multiple conditions.

```{r model-comparison-table}
bind_rows(
  broom.mixed::glance(model_ri_long_ml) |> mutate(model = "RI only",    setting = "Longitudinal"),
  broom.mixed::glance(model_rs_long_ml) |> mutate(model = "RI + Slope", setting = "Longitudinal"),
  broom.mixed::glance(model_ri_exp_ml)  |> mutate(model = "RI only",    setting = "Experimental"),
  broom.mixed::glance(model_rs_exp_ml)  |> mutate(model = "RI + Slope", setting = "Experimental")
) |>
  select(setting, model, AIC, BIC, logLik) |>
  knitr::kable(digits = 1,
               caption = "Model fit: random intercept vs. random intercept + slope")
```

## Recommended Reading

| Topic | Citation |
|-------|----------|
| Maximal random effects in experiments | Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language, 68*(3), 255–278. |
| Random slopes and cross-level interactions | Heisig, J. P., & Schaeffer, M. (2019). Why you should always include a random slope for the lower-level variable involved in a cross-level interaction. *European Sociological Review, 35*(2), 258–279. |
| Random slopes in experimental psychology | Judd, C. M., Westfall, J., & Kenny, D. A. (2012). Treating stimuli as a random factor in social psychology: A new and compelling case for mixed models. *Journal of Personality and Social Psychology, 103*(1), 54–69. |
| Practical guide to `lme4` | Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting linear mixed-effects models using lme4. *Journal of Statistical Software, 67*(1), 1–48. |
